### 2021-02-05 Manuscript.Rmd
Initial commit

---
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: no
    number_sections: false
    keep_tex: true
    includes:
      in_header: "preamble.tex"
bibliography: "references.bib"
abstract: The abstract (but it could also be a normal section).
title: "Paper title"
author:
  - Author 1^[Institution 1, email1]
  - Author 2^[Institution 2, email2]
---

```{r setup, echo=FALSE}
```

# Introduction

This paper is neat [@Ragsdale2020-hv].

@Ragsdale2020-hv is still a neat paper if we cite it like this.

```{r figure, echo=F, fig.cap="A figure", fig.align="center", out.width="50%"}
knitr::include_graphics("Model.png")
```

See Figure \@ref(fig:figure).

See Supplementary Figure \@ref(fig:figure).

\clearpage
\newpage

# References

### 2021-02-20 Manuscript.Rmd
Draft and analyses started!

@@
       in_header: "preamble.tex"
+    citation_package: natbib
 bibliography: "references.bib"
-abstract: The abstract (but it could also be a normal section).
-title: "Paper title"
+biblio-style: "genetics"
+title: Can we distinguish modes of selective interactions using linkage disequilibrium?
 author:
-  - Author 1^[Institution 1, email1]
-  - Author 2^[Institution 2, email2]
+header-includes:
+- |
+  ```{=latex}
+  \usepackage{authblk}
+  \author{Aaron P. Ragsdale \\ aaronpeaceragsdale@gmail.com}
+  \affil{Department of Human Genetics, McGill University, Montreal, Canada \\
+         National Laboratory of Genomics for Biodiversity, Irapuato, Mexico \\
+         }
+  ```
+date: \today
 ---
@@
 
+\begin{abstract} Selection acting on a mutation interferes and interacts with
+evolutionary processes at nearby loci, causing allele frequency and correlation
+patterns between pairs of selected mutations to deviate from expected
+single-locus dynamics. A number of recent studies have used patterns of linkage
+disequilibrium between selected variants to test for selective interference and
+epistatic interactions, with some disagreement over interpretation of
+observations. Interpreting this data is hindered by the relative lack of
+analytic or even numerical expectations for patterns of variation between pairs
+of loci under the combined effects of selection, dominance, epistasis, and
+demography. Here, I develop a numerical approach to compute the expected
+two-locus sampling distribution under diploid selection, with arbitrary
+epistasis and dominance, and variable recombination and population sizes. I use
+this to explore how different models of epistasis and dominance affect expected
+signed LD, including for non-steady-state demography relevant to human
+populations. Finally, I use whole-genome sequencing data to asses how well we
+can differentiate models of selective interference in practice, and find [[that
+while xxx, within classes/domains, we see...]]. \end{abstract}
+
 # Introduction
 
-This paper is neat [@Ragsdale2020-hv].
+Most new mutations that affect fitness are deleterious and tend to be
+eliminated from a population. The average number of generations that a
+deleterious mutation segregates in a population depends on the strength of
+selection against genomes that carry it, with very damaging mutations kept at
+low frequencies in the population and purged relatively rapidly. But in the
+time between mutation and fixation or loss, selected variants, both beneficial
+and damaging, can dramatically impact patterns of variation in nearby linked
+regions. This distortion away from neutral expectations has been well
+documented in practically every taxon and population for which we have genomic
+data. However, questions remain about the primary mode of interactions between
+multiple linked variants and their joint effects on genome-wide patterns of
+diversity.
 
-@Ragsdale2020-hv is still a neat paper if we cite it like this.
+In their foundational paper, @Hill1966-gv recognized that linked negatively
+selected variants reciprocally impede the efficacy of population to remove
+those mutations. In general, linked selection reduces the fixation probability
+of advantageous mutations and increases that of deleterious mutations, compared
+to expectations under single-locus dynamics [@Birky1988-jm]. Allele frequency
+dynamics of linked selected variants are also predicted to deviate from
+expectations without interference. Under a multiplicative fitness model, where
+the fitness reduction of a haplotype carrying multiple deleterious variants is
+equal to the product of the fitness reduction of each mutation independently,
 
-```{r figure, echo=F, fig.cap="A figure", fig.align="center", out.width="50%"}
-knitr::include_graphics("Model.png")
+One (!!) paragraph on epistasis, dominance, etc and overall importance. Thus,
+there is a lot of interest in disentangling and distinguishing between these
+processes.
+
+## Empirical observations
+
+The most direct way to test for interactions between linked selected variants
+is through deep mutation scanning experiments, in which many distinct mutations
+are induced within a target gene and fitness, or some protein function, is
+experimentally measured [@Romero2009-yi; @Steinberg2016-is]. For example, using
+the model system of the TEM-1 $\beta$-lactamase gene in *E. coli*,
+@Bershtein2006-bk found evidence for synergistic epistasis, where multiple
+deleterious mutations had a greater effect on fitness than the multiplication
+of the observed effects of the individual mutations. The scale of mutation
+scanning experiments continues to improve dramatically, promising greater
+resolution of the fitness landscape in such model systems.
+
+In most natural populations, directed mutation studies are not possible, and we
+must turn to population genetic approaches to infer selective interactions
+between observed mutations. Motivated by theory that linked negatively selected
+mutations will display negative LD due to interference [@Hill1966-gv], and that
+epistasis will drive expected LD away from zero, a number of recent studies
+have used patterns of LD within classes of putatively selected variants to
+infer modes of selective interactions. In a notable study from
+@Callahan2011-ac, pairs of nonsynonymous mutations were found to cluster more
+than expected along lineages in the *Drosophilid* species complex, and that
+those clustered mutations tended to preserve the charge of the protein and were
+in positive LD compared to pairs including synonymous mutations, suggesting
+that compensatory nonsynonymous variants were more generally tolerated.
+
+More recently, @Sohail2017-zq observed negative LD between loss-of-function
+variants in protein-coding genes (such as stop gains and losses, frameshifts,
+and other nonsense mutations) in both human and fruit fly populations, from
+which they proposed that widespread synergistic epistasis between these
+mutations. Within the past year or two, both @Sandler2020-vk and @Garcia2020-zn
+have reevaluated patterns of LD between coding variants in humans, fruit flies,
+and *Capsella grandiflora*, and suggested interference and dominance may
+instead be driving patterns of LD [@Garcia2020-zn] or questioned whether LD
+between loss-of-function variants is significantly different from zero
+[@Sandler2020-vk].
+
+A number of factors impede our interpretation of patterns of signed LD between
+coding variants. First, for strongly deleterious or loss-of-function variants,
+their rarity and low frequency means that statistical measurement of LD and
+other diversity measures are quite noisy. Second, comparisons are based on
+theory that are confined to limiting and simplistic assumptions, including
+steady-state demography, simple selection and interaction models, or unlinked
+loci. To generate predictions under more complex models, we rely on expensive
+forward simulations. These can be great for building intuition or testing
+inference methods, but do not efficiently provide expectations for quantities
+of interest across parameter regimes of interest. Theoretical and numerical
+studies of haplotype frequencies and LD under general selective interaction
+models would be of great benefit.
+
+## Existing theory and numerical methods
+
+In addition to impacting surrounding neutral variation, selected variants can
+interact and interfere with selection acting for or against other mutations. In
+their foundational paper, @Hill1966-gv recognized that linked negatively
+selected variants impede the efficacy of a population to remove those
+mutations. Generally, linked selection reduces the fixation probability of
+advantageous mutations and increases that of deleterious mutations, compared to
+expectations under single-locus dynamics [@Birky1988-jm; @Barton1995-mj]. The
+joint allele frequency dynamics of linked variants under selection are also
+predicted to deviate from expectations without interference. Since haplotypes
+that carry multiple deleterious mutations have lower fitness than haplotypes
+that carry just one or the other under a multiplicative fitness model, we
+should expect to see those mutation segregate on different haplotypes more
+often than together, leading to negative, or repulsion, linkage disequilibrium
+(LD), although the extent of LD depends non-trivially on the strength of
+selection and probability of recombination separating loci. Nonetheless, this
+provides a simple, testable hypothesis under the standard selection model of
+additive effects within loci and multiplicative fitness between: we should
+observe average signed LD to be negative between pairs of selected variants.
+
+Aside from selective interference under a multiplicative fitness model, where
+the relative fitness conferred by a haplotype carrying two deleterious
+mutations is equal to the product of the fitnesses of haplotypes carrying
+individual mutations, considerable attention has been given to
+non-multiplicative deviations in the fitness function, known as epistasis.
+Early work focused primarily on mutations of strong effect and explored models
+of both synergistic and diminishing returns (or antagonistic or threshold)
+epistasis under sexual and asexual reproduction, often in the limiting case of
+free recombination between deleterious variants. Generally, for sexually
+reproducing organisms, deleterious load is decreased under simple models of
+synergistic epistasis compared to multiplicative fitness [@Kimura1966-gc;
+@Kondrashov1995-va]. Pervasive epistasis has been invoked as an explanation for
+the evolutionary advantage of sex [@Kondrashov1982-sf; @Charlesworth1990-kw;
+@Barton1998-lu], as well as driving incompatibilities that lead to postzygotic
+isolation during the process of speciation [@Turelli2000-kz]. Within
+populations, epistasis is known to cause signed LD to deviate dramatically from
+zero, with negative synergistic epistasis leading to an excess of negative LD
+[@Kondrashov1982-sf; @Charlesworth1990-kw].
+
+While epistasis is the most commonly considered nonlinear effect influencing
+interactions between selected variants, non-additive effects *within* a
+locus may also cause deviations from our baseline models, either with or
+without epistasis. Dominance and multi-locus interactions are jointly important
+in shaping the expected equilibrium allele frequencies and mutation load due to
+strongly damaging disease mutations [@Clark1998-kq]. However, despite dominance
+and partial recessiveness being appreciated as important factors shaping linked
+variation [@Turelli2000-kz; @Zhao2016-bb], dominance has received less
+attention than epistasis in multi-locus selection models, and its impact on the
+joint segregation of negatively selected variants is poorly understood. There
+is ample evidence that a large fraction of selected mutations are at least
+partially recessive in humans [@xx], flies [@xx], and other species [@xx], and
+that average levels of dominance vary with the strength of selection [@xx].
+
+- Patterns of dominance
+  - @Huber2018-cp, @Haldane1930-tt, @Kacser1981-nc
+
+### @Ohta1969-ie and @Ohta1971-yd
+
+- Diffusion models for the neutral two-locus decay of LD $\sigma_d^2$
+
+### @Hallgrimsdottir2008-sq
+
+-
+
+### @Good2020-wj
+
+-
+
+- Neutrality, additive, dominance, epistasis, general diploid selection models
+- Analytic results: Older stuff, like some basic results from HR, through more
+  recent results in @Good2020-wj
+  - recursions under neutrality [@Golding1984-pu; @Hudson2001-sg]
+- Recent numerical approaches [@Ragsdale2017-gg; @Ragsdale2019-nt] and others that aren't me?
+- Resorting to discrete forward simulations (a few examples)
+
+In this paper, I develop a numerical approach to solve for the two-locus
+sampling distribution under a general diploid selection model with variable
+recombination and single-population size history. I use this model to describe
+how epistasis and dominance shape expected patterns of signed LD, under both
+steady-state and non-equilibrium demography, that have been used to test for
+interference and epistasis in population genomic data. I then turn to human
+sequencing data and compare patterns of LD for synonymous, missense, and
+loss-of-function mutations within and between protein-coding regions, and
+assess how well we can expect to discriminate between modes of selective
+interactions from genome-wide patterns of LD.
+
+# Methods
+
+## The two-locus sampling distribution with arbitrary selection
+
+### Drift and recombination
+
+### Selection models with epistasis and dominance
+
+### Low-order summaries of the sampling distribution
+
+## Numerical solution
+
+### Moment closure approach
+
+### Implementation
+
+- Numerical approach for solving the two-locus model with recombination,
+  single-population demography, and general selection models, following
+  @Jouganous2017-pq and @Ragsdale2019-nt.
+
+### Validation
+
+### Expected LD under inferred human demographic history
+
+## Analysis of human genomic data
+- Computing LD statistics commonly used for inference about selection models
+  (Figure \@ref(fig:fig1)), using @Ragsdale2020-nz
+- Data analysis of Thousand Genomes Project data
+  @1000_Genomes_Project_Consortium2015-zq
+- Domain information from @Stanek2020-pa
+
+**idea**: LD within genes partitioned by whether two mutations fall within
+domains or are in different domains (distance as a proxy? or can we get
+information about whether they lie in the same gene or not?)
+
+## Data and software availability
+
+- Thousand genomes VCFs and ancestral sequence
+  [@1000_Genomes_Project_Consortium2015-zq]
+- Protein domain information [@Stanek2020-pa]
+- Implementation as \texttt{moments.TwoLocus} at
+  \url{https://bitbucket.org/simongravel/moments}
+- Scripts to recreate analyses, figures, and compile this manuscript at
+  \url{https://github.com/apragsdale/two_locus_selection}
+
+# Results
+
+## Singed LD under steady-state demography
+
+- Expected LD under models of epistasis and dominance
+  at equilibrium (Figures \@ref(fig:fig2) and \@ref(fig:fig3))
+  - $\sigma_d^1$, $\sigma_d^2$, $r^2$, for all frequencies
+  - The same statistics conditioned on sample allele counts
+  - Relationship between "Hudson slice" and allele count-conditioned LD
+
+### Additive selection and epistasis
+
+### Simple dominance within a locus
+
+### General selection and gene-based dominance
+
+## Population size changes [[do what]]
+
+- Effect of demography on these same statistics
+  - Bottleneck with recovery
+  - Human-like size histories (using @Speidel2019-nj)
+
+## Data... main result in title
+
+- Present data from loss-of-function (LOF), synonymous, and nonsynonymous
+  variants from Thousand Genomes, both for pairs of loci within genes and for
+  pairs of loci between nearby genes. Do we see any differences, and do the
+  different patterns in African vs non-African populations tell us anything
+  about the possible selection model for different categories of variants?
+
+# Discussion
+
+- What can we say about epistasis and models of dominance using LD statistics?
+  Do we have power to distinguish models using low-order statistics of this
+  kind?
+- The role of numerical solutions to such models for population genetics
+  analyses. They bridge the gap between theory limited by analytic
+  intractability and expensive forward simulations such as \texttt{SLiM} and
+  \texttt{fwdpy11}.
+- Limitations and future directions:
+  - multiple populations
+  - strong selection leads to numerical instability
+
+# Acknowledgements
+- Simon Gravel
+- Kevin Thornton
+
+\clearpage
+\newpage
+
+# Figures:
+
+```{r fig1, echo=F, fig.cap="(ref:fig1cap)", fig.align="center"}
+knitr::include_graphics("figures/fig1.pdf")
 ```
 
-See Figure \@ref(fig:figure).
+(ref:fig1cap) \textbf{Additive selection and allele count-conditioned LD.} (A
+and B) The distribution of $AB$ haplotype counts in a sample size of 30, in
+which we observe 10 $A$ alleles at the left locus, and 10 $B$ alleles at the
+right locus. (C and D) The decay of normalized squared LD ($\sigma_d^2$) with
+scaled recombination distance for pairs of neutral and selected variants,
+respectively. The dashed line is the neutral expectation [@Ohta1971-yd]. (E and
+F) Similarly, the decay of $\sigma_d^1=\mathbb{E}[D]/ \mathbb{E}[p(1-p)q(1-q)]$
+for neutral and selected variants, respectively. Note the difference in scale
+between the two panels.
 
-See Supplementary Figure \@ref(fig:figure).
+```{r fig2, echo=F, fig.cap="(ref:fig2cap)", fig.align="center"}
+knitr::include_graphics("figures/fig2.pdf")
+```
+
+(ref:fig2cap) \textbf{Additive selection with epistasis.} (A) $D^2$, and (B) $D$.
+
+```{r fig3, echo=F, fig.cap="(ref:fig3cap)", fig.align="center"}
+knitr::include_graphics("figures/fig3.pdf")
+```
+
+(ref:fig3cap) \textbf{Dominance.} (A) $D^2$, and (B) $D$.
+
+4. Effects of demography on LD (bottleneck and recovery vs expansion, following
+   Tennessen model)
+
+5. LOF, missense, and synonymous variants, within and between genes.
+
+6. Missense variants in genes partitioned by conservation? Tolerance to
+   nonsynonymous changes?
 

### 2021-02-20 SupportingInformation.Rmd
Draft and analyses started!

---
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: no
    number_sections: false
    keep_tex: true
    includes:
      in_header: "sipreamble.tex"
bibliography: "references.bib"
title: "Paper title"
author:
  - Author 1^[Institution 1, email1]
  - Author 2^[Institution 2, email2]
---

```{r sisetup, echo=F}
options(tinytex.clean = FALSE)
```

# Supporting Information

\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thetable}{S\arabic{table}}
\setcounter{figure}{0}
\setcounter{table}{0}

\clearpage
\newpage

# Supporting References

### 2021-02-20 analysis/compute_LD_phased.py

@@ -1,0 +1,327 @@
+import numpy as np
+import gzip
+import moments.LD
+import sys
+import pickle
+
+
+lof_annotations = [
+    "frameshift_variant",
+    "splice_acceptor_variant",
+    "splice_donor_variant",
+    "start_lost",
+    "stop_gained",
+    "stop_lost",
+    "transcript_ablation",
+]
+
+import tarfile
+
+try:
+    print("already loaded ancestral seqs", len(ancestral_seqs))
+except NameError:
+    ancestral_seqs = {}
+    print("reading ancestral sequences")
+
+    tar = tarfile.open(
+        "/home/aaronragsdale/Data/1000G/human_ancestor/human_ancestor_GRCh37_e59.tar.bz2",
+        mode="r:bz2",
+    )
+    for member in tar.getmembers():
+        if member.isfile() is False:
+            # print(member)
+            continue
+        if member.get_info()["name"].endswith("fa"):
+            f = tar.extractfile(member)
+            ## can't get SeqIO.parse to read in the extracted file
+            lines = f.readlines()
+            chrom = lines[0].decode().split(":")[2]
+            print("found chrom", chrom)
+            ancestral_seqs[chrom] = "".join([l.decode().strip() for l in lines[1:]])
+            # for record in SeqIO.parse(x, "fasta"):
+            #    chrom = record.id.split(":")[2]
+            #    print(chrom)
+            #    ancestral_seqs[chrom] = record.seq
+    tar.close()
+
+
+def get_annotations(annot_file):
+    annots = {}  # pos: alt: (gene, csq)
+    for line in open(annot_file, "r"):
+        chrom, pos, ref, alt, gene, csq = line.split()
+        annots.setdefault(pos, {})
+        annots[pos][alt] = (gene, csq)
+    return annots
+
+
+def flip_gts(gts):
+    gts_out = []
+    for gt in gts:
+        if gt == "0|0":
+            gts_out.append("1|1")
+        elif gt == "0|1":
+            gts_out.append("1|0")
+        elif gt == "1|0":
+            gts_out.append("0|1")
+        elif gt == "1|1":
+            gts_out.append("0|0")
+        else:
+            raise ValueError("unexpected genotype", gt)
+    return gts_out
+
+
+def to_haplotypes(gts):
+    h = np.zeros(2 * len(gts))
+    for ii, g in enumerate(gts):
+        h[2 * ii] = int(g[0])
+        h[2 * ii + 1] = int(g[2])
[…]

### 2021-02-20 analysis/compute_LD_unphased.py

@@ -1,0 +1,326 @@
+import numpy as np
+import gzip
+import moments.LD
+import sys
+import pickle
+
+
+lof_annotations = [
+    "frameshift_variant",
+    "splice_acceptor_variant",
+    "splice_donor_variant",
+    "start_lost",
+    "stop_gained",
+    "stop_lost",
+    "transcript_ablation",
+]
+
+import tarfile
+
+try:
+    print("already loaded ancestral seqs", len(ancestral_seqs))
+except NameError:
+    ancestral_seqs = {}
+    print("reading ancestral sequences")
+
+    tar = tarfile.open(
+        "/home/aaronragsdale/Data/1000G/human_ancestor/human_ancestor_GRCh37_e59.tar.bz2",
+        mode="r:bz2",
+    )
+    for member in tar.getmembers():
+        if member.isfile() is False:
+            # print(member)
+            continue
+        if member.get_info()["name"].endswith("fa"):
+            f = tar.extractfile(member)
+            ## can't get SeqIO.parse to read in the extracted file
+            lines = f.readlines()
+            chrom = lines[0].decode().split(":")[2]
+            print("found chrom", chrom)
+            ancestral_seqs[chrom] = "".join([l.decode().strip() for l in lines[1:]])
+            # for record in SeqIO.parse(x, "fasta"):
+            #    chrom = record.id.split(":")[2]
+            #    print(chrom)
+            #    ancestral_seqs[chrom] = record.seq
+    tar.close()
+
+
+def get_annotations(annot_file):
+    annots = {}  # pos: alt: (gene, csq)
+    for line in open(annot_file, "r"):
+        chrom, pos, ref, alt, gene, csq = line.split()
+        annots.setdefault(pos, {})
+        annots[pos][alt] = (gene, csq)
+    return annots
+
+
+def flip_gts(gts):
+    gts_out = []
+    for gt in gts:
+        if gt == "0|0":
+            gts_out.append("1|1")
+        elif gt == "0|1":
+            gts_out.append("1|0")
+        elif gt == "1|0":
+            gts_out.append("0|1")
+        elif gt == "1|1":
+            gts_out.append("0|0")
+        else:
+            raise ValueError("unexpected genotype", gt)
+    return gts_out
+
+
+def to_genotypes(gts):
+    g = np.zeros(len(gts))
+    for ii, gg in enumerate(gts):
+        g[ii] = int(gg[0]) + int(gg[2])
+    return g
+
+
[…]

### 2021-02-20 analysis/compute_LD_unphased_domains.py

@@ -1,0 +1,336 @@
+import numpy as np
+import gzip
+import moments.LD
+import sys
+import pickle
+
+
+lof_annotations = [
+    "frameshift_variant",
+    "splice_acceptor_variant",
+    "splice_donor_variant",
+    "start_lost",
+    "stop_gained",
+    "stop_lost",
+    "transcript_ablation",
+]
+
+import tarfile
+
+try:
+    print("already loaded ancestral seqs", len(ancestral_seqs))
+except NameError:
+    ancestral_seqs = {}
+    print("reading ancestral sequences")
+
+    tar = tarfile.open(
+        "/home/aaronragsdale/Data/1000G/human_ancestor/human_ancestor_GRCh37_e59.tar.bz2",
+        mode="r:bz2",
+    )
+    for member in tar.getmembers():
+        if member.isfile() is False:
+            # print(member)
+            continue
+        if member.get_info()["name"].endswith("fa"):
+            f = tar.extractfile(member)
+            ## can't get SeqIO.parse to read in the extracted file
+            lines = f.readlines()
+            chrom = lines[0].decode().split(":")[2]
+            print("found chrom", chrom)
+            ancestral_seqs[chrom] = "".join([l.decode().strip() for l in lines[1:]])
+            # for record in SeqIO.parse(x, "fasta"):
+            #    chrom = record.id.split(":")[2]
+            #    print(chrom)
+            #    ancestral_seqs[chrom] = record.seq
+    tar.close()
+
+
+def get_annotations(annot_file):
+    annots = {}  # pos: alt: (gene, csq)
+    for line in open(annot_file, "r"):
+        chrom, pos, ref, alt, gene, csq = line.split()
+        annots.setdefault(pos, {})
+        annots[pos][alt] = (gene, csq)
+    return annots
+
+
+def flip_gts(gts):
+    gts_out = []
+    for gt in gts:
+        if gt == "0|0":
+            gts_out.append("1|1")
+        elif gt == "0|1":
+            gts_out.append("1|0")
+        elif gt == "1|0":
+            gts_out.append("0|1")
+        elif gt == "1|1":
+            gts_out.append("0|0")
+        else:
+            raise ValueError("unexpected genotype", gt)
+    return gts_out
+
+
+def to_genotypes(gts):
+    g = np.zeros(len(gts))
+    for ii, gg in enumerate(gts):
[…]

### 2021-02-20 analysis/data/supp/classify_coding_variants.py

@@ -1,0 +1,49 @@
+import gzip
+
+coding_annotations = [
+    "frameshift_variant",
+    "splice_acceptor_variant",
+    "splice_donor_variant",
+    "start_lost",
+    "stop_gained",
+    "stop_lost",
+    "transcript_ablation",
+    "missense",
+    "synonymous",
+]
+
+lof_annotations = coding_annotations[:-2]
+
+SNPs = ["A", "C", "G", "T"]
+
+for chrom in range(1, 23):
+    with gzip.open(
+        f"../../../../../Data/1000G/annotations/ALL.chr{chrom}.phase3_shapeit2_mvncall_integrated_v5.20130502.sites.annotation.vcf.gz",
+        "rb"
+    ) as fin:
+        with open(f"variants/coding_variant_consequences.chr{chrom}.txt", "w+") as fout:
+            for line in fin:
+                l = line.decode()
+                if not l.startswith("#"):
+                    coding = False
+                    for annot in coding_annotations:
+                        if annot in l:
+                            coding = True
+                    if coding:
+                        c, p, _, ref, alts, _, _, info = l.split()
+                        if ref not in SNPs:
+                            continue
+                        csqs = info.split("CSQ=")[1].split(";")[0].split(",")
+                        alts = alts.split(",")
+                        for alt in alts:
+                            if alt not in SNPs:
+                                continue
+                            found = False
+                            for csq in csqs:
+                                allele, gene, _, _, annot = csq.split("|")[:5]
+                                if allele == alt:
+                                    found = True
+                                    fout.write(f"{c}\t{p}\t{ref}\t{alt}\t{gene}\t{annot}\n")
+                            if not found:
+                                raise ValueError("uh oh")
+    print("finished with chromosome", chrom)

### 2021-02-20 analysis/data/supp/get_coding_variants.py

@@ -1,0 +1,31 @@
+import gzip
+
+coding_annotations = [
+    "frameshift_variant",
+    "splice_acceptor_variant",
+    "splice_donor_variant",
+    "start_lost",
+    "stop_gained",
+    "stop_lost",
+    "transcript_ablation",
+    "missense",
+    "synonymous",
+]
+
+for chrom in range(1, 23):
+    with gzip.open(
+        f"../../../../../Data/1000G/annotations/ALL.chr{chrom}.phase3_shapeit2_mvncall_integrated_v5.20130502.sites.annotation.vcf.gz",
+        "rb"
+    ) as fin:
+        with open(f"variants/coding_variants.chr{chrom}.txt", "w+") as fout:
+            for line in fin:
+                l = line.decode()
+                if not l.startswith("#"):
+                    coding = False
+                    for annot in coding_annotations:
+                        if annot in l:
+                            coding = True
+                    if coding:
+                        c, p = l.split()[:2]
+                        fout.write(f"{c}\t{p}\n")
+    print("finished with chromosome", chrom)

### 2021-02-20 analysis/data/supp/get_protein_domains.py

@@ -1,0 +1,43 @@
+# for each protein (ensembl ID), get list of regions based on prot2hg file
+from collections import defaultdict
+
+data = {}
+for i in range(1, 23):
+    data[i] = defaultdict(list)
+
+c = 0
+Ls = []
+
+chroms_to_keep = [str(i) for i in range(1, 23)]
+
+header = 0
+for line in open("prot2hg_1938_112019.csv", "r"):
+    if header == 0:
+        chr_col = line.split(";").index('"chrom"')
+        start_col = line.split(";").index('"chr_start"')
+        end_col = line.split(";").index('"chr_end"')
+        ensembl_col = line.split(";").index('"ensembl"')
+        type_col = line.split(";").index('"type"')
+        header = 1
+    else:
+        l = line.split(";")
+        t = l[type_col]
+        if t != '"Region"':
+            continue
+        chrom = l[chr_col]
+        chrom = chrom.split('"')[1][3:]
+        if chrom not in chroms_to_keep:
+            continue
+        start = int(l[start_col])
+        end = int(l[end_col])
+        ID = l[ensembl_col]
+        ID = ID.split('"')[1]
+        IDs = ID.split(",")
+        for ID in IDs:
+            data[int(chrom)][ID].append((start, end))
+        c += 1
+
+print("found", c, "domains")
+
+import pickle
+pickle.dump(data, open("domain_dict.bp", "wb+"))

### 2021-02-20 analysis/data/supp/subset_vcfs.sh

@@ -1,0 +1,25 @@
+#!/bin/bash
+
+# 1. for each population, subset to samples and coding variants
+# 2. trim alt alleles
+# 3. keep biallelic variable sites only within each population
+
+POP=$1
+
+echo "running for population ${POP}"
+
+# 1.
+for i in {1..22}; do echo $i; bcftools view ../../../../../Data/1000G/genotypes/ALL.chr${i}.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf.gz -R variants/coding_variants.chr${i}.txt -S samples/samples.${POP}.txt -v snps -Oz -o temp.${POP}.${i}.vcf.gz; bcftools index temp.${POP}.${i}.vcf.gz; done;
+
+# 2.
+echo "trimming alts"
+for i in {1..22}; do echo $i; bcftools view temp.${POP}.${i}.vcf.gz --trim-alt-alleles -c 1 -Oz -o temp.${POP}.2.${i}.vcf.gz; bcftools index temp.${POP}.2.${i}.vcf.gz; done;
+
+# 3.
+echo "keeping biallelic variable sites"
+for i in {1..22}; do echo $i; bcftools view temp.${POP}.2.${i}.vcf.gz -m2 -M2 -Oz -o ../vcfs/coding.${POP}.chr${i}.vcf.gz; done;
+
+# 4.
+echo "cleaning up temp.${POP} vcfs"
+rm temp.${POP}.*.vcf.gz*
+

### 2021-02-20 figures/fig1.py

@@ -1,0 +1,244 @@
+# Unconditioned vs allele count-conditioned LD.  Four panels, first two showing
+# Hudson slices in the with rho=1 and rho=10 for neutrality and with gamma=-5
+# at both loci. Second two panels show \sigma_d^2 and \sigma_d^1 for nuetrality
+# and using all SNPs with gamma=-5, as well as conditioning on doubletons,
+# tripletons, and few other frequencies.
+
+import moments.TwoLocus
+import matplotlib.pylab as plt
+import numpy as np
+import pickle
+
+# set font sizes
+import matplotlib
+
+plt.rcParams["legend.title_fontsize"] = "xx-small"
+matplotlib.rc("xtick", labelsize=6)
+matplotlib.rc("ytick", labelsize=6)
+matplotlib.rc("axes", labelsize=7)
+matplotlib.rc("axes", titlesize=7)
+matplotlib.rc("legend", fontsize=6)
+
+from bokeh.palettes import Colorblind
+
+colors = Colorblind[8]
+
+n0 = 60  # larger sample size for improved accuracy
+n = 30  # the projection size
+
+nA = 10
+nB = 10
+
+data_fname = "fig1_data.bp"
+try:
+    data = pickle.load(open(data_fname, "rb"))
+except IOError:
+    data = {}
+
+if "A" not in data.keys():
+    data["A"] = {}
+    # stores the nAB distribution with n=30, nA=10, nB=10 for nuetrality
+    # and gamma=-5, with rho=1 and rho=10
+    for rho in [1, 10]:
+        data["A"][rho] = {}
+        for gamma in [0, -5]:
+            sel_params = [2 * gamma, gamma, gamma]
+            F0 = moments.TwoLocus.Demographics.equilibrium(
+                n0, rho=rho, sel_params=sel_params
+            )
+            F = F0.project(n)
+            counts, pAB = moments.TwoLocus.Util.pAB(F, nA, nB)
+            data["A"][rho][gamma] = pAB
+            print("computed pAB for rho, gamma =", rho, ",", gamma)
+    pickle.dump(data, open(data_fname, "wb+"))
+
+if "C" not in data.keys():
[…]

### 2021-02-20 figures/fig2.py

@@ -1,0 +1,143 @@
+# Figure 2: Additive selection with epistasis.  In this epistasis model, sAB =
+# (1 + epsilon) * (sA + sB).  Smaller, single-column figure showing the effect
+# of synergistic epistasis for weakly and moderately selected variants.
+import sys
+import moments.TwoLocus
+import matplotlib.pylab as plt
+import numpy as np
+import pickle
+
+# set font sizes
+import matplotlib
+
+plt.rcParams["legend.title_fontsize"] = "xx-small"
+matplotlib.rc("xtick", labelsize=6)
+matplotlib.rc("ytick", labelsize=6)
+matplotlib.rc("axes", labelsize=7)
+matplotlib.rc("axes", titlesize=7)
+matplotlib.rc("legend", fontsize=6)
+
+from bokeh.palettes import Colorblind
+
+## got back and run with larger n0
+n0 = int(sys.argv[1])  # larger sample size for improved accuracy
+n = 30  # the projection size
+
+epsilons = [-0.5, 0.0, 0.5, 1.0]
+gammas = [-1, -10]
+conditions = ["all", 2, 3, 4, 10]
+
+data_fname = f"fig2_data_ns_{n0}_{n}.bp"
+try:
+    data = pickle.load(open(data_fname, "rb"))
+except IOError:
+    data = {}
+    # stores the sigma_d^2 and sigma_d^1 curves for all snps and
+    rhos = np.logspace(-2, np.log10(30), 20)
+    data["rhos"] = rhos
+    data["Ohta_Kimura"] = (5 + rhos / 2) / (11 + 13 * rhos / 2 + rhos ** 2 / 2)
+    for gamma in gammas:
+        for eps in epsilons:
+            print("Running with  gamma, eps = ", gamma, ",", eps, end=" ")
+            sel_params = moments.TwoLocus.Util.additive_epistasis(gamma, epsilon=eps)
+            data[(gamma, eps)] = {cond: {"sd1": [], "sd2": []} for cond in conditions}
+            for rho in rhos:
+                print(".", end="")
+                F0 = moments.TwoLocus.Demographics.equilibrium(
+                    n0, rho=rho, sel_params=sel_params
+                )
+                F = F0.project(n)
+                for cond in conditions:
+                    if cond == "all":
[…]

### 2021-02-20 figures/fig3.py

@@ -1,0 +1,138 @@
+# Figure 3: Dominance selection model.
+import sys
+import moments.TwoLocus
+import matplotlib.pylab as plt
+import numpy as np
+import pickle
+
+# set font sizes
+import matplotlib
+
+plt.rcParams["legend.title_fontsize"] = "xx-small"
+matplotlib.rc("xtick", labelsize=6)
+matplotlib.rc("ytick", labelsize=6)
+matplotlib.rc("axes", labelsize=7)
+matplotlib.rc("axes", titlesize=7)
+matplotlib.rc("legend", fontsize=6)
+
+from bokeh.palettes import Colorblind
+
+n0 = int(sys.argv[1])  # larger sample size for improved accuracy
+n = 30  # the projection size
+
+hs = [0.0, 0.1, 0.5]
+gammas = [-1, -5]
+conditions = ["all", 2, 3, 4, 10]
+
+data_fname = f"fig3_data_ns_{n0}_{n}.bp"
+try:
+    data = pickle.load(open(data_fname, "rb"))
+except IOError:
+    data = {}
+    data["ns"] = (n0, n)
+    # stores the sigma_d^2 and sigma_d^1 curves for all snps and
+    rhos = np.logspace(-2, np.log10(30), 20)
+    data["rhos"] = rhos
+    data["Ohta_Kimura"] = (5 + rhos / 2) / (11 + 13 * rhos / 2 + rhos ** 2 / 2)
+    for gamma in gammas:
+        for h in hs:
+            print("Running gamma, h = ", gamma, ",", h, end=" ", flush=True)
+            sel_params = moments.TwoLocus.Util.simple_dominance(gamma, h=h)
+            data[(gamma, h)] = {cond: {"sd1": [], "sd2": []} for cond in conditions}
+            for rho in rhos:
+                print(".", end="", flush=True)
+                F0 = moments.TwoLocus.Demographics.equilibrium(
+                    n0, rho=rho, sel_params_general=sel_params
+                )
+                F = F0.project(n)
+                for cond in conditions:
+                    if cond == "all":
+                        data[(gamma, h)][cond]["sd1"].append(F.D() / F.pi2())
+                        data[(gamma, h)][cond]["sd2"].append(F.D2() / F.pi2())
+                    else:
[…]

### 2021-02-20 figures/fig4.py

@@ -1,0 +1,382 @@
+# Figure 3: Demography and trajectories.
+import sys
+import moments.TwoLocus
+import matplotlib.pylab as plt
+import numpy as np
+import pickle
+
+# set font sizes
+import matplotlib
+
+plt.rcParams["legend.title_fontsize"] = "xx-small"
+matplotlib.rc("xtick", labelsize=6)
+matplotlib.rc("ytick", labelsize=6)
+matplotlib.rc("axes", labelsize=7)
+matplotlib.rc("axes", titlesize=7)
+matplotlib.rc("legend", fontsize=6)
+
+from bokeh.palettes import Colorblind
+
+try:
+    n = int(sys.argv[1])  # the sample size
+    gamma = float(sys.argv[2])
+    rho = float(sys.argv[3])
+except:
+    raise ValueError("inputs are n, gamma, rho in that order")
+
+def piecewise_constant(
+    nus,
+    Ts,
+    n,
+    rho=0.0,
+    gamma=0,
+    h=None,
+    eps=None,
+    track_stats=True,
+    spacing=3,
+    conditions=[],
+):
+    if track_stats:
+        stats = {"t": [], "sd1": {"all": [],}, "sd2": {"all": [],}}
+        for cond in conditions:
+            stats["sd1"][cond] = []
+            stats["sd2"][cond] = []
+    if gamma == 0:
+        sel_params = None
+        sel_params_general = None
+    elif h is None and eps is None:
+        sel_params = moments.TwoLocus.Util.additive_epistasis(gamma)
+        sel_params_general = None
+    elif h is not None:
+        sel_params = None
+        sel_params_general = moments.TwoLocus.Util.simple_dominance(gamma, h=h)
+    elif eps is not None:
+        sel_params = moments.TwoLocus.Util.additive_epistasis(gamma, epsilon=eps)
+        sel_params_general = None
+    F = moments.TwoLocus.Demographics.equilibrium(
+        n, rho=rho, sel_params=sel_params, sel_params_general=sel_params_general
+    )
+    if track_stats:
+        stats["t"].append(0)
[…]

### 2021-02-22 Manuscript.Rmd
ms edits and data figures

@@
 equal to the product of the fitness reduction of each mutation independently,
+we should expect to see those mutation segregate on different haplotypes more
+often than together, leading to negative, or repulsion, linkage disequilibrium
+(LD), although the extent of LD depends non-trivially on the strength of
+selection and probability of recombination separating loci [@Hill1966-gv].
 
-One (!!) paragraph on epistasis, dominance, etc and overall importance. Thus,
-there is a lot of interest in disentangling and distinguishing between these
-processes.
+Non-additive effects further complicate our evolutionary models, including
+dominance (i.e., interactions *within* loci) and epistasis (interactions
+*between* loci), and both are thought to be widespread phenomena. A large
+fraction of nonsynonymous coding mutations are thought to be at least partially
+recessive [@Huber2018-cp; @Agrawal2011-wb], with average levels of dominance
+correlating with strength of selection [@Kacser1981-nc]^[correct citation?],
+and dominance plays an important role in shaping expected equilibrium allele
+frequencies and mutation load of strongly damaging disease mutations
+[@Clark1998-kq].^[maybe also cite @Haldane1930-tt] On the other hand, epistasis
+differentially impacts deleterious load for asexually vs sexually reproducing
+organisms [@Kimura1966-gc; @Kondrashov1995-va] and has been invoked as an
+explanation for the evolutionary advantage of sex [@Kondrashov1982-sf;
+@Charlesworth1990-kw; @Barton1998-lu], and epistatic interactions may also
+drive incompatibilities that lead to postzygotic isolation during the process
+of speciation [@Turelli2000-kz]. Within populations, epistasis is known to
+cause signed LD to deviate dramatically from zero [@Kondrashov1995-va;
+@Charlesworth1990-kw]. However, despite patterns of dominance being appreciated
+as important factors shaping linked variation [@Turelli2000-kz; @Zhao2016-bb],
+and the evolutionary importance of epistatic interactions, we currently lack
+models for predicting correlation patterns between linked mutations under
+general selection models.^[See @McVean2000-ox].
+
+In this paper, I develop a numerical approach to solve for the two-locus
+sampling distribution under a general diploid selection model with variable
+recombination and single-population size history. I use this model to describe
+how epistasis and dominance shape expected patterns of signed LD, under both
+steady-state and non-equilibrium demography, that have been used to test for
+interference and epistasis in population genomic data. I then turn to human
+sequencing data and compare patterns of LD for synonymous, missense, and
+loss-of-function mutations within and between protein-coding genes and domains,
+and assess how well we can expect to discriminate between modes of selective
+interactions using genome-wide patterns of LD.
 
@@
 
-In addition to impacting surrounding neutral variation, selected variants can
-interact and interfere with selection acting for or against other mutations. In
-their foundational paper, @Hill1966-gv recognized that linked negatively
-selected variants impede the efficacy of a population to remove those
-mutations. Generally, linked selection reduces the fixation probability of
-advantageous mutations and increases that of deleterious mutations, compared to
-expectations under single-locus dynamics [@Birky1988-jm; @Barton1995-mj]. The
-joint allele frequency dynamics of linked variants under selection are also
-predicted to deviate from expectations without interference. Since haplotypes
-that carry multiple deleterious mutations have lower fitness than haplotypes
-that carry just one or the other under a multiplicative fitness model, we
-should expect to see those mutation segregate on different haplotypes more
-often than together, leading to negative, or repulsion, linkage disequilibrium
-(LD), although the extent of LD depends non-trivially on the strength of
-selection and probability of recombination separating loci. Nonetheless, this
-provides a simple, testable hypothesis under the standard selection model of
-additive effects within loci and multiplicative fitness between: we should
-observe average signed LD to be negative between pairs of selected variants.
+Many well-known properties of two-locus dynamics and equilibrium LD come from
+early work that centered on the multi-locus diffusion approximation
+[@Kimura1955-qe; @Hill1968-vu; @Ohta1969-ie; @Ohta1971-yd]. This includes the
+result that genome-wide averages of signed LD are expected to be zero under
+neutrality. Under a two-locus biallelic model, where the left locus allows
+alleles $A$ and $a$ and the right locus has alleles $B$ and $b$, the standard
+covariance measure of LD is defined as $D = f_{AB} - f_{A}f_{B}$ where $f_{AB}$
+is the haplotype frequency of double-derived types carrying both $A$ and $B$,
+and $f_{A}$ and $f_{B}$ are the marginal frequencies of the derived alleles at
+each locus. This covariance decays due to both drift and recombination at a
+rate proportional to the inverse of the effective population size and the
+distance separating loci: \[\mathbb{E}[D]_{t+1} = \left(1 - \frac{1}{2N_e(t)} -
+r \right)\mathbb{E}[D]_t.\] While $\mathbb{E}[D] = 0$, the variance of $D$ is
+non-zero, and @Ohta1971-yd presented their famed result that the variance of
+$D$ under neutrality and steady-state demography, normalized by the joint
+heterozygosity of the two loci, is \[\sigma_d^2 =
+\frac{\mathbb{E}[D^2]}{\mathbb{E}[f_A(1-f_A)f_B(1-f_B)]}\approx\frac{5 + \rho /
+2}{11 + 13\rho / 2 + \rho^2 / 2},\] where $\rho = 4N_e r$.
 
-Aside from selective interference under a multiplicative fitness model, where
-the relative fitness conferred by a haplotype carrying two deleterious
-mutations is equal to the product of the fitnesses of haplotypes carrying
-individual mutations, considerable attention has been given to
-non-multiplicative deviations in the fitness function, known as epistasis.
-Early work focused primarily on mutations of strong effect and explored models
-of both synergistic and diminishing returns (or antagonistic or threshold)
-epistasis under sexual and asexual reproduction, often in the limiting case of
-free recombination between deleterious variants. Generally, for sexually
-reproducing organisms, deleterious load is decreased under simple models of
-synergistic epistasis compared to multiplicative fitness [@Kimura1966-gc;
-@Kondrashov1995-va]. Pervasive epistasis has been invoked as an explanation for
-the evolutionary advantage of sex [@Kondrashov1982-sf; @Charlesworth1990-kw;
-@Barton1998-lu], as well as driving incompatibilities that lead to postzygotic
-isolation during the process of speciation [@Turelli2000-kz]. Within
-populations, epistasis is known to cause signed LD to deviate dramatically from
-zero, with negative synergistic epistasis leading to an excess of negative LD
-[@Kondrashov1982-sf; @Charlesworth1990-kw].
+Analytic progress beyond these well-known results has come haltingly. In the
+1980s, recursions were developed to compute the two-locus sampling distribution
+under neutrality, that is, the probability of observing given counts of
+two-locus haplotypes in a sample of size $n$ [@Golding1984-pu]. This approach
+would continue to be developed and later form the foundation for the inference
+of local recombination rates from population genetic data [@Hudson2001-sg;
+@McVean2004-gj].
 
-While epistasis is the most commonly considered nonlinear effect influencing
-interactions between selected variants, non-additive effects *within* a
-locus may also cause deviations from our baseline models, either with or
-without epistasis. Dominance and multi-locus interactions are jointly important
-in shaping the expected equilibrium allele frequencies and mutation load due to
-strongly damaging disease mutations [@Clark1998-kq]. However, despite dominance
-and partial recessiveness being appreciated as important factors shaping linked
-variation [@Turelli2000-kz; @Zhao2016-bb], dominance has received less
-attention than epistasis in multi-locus selection models, and its impact on the
-joint segregation of negatively selected variants is poorly understood. There
-is ample evidence that a large fraction of selected mutations are at least
-partially recessive in humans [@xx], flies [@xx], and other species [@xx], and
-that average levels of dominance vary with the strength of selection [@xx].
+To include selection, however, there have not been many advances beyond the
+Monte Carlo simulation approach taken by @Hill1966-gv, albeit with somewhat
+more powerful computational resources and more sophisticated software for
+performing flexible forward simulation (e.g., @Thornton2014-pn,
+@Haller2019-vm). Analytic results for two-locus distributions under selection
+are notoriously difficult, with a few notable flashes of progress.
+@McVean2007-xd considered the effect of a recent sweep on patterns of LD for
+loci nearby or spanning the locus under selection, and in an impressive recent
+paper, @Good2020-wj presented analytic solutions for patterns of LD between
+rare mutations under additive selection with epistasis. Nonetheless, such
+approaches are confined to steady-state demography and constrained selection
+models.
 
-- Patterns of dominance
-  - @Huber2018-cp, @Haldane1930-tt, @Kacser1981-nc
-
-### @Ohta1969-ie and @Ohta1971-yd
-
-- Diffusion models for the neutral two-locus decay of LD $\sigma_d^2$
-
-### @Hallgrimsdottir2008-sq
-
--
-
-### @Good2020-wj
-
--
-
-- Neutrality, additive, dominance, epistasis, general diploid selection models
-- Analytic results: Older stuff, like some basic results from HR, through more
-  recent results in @Good2020-wj
-  - recursions under neutrality [@Golding1984-pu; @Hudson2001-sg]
-- Recent numerical approaches [@Ragsdale2017-gg; @Ragsdale2019-nt] and others that aren't me?
-- Resorting to discrete forward simulations (a few examples)
-
-In this paper, I develop a numerical approach to solve for the two-locus
-sampling distribution under a general diploid selection model with variable
-recombination and single-population size history. I use this model to describe
-how epistasis and dominance shape expected patterns of signed LD, under both
-steady-state and non-equilibrium demography, that have been used to test for
-interference and epistasis in population genomic data. I then turn to human
-sequencing data and compare patterns of LD for synonymous, missense, and
-loss-of-function mutations within and between protein-coding regions, and
-assess how well we can expect to discriminate between modes of selective
-interactions from genome-wide patterns of LD.
+Numerical methods straddle the two approaches of expensive and noisy discrete
+simulation and analytic solutions, providing a more efficient and practical
+method to compute expectations of two-locus diversity measures under a wider
+range of parameters and demographic scenarios.^[cite other numerical
+approaches, and need to read: @Hallgrimsdottir2008-sq] @Ragsdale2017-gg solved
+the two-locus diffusion equation with additive selection at either locus,
+although that paper focused on the applicability of two-locus statistics to
+demographic inference. More recently, @Ragsdale2019-nt extended the
+@Hill1968-vu system for $\mathbb{E}[D^2]$ to compute arbitrary moments of the
+distribution of $D$ for any number of populations connected by migration and
+admixture. This paper also showed that such a moments-construction can be used
+to solve for the two-locus sampling distribution within a single population,
+though requires a moment-closure approximation for nonzero recombination and
+selection. In the [Methods], I extend this same approach to model arbitrary
+diploid selection, which encompasses dominance, epistasis, and other forms of
+selective interactions between two loci.
 
@@
 
+The two-locus sampling distribution is the direct analog to the single-locus
+site frequency spectrum (SFS) of a given sample size. Instead of describing the
+density or number of mutations with a given allele count out of $n$ samples,
+the two-locus distribution $\Psi_n$ stores the density or number of pairs of
+loci with observed haplotype counts, so that $\Psi_n(i, j, k)$ is the number
+of pairs where we see $i$ $AB$ haplotypes, $j$ $Ab$ haplotypes, and $k$ $aB$
+haplotypes. Here, we use upper case to denote the derived allele, so the number
+of doubly ancestral haplotypes $ab$ is $n-i-j-k$.
+
+- Diffusion equation for $\psi(x, y, z)$
+- Sampling to get $\Psi_n$
+- Point to @Ragsdale2019-nt supplement, with $\dot{\Psi}_n = \mathcal{D}\Psi_n \ldots$
+- Note that the selection operator
+
 ### Drift and recombination
@@
 
+### Moment closure and implementation
+
 ### Low-order summaries of the sampling distribution
 
-## Numerical solution
-
-### Moment closure approach
-
-### Implementation
-
-- Numerical approach for solving the two-locus model with recombination,
-  single-population demography, and general selection models, following
-  @Jouganous2017-pq and @Ragsdale2019-nt.
-
-### Validation
-
-### Expected LD under inferred human demographic history
+## Expected LD under inferred human demographic history
 
@@
 information about whether they lie in the same gene or not?)
+- see @Yeang2007-gj, @Ivankov2014-tn, @Taverner2020-lk
 
@@
 
-## Population size changes [[do what]]
+## The effect of population size changes on signed LD
 
 - Effect of demography on these same statistics
-  - Bottleneck with recovery
+
+  - Bottleneck with recovery and expansion
   - Human-like size histories (using @Speidel2019-nj)
+  - For each (in own column): \@ref(fig:fig4)
+    - Top panel: demography
+    - Next panel: additive selection and synergistic epistasis
+    - Next panel: dominance with h = 0, 0.2
+    - Next panel: gene-based dominance and compensatory mutations
 
-## Data... main result in title
+## Signed LD for mutations within and between genes
+
+## Large positive signed LD for missense mutations within annotated protein
+domains
 
@@
   variants from Thousand Genomes, both for pairs of loci within genes and for
-  pairs of loci between nearby genes. Do we see any differences, and do the
-  different patterns in African vs non-African populations tell us anything
-  about the possible selection model for different categories of variants?
+  pairs of loci between nearby genes \@ref(fig:fig5). Do we see any
+  differences, and do the different patterns in African vs non-African
+  populations tell us anything about the possible selection model for different
+  categories of variants?
+
+Some results:
+
+- Within genes, both synonymous and missense mutations have LD slightly greater
+  than zero, and are not significantly different from each other.
+- On the other hand, LOF mutations have negative LD, although the small number
+  and low frequency of LOF mutations in human populations means that the
+  measurement is quite noisy.
+- In fact, taking each population independently, all measurements have confidence
+  intervals that overlap with zero
+- But assuming independent measurements between populations (which is probably not
+  correct, as populations have shared history and thus shared diversity patterns),
+  can we say that synonymous and missense mutations have values significantly greater
+  than zero and LOF mutations have values less than zero?
+
+Within domains \@ref(fig:fig6):
+
+- Missense and synonymous mutations have positive LD when looking at pairs
+  within the same annotated domain.
+- Between domains, both are reduced to be roughly zero
+- Within domains, missense mutations have signed LD much larger than synonymous
+  mutations
 
@@
 
-(ref:fig3cap) \textbf{Dominance.} (A) $D^2$, and (B) $D$.
+(ref:fig3cap) **Dominance effects.** (A) $D^2$, and (B) $D$.
 
-4. Effects of demography on LD (bottleneck and recovery vs expansion, following
-   Tennessen model)
+```{r fig4, echo=F, fig.cap="(ref:fig4cap)", fig.align="center"}
+knitr::include_graphics("figures/fig4.pdf")
+```
 
-5. LOF, missense, and synonymous variants, within and between genes.
+(ref:fig4cap) **The effects of demography on signed LD.**
 
-6. Missense variants in genes partitioned by conservation? Tolerance to
-   nonsynonymous changes?
+```{r fig5, echo=F, fig.cap="(ref:fig5cap)", fig.align="center"}
+knitr::include_graphics("figures/fig5.pdf")
+```
+
+(ref:fig5cap) **LD within and between protein-coding genes.**
+
+```{r fig6, echo=F, fig.cap="(ref:fig6cap)", fig.align="center"}
+knitr::include_graphics("figures/fig6.pdf")
+```
+
+(ref:fig6cap) **LD within and between annotated domains in protein-coding
+genes.**
 

### 2021-02-22 SupportingInformation.Rmd
ms edits and data figures

@@
 
+- Tables with data presented in figures 5 and 6
+- Table with Tajima's D for syn, mis, lof mutations for all, within, or outside of domains
+
+\clearpage
+\newpage
+
+# Supplementary Tables
+
+```{r tabS1, echo=F, tab.align="center"}
+library(knitr)
+library(kableExtra)
+pops <- read.csv("tables/populations.txt", sep=";", header=TRUE)
+knitr::kable(pops, "latex", booktabs=T,
+    caption="Thousand Genomes Project population descriptions for populations used in this study.") %>%
+kable_styling(latex_options = "HOLD_position")
+```
+
+```{r tabS2, echo=F, tab.align="center"}
+library(knitr)
+library(kableExtra)
+taj <- read.csv("tables/tajimas_d.txt", sep=";", header=TRUE)
+knitr::kable(taj, "latex", booktabs=T, longtable=T,
+    linesep = c("", "", "", "", "", "", "", "", "\\addlinespace"),
+    caption="Tamija's $D$ for classes of coding mutations, both within annotated domains and outside of domains.") %>%
+kable_styling(latex_options = c("HOLD_position", "repeat_header"))
+```
+
+\clearpage
+\newpage
+
+# Supplementary Figures
+
+1) Accuracy of jackknife for varying sample size
+2) Repeat of figures 2, 3, and 4 but for average $D$ instead of $\sigma_d^2$
+3)
+
 \clearpage

### 2021-02-22 analysis/compute_LD_unphased.py

@@ -227,11 +227,13 @@ def compute_LD_by_distance(
     """
     genes_set = set(genes)
     bins = [(l, r) for l, r in zip(bp_bins[:-1], bp_bins[1:])]
-    stat_sums = {b: np.zeros(4) for b in bins}
-    num_pairs = {b: 0 for b in bins}
+    stat_sums = {b: {} for b in bins}
+    num_pairs = {b: {} for b in bins}
     for ii, (pos, gene) in enumerate(zip(positions, genes)):
         other_gene = genes != gene
         for b in bins:
+            stat_sums[b].setdefault(gene, np.zeros(4))
+            num_pairs[b].setdefault(gene, 0)
             to_keep = np.logical_and(
                 positions - positions[ii] > b[0], positions - positions[ii] <= b[1]
             )
@@ -243,11 +245,29 @@ def compute_LD_by_distance(
                 D2, Dz, pi2, D = moments.LD.Parsing.compute_pairwise_stats_between(
                     focal_haplotype[np.newaxis, :], compare_haplotypes, genotypes=True
                 )
-                stat_sums[b] += np.array(
+                stat_sums[b][gene] += np.array(
                     [np.sum(D2), np.sum(Dz), np.sum(pi2), np.sum(D)]
                 )
-                num_pairs[b] += len(compare_haplotypes)
+                num_pairs[b][gene] += len(compare_haplotypes)
     return stat_sums, num_pairs
+
+
+def bootstrap_over_genes(gene_data, gene_sets, reps=1000):
+    bs_sums = [np.zeros(4) for k in gene_sets["gene_sets"].keys()]
+    for gene in gene_data:
+        try:
+            bs_sums[gene_sets["gene_set_map"][gene]] += gene_data[gene]
+        except KeyError:
+            print(gene)
+            # not good...
+            bs_sums[-1] += gene_data[gene]
+
+    sigma_d1s = []
+    for i in range(reps):
+        choices = np.random.choice(range(len(bs_sums)), len(bs_sums))
+        bs_stats = np.sum([bs_sums[j] for j in choices], axis=0)
+        sigma_d1s.append(bs_stats[3] / bs_stats[2])
+    return np.std(sigma_d1s)
 
 
 if __name__ == "__main__":
@@ -257,15 +277,15 @@ if __name__ == "__main__":
[…]

### 2021-02-22 analysis/compute_LD_unphased_domains.py

@@ -276,6 +276,46 @@ def compute_LD_between_domains(positions, genes, G, dom):
     return gene_stat_sums, num_pairs_per_gene
 
 
+def bootstrap_over_genes(gene_data, gene_sets, reps=1000):
+    bs_sums = [np.zeros(4) for k in gene_sets["gene_sets"].keys()]
+    for gene in gene_data:
+        try:
+            bs_sums[gene_sets["gene_set_map"][gene]] += gene_data[gene]
+        except KeyError:
+            print(gene)
+            # not good...
+            bs_sums[-1] += gene_data[gene]
+
+    sigma_d1s = []
+    for i in range(reps):
+        choices = np.random.choice(range(len(bs_sums)), len(bs_sums))
+        bs_stats = np.sum([bs_sums[j] for j in choices], axis=0)
+        sigma_d1s.append(bs_stats[3] / bs_stats[2])
+    return np.std(sigma_d1s)
+
+
+def bootstrap_over_genes_meanD(gene_data, num_pairs, gene_sets, reps=1000):
+    bs_sums = [np.zeros(4) for k in gene_sets["gene_sets"].keys()]
+    num_sums = [0 for k in gene_sets["gene_sets"].keys()]
+    for gene in gene_data:
+        try:
+            bs_sums[gene_sets["gene_set_map"][gene]] += gene_data[gene]
+            num_sums[gene_sets["gene_set_map"][gene]] += num_pairs[gene]
+        except KeyError:
+            print(gene)
+            # not good...
+            bs_sums[-1] += gene_data[gene]
+            num_sums[-1] += num_pairs[gene]
+
+    mean_Ds = []
+    for i in range(reps):
+        choices = np.random.choice(range(len(bs_sums)), len(bs_sums))
+        bs_stats = np.sum([bs_sums[j] for j in choices], axis=0)
+        total_pairs = np.sum([num_sums[j] for j in choices])
+        mean_Ds.append(bs_stats[3] / total_pairs)
+    return np.std(mean_Ds)
+
+
 if __name__ == "__main__":
     POP = sys.argv[1]
 
@@ -333,4 +373,41 @@ if __name__ == "__main__":
         "num_pairs_between_domains": num_pairs_between,
     }
 
-    pickle.dump(outputs, open(f"parsed_data/{POP}.unphased.domains.bp", "wb+"))
[…]

### 2021-02-22 analysis/get_frqeuency_spectra.py

@@ -1,0 +1,233 @@
+# Get the frequency spectra for each mutation type, for all mutations
+# within a gene, for mutations falling in annotated domains, and for
+# mutations falling outside of annotated domains.
+
+import numpy as np
+import gzip
+import moments
+import sys
+import pickle
+
+lof_annotations = [
+    "frameshift_variant",
+    "splice_acceptor_variant",
+    "splice_donor_variant",
+    "start_lost",
+    "stop_gained",
+    "stop_lost",
+    "transcript_ablation",
+]
+
+import tarfile
+
+try:
+    print("already loaded ancestral seqs", len(ancestral_seqs))
+except NameError:
+    ancestral_seqs = {}
+    print("reading ancestral sequences")
+
+    tar = tarfile.open(
+        "/home/aaronragsdale/Data/1000G/human_ancestor/human_ancestor_GRCh37_e59.tar.bz2",
+        mode="r:bz2",
+    )
+    for member in tar.getmembers():
+        if member.isfile() is False:
+            # print(member)
+            continue
+        if member.get_info()["name"].endswith("fa"):
+            f = tar.extractfile(member)
+            ## can't get SeqIO.parse to read in the extracted file
+            lines = f.readlines()
+            chrom = lines[0].decode().split(":")[2]
+            print("found chrom", chrom)
+            ancestral_seqs[chrom] = "".join([l.decode().strip() for l in lines[1:]])
+            # for record in SeqIO.parse(x, "fasta"):
+            #    chrom = record.id.split(":")[2]
+            #    print(chrom)
+            #    ancestral_seqs[chrom] = record.seq
+    tar.close()
+
+
+def get_annotations(annot_file):
+    annots = {}  # pos: alt: (gene, csq)
+    for line in open(annot_file, "r"):
+        chrom, pos, ref, alt, gene, csq = line.split()
+        annots.setdefault(pos, {})
+        annots[pos][alt] = (gene, csq)
+    return annots
+
+
+def flip_gts(gts):
+    gts_out = []
+    for gt in gts:
+        if gt == "0|0":
+            gts_out.append("1|1")
+        elif gt == "0|1":
+            gts_out.append("1|0")
+        elif gt == "1|0":
+            gts_out.append("0|1")
+        elif gt == "1|1":
+            gts_out.append("0|0")
+        else:
+            raise ValueError("unexpected genotype", gt)
+    return gts_out
+
+
+def to_haplotypes(gts):
[…]

### 2021-02-22 figures/fig2.py

@@ -65,7 +65,7 @@ except IOError:
     pickle.dump(data, open(data_fname, "wb+"))
 
 
-fig = plt.figure(2, figsize=(3.25, 4.5))
+fig = plt.figure(2, figsize=(6.5, 4.5))
 fig.clf()
 
 rhos = data["rhos"]
@@ -73,9 +73,13 @@ ok = data["Ohta_Kimura"]
 
 markers = ["x", "+", "1", "2"]
 colors = Colorblind[8]
-ax1 = plt.subplot(2, 1, 1)
 
-ax1.plot(rhos, ok, "k--", lw=1.5, label=None)
+# top panels are sigma_d^2
+ax1 = plt.subplot(2, 2, 1)
+ax2 = plt.subplot(2, 2, 2)
+
+ax1.plot(rhos, ok, "k--", lw=1, label=None)
+ax2.plot(rhos, ok, "k--", lw=1, label=None)
 
 for ii, eps in enumerate(epsilons):
     ax1.plot(
@@ -85,59 +89,75 @@ for ii, eps in enumerate(epsilons):
         color=colors[ii],
         ms=4,
         lw=0.5,
-        label=rf"$\gamma={-1}, \epsilon={eps}$",
+        label=rf"$\epsilon={eps}$",
     )
 
 for ii, eps in enumerate(epsilons):
-    ax1.plot(
+    ax2.plot(
         rhos,
         data[(-10, eps)]["all"]["sd2"],
-        markers[ii] + ":",
+        markers[ii] + "-",
         color=colors[ii],
         ms=4,
         lw=0.5,
-        label=rf"$\gamma={-10}, \epsilon={eps}$",
+        label=rf"$\epsilon={eps}$",
     )
 
 ax1.set_xscale("log")
 ax1.set_yscale("log")
-#ax1.set_xlabel(r"$\rho$")
+ax2.set_xscale("log")
+ax2.set_yscale("log")
 ax1.set_ylabel(r"$\sigma_d^2$")
-#ax1.legend(frameon=False, ncol=2)
+ax1.legend()
+#ax2.legend()
+ax1.set_title(r"$\gamma = -1$")
+ax2.set_title(r"$\gamma = -10$")
 
-ax2 = plt.subplot(2, 1, 2)
+ax3 = plt.subplot(2, 2, 3)
+ax4 = plt.subplot(2, 2, 4)
+
+ax3.plot(rhos, 0 * ok, "k--", lw=1, label=None)
[…]

### 2021-02-22 figures/fig3.py

@@ -20,7 +20,7 @@ from bokeh.palettes import Colorblind
 n0 = int(sys.argv[1])  # larger sample size for improved accuracy
 n = 30  # the projection size
 
-hs = [0.0, 0.1, 0.5]
+hs = [0.0, 0.1, 0.5, 1.0]
 gammas = [-1, -5]
 conditions = ["all", 2, 3, 4, 10]
 
@@ -63,18 +63,19 @@ except IOError:
     pickle.dump(data, open(data_fname, "wb+"))
 
 
-fig = plt.figure(3, figsize=(3.25, 4.5))
+fig = plt.figure(3, figsize=(6.5, 4.5))
 fig.clf()
 
 rhos = data["rhos"]
 ok = data["Ohta_Kimura"]
 
-markers = ["x", "+", "1"]
+markers = ["x", "+", "1", "2"]
 colors = Colorblind[8]
 
-ax1 = plt.subplot(2, 1, 1)
+ax1 = plt.subplot(2, 2, 1)
+ax2 = plt.subplot(2, 2, 2)
 
-ax1.plot(rhos, ok, "k--", lw=1.5, label=None)
+ax1.plot(rhos, ok, "k--", lw=1, label=None)
 
 for ii, h in enumerate(hs):
     ax1.plot(
@@ -84,55 +85,60 @@ for ii, h in enumerate(hs):
         color=colors[ii],
         ms=4,
         lw=0.5,
-        label=rf"$\gamma={-1}, h={h}$",
+        label=rf"$h={h}$",
     )
-
-for ii, h in enumerate(hs):
-    ax1.plot(
-        rhos,
-        data[(-5, h)]["all"]["sd2"],
-        markers[ii] + ":",
-        color=colors[ii],
-        ms=4,
-        lw=0.5,
-        label=rf"$\gamma={-5}, h={h}$",
-    )
-
-ax2 = plt.subplot(2, 1, 2)
-
-ax2.plot(rhos, 0 * ok, "k--", lw=1.5, label=None)
 
 for ii, h in enumerate(hs):
     ax2.plot(
+        rhos,
+        data[(-5, h)]["all"]["sd2"],
+        markers[ii] + "-",
+        color=colors[ii],
+        ms=4,
+        lw=0.5,
+        label=rf"$h={h}$",
+    )
+
+ax3 = plt.subplot(2, 2, 3)
+ax4 = plt.subplot(2, 2, 4)
[…]

### 2021-02-22 figures/fig5.py

@@ -1,0 +1,194 @@
+# Figure 5: LD within and between genes
+import sys
+import moments.TwoLocus
+import matplotlib.pylab as plt
+import numpy as np
+import pickle
+
+# set font sizes
+import matplotlib
+
+plt.rcParams["legend.title_fontsize"] = "xx-small"
+matplotlib.rc("xtick", labelsize=6)
+matplotlib.rc("ytick", labelsize=6)
+matplotlib.rc("axes", labelsize=7)
+matplotlib.rc("axes", titlesize=7)
+matplotlib.rc("legend", fontsize=6)
+
+from bokeh.palettes import Colorblind
+
+data = {}
+
+bs = {}
+
+populations = {
+    "Africa": ["ESN", "GWD", "LWK", "MSL", "YRI"],
+    "Europe": ["CEU", "FIN", "GBR", "IBS", "TSI"],
+    "East Asia": ["CDX", "CHB", "CHS", "JPT", "KHV"],
+}
+
+for continent, pops in populations.items():
+    for pop in pops:
+        ld_pop = pickle.load(
+            open(f"../analysis/parsed_data/{pop}.unphased.within.between.bp", "rb")
+        )
+        data[pop] = ld_pop
+        bs_pop = pickle.load(
+            open(
+                f"../analysis/parsed_data/{pop}.unphased.within.between.sigmad1.bootstrap_std_err.bp",
+                "rb",
+            )
+        )
+        bs[pop] = bs_pop
+
+ms = 2
+lw = 0.5
+
+fig = plt.figure(5, figsize=(6.5, 4.5))
+fig.clf()
+
+markers = ["x", "+", "1"]
+colors = Colorblind[8]
+
+# plot all stats within genes
+ax1 = plt.subplot2grid((5, 4), (0, 0), rowspan=3, colspan=4)
+
+xx = np.concatenate(
+    (np.linspace(1, 15, 15), np.linspace(18, 32, 15), np.linspace(35, 49, 15))
+)
+yy = []
+err = []
+
+for continent in ["Africa", "Europe", "East Asia"]:
+    for pop in populations[continent]:
+        yy.append(
+            data[pop]["ld_within"]["synonymous"][3]
+            / data[pop]["ld_within"]["synonymous"][2]
+        )
+        yy.append(
+            data[pop]["ld_within"]["missense"][3]
[…]

### 2021-02-22 figures/fig6.py

@@ -1,0 +1,156 @@
+# Figure 6: LD within and between gene domains
+import sys
+import moments.TwoLocus
+import matplotlib.pylab as plt
+import numpy as np
+import pickle
+
+# set font sizes
+import matplotlib
+
+plt.rcParams["legend.title_fontsize"] = "xx-small"
+matplotlib.rc("xtick", labelsize=6)
+matplotlib.rc("ytick", labelsize=6)
+matplotlib.rc("axes", labelsize=7)
+matplotlib.rc("axes", titlesize=7)
+matplotlib.rc("legend", fontsize=6)
+
+from bokeh.palettes import Colorblind
+
+data = {}
+
+bs = {}
+
+populations = {
+    "Africa": ["ESN", "GWD", "LWK", "MSL", "YRI"],
+    "Europe": ["CEU", "FIN", "GBR", "IBS", "TSI"],
+    "East Asia": ["CDX", "CHB", "CHS", "JPT", "KHV"],
+}
+
+for continent, pops in populations.items():
+    for pop in pops:
+        ld_pop = pickle.load(
+            open(f"../analysis/parsed_data/{pop}.unphased.domains.bp", "rb")
+        )
+        data[pop] = ld_pop
+        bs_pop = pickle.load(
+            open(
+                f"../analysis/parsed_data/{pop}.unphased.domains.sigmad1.bootstrap_std_err.bp",
+                "rb",
+            )
+        )
+        bs[pop] = bs_pop
+
+ms = 2
+lw = 0.5
+
+fig = plt.figure(5, figsize=(6.5, 3))
+fig.clf()
+
+markers = ["x", "+", "1"]
+colors = Colorblind[8]
+
+# plot all stats within genes
+ax1 = plt.subplot(1, 1, 1)
+
+xx = np.concatenate(
+    (np.linspace(1, 20, 20), np.linspace(23, 42, 20), np.linspace(45, 64, 20))
+)
+yy = []
+err = []
+
+for continent in ["Africa", "Europe", "East Asia"]:
+    for pop in populations[continent]:
+        yy.append(
+            data[pop]["ld_within_domains"]["synonymous"][3]
+            / data[pop]["ld_within_domains"]["synonymous"][2]
+        )
+        yy.append(
+            data[pop]["ld_within_domains"]["missense"][3]
[…]

### 2021-02-22 tables/build_taj_d_table.py

@@ -1,0 +1,25 @@
+import pickle
+import moments
+import numpy as np
+
+pops = ["ESN", "GWD", "LWK", "MSL", "YRI", "CEU", "FIN", "GBR", "IBS", "TSI", "CDX", "CHB", "CHS", "JPT", "KHV"]
+
+with open("tajimas_d.txt", "w+") as fout:
+    fout.write("Population;Mutation type;Region;Tajima's $D$\n")
+    for pop in pops:
+        spectra = pickle.load(open(f"../analysis/parsed_data/{pop}.frequency_spectra.bp", "rb"))
+        Ds = {}
+        for annot in ["synonymous", "missense", "loss_of_function"]:
+            Ds[annot] = {}
+            for c in ["all", "in", "out"]:
+                Ds[annot][c] = moments.Spectrum(spectra[annot][c]).Tajima_D()
+        fout.write(f"{pop};Synonymous;All;{Ds['synonymous']['all']:.3f}\n")
+        fout.write(f";;In domain;{Ds['synonymous']['in']:.3f}\n")
+        fout.write(f";;Not in domain;{Ds['synonymous']['out']:.3f}\n")
+        fout.write(f";Missense;All;{Ds['missense']['all']:.3f}\n")
+        fout.write(f";;In domain;{Ds['missense']['in']:.3f}\n")
+        fout.write(f";;Not in domain;{Ds['missense']['out']:.3f}\n")
+        fout.write(f";Loss of function;All;{Ds['loss_of_function']['all']:.3f}\n")
+        fout.write(f";;In domain;{Ds['loss_of_function']['in']:.3f}\n")
+        fout.write(f";;Not in domain;{Ds['loss_of_function']['out']:.3f}\n")
+

### 2021-03-01 Manuscript.Rmd
edits throughout

--- 2021-02-05 Manuscript.Rmd
@@
       in_header: "preamble.tex"
+    citation_package: natbib
+    extra_dependencies: ["float"]
 bibliography: "references.bib"
-abstract: The abstract (but it could also be a normal section).
-title: "Paper title"
+biblio-style: "genetics"
+title: Can we distinguish modes of selective interactions using linkage disequilibrium?
+((or)) Mutation correlations in conserved domains reveal a heterogeneous landscape of selective interactions within genes ((or...?))
 author:
-  - Author 1^[Institution 1, email1]
-  - Author 2^[Institution 2, email2]
+header-includes:
+- |
+  ```{=latex}
+  \usepackage{authblk}
+  \author{Aaron P. Ragsdale \\ aaronpeaceragsdale@gmail.com}
+  \affil{Department of Human Genetics, McGill University, Montreal, Canada \\
+         National Laboratory of Genomics for Biodiversity, Irapuato, Mexico \\
+         }
+  ```
+date: \today
 ---
@@
 
+\begin{abstract}
+
+Selected mutations interfere and interact with evolutionary processes at nearby
+loci, distorting patterns of allele frequencies and correlations between pairs
+of mutations. A number of recent studies have used patterns of linkage
+disequilibrium between selected variants to test for selective interference and
+epistatic interactions, with some disagreement over interpretation of
+observations. Interpreting this data is hindered by the relative lack of
+analytic or even numerical expectations for patterns of variation between pairs
+of loci under the combined effects of selection, dominance, epistasis, and
+demography. Here, I develop a numerical approach to compute the expected
+two-locus sampling distribution under diploid selection, with arbitrary
+epistasis and dominance, and variable recombination and population sizes. I use
+this to explore how different models of epistasis and dominance affect expected
+signed LD, including for non-steady-state demography relevant to human
+populations. Finally, I use whole-genome sequencing data from humans to asses
+how well we can differentiate modes of selective interactions in practice. I
+find that observations of positive LD between missense mutations within genes
+is driven by strong positive correlations between pairs of mutations that fall
+within the same conserved domain, pointing to compensatory mutations or
+antagonistic epistasis as the prevailing mode of interaction within but not
+outside of conserved genic elements. The heterogeneous landscape of both
+strength of selection and selective interactions within protein-coding genes
+calls for more refined inferences of the joint distribution of fitness and
+interactive effects, and the methods presented here should prove useful in that
+pursuit.
+
+\end{abstract}
+
 # Introduction
 
-This paper is neat [@Ragsdale2020-hv].
+Most new mutations that affect fitness are deleterious and tend to be
+eliminated from a population. The amount of time that a deleterious mutation
+segregates in a population depends on the strength of selection against genomes
+that carry it, with very damaging mutations kept at low frequencies in the
+population and purged relatively rapidly. In the time between mutation and
+fixation or loss, selected variants, both beneficial and damaging, can
+dramatically impact patterns of variation in nearby linked regions. This
+distortion away from neutral expectations has been widely documented in
+sequencing data, but questions remain about the primary mode of interactions
+between multiple linked variants and their joint effects on genome-wide
+patterns of diversity.
 
-@Ragsdale2020-hv is still a neat paper if we cite it like this.
+In their foundational paper, @Hill1966-gv recognized that linked selected
+variants reciprocally impede the efficacy of selection at each locus. Linked
+selection reduces the fixation probability of advantageous mutations and
+increases that of deleterious mutations, compared to expectations under
+single-locus models [@Birky1988-jm]. Allele frequency dynamics of linked
+selected variants are also predicted to deviate from expectations without
+interference. Under a multiplicative fitness model, where the fitness reduction
+of a haplotype carrying multiple deleterious variants is equal to the product
+of the fitness reduction of each mutation independently, we should expect to
+see those mutation segregate on different haplotypes more often than together,
+leading to negative, or repulsion, linkage disequilibrium (LD), although the
+extent of LD depends non-trivially on the strength of selection and probability
+of recombination separating loci [@Hill1966-gv; @McVean2000-ox].
 
-```{r figure, echo=F, fig.cap="A figure", fig.align="center", out.width="50%"}
-knitr::include_graphics("Model.png")
+Non-additive effects, including dominance (i.e., interactions *within* loci)
+and epistasis (interactions *between* loci), further complicate our
+evolutionary models. A large fraction of nonsynonymous coding mutations are
+thought to be at least partially recessive [@Agrawal2011-w; b@Huber2018-cp],
+with average levels of dominance correlating with strength of selection
+[@Kacser1981-nc], and dominance plays an important role in shaping expected
+equilibrium allele frequencies and mutation load of strongly damaging disease
+mutations [@Clark1998-kq]. On the other hand, epistasis differentially impacts
+the deleterious load for asexually and sexually reproducing organisms
+[@Kimura1966-gc; @Kondrashov1995-va], has been invoked as an explanation for
+the evolutionary advantage of sex [@Kondrashov1982-sf; @Charlesworth1990-kw;
+@Barton1998-lu], and can drive incompatibilities that lead to postzygotic
+isolation during the process of speciation [@Turelli2000-kz]. Within
+populations, epistasis is known to cause signed LD to deviate dramatically from
+zero [@Charlesworth1990-kw; @Kondrashov1995-va]. However, despite dominance
+being appreciated as important factors shaping linked variation
+[@Turelli2000-kz; @Zhao2016-bb] and the evolutionary importance of epistatic
+interactions, we currently lack models for predicting correlation patterns
+between linked mutations under general selection models.
+
+In this paper, I develop a numerical approach to solve for the two-locus
+sampling distribution under a general diploid selection model with variable
+recombination and single-population size history. I use this model to describe
+how epistasis and dominance shape expected patterns of signed LD, under both
+steady-state and non-equilibrium demography, that have been used to test for
+interference and epistasis in population genomic data. I then turn to human
+sequencing data and compare patterns of LD for synonymous, missense, and
+loss-of-function mutations within and between protein-coding genes and
+annotated domains. I show that while synonymous and missense variants display
+similar slightly positive average LD within genes, for missense mutations this
+signal is driven by correlations between pairs of mutations within, but not
+between, protein-coding domains, suggesting an importance for antagonistic
+epistasis or a prevalence of compensatory nonsynonymous mutations within
+conserved elements. Finally, I discuss how well we can expect to discriminate
+between modes of selective interactions using genome-wide patterns of LD.
+
+## Empirical observations
+
+The most direct way to test for interactions between linked selected variants
+is through deep mutation scanning experiments, in which many distinct mutations
+are introduced within a target gene and then organismal fitness or some protein
+function is experimentally measured [@Romero2009-yi; @Bank2015-vq;
+@Puchta2016-rx; @Steinberg2016-is]. For example, using the model system of the
+TEM-1 $\beta$-lactamase gene in *E. coli*, @Bershtein2006-bk found evidence for
+synergistic epistasis, where multiple deleterious mutations had a greater
+effect on fitness than the multiplication of the observed effects of the
+individual mutations. The scale of mutation scanning experiments continues to
+improve dramatically, promising greater resolution of the fitness landscape in
+such model systems which can be compared to evolutionary models
+[@Otwinowski2018-zk].
+
+In most natural populations, directed mutational studies are not possible, and
+we must turn to population genetic approaches to infer selective interactions
+between observed mutations. Motivated by theory that linked negatively selected
+mutations will display negative LD due to interference [@Hill1966-gv], and that
+epistasis will drive expected LD away from zero, a number of recent studies
+have used patterns of LD within classes of putatively selected variants to
+infer modes of selective interactions. In a notable study from
+@Callahan2011-ac, pairs of nonsynonymous mutations were found to cluster more
+than expected along lineages in the *Drosophilid* species complex, and that
+those clustered mutations tended to preserve the charge of the protein and were
+in positive LD compared to pairs involving synonymous mutations, suggesting
+that compensatory nonsynonymous variants are tolerated and maintained. More
+recently, @Taverner2020-lk replicated this finding across a diverse set of
+genera, showing that such epistatic interactions are important for protein
+evolution.
+
+@Sohail2017-zq observed negative LD between loss-of-function variants in
+protein-coding genes (such as stop gains and losses, frameshifts, and other
+nonsense mutations) in both human and fruit fly populations, which they
+interpreted as evidence for widespread synergistic epistasis between these
+mutations. Within the past year or two, both @Sandler2020-vk and @Garcia2020-zn
+have reevaluated patterns of LD between coding variants in humans, fruit flies,
+and *Capsella grandiflora*, and suggested interference and dominance may
+instead be driving patterns of LD [@Garcia2020-zn] or questioned whether LD
+between loss-of-function variants is significantly different from zero
+[@Sandler2020-vk].
+
+A number of factors impede our interpretation of patterns of signed LD between
+coding variants. First, for strongly deleterious or loss-of-function variants,
+their rarity and low frequency means that statistical measurement of LD and
+other diversity measures are very noisy. Second, comparisons are based on
+theory that are confined to limiting and simplistic assumptions, including
+steady-state demography, simple selection and interaction models, or unlinked
+loci. To generate predictions under more complex models, we rely on expensive
+forward simulations. These can be great for building intuition or testing
+inference methods, but do not efficiently provide expectations for quantities
+of interest across the range of relevant parameters. Analytical and numerical
+methods for expected haplotype frequencies and LD under general selective
+interaction models are thus crucial for interpreting patterns of variation
+observed in data.
+
+# Results
+
+## Expected signed LD under steady-state demography
+
+In the Methods, I expand on the moment system developed in @Ragsdale2019-nt to
+compute the expected sampling distribution $\Psi_n$ of two-locus haplotypes
+under a general model of selective interactions. This sampling distribution, or
+two-locus haplotype frequency spectrum, stores the expected density or observed
+counts of pairs of loci with each possible haplotype frequency configuration in
+a sample of size $n$. In many cases it is simpler to visualize summaries (such
+as $\E[D]$) or slices of the distribution (Figure \@ref(fig:fig1)) instead of
+the full three-dimensional distribution. Here, I focus on low-order LD
+statistics, such as $\E[D]$ and $\E[D^2]$, and their decay with recombination
+distance as these are statistics that are commonly used to test for
+interactions between loci. Instead of unnormalized $\E[D^2]$ and $\E[D]$, I
+consider expectations for $\sigma_d^2 = \E[D^2]/\E[p(1-p)q(1-q)]$ and
+$\sigma_d^1 = \E[D]/\E[p(1-p)q(1-q)$. These statistics have two benefits:
+first, the mutation rates cancel so that expectations are robust to assumptions
+about the per-base mutation rate, and second, we have analytic expectations for
+these quantities under neutrality at steady state for comparison
+[@Ohta1971-yd].
+
+I first consider the case of additive selection and epistasis, followed by
+dominance acting within loci but no epistatic effects, and then describe a
+general diploid selection model and consider simple simple gene-based dominance
+effects. In what follows I focus on parameters where the strengths of selection
+and dominance at each locus are equal, but note that the methods and
+implementation allow for arbitrary and unequal selection and dominance at the
+two loci.
+
+### Additive selection and epistasis
+
+For additive mutations (i.e. $h=1/2$) and no epistasis, we recover the well
+known Hill-Robertson interference result of negative LD between selected
+mutations, which is strongest for pairs for mutations that have selection
+coefficients $\gamma=O(1)$, or $s \approx 1/2N_e$ (Figure \@ref(fig:fig1)D).
+For strongly deleterious mutations, LD is close to zero even with tight
+linkage, as they almost always segregate at low enough frequencies that they
+are unlikely to interfere with each other [@McVean2000-ox].
+
+With epistasis, mean signed LD is large for both weakly and strongly selected
+variants, with sign depending on the direction of epistatic interactions
+(Figure \@ref(fig:fig2)). Synergistic epistasis results in negative LD while
+antagonistic epistasis results positive LD, and large positive LD can occur
+even when the epistatic effect is relatively weak. Epistasis-induced LD can
+extend over long distances, especially for strongly deleterious mutations. For
+example, in Figure \@ref(fig:fig1)D with $\gamma=-10$, which would be
+considered moderately strong selection, large mean LD extends well beyond 100
+kilobases, with more strongly deleterious interacting mutations expected to
+show large signed LD over much larger recombination distances.
+
+### Simple dominance within a locus
+
+The effects of epistasis between variants on LD has been widely documented.
+However, the role that non-additive selection within a locus affects
+interference and LD between loci has received less attention. @Garcia2020-zn
+used large-scale forward simulation to explore the effect of dominance on
+patterns of LD, showing that LD depends on both the magnitude of the fitness
+and dominance effects in a nonlinear way, but note that expensive large-scale
+simulations are needed to explore these effects as no closed form approach to
+computing those expectations exist. The method presented here provides such a
+closed form approach, so that expectations of signed LD and other two-locus
+statistics can be computed more efficiently and accurately.
+
+The combined effect of the strength of selection and dominance on interference
+is indeed nontrivial, and some parameter regimes can cause strong negative LD
+between pairs of negatively selected variants. For strong selection, pairs of
+recessive mutations ($h<1/2$) can have large negative signed LD (Figure
+@\ref(fig:fig3)D), although it decays rapidly with increasing distance between
+loci and is roughly zero for $\rho > 1$. For weakly selected mutations with
+$\gamma=-1$, we see that there is no monotonic effect of the level of dominance
+on negative LD, with both recessive and dominant pairs of mutations having more
+negative LD than the additive case.
+
+### General selection and gene-based dominance
+
+While dominance and epistasis are well-characterized models of selection, a
+large family of selection models can be specified by the assigning unique
+selection coefficients to possible diploid pairs of haplotypes. If we assume
+the diploid genotype that is homozygous for the ancestral allele ($ab/ab$) has
+fitness 1, then there are nine other possible diploid two-locus genotypes that
+could be given unique fitnesses (Table \@ref(tab:selmodels)), noting that
+$AB/ab$ and $Ab/aB$ genotypes can have differing fitness effects.
+
+The case with $s_{AB/ab}\not=s_{Ab/aB}$ could arise in a scenario where a
+mutation at either locus within a haplotype impacts some functional region or
+element, but a diploid carrying at least one copy that is free of mutations has
+minimal fitness loss. This is analogous to the single-locus dominance case,
+where $s_{Aa} < \frac{1}{2}s_{AA}$. In this "gene-based dominance" scenario
+[e.g., @Sanjak2017-zw], an $AB/ab$ genotype has higher fitness than an $Ab/aB$
+type (a simple implementation of this model is given in the final column in
+Table \@ref(tab:selmodels)). Such a gene-based fitness model gives similar
+expected positive signed LD to the model of antagonistic epistasis (compare
+Figure \@ref(fig:fig4)E-F to G-H), although the interpretation of those two
+models can vary. It does show that with a highly parameterized space of
+possible general diploid selection models, multiple models with different
+biological interpretations can give similar patterns of expected $D$.
+
+## The effect of population size changes on signed LD
+
+The moment system for $\Psi_n$ readily incorporates variable population size or
+mutation and recombination rates or selection parameters that change over time.
+Here I focus on non-equilibrium population size history and consider scenarios
+relevant to human demographic history, including bottlenecks and expansions. I
+explore two simple models (Figure \@ref(fig:fig4)A), one with an instantaneous
+expansion and another with a bottleneck followed by recovery. I also consider
+two demographic histories inferred using genome-wide genealogy reconstruction
+[@Speidel2019-nj] applied to the @1000_Genomes_Project_Consortium2015-zq
+dataset, and focus on size histories for the Yoruba from Ibidan, Nigeria (YRI)
+and Utahns of North and West European ancestry (CEU) (Figure \@ref(fig:fig4)B).
+
+For each of the four histories, Figure \@ref(fig:fig4)C-H shows the dynamics of
+$\sigma_d^1$ for a given parameterization of two-locus selection, including
+synergistic and antagonistic epistasis, dominance within loci, and gene-based
+dominance. In general across each selection model, population size expansions
+are not expected to strongly affect $\sigma_d^1$, whether that expansion occurs
+deeper in the past as in the simple expansion model or rapid expansion more
+recently, as in the YRI. On the other hand, population size reductions push
+signed LD to more extreme values and subsequent recoveries or expansion again
+reduce the magnitude of LD. Under no selection condition tested here do
+population size changes cause expected LD to change sign, showing that while
+the magnitude of deviation of LD from zero is sensitive to population size
+history, interpreting general patterns of the observed sign of LD in data
+should not be strongly affected by population size history.
+
+## Signed LD within protein-coding genes
+
+In what follows, I turn to empirical measurements of LD within human
+populations, focusing on mutations within protein-coding regions. I first look
+at average values of LD across genes for synonymous, missense, and
+loss-of-function mutations, and then turn to patterns of LD partitioned by
+functional elements and distances separating mutations within genes.
+
+Synonymous and missense mutations show similar levels of positive LD when
+considering pairs of mutations within the same gene. Loss-of-function mutations
+instead have negative LD, pointing to differing modes of selective interactions
+for loss-of-function and missense mutations (Figure \@ref(fig:fig5)). Within
+each population, large measurement noise gives 95\% confidence intervals that
+overlap with zero for each mutation class, although the pattern is remarkably
+consistent across African, European, and East Asian populations in the Thousand
+Genomes dataset. Comparing mean LD within genes between African and Eurasian
+populations, LD in Eurasian populations is somewhat larger on average, that
+is, more positive for missense mutations and more negative for loss-of-function
+mutations. This is in agreement with differences in expectations between
+populations that have or have not gone through a bottleneck in their recent
+past (Figure \@ref(fig:fig4)).
+
+For pairs of missense mutations in different protein-coding genes, signed LD
+decays to zero fairly rapidly and is lower than signed LD within genes even for
+mutations that are more tightly linked (distance less that 50 kb). The
+similarity in signed LD observed between missense and synonymous mutations
+would suggest that interference between missense mutations is minimal. However,
+below I show that interactive effects differ dramatically between
+mutations that lie within annotated functional domains and mutations outside of
+functional domains, with these opposing effects effectively canceling out when
+averaging LD over entire genes. Due to the rarity of loss-of-function
+mutations, I focus only on comparisons between synonymous and missense
+mutations when looking at finer partitions of mutations within genes.
+
+### Large positive LD between pairs of missense mutations in conserved domains
+
+To further explore the observation of positive LD between both synonymous and
+missense mutations within genes, I examine the role that annotated conserved
+domains in protein-coding genes play in driving this signal. Note that the low
+density of loss-of-function mutations within annotated domains in humans
+precludes any analysis of LD between those mutations at the level of
+sub-elements within genes. Protein-coding domains are conserved elements of
+genes, often associated with some known functional or structural feature of a
+protein. As such, we would expect that purifying selection is stronger on
+average within domains than within the same gene but outside of a conserved
+domains. Indeed, the site-frequency spectrum (SFS) is skewed to lower
+frequencies for both missense and loss-of-function mutations within domains
+when compared to the same classes of mutations outside of domains, with much
+more negative values of Tajima's D within domains (Table \@ref(tab:tajimasD)).
+On the other hand, no difference is observed for synonymous mutations within or
+outside domains, suggesting roughly equivalent effects of selection (either
+direct or linked) on synonymous variation.
+
+For pairs of missense mutations that both fall within the same functional
+domain, we observe large positive LD, roughly an order of magnitude larger than
+pairs of synonymous mutations that both fall within the same domain (Figure
+\@ref(fig:fig6)). This difference between pairs of missense and synonymous
+variants within domains is observed for both tightly linked ($< 10$bps) and
+more distant ($>1$kb) pairs (Figure \@ref(fig:lddecaydomain)). Assuming most
+synonymous mutations are neutral, we would expect signed LD between missense
+mutations to be less than that between synonymous mutations under predictions
+from models of either Hill-Robertson interference, synergistic epistasis, or
+dominance (Figures \@ref(fig:fig1)-\@ref(fig:fig3)). This observation is
+opposite to those expectations, suggesting a different prevailing interactive
+effect within domains than those that are commonly tested or modeled.
+
+The large positive signal of LD for missense mutations within the same
+domain does not extend to pairs of missense mutations that are in
+different domains. Pairs of missense and synonymous mutations show
+nearly equal levels of LD close to zero across domains, with missense mutations
+even slightly more negative than synonymous mutations (Figure \@ref(fig:fig6)).
+The interactive effect driving large LD within domains is therefore likely
+domain-specific. However, the average distance between pairs of mutations
+within domains is much smaller than between domains, so this observation may be
+primarily driven by the higher recombination rate between pairs of mutations
+across distinct domains.
+
+### Reduced LD between pairs of missense mutations outside of conserved domains
+
+I also measure LD between pairs of mutations that both fall outside of
+annotated domains. For such mutations, I consider averages across all pairs
+outside of domains, as well as subsets of pairs matched to the same distances
+for pairs of mutation within and between domains (Figures \@ref(fig:fig6)B-D and
+\@ref(fig:domainspops)). Averaging over all pairs of mutations within the same
+gene but outside domains, both synonymous and missense mutations had LD
+slightly above zero, with synonymous mutations observed to have slightly higher
+signed LD than missense mutations. This pattern is primarily driven by tightly
+linked variants. For pairs of mutations at distances matched to those within
+the domains, synonymous mutations have much larger positive LD than
+missense mutations, a signal that vanishes for pairs matched to distances
+between domains. For larger distances, both synonymous and missense mutations
+have LD approximately zero.
+
+The role that tightly linked variants have in driving this signal can be
+clearly seen in the decay of signed LD with distance (Figure
+\@ref(fig:lddecay)). Both pairs of synonymous and missense mutations with
+distances greater than a few hundred base pairs have LD that fluctuates around
+zero. But for shorter distances, synonymous variants have large positive LD,
+while missense mutations have much lower LD except for pairs within the
+shortest distance bin ($<10$ bp). This reduction in LD is consistent with
+Hill-Robertson interference between linked missense mutations instead of
+epistatic effects, as interference between either additive or recessive
+mutations without epistasis results in signed LD that decays rapidly by
+recombination distance $\rho \approx 1$.
+
+# Discussion
+
+Previous theoretical and simulation studies have shown that interference and
+interactions between selected mutations reduces the efficacy of selection at
+linked loci [@Hill1968-vu; @Birky1988-jm; @Barton1995-mj; @McVean2000-ox],
+impacting substitution rates, the deleterious mutation load, and dynamics of
+segregating mutations. Interference between moderately deleterious mutations
+and synergistic epistasis are both expected to cause negative LD between
+selected mutations, which can be readily tested in population genetic data
+[@Sohail2017-zq; @Sandler2020-vk; @Garcia2020-zn]. Here, I first used a closed
+numerical approach to generate expectations for signed LD under a wide range of
+selective scenarios, and then compared patterns of LD in human populations
+between classes of coding mutations within genes using unbiased estimators for
+LD from unphased genotypes.
+
+By taking broad genome- or even gene-wide surveys of LD within functional
+classes of mutations, patterns of interactive effects that occur within
+functional sub-elements of genes can be missed. However, at the gene-wide
+level, we already find large differences in the interactions between pairs of
+missense and loss-of-function mutations. While LD between pairs of missense
+mutations does not appear to be different from pairs of synonymous variants, LD
+between loss-of-function variants is substantially more negative.
+Hill-Robertson effects under additive selection are expected to be strongest
+for slightly to moderately deleterious variants (with $s\sim 1/N_e$), as
+strongly deleterious mutations are not expected to interfere with one another
+[@McVean2000-ox]. Negative synergistic epistasis between strongly deleterious
+mutations does produce large negative deviations of mean LD. However, strongly
+deleterious *recessive* mutations can also produce this pattern, making it
+difficult to distinguish these effects with certainty. This difficulty is
+compounded by the large measurement noise for $\E[D]$, especially for
+loss-of-function variants for which only a few hundred pairs are found within
+genes in each population.
+
+## Non-uniform interactive effects of selected mutations within genes
+
+Positive average LD between missense and synonymous mutations has previously
+been reported in humans and other species [@Sohail2017-zq; @Garcia2020-zn;
+@Sandler2020-vk], and the similarity of their gene-wide LD might suggest that
+interference or interactions between missense mutations are minimal, or at
+least no stronger than those between synonymous mutations. However, averaging
+over all observed pairs of mutations within a gene completely masks different
+interactive effects in functional subunits of genes, effects that even drive LD
+in opposite directions. Nonsynonymous mutations found within protein domains,
+identified as conserved subunits of a gene with some structural or functional
+role, are more strongly selected against on average, but also have increased
+signed LD over synonymous mutations at the same distances and also falling
+within domains. Missense mutation outside of domains but at the same distances
+as those within domains have more negative LD, both compared to synonymous
+mutations outside of domains at the same distance and to mutations within
+domains. Neither dominance effects, synergistic epistasis, nor Hill-Robertson
+interference are expected to result in positive LD, so that some other
+interactive effect must be driving this signal within domains.
+
+There are a number of possible interaction scenarios that can result in
+positive signed LD between linked loci. One possible explanation is a
+prevalence of pairs of compensatory mutations that are tolerated to
+co-segregate at high frequencies within conserved domains [@Yeang2007-gj;
+@Ivankov2014-tn]. @Callahan2011-ac and @Taverner2020-lk have proposed such a
+mechanism to explain observed clusters of mutations in lineages of *Drosophila*
+and other species. Another possibility is a model of antagonistic, or
+diminishing returns epistasis, in which a single amino acid-changing mutation
+within a domain damages the functionality of that subunit, but additional
+mutations within that same domain reduces fitness by a factor less than the
+first mutation. A third possibility, related to the antagonistic epistasis
+case, is that selection acts on the functional domain as a unit instead of at
+loci within the domain individually (such as gene-based dominance). In this
+scenario, double heterozygotes have different fitnesses depending on whether
+the mutations are found on same haplotype or on different haplotypes. Each of
+these scenarios lead to positive LD.
+
+We do not see an increase in LD between mutations that are found in different
+annotated domains within the same gene, which are on average at much larger
+recombination distances. Nor do we find increased LD outside of domains.
+Rather, for pairs of mutations outside of domains but at the same distance as
+those within the same domain, missense variants have considerably lower LD than
+than synonymous variants. This difference between synonymous and missense pairs
+of variants largely disappears for SNPs separated by more than a few hundred or
+a thousand base pairs (Figure \@ref(fig:lddecay)), which suggests that
+Hill-Robertson interference is the primary mode of interaction between
+mutations falling outside of domains, as epistasis is expected to impact LD
+over larger distances than what is observed.
+
+Taken together, selection on segregating variants within protein-coding genes
+is nonuniform, with both the overall strength of selection and interactions
+between variants differing between annotated elements. Missense and
+loss-of-function mutations within conserved domains are subject to stronger
+selection, skewing the SFS to lower frequencies. However, typical approaches
+for inferring the distribution of fitness effects from population genomic data
+average over these differences by considering all nonsynonymous mutations
+together [@Boyko2008-zk; @Kim2017-xo]. It would be straightforward to adapt
+those methods to infer more detailed and accurate representation of the
+heterogeneous effects of new mutations within genes by partitioning by missense
+and loss-of-function classes as well as by annotated domains. Similarly, the
+different modes of interaction between mutations across annotated domains mean
+that our standard models and simulation approaches may be too simple and will
+not capture the evolutionary trajectories and patterns of diversity that differ
+at finer scales.
+
+## Clustered mutations may drive positive LD among synonymous mutations
+
+In a single randomly mating population, neutral mutations are expected to have
+average signed LD of zero, but across all populations analyzed here, we observe
+positive LD between synonymous mutations. While selection on some subset of
+synonymous variants is possible, it should be much weaker on average than
+between missense mutations, and any interference between selected synonymous
+variants should lead to negative LD. @Sandler2020-vk proposed that spatial
+population structure is responsible for the increase of LD for both synonymous
+and missense variants, as every natural population lives in some geography that
+will necessarily induce some structure, and they used forward simulations to
+demonstrate this effect. However, this increase in LD, at least in simulations,
+extends over larger distances separating base-pairs than we
+observe in the data. Instead, positive LD between synonymous mutations
+decays to zero within just a few hundred base-pairs (Figures \@ref(fig:lddecay)
+and \@ref(fig:lddecaydomain)). We might also expect that differences in
+migration and admixture histories between human populations would result in
+different levels of positive synonymous LD, but the patterns are fairly
+consistent across populations.
+
+An alternative explanation for positive LD between neutral variants separated
+by short distances is a nonrandom mutational process, where clusters of
+mutations occur simultaneously in the same mutational event. Such
+multinucleotide mutations have been shown to affect patterns of linkage
+disequilibrium over short distances, on the order of 10s to 100s of base pairs
+[@Harris2014-zg], and clustered mutational events may be common in humans
+[@Besenbacher2016-ac]. Thus, positive LD may be the baseline expectation for
+tightly linked variants primarily due to clustered mutations, and subsequent
+selective processes and interactions will cause LD to deviate from that
+expectation, as seen for missense mutations across different sub-elements of
+genes. Similarly, it may be more appropriate to compare the negative LD
+observed between linked loss-of-function variants to that positive baseline
+expectation instead of zero, implying that they would be more recessive or have
+stronger synergistic epistasis than if the neutral expectation were zero.
+
+## Challenges to distinguishing highly parameterized models of selective interactions
+
+In exploring or inferring selective models across multiple mutations, the
+number of plausible selection scenarios becomes quite large as we relax
+assumptions about additive and multiplicative fitness functions. The inclusion
+of dominance effects, epistasis, or other interactions leads to a rapid
+increase in the number of parameters to consider. This makes performing forward
+simulations that span the range of all such selective interaction scenarios
+burdensome.
+
+Instead, closed, numerical approaches to compute expectations for two-locus
+statistics under a wide range of selection models, as presented here, allows us
+to explore this highly parameterized space of models far more efficiently, and
+it opens the possibility for performing likelihood-based inference using signed
+LD or other two-locus summaries of the data. For example, inferring the joint
+distribution of dominance and selection is underpowered using the SFS. Because
+signed LD is sensitive to the levels of dominance (Figure \@ref(fig:fig3)),
+inferring the distribution of fitness effects with dominance may become
+feasible using the joint distribution of allele frequencies and LD.
+
+Using a single low-order summary of signed LD, such as $\E[D]$ or $\sigma_d^1$,
+is likely insufficient to confidently discriminate modes of selective
+interactions between linked mutations. Both dominance and epistasis can result
+in strongly negative LD, although the expected decay of LD under those
+scenarios can differ. Similarly, multiple interaction models can lead to
+positive signed LD. Among all interaction models, the extent of LD and rate of
+its decay also depends on the underlying distribution of selection coefficients
+among a class of mutations, which are unknown for a given pair of mutations, so
+that we must integrate over a distribution of fitness effects. This DFE,
+however, will have been inferred under a simple set of assumptions, such as
+additivity and interchangeability between sites within a gene, potentially
+biasing any inference using this DFE to learn about other patterns of
+selection.
+
+Finally, exploring additional summaries of the two-locus sampling distribution
+could provide additional power to distinguish between interaction models. The
+method developed here allows us to compute the full sampling distribution for a
+given sample size, and expectations for a large family of informative two-locus
+statistics can be computed directly from this distribution. These expectations
+can then be compared directly to empirical observations, which can be taken
+from either phased or unphased [@Ragsdale2020-nz]. Additional work performing
+such an analysis is therefore warranted, which should provide a path forward
+for distinguishing between modes of selective interactions.
+
+# Methods
+
+## Existing theory and numerical methods
+
+Many well-known properties of two-locus dynamics and equilibrium LD come from
+early work on the multi-locus diffusion approximation [@Kimura1955-qe;
+@Hill1968-vu; @Ohta1969-ie; @Ohta1971-yd]. This includes the result that
+genome-wide averages of signed LD are expected to be zero under neutrality.
+Under a two-locus biallelic model, where the left locus allows alleles $A$ and
+$a$ and the right locus has alleles $B$ and $b$, the standard covariance
+measure of LD is defined as $D = f_{AB} - f_{A}f_{B}$ where $f_{AB}$ is the
+haplotype frequency of double-derived types carrying both $A$ and $B$, and
+$f_{A}$ and $f_{B}$ are the marginal frequencies of the derived alleles at each
+locus. This covariance decays due to both drift and recombination at a rate
+proportional to the inverse of the effective population size and the distance
+separating loci: \[\E[D]_{t+1} = \left(1 - \frac{1}{2N_e(t)} - r
+\right)\E[D]_t.\] While $\E[D] = 0$, the variance of $D$ is non-zero, and
+@Ohta1971-yd presented their groundbreaking result that the variance of $D$
+under neutrality and steady-state demography, normalized by the joint
+heterozygosity of the two loci, is \begin{equation} (\#eq:ohta) \sigma_d^2 =
+\frac{\E[D^2]}{\E[f_A(1-f_A)f_B(1-f_B)]}\approx\frac{5 + \rho / 2}{11 + 13\rho
+/ 2 + \rho^2 / 2}, \end{equation} where $\rho = 4N_e r$.
+
+Analytic progress beyond these results has come haltingly. In the
+1980s, recursions were developed to compute the two-locus sampling distribution
+under neutrality, that is, the probability of observing given counts of
+two-locus haplotypes in a sample of size $n$ [@Golding1984-pu]. This approach
+would continue to be developed and later form the foundation for the inference
+of local recombination rates from population genetic data [@Hudson2001-sg;
+@McVean2004-gj].
+
+To include selection, however, there have been relatively few advances beyond
+the Monte Carlo simulation approach taken by @Hill1966-gv, albeit we now have
+access to somewhat more powerful computational resources and more sophisticated
+software for performing flexible forward simulation (e.g., @Thornton2014-pn,
+@Haller2019-vm). Analytic results for two-locus distributions under selection
+are notoriously difficult, with a few notable flashes of progress.
+@McVean2007-xd considered the effect of a recent sweep on patterns of LD for
+loci nearby or spanning the locus under selection, and in an impressive recent
+paper, @Good2020-wj presented analytic solutions for patterns of LD between
+rare mutations under additive selection with epistasis. Nonetheless, such
+approaches are confined to steady-state demography and constrained selection
+models.
+
+Numerical methods straddle the two approaches of expensive and noisy discrete
+simulation and analytic solutions, providing a more efficient and practical
+method to compute expectations of two-locus diversity measures under a wider
+range of parameters and demographic scenarios. @Ragsdale2017-gg used a finite
+differencing approach to numerically solve the two-locus diffusion equation
+with additive selection at either locus, although that paper focused on the
+applicability of two-locus statistics to demographic inference. More recently,
+@Ragsdale2019-nt extended the @Hill1968-vu system for $\E[D^2]$ to compute
+arbitrary moments of the distribution of $D$ for any number of populations
+connected by migration and admixture. This paper also showed that such a
+moments-construction can be used to solve for the two-locus sampling
+distribution within a single population, though requires a moment-closure
+approximation for nonzero recombination and selection. Below, I
+extend this same approach to model arbitrary diploid selection, which
+encompasses dominance, epistasis, and other forms of selective interactions
+between two loci.
+
+## The two-locus sampling distribution with arbitrary selection
+
+The two-locus sampling distribution is the direct analog to the single-locus
+site frequency spectrum (SFS) of a given sample size. Instead of describing the
+density or number of mutations with a given allele count out of $n$ samples,
+the two-locus distribution $\Psi_n$ stores the density or number of pairs of
+loci with observed haplotype counts, so that $\Psi_n(i, j, k)$ is the number of
+pairs where with $i$ copies of the $AB$ haplotype, $j$ of type $Ab$, $k$ of
+type $aB$, and $n-i-j-k$ of type $ab$. The size of $\Psi_n$ grows rapidly, with
+$O(n^3)$ entries, which practically limits computational approaches to moderate
+sample sizes and a single population.
+
+Under neutrality, a number of approaches exist to compute $\Psi_n$, including
+the recursion due to @Golding1984-pu and @Ethier1990-tb, or more recent
+numerical approaches to the two-locus coalescent [@Kamm2016-ag] or diffusion
+approximation [@Ragsdale2017-gg]. Practically, selection can only be included
+using the forward-in-time diffusion equation [@Kimura1955-qe; @Hill1966-gv],
+where a standard approach is to first solve for the continuous distribution
+$\psi$ of the density of two-locus haplotype configurations in the full
+population, and then integrate $\psi$ against the multinomial sampling function
+to obtain $\Psi_n$.
+
+Alternatively, @Ragsdale2019-nt showed that there exists a system of ordinary
+differential equations directly on the entries of $\Psi_n$, although that paper
+focused primarily on neutral dynamics and considered a model with selection at
+only one of the two loci. I briefly summarize this general approach below, but
+refer readers to that previous paper for detailed derivations of the drift,
+recombination, and mutation terms and the moment-closure approximation.
+Instead, here I focus on generalizing the selection operator to include
+epistasis, dominance, and other forms of two-locus interactions.
+
+### Moment equation for $\Psi_n$
+
+The system of linear ordinary differential equations for the entries of
+$\Psi_n$takes the form
+\begin{equation}
+(\#eq:system)
+{\Psi}_n^{t+1}(i, j, k; t) = 
+\mathcal{D}_{N(t)}\Psi_n^t
++ \mathcal{R}_{r}\Psi_{n+1}^t
++ \mathcal{U}_{u}\Psi_n^t
++ \mathcal{S}_{s_A, s_B, \ldots, h_A, h_B, \ldots}\Psi_{n+2}^t.
+\end{equation}
+Here, $\mathcal{D}_{N(t)}$ is a sparse linear operator accounting for drift
+with population size $N(t)$, $\mathcal{R}$ accounts for recombination
+with per-generation recombination probability $r$ between the two loci,
+$\mathcal{U}$ accounts for mutation, either under an infinite sites or
+biallelic reversible mutation model, and $\mathcal{S}$ accounts for
+selection.
+
+The moment system for $\Psi_n$ can be derived directly from the diffusion
+approximation, or it can be found through a more intuitive process of tracking
+the dynamics of allelic states of a sample of size of $n$ from the full
+population. We assume $n \ll N_e$, and $r$ and $s$ are $O(1/Ne)$ so that
+multiple coalescence, recombination, or selective events are rare in any given
+generation. For example, for drift events, we account for the probability that
+two lineages within the $n$ tracked lineages at time $t+1$ coalesced in the
+previous generation and, given a coalescence event, the probability that
+haplotype counts changed due to this coalescence event [see Supporting
+Information; @Jouganous2017-pq; @Ragsdale2019-nt]. In typical diffusion
+approximation fashion, we multiple through by $2N_{ref}$ so that time
+is measured in $2N_{ref}$ generations, and we consider scaled parameters
+$\rho = 4Nr$, $\theta = 4Nu$, and $\gamma=2Ns$.
+
+### Moment closure
+
+In the absence of selection and for fully linked loci (i.e., $\rho=0$), the
+system is closed and can be solved exactly. However, for nonzero recombination
+or selection, the entries of $\Psi_n$ rely on the slightly larger frequency
+spectra with sample sizes $n+1$ (for recombination and additive selection) or
+$n+2$ (for non-additive selection). This is because if a recombination event
+occurs within one of $n$ lineages being tracked by $\Psi_n$, we need to draw an
+additional lineage from the full population to recombine with that chosen
+lineage, thus requiring $\Psi_{n+1}^t$ to find $\Psi_n^{t+1}$. Selection events
+similarly require extra lineages from the full population, which replace a
+chosen lineage that fails to reproduce with probability proportional to its
+selection coefficient, assuming negative selection.
+
+This requirement of extra lineages for nonzero recombination and selection
+means that the system in \@ref(eq:system) is not closed.
+This means we need a moment-closure approximation to solve for $\Psi_n$. As in
+@Ragsdale2019-nt, a jackknife approximation is used to estimate $\Psi_{n+l}$,
+for $l=1$ or $2$, from $\Psi_n$ [@Jouganous2017-pq], so that
+$\hat{\Psi}_{n+l}(i, j, k) = \mathcal{J}_{n, l}\Psi_n$, although other closure approximations are possible.
+This gives us a closed approximated system,
+\begin{equation}
+(\#eq:closed-system)
+\dot{\hat\Psi}_n(i, j, k; t) = 
+\mathcal{D}_{\nu(t)}\hat\Psi_n(t)
++ \mathcal{R}_{\rho}\mathcal{J}_{n, 1}\hat\Psi_{n}(t)
++ \mathcal{U}_{\theta}\hat\Psi_n(t)
++ \mathcal{S}_{\gamma, h}\mathcal{J}_{n, 2}\hat\Psi_{n}(t).
+\end{equation}
+
+The jackknife approximation, which approximates an entry $\Psi_{n+l}(i,j,k)$
+using nearby entries in $\Psi_n$, is more accurate for larger sample sizes,
+creating a tension between efficiency and accuracy: larger sample sizes result
+in more accurate solutions, as error in the jackknife is diminished, but
+computational complexity also grows rapidly in the number of entries of
+$\Psi_n$, which is $O(n^3)$. Accuracy and runtime are shown in Figure
+\@ref(fig:jackknife). In the results presented in this paper, sample sizes
+between $n=30$ and $n=80$ were used.
+
+Derivations for the drift, recombination, and mutation operators and the
+jackknife moment-closure approximation can be found in section S1.3 of
+@Ragsdale2019-nt, and I repeat the main results in the Supporting Information
+of this paper.
+
+### Selection models with epistasis and dominance
+
+To model selection, we consider a model where we draw lineages uniformly from
+the previous generation, but keep lineages with probability proportional to
+their fitness. In the absence of dominance, the selection model reduces to a
+haploid model, with acceptance and rejection probabilities depending on the
+fitnesses of each haploid copy, where haplotype $Ab$ has fitness $1 + s_A$,
+$aB$ has fitness $1 + s_B$, and $AB$ has fitness $s_{AB}$. We assume the doubly
+ancestral haplotype $ab$ has fitness $1$, so fitnesses are relative to that of
+$ab$ haplotypes. The standard multiplicative fitness function assumes that
+$s_{AB} \approx s_A + s_B$, and a general epistasis function can be written as
+$s_{AB} = (s_A + s_B) (1 + \epsilon)$, so that $\epsilon>0$ implies synergistic
+epistasis, while $\epsilon < 0$ implies antagonistic epistasis.
+
+To obtain the recursion equation under selection we consider drawing $n$
+lineages from generation $t$, and this sampling has an expected distribution of
+haplotype counts given by $\Psi_n^t$. However, assuming $s\leq0$ for each
+haplotype, each of those sampled lineages as probability of being rejected
+equal to the relative fitness of its haplotype state. If a lineage is rejected,
+a replacement is drawn from among the population. Under the assumption that
+$ns_{\cdot\cdot} \ll 1$, the probability that more than two selection events
+occur in any given generation is negligibly small, so that the case of multiple
+simultaneous rejections can be ignored, and $Psi_n^{t+1}$ relies only on
+$\Psi_n^t$ and $\Psi_{n+1}^t$ for additive selection. The full selection
+operator $\mathcal{S}$ is given in the Supporting Information.
+
+To account for dominance within loci, or other general forms of two-locus
+selection, the selection operator no longer reduces to individual haplotypes,
+but instead we need to know the state of two-locus genotypes. For example, the
+fitness of an individual carrying an $Ab$ haplotype depends on whether their
+second haplotype is $ab$, $Ab$, $aB$, or $AB$. We can therefore assign a
+selection coefficient to each possible diploid configuration, $s_{AB/AB}$,
+$s_{AB/Ab}$, and so on. Assuming that the doubly homozygous ancestral $ab/ab$
+genootype has relative fitness 1, this gives nine possible unique selection
+coefficients in the most general two-locus selection model. Note that $AB/ab$
+and $Ab/aB$ genotypes need not have the same selection coefficient, which
+allows for simulation under a gene-based model of dominance (Table
+\@ref(tab:selmodels)).
+
+The general selection operator follows the same approach as the haploid
+selection operator with epistasis described above. Now, in the case of a
+selection event rejecting a lineage within our tracked samples, we need to draw
+not only the replacement lineage from the full population but also a second
+haplotype from the full population to form the diploid genotype, as this
+determines the probability that we reject the focal haplotype. We again assume
+that $ns \ll 1$ for all genotype selection coefficients, so that we may assume
+at most a single selection event occurs in any given generation. This means
+that to find $\Psi_n^{t+1}$ under a general two-locus selection model, which
+encompasses dominance within either locus, gene-based dominance, or a
+combination or dominance and epistatic effects, we need $\Psi_{n+2}^t$. Again,
+a full derivation and expressions for the general selection operator are given
+in the Supporting Information.
+
+### Low-order summaries of the sampling distribution
+
+From $\Psi_n$, expectations for any two-locus statistic can be found by
+downsampling to the appropriate sample size. For example, to compute $\E[D]$,
+the sum is taken over all haplotype configurations $\mathbf{n} = (n_{AB},
+n_{Ab}, n_{aB}, n_{ab})$, weighted by the density $\Psi_n$ for that
+configuration:
+\begin{equation}
+\E[D] = \sum_{\mathbf{n}} \Psi_n(\mathbf{n})
+\frac{n_{AB}n_{ab} - n_{Ab}n_{aB}}{n(n-1)}.
+\end{equation}
+For large sample sizes, this is approximately equal to computing $D$ by taking
+the maximum likelihood estimate for each allele frequency $f_i = n_i / n$, but
+the maximum likelihood estimate will be noticeably biased for small to moderate
+sample sizes. Other low-order two-locus statistics can be computed using the
+same approach, as implemented in *moments* following @Ragsdale2020-nz, which
+can be compared across sample sizes and between estimates from phased or
+unphased data. In this paper, I focus on $\sigma_d^2 = \E[D^2] /
+\E[p(1-p)q(1-q)]$ and $\sigma_d^1 = \E[D²] / \E[p(1-p)q(1-q)]$, averaged over
+alleles at all frequencies. However, allele-frequency conditioned statistics
+(e.g., keeping only loci below some frequency threshold [@Good2020-wj]) can be
+considered using this same approach.
+
+### Simulations of non-steady-state demography
+
+## Analysis of human genomic data
+
+Using the annotated variant call format (VCF) files from the phase 3
+@1000_Genomes_Project_Consortium2015-zq (Thousand Genomes) data release to
+subset the genotype VCFs of the same release to autosomal variants that were
+annotated as either synonymous or nonsynonymous (including both missense and
+more damaging mutations such as stop gains, stop losses, and so on). I further
+subset to samples within each non-admixed population in the African, European,
+and East Asian continental groups. Signed LD is sensitive to ancestral state
+identification, so I only kept sites for which ancestral alleles were estimated
+with high confidence in both the VCF info fields and the Thousand Genomes human
+ancestor reconstructed from a phylogeny of six primates.
+
+In addition to ancestral state misidentification, measured LD is sensitive to
+phasing error, so I computed LD statistics using unphased genotypes following
+@Ragsdale2020-nz. This approach provides unbiased estimates for pairwise LD,
+under the assumption of a random mating population. I considered pairs of
+mutations within the same mutation class (synonymous, missense, and
+loss-of-function) either within the same gene, in different genes within
+distance bins up to 500kb, or inside or outside of annotated domains within the
+same protein-coding genes. I used a dataset of all known conserved domains in
+humans compiled by @Stanek2020-pa.
+
+## Data and software availability
+
+All data and software used in this paper are publicly available and open source.
+I downloaded the Thousand Genomes annotations and genotypes VCFs from the ftp
+server at \url{ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/}, and
+the Thousand Genomes human ancestor fasta file from
+\url{ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase1/analysis_results/supporting/ancestral_alignments/}.
+Protein domain information from @Stanek2020-pa was downloaded from
+\url{http://prot2hg.com/dbdownload.php}.
+
+Implementation of moment equations to compute expectations for two-locus and
+linkage disequilibrium statistics are implemented in Python using Numpy
+[@Harris2020-pc] and sparse matrix solvers in Scipy [@Virtanen2020-kr]. These
+methods are packaged within *moments*, and analyses here were performed using
+version 1.1.0 (\url{https://bitbucket.org/simongravel/moments/src/master/}).
+Scripts to run all analyses, recreate figures, and compile this manuscript are
+available at \url{https://github.com/apragsdale/two_locus_selection}.
+
+# Acknowledgements
+- Simon Gravel
+- Kevin Thornton
+- Alex Diaz-Papkovich
+- Mashaal Sohail
+
+\clearpage
+\newpage
+
+# Figures
+
+```{r fig1, echo=F, fig.cap="(ref:fig1cap)", fig.align="center"}
+knitr::include_graphics("figures/fig1.pdf")
 ```
 
-See Figure \@ref(fig:figure).
+(ref:fig1cap) \textbf{The two-locus sampling distribution and additive
+selection.} (A and B) The distribution of $AB$ haplotype counts in a sample
+size of 30, in which we observe 10 $A$ alleles at the left locus, and 10 $B$
+alleles at the right locus. (C) The decay of normalized squared LD
+($\sigma_d^2$) with recombination distance ($\rho=4N_er$) for pairs of variants
+with a given selection coefficient. The dashed line is the neutral expectation
+from @Ohta1971-yd, given in Equation \@ref(eq:ohta). (D) The decay of
+$\sigma_d^1=\E[D]/ \E[p(1-p)q(1-q)]$ for pairs of variants. The
+negative LD effect of Hill-Robertson interference is strongest for intermediate
+selection, with $\gamma\sim-1$ to $-5$.
 
-See Supplementary Figure \@ref(fig:figure).
+```{r fig2, echo=F, fig.cap="(ref:fig2cap)", fig.align="center"}
+knitr::include_graphics("figures/fig2.pdf")
+```
+
+(ref:fig2cap) \textbf{Additive selection with epistasis.} (A, B) Synergistic
+epistasis decreases $D^2$, while antagonistic epistasis increases $D^2$. (C, D)
+Epistasis expected signed LD away from zero, an effect which can extend over
+long recombination distances. Here, $\gamma$ is the selection coefficient for
+derived alleles $A$ and $B$, assuming selection is equal at the two loci.
+$\epsilon=0$ is the case of pure additive selection with no epistasis, and
+$\gamma_{AB}=(\gamma_A + \gamma_B)(1+\epsilon)$. Dashed lines are expectations
+under neutrality for both $\sigma_d^2$ and $\sigma_d^1$.
+
+```{r fig3, echo=F, fig.cap="(ref:fig3cap)", fig.align="center"}
+knitr::include_graphics("figures/fig3.pdf")
+```
+
+(ref:fig3cap) **Dominance effects.** (A-B) Pairs of recessive mutations
+($h<0.5$) have higher squared LD than additive mutations, while pairs of
+dominant alleles have lower expected $D^2$. (C-D) Pairs of recessive, strongly
+deleterious mutations can have large negative LD, although this correlation
+decays more rapidly with larger recombination than LD under epistasis (Figure
+\@ref(fig:fig2)D). Notably, there is no simple relationship between strength of
+selection and dominance, particularly when $s$ is $O(1/N)$.
+
+```{r fig4, echo=F, fig.cap="(ref:fig4cap)", fig.align="center"}
+knitr::include_graphics("figures/fig4.pdf")
+```
+
+(ref:fig4cap) **The effects of demography on signed LD.** Left column:
+simulations under two toy models, one with an instantaneous expansion in the
+past, and the other with a 10-fold reduction in size followed by recovery.
+Right column: simulations under population size histories inferred using Relate
+[@Speidel2019-nj] applied to populations from
+@1000_Genomes_Project_Consortium2015-zq. In all simulations, $\gamma$ is set to
+$-2$ at both loci. (A-B) Model illustrations. Note that the differences in
+scale. (C-D) Additive selection with no epistasis, showing fully linked
+($\rho=0$) and more loosely linked ($\rho=5$) pairs. Bottleneck events result
+in larger negative values of signed LD, but population size changes are not
+expected to change the sign of mean LD. (E-F) Synergistic ($\epsilon > 0$) and
+antagonistic ($\epsilon < 0$) epistasis. Again, bottleneck events result in
+larger signed LD. Here, $\epsilon=1/2$ is a model with $\gamma_A=-2$,
+$\gamma_B=-2$ and $\gamma_{AB}=-6$, and $\epsilon=-1/2$ is a model with
+$\gamma_A=-2$, $\gamma_B=-2$ and $\gamma_{AB}=-2$. (G-H) Dominance within loci
+results in negative LD, while dominance acting at the gene level results in
+positive LD. Again, bottlenecks result in larger average LD in both cases.
+
+```{r fig5, echo=F, fig.cap="(ref:fig5cap)", fig.align="center"}
+knitr::include_graphics("figures/fig5.pdf")
+```
+
+(ref:fig5cap) **LD within and between protein-coding genes.** (A-B) Pairs of
+variants in coding regions partitioned by classes of mutations.
+Loss-of-function mutations are defined as stop and start gains and losses,
+frameshifts, and other nonsense mutations. In all populations, we observe
+positive mean signed LD, although the confidence interval of each measurement
+overlaps with zero. Similarly, we observe mean negative LD for pairs of
+loss-of-function variants within genes. There are only $\sim200-300$ pairs of
+loss of function variants that fall within the same gene across the populations
+tested here, causing large uncertainty in each measurement. Gray lines indicate
+2 standard errors of the measurement, via bootstrap. (C-F) While pairs of
+synonymous and missense mutations within genes are consistently positive, pairs
+between genes are not consistent in their sign across populations. LD for all
+populations analyzed here are shown in Figure \@ref(fig:genespops).
+
+```{r fig6, echo=F, fig.cap="(ref:fig6cap)", fig.align="center"}
+knitr::include_graphics("figures/fig6.pdf")
+```
+
+(ref:fig6cap) **LD within and between annotated domains in protein-coding
+genes.** Considering only variants that fall within annotated domains in
+protein-coding genes, pairs of mutations that fall within the same gene display
+larger positive LD than pairs that lie in different domains. This pattern is
+most striking for missense mutations, for which pairs within domains have
+average signed LD that is consistently larger than for pairs that span two
+domains, or for pairs of synonymous mutations that are within the same domain.
+The slightly positive LD that we observe for missense and synonymous mutations
+gene-wide (Figure \@ref(fig:fig4)A) is largely driven by the subset of pairs
+mutations that fall within the same annotated domain. LD for all populations
+analyzed here are shown in Figure \@ref(fig:domainspops).
 

### 2021-03-01 SupportingInformation.Rmd
edits throughout

--- 2021-02-20 SupportingInformation.Rmd
@@
       in_header: "sipreamble.tex"
+    extra_dependencies: ["float"]
 bibliography: "references.bib"
-title: "Paper title"
+title: "Supporting Information for [[main paper title]]"
 author:
-  - Author 1^[Institution 1, email1]
-  - Author 2^[Institution 2, email2]
+  - Aaron P. Ragsdale^[aaronpeaceragsdale@gmail.com]
+date:
+  - \today
 ---
@@
 
+Yet to include:
+- Tables with data presented in figures 5 and 6
+
+## The diffusion equation, etc
+
+\[ 
+\frac{\partial \psi}{\partial t} = 
+\frac{1}{2}\mathop{\sum\sum}_{1\leq i, j \leq 3} 
+\frac{\partial^2}{\partial x_i \partial x_j} 
+\left[\frac{x_i(\delta_{i=j}-x_j)\psi}{N(t)}\right] 
+-\frac{\rho}{2}\left(\frac{\partial}{\partial x}\right) 
+\]
+
+## Moment equation and evolution operators
+
+### Drift
+
+### Recombination
+
+### Mutation
+
+### General selection
+
+### Moment closure via jackknife approximation
+
+## Data analysis
+
+### Grouping Thousand Genomes populations based on clustering
+
+Citation for Alex's work:
+\url{https://www.abstractsonline.com/pp8/#!/9070/presentation/2384}
+
+\clearpage
+\newpage
+
+# Supplementary Tables
+
+```{r selmodels, echo=F, tab.align="center"}
+library(knitr)
+library(kableExtra)
+pops <- read.csv("tables/sel_models.txt", sep=";", header=TRUE, check.names=FALSE)
+knitr::kable(pops, "latex", booktabs=T, escape=F,
+    caption="General selection model for diploids and dominance models.") %>%
+kable_styling(latex_options = "HOLD_position")
+```
+
+```{r epistasis, echo=F, tab.align="center"}
+library(knitr)
+library(kableExtra)
+pops <- read.csv("tables/epistasis_model.txt", sep=";", header=TRUE, check.names=FALSE)
+knitr::kable(pops, "latex", booktabs=T, escape=F,
+    caption="Haploid epistasis model.") %>%
+kable_styling(latex_options = "HOLD_position")
+```
+
+```{r 1kgpops, echo=F, tab.align="center"}
+library(knitr)
+library(kableExtra)
+pops <- read.csv("tables/populations.txt", sep=";", header=TRUE, check.names=FALSE)
+knitr::kable(pops, "latex", booktabs=T,
+    caption="Thousand Genomes Project population descriptions for populations used in this study.") %>%
+kable_styling(latex_options = "HOLD_position")
+```
+
+```{r tajimasD, echo=F, tab.align="center"}
+library(knitr)
+library(kableExtra)
+taj <- read.csv("tables/tajimas_d.txt", sep=";", header=TRUE, check.names=FALSE)
+knitr::kable(taj, "latex", booktabs=T, longtable=T,
+    linesep = c("", "", "", "", "", "", "", "", "\\addlinespace"),
+    caption="Tamija's $D$ for classes of coding mutations, both within annotated domains and outside of domains.") %>%
+kable_styling(latex_options = c("HOLD_position", "repeat_header"))
+```
+
+\clearpage
+\newpage
+
+# Supplementary Figures
+
+1) Accuracy of jackknife for varying sample size
+2) Repeat of figures 2, 3, and 4 but for average $D$ instead of $\sigma_d^2$
+3)
+
+```{r genespops, echo=F, fig.cap="(ref:figgenespopscap)", fig.align="center"}
+knitr::include_graphics("figures/supp_gene_wide_all_pops.pdf")
+```
+
+(ref:figgenespopscap) **LD within and between protein-coding genes.** This is
+the caption.
+
+```{r domainspops, echo=F, fig.cap="(ref:figdomainspopscap)", fig.align="center"}
+knitr::include_graphics("figures/supp_domain_all_pops.pdf")
+```
+
+(ref:figdomainspopscap) **LD within and between coding domains and pairs
+outside domains at matched distances.** This is the caption.
+
+```{r lddecay, echo=F, fig.cap="(ref:figlddecay)", fig.align="center"}
+knitr::include_graphics("figures/supp_ld_decay.pdf")
+```
+
+(ref:figlddecay) **LD decay for synonymous and missense mutations for pairs
+of mutations that fall outside of domains.**
+
+```{r lddecaydomain, echo=F, fig.cap="(ref:figlddecaydomain)", fig.align="center"}
+knitr::include_graphics("figures/supp_ld_decay_in_domain.pdf")
+```
+
+(ref:figlddecaydomain) **LD decay for synonymous and missense mutations for pairs
+of mutations that fall inside the same domains.**
+
 \clearpage

### 2021-03-01 Manuscript.Rmd
minor edits throughout

@@
 and synergistic epistasis are both expected to cause negative LD between
-selected mutations, which can be readily tested in population genetic data
+selected mutations, which can be readily tested using population genetic data
 [@Sohail2017-zq; @Sandler2020-vk; @Garcia2020-zn]. Here, I first used a closed
@@
 
-By taking broad genome- or even gene-wide surveys of LD within functional
-classes of mutations, patterns of interactive effects that occur within
+By taking broad genome- or gene-wide surveys of LD across functional classes of
+mutations, heterogeneous patterns of interactive effects that occur within
 functional sub-elements of genes can be missed. However, at the gene-wide
-level, we already find large differences in the interactions between pairs of
-missense and loss-of-function mutations. While LD between pairs of missense
-mutations does not appear to be different from pairs of synonymous variants, LD
-between loss-of-function variants is substantially more negative.
-Hill-Robertson effects under additive selection are expected to be strongest
-for slightly to moderately deleterious variants (with $s\sim 1/N_e$), as
-strongly deleterious mutations are not expected to interfere with one another
-[@McVean2000-ox]. Negative synergistic epistasis between strongly deleterious
-mutations does produce large negative deviations of mean LD. However, strongly
-deleterious *recessive* mutations can also produce this pattern, making it
-difficult to distinguish these effects with certainty. This difficulty is
-compounded by the large measurement noise for $\E[D]$, especially for
-loss-of-function variants for which only a few hundred pairs are found within
-genes in each population.
+level, we already find large differences in interactions between pairs of
+missense and loss-of-function mutations.
+
+From gene-wide averages, LD between pairs of missense mutations does not appear
+to be different from pairs of synonymous variants, but LD between
+loss-of-function variants is substantially more negative. Hill-Robertson
+effects under additive selection are expected to be strongest for slightly to
+moderately deleterious variants (with $s\sim 1/N_e$), as strongly deleterious
+mutations are not expected to interfere with one another [@McVean2000-ox].
+However, loss-of-function mutations do not typically fall within this regime,
+and inference of the distribution of fitness effects (DFE) for new
+loss-of-function variants shows that a large majority are strongly deleterious
+(Supporting Information). Instead, negative synergistic epistasis between
+strongly deleterious mutations does produce large negative deviations of mean
+LD. Strongly deleterious *recessive* mutations can also produce this pattern,
+making it difficult to distinguish these effects with certainty. This
+difficulty is compounded by the large measurement noise for $\E[D]$, especially
+for loss-of-function variants for which only a few hundred pairs are found
+within genes in each population. If loss-of-function variants are in fact
+recessive, even partially, this can have a large impact on the inferences of
+the DFE (Supporting Information).
 
@@
 least no stronger than those between synonymous mutations. However, averaging
-over all observed pairs of mutations within a gene completely masks different
-interactive effects in functional subunits of genes, effects that even drive LD
-in opposite directions. Nonsynonymous mutations found within protein domains,
-identified as conserved subunits of a gene with some structural or functional
-role, are more strongly selected against on average, but also have increased
-signed LD over synonymous mutations at the same distances and also falling
-within domains. Missense mutation outside of domains but at the same distances
-as those within domains have more negative LD, both compared to synonymous
-mutations outside of domains at the same distance and to mutations within
-domains. Neither dominance effects, synergistic epistasis, nor Hill-Robertson
-interference are expected to result in positive LD, so that some other
-interactive effect must be driving this signal within domains.
+over all observed pairs of mutations within a gene masks element-specific
+interactive effects that drive LD in opposite directions. Nonsynonymous
+mutations found within protein domains, identified as conserved subunits of a
+gene with some structural or functional role, are more strongly selected
+against on average, but also have increased signed LD over synonymous mutations
+at the same distances within domains. Missense mutation outside of domains but
+at the same distances as those within domains have more negative LD, both
+compared to distance-matched synonymous mutations outside of domains and to
+mutations within domains. Neither dominance effects, synergistic epistasis, nor
+Hill-Robertson interference are expected to result in positive LD, so that some
+other interactive effect must be driving this positive LD signal.
 
 There are a number of possible interaction scenarios that can result in
-positive signed LD between linked loci. One possible explanation is a
+positive signed LD between linked loci. One explanation is a
 prevalence of pairs of compensatory mutations that are tolerated to
@@
 scenario, double heterozygotes have different fitnesses depending on whether
-the mutations are found on same haplotype or on different haplotypes. Each of
-these scenarios lead to positive LD.
+the mutations are found on same haplotype or on different haplotypes.
 
@@
 
-## Clustered mutations may drive positive LD among synonymous mutations
+## Clustered mutations can account for positive LD among synonymous mutations
 
@@
 population structure is responsible for the increase of LD for both synonymous
-and missense variants, as every natural population lives in some geography that
-will necessarily induce some structure, and they used forward simulations to
-demonstrate this effect. However, this increase in LD, at least in simulations,
-extends over larger distances separating base-pairs than we
-observe in the data. Instead, positive LD between synonymous mutations
-decays to zero within just a few hundred base-pairs (Figures \@ref(fig:lddecay)
-and \@ref(fig:lddecaydomain)). We might also expect that differences in
-migration and admixture histories between human populations would result in
-different levels of positive synonymous LD, but the patterns are fairly
-consistent across populations.
+and missense variants, arguing that every natural population lives in some
+geography that will necessarily induce some structure, and they used forward
+simulations to demonstrate this effect. However, this increase in LD, at least
+in simulations of spatial structure and migration, typically extends over larger
+distances separating base-pairs than we observe in the data. Instead, positive
+LD between synonymous mutations decays to zero within just a few hundred
+base-pairs (Figures \@ref(fig:lddecay) and \@ref(fig:lddecaydomain)).
 
-An alternative explanation for positive LD between neutral variants separated
-by short distances is a nonrandom mutational process, where clusters of
+Nonrandom mutational processes provide an alternative explanation for positive
+LD between neutral variants separated by short distances, in which clusters of
 mutations occur simultaneously in the same mutational event. Such
-multinucleotide mutations have been shown to affect patterns of linkage
-disequilibrium over short distances, on the order of 10s to 100s of base pairs
-[@Harris2014-zg], and clustered mutational events may be common in humans
-[@Besenbacher2016-ac]. Thus, positive LD may be the baseline expectation for
-tightly linked variants primarily due to clustered mutations, and subsequent
-selective processes and interactions will cause LD to deviate from that
-expectation, as seen for missense mutations across different sub-elements of
-genes. Similarly, it may be more appropriate to compare the negative LD
-observed between linked loss-of-function variants to that positive baseline
-expectation instead of zero, implying that they would be more recessive or have
-stronger synergistic epistasis than if the neutral expectation were zero.
+multinucleotide mutations have been shown to affect patterns of LD over short
+distances, on the order of 10s to 100s of base pairs [@Harris2014-zg], and
+clustered mutational events may be common in humans [@Besenbacher2016-ac].
+Thus, positive LD may be the baseline expectation for tightly linked variants
+primarily due to clustered mutations, and only subsequent selective processes and
+interactions will cause LD to deviate from that expectation, as seen for
+missense mutations across different sub-elements of genes. Similarly, it may be
+more appropriate to compare the negative LD observed between linked
+loss-of-function variants to that positive baseline expectation instead of
+zero, implying that they would be more recessive or have stronger synergistic
+epistasis than if the neutral expectation were zero.
 
@@
 
-In exploring or inferring selective models across multiple mutations, the
-number of plausible selection scenarios becomes quite large as we relax
-assumptions about additive and multiplicative fitness functions. The inclusion
-of dominance effects, epistasis, or other interactions leads to a rapid
-increase in the number of parameters to consider. This makes performing forward
-simulations that span the range of all such selective interaction scenarios
-burdensome.
+In exploring or inferring multi-locus selective models, the number of plausible
+selection scenarios becomes quite large as we relax the strict assumptions of
+additive and multiplicative fitness effects. The inclusion of dominance
+effects, epistasis, or other interactions leads to a rapid increase in the
+number of parameters to consider. This makes performing forward simulations
+that span the range of all such selective interaction scenarios burdensome.
 
 Instead, closed, numerical approaches to compute expectations for two-locus
-statistics under a wide range of selection models, as presented here, allows us
+statistics under a wide range of selection models, as presented here, allow us
 to explore this highly parameterized space of models far more efficiently, and
@@
 signed LD is sensitive to the levels of dominance (Figure \@ref(fig:fig3)),
-inferring the distribution of fitness effects with dominance may become
-feasible using the joint distribution of allele frequencies and LD.
+inferring the DFE with dominance may become feasible using the joint
+distribution of allele frequencies and LD.
 

### 2021-03-02 Manuscript.Rmd
add supp figures and msl dfe and mnm analysis

@@
 biblio-style: "genetics"
-title: Can we distinguish modes of selective interactions using linkage disequilibrium?
-((or)) Mutation correlations in conserved domains reveal a heterogeneous landscape of selective interactions within genes ((or...?))
+title: Linkage disequilibrium in conserved domains reveals opposing selective interactions in genes
 author:
@@
 eliminated from a population. The amount of time that a deleterious mutation
-segregates in a population depends on the strength of selection against genomes
-that carry it, with very damaging mutations kept at low frequencies in the
-population and purged relatively rapidly. In the time between mutation and
-fixation or loss, selected variants, both beneficial and damaging, can
-dramatically impact patterns of variation in nearby linked regions. This
-distortion away from neutral expectations has been widely documented in
-sequencing data, but questions remain about the primary mode of interactions
-between multiple linked variants and their joint effects on genome-wide
-patterns of diversity.
+segregates depends on the strength of selection against genomes that carry it,
+with very damaging mutations kept at low frequencies and purged relatively
+rapidly. In the time between mutation and fixation or loss, selected variants,
+both beneficial and damaging, can dramatically impact patterns of variation in
+nearby linked regions. This distortion away from neutral expectations has been
+widely documented in sequencing data, but questions remain about the primary
+mode of interactions between multiple linked variants and their joint effects
+on genome-wide patterns of diversity.
 
@@
 evolutionary models. A large fraction of nonsynonymous coding mutations are
-thought to be at least partially recessive [@Agrawal2011-w; b@Huber2018-cp],
+thought to be at least partially recessive [@Agrawal2011-wb; @Huber2018-cp],
 with average levels of dominance correlating with strength of selection
@@
 loci. To generate predictions under more complex models, we rely on expensive
-forward simulations. These can be great for building intuition or testing
+forward simulations. Such simulations can help build intuition and test
 inference methods, but do not efficiently provide expectations for quantities
@@
 consider expectations for $\sigma_d^2 = \E[D^2]/\E[p(1-p)q(1-q)]$ and
-$\sigma_d^1 = \E[D]/\E[p(1-p)q(1-q)$. These statistics have two benefits:
-first, the mutation rates cancel so that expectations are robust to assumptions
-about the per-base mutation rate, and second, we have analytic expectations for
-these quantities under neutrality at steady state for comparison
-[@Ohta1971-yd].
+$\sigma_d^1 = \E[D]/\E[p(1-p)q(1-q)]$. These statistics have two benefits:
+first, the mutation rate cancels so that expectations are robust to assumptions
+about the per-base mutation rate, and second, analytic expectations for these
+quantities are available under neutrality at constant population size for
+comparison [@Ohta1971-yd].
 
@@
 dominance acting within loci but no epistatic effects, and then describe a
-general diploid selection model and consider simple simple gene-based dominance
+general diploid selection model and consider simple gene-based dominance
 effects. In what follows I focus on parameters where the strengths of selection
@@
 example, in Figure \@ref(fig:fig1)D with $\gamma=-10$, which would be
-considered moderately strong selection, large mean LD extends well beyond 100
-kilobases, with more strongly deleterious interacting mutations expected to
-show large signed LD over much larger recombination distances.
+considered moderate, large mean LD extends well beyond 100 kilobases, with more
+strongly deleterious interacting mutations expected to show large signed LD
+over much larger recombination distances.
 
-### Simple dominance within a locus
+### Dominance and interference
 
 The effects of epistasis between variants on LD has been widely documented.
-However, the role that non-additive selection within a locus affects
-interference and LD between loci has received less attention. @Garcia2020-zn
-used large-scale forward simulation to explore the effect of dominance on
-patterns of LD, showing that LD depends on both the magnitude of the fitness
-and dominance effects in a nonlinear way, but note that expensive large-scale
-simulations are needed to explore these effects as no closed form approach to
-computing those expectations exist. The method presented here provides such a
-closed form approach, so that expectations of signed LD and other two-locus
-statistics can be computed more efficiently and accurately.
+However, the effect of non-additive selection within a locus on interference
+has received less attention. @Garcia2020-zn used large-scale forward simulation
+to explore the effect of dominance on patterns of LD, showing that LD depends
+on both the magnitude of the fitness and dominance effects in a nonlinear way,
+but note that expensive large-scale simulations are needed to explore these
+effects as no closed form approach to computing those expectations exist. The
+method presented here provides such a closed form approach, so that
+expectations of signed LD and other two-locus statistics can be computed more
+efficiently and accurately.
 
@@
 recessive mutations ($h<1/2$) can have large negative signed LD (Figure
-@\ref(fig:fig3)D), although it decays rapidly with increasing distance between
+\@ref(fig:fig3)D), although it decays rapidly with increasing distance between
 loci and is roughly zero for $\rho > 1$. For weakly selected mutations with
@@
 
-The moment system for $\Psi_n$ readily incorporates variable population size or
-mutation and recombination rates or selection parameters that change over time.
-Here I focus on non-equilibrium population size history and consider scenarios
-relevant to human demographic history, including bottlenecks and expansions. I
-explore two simple models (Figure \@ref(fig:fig4)A), one with an instantaneous
-expansion and another with a bottleneck followed by recovery. I also consider
-two demographic histories inferred using genome-wide genealogy reconstruction
-[@Speidel2019-nj] applied to the @1000_Genomes_Project_Consortium2015-zq
-dataset, and focus on size histories for the Yoruba from Ibidan, Nigeria (YRI)
-and Utahns of North and West European ancestry (CEU) (Figure \@ref(fig:fig4)B).
+The moment system for $\Psi_n$ readily incorporates variable population size as
+well as mutation and recombination rates and selection parameters that change
+over time. Here I focus on non-equilibrium population size history and consider
+scenarios relevant to human demographic history, including bottlenecks and
+expansions. I explore two simple models (Figure \@ref(fig:fig4)A), one with an
+instantaneous expansion and another with a bottleneck followed by recovery. I
+also consider two demographic histories inferred using genome-wide genealogy
+reconstruction [@Speidel2019-nj] applied to the
+@1000_Genomes_Project_Consortium2015-zq dataset, and focus on size histories
+for the Yoruba from Ibidan, Nigeria (YRI) and Utahns of North and West European
+ancestry (CEU) (Figure \@ref(fig:fig4)B).
 
@@
 \@ref(fig:domainspops)). Averaging over all pairs of mutations within the same
-gene but outside domains, both synonymous and missense mutations had LD
+gene but outside domains, both synonymous and missense mutations have LD
 slightly above zero, with synonymous mutations observed to have slightly higher
@@
 Previous theoretical and simulation studies have shown that interference and
-interactions between selected mutations reduces the efficacy of selection at
-linked loci [@Hill1968-vu; @Birky1988-jm; @Barton1995-mj; @McVean2000-ox],
-impacting substitution rates, the deleterious mutation load, and dynamics of
-segregating mutations. Interference between moderately deleterious mutations
-and synergistic epistasis are both expected to cause negative LD between
-selected mutations, which can be readily tested using population genetic data
+interactions between selected mutations reduce the efficacy of selection at
+linked loci, impacting substitution rates, the deleterious mutation load, and
+dynamics of segregating mutations [@Hill1968-vu; @Birky1988-jm; @Barton1995-mj;
+@McVean2000-ox]. Interference between moderately deleterious mutations and
+synergistic epistasis are both expected to cause negative LD between selected
+mutations, which can be readily tested using population genetic data
 [@Sohail2017-zq; @Sandler2020-vk; @Garcia2020-zn]. Here, I first used a closed
@@
 within genes in each population. If loss-of-function variants are in fact
-recessive, even partially, this can have a large impact on the inferences of
+recessive, even partially, this can impact inferences of
 the DFE (Supporting Information).
 
-## Non-uniform interactive effects of selected mutations within genes
+## Non-uniform interactions between selected mutations within genes
 
@@
 
-## Clustered mutations can account for positive LD among synonymous mutations
+## Clustered mutations account for positive LD among synonymous mutations
 
@@
 clustered mutational events may be common in humans [@Besenbacher2016-ac].
-Thus, positive LD may be the baseline expectation for tightly linked variants
-primarily due to clustered mutations, and only subsequent selective processes and
-interactions will cause LD to deviate from that expectation, as seen for
-missense mutations across different sub-elements of genes. Similarly, it may be
-more appropriate to compare the negative LD observed between linked
-loss-of-function variants to that positive baseline expectation instead of
-zero, implying that they would be more recessive or have stronger synergistic
+Indeed, a simple exponential model in which the fraction of pairs of mutations
+arising arising from a multinucleotide event decays with distance fits the
+observed decay of $\sigma_d^1$ between synonymous mutations (Supporting
+Information). We may therefore treat positive LD as the baseline expectation
+for tightly linked variants primarily due to clustered mutations, so that
+subsequent selective processes and interactions cause LD to deviate from that
+expectation, as seen for missense mutations across different sub-elements of
+genes. It may also be more appropriate to compare the negative LD observed
+between linked loss-of-function variants to that positive baseline expectation
+instead of zero, so that they are more recessive or have stronger synergistic
 epistasis than if the neutral expectation were zero.
@@
 selection scenarios becomes quite large as we relax the strict assumptions of
-additive and multiplicative fitness effects. The inclusion of dominance
+additivity and multiplicative fitness effects. The inclusion of dominance
 effects, epistasis, or other interactions leads to a rapid increase in the
@@
 can then be compared directly to empirical observations, which can be taken
-from either phased or unphased [@Ragsdale2020-nz]. Additional work performing
-such an analysis is therefore warranted, which should provide a path forward
-for distinguishing between modes of selective interactions.
+from either phased or unphased data [@Ragsdale2020-nz]. Additional work
+performing such an analysis is therefore warranted, which should provide a path
+forward for distinguishing between modes of selective interactions.
 
@@
 @Haller2019-vm). Analytic results for two-locus distributions under selection
-are notoriously difficult, with a few notable flashes of progress.
-@McVean2007-xd considered the effect of a recent sweep on patterns of LD for
+are notoriously difficult, with a few notable flashes of progress. For example,
+@McVean2007-xd considered the effect of a recent sweep on patterns of LD in
 loci nearby or spanning the locus under selection, and in an impressive recent
@@
 rare mutations under additive selection with epistasis. Nonetheless, such
-approaches are confined to steady-state demography and constrained selection
-models.
+approaches are typically confined to steady-state demography and constrained
+selection models.
 
-Numerical methods straddle the two approaches of expensive and noisy discrete
-simulation and analytic solutions, providing a more efficient and practical
-method to compute expectations of two-locus diversity measures under a wider
-range of parameters and demographic scenarios. @Ragsdale2017-gg used a finite
+Numerical methods inhabit the space in between expensive discrete simulations
+and limited analytic solutions, providing a more efficient and practical method
+to compute expectations of two-locus diversity measures under a wider range of
+parameters and demographic scenarios. @Ragsdale2017-gg used a finite
 differencing approach to numerically solve the two-locus diffusion equation
@@
 moments-construction can be used to solve for the two-locus sampling
-distribution within a single population, though requires a moment-closure
-approximation for nonzero recombination and selection. Below, I
-extend this same approach to model arbitrary diploid selection, which
-encompasses dominance, epistasis, and other forms of selective interactions
-between two loci.
+distribution within a single population, though it requires a moment-closure
+approximation for nonzero recombination and selection. Below, I extend this
+approach to model arbitrary diploid selection, which encompasses dominance,
+epistasis, and other forms of selective interactions between two loci.
 
@@
 loci with observed haplotype counts, so that $\Psi_n(i, j, k)$ is the number of
-pairs where with $i$ copies of the $AB$ haplotype, $j$ of type $Ab$, $k$ of
-type $aB$, and $n-i-j-k$ of type $ab$. The size of $\Psi_n$ grows rapidly, with
-$O(n^3)$ entries, which practically limits computational approaches to moderate
-sample sizes and a single population.
+pairs for which we observe $i$ copies of the $AB$ haplotype, $j$ of type $Ab$,
+$k$ of type $aB$, and $n-i-j-k$ of type $ab$. The size of $\Psi_n$ grows
+rapidly, with $O(n^3)$ entries, which practically limits computational
+approaches to moderate sample sizes and a single population.
 
@@
 This requirement of extra lineages for nonzero recombination and selection
-means that the system in \@ref(eq:system) is not closed.
-This means we need a moment-closure approximation to solve for $\Psi_n$. As in
-@Ragsdale2019-nt, a jackknife approximation is used to estimate $\Psi_{n+l}$,
-for $l=1$ or $2$, from $\Psi_n$ [@Jouganous2017-pq], so that
-$\hat{\Psi}_{n+l}(i, j, k) = \mathcal{J}_{n, l}\Psi_n$, although other closure approximations are possible.
-This gives us a closed approximated system,
+means that the system in \@ref(eq:system) is not closed, so that we need a
+moment-closure approximation to solve for $\Psi_n$. As in @Ragsdale2019-nt, a
+jackknife approximation is used to estimate $\Psi_{n+l}$, for $l=1$ or $2$,
+from $\Psi_n$ [@Jouganous2017-pq], so that $\hat{\Psi}_{n+l}(i, j, k) =
+\mathcal{J}_{n, l}\Psi_n$, although other closure approximations are possible.
+This emits a closed approximate system,
 \begin{equation}
@@
 \@ref(fig:jackknife). In the results presented in this paper, sample sizes
-between $n=30$ and $n=80$ were used.
+between $n=30$ and $n=80$ are used.
 
@@
 
-To model selection, we consider a model where we draw lineages uniformly from
+To include selection, we consider a model where we draw lineages uniformly from
 the previous generation, but keep lineages with probability proportional to
-their fitness. In the absence of dominance, the selection model reduces to a
-haploid model, with acceptance and rejection probabilities depending on the
-fitnesses of each haploid copy, where haplotype $Ab$ has fitness $1 + s_A$,
-$aB$ has fitness $1 + s_B$, and $AB$ has fitness $s_{AB}$. We assume the doubly
+their fitness. In the absence of dominance, selection reduces to a haploid
+model, with acceptance and rejection probabilities depending on the fitnesses
+of each haploid copy, where haplotype $Ab$ has fitness $1 + s_{A}$, $aB$ has
+fitness $1 + s_{B}$, and $AB$ has fitness $1 + s_{AB}$. We assume the doubly
 ancestral haplotype $ab$ has fitness $1$, so fitnesses are relative to that of
 $ab$ haplotypes. The standard multiplicative fitness function assumes that
-$s_{AB} \approx s_A + s_B$, and a general epistasis function can be written as
-$s_{AB} = (s_A + s_B) (1 + \epsilon)$, so that $\epsilon>0$ implies synergistic
-epistasis, while $\epsilon < 0$ implies antagonistic epistasis.
+$s_{AB} \approx s_{A} + s_{B}$, and a general epistasis function can be written
+as $s_{AB} = (s_{A} + s_{B}) (1 + \epsilon)$, so that $\epsilon > 0$ implies
+synergistic epistasis, while $\epsilon < 0$ implies antagonistic epistasis.
 
@@
 - Mashaal Sohail
+- Matthias Steinrücken
 

### 2021-03-02 SupportingInformation.Rmd
add supp figures and msl dfe and mnm analysis

@@
 bibliography: "references.bib"
-title: "Supporting Information for [[main paper title]]"
+title: Supporting Information for 
 author:
@@
 \renewcommand{\thetable}{S\arabic{table}}
+\renewcommand{\theequation}{S\arabic{equation}}
 \setcounter{figure}{0}
 \setcounter{table}{0}
+\setcounter{equation}{0}
 
-Yet to include:
-- Tables with data presented in figures 5 and 6
+## The diffusion equation and moment system for the two-locus sampling distribution
 
-## The diffusion equation, etc
-
-\[ 
-\frac{\partial \psi}{\partial t} = 
+The two-locus diffusion equation with additive selection was first described by
+@Kimura1955-qe studied extensively in the 1960s and 70s, including by
+@Hill1966-gv and @Ohta1969-ie. The continuous distribution $\psi(x_1, x_2,
+x_3)$ of haplotype frequencies in a population, where $x_1$ is the frequency of
+$AB$, $x_2$ of $Ab$, and $x_3$ of $aB$, is governed by the multi-dimensional
+Fokker-Planck equation:
+\begin{align} \label{eq:diffeq}
+\frac{\partial \psi}{\partial \tau} = &
 \frac{1}{2}\mathop{\sum\sum}_{1\leq i, j \leq 3} 
 \frac{\partial^2}{\partial x_i \partial x_j} 
-\left[\frac{x_i(\delta_{i=j}-x_j)\psi}{N(t)}\right] 
--\frac{\rho}{2}\left(\frac{\partial}{\partial x}\right) 
-\]
+\left[\frac{x_i(\delta_{i=j}-x_j)\psi}{\nu(\tau)}\right] \\\nonumber
+& -\frac{\rho}{2}\left(-\frac{\partial}{\partial x_1} D\psi
+  + \frac{\partial}{\partial x_2} D\psi
+  + \frac{\partial}{\partial x_3} D\psi\right) \\\nonumber
+& - \frac{\gamma_A}{2}\left[
+  \frac{\partial}{\partial x_1} x_1(1-x_1-x_2)\psi
+  + \frac{\partial}{\partial x_2} x_2(1-x_1-x_2)\psi
+  - \frac{\partial}{\partial x_3} x_3(x_1+x_2)\psi
+  \right] \\\nonumber
+& -\frac{\gamma_B}{2}\left[
+  \frac{\partial}{\partial x_1} x_1(1-x_1-x_3)\psi
+  - \frac{\partial}{\partial x_2} x_2(x_1+x_3)\psi
+  + \frac{\partial}{\partial x_3} x_3(1-x_1-x_3)\psi
+  \right].
+\end{align}
+$D$ is the standard covariance measure of linkage disequilibrium, \[D=x_1 -
+(x_1+x_2)(x_1+x_3) = x_1 x_4 - x_2 x_3,\] $\gamma_A$ and $\gamma_B$ are the
+scaled selection coefficients at the left and right locus, and $\rho$ is the
+scaled recombination rate between the two loci. Time $\tau$ is measured in
+$2N_e$ generations.
 
-## Moment equation and evolution operators
+Given a function $\psi$ that solves Equation \ref{eq:diffeq}, the two-locus
+sampling distribution for a sample size of $n$ haploids can be found by
+integrating $\Psi$ against the multinomial sampling function, so that
+\begin{align} \label{eq:multinomial}
+\Psi_n(i, j, k) =
+{n \choose{i, j, k, n-i-j-k}}
+\mathop{\mathop{\int\int\int}_{x_1, x_2, x_3 \geq0}}_{x_1+x_2+x_3\leq1}
+\psi(x_1, x_2, x_3) x_1^i x_2^j x_3^k (1-x_1-x_2-x_3)^{n-i-j-k}
+dx_1 dx_2 dx_3.
+\end{align}
 
-### Drift
+- instead of (numerically) solving for $\psi$ and using \@ref(eq:multinomial},
+  we can apply the multinomial sampling formula directly to the PDE
+- what we get instead is a system of differential equations on the entries of
+  $\Psi_n$
+- we can then solve for these noncanonical moments of $\psi$, which are the
+  objects of interest in any case, and avoid having to solve this numerical PDE
+  with unpleasant boundary conditions
+- the same system can be found by deriving recursion from considering tracking
+  $n$ lineages within the population, in the vein of @Wright1931-xx
 
-### Recombination
+### Drift, mutation, recombination, and moment closure
 
-### Mutation
+- point to supplement of @Ragsdale2019-nt
 
-### General selection
-
-### Moment closure via jackknife approximation
+### Selection
 
@@
 
+### DFE for missense and LOF variants
+
+Loss-of-function (LOF) variants show a dramatic skew toward low-frequency
+variants across all human populations (Table \@ref(tab:tajimasD)). Here, using
+the folded SFS for synonymous, missense, and LOF mutations across all autosomal
+genes, I infer DFEs for missense and LOF mutations independently. I consider a
+few different dominance coefficients to explore the effect of the assumed
+recessivity of the two classes of mutations.
+
+The standard SFS approach to fitting the DFE involves first inferring a
+demographic history for the population using putatively neutral variants (here,
+synonymous mutations), and then fixing that demography and fitting a
+parameterized function for the distribution of selection coefficients for new
+mutations for the selected classes. DFE inference also requires an estimate for
+the total mutation rate of the different mutation classes, as much of the
+signal for strongly selected mutations comes from observing fewer mutations
+than expected given a known mutation rate (with the assumption that selection
+purges some fraction of strongly deleterious mutations which are unseen in the
+sample). Here, I fit demography and DFEs to the SFS from the Mende in Sierra
+Leone (MSL) using *moments* [@Jouganous2017-pq].
+
+I used the mutation model from @Karczewski2020-le to estimate the total
+mutation rate across autosomal genes ($uL$, where $u$ is the per-base mutation
+rate, and $L$ is the total length of the coding genome). These values were
+$(0.1442, 0.3426, 0.0256)$ for synonymous, missense, and LOF mutations,
+respectively. Roughly two thirds of new mutations in coding regions are
+expected to be missense mutations, while only 5\% of new mutations are LOF. I
+fit a demographic model to the synonymous variants, which included a population
+expansion in the deeper past and exponential growth in the recent past (Figure
+\@ref(fig:msldfe)A). Using the inferred optimal scaled mutation rate,
+$\theta=4N_euL$, I estimated $Ne\approx12,300$, and assuming an average
+generation time of 29 years I converted the inferred genetic units to physical
+units. The best-fit model had a roughly two-fold expansion 400 thousand years
+ago, and then exponential growth over the past 20-30 thousand years, with a
+current effective size of $\sim 63,000$.
+
+Under this demographic model, I fit a gamma distribution for the distribution
+of fitness effects to missense and LOF mutations (Table \@ref(tab:msldfe)). For
+each fit, I fixed the scaled mutation rate for each mutation class, so that
+$\theta_{mis} = \frac{u_{mis}}{u_{syn}} \hat\theta_{syn}$ and $\theta_{lof} =
+\frac{u_{lof}}{u_{syn}} \hat\theta_{syn}$, where values of $u$ were found using
+the GNOMAD mutation model [@Karczewski2020-le]. I tested three values for the
+dominance coefficient $h$, $0$, $0.2$ and $0.5$. For missense mutations, $h=0$
+gave a poor fit to the data, and $h=0.5$ fit best among the three tested
+dominance coefficients. For LOF variants, $h=0$ also fit poorly, but $h=0.2$
+and $h=0.5$ gave similar likelihoods, highlighting that inferring dominance
+using the SFS is poorly constrained. Regardless of the dominance coefficient
+assumed, however, the vast majority of LOF variants were inferred to be
+strongly deleterious, with only $\sim10\%$ of new mutations having selection
+coefficients on the order $1/N_e$ or less.
+
+### Multinucleotide mutations and positive LD between linked synonymous variants
+
+Multinucleotide mutations (MNMs) are complex mutational events that result in
+multiple mutations occurring on the same haplotype background in a single
+generation. Because MNMs fall on the same haplotype, those mutations will be in
+positive LD, and LD between those pairs that are very tightly linked will not
+be broken down all that rapidly. MNMs are expected to occur over relatively short
+distances, on the order of 10s or 100s of base pairs, making them a likely culprit
+of the observed positive LD among synonymous mutations at short distances.
+
+Multinucleotide mutations can be easily incorporated into the moment system with
+a simple adjustment to the mutation operator. Instead of all mutations occurring
+independently in haplotypes with mutations already segregating at the other locus,
+some fraction of new mutations could instead occur spontaneously and create a new
+pair of mutations with initial counts $n_{AB} = 1$ and $n_{ab} = n-1$.
+
+Here, I fit a simple exponential model for the fraction of new mutations at a
+given distance that arose through a MNM event, so that $P(MNM | d) =
+Ae^{-\lambda d}$, where $d$ is the distance separating pairs of mutations. I
+considered all synonymous mutations within genes in the MSL population and used
+the same population size history model as inferred in the DFE section above for
+a demographic control. This left two parameters to be fit, $A$ and $\lambda$,
+which I fit to the binned decay curve of $\sigma_d^1$. I needed to assume an
+average per-base recombination rate $r$ across gene regions, and tested a
+number of values between $10^{-9}$ and $2\times 10^{-8}$. The optimization was
+insensitive to the chosen value or $r$, because the decay of positive LD occurs
+rapidly. For any plausible value of $r$, the $\sigma_d^1$ decays to zero well
+before distances between pairs have scaled recombination rates $\rho=4 N_e r
+d$, and expected statistics for $\rho \ll 1$ vary only negligibly.
+
+In fitting the LD decay of $\sigma_d^1$, the best fit parameters were
+$A=0.132$, and $\lambda=0.0103$. An exponential scaling of $0.01$ implies that
+the vast majority of new mutations do not occur via MNMs for distances greater
+than 200 bp, though a substantial fraction ($10-15\%$) occur via MNMs for very
+tightly linked loci with distances on the order $0-50$ base pairs. It is
+important to note that this does not mean that $10-15\%$ of new mutations occur
+via MNMs, since this fraction is contitioned on two mutations occurring at
+short distances.
+
 ### Grouping Thousand Genomes populations based on clustering
+
+Testing grouping populations that cluster together as a single population.
+
+- Does not reduce CIs
 
@@
 
+```{r msldfe, echo=F, tab.align="center"}
+library(knitr)
+library(kableExtra)
+dfes <- read.csv("tables/msl_dfes.txt", sep=";", header=TRUE, check.names=FALSE)
+knitr::kable(dfes, "latex", booktabs=T, escape=FALSE,
+    linesep = c("", "", "\\addlinespace"),
+    caption="DFEs inferred for missense and loss-of-function variants in MSL for varying values of $h$.
+General patterns are consistent across different chosen values of $h$, although for $h=0$ results in poorer fits for both missense and LOF variants.
+Columns to the right of the log-likelihood (LL) column show proportions of new mutations with $|s|$ in each given bin.") %>%
+kable_styling(latex_options = c("HOLD_position", "repeat_header"))
+```
+
 \clearpage
@@
 
-1) Accuracy of jackknife for varying sample size
-2) Repeat of figures 2, 3, and 4 but for average $D$ instead of $\sigma_d^2$
-3)
+```{r jackknife, echo=F, fig.cap="(ref:figjackknife)", fig.align="center"}
+knitr::include_graphics("figures/jackknife.pdf")
+```
+
+(ref:figjackknife) **Accuracy of the jackknife approximation and runtime**
+Small sample sizes can lead to large error in the closure approximation for
+larger recombination distances or selection coefficients. Generally, the
+jackknife approximation breaks down for recombination rates greater than
+$\rho\approx30$. While increasing sample size leads to more accurate solutions,
+it comes at the cost of both increased runtime and memory usage. Most analyses
+performed in this paper used $n$ between 40 and 70.
 
@@
 
-(ref:figgenespopscap) **LD within and between protein-coding genes.** This is
-the caption.
+(ref:figgenespopscap) **LD within and between protein-coding genes.**
 
@@
 (ref:figdomainspopscap) **LD within and between coding domains and pairs
-outside domains at matched distances.** This is the caption.
+outside domains at matched distances.**
+
+```{r lddecaydomain, echo=F, fig.cap="(ref:figlddecaydomain)", fig.align="center"}
+knitr::include_graphics("figures/supp_ld_decay_in_domain.pdf")
+```
+
+(ref:figlddecaydomain) **LD decay for synonymous and missense mutations for pairs
+of mutations that fall inside the same domains.**
 
@@
 
-```{r lddecaydomain, echo=F, fig.cap="(ref:figlddecaydomain)", fig.align="center"}
-knitr::include_graphics("figures/supp_ld_decay_in_domain.pdf")
+```{r msldemogdfe, echo=F, fig.cap="(ref:figmsldfe)", fig.align="center"}
+knitr::include_graphics("figures/msl_demography_dfes.pdf")
 ```
 
-(ref:figlddecaydomain) **LD decay for synonymous and missense mutations for pairs
-of mutations that fall inside the same domains.**
+(ref:figmsldfe) **Demography and DFE for MSL.** A demographic model was fit to
+the folded synonymous SFS, and DFEs were fit to missense and loss-of-function
+SFS. Shown here are DFEs fit with $h=0.5$.
+
+```{r mslmnms, echo=F, fig.cap="(ref:figmslmnm)", fig.align="center"}
+knitr::include_graphics("figures/msl_mnms.pdf")
+```
+
+(ref:figmslmnm) **Optimization of fraction of new mutations arising via
+multinucleotide mutations by distance.** A simple exponential function was fit
+to describe the probability that a pair of mutations arose through a MSM event
+at a given distance $d$, as $Ae^{-\lambda d}$. Across all recombination rates
+tested, the best fit parameters were $A=0.13$ and $\lambda=0.010$.
 

### 2021-03-02 analysis/compute_LD_unphased_all_decay.py

@@ -1,0 +1,302 @@
+import numpy as np
+import gzip
+import moments.LD
+import sys
+import pickle
+
+
+lof_annotations = [
+    "frameshift_variant",
+    "splice_acceptor_variant",
+    "splice_donor_variant",
+    "start_lost",
+    "stop_gained",
+    "stop_lost",
+    "transcript_ablation",
+]
+
+import tarfile
+
+try:
+    print("already loaded ancestral seqs", len(ancestral_seqs))
+except NameError:
+    ancestral_seqs = {}
+    print("reading ancestral sequences")
+
+    tar = tarfile.open(
+        "/home/aaronragsdale/Data/1000G/human_ancestor/human_ancestor_GRCh37_e59.tar.bz2",
+        mode="r:bz2",
+    )
+    for member in tar.getmembers():
+        if member.isfile() is False:
+            # print(member)
+            continue
+        if member.get_info()["name"].endswith("fa"):
+            f = tar.extractfile(member)
+            ## can't get SeqIO.parse to read in the extracted file
+            lines = f.readlines()
+            chrom = lines[0].decode().split(":")[2]
+            print("found chrom", chrom)
+            ancestral_seqs[chrom] = "".join([l.decode().strip() for l in lines[1:]])
+            # for record in SeqIO.parse(x, "fasta"):
+            #    chrom = record.id.split(":")[2]
+            #    print(chrom)
+            #    ancestral_seqs[chrom] = record.seq
+    tar.close()
+
+
+def get_annotations(annot_file):
+    annots = {}  # pos: alt: (gene, csq)
+    for line in open(annot_file, "r"):
+        chrom, pos, ref, alt, gene, csq = line.split()
+        annots.setdefault(pos, {})
+        annots[pos][alt] = (gene, csq)
+    return annots
+
+
+def flip_gts(gts):
+    gts_out = []
+    for gt in gts:
+        if gt == "0|0":
+            gts_out.append("1|1")
+        elif gt == "0|1":
+            gts_out.append("1|0")
+        elif gt == "1|0":
+            gts_out.append("0|1")
+        elif gt == "1|1":
+            gts_out.append("0|0")
+        else:
+            raise ValueError("unexpected genotype", gt)
+    return gts_out
+
+
+def to_genotypes(gts):
+    g = np.zeros(len(gts))
+    for ii, gg in enumerate(gts):
+        g[ii] = int(gg[0]) + int(gg[2])
+    return g
+
+
+SNPs = ["A", "C", "G", "T"]
+
+
[…]

### 2021-03-02 analysis/fit_MSL_clustered_mutations.py

@@ -1,0 +1,209 @@
+# Fit a model of clustered mutations to the MSL data, using sigma_d^1 and sigma_d^2
+
+import os, sys
+import scipy.optimize
+import pickle
+import numpy as np
+import moments.TwoLocus
+
+data = pickle.load(open("parsed_data/MSL.unphased.decay.all.bp", "rb"))
+
+num_bins = 12
+
+bins = data["bins"][:num_bins]
+bin_mids = np.mean(bins, axis=1)
+
+sd1 = np.array([v[3] / v[2] for v in data["ld"]["synonymous"].values()][:num_bins])
+sd2 = np.array([v[0] / v[2] for v in data["ld"]["synonymous"].values()][:num_bins])
+v1 = np.array(
+    [v ** 2 for v in data["standard_errors"]["synonymous"]["sigmad1"].values()][
+        :num_bins
+    ]
+)
+v2 = np.array(
+    [v ** 2 for v in data["standard_errors"]["synonymous"]["sigmad2"].values()][
+        :num_bins
+    ]
+)
+
+demo_params = [2.194, 5.145, 0.579, 0.0400]
+Ne = 12314.8
+
+n = 40
+
+
+def demo_func(demo_params, rho, n=30, mnm=False):
+    F = moments.TwoLocus.Integration.steady_state(n, rho=rho, clustered_mutations=mnm)
+    F.integrate(demo_params[0], demo_params[2], rho=rho, clustered_mutations=mnm)
+    nu_func = lambda t: demo_params[0] * np.exp(
+        np.log(demo_params[1] / demo_params[0]) * t / demo_params[3]
+    )
+    F.integrate(nu_func, demo_params[3], rho=rho, clustered_mutations=mnm)
+    return F
+
+
+def cache_spectra(demo_params, Ne, bin_mids, r):
+    try:
+        cache = pickle.load(
+            open(f"data/supp/mnm_inference/MSL.cache.n.{n}.r.{r}.bp", "rb")
+        )
+        for k in ["mnm", "ism"]:
+            for b in bin_mids:
+                if b not in cache[k].keys():
+                    rho = 4 * Ne * b * r
+                    print("recomputing", rho)
+                    if k == "mnm":
+                        F = demo_func(demo_params, rho, n=n, mnm=True)
+                        cache["mnm"][b] = F
+                    elif k == "ism":
+                        F = demo_func(demo_params, rho, n=n, mnm=False)
[…]

### 2021-03-02 analysis/fit_MSL_demography_DFE.py

@@ -1,0 +1,344 @@
+# Using the folded SFS in MSL, we fit a demographic history to the synonymous
+# data, which has an ancient expansion, and recent exponential growth.
+# We then fit DFEs to both missense mutations and LOF mutations. For missense
+# mutations, we assume h=0.2, h=0.5, and for LOF, we assume h=0, 0.2, 0.5.
+
+
+## import modules and data
+
+import moments
+import pickle
+import numpy as np
+import scipy.stats
+
+data = pickle.load(open("parsed_data/MSL.frequency_spectra.folded.bp", "rb"))
+
+fs_syn = data["synonymous"]["all"].fold()
+fs_mis = data["missense"]["all"].fold()
+fs_lof = data["loss_of_function"]["all"].fold()
+
+## total mutation rates from Karczewski et al (GNOMAD mutation model)
+# note that these values are u * L, where L is the total length of coding
+# regions across autosomes
+u_syn = 0.1442
+u_mis = 0.3426
+u_lof = 0.0256
+
+
+## Demography inference
+
+
+def model_func(params, ns):
+    nuA, nuF, TA, TF = params
+    fs = moments.Demographics1D.snm(ns)
+    fs.integrate([nuA], TA)
+    nu_func = lambda t: [nuA * np.exp(np.log(nuF / nuA) * t / TF)]
+    fs.integrate(nu_func, TF)
+    return fs
+
+
+p_guess = [2.0, 10.0, 0.3, 0.01]
+lower_bound = [1e-3, 1e-3, 1e-3, 1e-3]
+upper_bound = [10, 100, 1, 1]
+p_guess = moments.Misc.perturb_params(
+    p_guess, lower_bound=lower_bound, upper_bound=upper_bound
+)
+
+
+demo_params = moments.Inference.optimize_log_fmin(
+    p_guess, fs_syn, model_func, lower_bound=lower_bound, upper_bound=upper_bound
+)
+
+
+model = model_func(demo_params, fs_syn.sample_sizes)
+opt_syn_theta = moments.Inference.optimal_sfs_scaling(model, fs_syn)
+Ne = opt_syn_theta / u_syn / 4
+
+print("optimal demog. parameters:", [float(f"{o:.4f}") for o in demo_params])
[…]

### 2021-03-02 analysis/get_frequency_spectra.py

@@ -1,0 +1,239 @@
+# Get the frequency spectra for each mutation type, for all mutations
+# within a gene, for mutations falling in annotated domains, and for
+# mutations falling outside of annotated domains.
+
+import numpy as np
+import gzip
+import moments
+import sys
+import pickle
+
+lof_annotations = [
+    "frameshift_variant",
+    "splice_acceptor_variant",
+    "splice_donor_variant",
+    "start_lost",
+    "stop_gained",
+    "stop_lost",
+    "transcript_ablation",
+]
+
+import tarfile
+
+try:
+    print("already loaded ancestral seqs", len(ancestral_seqs))
+except NameError:
+    ancestral_seqs = {}
+    print("reading ancestral sequences")
+
+    tar = tarfile.open(
+        "/home/aaronragsdale/Data/1000G/human_ancestor/human_ancestor_GRCh37_e59.tar.bz2",
+        mode="r:bz2",
+    )
+    for member in tar.getmembers():
+        if member.isfile() is False:
+            # print(member)
+            continue
+        if member.get_info()["name"].endswith("fa"):
+            f = tar.extractfile(member)
+            ## can't get SeqIO.parse to read in the extracted file
+            lines = f.readlines()
+            chrom = lines[0].decode().split(":")[2]
+            print("found chrom", chrom)
+            ancestral_seqs[chrom] = "".join([l.decode().strip() for l in lines[1:]])
+            # for record in SeqIO.parse(x, "fasta"):
+            #    chrom = record.id.split(":")[2]
+            #    print(chrom)
+            #    ancestral_seqs[chrom] = record.seq
+    tar.close()
+
+
+def get_annotations(annot_file):
+    annots = {}  # pos: alt: (gene, csq)
+    for line in open(annot_file, "r"):
+        chrom, pos, ref, alt, gene, csq = line.split()
+        annots.setdefault(pos, {})
+        annots[pos][alt] = (gene, csq)
+    return annots
+
+
+def flip_gts(gts):
+    gts_out = []
+    for gt in gts:
+        if gt == "0|0":
+            gts_out.append("1|1")
+        elif gt == "0|1":
+            gts_out.append("1|0")
+        elif gt == "1|0":
+            gts_out.append("0|1")
+        elif gt == "1|1":
+            gts_out.append("0|0")
+        else:
+            raise ValueError("unexpected genotype", gt)
+    return gts_out
+
+
+def to_haplotypes(gts):
+    h = np.zeros(2 * len(gts))
[…]

### 2021-03-02 analysis/get_frequency_spectra_folded.py

@@ -2,6 +2,8 @@
 # within a gene, for mutations falling in annotated domains, and for
 # mutations falling outside of annotated domains.
 
+# Ignore ancestral information, so keeping all SNPs, and return folded SFS
+
 import numpy as np
 import gzip
 import moments
@@ -18,35 +20,6 @@ lof_annotations = [
     "transcript_ablation",
 ]
 
-import tarfile
-
-try:
-    print("already loaded ancestral seqs", len(ancestral_seqs))
-except NameError:
-    ancestral_seqs = {}
-    print("reading ancestral sequences")
-
-    tar = tarfile.open(
-        "/home/aaronragsdale/Data/1000G/human_ancestor/human_ancestor_GRCh37_e59.tar.bz2",
-        mode="r:bz2",
-    )
-    for member in tar.getmembers():
-        if member.isfile() is False:
-            # print(member)
-            continue
-        if member.get_info()["name"].endswith("fa"):
-            f = tar.extractfile(member)
-            ## can't get SeqIO.parse to read in the extracted file
-            lines = f.readlines()
-            chrom = lines[0].decode().split(":")[2]
-            print("found chrom", chrom)
-            ancestral_seqs[chrom] = "".join([l.decode().strip() for l in lines[1:]])
-            # for record in SeqIO.parse(x, "fasta"):
-            #    chrom = record.id.split(":")[2]
-            #    print(chrom)
-            #    ancestral_seqs[chrom] = record.seq
-    tar.close()
-
 
 def get_annotations(annot_file):
     annots = {}  # pos: alt: (gene, csq)
@@ -55,22 +28,6 @@ def get_annotations(annot_file):
         annots.setdefault(pos, {})
         annots[pos][alt] = (gene, csq)
     return annots
-
-
-def flip_gts(gts):
-    gts_out = []
-    for gt in gts:
-        if gt == "0|0":
-            gts_out.append("1|1")
-        elif gt == "0|1":
-            gts_out.append("1|0")
-        elif gt == "1|0":
-            gts_out.append("0|1")
-        elif gt == "1|1":
-            gts_out.append("0|0")
-        else:
-            raise ValueError("unexpected genotype", gt)
-    return gts_out
 
 
 def to_haplotypes(gts):
@@ -96,6 +53,8 @@ def load_genotype_matrix(chrom, pop, haplotypes=True):
     annotations = []
     genes = []
     G = []
+    kept = 0
+    discarded = 0
[…]

### 2021-03-02 figures/jackknife.py

@@ -1,0 +1,103 @@
+import moments.TwoLocus
+import numpy as np
+import matplotlib.pylab as plt
+
+import matplotlib
+plt.rcParams["legend.title_fontsize"] = "xx-small"
+matplotlib.rc("xtick", labelsize=6)
+matplotlib.rc("ytick", labelsize=6)
+matplotlib.rc("axes", labelsize=7)
+matplotlib.rc("axes", titlesize=7)
+matplotlib.rc("legend", fontsize=6)
+
+from matplotlib.ticker import ScalarFormatter
+ns = [10, 15, 20, 25, 30, 35, 40, 50, 60, 70]
+t_jk = np.array(
+    [1.01, 1.48, 3.03, 7.65, 20.58, 47.07, 102.44, 412.94, 1192.44, 3526.23]
+)
+mem_jk = np.array(
+    [131828, 135580, 141608, 157068, 187996, 253480, 345692, 700044, 1601448, 3543684]
+)
+t_no_jk = np.array(
+    [0.87, 0.93, 1.02, 1.43, 2.21, 4.24, 8.31, 26.66, 77.12, 280.66]
+)
+mem_no_jk = np.array(
+    [130864, 133832, 140784, 156116, 185932, 247812, 341120, 692348, 1598728, 3541400]
+)
+
+
+
+rhos_ok = np.logspace(-1, 2, 50)
+ohta_kimura = (5 + rhos_ok / 2) / (11 + 13 * rhos_ok / 2 + rhos_ok ** 2 / 2)
+rhos = np.logspace(-1, 2, 11)
+ok_compare = (5 + rhos / 2) / (11 + 13 * rhos / 2 + rhos ** 2 / 2)
+# precomputed using `F = moments.TwoLocus.Demographics.equilibrium(n, rho=rho)`
+# and then `F.D2() / F.pi2()`
+ld_curve_moments = {
+    20: [0.4332, 0.4138, 0.3799, 0.3264, 0.2547, 0.1774, 0.1108, 0.0634, 0.0339, 0.0172, 0.0045],
+    30: [0.4332, 0.4139, 0.3801, 0.3269, 0.2556, 0.1786, 0.1121, 0.0646, 0.035, 0.0248, 0.0074],
+    50: [0.4333, 0.414, 0.3803, 0.3272, 0.2562, 0.1794, 0.1128, 0.0652, 0.0356, 0.0186, 0.2883],
[…]

### 2021-03-03 Manuscript.Rmd
add simulation section to methods

@@
 To obtain the recursion equation under selection we consider drawing $n$
-lineages from generation $t$, and this sampling has an expected distribution of
-haplotype counts given by $\Psi_n^t$. However, assuming $s\leq0$ for each
-haplotype, each of those sampled lineages as probability of being rejected
+lineages from generation $t$, which has has an expected sampling distribution
+of haplotype counts given by $\Psi_n^t$. However, assuming $s\leq0$ for each
+haplotype, each of those sampled lineages has probability of being rejected
 equal to the relative fitness of its haplotype state. If a lineage is rejected,
-a replacement is drawn from among the population. Under the assumption that
-$ns_{\cdot\cdot} \ll 1$, the probability that more than two selection events
-occur in any given generation is negligibly small, so that the case of multiple
-simultaneous rejections can be ignored, and $Psi_n^{t+1}$ relies only on
-$\Psi_n^t$ and $\Psi_{n+1}^t$ for additive selection. The full selection
-operator $\mathcal{S}$ is given in the Supporting Information.
+a replacement is drawn from the full population. Under the assumption that
+$ns_{\cdot\cdot} \ll 1$, the probability that more than one selection event
+occurs in any given generation is negligibly small, so that the case of
+multiple simultaneous rejections can be ignored. Then $Psi_n^{t+1}$ relies only
+on $\Psi_n^t$ and $\Psi_{n+1}^t$ for additive selection. The full selection
+operator $\mathcal{S}$ for additive selection is given in the Supporting
+Information.
 
@@
 unphased data. In this paper, I focus on $\sigma_d^2 = \E[D^2] /
-\E[p(1-p)q(1-q)]$ and $\sigma_d^1 = \E[D²] / \E[p(1-p)q(1-q)]$, averaged over
-alleles at all frequencies. However, allele-frequency conditioned statistics
-(e.g., keeping only loci below some frequency threshold [@Good2020-wj]) can be
-considered using this same approach.
+\E[p(1-p)q(1-q)]$ and $\sigma_d^1 = \E[D] / \E[p(1-p)q(1-q)]$, averaged over
+pairs of variants at all frequencies. However, allele-frequency conditioned
+statistics (such as keeping only loci below some frequency threshold as in
+@Good2020-wj) can be considered using this same approach.
 
@@
 
+I considered four variable population size histories, two simple toy models and
+two inferred from human populations in African and Europe. For each size
+history scenario, I tracked the evolution of $\Psi_n(t)$ for varying selection
+models, and Figure \@ref(fig:fig4) plots the trajectory of $\sigma_d^1$ over
+time. The selection strength at both loci was fixed at $\gamma=2$ for all
+scenarios, and I considered additive selection without epistasis for fully
+linked ($\rho=0$) and more loosely linked ($\rho=5$) loci. With $\rho=1$, I
+also considered positive and negative epistasis ($\epsilon=0.5$ and $-0.5$),
+recessive loci ($h=0.1$), and a gene-based dominance model with $h=0.1$.
+
+The simple size change models both had ancestral $N_e=10,000$, with one a
+$3$-fold population expansion that occurs 3,000 generations ago, and the other
+a $10$-fold reduction 2,000 generations ago followed by an expansion to double
+the ancestral size $1,000$ generations ago. The size histories for YRI and CEU
+were inferred using Relate [@Speidel2019-nj] applied to the phased autosomal
+data from @1000_Genomes_Project_Consortium2015-zq, using default parameters as
+recommended in the Relate tutorial. Relate returns estimates of coalescence
+rates given time bins, and population sizes are estimated as their inverses.
+Estimates for population sizes in the very recent past ($<2,000$ years)
+diverged, so I truncated the history over this time period and assumed a
+constant size from the most recent non-diverged bin.
+
 ## Analysis of human genomic data
@@
 Using the annotated variant call format (VCF) files from the phase 3
-@1000_Genomes_Project_Consortium2015-zq (Thousand Genomes) data release to
+@1000_Genomes_Project_Consortium2015-zq (Thousand Genomes) data release, I
 subset the genotype VCFs of the same release to autosomal variants that were
@@
 @Ragsdale2020-nz. This approach provides unbiased estimates for pairwise LD,
-under the assumption of a random mating population. I considered pairs of
+under the assumption that individuals are not inbred. I considered pairs of
 mutations within the same mutation class (synonymous, missense, and
 loss-of-function) either within the same gene, in different genes within
-distance bins up to 500kb, or inside or outside of annotated domains within the
-same protein-coding genes. I used a dataset of all known conserved domains in
-humans compiled by @Stanek2020-pa.
+distance bins up to 500kb, and inside or outside of annotated domains within
+the same protein-coding genes. I used a dataset of all known conserved domains
+in humans compiled by @Stanek2020-pa.
 
@@
 methods are packaged within *moments*, and analyses here were performed using
-version 1.1.0 (\url{https://bitbucket.org/simongravel/moments/src/master/}).
-Scripts to run all analyses, recreate figures, and compile this manuscript are
-available at \url{https://github.com/apragsdale/two_locus_selection}.
+*moments* version 1.1.0, available from
+\url{https://bitbucket.org/simongravel/moments/src/master/} with extensive
+documentation at \url{https://moments.readthedocs.org}. Scripts to run all
+analyses, recreate figures, and compile this manuscript are available at
+\url{https://github.com/apragsdale/two_locus_selection}.
 
@@
 results in negative LD, while dominance acting at the gene level results in
-positive LD. Again, bottlenecks result in larger average LD in both cases.
+positive LD. Here, $h=0.1$. Again, bottlenecks result in larger average LD in
+both cases.
 

### 2021-03-07 Manuscript.Rmd
fill in supplement

@@
 biblio-style: "genetics"
-title: Linkage disequilibrium in conserved domains reveals opposing selective interactions in genes
+title: Can we distinguish modes of selective interactions using linkage disequilibrium?
 author:
@@
 both beneficial and damaging, can dramatically impact patterns of variation in
-nearby linked regions. This distortion away from neutral expectations has been
-widely documented in sequencing data, but questions remain about the primary
-mode of interactions between multiple linked variants and their joint effects
-on genome-wide patterns of diversity.
+nearby linked regions. This distortion away from neutral expectations is now
+empirically documented using sequencing data from an ever-growing set of study
+systems, but questions remain about the primary mode of interactions between
+multiple linked variants and their joint effects on genome-wide patterns of
+diversity.
 

### 2021-03-07 SupportingInformation.Rmd
fill in supplement

@@
 
-# Supporting Information
-
 \renewcommand{\thefigure}{S\arabic{figure}}
@@
 
-## The diffusion equation and moment system for the two-locus sampling distribution
+# The diffusion equation and moment system for the two-locus sampling distribution
 
@@
 scaled recombination rate between the two loci. Time $\tau$ is measured in
-$2N_e$ generations.
+$2N_e$ generations, and $\nu(\tau)$ is the population size relative
+to the ancestral size or some reference size at time $\tau$.
 
@@
 
-- instead of (numerically) solving for $\psi$ and using \@ref(eq:multinomial},
-  we can apply the multinomial sampling formula directly to the PDE
-- what we get instead is a system of differential equations on the entries of
-  $\Psi_n$
-- we can then solve for these noncanonical moments of $\psi$, which are the
-  objects of interest in any case, and avoid having to solve this numerical PDE
-  with unpleasant boundary conditions
-- the same system can be found by deriving recursion from considering tracking
-  $n$ lineages within the population, in the vein of @Wright1931-xx
+In the method-of-moments approach, instead of solving the differential equation
+for $\psi$, we instead integrate both sides of the differential equation
+against the multinomial sampling function for a given sampling configuration
+$(i, j, k)$. On the left side, we get $\partial_t \Psi_n(i, j, k)$, and on the
+right we obtain, after some simple integration by parts and somewhat tedious
+simplification, terms for drift, recombination, and selection that can be
+written as sparse linear operators of $\Psi_n$. Written compactly, this takes
+the form
+\begin{equation}
+\partial_\tau \Psi_n = 
+\frac{1}{2\nu(\tau)}\mathcal{D}_{n}\Psi_n
++ \frac{\rho}{2}\mathcal{R}_{n}\Psi_n
++ \frac{\theta}{2}\mathcal{U}_{n}\Psi_n
++ \mathcal{S}_{n, \mathbf{\gamma}, \mathbf{h}}\Psi_n.
+\end{equation}.
 
-### Drift, mutation, recombination, and moment closure
+Alternatively, we arrive at this same linear system of equations by considering
+tracking the expected sampling distribution over $n$ lineages within the full
+population and how that changes over time by drawing lineages from one
+generation to the next in the style of @Wright1931-wy. Both @Jouganous2017-pq,
+for the single-locus SFS, and @Ragsdale2019-nt drew this connection in detail,
+so I refer readers to that previous work for a fuller description of those
+derivations and discussion. In the next section I repeat the results for
+$\mathcal{D}$, $\mathcal{R}$, and $\mathcal{U}$, briefly describe the moment
+closure approximation (which is the same as presented in @Ragsdale2019-nt), and
+then describe the selection operator $\mathcal{S}$ for selection with
+epistasis, both with and without dominance.
 
-- point to supplement of @Ragsdale2019-nt
+## Drift, mutation, recombination, and moment closure
 
-### Selection
+### Drift
 
-## Data analysis
+Drift for an entry $(i, j, k)$ depends only on $\Psi_n$ and therefore closes.
+The entries of $\mathcal{D}$ are found by considering the possibility of a
+coalescence event occurring within a given generation within $n$ lineages in
+the full population. If $n \ll N$, we can safely assume that at most a single
+such event occurs in any given generation.
 
-### DFE for missense and LOF variants
+\begin{align}
+\mathcal{D}_{n}(i, j, k)\Psi_n =
+& (i - 1) (n - i - j - k + 1) \Psi_n(i - 1, j, k) \\\nonumber
+& + (i + 1) (n - i - j - k - 1) \Psi_n(i + 1, j, k) \\\nonumber
+& + (i - 1) (k + 1) \Psi_n(i - 1, j, k + 1) \\\nonumber
+& + (i + 1) (k - 1) \Psi_n(i + 1, j, k - 1) \\\nonumber
+& + (i - 1) (j + 1) \Psi_n(i - 1, j + 1, k) \\\nonumber
+& + (i + 1) (j - 1) \Psi_n(i + 1, j - 1, k) \\\nonumber
+& + (j - 1) (n - i - j - k + 1) \Psi_n(i, j - 1, k) \\\nonumber
+& + (j + 1) (n - i - j - k - 1) \Psi_n(i, j + 1, k) \\\nonumber
+& + (j - 1) (k + 1) \Psi_n(i, j - 1, k + 1) \\\nonumber
+& + (j + 1) (k - 1) \Psi_n(i, j + 1, k - 1) \\\nonumber
+& + (k - 1) (n - i - j - k + 1) \Psi_n(i, j, k - 1) \\\nonumber
+& + (k + 1) (n - i - j - k - 1) \Psi_n(i, j, k + 1) \\\nonumber
+& - 2\left( i ( n - i - j - k) + i k + i j + j (n - i - j - k) + j k + k(n - i - j - k)\right)\Psi_n(i, j, k)
+\end{align}
+
+### Recombination
+
+If a lineage in our sample of size $n$ recombines in a given generation, which
+occurs with probability $nr$, we need to draw an extra lineage from the full
+population for it to recombine with. This means we need $\Psi_{n+1}$ in the
+previous generation. After drawing that extra lineage, $\Psi_n$ changes as we
+draw one of the two recombinant types (each with probability $1/2$$ instead of
+the lineage that was chosen to recombine.
+
+\begin{align}
+\mathcal{R}_{n}(i, j, k)\Psi_n = &
+\frac{(i + 1)(n - i - j - k + 1)}{n + 1}\Psi_{n+1}(i + 1, j - 1, k) \\\nonumber
+& + \frac{(i + 1)(n - i - j - k + 1)}{n + 1}\Psi_{n+1}(i + 1, j, k - 1) \\\nonumber
+& + \frac{(j + 1)(k + 1}{n + 1)}\Psi_{n+1}(i - 1, j + 1, k + 1) \\\nonumber
+& + \frac{(j + 1)(k + 1}{n + 1)}\Psi_{n+1}(i, j + 1, k + 1) \\\nonumber
+& - \frac{(i + 1)(n - i - j - k)}{n + 1}\Psi_{n+1}(i + 1, j, k) \\\nonumber
+& - \frac{(j + 1) k}{n + 1}\Psi_{n+1}(i, j + 1, k) \\\nonumber
+& - \frac{j (k + 1)}{n + 1}\Psi_{n+1}(i, j, k + 1) \\\nonumber
+& - \frac{i (n - i - j - k + 1}{n + 1}\Psi_{n+1}(i, j, k)
+\end{align}
+
+### Mutation
+
+We assume an infinite sites mutation (ISM) model where new mutations occur at
+previously unmutated loci. In the two-locus ISM model, two-locus pairs of
+variable loci arise when a mutation occurs at one locus when the other locus is
+already variable. Thus, new mutations at the $B/b$ locus occur against the
+single-locus allele frequency distribution $\Phi_{n,A}$, and new mutations at
+the $A/a$ locus occur against $\Phi_{n, B}$, which are found via the single-locus
+system from @Jouganous2017-pq.
+
+\begin{align}
+\mathcal{U}_{n}(i, j, k)\Psi_n = & 
+(j + 1)\frac{\theta_B}{2} \Phi_{n, A}(j+1) \delta_{i=1, k=0} \\\nonumber
+& + (n - j)\frac{\theta_B}{2} \Phi_{n, A}(j) \delta_{i=0, k=1} \\\nonumber
+& + (i + 1)\frac{\theta_A}{2} \Phi_{n, B}(i+1) \delta_{i=1, j=0} \\\nonumber
+& + (n - i)\frac{\theta_A}{2} \Phi_{n, B}(i) \delta_{i=0, j=1} \\\nonumber
+\end{align}
+
+### Jackknife moment closure approximation
+
+We use a jackknife approximation to write the entries of $\Psi_{n+1}$ and
+$\Psi_{n+2}$ as linear combinations of entries in $\Psi_n$. The general
+strategy is to assume the underlying continuous distribution $\psi(x, y, z)$
+can be approximated locally as a quadratic, and then use entries in $\Psi_n$
+that are close in frequency to a given entry in $\Psi_{n+l}$ to estimate the
+coefficients of that quadratic using the multinomial sampling formula. Then
+this quadratic local approximation to $\psi$ can be used to compute
+$\Psi_{n+l}(i, j, k)$ using Eq. \@ref(eq:multinomial). Readers should refer to
+section S1.3.5 in the Supporting material for @Ragsdale2019-nt for details.
+
+## Selection
+
+First consider the case of no dominance, so that the haplotypes $Ab$, $aB$,
+and $AB$ have selection coefficients $s_{Ab}$, $s_{aB}$, and $s_{AB}$,
+respectively. Note that the case with $s_{AB} = s_{Ab} + s_{aB}$ implies no
+epistasis between the $A/a$ and $B/b$ loci. Here, we assume all selection
+coefficients are negative. In a given generation, a selection event could
+occur in which a haplotype is rejected (selected against) with probability
+proportional to its selection coefficient, $-s$. We then draw an extra lineage
+from the full population to replace that rejected lineage.
+
+For example, the probability that an $AB$ haplotype is selected against and
+replaced by an $Ab$ haplotype is \[-ns_{AB} \frac{i}{n+1}{j+1}{n}\Psi_{n+1}(i,
+j + 1, k),\] where the additional $j + 1$ lineage in a sample of size $n + 1$
+accounts drawing that extra $Ab$ haplotype. Taking all such selective events
+together, for additive selection we get
+\begin{align}
+\mathcal{S}_{n}(i, j, k)\Psi_n = &
+\frac{i+1}{n+1}\left(-s_{AB}(n-i) + s_{Ab}j
++ s_{aB}k\right)\Psi_{n+1}(i+1, j, k) \\\nonumber
+& +\frac{j+1}{n+1}\left(s_{AB}i - s_{Ab}(n-j)
++ s_{aB}k\right)\Psi_{n+1}(i, j+1, k) \\\nonumber
+& +\frac{k+1}{n+1}\left(s_{AB}i + s_{Ab}j
+- s_{aB}(n-k)\right)\Psi_{n+1}(i, j, k+1) \\\nonumber
+& + \frac{n-i-j-k+1}{n+1}\left(s_{AB}i + s_{Ab}j
++ s_{aB}k\right)\Psi_{n+1}(i, j, k)
+\end{align}
+
+For a general diploid selection model, the idea is nearly the same, but we need
+to draw an extra lineage to determine the fitness of a diploid individual. For
+example, the probability that an $AB$ haplotype is paired with an additional
+lineage $Ab$ and selected against, and then replaced by an $aB$ haplotype is
+\[-ns_{AB/Ab}\frac{i}{n+2}\frac{j+1}{n+1}\frac{k+1}{n}\Psi_{n+2}(i, j+1,
+k+1).\] There are now many more possible selective events to consider, but
+after accounting for all possible diploid pairs and replacements (90 in total)
+and simplifying, we find
+\begin{align}
+\mathcal{S}_{n}(i, j, k)\Psi_n = &
+\frac{n-i-j-k+2}{n+2}\frac{n-i-j-k+1}{n+1}\left(
+  s_{AB/ab}i 
+  + s_{Ab/ab}j 
+  + s_{aB/ab}k
+\right)\Psi_{n+2}(i,j,k) \\\nonumber
+& + \frac{i+1}{n+2}\frac{n-i-j-k+1}{n+1}(
+  s_{AB/AB}i
+  + s_{AB/Ab}j
+  + s_{AB/aB}k
+  + s_{Ab/ab}j \\\nonumber & \hspace{20pt}
+  + s_{aB/ab}k
+  - s_{AB/ab}(n + j + k)
+)\Psi_{n+2}(i+1, j, k) \\\nonumber
+& + \frac{i+2}{n+2}\frac{i+1}{n+1}(
+  s_{AB/Ab}j
+  + s_{AB/aB}k
+  + s_{AB/ab}(n-i-j-k)
+  - s_{AB/AB}(n-i)
+)\Psi_{n+2}(i+2, j, k) \\\nonumber
+& + \frac{i+1}{n+2}\frac{j+1}{n+2}(
+  s_{AB/AB}i
+  + s_{AB/aB}k
+  + s_{AB/ab}(n-i-j-k)
+  + s_{Ab/Ab}j \\\nonumber & \hspace{20pt}
+  + s_{Ab/aB}k
+  + s_{Ab/ab}(n-i-j-k)
+  - s_{AB/Ab}(2n-i-j)
+)\Psi_{n+2}(i+1, j+1, k) \\\nonumber
+& + \frac{i+1}{n+2}\frac{k+1}{n+1}(
+  s_{AB/AB}i
+  + s_{AB/Ab}j
+  + s_{AB/ab}(n-i-j-k)
+  + s_{Ab/aB}j \\\nonumber & \hspace{20pt}
+  + s_{aB/aB}k
+  + s_{aB/ab}(n-i-j-k)
+  - s_{AB/aB}(2n-i-k)
+)\Psi_{n+2}(i+1, j, k+1) \\\nonumber
+& + \frac{j+1}{n+2}\frac{n-i-j-k+1}{n+1}(
+  s_{AB/Ab}i
+  + s_{AB/ab}i
+  + s_{Ab/Ab}j
+  + s_{Ab/aB}k \\\nonumber & \hspace{20pt}
+  + s_{aB/ab}k
+  -s_{Ab/ab}(n+i+k)
+)\Psi_{n+2}(i, j+1, k) \\\nonumber
+& + \frac{j+2}{n+2}\frac{j+1}{n+1}(
+s_{AB/Ab}i + s_{Ab/aB}k + s_{Ab/ab}(n-i-j-k) - s_{Ab/Ab}(n-j)
+)\Psi_{n+2}(i, j+2, k) \\\nonumber
+& + \frac{j+1}{n+2}\frac{k+1}{n+1}(
+  s_{AB/Ab}i
+  + s_{AB/aB}i
+  + s_{Ab/Ab}j
+  + s_{Ab/ab}(n-i-j-k) \\\nonumber & \hspace{20pt}
+  + s_{aB/aB}k
+  + s_{aB/ab}(n-i-j-k)
+  - s_{Ab/aB}(2n - j - k)
+)\Psi_{n+2}(i, j+1, k+1) \\\nonumber
+& + \frac{k+1}{n+2}\frac{n-i-j-k+1}{n+1}(
+  s_{AB/aB}i
+  + s_{AB/ab}i
+  + s_{Ab/aB}j
+  + s_{Ab/ab}j \\\nonumber & \hspace{20pt}
+  + s_{aB/aB}k
+  - s_{aB/ab}(n + i + j)
+)\Psi_{n+2}(i, j, k+1) \\\nonumber
+& + \frac{k+2}{n+2}\frac{k+1}{n+1}(
+  s_{AB/aB}i
+  + s_{Ab/aB}j
+  + s_{aB/ab}(n-i-j-k)
+  - s_{aB/aB}(n - k)
+)\Psi_{n+2}(i, j, k+2).
+\end{align}
+Multiplying through by $2N_{ref}$ gives us selection operators in terms of
+$\gamma$ instead of $s$.
+
+# Data analysis
+
+## DFE for missense and LOF variants
 
@@
 
-### Multinucleotide mutations and positive LD between linked synonymous variants
+## Multinucleotide mutations and positive LD between linked synonymous variants
 
@@
 $A=0.132$, and $\lambda=0.0103$. An exponential scaling of $0.01$ implies that
-the vast majority of new mutations do not occur via MNMs for distances greater
-than 200 bp, though a substantial fraction ($10-15\%$) occur via MNMs for very
-tightly linked loci with distances on the order $0-50$ base pairs. It is
-important to note that this does not mean that $10-15\%$ of new mutations occur
-via MNMs, since this fraction is contitioned on two mutations occurring at
-short distances.
+the vast majority of new mutations *pairs* do not occur via MNMs for distances
+greater than 200 bp, though a substantial fraction ($10-15\%$) occur via MNMs
+for very tightly linked loci with distances on the order $0-50$ base pairs. It
+is important to note that this does not mean that $10-15\%$ of new mutations
+occur via MNMs, since this fraction is contitioned on two mutations occurring
+at short distances.
 
-### Grouping Thousand Genomes populations based on clustering
+## Grouping Thousand Genomes populations based on clustering
 
-Testing grouping populations that cluster together as a single population.
+The large confidence intervals for measurements of signed LD could be driven by
+either averaging over relatively few observed pairs of mutations, or due to
+small sample sizes that make each individual measurement noisy estimates of the
+LD for that pair of mutations in the full population. To explore the underlying
+cause of measurement uncertainty in the @1000_Genomes_Project_Consortium2015-zq
+data, I considered larger sets of samples by combining populations that
+consistently cluster together in PCA and UMAP space and have low
+differentiation [@Diaz-Papkovich2020-ee]. I took combinations of CEU/GBR,
+CHB/CHS, CDX/KHV, and MSL/GWD. While recognizing that residual population
+structure in these population combinations could alter expected LD statistics
+compared to the respective single-population estimates, I was more interested
+in the effect that increasing the sample sizes would have on estimated
+measurement error.
 
-- Does not reduce CIs
-
-Citation for Alex's work:
-\url{https://www.abstractsonline.com/pp8/#!/9070/presentation/2384}
+Across each of the four combinations tested, confidence intervals were roughly
+equivalent to those of each of the individual populations. This suggests that
+the limiting factor to accurate LD measurement is not sample size but rather
+the overall levels of diversity and number of pairs of mutations that we
+compare. $\mathbb{E}[D]$ is most affected by common variants, and the sample
+sizes of the Thousand Genomes Project data are likely sufficient to accurately
+estimate common allele frequencies. Adding additional samples will increase the
+number of rare variants that we observe, but rare variants have minimal impact
+on $\sigma_d^1$. Thus, the accuracy of estimates of $\sigma_d^1$ is more
+fundamentally limited by evolutionary history and genome biology (i.e. past
+population sizes, mutation and recombination rates) than by sample sizes.
 

### 2021-03-08 Manuscript.Rmd
edits

@@
 how well we can differentiate modes of selective interactions in practice. I
-find that observations of positive LD between missense mutations within genes
-is driven by strong positive correlations between pairs of mutations that fall
-within the same conserved domain, pointing to compensatory mutations or
-antagonistic epistasis as the prevailing mode of interaction within but not
-outside of conserved genic elements. The heterogeneous landscape of both
-strength of selection and selective interactions within protein-coding genes
-calls for more refined inferences of the joint distribution of fitness and
-interactive effects, and the methods presented here should prove useful in that
-pursuit.
+find that positive LD between missense mutations within genes is driven by
+strong positive correlations between pairs of mutations that fall within the
+same conserved domain, pointing to compensatory mutations or antagonistic
+epistasis as the prevailing mode of interaction within but not outside of
+conserved genic elements. The heterogeneous landscape of both strength of
+selection and selective interactions within protein-coding genes calls for more
+refined inferences of the joint distribution of fitness and interactive
+effects, and the methods presented here should prove useful in that pursuit.
 
@@
 selection reduces the fixation probability of advantageous mutations and
-increases that of deleterious mutations, compared to expectations under
+increases that of deleterious mutations compared to expectations under
 single-locus models [@Birky1988-jm]. Allele frequency dynamics of linked
@@
 [@Kacser1981-nc], and dominance plays an important role in shaping expected
-equilibrium allele frequencies and mutation load of strongly damaging disease
-mutations [@Clark1998-kq]. On the other hand, epistasis differentially impacts
-the deleterious load for asexually and sexually reproducing organisms
+equilibrium allele frequencies and the mutation load of strongly damaging
+disease mutations [@Clark1998-kq]. On the other hand, epistasis differentially
+impacts the deleterious load for asexually and sexually reproducing organisms
 [@Kimura1966-gc; @Kondrashov1995-va], has been invoked as an explanation for
@@
 populations, epistasis is known to cause signed LD to deviate dramatically from
-zero [@Charlesworth1990-kw; @Kondrashov1995-va]. However, despite dominance
-being appreciated as important factors shaping linked variation
-[@Turelli2000-kz; @Zhao2016-bb] and the evolutionary importance of epistatic
-interactions, we currently lack models for predicting correlation patterns
-between linked mutations under general selection models.
+zero [@Charlesworth1990-kw; @Kondrashov1995-va]. However, despite appreciation
+of the effect of dominance on linked variation [@Turelli2000-kz; @Zhao2016-bb]
+and the evolutionary importance of epistatic interactions, we currently lack
+models for predicting correlation patterns between linked mutations under
+general selection models.
 
@@
 
-In most natural populations, directed mutational studies are not possible, and
+Directed mutational studies are not possible in most natural populations, and
 we must turn to population genetic approaches to infer selective interactions
@@
 other diversity measures are very noisy. Second, comparisons are based on
-theory that are confined to limiting and simplistic assumptions, including
-steady-state demography, simple selection and interaction models, or unlinked
-loci. To generate predictions under more complex models, we rely on expensive
-forward simulations. Such simulations can help build intuition and test
-inference methods, but do not efficiently provide expectations for quantities
-of interest across the range of relevant parameters. Analytical and numerical
-methods for expected haplotype frequencies and LD under general selective
-interaction models are thus crucial for interpreting patterns of variation
-observed in data.
+theory with limiting assumptions, such as steady-state demography, simple
+selection and interaction models, or unlinked loci. To generate predictions
+under more complex models, we rely on expensive forward simulations. Such
+simulations can help build intuition and test inference methods, but do not
+efficiently provide expectations for quantities of interest across the range of
+relevant parameters. Analytical and numerical methods for expected haplotype
+frequencies and LD under general selective interaction models are thus crucial
+for interpreting patterns of variation observed in data.
 
@@
 effects. In what follows I focus on parameters where the strengths of selection
-and dominance at each locus are equal, but note that the methods and
-implementation allow for arbitrary and unequal selection and dominance at the
-two loci.
+and dominance at each locus are equal, but note that the methods presented here
+allow for arbitrary and unequal selection and dominance at the two loci.
 
@@
 In a single randomly mating population, neutral mutations are expected to have
-average signed LD of zero, but across all populations analyzed here, we observe
-positive LD between synonymous mutations. While selection on some subset of
+average signed LD of zero, but across all populations analyzed here, LD between
+synonymous mutations is positive. While selection on some subset of
 synonymous variants is possible, it should be much weaker on average than
@@
 variants should lead to negative LD. @Sandler2020-vk proposed that spatial
-population structure is responsible for the increase of LD for both synonymous
-and missense variants, arguing that every natural population lives in some
-geography that will necessarily induce some structure, and they used forward
-simulations to demonstrate this effect. However, this increase in LD, at least
-in simulations of spatial structure and migration, typically extends over larger
-distances separating base-pairs than we observe in the data. Instead, positive
-LD between synonymous mutations decays to zero within just a few hundred
-base-pairs (Figures \@ref(fig:lddecay) and \@ref(fig:lddecaydomain)).
+population structure may be responsible for the increase of LD for both synonymous
+and missense variants. They argued that every natural population lives in some
+geography that will necessarily induce some structure, and that migration, admixture,
+or spatial structure could increase signed LD above zero.
+
+However, under standard population genetic assumptions of constant mutation and
+recombination rates and independent mutation events, migration and admixture
+does not result in $\E[D]$ to be nonzero. For a given set of mutations with
+differences in allele frequencies between source populations, $D$ will be
+nonzero in an admixed population between those mutations even when $D$ is zero
+in the sources [@Cavalli-Sforza1971-jv]. However, genome averages of allele
+frequency differences in the source populations are zero, as they can be either
+positive or negative, and resulting $\E[D]$ is still zero genome-wide.
+Furthermore, @Sandler2020-vk used forward simulations with an explicit spatial
+model and a multiplicative fitness function to explore this effect, and the
+positive LD they observed under these scenarios extends over larger distances
+than observed in the data. Instead, positive LD between synonymous mutations
+decays to zero within just a few hundred base (Figures \@ref(fig:lddecay) and
+\@ref(fig:lddecaydomain)).
 

### 2021-03-08 SupportingInformation.Rmd
edits

@@
 bibliography: "references.bib"
-title: Supporting Information for 
+title: Supporting Information for "Can we distinguish model of selective interactions using linkage disequilibrium?"
 author:
-  - Aaron P. Ragsdale^[aaronpeaceragsdale@gmail.com]
+  - Aaron P. Ragsdale
 date:

### 2021-03-13 Manuscript.Rmd
asses to assess
edits and readme

@@
 Selected mutations interfere and interact with evolutionary processes at nearby
-loci, distorting patterns of allele frequencies and correlations between pairs
+loci, distorting allele frequency trajectories and correlations between pairs
 of mutations. A number of recent studies have used patterns of linkage
 disequilibrium between selected variants to test for selective interference and
-epistatic interactions, with some disagreement over interpretation of
-observations. Interpreting this data is hindered by the relative lack of
-analytic or even numerical expectations for patterns of variation between pairs
-of loci under the combined effects of selection, dominance, epistasis, and
-demography. Here, I develop a numerical approach to compute the expected
-two-locus sampling distribution under diploid selection, with arbitrary
-epistasis and dominance, and variable recombination and population sizes. I use
-this to explore how different models of epistasis and dominance affect expected
-signed LD, including for non-steady-state demography relevant to human
-populations. Finally, I use whole-genome sequencing data from humans to asses
-how well we can differentiate modes of selective interactions in practice. I
-find that positive LD between missense mutations within genes is driven by
-strong positive correlations between pairs of mutations that fall within the
-same conserved domain, pointing to compensatory mutations or antagonistic
-epistasis as the prevailing mode of interaction within but not outside of
-conserved genic elements. The heterogeneous landscape of both strength of
-selection and selective interactions within protein-coding genes calls for more
-refined inferences of the joint distribution of fitness and interactive
-effects, and the methods presented here should prove useful in that pursuit.
+epistatic interactions, with some disagreement over the interpretation of
+observations from data. Interpreting this data is hindered by the relative lack
+of analytic or even numerical expectations for patterns of variation between
+pairs of loci under the combined effects of selection, dominance, epistasis,
+and demography. Here, I develop a numerical approach to compute the expected
+two-locus sampling distribution under diploid selection with arbitrary
+epistasis and dominance, recombination, and variable population size. I use
+this to explore how epistasis and dominance affect expected signed LD,
+including for non-steady-state demography relevant to human populations.
+Finally, I use whole-genome sequencing data from humans to assess how well we
+can differentiate modes of selective interactions in practice. I find that
+positive LD between missense mutations within genes is driven by strong
+positive correlations between pairs of mutations that fall within the same
+conserved domain, pointing to compensatory mutations or antagonistic epistasis
+as the prevailing mode of interaction within but not outside of conserved genic
+elements. The heterogeneous landscape of both mutational fitness effects and
+selective interactions within protein-coding genes calls for more refined
+inferences of the joint distribution of fitness and interactive effects, and
+the methods presented here should prove useful in that pursuit.
 
@@
 increases that of deleterious mutations compared to expectations under
-single-locus models [@Birky1988-jm]. Allele frequency dynamics of linked
-selected variants are also predicted to deviate from expectations without
-interference. Under a multiplicative fitness model, where the fitness reduction
-of a haplotype carrying multiple deleterious variants is equal to the product
-of the fitness reduction of each mutation independently, we should expect to
-see those mutation segregate on different haplotypes more often than together,
-leading to negative, or repulsion, linkage disequilibrium (LD), although the
-extent of LD depends non-trivially on the strength of selection and probability
-of recombination separating loci [@Hill1966-gv; @McVean2000-ox].
+single-locus models [@Birky1988-jm]. Allele frequency dynamics and correlations
+of linked selected variants are also predicted to deviate from expectations
+without interference. Under a multiplicative fitness model, where the fitness
+reduction of a haplotype carrying multiple deleterious variants is equal to the
+product of the fitness reduction of each mutation independently, we should
+expect to see those mutation segregate on different haplotypes more often than
+together, leading to negative, or repulsion, linkage disequilibrium (LD),
+although the extent of LD depends non-trivially on the strength of selection
+and probability of recombination separating loci [@Hill1966-gv;
+@McVean2000-ox].
 
@@
 improve dramatically, promising greater resolution of the fitness landscape in
-such model systems which can be compared to evolutionary models
+such model systems which can be compared to evolutionary theory
 [@Otwinowski2018-zk].
@@
 A number of factors impede our interpretation of patterns of signed LD between
-coding variants. First, for strongly deleterious or loss-of-function variants,
-their rarity and low frequency means that statistical measurement of LD and
-other diversity measures are very noisy. Second, comparisons are based on
+coding variants. First, for strongly deleterious or loss-of-function mutations,
+their rarity and low allele frequency mean that statistical measurement of LD
+and other diversity measures are very noisy. Second, comparisons are based on
 theory with limiting assumptions, such as steady-state demography, simple
@@
 compute the expected sampling distribution $\Psi_n$ of two-locus haplotypes
-under a general model of selective interactions. This sampling distribution, or
-two-locus haplotype frequency spectrum, stores the expected density or observed
-counts of pairs of loci with each possible haplotype frequency configuration in
-a sample of size $n$. In many cases it is simpler to visualize summaries (such
-as $\E[D]$) or slices of the distribution (Figure \@ref(fig:fig1)) instead of
-the full three-dimensional distribution. Here, I focus on low-order LD
-statistics, such as $\E[D]$ and $\E[D^2]$, and their decay with recombination
-distance as these are statistics that are commonly used to test for
-interactions between loci. Instead of unnormalized $\E[D^2]$ and $\E[D]$, I
-consider expectations for $\sigma_d^2 = \E[D^2]/\E[p(1-p)q(1-q)]$ and
-$\sigma_d^1 = \E[D]/\E[p(1-p)q(1-q)]$. These statistics have two benefits:
+under a general model of selective interactions. This sampling distribution
+stores the expected density or observed counts of pairs of biallelic loci with
+each possible haplotype frequency configuration in a sample of size $n$. In
+many cases it is simpler to visualize summaries (such as $\E[D]$) or slices of
+the distribution (Figure \@ref(fig:fig1)) instead of the full three-dimensional
+distribution. Here, I focus on low-order LD statistics, such as $\E[D]$ and
+$\E[D^2]$, and their decay with recombination distance as these are statistics
+that are commonly used to test for interactions between loci. Instead of
+unnormalized $\E[D^2]$ and $\E[D]$, I consider expectations for $\sigma_d^2 =
+\E[D^2]/\E[p(1-p)q(1-q)]$ and $\sigma_d^1 = \E[D]/\E[p(1-p)q(1-q)]$, where $p$
+and $q$ are the derived allele frequencies at the left ($A/a$) and right
+($B/b$) loci, respectively. These normalized statistics have two benefits:
 first, the mutation rate cancels so that expectations are robust to assumptions
@@
 
-For additive mutations (i.e. $h=1/2$) and no epistasis, we recover the well
-known Hill-Robertson interference result of negative LD between selected
-mutations, which is strongest for pairs for mutations that have selection
-coefficients $\gamma=O(1)$, or $s \approx 1/2N_e$ (Figure \@ref(fig:fig1)D).
-For strongly deleterious mutations, LD is close to zero even with tight
-linkage, as they almost always segregate at low enough frequencies that they
-are unlikely to interfere with each other [@McVean2000-ox].
+For mutations under additive selection (i.e. $h=1/2$) and no epistasis, we
+recover the well known Hill-Robertson interference result of negative LD
+between selected mutations, which is strongest for pairs for mutations that
+have selection coefficients $\gamma=O(1)$, or $s \approx 1/2N_e$ (Figure
+\@ref(fig:fig1)D). For strongly deleterious mutations, LD is close to zero even
+with tight linkage, as they almost always segregate at low enough frequencies
+that they are unlikely to interfere with each other [@McVean2000-ox].
 
@@
 (Figure \@ref(fig:fig2)). Synergistic epistasis results in negative LD while
-antagonistic epistasis results positive LD, and large positive LD can occur
+antagonistic epistasis results in positive LD, and large positive LD can occur
 even when the epistatic effect is relatively weak. Epistasis-induced LD can
 extend over long distances, especially for strongly deleterious mutations. For
-example, in Figure \@ref(fig:fig1)D with $\gamma=-10$, which would be
-considered moderate, large mean LD extends well beyond 100 kilobases, with more
-strongly deleterious interacting mutations expected to show large signed LD
-over much larger recombination distances.
+example, in Figure \@ref(fig:fig2)D even moderately deleterious mutations with
+population-size-scaled selection coefficients of $\gamma=-10$ show large mean
+LD that extends well beyond 100 kilobases. More strongly deleterious
+interacting mutations are expected to show large signed LD over much larger
+recombination distances.
 
@@
 
-The effects of epistasis between variants on LD has been widely documented.
-However, the effect of non-additive selection within a locus on interference
-has received less attention. @Garcia2020-zn used large-scale forward simulation
-to explore the effect of dominance on patterns of LD, showing that LD depends
-on both the magnitude of the fitness and dominance effects in a nonlinear way,
-but note that expensive large-scale simulations are needed to explore these
-effects as no closed form approach to computing those expectations exist. The
-method presented here provides such a closed form approach, so that
-expectations of signed LD and other two-locus statistics can be computed more
-efficiently and accurately.
+The effects of epistasis between variants on LD have been widely studied.
+However, the combined effect of non-additive selection within a locus and
+interference has received less attention. @Garcia2020-zn used large-scale
+forward simulations to explore the effect of dominance on patterns of LD,
+showing that LD depends on both the magnitude of the selection and dominance
+coefficients in a nonlinear way, but note that expensive large-scale
+simulations are needed to explore these effects as no closed form approach to
+computing those expectations exist. The method presented here provides such a
+closed form approach, so that expectations of signed LD and other two-locus
+statistics can be computed more efficiently and accurately.
 
@@
 \@ref(fig:fig3)D), although it decays rapidly with increasing distance between
-loci and is roughly zero for $\rho > 1$. For weakly selected mutations with
+loci and is roughly zero for $\rho \gg 1$. For weakly selected mutations with
 $\gamma=-1$, we see that there is no monotonic effect of the level of dominance
@@
 
-The case with $s_{AB/ab}\not=s_{Ab/aB}$ could arise in a scenario where a
+The case with $s_{AB/ab}\not=s_{Ab/aB}$ can arise in a scenario where a
 mutation at either locus within a haplotype impacts some functional region or
-element, but a diploid carrying at least one copy that is free of mutations has
-minimal fitness loss. This is analogous to the single-locus dominance case,
-where $s_{Aa} < \frac{1}{2}s_{AA}$. In this "gene-based dominance" scenario
-[e.g., @Sanjak2017-zw], an $AB/ab$ genotype has higher fitness than an $Ab/aB$
-type (a simple implementation of this model is given in the final column in
-Table \@ref(tab:selmodels)). Such a gene-based fitness model gives similar
-expected positive signed LD to the model of antagonistic epistasis (compare
-Figure \@ref(fig:fig4)E-F to G-H), although the interpretation of those two
-models can vary. It does show that with a highly parameterized space of
-possible general diploid selection models, multiple models with different
-biological interpretations can give similar patterns of expected $D$.
+element, but a diploid individual carrying at least one copy that is free of
+mutations has minimal fitness loss. This is analogous to the single-locus
+dominance case, where $s_{Aa} < \frac{1}{2}s_{AA}$. In this "gene-based
+dominance" scenario [e.g., @Sanjak2017-zw], an $AB/ab$ genotype has higher
+fitness than an $Ab/aB$ type (a simple implementation of this model is given in
+the final column in Table \@ref(tab:selmodels)). Such a gene-based fitness
+model gives similar expected positive signed LD to the model of antagonistic
+epistasis (compare Figure \@ref(fig:fig4)E-F to G-H), although the
+interpretation of those two models can differ. It shows that with a highly
+parameterized space of possible general diploid selection models, multiple
+models with different biological interpretations can give similar patterns of
+expected $D$.
 

### 2021-03-13 Manuscript.Rmd
add eric and matthias, some edits

@@
 
-In what follows, I turn to empirical measurements of LD within human
-populations, focusing on mutations within protein-coding regions. I first look
-at average values of LD across genes for synonymous, missense, and
-loss-of-function mutations, and then turn to patterns of LD partitioned by
-functional elements and distances separating mutations within genes.
+In what follows I examine LD from human population sequencing data, focusing on
+mutations within protein-coding regions. I first look at average values of LD
+across genes for pairs of synonymous, missense, and loss-of-function mutations
+and then turn to patterns of LD partitioned by functional elements and
+distances separating mutations within genes.
 
@@
 considering pairs of mutations within the same gene. Loss-of-function mutations
-instead have negative LD, pointing to differing modes of selective interactions
-for loss-of-function and missense mutations (Figure \@ref(fig:fig5)). Within
-each population, large measurement noise gives 95\% confidence intervals that
-overlap with zero for each mutation class, although the pattern is remarkably
-consistent across African, European, and East Asian populations in the Thousand
-Genomes dataset. Comparing mean LD within genes between African and Eurasian
-populations, LD in Eurasian populations is somewhat larger on average, that
-is, more positive for missense mutations and more negative for loss-of-function
-mutations. This is in agreement with differences in expectations between
-populations that have or have not gone through a bottleneck in their recent
-past (Figure \@ref(fig:fig4)).
-
-For pairs of missense mutations in different protein-coding genes, signed LD
-decays to zero fairly rapidly and is lower than signed LD within genes even for
-mutations that are more tightly linked (distance less that 50 kb). The
-similarity in signed LD observed between missense and synonymous mutations
+instead have more negative LD, suggesting differing modes of selective
+interactions for loss-of-function and missense mutations (Figure
+\@ref(fig:fig5)). Within each population, large measurement noise gives 95\%
+confidence intervals that overlap with zero across mutation class, although the
+pattern is remarkably consistent across African, European, and East Asian
+populations in the Thousand Genomes dataset. Comparing mean LD within genes in
+African and Eurasian populations, LD in Eurasian populations is somewhat larger
+on average, that is, more positive for missense mutations and more negative for
+loss-of-function mutations. This is in agreement with differences in
+expectations between populations that have or have not gone through a
+bottleneck in their recent past (Figure \@ref(fig:fig4)). For pairs of missense
+mutations in different protein-coding genes, signed LD decays to zero fairly
+rapidly and is lower than signed LD within genes even for mutations that are
+linked (distance less that 50 kb).
+
+The similarity in signed LD observed between missense and synonymous mutations
 would suggest that interference between missense mutations is minimal. However,
-below I show that interactive effects differ dramatically between
-mutations that lie within annotated functional domains and mutations outside of
+below I show that interactive effects differ dramatically between mutations
+that lie within annotated functional domains and mutations outside of
 functional domains, with these opposing effects effectively canceling out when
@@
 density of loss-of-function mutations within annotated domains in humans
-precludes any analysis of LD between those mutations at the level of
-sub-elements within genes. Protein-coding domains are conserved elements of
-genes, often associated with some known functional or structural feature of a
-protein. As such, we would expect that purifying selection is stronger on
-average within domains than within the same gene but outside of a conserved
-domains. Indeed, the site-frequency spectrum (SFS) is skewed to lower
-frequencies for both missense and loss-of-function mutations within domains
-when compared to the same classes of mutations outside of domains, with much
-more negative values of Tajima's D within domains (Table \@ref(tab:tajimasD)).
-On the other hand, no difference is observed for synonymous mutations within or
-outside domains, suggesting roughly equivalent effects of selection (either
-direct or linked) on synonymous variation.
+precludes analysis of LD between those mutations at the level of sub-elements
+within genes. Protein-coding domains are conserved elements of genes, often
+associated with some known functional or structural feature of a protein. As
+such, we would expect that purifying selection is stronger on average within
+domains than within the same gene but outside of a conserved domains. Indeed,
+the site-frequency spectrum (SFS) is skewed to lower frequencies for both
+missense and loss-of-function mutations within domains when compared to the
+same classes of mutations outside of domains, with much more negative values of
+Tajima's D within domains (Table \@ref(tab:tajimasD)). On the other hand, no
+difference is observed for synonymous mutations whether within or outside
+domains, suggesting roughly equivalent effects of selection (either direct or
+linked) on synonymous variation.
 
@@
 opposite to those expectations, suggesting a different prevailing interactive
-effect within domains than those that are commonly tested or modeled.
+effect between nonsynonymous mutations within domains than those that are
+commonly tested or modeled.
 
@@
 within domains is much smaller than between domains, so this observation may be
-primarily driven by the higher recombination rate between pairs of mutations
+primarily driven by the higher recombination distances between pairs of mutations
 across distinct domains.
@@
 I also measure LD between pairs of mutations that both fall outside of
-annotated domains. For such mutations, I consider averages across all pairs
-outside of domains, as well as subsets of pairs matched to the same distances
-for pairs of mutation within and between domains (Figures \@ref(fig:fig6)B-D and
-\@ref(fig:domainspops)). Averaging over all pairs of mutations within the same
-gene but outside domains, both synonymous and missense mutations have LD
-slightly above zero, with synonymous mutations observed to have slightly higher
-signed LD than missense mutations. This pattern is primarily driven by tightly
-linked variants. For pairs of mutations at distances matched to those within
-the domains, synonymous mutations have much larger positive LD than
-missense mutations, a signal that vanishes for pairs matched to distances
-between domains. For larger distances, both synonymous and missense mutations
-have LD approximately zero.
+annotated domains. For such mutations, I consider averages across all pairs of
+mutations outside of domains, as well as subsets of pairs matched to the same
+distances as pairs of mutation within and between domains as surveyed in the
+previous section (Figures \@ref(fig:fig6)B-D and \@ref(fig:domainspops)).
+Averaging over all pairs of mutations within the same gene but outside domains,
+both synonymous and missense mutations have LD slightly above zero, with
+synonymous mutations observed to have slightly higher signed LD than missense
+mutations. This pattern is primarily driven by tightly linked variants. For
+pairs of mutations at distances matched to those within the domains, synonymous
+mutations have much larger positive LD than missense mutations, a signal that
+vanishes for pairs matched to distances between domains. For larger distances,
+both synonymous and missense mutations have LD approximately zero.
 
@@
 \@ref(fig:lddecay)). Both pairs of synonymous and missense mutations with
-distances greater than a few hundred base pairs have LD that fluctuates around
-zero. But for shorter distances, synonymous variants have large positive LD,
-while missense mutations have much lower LD except for pairs within the
-shortest distance bin ($<10$ bp). This reduction in LD is consistent with
-Hill-Robertson interference between linked missense mutations instead of
-epistatic effects, as interference between either additive or recessive
-mutations without epistasis results in signed LD that decays rapidly by
-recombination distance $\rho \approx 1$.
+distances greater than a few hundred base pairs have average LD that fluctuates
+around zero. But for shorter distances, synonymous variants have large positive
+LD, while missense mutations have much lower LD except for pairs within the
+shortest distance bin ($<10$ bp). This reduction in LD for missense mutations
+compared to synonymous mutations is consistent with Hill-Robertson interference
+between linked selected mutations instead of epistatic effects, as interference
+between either additive or recessive mutations but without epistasis results in
+signed LD that decays rapidly by recombination distances of $\rho \approx 1$.
 
@@
 difficulty is compounded by the large measurement noise for $\E[D]$, especially
-for loss-of-function variants for which only a few hundred pairs are found
-within genes in each population. If loss-of-function variants are in fact
-recessive, even partially, this can impact inferences of
-the DFE (Supporting Information).
+for loss-of-function variants for which only a few hundred within-gene pair in
+human populations.
 
@@
 Hill-Robertson interference are expected to result in positive LD, so that some
-other interactive effect must be driving this positive LD signal.
+other interactive effect must be driving this positive LD signal within
+conserved domains.
 
 There are a number of possible interaction scenarios that can result in
-positive signed LD between linked loci. One explanation is a
-prevalence of pairs of compensatory mutations that are tolerated to
-co-segregate at high frequencies within conserved domains [@Yeang2007-gj;
-@Ivankov2014-tn]. @Callahan2011-ac and @Taverner2020-lk have proposed such a
-mechanism to explain observed clusters of mutations in lineages of *Drosophila*
-and other species. Another possibility is a model of antagonistic, or
-diminishing returns epistasis, in which a single amino acid-changing mutation
-within a domain damages the functionality of that subunit, but additional
-mutations within that same domain reduces fitness by a factor less than the
-first mutation. A third possibility, related to the antagonistic epistasis
-case, is that selection acts on the functional domain as a unit instead of at
-loci within the domain individually (such as gene-based dominance). In this
-scenario, double heterozygotes have different fitnesses depending on whether
-the mutations are found on same haplotype or on different haplotypes.
+positive signed LD between linked loci. One explanation is a prevalence of
+pairs of compensatory mutations that are tolerated to co-segregate at high
+frequencies within conserved domains [@Yeang2007-gj; @Ivankov2014-tn].
+@Callahan2011-ac and @Taverner2020-lk have proposed such a mechanism to explain
+observed clusters of mutations in lineages of *Drosophila* and other species.
+Another possibility is a model of antagonistic, or diminishing returns
+epistasis, in which a single amino acid-changing mutation within a domain
+damages the functionality of that subunit, but additional mutations within that
+same domain reduce fitness by a factor less than the first mutation. A third
+possibility, related to the antagonistic epistasis case, is that selection acts
+on the functional domain as a unit instead of at loci within the domain
+individually (such as gene-based dominance). In this scenario, double
+heterozygotes have different fitnesses depending on whether the mutations are
+found on same haplotype or on different haplotypes.
 
@@
 loss-of-function mutations within conserved domains are subject to stronger
-selection, skewing the SFS to lower frequencies. However, typical approaches
-for inferring the distribution of fitness effects from population genomic data
+selection, skewing the SFS to lower frequencies. Typical approaches for
+inferring the distribution of fitness effects from population genomic data
 average over these differences by considering all nonsynonymous mutations
-together [@Boyko2008-zk; @Kim2017-xo]. It would be straightforward to adapt
-those methods to infer more detailed and accurate representation of the
-heterogeneous effects of new mutations within genes by partitioning by missense
-and loss-of-function classes as well as by annotated domains. Similarly, the
+together [@Boyko2008-zk; @Kim2017-xo], but if loss-of-function variants are in
+fact recessive, even partially, this can impact inferences of the DFE
+(Supporting Information). It would be straightforward to adapt DFE-inference
+methods to infer more detailed and accurate representation of the heterogeneous
+effects of new mutations within genes by partitioning by missense and
+loss-of-function classes as well as by annotated domains. Similarly, the
 different modes of interaction between mutations across annotated domains mean
@@
 average signed LD of zero, but across all populations analyzed here, LD between
-synonymous mutations is positive. While selection on some subset of
-synonymous variants is possible, it should be much weaker on average than
-between missense mutations, and any interference between selected synonymous
-variants should lead to negative LD. @Sandler2020-vk proposed that spatial
-population structure may be responsible for the increase of LD for both synonymous
-and missense variants. They argued that every natural population lives in some
-geography that will necessarily induce some structure, and that migration, admixture,
-or spatial structure could increase signed LD above zero.
-
-However, under standard population genetic assumptions of constant mutation and
-recombination rates and independent mutation events, migration and admixture
-does not result in $\E[D]$ to be nonzero. For a given set of mutations with
-differences in allele frequencies between source populations, $D$ will be
-nonzero in an admixed population between those mutations even when $D$ is zero
-in the sources [@Cavalli-Sforza1971-jv]. However, genome averages of allele
-frequency differences in the source populations are zero, as they can be either
-positive or negative, and resulting $\E[D]$ is still zero genome-wide.
-Furthermore, @Sandler2020-vk used forward simulations with an explicit spatial
-model and a multiplicative fitness function to explore this effect, and the
-positive LD they observed under these scenarios extends over larger distances
-than observed in the data. Instead, positive LD between synonymous mutations
-decays to zero within just a few hundred base (Figures \@ref(fig:lddecay) and
-\@ref(fig:lddecaydomain)).
+synonymous mutations is positive. While selection on some subset of synonymous
+variants is possible, it should be much weaker on average than between missense
+mutations, and any interference between selected synonymous variants should
+lead to negative LD. @Sandler2020-vk proposed that spatial population structure
+may be responsible for the increase of LD for both synonymous and missense
+variants. They argued that every natural population lives in some geography
+that will necessarily induce some structure, and that migration, admixture, or
+spatial structure could increase average signed LD above zero.
+
+Under standard population genetic assumptions of constant mutation and
+recombination rates and independent mutation events, however, migration and
+admixture do not result in nonzero $\E[D]$. Rather, for a set of mutations with
+*given* differences in allele frequencies between source populations, $D$ can
+be nonzero in an admixed population between those particular mutations even
+when $D$ is zero in the source populations [@Cavalli-Sforza1971-jv]. But this
+result relies on conditioning on directional differences in allele frequencies
+between populations. Expected mean allele frequency difference in the source
+populations is zero ($\E[(p_1-p_2)(q_1-q_2)] = 0$), as differences are equally
+likely to be positive or negative, and resulting $\E[D]$ is still zero when
+taking genome-wide averages. Furthermore, @Sandler2020-vk used forward
+simulations with an explicit spatial model and a multiplicative fitness
+function to explore the joint effect of structure and interference, and the
+positive LD they observed under these scenarios extends over much larger
+distances than observed in the data. Instead, positive LD between synonymous
+mutations decays to zero within just a few hundred base (Figures
+\@ref(fig:lddecay) and \@ref(fig:lddecaydomain)).
 
@@
 Indeed, a simple exponential model in which the fraction of pairs of mutations
-arising arising from a multinucleotide event decays with distance fits the
-observed decay of $\sigma_d^1$ between synonymous mutations (Supporting
+arising from a multinucleotide mutation event decays with distance fits the
+observed decay of $\sigma_d^1$ between synonymous variants (Supporting
 Information). We may therefore treat positive LD as the baseline expectation
@@
 between linked loss-of-function variants to that positive baseline expectation
-instead of zero, so that they are more recessive or have stronger synergistic
-epistasis than if the neutral expectation were zero.
+instead of zero, which would imply that they are more recessive or have
+stronger synergistic epistasis than from inferences assuming a neutral
+expectation of zero.
 
@@
 Instead, closed, numerical approaches to compute expectations for two-locus
-statistics under a wide range of selection models, as presented here, allow us
-to explore this highly parameterized space of models far more efficiently, and
-it opens the possibility for performing likelihood-based inference using signed
-LD or other two-locus summaries of the data. For example, inferring the joint
-distribution of dominance and selection is underpowered using the SFS. Because
+statistics under a wide range of selection models, as presented here or in
+@Ragsdale2019-nt and @Friedlander2021-bs, allow us to explore this highly
+parameterized space of models far more efficiently, and it opens the
+possibility for performing likelihood-based inference using signed LD or other
+two-locus summaries of the data. For example, inferring the joint distribution
+of dominance and selection is underpowered using the SFS alone, but because
 signed LD is sensitive to the levels of dominance (Figure \@ref(fig:fig3)),
-inferring the DFE with dominance may become feasible using the joint
-distribution of allele frequencies and LD.
+inferring the DFE with dominance may be feasible using the joint distribution
+of allele frequencies and LD.
 
@@
 additivity and interchangeability between sites within a gene, potentially
-biasing any inference using this DFE to learn about other patterns of
-selection.
+biasing any inference using previously inferred DFEs to learn about patterns of
+interactions. Again, this underlies the need to jointly infer strengths and
+interactions of selected variants.
 
@@
 arbitrary moments of the distribution of $D$ for any number of populations
-connected by migration and admixture. This paper also showed that such a
+connected by migration and admixture. They also showed that such a
 moments-construction can be used to solve for the two-locus sampling
@@
 approach to model arbitrary diploid selection, which encompasses dominance,
-epistasis, and other forms of selective interactions between two loci.
+epistasis, and other forms of selective interactions between two loci. In a
+concurrent study to this paper, @Friedlander2021-bs developed numerical
+solutions to the same moments system, which they used to describe selected
+haplotype trajectories and the distortion of neutral diversity at loci variably
+linked to beneficial alleles that sweep to high frequencies under
+non-equilibrium demography.
 
@@
 from $\Psi_n$ [@Jouganous2017-pq], so that $\hat{\Psi}_{n+l}(i, j, k) =
-\mathcal{J}_{n, l}\Psi_n$, although other closure approximations are possible.
-This emits a closed approximate system,
+\mathcal{J}_{n, l}\Psi_n$, although other accurate closure approximations are
+possible [@Friedlander2021-bs]. This emits a closed approximate system,
 \begin{equation}

### 2021-03-14 Manuscript.Rmd
edits to methods and figures

@@
 Under a two-locus biallelic model, where the left locus allows alleles $A$ and
-$a$ and the right locus has alleles $B$ and $b$, the standard covariance
-measure of LD is defined as $D = f_{AB} - f_{A}f_{B}$ where $f_{AB}$ is the
+$a$ and the right locus allows alleles $B$ and $b$, the standard covariance
+measure of LD is defined as $D = f_{AB} - f_{A}f_{B}$, where $f_{AB}$ is the
 haplotype frequency of double-derived types carrying both $A$ and $B$, and
@@
 proportional to the inverse of the effective population size and the distance
-separating loci: \[\E[D]_{t+1} = \left(1 - \frac{1}{2N_e(t)} - r
+separating loci [@Hill1968-vu]: \[\E[D]_{t+1} = \left(1 - \frac{1}{2N_e(t)} - r
 \right)\E[D]_t.\] While $\E[D] = 0$, the variance of $D$ is non-zero, and
@@
 To include selection, however, there have been relatively few advances beyond
-the Monte Carlo simulation approach taken by @Hill1966-gv, albeit we now have
-access to somewhat more powerful computational resources and more sophisticated
-software for performing flexible forward simulation (e.g., @Thornton2014-pn,
-@Haller2019-vm). Analytic results for two-locus distributions under selection
-are notoriously difficult, with a few notable flashes of progress. For example,
-@McVean2007-xd considered the effect of a recent sweep on patterns of LD in
-loci nearby or spanning the locus under selection, and in an impressive recent
-paper, @Good2020-wj presented analytic solutions for patterns of LD between
-rare mutations under additive selection with epistasis. Nonetheless, such
-approaches are typically confined to steady-state demography and constrained
-selection models.
+the Monte Carlo simulation approach taken by @Hill1966-gv, albeit now with more
+powerful computational resources and sophisticated software for performing
+flexible forward simulation (e.g., @Thornton2014-pn, @Haller2019-vm). Analytic
+results for two-locus distributions under selection are notoriously difficult,
+with a few notable flashes of progress. For example, @McVean2007-xd considered
+the effect of a recent sweep on patterns of LD between neutral loci near the
+locus under selection, and in a recent paper, @Good2020-wj presented analytic
+solutions for patterns of LD between rare mutations under additive selection
+with epistasis. Nonetheless, such approaches are typically confined to
+steady-state demography and constrained selection models.
 
@@
 numerical approaches to the two-locus coalescent [@Kamm2016-ag] or diffusion
-approximation [@Ragsdale2017-gg]. Practically, selection can only be included
-using the forward-in-time diffusion equation [@Kimura1955-qe; @Hill1966-gv],
-where a standard approach is to first solve for the continuous distribution
-$\psi$ of the density of two-locus haplotype configurations in the full
-population, and then integrate $\psi$ against the multinomial sampling function
-to obtain $\Psi_n$.
+approximation [@Ragsdale2017-gg]. Selection is most easily included using the
+forward-in-time diffusion equation [@Kimura1955-qe; @Hill1966-gv], where a
+standard approach is to first solve for the continuous distribution $\psi$ of
+the density of two-locus haplotype configurations in the full population, and
+then integrate $\psi$ against the multinomial sampling function to obtain
+$\Psi_n$.
 
@@
 The system of linear ordinary differential equations for the entries of
-$\Psi_n$takes the form
+$\Psi_n$ takes the form
 \begin{equation}
@@
 population. We assume $n \ll N_e$, and $r$ and $s$ are $O(1/Ne)$ so that
-multiple coalescence, recombination, or selective events are rare in any given
-generation. For example, for drift events, we account for the probability that
-two lineages within the $n$ tracked lineages at time $t+1$ coalesced in the
-previous generation and, given a coalescence event, the probability that
-haplotype counts changed due to this coalescence event [see Supporting
-Information; @Jouganous2017-pq; @Ragsdale2019-nt]. In typical diffusion
-approximation fashion, we multiple through by $2N_{ref}$ so that time
-is measured in $2N_{ref}$ generations, and we consider scaled parameters
-$\rho = 4Nr$, $\theta = 4Nu$, and $\gamma=2Ns$.
+multiple coalescence, recombination, or selective events within the $n$
+lineages are rare in any given generation. For example, for drift events, we
+account for the probability that two lineages within the $n$ tracked lineages
+at time $t+1$ coalesced in the previous generation and, given a coalescence
+event, the probability that haplotype counts changed due to this coalescence
+event [see Supporting Information; @Jouganous2017-pq; @Ragsdale2019-nt]. In
+typical diffusion approximation fashion, we multiple through by $2N_{ref}$ so
+that time is measured in $2N_{ref}$ generations, and we consider scaled
+parameters $\rho = 4Nr$, $\theta = 4Nu$, and $\gamma=2Ns$.
 
@@
 system is closed and can be solved exactly. However, for nonzero recombination
-or selection, the entries of $\Psi_n$ rely on the slightly larger frequency
-spectra with sample sizes $n+1$ (for recombination and additive selection) or
-$n+2$ (for non-additive selection). This is because if a recombination event
-occurs within one of $n$ lineages being tracked by $\Psi_n$, we need to draw an
-additional lineage from the full population to recombine with that chosen
-lineage, thus requiring $\Psi_{n+1}^t$ to find $\Psi_n^{t+1}$. Selection events
-similarly require extra lineages from the full population, which replace a
-chosen lineage that fails to reproduce with probability proportional to its
-selection coefficient, assuming negative selection.
+or selection, the entries of $\Psi_n$ rely on the slightly sampling
+distributions with sample sizes $n+1$ (for recombination and additive
+selection) or $n+2$ (for non-additive selection). This is because if a
+recombination event occurs within one of $n$ lineages being tracked by
+$\Psi_n$, we need to draw an additional lineage from the full population to
+recombine with that chosen lineage, thus requiring $\Psi_{n+1}^t$ to find
+$\Psi_n^{t+1}$. Selection events similarly require extra lineages from the full
+population, which replace a chosen lineage that fails to reproduce with
+probability proportional to its selection coefficient, assuming negative
+selection.
 
@@
 jackknife approximation is used to estimate $\Psi_{n+l}$, for $l=1$ or $2$,
-from $\Psi_n$ [@Jouganous2017-pq], so that $\hat{\Psi}_{n+l}(i, j, k) =
-\mathcal{J}_{n, l}\Psi_n$, although other accurate closure approximations are
-possible [@Friedlander2021-bs]. This emits a closed approximate system,
+from $\Psi_n$ [following the single-locus closure introduced in
+@Jouganous2017-pq], so that $\hat{\Psi}_{n+l}(i, j, k) = \mathcal{J}_{n,
+l}\Psi_n$, although other accurate closure approximations are possible
+[@Friedlander2021-bs]. This emits a closed approximate system,
 \begin{equation}
@@
 The jackknife approximation, which approximates an entry $\Psi_{n+l}(i,j,k)$
-using nearby entries in $\Psi_n$, is more accurate for larger sample sizes,
-creating a tension between efficiency and accuracy: larger sample sizes result
-in more accurate solutions, as error in the jackknife is diminished, but
-computational complexity also grows rapidly in the number of entries of
-$\Psi_n$, which is $O(n^3)$. Accuracy and runtime are shown in Figure
-\@ref(fig:jackknife). In the results presented in this paper, sample sizes
-between $n=30$ and $n=80$ are used.
+using nearby entries in $\Psi_n$, is more accurate for larger sample sizes
+(Figure \@ref(fig:jackknife)), creating a tension between efficiency and
+accuracy: larger sample sizes result in more accurate solutions, as error in
+the jackknife is diminished, but computational complexity also grows rapidly in
+the number of entries of $\Psi_n$, which is $O(n^3)$. Accuracy and runtime are
+shown in Figure \@ref(fig:jackknife). In the results presented in this paper,
+sample sizes between $n=30$ and $n=80$ are used.
 
@@
 $ab$ haplotypes. The standard multiplicative fitness function assumes that
-$s_{AB} \approx s_{A} + s_{B}$, and a general epistasis function can be written
-as $s_{AB} = (s_{A} + s_{B}) (1 + \epsilon)$, so that $\epsilon > 0$ implies
+$s_{AB} \approx s_{A} + s_{B}$, and the effect of epistasis can be written as
+$s_{AB} = (s_{A} + s_{B}) (1 + \epsilon)$, so that $\epsilon > 0$ implies
 synergistic epistasis, while $\epsilon < 0$ implies antagonistic epistasis.
@@
 of haplotype counts given by $\Psi_n^t$. However, assuming $s\leq0$ for each
-haplotype, each of those sampled lineages has probability of being rejected
-equal to the relative fitness of its haplotype state. If a lineage is rejected,
-a replacement is drawn from the full population. Under the assumption that
-$ns_{\cdot\cdot} \ll 1$, the probability that more than one selection event
-occurs in any given generation is negligibly small, so that the case of
+derived haplotype, each of those sampled lineages has probability of being
+rejected equal to the relative fitness of its haplotype state. If a lineage is
+rejected, a replacement is drawn from the full population. Under the assumption
+that $ns_{\cdot\cdot} \ll 1$, the probability that more than one selection
+event occurs in any given generation is negligibly small, so that the case of
 multiple simultaneous rejections can be ignored. Then $Psi_n^{t+1}$ relies only
@@
 $s_{AB/Ab}$, and so on. Assuming that the doubly homozygous ancestral $ab/ab$
-genootype has relative fitness 1, this gives nine possible unique selection
+genotype has relative fitness 1, this gives nine possible unique selection
 coefficients in the most general two-locus selection model. Note that $AB/ab$
@@
 identification, so I only kept sites for which ancestral alleles were estimated
-with high confidence in both the VCF info fields and the Thousand Genomes human
+with high confidence in both the VCF info field and the Thousand Genomes human
 ancestor reconstructed from a phylogeny of six primates.
@@
 # Acknowledgements
+- Discussions: Simon Gravel, Alex Diaz-Papkovich, Matthias Steinrücken, Eric Friedlander
+- Comments on ms:
+
+Asked from comments from
+- Kevin Thornton
+- Mashaal Sohail
 - Simon Gravel
-- Kevin Thornton
-- Alex Diaz-Papkovich
-- Mashaal Sohail
 - Matthias Steinrücken
@@
 from @Ohta1971-yd, given in Equation \@ref(eq:ohta). (D) The decay of
-$\sigma_d^1=\E[D]/ \E[p(1-p)q(1-q)]$ for pairs of variants. The
-negative LD effect of Hill-Robertson interference is strongest for intermediate
-selection, with $\gamma\sim-1$ to $-5$.
+$\sigma_d^1=\E[D]/ \E[p(1-p)q(1-q)]$ for pairs of variants. The negative LD
+effect of Hill-Robertson interference is strongest for intermediate selection,
+with $\gamma\sim-1$ to $-5$, where $\gamma=4N_es$.
 
@@
 Epistasis expected signed LD away from zero, an effect which can extend over
-long recombination distances. Here, $\gamma$ is the selection coefficient for
+large recombination distances. Here, $\gamma$ is the selection coefficient for
 derived alleles $A$ and $B$, assuming selection is equal at the two loci.
@@
 @1000_Genomes_Project_Consortium2015-zq. In all simulations, $\gamma$ is set to
-$-2$ at both loci. (A-B) Model illustrations. Note that the differences in
+$-2$ at both loci. (A-B) Model illustrations. Note the differences in
 scale. (C-D) Additive selection with no epistasis, showing fully linked
@@
 variants in coding regions partitioned by classes of mutations.
-Loss-of-function mutations are defined as stop and start gains and losses,
-frameshifts, and other nonsense mutations. In all populations, we observe
-positive mean signed LD, although the confidence interval of each measurement
-overlaps with zero. Similarly, we observe mean negative LD for pairs of
-loss-of-function variants within genes. There are only $\sim200-300$ pairs of
-loss of function variants that fall within the same gene across the populations
-tested here, causing large uncertainty in each measurement. Gray lines indicate
-2 standard errors of the measurement, via bootstrap. (C-F) While pairs of
+Loss-of-function mutations are defined as stop and start gains and losses and
+other nonsense mutations of high impact. In all populations, we observe
+positive mean signed LD for missense and synonymous variants, although the
+confidence interval of each measurement overlaps with zero. Similarly, we
+observe mean negative LD for pairs of loss-of-function variants within genes.
+There are only $\sim200-300$ pairs of loss-of-function variants that fall
+within the same gene across the populations tested here, causing large
+uncertainty in each measurement. Gray lines indicate 2 standard errors of the
+measurement, computed by bootstrapping of sets of genes. (C-F) While pairs of
 synonymous and missense mutations within genes are consistently positive, pairs
@@
 (ref:fig6cap) **LD within and between annotated domains in protein-coding
-genes.** Considering only variants that fall within annotated domains in
+genes.** (A) Considering only variants that fall within annotated domains in
 protein-coding genes, pairs of mutations that fall within the same gene display
@@
 domains, or for pairs of synonymous mutations that are within the same domain.
-The slightly positive LD that we observe for missense and synonymous mutations
-gene-wide (Figure \@ref(fig:fig4)A) is largely driven by the subset of pairs
-mutations that fall within the same annotated domain. LD for all populations
-analyzed here are shown in Figure \@ref(fig:domainspops).
+(B-D) The slightly positive LD that we observe for missense and synonymous
+mutations gene-wide (Figure \@ref(fig:fig4)A) is largely driven by the subset
+of pairs mutations that fall within the same annotated domain. LD for all
+populations analyzed here are shown in Figure \@ref(fig:domainspops).
 

### 2021-03-14 figures/fig2.py

@@ -155,9 +155,9 @@ ax3.set_ylim(-1.2, 1.2)
 ax4.set_ylim(ax3.get_ylim())
 
 fig.tight_layout()
-#fig.subplots_adjust(hspace=0.5, top=0.98)
-
-#fig.text(0.05, 0.97, "A", fontsize=8, ha="center", va="center")
-#fig.text(0.05, 0.45, "B", fontsize=8, ha="center", va="center")
+fig.text(0.02, 0.98, "A", fontsize=8, ha="center", va="center")
+fig.text(0.52, 0.98, "B", fontsize=8, ha="center", va="center")
+fig.text(0.02, 0.48, "C", fontsize=8, ha="center", va="center")
+fig.text(0.52, 0.48, "D", fontsize=8, ha="center", va="center")
 
 plt.savefig(f"fig2_n0_{n0}.pdf")

### 2021-03-14 figures/fig3.py

@@ -143,7 +143,9 @@ ax1.set_title(r"$\gamma=-1$")
 ax2.set_title(r"$\gamma=-5$")
 
 fig.tight_layout()
-#fig.text(0.05, 0.97, "A", fontsize=8, ha="center", va="center")
-#fig.text(0.05, 0.47, "B", fontsize=8, ha="center", va="center")
+fig.text(0.02, 0.98, "A", fontsize=8, ha="center", va="center")
+fig.text(0.52, 0.98, "B", fontsize=8, ha="center", va="center")
+fig.text(0.02, 0.48, "C", fontsize=8, ha="center", va="center")
+fig.text(0.52, 0.48, "D", fontsize=8, ha="center", va="center")
 
 plt.savefig(f"fig3_n0_{n0}.pdf")

### 2021-03-14 figures/fig4.py

@@ -332,7 +332,7 @@ plot_relate_curve(
     gen=1,
 )
 
-ax1.legend(frameon=False)
+ax1.legend()
 # ax1.set_xscale("log")
 # ax1.set_yscale("log")
 ax1.set_xlim(4000, 0)
@@ -344,7 +344,7 @@ ax2 = plt.subplot(4, 2, 2)
 plot_relate_curve(ax2, t, N0, line_style="-", lw=1, color=colors[0], label=pop0)
 plot_relate_curve(ax2, t, N1, line_style="--", lw=1, color=colors[1], label=pop1)
 
-ax2.legend(frameon=False)
+ax2.legend()
 ax2.set_xscale("log")
 ax2.set_yscale("log")
 ax2.set_xlim(1e6, 1e3)
@@ -370,7 +370,7 @@ for rho, c in zip([0, 5], [colors[2], colors[3]]):
             label=rf"{pop[:3]}, $\rho={rho}$",
         )
 
-ax3.legend(frameon=False, ncol=2)
+ax3.legend(ncol=2)
 ax3.set_xlim(ax1.get_xlim())
 ax3.set_ylabel(f"$\sigma_d^1$")
 ax3.set_title("Additive selection, no epistasis")
@@ -392,7 +392,7 @@ for rho, c in zip([0, 5], [colors[2], colors[3]]):
 ax4.set_xlim(ax2.get_xlim())
 ax4.set_title("Additive selection, no epistasis")
 ax4.set_xscale("log")
-ax4.legend(frameon=False, ncol=2)
+ax4.legend(ncol=2)
 
 ax5 = plt.subplot(4, 2, 5)
 
@@ -408,7 +408,7 @@ for eps, c in zip([-0.5, 0.5], [colors[4], colors[5]]):
             label=rf"{pop[:3]}, $\epsilon={eps}$",
         )
 
-ax5.legend(frameon=False, ncol=2)
+ax5.legend(ncol=2)
 ax5.set_xlim(ax1.get_xlim())
 ax5.set_ylabel(f"$\sigma_d^1$")
 ax5.set_title(r"Epistasis (with $\rho=1$)")
@@ -430,7 +430,7 @@ for eps, c in zip([-0.5, 0.5], [colors[4], colors[5]]):
 ax6.set_xlim(ax2.get_xlim())
 ax6.set_title(r"Epistasis (with $\rho=1$)")
 ax6.set_xscale("log")
-ax6.legend(frameon=False, ncol=2)
[…]

### 2021-03-14 figures/fig5.py

@@ -224,4 +224,11 @@ plot_between(pop, data, bs, ax5)
 
 fig.tight_layout()
 fig.subplots_adjust(hspace=0.3)
+fig.text(0.02, 0.95, "A", fontsize=8, ha="center", va="center")
+fig.text(0.02, 0.70, "B", fontsize=8, ha="center", va="center")
+fig.text(0.06, 0.38, "C", fontsize=8, ha="center", va="center")
+fig.text(0.30, 0.38, "D", fontsize=8, ha="center", va="center")
+fig.text(0.54, 0.38, "E", fontsize=8, ha="center", va="center")
+fig.text(0.77, 0.38, "F", fontsize=8, ha="center", va="center")
+
 plt.savefig(f"fig5.pdf")

### 2021-03-14 figures/fig6.py

@@ -311,5 +311,9 @@ ax4 = plt.subplot2grid((5, 3), (3, 2), rowspan=2)
 plot_inside_outside("KHV", ax4)
 
 fig.tight_layout()
+fig.text(0.02, 0.95, "A", fontsize=8, ha="center", va="center")
+fig.text(0.05, 0.38, "B", fontsize=8, ha="center", va="center")
+fig.text(0.38, 0.38, "C", fontsize=8, ha="center", va="center")
+fig.text(0.70, 0.38, "D", fontsize=8, ha="center", va="center")
 
-plt.savefig(f"fig6_multipanel.pdf")
+plt.savefig(f"fig6.pdf")

### 2021-03-23 Manuscript.Rmd
consolidate figures, incorporate comments

@@
 positive LD between missense mutations within genes is driven by strong
-positive correlations between pairs of mutations that fall within the same
-conserved domain, pointing to compensatory mutations or antagonistic epistasis
-as the prevailing mode of interaction within but not outside of conserved genic
-elements. The heterogeneous landscape of both mutational fitness effects and
-selective interactions within protein-coding genes calls for more refined
-inferences of the joint distribution of fitness and interactive effects, and
-the methods presented here should prove useful in that pursuit.
+positive allele-frequency correlations between pairs of mutations that fall
+within the same conserved domain, pointing to compensatory mutations or
+antagonistic epistasis as the prevailing mode of interaction within but not
+outside of conserved genic elements. The heterogeneous landscape of both
+mutational fitness effects and selective interactions within protein-coding
+genes calls for more refined inferences of the joint distribution of fitness
+and interactive effects, and the methods presented here should prove useful in
+that pursuit.
 
@@
 In their foundational paper, @Hill1966-gv recognized that linked selected
-variants reciprocally impede the efficacy of selection at each locus. Linked
-selection reduces the fixation probability of advantageous mutations and
-increases that of deleterious mutations compared to expectations under
-single-locus models [@Birky1988-jm]. Allele frequency dynamics and correlations
-of linked selected variants are also predicted to deviate from expectations
-without interference. Under a multiplicative fitness model, where the fitness
-reduction of a haplotype carrying multiple deleterious variants is equal to the
-product of the fitness reduction of each mutation independently, we should
-expect to see those mutation segregate on different haplotypes more often than
-together, leading to negative, or repulsion, linkage disequilibrium (LD),
-although the extent of LD depends non-trivially on the strength of selection
-and probability of recombination separating loci [@Hill1966-gv;
-@McVean2000-ox].
+variants reciprocally impede the efficacy of selection at each locus, a process
+known as selective interference. Linked selection reduces the fixation
+probability of advantageous mutations and increases that of deleterious
+mutations compared to expectations under single-locus models [@Birky1988-jm].
+Allele frequency dynamics and correlations of linked selected variants are also
+predicted to deviate from expectations without interference. Under a
+multiplicative fitness model, where the fitness reduction of a genome carrying
+multiple deleterious variants is equal to the product of the fitness reduction
+of each mutation independently, we expect to see net linkage disequilibrium
+(LD) equal to zero for unlinked sites [@Kondrashov1995-va]. But for linked
+loci, those mutations are expected to segregate on different haplotypes more
+often than together, leading to negative, or repulsion, LD, although the extent
+of LD depends non-trivially on the strength of selection and probability of
+recombination separating loci [@Hill1966-gv; @McVean2000-ox].
 
@@
 protein-coding genes (such as stop gains and losses, frameshifts, and other
-nonsense mutations) in both human and fruit fly populations, which they
-interpreted as evidence for widespread synergistic epistasis between these
-mutations. Within the past year or two, both @Sandler2020-vk and @Garcia2020-zn
-have reevaluated patterns of LD between coding variants in humans, fruit flies,
-and *Capsella grandiflora*, and suggested interference and dominance may
-instead be driving patterns of LD [@Garcia2020-zn] or questioned whether LD
-between loss-of-function variants is significantly different from zero
+nonsense mutations) in both human and fruit fly populations. They interpreted
+this as evidence for widespread synergistic epistasis between these mutations,
+in which the fitness reduction of multiple mutations is greater than the
+product of that of each individual mutations independently. Within the past
+year or two, both @Sandler2020-vk and @Garcia2020-zn have reevaluated patterns
+of LD between coding variants in humans, fruit flies, and *Capsella
+grandiflora*, and suggested interference and dominance may instead be driving
+patterns of LD [@Garcia2020-zn] or questioned whether LD between
+loss-of-function variants is significantly different from zero
 [@Sandler2020-vk].
@@
 many cases it is simpler to visualize summaries (such as $\E[D]$) or slices of
-the distribution (Figure \@ref(fig:fig1)) instead of the full three-dimensional
-distribution. Here, I focus on low-order LD statistics, such as $\E[D]$ and
-$\E[D^2]$, and their decay with recombination distance as these are statistics
-that are commonly used to test for interactions between loci. Instead of
-unnormalized $\E[D^2]$ and $\E[D]$, I consider expectations for $\sigma_d^2 =
-\E[D^2]/\E[p(1-p)q(1-q)]$ and $\sigma_d^1 = \E[D]/\E[p(1-p)q(1-q)]$, where $p$
-and $q$ are the derived allele frequencies at the left ($A/a$) and right
-($B/b$) loci, respectively. These normalized statistics have two benefits:
-first, the mutation rate cancels so that expectations are robust to assumptions
-about the per-base mutation rate, and second, analytic expectations for these
-quantities are available under neutrality at constant population size for
-comparison [@Ohta1971-yd].
+the distribution (Figures \@ref(fig:fig1) and \@ref(fig:hudson) ) instead of
+the full three-dimensional distribution. Here, I focus on low-order LD
+statistics, such as $\E[D]$ and $\E[D^2]$, and their decay with recombination
+distance as these are statistics that are commonly used to test for
+interactions between loci. Instead of unnormalized $\E[D^2]$ and $\E[D]$, I
+consider expectations for $\sigma_d^2 = \E[D^2]/\E[p(1-p)q(1-q)]$ and
+$\sigma_d^1 = \E[D]/\E[p(1-p)q(1-q)]$, where $p$ and $q$ are the derived allele
+frequencies at the left ($A/a$) and right ($B/b$) loci, respectively. These
+normalized statistics have two benefits: first, the mutation rate cancels so
+that expectations are robust to assumptions about the per-base mutation rate,
+and second, analytic expectations for these quantities are available under
+neutrality at constant population size for comparison [@Ohta1971-yd].
 
@@
 
+```{r fig1, echo=F, fig.cap="(ref:fig1cap)", fig.align="center", fig.pos="tbp!"}
+knitr::include_graphics("figures/fig1.pdf")
+```
+
+(ref:fig1cap) \textbf{Additive selection and epistasis.} (A and B) The
+distribution at stationarity of $AB$ haplotype counts in a sample size of 30,
+in which we observe 10 $A$ alleles at the left locus, and 10 $B$ alleles at the
+right locus. (C) The decay of normalized squared LD ($\sigma_d^2$) with
+recombination distance ($\rho=4N_er$) for pairs of variants with a given
+selection coefficient. The dashed line is the neutral expectation from
+@Ohta1971-yd, given in Equation \@ref(eq:ohta). (D) The decay of
+$\sigma_d^1=\E[D]/ \E[p(1-p)q(1-q)]$ for pairs of variants. The negative LD
+effect of Hill-Robertson interference is strongest for intermediate selection,
+with $\gamma\sim-1$ to $-5$, where $\gamma=4N_es$. **[The text does not refer
+to specific panels, 1c should give the equation for $\sigma_d^2$, and not sure
+what we get from 1a and 1b. Define rho in caption, and place other parameters
+in parentheses]** **[Move a, b to supp, consolidate c and d with fig 2]** And
+from figure 2 previously ...(A, B) Synergistic epistasis decreases
+$\sigma_d^2$, while antagonistic epistasis increases $\sigma_d^2$, though both
+are decreased in comparison to neutral expectations. (C, D) Epistasis expected
+signed LD away from zero, an effect which can extend over large recombination
+distances. Here, $\gamma$ is the selection coefficient for derived alleles $A$
+and $B$, assuming selection is equal at the two loci. $\epsilon=0$ is the case
+of pure additive selection with no epistasis, and $\gamma_{AB}=(\gamma_A +
+\gamma_B)(1+\epsilon)$. Dashed lines are expectations under neutrality for both
+$\sigma_d^2$ and $\sigma_d^1$. **[More explicitly define that positive e is
+synergistic, negative is antagonistic, and define formula in the methods]**
+**[caption for C, D reads a little trailing]**
+
 ### Additive selection and epistasis
@@
 recover the well known Hill-Robertson interference result of negative LD
-between selected mutations, which is strongest for pairs for mutations that
-have selection coefficients $\gamma=O(1)$, or $s \approx 1/2N_e$ (Figure
-\@ref(fig:fig1)D). For strongly deleterious mutations, LD is close to zero even
+between selected mutations, which is strongest for pairs of mutations that have
+selection coefficients $\gamma = O(1)$, or $s \approx 1/2N_e$ (Figure
+\@ref(fig:fig1)B). For strongly deleterious mutations, LD is close to zero even
 with tight linkage, as they almost always segregate at low enough frequencies
@@
 variants, with sign depending on the direction of epistatic interactions
-(Figure \@ref(fig:fig2)). Synergistic epistasis results in negative LD while
+(Figure \@ref(fig:fig1)). Synergistic epistasis results in negative LD while
 antagonistic epistasis results in positive LD, and large positive LD can occur
-even when the epistatic effect is relatively weak. Epistasis-induced LD can
-extend over long distances, especially for strongly deleterious mutations. For
-example, in Figure \@ref(fig:fig2)D even moderately deleterious mutations with
-population-size-scaled selection coefficients of $\gamma=-10$ show large mean
-LD that extends well beyond 100 kilobases. More strongly deleterious
-interacting mutations are expected to show large signed LD over much larger
-recombination distances.
-
-### Dominance and interference
+even when the epistatic effect is relatively weak **[example in supp]**.
+Epistasis-induced LD can extend over long distances, especially for strongly
+deleterious mutations. For example, in Figure \@ref(fig:fig1)F even moderately
+deleterious mutations with population-size-scaled selection coefficients of
+$\gamma=-10$ show large mean LD that extends well beyond 100 kilobases. More
+strongly deleterious interacting mutations are expected to show large signed LD
+over much larger recombination distances.
+
+### Dominance
 
@@
 recessive mutations ($h<1/2$) can have large negative signed LD (Figure
-\@ref(fig:fig3)D), although it decays rapidly with increasing distance between
+\@ref(fig:fig2)D), although it decays rapidly with increasing distance between
 loci and is roughly zero for $\rho \gg 1$. For weakly selected mutations with
@@
 
+```{r fig2, echo=F, fig.cap="(ref:fig2cap)", fig.align="center", fig.pos="tbp!"}
+knitr::include_graphics("figures/fig2.pdf")
+```
+
+(ref:fig2cap) **Dominance effects.** (A-B) Pairs of recessive mutations
+($h<0.5$) have higher squared LD than additive mutations, while pairs of
+dominant alleles have lower expected $D^2$. (C-D) Pairs of recessive, strongly
+deleterious mutations can have large negative LD, although this correlation
+decays more rapidly with larger recombination than LD under epistasis (Figure
+\@ref(fig:fig1)D and F). Notably, there is no simple relationship between strength of
+selection and dominance, particularly when $s$ is $O(1/N)$. **[Epistasis is
+zero in this setup, and rearrange panels to match figure 1]**
+
 ### General selection and gene-based dominance
 
+**[Struggled a bit with this section, make clearer and easier to parse]**
+
 While dominance and epistasis are well-characterized models of selection, a
-large family of selection models can be specified by the assigning unique
+large family of selection models can be specified by assigning unique
 selection coefficients to possible diploid pairs of haplotypes. If we assume
@@
 element, but a diploid individual carrying at least one copy that is free of
-mutations has minimal fitness loss. This is analogous to the single-locus
-dominance case, where $s_{Aa} < \frac{1}{2}s_{AA}$. In this "gene-based
-dominance" scenario [e.g., @Sanjak2017-zw], an $AB/ab$ genotype has higher
-fitness than an $Ab/aB$ type (a simple implementation of this model is given in
-the final column in Table \@ref(tab:selmodels)). Such a gene-based fitness
-model gives similar expected positive signed LD to the model of antagonistic
-epistasis (compare Figure \@ref(fig:fig4)E-F to G-H), although the
-interpretation of those two models can differ. It shows that with a highly
-parameterized space of possible general diploid selection models, multiple
-models with different biological interpretations can give similar patterns of
-expected $D$.
+mutations has minimal fitness loss. In this "gene-based dominance" scenario
+[e.g., @Sanjak2017-zw], an $AB/ab$ genotype has higher fitness than an $Ab/aB$
+type (a simple implementation of this model is given in the final column in
+Table \@ref(tab:selmodels)). Such a gene-based fitness model gives similar
+expected positive signed LD to the model of antagonistic epistasis (compare
+Figure \@ref(fig:fig3)E-F to G-H), although the interpretation of those two
+models can differ. It shows that with a highly parameterized space of possible
+general diploid selection models, multiple models with different biological
+interpretations can give similar patterns of expected $D$.
 
@@
 scenarios relevant to human demographic history, including bottlenecks and
-expansions. I explore two simple models (Figure \@ref(fig:fig4)A), one with an
+expansions. I explore two simple models (Figure \@ref(fig:fig3)A), one with an
 instantaneous expansion and another with a bottleneck followed by recovery. I
@@
 for the Yoruba from Ibidan, Nigeria (YRI) and Utahns of North and West European
-ancestry (CEU) (Figure \@ref(fig:fig4)B).
-
-For each of the four histories, Figure \@ref(fig:fig4)C-H shows the dynamics of
+ancestry (CEU) (Figure \@ref(fig:fig3)B).
+
+For each of the four histories, Figure \@ref(fig:fig3)C-H shows the dynamics of
 $\sigma_d^1$ for a given parameterization of two-locus selection, including
@@
 
+```{r fig3, echo=F, fig.cap="(ref:fig3cap)", fig.align="center", fig.pos="tbp!"}
+knitr::include_graphics("figures/fig3.pdf")
+```
+
+(ref:fig3cap) **The effects of demography on signed LD.** Left column:
+simulations under two toy models, one with an instantaneous expansion in the
+past, and the other with a 10-fold reduction in size followed by recovery.
+Right column: simulations under population size histories inferred using Relate
+[@Speidel2019-nj] applied to populations from
+@1000_Genomes_Project_Consortium2015-zq. In all simulations, $\gamma$ is set to
+$-2$ at both loci. (A-B) Model illustrations. Note the differences in
+scale. (C-D) Additive selection with no epistasis, showing fully linked
+($\rho=0$) and more loosely linked ($\rho=5$) pairs. Bottleneck events result
+in larger negative values of signed LD, but population size changes are not
+expected to change the sign of mean LD. (E-F) Synergistic ($\epsilon > 0$) and
+antagonistic ($\epsilon < 0$) epistasis. Again, bottleneck events result in
+larger signed LD. Here, $\epsilon=1/2$ is a model with $\gamma_A=-2$,
+$\gamma_B=-2$ and $\gamma_{AB}=-6$, and $\epsilon=-1/2$ is a model with
+$\gamma_A=-2$, $\gamma_B=-2$ and $\gamma_{AB}=-2$. (G-H) Dominance within loci
+results in negative LD, while dominance acting at the gene level results in
+positive LD. Here, $h=0.1$. Again, bottlenecks result in larger average LD in
+both cases.
+
 ## Signed LD within protein-coding genes
@@
 interactions for loss-of-function and missense mutations (Figure
-\@ref(fig:fig5)). Within each population, large measurement noise gives 95\%
+\@ref(fig:fig4)). Within each population, large measurement noise gives 95\%
 confidence intervals that overlap with zero across mutation class, although the
@@
 expectations between populations that have or have not gone through a
-bottleneck in their recent past (Figure \@ref(fig:fig4)). For pairs of missense
+bottleneck in their recent past (Figure \@ref(fig:fig3)). For pairs of missense
 mutations in different protein-coding genes, signed LD decays to zero fairly
@@
 linked (distance less that 50 kb).
+
+```{r fig4, echo=F, fig.cap="(ref:fig4cap)", fig.align="center", fig.pos="tbp!"}
+knitr::include_graphics("figures/fig4.pdf")
+```
+
+(ref:fig4cap) **LD within and between protein-coding genes.** (A-B) Pairs of
+variants in coding regions partitioned by classes of mutations.
+Loss-of-function mutations are defined as stop and start gains and losses and
+other nonsense mutations of high impact. In all populations, we observe
+positive mean signed LD for missense and synonymous variants, although the
+confidence interval of each measurement overlaps with zero. Similarly, we
+observe mean negative LD for pairs of loss-of-function variants within genes.
+There are only $\sim200-300$ pairs of loss-of-function variants that fall
+within the same gene across the populations tested here, causing large
+uncertainty in each measurement. Gray lines indicate 2 standard errors of the
+measurement, computed by bootstrapping of sets of genes. (C-F) While pairs of
+synonymous and missense mutations within genes are consistently positive, pairs
+between genes are not consistent in their sign across populations. LD for all
+populations analyzed here is shown in Figure \@ref(fig:genespops).
 
@@
 missense mutations within genes, I examine the role that annotated conserved
-domains in protein-coding genes play in driving this signal. Note that the low
-density of loss-of-function mutations within annotated domains in humans
-precludes analysis of LD between those mutations at the level of sub-elements
-within genes. Protein-coding domains are conserved elements of genes, often
-associated with some known functional or structural feature of a protein. As
-such, we would expect that purifying selection is stronger on average within
-domains than within the same gene but outside of a conserved domains. Indeed,
-the site-frequency spectrum (SFS) is skewed to lower frequencies for both
-missense and loss-of-function mutations within domains when compared to the
-same classes of mutations outside of domains, with much more negative values of
-Tajima's D within domains (Table \@ref(tab:tajimasD)). On the other hand, no
-difference is observed for synonymous mutations whether within or outside
-domains, suggesting roughly equivalent effects of selection (either direct or
-linked) on synonymous variation.
+domains in protein-coding genes play in driving this signal. Protein-coding
+domains are conserved elements of genes, often associated with some known
+functional or structural feature of a protein. As such, we would expect that
+purifying selection is stronger on average within domains than within the same
+gene but outside of a conserved domains. Indeed, the site-frequency spectrum
+(SFS) is skewed to lower frequencies for both missense and loss-of-function
+mutations within domains when compared to the same classes of mutations outside
+of domains, with much more negative values of Tajima's D within domains (Table
+\@ref(tab:tajimasD)). On the other hand, no difference is observed for
+synonymous mutations whether within or outside domains, suggesting roughly
+equivalent effects of selection (either direct or linked) on synonymous
+variation.
+
+```{r fig5, echo=F, fig.cap="(ref:fig5cap)", fig.align="center", fig.pos="tbp!"}
+knitr::include_graphics("figures/fig5.pdf")
+```
+
+(ref:fig5cap) **LD within and between annotated domains in protein-coding
+genes.** (A) Considering only variants that fall within annotated domains in
+protein-coding genes, pairs of mutations that fall within the same gene display
+larger positive LD than pairs that lie in different domains. This pattern is
+most striking for missense mutations, for which pairs within domains have
+average signed LD that is consistently larger than for pairs that span two
+domains, or for pairs of synonymous mutations that are within the same domain.
+(B-D) The slightly positive LD that we observe for missense and synonymous
+mutations gene-wide (Figure \@ref(fig:fig3)A) is largely driven by the subset
+of pairs mutations that fall within the same annotated domain. LD for all
+populations analyzed here isshown in Figure \@ref(fig:domainspops).
 
@@
 pairs of synonymous mutations that both fall within the same domain (Figure
-\@ref(fig:fig6)). This difference between pairs of missense and synonymous
+\@ref(fig:fig5)). This difference between pairs of missense and synonymous
 variants within domains is observed for both tightly linked ($< 10$bps) and
@@
 from models of either Hill-Robertson interference, synergistic epistasis, or
-dominance (Figures \@ref(fig:fig1)-\@ref(fig:fig3)). This observation is
+dominance (Figures \@ref(fig:fig1) and \@ref(fig:fig2)). This observation is
 opposite to those expectations, suggesting a different prevailing interactive
@@
 nearly equal levels of LD close to zero across domains, with missense mutations
-even slightly more negative than synonymous mutations (Figure \@ref(fig:fig6)).
+even slightly more negative than synonymous mutations (Figure \@ref(fig:fig5)).
 The interactive effect driving large LD within domains is therefore likely
@@
 distances as pairs of mutation within and between domains as surveyed in the
-previous section (Figures \@ref(fig:fig6)B-D and \@ref(fig:domainspops)).
+previous section (Figures \@ref(fig:fig5)B-D and \@ref(fig:domainspops)).
 Averaging over all pairs of mutations within the same gene but outside domains,
@@
 those within the same domain, missense variants have considerably lower LD than
-than synonymous variants. This difference between synonymous and missense pairs
-of variants largely disappears for SNPs separated by more than a few hundred or
-a thousand base pairs (Figure \@ref(fig:lddecay)), which suggests that
+synonymous variants. This difference between synonymous and missense pairs of
+variants largely disappears for SNPs separated by more than a few hundred or a
+thousand base pairs (Figure \@ref(fig:lddecay)). This suggests that
 Hill-Robertson interference is the primary mode of interaction between
 mutations falling outside of domains, as epistasis is expected to impact LD
-over larger distances than what is observed.
+over larger distances than what is observed, although the strength of epistasis
+is also likely to be a function of distance between mutations.
 
@@
 ## Clustered mutations account for positive LD among synonymous mutations
+
+**[Mashaal points out that simulations with structure leading to inbreeding, in which subpopulations have differing load, can lead to positive LD]**
 
@@
 
+**[Need to point out that confidence intervals are wide, and while the signal
+is consistent across populations, shared evolutionary history makes formal
+testing of significance difficult.]**
+
 In exploring or inferring multi-locus selective models, the number of plausible
@@
 of dominance and selection is underpowered using the SFS alone, but because
-signed LD is sensitive to the levels of dominance (Figure \@ref(fig:fig3)),
+signed LD is sensitive to the levels of dominance (Figure \@ref(fig:fig2)),
 inferring the DFE with dominance may be feasible using the joint distribution
@@
 To obtain the recursion equation under selection we consider drawing $n$
-lineages from generation $t$, which has has an expected sampling distribution
+lineages from generation $t$, which has an expected sampling distribution
 of haplotype counts given by $\Psi_n^t$. However, assuming $s\leq0$ for each
@@
 history scenario, I tracked the evolution of $\Psi_n(t)$ for varying selection
-models, and Figure \@ref(fig:fig4) plots the trajectory of $\sigma_d^1$ over
+models, and Figure \@ref(fig:fig3) plots the trajectory of $\sigma_d^1$ over
 time. The selection strength at both loci was fixed at $\gamma=2$ for all
@@
 rates given time bins, and population sizes are estimated as their inverses.
-Estimates for population sizes in the very recent past ($<2,000$ years)
-diverged, so I truncated the history over this time period and assumed a
-constant size from the most recent non-diverged bin.
+Estimates using Relate for population sizes in the very recent past ($<3,000$
+years, or $\approx 100$ generations) diverged, so I truncated the history over
+this time period and assumed a constant size from the most recent non-diverged
+bin.
 
@@
 annotated as either synonymous or nonsynonymous (including both missense and
-more damaging mutations such as stop gains, stop losses, and so on). I further
-subset to samples within each non-admixed population in the African, European,
-and East Asian continental groups. Signed LD is sensitive to ancestral state
-identification, so I only kept sites for which ancestral alleles were estimated
-with high confidence in both the VCF info field and the Thousand Genomes human
-ancestor reconstructed from a phylogeny of six primates.
+more damaging mutations such as stop gains, stop losses, and start losses). I
+further subset to samples within each non-admixed population in the African,
+European, and East Asian continental groups. Signed LD is sensitive to
+ancestral state identification, so I only kept sites for which ancestral
+alleles were estimated with high confidence in both the VCF info field and the
+Thousand Genomes human ancestor reconstructed from a phylogeny of six primates.
 
@@
 distance bins up to 500kb, and inside or outside of annotated domains within
-the same protein-coding genes. I used a dataset of all known conserved domains
-in humans compiled by @Stanek2020-pa.
+the same protein-coding genes. I used a dataset of annotated protein domains
+mapped to hg19 compiled by @Stanek2020-pa to determine if a given mutation
+falls within a domain or not.
 
@@
 # Acknowledgements
-- Discussions: Simon Gravel, Alex Diaz-Papkovich, Matthias Steinrücken, Eric Friedlander
-- Comments on ms:
-
-Asked from comments from
-- Kevin Thornton
-- Mashaal Sohail
-- Simon Gravel
-- Matthias Steinrücken
-
-\clearpage
-\newpage
-
-# Figures
-
-```{r fig1, echo=F, fig.cap="(ref:fig1cap)", fig.align="center"}
-knitr::include_graphics("figures/fig1.pdf")
-```
-
-(ref:fig1cap) \textbf{The two-locus sampling distribution and additive
-selection.} (A and B) The distribution of $AB$ haplotype counts in a sample
-size of 30, in which we observe 10 $A$ alleles at the left locus, and 10 $B$
-alleles at the right locus. (C) The decay of normalized squared LD
-($\sigma_d^2$) with recombination distance ($\rho=4N_er$) for pairs of variants
-with a given selection coefficient. The dashed line is the neutral expectation
-from @Ohta1971-yd, given in Equation \@ref(eq:ohta). (D) The decay of
-$\sigma_d^1=\E[D]/ \E[p(1-p)q(1-q)]$ for pairs of variants. The negative LD
-effect of Hill-Robertson interference is strongest for intermediate selection,
-with $\gamma\sim-1$ to $-5$, where $\gamma=4N_es$.
-
-```{r fig2, echo=F, fig.cap="(ref:fig2cap)", fig.align="center"}
-knitr::include_graphics("figures/fig2.pdf")
-```
-
-(ref:fig2cap) \textbf{Additive selection with epistasis.} (A, B) Synergistic
-epistasis decreases $D^2$, while antagonistic epistasis increases $D^2$. (C, D)
-Epistasis expected signed LD away from zero, an effect which can extend over
-large recombination distances. Here, $\gamma$ is the selection coefficient for
-derived alleles $A$ and $B$, assuming selection is equal at the two loci.
-$\epsilon=0$ is the case of pure additive selection with no epistasis, and
-$\gamma_{AB}=(\gamma_A + \gamma_B)(1+\epsilon)$. Dashed lines are expectations
-under neutrality for both $\sigma_d^2$ and $\sigma_d^1$.
-
-```{r fig3, echo=F, fig.cap="(ref:fig3cap)", fig.align="center"}
-knitr::include_graphics("figures/fig3.pdf")
-```
-
-(ref:fig3cap) **Dominance effects.** (A-B) Pairs of recessive mutations
-($h<0.5$) have higher squared LD than additive mutations, while pairs of
-dominant alleles have lower expected $D^2$. (C-D) Pairs of recessive, strongly
-deleterious mutations can have large negative LD, although this correlation
-decays more rapidly with larger recombination than LD under epistasis (Figure
-\@ref(fig:fig2)D). Notably, there is no simple relationship between strength of
-selection and dominance, particularly when $s$ is $O(1/N)$.
-
-```{r fig4, echo=F, fig.cap="(ref:fig4cap)", fig.align="center"}
-knitr::include_graphics("figures/fig4.pdf")
-```
-
-(ref:fig4cap) **The effects of demography on signed LD.** Left column:
-simulations under two toy models, one with an instantaneous expansion in the
-past, and the other with a 10-fold reduction in size followed by recovery.
-Right column: simulations under population size histories inferred using Relate
-[@Speidel2019-nj] applied to populations from
-@1000_Genomes_Project_Consortium2015-zq. In all simulations, $\gamma$ is set to
-$-2$ at both loci. (A-B) Model illustrations. Note the differences in
-scale. (C-D) Additive selection with no epistasis, showing fully linked
-($\rho=0$) and more loosely linked ($\rho=5$) pairs. Bottleneck events result
-in larger negative values of signed LD, but population size changes are not
-expected to change the sign of mean LD. (E-F) Synergistic ($\epsilon > 0$) and
-antagonistic ($\epsilon < 0$) epistasis. Again, bottleneck events result in
-larger signed LD. Here, $\epsilon=1/2$ is a model with $\gamma_A=-2$,
-$\gamma_B=-2$ and $\gamma_{AB}=-6$, and $\epsilon=-1/2$ is a model with
-$\gamma_A=-2$, $\gamma_B=-2$ and $\gamma_{AB}=-2$. (G-H) Dominance within loci
-results in negative LD, while dominance acting at the gene level results in
-positive LD. Here, $h=0.1$. Again, bottlenecks result in larger average LD in
-both cases.
-
-```{r fig5, echo=F, fig.cap="(ref:fig5cap)", fig.align="center"}
-knitr::include_graphics("figures/fig5.pdf")
-```
-
-(ref:fig5cap) **LD within and between protein-coding genes.** (A-B) Pairs of
-variants in coding regions partitioned by classes of mutations.
-Loss-of-function mutations are defined as stop and start gains and losses and
-other nonsense mutations of high impact. In all populations, we observe
-positive mean signed LD for missense and synonymous variants, although the
-confidence interval of each measurement overlaps with zero. Similarly, we
-observe mean negative LD for pairs of loss-of-function variants within genes.
-There are only $\sim200-300$ pairs of loss-of-function variants that fall
-within the same gene across the populations tested here, causing large
-uncertainty in each measurement. Gray lines indicate 2 standard errors of the
-measurement, computed by bootstrapping of sets of genes. (C-F) While pairs of
-synonymous and missense mutations within genes are consistently positive, pairs
-between genes are not consistent in their sign across populations. LD for all
-populations analyzed here are shown in Figure \@ref(fig:genespops).
-
-```{r fig6, echo=F, fig.cap="(ref:fig6cap)", fig.align="center"}
-knitr::include_graphics("figures/fig6.pdf")
-```
-
-(ref:fig6cap) **LD within and between annotated domains in protein-coding
-genes.** (A) Considering only variants that fall within annotated domains in
-protein-coding genes, pairs of mutations that fall within the same gene display
-larger positive LD than pairs that lie in different domains. This pattern is
-most striking for missense mutations, for which pairs within domains have
-average signed LD that is consistently larger than for pairs that span two
-domains, or for pairs of synonymous mutations that are within the same domain.
-(B-D) The slightly positive LD that we observe for missense and synonymous
-mutations gene-wide (Figure \@ref(fig:fig4)A) is largely driven by the subset
-of pairs mutations that fall within the same annotated domain. LD for all
-populations analyzed here are shown in Figure \@ref(fig:domainspops).
-
-\clearpage
-\newpage
+
+I am grateful to Alex Diaz-Papkovich, Eric Friedlander, Simon Gravel, Mashaal
+Sohail, Matthias Steinrücken, and Kevin Thornton for helpful discussions and
+feedback on earlier versions of this manuscript that improved this work.
 

### 2021-03-23 figures/fig1.py

@@ -31,16 +31,16 @@ nB = 10
 
 data_fname = f"fig1_data_ns_{n0}_{n}.bp"
 try:
-    data = pickle.load(open(data_fname, "rb"))
+    data1 = pickle.load(open(data_fname, "rb"))
 except IOError:
-    data = {}
+    data1 = {}
 
-if "A" not in data.keys():
-    data["A"] = {}
+if "A" not in data1.keys():
+    data1["A"] = {}
     # stores the nAB distribution with n=30, nA=10, nB=10 for nuetrality
     # and gamma=-5, with rho=1 and rho=10
     for rho in [1, 10]:
-        data["A"][rho] = {}
+        data1["A"][rho] = {}
         for gamma in [0, -5]:
             sel_params = [2 * gamma, gamma, gamma]
             F0 = moments.TwoLocus.Demographics.equilibrium(
@@ -48,23 +48,23 @@ if "A" not in data.keys():
             )
             F = F0.project(n)
             counts, pAB = moments.TwoLocus.Util.pAB(F, nA, nB)
-            data["A"][rho][gamma] = pAB
+            data1["A"][rho][gamma] = pAB
             print("computed pAB for rho, gamma =", rho, ",", gamma)
-    pickle.dump(data, open(data_fname, "wb+"))
+    pickle.dump(data1, open(data_fname, "wb+"))
 
 gammas = [0, -0.1, -1, -5, -10]
 
-if "C" not in data.keys():
+if "C" not in data1.keys():
     rhos = np.logspace(-2, np.log10(30), 20)
-    data["C"] = {}
-    data["C"]["rhos"] = rhos
-    data["C"]["Ohta_Kimura"] = (5 + rhos / 2) / (11 + 13 * rhos / 2 + rhos ** 2 / 2)
+    data1["C"] = {}
+    data1["C"]["rhos"] = rhos
+    data1["C"]["Ohta_Kimura"] = (5 + rhos / 2) / (11 + 13 * rhos / 2 + rhos ** 2 / 2)
     # LD curves for gamma = 0, -5, acorss rhos, for "all" SNPs, and nA=nB=2, 3, 4, 10
     for gamma in gammas:
         sel_params = [2 * gamma, gamma, gamma]
-        data["C"][gamma] = {}
+        data1["C"][gamma] = {}
         for cond in ["all", 2, 3, 4, 10]:
[…]

### 2021-03-23 figures/fig2.py

@@ -1,6 +1,4 @@
-# Figure 2: Additive selection with epistasis.  In this epistasis model, sAB =
-# (1 + epsilon) * (sA + sB).  Smaller, single-column figure showing the effect
-# of synergistic epistasis for weakly and moderately selected variants.
+# Figure 3: Dominance selection model.
 import sys
 import moments.TwoLocus
 import matplotlib.pylab as plt
@@ -19,12 +17,11 @@ matplotlib.rc("legend", fontsize=6)
 
 from bokeh.palettes import Colorblind
 
-## got back and run with larger n0
-n0 = int(sys.argv[1])  # larger sample size for improved accuracy
+n0 = 70  # larger sample size for improved accuracy
 n = 30  # the projection size
 
-epsilons = [-0.5, 0.0, 0.5, 1.0]
-gammas = [-1, -10]
+hs = [0.0, 0.1, 0.5, 1.0]
+gammas = [-1, -5]
 conditions = ["all", 2, 3, 4, 10]
 
 data_fname = f"fig2_data_ns_{n0}_{n}.bp"
@@ -32,40 +29,41 @@ try:
     data = pickle.load(open(data_fname, "rb"))
 except IOError:
     data = {}
+    data["ns"] = (n0, n)
     # stores the sigma_d^2 and sigma_d^1 curves for all snps and
     rhos = np.logspace(-2, np.log10(30), 20)
     data["rhos"] = rhos
     data["Ohta_Kimura"] = (5 + rhos / 2) / (11 + 13 * rhos / 2 + rhos ** 2 / 2)
     for gamma in gammas:
-        for eps in epsilons:
-            print("Running with  gamma, eps = ", gamma, ",", eps, end=" ")
-            sel_params = moments.TwoLocus.Util.additive_epistasis(gamma, epsilon=eps)
-            data[(gamma, eps)] = {cond: {"sd1": [], "sd2": []} for cond in conditions}
+        for h in hs:
+            print("Running gamma, h = ", gamma, ",", h, end=" ", flush=True)
+            sel_params = moments.TwoLocus.Util.simple_dominance(gamma, h=h)
+            data[(gamma, h)] = {cond: {"sd1": [], "sd2": []} for cond in conditions}
             for rho in rhos:
-                print(".", end="")
[…]

### 2021-03-23 figures/fig3.py

@@ -1,4 +1,11 @@
-# Figure 3: Dominance selection model.
+# Figure 4: Demography and trajectories.
+# Two columns, left is simple size changes: expansion and bottleneck+recovery
+# Right column has human-like domographic history inferred by Relate
+# Rows: all with rho = 1
+# 1: illustration of demography
+# 2: (a) Additive selection w/ gamma=-1, (b) negative epistasis, gamma=-1, (c) negative epistasis, gamma=-10
+# 3: (a) Simple dominance, gamma=-1, (b) simple dominance, gamma=-10
+# 4: (a) Some gene-based selection model
 import sys
 import moments.TwoLocus
 import matplotlib.pylab as plt
@@ -17,135 +24,469 @@ matplotlib.rc("legend", fontsize=6)
 
 from bokeh.palettes import Colorblind
 
-n0 = int(sys.argv[1])  # larger sample size for improved accuracy
-n = 30  # the projection size
+n = 30  # the sample size
 
-hs = [0.0, 0.1, 0.5, 1.0]
-gammas = [-1, -5]
-conditions = ["all", 2, 3, 4, 10]
 
-data_fname = f"fig3_data_ns_{n0}_{n}.bp"
+def piecewise_constant(
+    nus,
+    Ts,
+    n,
+    rho=0.0,
+    gamma=0,
+    h=None,
+    dom="simple",
+    eps=None,
+    track_stats=True,
+    spacing=3,
+    conditions=[],
+):
+    if track_stats:
+        stats = {"t": [], "sd1": {"all": [],}, "sd2": {"all": [],}}
+        for cond in conditions:
+            stats["sd1"][cond] = []
+            stats["sd2"][cond] = []
+    if gamma == 0:
+        sel_params = None
+        sel_params_general = None
+    elif h is None and eps is None:
+        sel_params = moments.TwoLocus.Util.additive_epistasis(gamma)
+        sel_params_general = None
+    elif h is not None:
+        sel_params = None
+        if dom == "simple":
+            sel_params_general = moments.TwoLocus.Util.simple_dominance(gamma, h=h)
+        elif dom == "gene":
+            sel_params_general = moments.TwoLocus.Util.gene_based_dominance(gamma, h=h)
+    elif eps is not None:
[…]

### 2021-03-23 figures/fig4.py

@@ -1,11 +1,4 @@
-# Figure 4: Demography and trajectories.
-# Two columns, left is simple size changes: expansion and bottleneck+recovery
-# Right column has human-like domographic history inferred by Relate
-# Rows: all with rho = 1
-# 1: illustration of demography
-# 2: (a) Additive selection w/ gamma=-1, (b) negative epistasis, gamma=-1, (c) negative epistasis, gamma=-10
-# 3: (a) Simple dominance, gamma=-1, (b) simple dominance, gamma=-10
-# 4: (a) Some gene-based selection model
+# Figure 5: LD within and between genes
 import sys
 import moments.TwoLocus
 import matplotlib.pylab as plt
@@ -24,469 +17,218 @@ matplotlib.rc("legend", fontsize=6)
 
 from bokeh.palettes import Colorblind
 
-n = 30  # the sample size
+data = {}
 
+bs = {}
 
-def piecewise_constant(
-    nus,
-    Ts,
-    n,
-    rho=0.0,
-    gamma=0,
-    h=None,
-    dom="simple",
-    eps=None,
-    track_stats=True,
-    spacing=3,
-    conditions=[],
-):
-    if track_stats:
-        stats = {"t": [], "sd1": {"all": [],}, "sd2": {"all": [],}}
-        for cond in conditions:
-            stats["sd1"][cond] = []
-            stats["sd2"][cond] = []
-    if gamma == 0:
-        sel_params = None
-        sel_params_general = None
-    elif h is None and eps is None:
-        sel_params = moments.TwoLocus.Util.additive_epistasis(gamma)
-        sel_params_general = None
-    elif h is not None:
-        sel_params = None
-        if dom == "simple":
-            sel_params_general = moments.TwoLocus.Util.simple_dominance(gamma, h=h)
-        elif dom == "gene":
-            sel_params_general = moments.TwoLocus.Util.gene_based_dominance(gamma, h=h)
-    elif eps is not None:
-        sel_params = moments.TwoLocus.Util.additive_epistasis(gamma, epsilon=eps)
-        sel_params_general = None
-    F = moments.TwoLocus.Demographics.equilibrium(
[…]

### 2021-03-23 figures/fig5.py

@@ -1,4 +1,4 @@
-# Figure 5: LD within and between genes
+# Figure 6: LD within and between gene domains
 import sys
 import moments.TwoLocus
 import matplotlib.pylab as plt
@@ -30,12 +30,12 @@ populations = {
 for continent, pops in populations.items():
     for pop in pops:
         ld_pop = pickle.load(
-            open(f"../analysis/parsed_data/{pop}.unphased.within.between.bp", "rb")
+            open(f"../analysis/parsed_data/{pop}.unphased.domains.bp", "rb")
         )
         data[pop] = ld_pop
         bs_pop = pickle.load(
             open(
-                f"../analysis/parsed_data/{pop}.unphased.within.between.sigmad1.bootstrap_std_err.bp",
+                f"../analysis/parsed_data/{pop}.unphased.domains.sigmad1.bootstrap_std_err.bp",
                 "rb",
             )
         )
@@ -43,19 +43,165 @@ for continent, pops in populations.items():
 
 ms = 2
 lw = 0.5
+markers = ["x", "+", "1"]
+colors = Colorblind[8]
+
+def plot_inside_outside(pop, ax, legend=False, ylabel=False, xlabel=True):
+    # plot missense and synonymous for each, in the order:
+    # gene-wide, within domains, between domains, all outside domains, matched to
+    #   distances within, matched to distances between
+    to_plot = []
+    err = []
+    # add gene-wide values
+    data_all = pickle.load(
+        open(f"../analysis/parsed_data/{pop}.unphased.within.between.bp", "rb")
+    )
+    bs_all = pickle.load(
+        open(
+            f"../analysis/parsed_data/{pop}.unphased.within.between.sigmad1.bootstrap_std_err.bp",
+            "rb",
+        )
+    )
+    to_plot.append(
+        data_all["ld_within"]["synonymous"][3] / data_all["ld_within"]["synonymous"][2]
+    )
+    to_plot.append(
+        data_all["ld_within"]["missense"][3] / data_all["ld_within"]["missense"][2]
+    )
+    err.append(bs_all["bs_within"]["synonymous"])
[…]

### 2021-03-23 figures/fig6.py

@@ -1,319 +1,0 @@
-# Figure 6: LD within and between gene domains
-import sys
-import moments.TwoLocus
-import matplotlib.pylab as plt
-import numpy as np
-import pickle
-
-# set font sizes
-import matplotlib
-
-plt.rcParams["legend.title_fontsize"] = "xx-small"
-matplotlib.rc("xtick", labelsize=6)
-matplotlib.rc("ytick", labelsize=6)
-matplotlib.rc("axes", labelsize=7)
-matplotlib.rc("axes", titlesize=7)
-matplotlib.rc("legend", fontsize=6)
-
-from bokeh.palettes import Colorblind
-
-data = {}
-
-bs = {}
-
-populations = {
-    "Africa": ["ESN", "GWD", "LWK", "MSL", "YRI"],
-    "Europe": ["CEU", "FIN", "GBR", "IBS", "TSI"],
-    "East Asia": ["CDX", "CHB", "CHS", "JPT", "KHV"],
-}
-
-for continent, pops in populations.items():
-    for pop in pops:
-        ld_pop = pickle.load(
-            open(f"../analysis/parsed_data/{pop}.unphased.domains.bp", "rb")
-        )
-        data[pop] = ld_pop
-        bs_pop = pickle.load(
-            open(
-                f"../analysis/parsed_data/{pop}.unphased.domains.sigmad1.bootstrap_std_err.bp",
-                "rb",
-            )
-        )
-        bs[pop] = bs_pop
-
-ms = 2
-lw = 0.5
-markers = ["x", "+", "1"]
-colors = Colorblind[8]
-
-def plot_inside_outside(pop, ax, legend=False, ylabel=False, xlabel=True):
-    # plot missense and synonymous for each, in the order:
-    # gene-wide, within domains, between domains, all outside domains, matched to
-    #   distances within, matched to distances between
-    to_plot = []
-    err = []
-    # add gene-wide values
-    data_all = pickle.load(
-        open(f"../analysis/parsed_data/{pop}.unphased.within.between.bp", "rb")
-    )
-    bs_all = pickle.load(
-        open(
-            f"../analysis/parsed_data/{pop}.unphased.within.between.sigmad1.bootstrap_std_err.bp",
-            "rb",
-        )
-    )
-    to_plot.append(
[…]

### 2021-03-23 figures/hudson_comparison.py

@@ -1,0 +1,134 @@
+# Unconditioned vs allele count-conditioned LD.  Four panels, first two showing
+# Hudson slices in the with rho=1 and rho=10 for neutrality and with gamma=-5
+# at both loci. Second two panels show \sigma_d^2 and \sigma_d^1 for nuetrality
+# and using all SNPs with gamma=-5, as well as conditioning on doubletons,
+# tripletons, and few other frequencies.
+
+import moments.TwoLocus
+import matplotlib.pylab as plt
+import numpy as np
+import pickle
+
+# set font sizes
+import matplotlib
+
+plt.rcParams["legend.title_fontsize"] = "xx-small"
+matplotlib.rc("xtick", labelsize=6)
+matplotlib.rc("ytick", labelsize=6)
+matplotlib.rc("axes", labelsize=7)
+matplotlib.rc("axes", titlesize=7)
+matplotlib.rc("legend", fontsize=6)
+
+from bokeh.palettes import Colorblind
+
+colors = Colorblind[8]
+
+n0 = 40  # larger sample size for improved accuracy
+n = 30  # the projection size
+
+nA = 10
+nB = 10
+
+data_fname = f"fig1_data_ns_{n0}_{n}.bp"
+try:
+    data = pickle.load(open(data_fname, "rb"))
+except IOError:
+    data = {}
+
+if "A" not in data.keys():
+    data["A"] = {}
+    # stores the nAB distribution with n=30, nA=10, nB=10 for nuetrality
+    # and gamma=-5, with rho=1 and rho=10
+    for rho in [1, 10]:
+        data["A"][rho] = {}
+        for gamma in [0, -5]:
+            sel_params = [2 * gamma, gamma, gamma]
+            F0 = moments.TwoLocus.Demographics.equilibrium(
+                n0, rho=rho, sel_params=sel_params
+            )
+            F = F0.project(n)
+            counts, pAB = moments.TwoLocus.Util.pAB(F, nA, nB)
+            data["A"][rho][gamma] = pAB
+            print("computed pAB for rho, gamma =", rho, ",", gamma)
+    pickle.dump(data, open(data_fname, "wb+"))
+
+gammas = [0, -0.1, -1, -5, -10]
+
+if "C" not in data.keys():
[…]

### 2021-03-24 Manuscript.Rmd
ready for last pass of edits

@@
 
-(ref:fig1cap) \textbf{Additive selection and epistasis.} (A and B) The
-distribution at stationarity of $AB$ haplotype counts in a sample size of 30,
-in which we observe 10 $A$ alleles at the left locus, and 10 $B$ alleles at the
-right locus. (C) The decay of normalized squared LD ($\sigma_d^2$) with
-recombination distance ($\rho=4N_er$) for pairs of variants with a given
-selection coefficient. The dashed line is the neutral expectation from
-@Ohta1971-yd, given in Equation \@ref(eq:ohta). (D) The decay of
-$\sigma_d^1=\E[D]/ \E[p(1-p)q(1-q)]$ for pairs of variants. The negative LD
-effect of Hill-Robertson interference is strongest for intermediate selection,
-with $\gamma\sim-1$ to $-5$, where $\gamma=4N_es$. **[The text does not refer
-to specific panels, 1c should give the equation for $\sigma_d^2$, and not sure
-what we get from 1a and 1b. Define rho in caption, and place other parameters
-in parentheses]** **[Move a, b to supp, consolidate c and d with fig 2]** And
-from figure 2 previously ...(A, B) Synergistic epistasis decreases
-$\sigma_d^2$, while antagonistic epistasis increases $\sigma_d^2$, though both
-are decreased in comparison to neutral expectations. (C, D) Epistasis expected
-signed LD away from zero, an effect which can extend over large recombination
-distances. Here, $\gamma$ is the selection coefficient for derived alleles $A$
-and $B$, assuming selection is equal at the two loci. $\epsilon=0$ is the case
-of pure additive selection with no epistasis, and $\gamma_{AB}=(\gamma_A +
-\gamma_B)(1+\epsilon)$. Dashed lines are expectations under neutrality for both
-$\sigma_d^2$ and $\sigma_d^1$. **[More explicitly define that positive e is
-synergistic, negative is antagonistic, and define formula in the methods]**
-**[caption for C, D reads a little trailing]**
+(ref:fig1cap) \textbf{Additive selection and epistasis.} Left panels show
+expectations for the decay of $\sigma_d^2 = \E[D^2] / \E[p(1-p)q(1-q)]$ with
+recombination distance, and right panels show expectations for $\sigma_d^1 =
+\E[D] / \E[p(1-p)q(1-q)]$. The scaled recombination rate is given by
+$\rho=4N_er$, where $r$ is the probability of recombination between two loci.
+All expectations are at stationarity, and dashed lines are expectations under
+neutrality (Equation \@ref(eq:ohta)). (A-B) Additive selection, with equal
+selection strength at both loci. Stronger selection reduces $\sigma_d^2$ (A),
+and in (B), we see that the negative LD effect of Hill-Robertson interference
+is strongest for intermediate selection ($\gamma \approx -1$ to $-5$, where
+$\gamma=4N_es$). (C-F) Epistasis with additive selection. Synergistic epistasis
+(\(\epsilon < 0\)) decreases $\sigma_d^2$ while antagonistic epistasis
+(\(\epsilon > 0\)) increases it compared to expectations under a multiplicative
+fitness model ($\epsilon=0$). Epistasis also pushes signed LD ($\sigma_d^1$)
+away from zero, with synergistic epistasis decreasing LD for negatively
+selected mutations and antagonistic epistasis increasing LD. The deviation of
+$\sigma_d^1$ from zero under epistasis can extend for long recombination
+distances, particularly for stronger selection.
 
@@
 antagonistic epistasis results in positive LD, and large positive LD can occur
-even when the epistatic effect is relatively weak **[example in supp]**.
-Epistasis-induced LD can extend over long distances, especially for strongly
-deleterious mutations. For example, in Figure \@ref(fig:fig1)F even moderately
-deleterious mutations with population-size-scaled selection coefficients of
-$\gamma=-10$ show large mean LD that extends well beyond 100 kilobases. More
-strongly deleterious interacting mutations are expected to show large signed LD
-over much larger recombination distances.
+even when the epistatic effect is relatively weak. Epistasis-induced LD can
+extend over long distances, especially for strongly deleterious mutations. For
+example, in Figure \@ref(fig:fig1)F even moderately deleterious mutations with
+population-size-scaled selection coefficients of $\gamma=-10$ show large mean
+LD that extends well beyond 100 kilobases. More strongly deleterious
+interacting mutations are expected to show large signed LD over much larger
+recombination distances.
 
@@
 
-(ref:fig2cap) **Dominance effects.** (A-B) Pairs of recessive mutations
+(ref:fig2cap) **Dominance effects.** (A and C) Pairs of recessive mutations
 ($h<0.5$) have higher squared LD than additive mutations, while pairs of
-dominant alleles have lower expected $D^2$. (C-D) Pairs of recessive, strongly
+dominant mutations have lower expected $D^2$. (B and D) Pairs of recessive
 deleterious mutations can have large negative LD, although this correlation
 decays more rapidly with larger recombination than LD under epistasis (Figure
-\@ref(fig:fig1)D and F). Notably, there is no simple relationship between strength of
-selection and dominance, particularly when $s$ is $O(1/N)$. **[Epistasis is
-zero in this setup, and rearrange panels to match figure 1]**
+\@ref(fig:fig1)D and F). Notably, there is no simple relationship between
+strength of selection and dominance, particularly when $s$ is $O(1/N)$. Note
+that this setup does not include epistatic effects (i.e., $\epsilon=0$).
 
 ### General selection and gene-based dominance
-
-**[Struggled a bit with this section, make clearer and easier to parse]**
 
@@
 large family of selection models can be specified by assigning unique
-selection coefficients to possible diploid pairs of haplotypes. If we assume
-the diploid genotype that is homozygous for the ancestral allele ($ab/ab$) has
+selection coefficients to each possible diploid pairs of haplotypes. If we assume
+the diploid genotype that is homozygous for the ancestral alleles ($ab/ab$) has
 fitness 1, then there are nine other possible diploid two-locus genotypes that
@@
 Figure \@ref(fig:fig3)E-F to G-H), although the interpretation of those two
-models can differ. It shows that with a highly parameterized space of possible
-general diploid selection models, multiple models with different biological
-interpretations can give similar patterns of expected $D$.
+models can differ. With a highly parameterized space of possible general
+diploid selection models, multiple models with different biological
+interpretations can give similar patterns of expected signed LD.
 
@@
 interactions for loss-of-function and missense mutations (Figure
-\@ref(fig:fig4)). Within each population, large measurement noise gives 95\%
+\@ref(fig:fig4)). Within each population, measurement noise gives 95\%
 confidence intervals that overlap with zero across mutation class, although the
@@
 
-**[Mashaal points out that simulations with structure leading to inbreeding, in which subpopulations have differing load, can lead to positive LD]**
-
 In a single randomly mating population, neutral mutations are expected to have
@@
 mutations, and any interference between selected synonymous variants should
-lead to negative LD. @Sandler2020-vk proposed that spatial population structure
-may be responsible for the increase of LD for both synonymous and missense
-variants. They argued that every natural population lives in some geography
-that will necessarily induce some structure, and that migration, admixture, or
-spatial structure could increase average signed LD above zero.
-
-Under standard population genetic assumptions of constant mutation and
-recombination rates and independent mutation events, however, migration and
-admixture do not result in nonzero $\E[D]$. Rather, for a set of mutations with
-*given* differences in allele frequencies between source populations, $D$ can
-be nonzero in an admixed population between those particular mutations even
-when $D$ is zero in the source populations [@Cavalli-Sforza1971-jv]. But this
-result relies on conditioning on directional differences in allele frequencies
-between populations. Expected mean allele frequency difference in the source
+lead to negative LD.
+
+One proposal is that spatial population structure may be responsible for the
+increase of LD observed between synonymous variants, as every natural
+population lives in some geography that will necessarily induce some structure.
+However, neutral evolution alone cannot create this effect. Under standard
+population genetic assumptions of constant mutation and recombination rates and
+independent mutation events, migration, admixture, and spatial structure do not
+result in nonzero $\E[D]$. Rather, for a set of mutations with *given*
+differences in allele frequencies between source populations, $D$ can be
+nonzero in an admixed population between those particular mutations even when
+$D$ is zero in the source populations [@Cavalli-Sforza1971-jv]. But this result
+relies on conditioning on directional differences in allele frequencies between
+populations. Expected mean allele frequency difference in the source
 populations is zero ($\E[(p_1-p_2)(q_1-q_2)] = 0$), as differences are equally
 likely to be positive or negative, and resulting $\E[D]$ is still zero when
-taking genome-wide averages. Furthermore, @Sandler2020-vk used forward
-simulations with an explicit spatial model and a multiplicative fitness
-function to explore the joint effect of structure and interference, and the
-positive LD they observed under these scenarios extends over much larger
-distances than observed in the data. Instead, positive LD between synonymous
-mutations decays to zero within just a few hundred base (Figures
-\@ref(fig:lddecay) and \@ref(fig:lddecaydomain)).
+taking genome-wide averages. Instead, @Sohail2017-zq and @Sandler2020-vk used
+forward simulations under multi-population or explicitly spatial models and a
+multiplicative fitness function to explore the joint effect of structure,
+assortative mating, and interference, scenarios that can lead to nonzero LD.
+@Sandler2020-vk find positive LD among neutral mutations in their simulation,
+although it extends over much larger distances than observed in the data.
+Instead, positive LD in data between synonymous mutations decays to zero within
+just a few hundred bases (see Figures \@ref(fig:lddecay) and
+\@ref(fig:lddecaydomain), and Figure 5 from @Sandler2020-vk).
 
@@
 
-**[Need to point out that confidence intervals are wide, and while the signal
-is consistent across populations, shared evolutionary history makes formal
-testing of significance difficult.]**
-
-In exploring or inferring multi-locus selective models, the number of plausible
-selection scenarios becomes quite large as we relax the strict assumptions of
-additivity and multiplicative fitness effects. The inclusion of dominance
-effects, epistasis, or other interactions leads to a rapid increase in the
-number of parameters to consider. This makes performing forward simulations
-that span the range of all such selective interaction scenarios burdensome.
+When partitioning measurements of LD by mutation classes or regions within
+genes, the decreasing number of pairwise comparisons leads to large estimated
+measurement noise. Within each population, confidence intervals of observed
+$\sigma_d^1$ often overlap with zero or overlap with that of other classes of
+mutations. While observed patterns are remarkably consistent across the 15
+populations considered here, their joint evolutionary histories make formal
+testing of significance difficult due to shared variation, so that cannot be
+treated as independent measurements. Extensive forward simulations will likely
+be required to more thoroughly assess significance.
+
+From a modeling perspective, in exploring or inferring multi-locus selective
+models, the number of plausible selection scenarios becomes quite large as we
+relax the strict assumptions of additivity and multiplicative fitness effects.
+The inclusion of dominance effects, epistasis, or other interactions leads to a
+rapid increase in the number of parameters to consider. This makes performing
+forward simulations that span the range of all such selective interaction
+scenarios burdensome.
 
@@
 $ab$ haplotypes. The standard multiplicative fitness function assumes that
-$s_{AB} \approx s_{A} + s_{B}$, and the effect of epistasis can be written as
-$s_{AB} = (s_{A} + s_{B}) (1 + \epsilon)$, so that $\epsilon > 0$ implies
+$s_{AB} \approx s_{A} + s_{B}$, and a model for epistasis can be written as
+\[s_{AB} = (s_{A} + s_{B}) (1 + \epsilon),\] so that $\epsilon > 0$ implies
 synergistic epistasis, while $\epsilon < 0$ implies antagonistic epistasis.

### 2021-03-24 SupportingInformation.Rmd
consolidate figures, incorporate comments
ready for last pass of edits

@@
 
+```{r hudson, echo=F, fig.cap="(ref:fighudson)", fig.align="center"}
+knitr::include_graphics("figures/hudson_comparison.pdf")
+```
+
+(ref:fighudson) The distribution at stationarity of AB haplotype counts in a
+sample size of 30, in which we observe 10 A alleles at the left locus, and 10 B
+alleles at the right locus.
+
 ```{r jackknife, echo=F, fig.cap="(ref:figjackknife)", fig.align="center"}

### 2021-03-25 Manuscript.Rmd
edits throughout

@@
 of mutations. A number of recent studies have used patterns of linkage
-disequilibrium between selected variants to test for selective interference and
-epistatic interactions, with some disagreement over the interpretation of
-observations from data. Interpreting this data is hindered by the relative lack
-of analytic or even numerical expectations for patterns of variation between
-pairs of loci under the combined effects of selection, dominance, epistasis,
-and demography. Here, I develop a numerical approach to compute the expected
-two-locus sampling distribution under diploid selection with arbitrary
+disequilibrium (LD) between selected variants to test for selective
+interference and epistatic interactions, with some disagreement over
+interpreting observations from data. Interpretation is hindered by the relative
+lack of analytic or even numerical expectations for patterns of variation
+between pairs of loci under the combined effects of selection, dominance,
+epistasis, and demography. Here, I develop a numerical approach to compute the
+expected two-locus sampling distribution under diploid selection with arbitrary
 epistasis and dominance, recombination, and variable population size. I use
@@
 
-Non-additive effects, including dominance (i.e., interactions *within* loci)
+Non-additive effects, including dominance (i.e., interactions *within* a locus)
 and epistasis (interactions *between* loci), further complicate our
@@
 disease mutations [@Clark1998-kq]. On the other hand, epistasis differentially
-impacts the deleterious load for asexually and sexually reproducing organisms
+impacts the deleterious load in asexually and sexually reproducing organisms
 [@Kimura1966-gc; @Kondrashov1995-va], has been invoked as an explanation for
@@
 improve dramatically, promising greater resolution of the fitness landscape in
-such model systems which can be compared to evolutionary theory
+such model systems that can be compared to evolutionary theory
 [@Otwinowski2018-zk].
@@
 we must turn to population genetic approaches to infer selective interactions
-between observed mutations. Motivated by theory that linked negatively selected
-mutations will display negative LD due to interference [@Hill1966-gv], and that
-epistasis will drive expected LD away from zero, a number of recent studies
-have used patterns of LD within classes of putatively selected variants to
-infer modes of selective interactions. In a notable study from
-@Callahan2011-ac, pairs of nonsynonymous mutations were found to cluster more
-than expected along lineages in the *Drosophilid* species complex, and that
-those clustered mutations tended to preserve the charge of the protein and were
-in positive LD compared to pairs involving synonymous mutations, suggesting
-that compensatory nonsynonymous variants are tolerated and maintained. More
-recently, @Taverner2020-lk replicated this finding across a diverse set of
-genera, showing that such epistatic interactions are important for protein
-evolution.
+between observed segregating polymorphisms. Motivated by theory that linked
+negatively selected mutations will display negative LD due to interference
+[@Hill1966-gv], and that epistasis will drive expected LD away from zero, a
+number of recent studies have used patterns of LD within classes of putatively
+selected variants to infer modes of selective interactions. In a notable study
+from @Callahan2011-ac, pairs of nonsynonymous mutations were found to cluster
+more than expected along lineages in the *Drosophilid* species complex, and
+that those clustered mutations tended to preserve the charge of the protein and
+were in positive LD compared to pairs involving synonymous mutations,
+suggesting that compensatory nonsynonymous variants are tolerated and
+maintained. More recently, @Taverner2020-lk replicated this finding across a
+diverse set of genera, showing that such epistatic interactions are important
+for protein evolution.
 
@@
 under more complex models, we rely on expensive forward simulations. Such
-simulations can help build intuition and test inference methods, but do not
-efficiently provide expectations for quantities of interest across the range of
-relevant parameters. Analytical and numerical methods for expected haplotype
-frequencies and LD under general selective interaction models are thus crucial
-for interpreting patterns of variation observed in data.
+simulations can help build intuition and be used to test inference methods, but
+they do not efficiently provide expectations for quantities of interest across
+the range of relevant parameters. Analytical and numerical methods for expected
+haplotype frequencies and LD under general selective interaction models are
+thus crucial for interpreting patterns of variation observed in data.
 
@@
 each possible haplotype frequency configuration in a sample of size $n$. In
-many cases it is simpler to visualize summaries (such as $\E[D]$) or slices of
-the distribution (Figures \@ref(fig:fig1) and \@ref(fig:hudson) ) instead of
-the full three-dimensional distribution. Here, I focus on low-order LD
-statistics, such as $\E[D]$ and $\E[D^2]$, and their decay with recombination
-distance as these are statistics that are commonly used to test for
-interactions between loci. Instead of unnormalized $\E[D^2]$ and $\E[D]$, I
-consider expectations for $\sigma_d^2 = \E[D^2]/\E[p(1-p)q(1-q)]$ and
-$\sigma_d^1 = \E[D]/\E[p(1-p)q(1-q)]$, where $p$ and $q$ are the derived allele
-frequencies at the left ($A/a$) and right ($B/b$) loci, respectively. These
-normalized statistics have two benefits: first, the mutation rate cancels so
-that expectations are robust to assumptions about the per-base mutation rate,
-and second, analytic expectations for these quantities are available under
-neutrality at constant population size for comparison [@Ohta1971-yd].
+many cases it is simpler to visualize summaries such as the expectation or
+variance of $D$ or slices of the distribution (e.g., Figures \@ref(fig:fig1),
+\@ref(fig:fig2) and \@ref(fig:hudson)) instead of the full three-dimensional
+distribution. Here I focus on low-order LD statistics including $\E[D]$ and
+$\E[D^2]$ and their decay with recombination distance, as these are statistics
+that are commonly used to test for interactions between loci. Instead of
+unnormalized $\E[D^2]$ and $\E[D]$ I consider expectations for $\sigma_d^2 =
+\E[D^2]/\E[p(1-p)q(1-q)]$ and $\sigma_d^1 = \E[D]/\E[p(1-p)q(1-q)]$, where $p$
+and $q$ are the derived allele frequencies at the left ($A/a$) and right
+($B/b$) loci, respectively. These normalized statistics have two benefits:
+first, the mutation rate cancels so that expectations are robust to assumptions
+about the per-base mutation rate, and second, analytic expectations for these
+quantities are available under neutrality at constant population size for
+comparison [@Ohta1971-yd].
 
@@
 (\(\epsilon > 0\)) increases it compared to expectations under a multiplicative
-fitness model ($\epsilon=0$). Epistasis also pushes signed LD ($\sigma_d^1$)
-away from zero, with synergistic epistasis decreasing LD for negatively
-selected mutations and antagonistic epistasis increasing LD. The deviation of
+fitness model ($\epsilon=0$). (See Methods for the definition of the
+epistasis model.) Epistasis also pushes signed LD ($\sigma_d^1$) away from
+zero, with synergistic epistasis decreasing LD for negatively selected
+mutations and antagonistic epistasis increasing LD. The deviation of
 $\sigma_d^1$ from zero under epistasis can extend for long recombination
@@
 
-The effects of epistasis between variants on LD have been widely studied.
-However, the combined effect of non-additive selection within a locus and
-interference has received less attention. @Garcia2020-zn used large-scale
-forward simulations to explore the effect of dominance on patterns of LD,
-showing that LD depends on both the magnitude of the selection and dominance
-coefficients in a nonlinear way, but note that expensive large-scale
-simulations are needed to explore these effects as no closed form approach to
-computing those expectations exist. The method presented here provides such a
-closed form approach, so that expectations of signed LD and other two-locus
-statistics can be computed more efficiently and accurately.
+The effects of epistasis on LD have been widely studied. However, the combined
+effect of non-additive selection within a locus and interference has received
+less attention. @Garcia2020-zn used large-scale forward simulations to explore
+the effect of dominance on patterns of LD, showing that LD depends on both the
+magnitude of the selection and dominance coefficients in a nonlinear way, but
+note that expensive large-scale simulations are needed to explore these effects
+as no closed form approach to computing those expectations exist. The method
+presented here provides such a closed form approach, so that expectations of
+signed LD and other two-locus statistics can be computed more efficiently and
+accurately.
 
@@
 
-While dominance and epistasis are well-characterized models of selection, a
-large family of selection models can be specified by assigning unique
-selection coefficients to each possible diploid pairs of haplotypes. If we assume
-the diploid genotype that is homozygous for the ancestral alleles ($ab/ab$) has
-fitness 1, then there are nine other possible diploid two-locus genotypes that
-could be given unique fitnesses (Table \@ref(tab:selmodels)), noting that
-$AB/ab$ and $Ab/aB$ genotypes can have differing fitness effects.
-
-The case with $s_{AB/ab}\not=s_{Ab/aB}$ can arise in a scenario where a
+Beyond our standard models of epistasis and dominance, a large family of
+selection models can be specified by assigning unique fitness effects to each
+possible diploid pair of haplotypes. If we assume the diploid genotype that is
+homozygous for the ancestral alleles ($ab/ab$) has fitness 1, then there are
+nine other possible diploid two-locus genotypes that could be given unique
+fitnesses (Table \@ref(tab:selmodels)), noting that $AB/ab$ and $Ab/aB$
+genotypes can have differing selection coefficients.
+
+The case with $s_{AB/ab} \not= s_{Ab/aB}$ can arise in a scenario where a
 mutation at either locus within a haplotype impacts some functional region or
@@
 instantaneous expansion and another with a bottleneck followed by recovery. I
-also consider two demographic histories inferred using genome-wide genealogy
-reconstruction [@Speidel2019-nj] applied to the
+also consider two demographic histories inferred using genome-wide gene
+genealogy reconstruction [@Speidel2019-nj] applied to the
 @1000_Genomes_Project_Consortium2015-zq dataset, and focus on size histories
@@
 to be different from pairs of synonymous variants, but LD between
-loss-of-function variants is substantially more negative. Hill-Robertson
-effects under additive selection are expected to be strongest for slightly to
-moderately deleterious variants (with $s\sim 1/N_e$), as strongly deleterious
-mutations are not expected to interfere with one another [@McVean2000-ox].
-However, loss-of-function mutations do not typically fall within this regime,
-and inference of the distribution of fitness effects (DFE) for new
-loss-of-function variants shows that a large majority are strongly deleterious
-(Supporting Information). Instead, negative synergistic epistasis between
-strongly deleterious mutations does produce large negative deviations of mean
-LD. Strongly deleterious *recessive* mutations can also produce this pattern,
+loss-of-function variants is more negative. Hill-Robertson effects under
+additive selection are expected to be strongest for slightly to moderately
+deleterious variants (with $s\sim 1/N_e$), as strongly deleterious mutations
+are not expected to interfere with one another [@McVean2000-ox]. However,
+loss-of-function mutations do not typically fall within this regime, and
+inference of the distribution of fitness effects (DFE) for new loss-of-function
+variants shows that a large majority are strongly deleterious (Supporting
+Information). Instead, negative synergistic epistasis between strongly
+deleterious mutations does produce large negative deviations of mean LD.
+Strongly deleterious *recessive* mutations can also produce this pattern,
 making it difficult to distinguish these effects with certainty. This
 difficulty is compounded by the large measurement noise for $\E[D]$, especially
-for loss-of-function variants for which only a few hundred within-gene pair in
-human populations.
+for loss-of-function variants for which only a few hundred within-gene pairs
+exist in human population data considered.
 
@@
 on the functional domain as a unit instead of at loci within the domain
-individually (such as gene-based dominance). In this scenario, double
-heterozygotes have different fitnesses depending on whether the mutations are
-found on same haplotype or on different haplotypes.
+individually (such as under a model of gene-based dominance). In this scenario,
+double heterozygotes have different fitnesses depending on whether the
+mutations are found on same haplotype or on different haplotypes.
 
@@
 recombination distances. Nor do we find increased LD outside of domains.
-Rather, for pairs of mutations outside of domains but at the same distance as
-those within the same domain, missense variants have considerably lower LD than
-synonymous variants. This difference between synonymous and missense pairs of
-variants largely disappears for SNPs separated by more than a few hundred or a
-thousand base pairs (Figure \@ref(fig:lddecay)). This suggests that
+Rather, for pairs of mutations outside of domains but matched to the same
+distance as those within the same domain, missense variants have considerably
+lower LD than synonymous variants. This difference between synonymous and
+missense pairs of variants largely disappears for SNPs separated by more than a
+few hundred base pairs (Figure \@ref(fig:lddecay)). This suggests that
 Hill-Robertson interference is the primary mode of interaction between
@@
 population lives in some geography that will necessarily induce some structure.
-However, neutral evolution alone cannot create this effect. Under standard
-population genetic assumptions of constant mutation and recombination rates and
-independent mutation events, migration, admixture, and spatial structure do not
-result in nonzero $\E[D]$. Rather, for a set of mutations with *given*
-differences in allele frequencies between source populations, $D$ can be
-nonzero in an admixed population between those particular mutations even when
-$D$ is zero in the source populations [@Cavalli-Sforza1971-jv]. But this result
-relies on conditioning on directional differences in allele frequencies between
-populations. Expected mean allele frequency difference in the source
-populations is zero ($\E[(p_1-p_2)(q_1-q_2)] = 0$), as differences are equally
-likely to be positive or negative, and resulting $\E[D]$ is still zero when
-taking genome-wide averages. Instead, @Sohail2017-zq and @Sandler2020-vk used
-forward simulations under multi-population or explicitly spatial models and a
-multiplicative fitness function to explore the joint effect of structure,
-assortative mating, and interference, scenarios that can lead to nonzero LD.
-@Sandler2020-vk find positive LD among neutral mutations in their simulation,
-although it extends over much larger distances than observed in the data.
-Instead, positive LD in data between synonymous mutations decays to zero within
-just a few hundred bases (see Figures \@ref(fig:lddecay) and
-\@ref(fig:lddecaydomain), and Figure 5 from @Sandler2020-vk).
+However, neutral evolution in structured populations alone cannot create this
+effect. Under standard population genetic assumptions of constant mutation and
+recombination rates and independent mutation events, migration, admixture, and
+spatial structure do not result in nonzero $\E[D]$. Rather, for a set of
+mutations with *given* differences in allele frequencies between source
+populations, $D$ can be nonzero in an admixed population between those
+particular mutations even when $D$ is zero in the source populations
+[@Cavalli-Sforza1971-jv]. But this result relies on conditioning on directional
+differences in allele frequencies between populations. Expected mean allele
+frequency difference in the source populations is zero ($\E[(p_1-p_2)(q_1-q_2)]
+= 0$), as differences are equally likely to be positive or negative, and
+resulting $\E[D]$ is still zero when taking genome-wide averages. Instead,
+@Sohail2017-zq and @Sandler2020-vk used forward simulations under
+multi-population or explicitly spatial models and a multiplicative fitness
+function to explore the joint effect of structure, assortative mating, and
+interference, scenarios that can lead to nonzero LD. @Sandler2020-vk find
+positive LD among neutral mutations in a simulation with admixture and linked
+selected mutations, although it extends over much larger distances than
+observed in the data. Instead, positive LD in data between synonymous mutations
+decays to zero within just a few hundred bases (see Figures \@ref(fig:lddecay)
+and \@ref(fig:lddecaydomain), and Figure 5 from @Sandler2020-vk).
 
@@
 the dynamics of allelic states of a sample of size of $n$ from the full
-population. We assume $n \ll N_e$, and $r$ and $s$ are $O(1/Ne)$ so that
+population. We assume $n \ll N_e$, and $r$ and $s$ are $O(1/N_e)$ so that
 multiple coalescence, recombination, or selective events within the $n$
@@
 system is closed and can be solved exactly. However, for nonzero recombination
-or selection, the entries of $\Psi_n$ rely on the slightly sampling
+or selection, the entries of $\Psi_n$ rely on the slightly larger sampling
 distributions with sample sizes $n+1$ (for recombination and additive
@@
 The jackknife approximation, which approximates an entry $\Psi_{n+l}(i,j,k)$
-using nearby entries in $\Psi_n$, is more accurate for larger sample sizes
-(Figure \@ref(fig:jackknife)), creating a tension between efficiency and
-accuracy: larger sample sizes result in more accurate solutions, as error in
-the jackknife is diminished, but computational complexity also grows rapidly in
-the number of entries of $\Psi_n$, which is $O(n^3)$. Accuracy and runtime are
-shown in Figure \@ref(fig:jackknife). In the results presented in this paper,
-sample sizes between $n=30$ and $n=80$ are used.
-
-Derivations for the drift, recombination, and mutation operators and the
-jackknife moment-closure approximation can be found in section S1.3 of
-@Ragsdale2019-nt, and I repeat the main results in the Supporting Information
-of this paper.
+using nearby entries in $\Psi_n$, is more accurate for larger sample sizes,
+creating a tension between efficiency and accuracy: larger sample sizes result
+in more accurate solutions, as error in the jackknife is diminished, but
+computational complexity also grows rapidly in the number of entries of
+$\Psi_n$, which is $O(n^3)$. Accuracy and runtime are shown in Figure
+\@ref(fig:jackknife). In the results presented in this paper, sample sizes
+between $n=30$ and $n=80$ are used. Derivations for the drift, recombination,
+and mutation operators and the jackknife moment-closure approximation can be
+found in section S1.3 of @Ragsdale2019-nt, and I repeat the main results in the
+Supporting Information of this paper.
 
@@
 To obtain the recursion equation under selection we consider drawing $n$
-lineages from generation $t$, which has an expected sampling distribution
-of haplotype counts given by $\Psi_n^t$. However, assuming $s\leq0$ for each
+lineages from generation $t$, which has an expected sampling distribution of
+haplotype counts given by $\Psi_n^t$. However, assuming $s\leq0$ for each
 derived haplotype, each of those sampled lineages has probability of being
-rejected equal to the relative fitness of its haplotype state. If a lineage is
-rejected, a replacement is drawn from the full population. Under the assumption
-that $ns_{\cdot\cdot} \ll 1$, the probability that more than one selection
-event occurs in any given generation is negligibly small, so that the case of
-multiple simultaneous rejections can be ignored. Then $Psi_n^{t+1}$ relies only
-on $\Psi_n^t$ and $\Psi_{n+1}^t$ for additive selection. The full selection
-operator $\mathcal{S}$ for additive selection is given in the Supporting
-Information.
+rejected equal to the selection coefficient assigned to its haplotype state. If
+a lineage is rejected, a replacement is drawn from the full population. Under
+the assumption that $ns_{\cdot\cdot} \ll 1$, the probability that more than one
+selection event occurs in any given generation is negligibly small, so that the
+case of multiple simultaneous rejections can be ignored. Then $\Psi_n^{t+1}$
+relies only on $\Psi_n^t$ and $\Psi_{n+1}^t$ for additive selection. The full
+selection operator $\mathcal{S}$ for additive selection is given in the
+Supporting Information.
 
@@
 second haplotype is $ab$, $Ab$, $aB$, or $AB$. We can therefore assign a
-selection coefficient to each possible diploid configuration, $s_{AB/AB}$,
-$s_{AB/Ab}$, and so on. Assuming that the doubly homozygous ancestral $ab/ab$
+selection coefficient to each possible diploid configuration, $s_{Ab/ab}$,
+$s_{Ab/Ab}$, and so on. Assuming that the doubly homozygous ancestral $ab/ab$
 genotype has relative fitness 1, this gives nine possible unique selection
@@
 the maximum likelihood estimate for each allele frequency $f_i = n_i / n$, but
-the maximum likelihood estimate will be noticeably biased for small to moderate
-sample sizes. Other low-order two-locus statistics can be computed using the
-same approach, as implemented in *moments* following @Ragsdale2020-nz, which
-can be compared across sample sizes and between estimates from phased or
-unphased data. In this paper, I focus on $\sigma_d^2 = \E[D^2] /
-\E[p(1-p)q(1-q)]$ and $\sigma_d^1 = \E[D] / \E[p(1-p)q(1-q)]$, averaged over
-pairs of variants at all frequencies. However, allele-frequency conditioned
-statistics (such as keeping only loci below some frequency threshold as in
-@Good2020-wj) can be considered using this same approach.
+the maximum likelihood-based estimate will be noticeably biased for small to
+moderate sample sizes. Other low-order two-locus statistics can be computed
+using the same approach, as implemented in *moments* following
+@Ragsdale2020-nz, which can be compared across sample sizes and between
+estimates from phased or unphased data. In this paper, I focus on $\sigma_d^2 =
+\E[D^2] / \E[p(1-p)q(1-q)]$ and $\sigma_d^1 = \E[D] / \E[p(1-p)q(1-q)]$,
+averaged over pairs of variants at all frequencies. However, allele-frequency
+conditioned statistics (such as keeping only loci below some frequency
+threshold as in @Good2020-wj) can be considered using this same approach.
 
@@
 models, and Figure \@ref(fig:fig3) plots the trajectory of $\sigma_d^1$ over
-time. The selection strength at both loci was fixed at $\gamma=2$ for all
+time. The selection strength at both loci was fixed at $\gamma=-2$ for all
 scenarios, and I considered additive selection without epistasis for fully
@@
 recommended in the Relate tutorial. Relate returns estimates of coalescence
-rates given time bins, and population sizes are estimated as their inverses.
-Estimates using Relate for population sizes in the very recent past ($<3,000$
-years, or $\approx 100$ generations) diverged, so I truncated the history over
-this time period and assumed a constant size from the most recent non-diverged
-bin.
+rates within specified time bins, and population sizes are estimated as their
+inverses. Estimates using Relate for population sizes in the very recent past
+($<3,000$ years, or $\approx 100$ generations) diverged, so I truncated the
+history over this time period and assumed a constant size from the most recent
+non-diverged bin.
 
@@
 further subset to samples within each non-admixed population in the African,
-European, and East Asian continental groups. Signed LD is sensitive to
-ancestral state identification, so I only kept sites for which ancestral
-alleles were estimated with high confidence in both the VCF info field and the
-Thousand Genomes human ancestor reconstructed from a phylogeny of six primates.
+European, and East Asian continental groups (Table \@ref(tab:1kgpops)). Signed
+LD is sensitive to ancestral state misidentification, so I only kept sites for
+which ancestral alleles were estimated with high confidence in both the VCF
+info field and the Thousand Genomes human ancestor reconstructed from a
+phylogeny of six primates.
 
@@
 the same protein-coding genes. I used a dataset of annotated protein domains
-mapped to hg19 compiled by @Stanek2020-pa to determine if a given mutation
-falls within a domain or not.
+mapped to the hg19 human reference build compiled by @Stanek2020-pa to
+determine if a given mutation falls within a domain or not.
 
@@
 
-I am grateful to Alex Diaz-Papkovich, Eric Friedlander, Simon Gravel, Mashaal
-Sohail, Matthias Steinrücken, and Kevin Thornton for helpful discussions and
-feedback on earlier versions of this manuscript that improved this work.
+I thank Alex Diaz-Papkovich, Eric Friedlander, Simon Gravel, Mashaal Sohail,
+Matthias Steinrücken, and Kevin Thornton for helpful discussions and feedback
+on earlier versions of this manuscript.
 

### 2021-03-25 SupportingInformation.Rmd
edits throughout

--- 2021-03-07 SupportingInformation.Rmd
@@
 bibliography: "references.bib"
-title: Supporting Information for 
+title: Supporting Information for "Can we distinguish modes of selective interactions using linkage disequilibrium?"
 author:
-  - Aaron P. Ragsdale^[aaronpeaceragsdale@gmail.com]
+  - Aaron P. Ragsdale^[\url{aaronpeaceragsdale@gmail.com}]
 date:
@@
 The two-locus diffusion equation with additive selection was first described by
-@Kimura1955-qe studied extensively in the 1960s and 70s, including by
-@Hill1966-gv and @Ohta1969-ie. The continuous distribution $\psi(x_1, x_2,
-x_3)$ of haplotype frequencies in a population, where $x_1$ is the frequency of
-$AB$, $x_2$ of $Ab$, and $x_3$ of $aB$, is governed by the multi-dimensional
+@Kimura1955-qe and studied extensively in the 1960s and 70s [e.g.,
+@Hill1966-gv; @Ohta1969-ie]. The continuous distribution $\psi(x_1, x_2, x_3)$
+of haplotype frequencies in a population, where $x_1$ is the frequency of $AB$,
+$x_2$ of $Ab$, and $x_3$ of $aB$, is governed by the multi-dimensional
 Fokker-Planck equation:
@@
 (x_1+x_2)(x_1+x_3) = x_1 x_4 - x_2 x_3,\] $\gamma_A$ and $\gamma_B$ are the
-scaled selection coefficients at the left and right locus, and $\rho$ is the
-scaled recombination rate between the two loci. Time $\tau$ is measured in
-$2N_e$ generations, and $\nu(\tau)$ is the population size relative
+(additive) scaled selection coefficients at the left and right locus, and
+$\rho$ is the scaled recombination rate between the two loci. Time $\tau$ is
+measured in $2N_e$ generations, and $\nu(\tau)$ is the population size relative
 to the ancestral size or some reference size at time $\tau$.
@@
 + \frac{\theta}{2}\mathcal{U}_{n}\Psi_n
-+ \mathcal{S}_{n, \mathbf{\gamma}, \mathbf{h}}\Psi_n.
++ \mathcal{S}_{n, \boldsymbol{\gamma}, \mathbf{h}}\Psi_n.
 \end{equation}.
 
-Alternatively, we arrive at this same linear system of equations by considering
-tracking the expected sampling distribution over $n$ lineages within the full
-population and how that changes over time by drawing lineages from one
-generation to the next in the style of @Wright1931-wy. Both @Jouganous2017-pq,
-for the single-locus SFS, and @Ragsdale2019-nt drew this connection in detail,
-so I refer readers to that previous work for a fuller description of those
-derivations and discussion. In the next section I repeat the results for
+Alternatively, we arrive at this same linear system of equations by tracking
+the expected sampling distribution over $n$ lineages within the full population
+and how that changes over time by drawing lineages from one generation to the
+next in the style of @Wright1931-wy. Both @Jouganous2017-pq, for the
+single-locus SFS, and @Ragsdale2019-nt drew this connection in detail, so I
+refer readers to those previous papers for a fuller description of those
+derivations and discussion. In the next section I simply repeat the results for
 $\mathcal{D}$, $\mathcal{R}$, and $\mathcal{U}$, briefly describe the moment
@@
 previous generation. After drawing that extra lineage, $\Psi_n$ changes as we
-draw one of the two recombinant types (each with probability $1/2$$ instead of
+draw one of the two recombinant types (each with probability $1/2$) instead of
 the lineage that was chosen to recombine.
@@
 $\Psi_{n+2}$ as linear combinations of entries in $\Psi_n$. The general
-strategy is to assume the underlying continuous distribution $\psi(x, y, z)$
-can be approximated locally as a quadratic, and then use entries in $\Psi_n$
-that are close in frequency to a given entry in $\Psi_{n+l}$ to estimate the
-coefficients of that quadratic using the multinomial sampling formula. Then
-this quadratic local approximation to $\psi$ can be used to compute
-$\Psi_{n+l}(i, j, k)$ using Eq. \@ref(eq:multinomial). Readers should refer to
-section S1.3.5 in the Supporting material for @Ragsdale2019-nt for details.
+strategy is to assume the underlying continuous distribution $\psi(x_1, x_2,
+x_3)$ can be approximated locally as a quadratic, and then use entries in
+$\Psi_n$ that are close in frequency to a given entry in $\Psi_{n+l}$ to
+estimate the coefficients of that quadratic using the multinomial sampling
+formula. Then this quadratically local approximation to $\psi$ can be used to
+compute $\Psi_{n+l}(i, j, k)$ using Eq. \@ref(eq:multinomial). Readers should
+refer to section S1.3.5 in the Supporting material for @Ragsdale2019-nt for
+details.
 
@@
 and simplifying, we find
+
 \begin{align}
@@
 the folded SFS for synonymous, missense, and LOF mutations across all autosomal
-genes, I infer DFEs for missense and LOF mutations independently. I consider a
-few different dominance coefficients to explore the effect of the assumed
-recessivity of the two classes of mutations.
+genes, I inferred DFEs for missense and LOF mutations independently. I
+considered a few different dominance coefficients to explore the effect of the
+assumed recessivity of the two classes of mutations.
 
@@
 purges some fraction of strongly deleterious mutations which are unseen in the
-sample). Here, I fit demography and DFEs to the SFS from the Mende in Sierra
-Leone (MSL) using *moments* [@Jouganous2017-pq].
+sample). Here, I fit demography and DFEs to the folded SFS from the Mende in
+Sierra Leone (MSL) using *moments* version 1.1.0 [@Jouganous2017-pq].
 
@@
 mutation rate across autosomal genes ($uL$, where $u$ is the per-base mutation
-rate, and $L$ is the total length of the coding genome). These values were
-$(0.1442, 0.3426, 0.0256)$ for synonymous, missense, and LOF mutations,
-respectively. Roughly two thirds of new mutations in coding regions are
-expected to be missense mutations, while only 5\% of new mutations are LOF. I
-fit a demographic model to the synonymous variants, which included a population
-expansion in the deeper past and exponential growth in the recent past (Figure
-\@ref(fig:msldfe)A). Using the inferred optimal scaled mutation rate,
-$\theta=4N_euL$, I estimated $Ne\approx12,300$, and assuming an average
+rate of a given mutation class, and $L$ is the total length of the coding
+genome). These values were $(0.1442, 0.3426, 0.0256)$ for synonymous, missense,
+and LOF mutations, respectively. Roughly two thirds of new mutations in coding
+regions are expected to be missense mutations, while only 5\% of new mutations
+are LOF. I fit a demographic model to the synonymous variants, which included a
+population expansion in the deeper past and exponential growth in the recent
+past (Figure \@ref(fig:msldemogdfe)A). Using the inferred optimal scaled mutation
+rate, $\theta=4N_euL$, I estimated $Ne\approx12,300$, and assuming an average
 generation time of 29 years I converted the inferred genetic units to physical
@@
 ago, and then exponential growth over the past 20-30 thousand years, with a
-current effective size of $\sim 63,000$.
+current effective size of $\approx 63,000$.
 
@@
 the GNOMAD mutation model [@Karczewski2020-le]. I tested three values for the
-dominance coefficient $h$, $0$, $0.2$ and $0.5$. For missense mutations, $h=0$
+dominance coefficient $h$: $0$, $0.2$ and $0.5$. For missense mutations, $h=0$
 gave a poor fit to the data, and $h=0.5$ fit best among the three tested
@@
 positive LD, and LD between those pairs that are very tightly linked will not
-be broken down all that rapidly. MNMs are expected to occur over relatively short
-distances, on the order of 10s or 100s of base pairs, making them a likely culprit
-of the observed positive LD among synonymous mutations at short distances.
+be broken down all that rapidly. MNMs are expected to occur over relatively
+short distances, on the order of 10s or 100s of base pairs, making them a
+likely culprit of the observed positive LD among synonymous mutations at short
+distances.
 
-Multinucleotide mutations can be easily incorporated into the moment system with
-a simple adjustment to the mutation operator. Instead of all mutations occurring
-independently in haplotypes with mutations already segregating at the other locus,
-some fraction of new mutations could instead occur spontaneously and create a new
-pair of mutations with initial counts $n_{AB} = 1$ and $n_{ab} = n-1$.
+MNM events can be easily incorporated into the moment system with a simple
+adjustment to the mutation operator. Instead of all mutations occurring
+independently in haplotypes with mutations already segregating at the other
+locus, some fraction of new mutations could instead occur spontaneously and
+create a new pair of mutations with initial counts $n_{AB} = 1$ and $n_{ab} =
+n-1$.
 
@@
 number of values between $10^{-9}$ and $2\times 10^{-8}$. The optimization was
-insensitive to the chosen value or $r$, because the decay of positive LD occurs
-rapidly. For any plausible value of $r$, the $\sigma_d^1$ decays to zero well
-before distances between pairs have scaled recombination rates $\rho=4 N_e r
-d$, and expected statistics for $\rho \ll 1$ vary only negligibly.
+insensitive to the chosen value of $r$, because the decay of positive LD occurs
+rapidly. For any plausible value of $r$, $\sigma_d^1$ decays to zero well
+before distances between pairs have scaled recombination rates $\rho=4 N_e r d
+\sim 1$, and expected statistics for $\rho \ll 1$ vary only negligibly.
 
@@
 is important to note that this does not mean that $10-15\%$ of new mutations
-occur via MNMs, since this fraction is contitioned on two mutations occurring
+occur via MNMs, since this fraction is conditioned on two mutations occurring
 at short distances.
@@
 either averaging over relatively few observed pairs of mutations, or due to
-small sample sizes that make each individual measurement noisy estimates of the
-LD for that pair of mutations in the full population. To explore the underlying
-cause of measurement uncertainty in the @1000_Genomes_Project_Consortium2015-zq
-data, I considered larger sets of samples by combining populations that
-consistently cluster together in PCA and UMAP space and have low
-differentiation [@Diaz-Papkovich2020-ee]. I took combinations of CEU/GBR,
-CHB/CHS, CDX/KHV, and MSL/GWD. While recognizing that residual population
-structure in these population combinations could alter expected LD statistics
-compared to the respective single-population estimates, I was more interested
-in the effect that increasing the sample sizes would have on estimated
-measurement error.
+small sample sizes that make each individual measurement a noisy estimate of
+the LD for that pair of mutations in the full population. To explore the
+underlying cause of measurement uncertainty in the
+@1000_Genomes_Project_Consortium2015-zq data, I considered larger sets of
+samples by combining populations that consistently cluster together in PCA and
+UMAP space and have low differentiation [@Diaz-Papkovich2020-ee]. I took
+combinations of CEU/GBR, CHB/CHS, CDX/KHV, and MSL/GWD. While recognizing that
+residual population structure in these population combinations could alter
+expected LD statistics compared to the respective single-population estimates,
+I was more interested in the effect that increasing the sample sizes would have
+on estimated measurement error.
 
@@
     caption="DFEs inferred for missense and loss-of-function variants in MSL for varying values of $h$.
-General patterns are consistent across different chosen values of $h$, although for $h=0$ results in poorer fits for both missense and LOF variants.
+General patterns are consistent across different chosen values of $h$, although $h=0$ results in poorer fits for both missense and LOF variants.
 Columns to the right of the log-likelihood (LL) column show proportions of new mutations with $|s|$ in each given bin.") %>%
@@
 # Supplementary Figures
+
+```{r hudson, echo=F, fig.cap="(ref:fighudson)", fig.align="center"}
+knitr::include_graphics("figures/hudson_comparison.pdf")
+```
+
+(ref:fighudson) The distribution at stationarity of AB haplotype counts in a
+sample size of 30, in which we observe 10 $A$ alleles at the left locus, and 10
+$B$ alleles at the right locus.
 
@@
 the folded synonymous SFS, and DFEs were fit to missense and loss-of-function
-SFS. Shown here are DFEs fit with $h=0.5$.
+SFS. Shown here are DFEs fit with $h=0.5$. See Table \@ref(tab:msldfe) for
+inferred best-fit parameters.
 

### 2021-11-11 Manuscript.Rmd
discrete simulations

@@
   ```
+- \usepackage{lineno}
+- \linenumbers
 date: \today

### 2021-11-11 figures/fig1.py

@@ -278,3 +278,4 @@ fig.text(0.02, 0.32, "E", fontsize=8, ha="center", va="center")
 fig.text(0.52, 0.32, "F", fontsize=8, ha="center", va="center")
 
 plt.savefig(f"fig1_combo.pdf")
+plt.savefig(f"fig1_combo.png", dpi=300)

### 2021-11-11 figures/fig2.py

@@ -141,8 +141,8 @@ ax2.set_ylabel(r"$\sigma_d^1$")
 ax3.set_ylabel(r"$\sigma_d^2$")
 ax4.set_ylabel(r"$\sigma_d^1$")
 
-ax3.set_xlabel(r"$\rho$")
-ax4.set_xlabel(r"$\rho$")
+ax3.set_xlabel(r"$\rho = 4 N_e r$ (recomb. distance)")
+ax4.set_xlabel(r"$\rho = 4 N_e r$ (recomb. distance)")
 
 ax1.set_title(r"$\gamma=-1$")
 ax2.set_title(r"$\gamma=-1$")
@@ -155,4 +155,5 @@ fig.text(0.52, 0.98, "B", fontsize=8, ha="center", va="center")
 fig.text(0.02, 0.48, "C", fontsize=8, ha="center", va="center")
 fig.text(0.52, 0.48, "D", fontsize=8, ha="center", va="center")
 
-plt.savefig(f"fig2.pdf")
+#plt.savefig(f"fig2.pdf")
+plt.savefig(f"fig2.png", dpi=300)

### 2021-11-11 figures/fig5.py

@@ -316,4 +316,5 @@ fig.text(0.05, 0.38, "B", fontsize=8, ha="center", va="center")
 fig.text(0.38, 0.38, "C", fontsize=8, ha="center", va="center")
 fig.text(0.70, 0.38, "D", fontsize=8, ha="center", va="center")
 
-plt.savefig(f"fig6.pdf")
+#plt.savefig(f"fig6.pdf")
+plt.savefig(f"fig6.png", dpi=300)

### 2021-11-11 simulations/discrete/compare_simulation_to_expectation.py

@@ -1,0 +1,87 @@
+import numpy as np
+import moments.TwoLocus
+import pickle
+import matplotlib.pylab as plt
+
+
+def get_residuals(model, data):
+    """
+    Return an array of Anscombe residuals, unordered.
+    """
+    resids = moments.Inference.Anscombe_Poisson_residual(
+        model[model.mask == False], data[data.mask == False]
+    )
+    return resids[resids.mask == False].data
+
+
+def plot_residuals(resids, ax=None):
+    """
+    Plot a histogram of residuals.
+    """
+    if ax is None:
+        fig = plt.figure()
+        ax = plt.subplot(1, 1, 1)
+
+    ax.hist(resids, bins=20, density=True)
+    ax.set_xlabel("Residuals")
+    ax.set_ylabel("Density")
+
+
+def plot_pAB_comparison(model, data, nA=None, nB=None, ax=None):
+    if nA is None or nB is None:
+        raise ValueError("must pass nA and nB")
+    if ax is None:
+        fig = plt.figure()
+        ax = plt.subplot(1, 1, 1)
+
+    nAB, pAB_model = moments.TwoLocus.Util.pAB(model, nA, nB)
+    nAB, pAB_data = moments.TwoLocus.Util.pAB(data, nA, nB)
+    ax.bar(nAB - 0.2, pAB_model / pAB_model.sum(), width=0.35, label="E[pAB]")
+    ax.bar(nAB + 0.2, pAB_data / pAB_data.sum(), width=0.35, label="O[pAB]")
+
+    ax.set_title(f"$n_A={nA}, n_B={nB}$")
+    ax.set_ylabel("Probability")
+    ax.set_xlabel("$n_{AB}$")
+    ax.set_xticks(nAB)
+
+
+subplots_dim = (2, 4)
+
+
+def plot_grid(model, data, args):
+    fig = plt.figure()
+
+    # flat residuals histogram
+    ax1 = plt.subplot2grid(subplots_dim, (0, 0), colspan=2)
+    resids = get_residuals(model, data)
+    plot_residuals(resids, ax=ax1)
+
+    # nA = 1, nB = 1
+    ax2 = plt.subplot2grid(subplots_dim, (0, 2))
+    plot_pAB_comparison(model, data, nA=1, nB=1, ax=ax2)
+
+    # nA = 2, nB = 3
[…]

### 2021-11-11 simulations/discrete/run_discrete_simulation.py

@@ -1,0 +1,246 @@
+# For a given population size, selection coefficient, dominance coefficient or
+# epistasis coefficient, and sample size, run some number of
+
+
+import moments.TwoLocus
+import numpy as np
+import sys, os
+import pickle
+
+
+import argparse
+from datetime import datetime
+
+
+def eprint(*args, **kwargs):
+    print(*args, file=sys.stderr, **kwargs)
+    sys.stderr.flush()
+
+
+def current_time():
+    return " [" + datetime.strftime(datetime.now(), "%Y-%m-%d %H:%M:%S") + "]"
+
+
+def make_parser():
+    ADHF = argparse.ArgumentDefaultsHelpFormatter
+    parser = argparse.ArgumentParser("run_discrete_comparison.py", formatter_class=ADHF)
+    parser.add_argument(
+        "--population_size",
+        "-N",
+        type=int,
+        default=5000,
+        help="Diploid population size, defaults to 5,000.",
+    )
+    parser.add_argument(
+        "--selection_coefficient",
+        "-s",
+        type=float,
+        default=0,
+        help="gamma = 2 * Ne * s",
+    )
+    parser.add_argument(
+        "--epistasis_coefficient",
+        "-e",
+        type=float,
+        default=0,
+        help="The epistasis coefficient, so that s_{AB} = s_A + s_B + ?",
+    )
+    parser.add_argument(
+        "--sample_size",
+        "-n",
+        type=int,
+        default=50,
+        help="The sample size to draw each sampling generation",
+    )
+    parser.add_argument(
+        "--mutation_rate",
+        "-u",
+        type=float,
+        default=5e-4,
+    )
+    parser.add_argument(
+        "--recombination_rate",
+        "-r",
+        type=float,
+        default=0,
+    )
+    parser.add_argument(
+        "--num_replicates",
+        type=int,
+        default=10000,
+    )
+    parser.add_argument(
+        "--spacing",
+        type=int,
+        default=10,
+    )
+    parser.add_argument(
+        "--out",
+        "-o",
[…]

### 2022-02-22 Manuscript.Rmd
add numerical checks

--- 2021-03-25 Manuscript.Rmd
@@
 biblio-style: "genetics"
-title: Can we distinguish modes of selective interactions using linkage disequilibrium?
+title: Working title -- A heterogeneous landscape of selection and epistasis in protein-coding genes
 author:
@@
   \usepackage{authblk}
-  \author{Aaron P. Ragsdale \\ aaronpeaceragsdale@gmail.com}
-  \affil{Department of Human Genetics, McGill University, Montreal, Canada \\
+  \author{Aaron P. Ragsdale \\ apragsdale@wisc.edu}
+  \affil{Department of Integrative Biology, University of Wisconsin--Madison, WI \\
+         Department of Human Genetics, McGill University, Montreal, Canada \\
          National Laboratory of Genomics for Biodiversity, Irapuato, Mexico \\
@@
 Selected mutations interfere and interact with evolutionary processes at nearby
-loci, distorting allele frequency trajectories and correlations between pairs
+loci, distorting allele frequency trajectories and creating correlations between pairs
 of mutations. A number of recent studies have used patterns of linkage
@@
 Most new mutations that affect fitness are deleterious and tend to be
-eliminated from a population. The amount of time that a deleterious mutation
+eliminated from a population **Citations**. The amount of time that a deleterious mutation
 segregates depends on the strength of selection against genomes that carry it,
@@
 both beneficial and damaging, can dramatically impact patterns of variation in
-nearby linked regions. This distortion away from neutral expectations is now
+nearby linked regions **Citations**.
+This distortion away from neutral expectations is now
 empirically documented using sequencing data from an ever-growing set of study
@@
 between, protein-coding domains, suggesting an importance for antagonistic
+(or "diminishing-returns")
 epistasis or a prevalence of compensatory nonsynonymous mutations within
@@
 function is experimentally measured [@Romero2009-yi; @Bank2015-vq;
-@Puchta2016-rx; @Steinberg2016-is]. For example, using the model system of the
-TEM-1 $\beta$-lactamase gene in *E. coli*, @Bershtein2006-bk found evidence for
-synergistic epistasis, where multiple deleterious mutations had a greater
-effect on fitness than the multiplication of the observed effects of the
-individual mutations. The scale of mutation scanning experiments continues to
+@Puchta2016-rx; @Steinberg2016-is].
+For example, using the model system of the TEM-1 $\beta$-lactamase gene in
+*E. coli*, @Bershtein2006-bk found evidence for synergistic epistasis,
+in which the combined effect of multiple deleterious mutations 
+on individual fitness was larger than would be expected from multiplying
+the independent observed effects of each individual mutation.
+The scale of mutation scanning experiments continues to
 improve dramatically, promising greater resolution of the fitness landscape in
@@
 stores the expected density or observed counts of pairs of biallelic loci with
-each possible haplotype frequency configuration in a sample of size $n$. In
-many cases it is simpler to visualize summaries such as the expectation or
+each possible haplotype frequency configuration in a sample of size $n$.
+Below, we compute expectations for $\Psi_n$ under varying scenarios of
+selection and interaction between pairs of loci.
+It is not possible in this framework to include the effects of additional
+selection mutations, such as background selection due to many linked variants.
+Individual-based forward simulations are still needed for such scenarios
+(e.g., Figure \@ref(fig:fwdpy11)).
+
+In many cases it is simpler to visualize summaries such as the expectation or
 variance of $D$ or slices of the distribution (e.g., Figures \@ref(fig:fig1),
@@
 $\E[D^2]$ and their decay with recombination distance, as these are statistics
-that are commonly used to test for interactions between loci. Instead of
-unnormalized $\E[D^2]$ and $\E[D]$ I consider expectations for $\sigma_d^2 =
-\E[D^2]/\E[p(1-p)q(1-q)]$ and $\sigma_d^1 = \E[D]/\E[p(1-p)q(1-q)]$, where $p$
-and $q$ are the derived allele frequencies at the left ($A/a$) and right
-($B/b$) loci, respectively. These normalized statistics have two benefits:
+that are commonly used to test for interactions between loci.
+For pairs of biallelic loci, with alleles $A/a$ at the left locus and $B/b$ at the
+right locus, $D=f_{AB}-f_A f_B$ is the standard covariance measure of LD,
+where $f_{AB}$ is the frequency of haplotypes carrying both $A$ and $B$, and
+$f_A$ and $f_B$ are the marginal allele frequencies of $A$ and $B$.
+
+Instead of unnormalized $\E[D^2]$ and $\E[D]$ I consider expectations for
+$\sigma_d^2 = \E[D^2]/\E[f_A(1-f_A)f_B(1-f_B)]$ and
+$\sigma_d^1 = \E[D]/\E[f_A(1-f_A)f_B(1-f_B)]$.
+Such normalized statistics have two benefits:
 first, the mutation rate cancels so that expectations are robust to assumptions
@@
 
-(ref:fig1cap) \textbf{Additive selection and epistasis.} Left panels show
+(ref:fig1cap) \textbf{Additive selection and epistasis.}
+Left panels show
 expectations for the decay of $\sigma_d^2 = \E[D^2] / \E[p(1-p)q(1-q)]$ with
@@
 $\gamma=4N_es$). (C-F) Epistasis with additive selection. Synergistic epistasis
-(\(\epsilon < 0\)) decreases $\sigma_d^2$ while antagonistic epistasis
-(\(\epsilon > 0\)) increases it compared to expectations under a multiplicative
+(\(\epsilon > 0\)) decreases $\sigma_d^2$ while antagonistic epistasis
+(\(\epsilon < 0\)) increases $\sigma_d^2$
+compared to expectations under a multiplicative
 fitness model ($\epsilon=0$). (See Methods for the definition of the
@@
 variants, with sign depending on the direction of epistatic interactions
-(Figure \@ref(fig:fig1)). Synergistic epistasis results in negative LD while
-antagonistic epistasis results in positive LD, and large positive LD can occur
+(Figure \@ref(fig:fig1)).
+Synergistic epistasis (in which the effect of two mutations together is larger than
+the product of each individual mutation's effect) results in negative LD while
+antagonistic epistasis (in which the combined effect is less than the product of
+independent effects) results in positive LD, and large positive LD can occur
 even when the epistatic effect is relatively weak. Epistasis-induced LD can
@@
 population-size-scaled selection coefficients of $\gamma=-10$ show large mean
-LD that extends well beyond 100 kilobases. More strongly deleterious
-interacting mutations are expected to show large signed LD over much larger
-recombination distances.
+LD that extends to values of $\rho$ much greater than 1 (for humans, assuming
+roughly 1 cM/Mb, this is on the order of 100 kilobases or more).
+More strongly deleterious interacting mutations are expected to show large
+signed LD over much larger recombination distances.
 
@@
 the effect of dominance on patterns of LD, showing that LD depends on both the
-magnitude of the selection and dominance coefficients in a nonlinear way, but
+magnitude of the selection and dominance coefficients in a nonlinear way, but they
 note that expensive large-scale simulations are needed to explore these effects
-as no closed form approach to computing those expectations exist. The method
-presented here provides such a closed form approach, so that expectations of
-signed LD and other two-locus statistics can be computed more efficiently and
-accurately.
+as no closed form approach to computing those expectations exist.
+The numerical method presented here provides such a closed form approach, so that expectations of signed LD and other two-locus statistics can be computed more
+efficiently and accurately.
 
@@
 
-(ref:fig2cap) **Dominance effects.** (A and C) Pairs of recessive mutations
-($h<0.5$) have higher squared LD than additive mutations, while pairs of
-dominant mutations have lower expected $D^2$. (B and D) Pairs of recessive
+(ref:fig2cap) **The effects of dominance on LD.**
+(A and C) Pairs of recessive mutations
+($h<0.5$) have higher expected squared LD than additive mutations, while pairs of
+dominant mutations have lower expected $\sigma_d^2$.
+(B and D) Pairs of recessive
 deleterious mutations can have large negative LD, although this correlation
@@
 strength of selection and dominance, particularly when $s$ is $O(1/N)$. Note
-that this setup does not include epistatic effects (i.e., $\epsilon=0$).
+that the expected values shown here do not include epistatic effects
+(i.e., $\epsilon=0$).
 
@@
 early work on the multi-locus diffusion approximation [@Kimura1955-qe;
-@Hill1968-vu; @Ohta1969-ie; @Ohta1971-yd]. This includes the result that
+@Hill1968-vu; @Ohta1969-ie; @Ohta1971-yd].
+This includes the result that
 genome-wide averages of signed LD are expected to be zero under neutrality.
@@
 
-Analytic progress beyond these results has come haltingly. In the
-1980s, recursions were developed to compute the two-locus sampling distribution
-under neutrality, that is, the probability of observing given counts of
-two-locus haplotypes in a sample of size $n$ [@Golding1984-pu]. This approach
-would continue to be developed and later form the foundation for the inference
-of local recombination rates from population genetic data [@Hudson2001-sg;
-@McVean2004-gj].
+Analytic progress beyond these results has come haltingly. In the 1980s,
+recursions were developed to compute the two-locus sampling distribution under
+neutrality, that is, the probability of observing given counts of two-locus
+haplotypes in a sample of size $n$ [@Golding1984-pu]. This approach would
+continue to be developed and later form the foundation for the inference of
+local recombination rates from population genetic data [@Hudson2001-sg;
+@McVean2004-gj]. More recently, @Song2007-qk computed $\E[r^2]$ using a
+diffusion approximation approach, although their solution involves the
+summation of infinitely many terms and is restricted to neutrality and
+steady-state demography.
 

### 2022-02-22 simulations/discrete/compare_simulation_to_expectation.py

@@ -1,7 +1,48 @@
 import numpy as np
+import argparse, sys
 import moments.TwoLocus
 import pickle
 import matplotlib.pylab as plt
+
+
+def make_parser():
+    ADHF = argparse.ArgumentDefaultsHelpFormatter
+    parser = argparse.ArgumentParser("compare_simulation_to_expectation.py", formatter_class=ADHF)
+    parser.add_argument(
+        "--population_size",
+        "-N",
+        type=int,
+        default=5000,
+        help="Diploid population size, defaults to 5,000.",
+    )
+    parser.add_argument(
+        "--selection_coefficient",
+        "-s",
+        type=float,
+        default=0,
+        help="gamma = -2 * Ne * s. Note that we set the abs value",
+    )
+    parser.add_argument(
+        "--epistasis_coefficient",
+        "-e",
+        type=float,
+        default=0,
+        help="The epistasis coefficient, so that s_{AB} = s_A + s_B + ?",
+    )
+    parser.add_argument(
+        "--sample_size",
+        "-n",
+        type=int,
+        default=50,
+        help="The sample size to draw each sampling generation",
+    )
+    parser.add_argument(
+        "--recombination_rate",
+        "-r",
+        type=float,
+        default=0,
+    )
+    return parser
 
 
 def get_residuals(model, data):
@@ -27,7 +68,7 @@ def plot_residuals(resids, ax=None):
     ax.set_ylabel("Density")
 
 
-def plot_pAB_comparison(model, data, nA=None, nB=None, ax=None):
+def plot_pAB_comparison(model, data, nA=None, nB=None, ax=None, legend=False):
     if nA is None or nB is None:
         raise ValueError("must pass nA and nB")
     if ax is None:
@@ -35,12 +76,16 @@ def plot_pAB_comparison(model, data, nA=None, nB=None, ax=None):
         ax = plt.subplot(1, 1, 1)
 
     nAB, pAB_model = moments.TwoLocus.Util.pAB(model, nA, nB)
+    # variance equals expectation, assuming poisson
+    pAB_error = np.sqrt(pAB_model) * 1.96
[…]

### 2022-02-22 simulations/discrete/run_discrete_simulation.py

@@ -23,7 +23,7 @@ def current_time():
 
 def make_parser():
     ADHF = argparse.ArgumentDefaultsHelpFormatter
-    parser = argparse.ArgumentParser("run_discrete_comparison.py", formatter_class=ADHF)
+    parser = argparse.ArgumentParser("run_discrete_simulation.py", formatter_class=ADHF)
     parser.add_argument(
         "--population_size",
         "-N",

### 2022-02-22 simulations/fwdpy11/BGS_effects_on_LD.py

@@ -14,6 +14,9 @@ from collections import defaultdict
 import copy
 import pickle
 import moments.LD
+import matplotlib.pylab as plt
+import msprime, tskit
+import gzip
 
 
 def eprint(*args, **kwargs):
@@ -34,19 +37,19 @@ def make_parser():
         "--population_size",
         "-N",
         type=int,
-        default=5000,
+        default=1000,
         help="Diploid population size, defaults to 5,000.",
     )
     optional.add_argument(
         "--recombination_rate",
         type=float,
-        default=2e-8,
+        default=1e-8,
         help="Per-base recombination rate, defaults to 2e-8.",
     )
     optional.add_argument(
         "--mutation_rate",
         type=float,
-        default=2e-8,
+        default=1e-8,
         help="Per-base mutation rate, defaults to 2e-8.",
     )
     optional.add_argument(
@@ -69,13 +72,6 @@ def make_parser():
         help="Selection coefficient for selected variants, defaults to 0.001.",
     )
     optional.add_argument(
-        "--epistasis_coefficient",
-        "-e",
-        type=float,
-        default=0,
-        help="Epistasis coefficient, sAB = (sA + sB) * (1 + e), defaults to 0.",
-    )
-    optional.add_argument(
         "--dominance_coefficient",
         "-d",
         type=float,
@@ -86,19 +82,19 @@ def make_parser():
         "--sample_size",
         "-n",
         type=int,
-        default=100,
+        default=25,
         help="The sample size for sampling the genotype matrix.",
     )
     optional.add_argument(
         "--sample_spacing",
         type=int,
-        default=50,
+        default=100,
         help="Spacing for how often to preserve ancient samples.",
     )
     optional.add_argument(
         "--simulation_length",
         type=int,
-        default=1,
+        default=100,
         help="Recorded simulation time, in units of 2N generations, defaults to 1.",
     )
     optional.add_argument(
@@ -123,13 +119,12 @@ def setup_simulation(args):
             L,
[…]

### 2022-03-23 analysis/compute_LD_unphased_between_domains_freq_matching.py
working on revisions

@@ -1,0 +1,906 @@
+import numpy as np
+import itertools
+import gzip
+import moments.LD
+import sys
+import pickle
+from collections import defaultdict
+
+
+lof_annotations = [
+    "frameshift_variant",
+    "splice_acceptor_variant",
+    "splice_donor_variant",
+    "start_lost",
+    "stop_gained",
+    "stop_lost",
+    "transcript_ablation",
+]
+
+import tarfile
+
+try:
+    print("already loaded ancestral seqs", len(ancestral_seqs))
+except NameError:
+    ancestral_seqs = {}
+    print("reading ancestral sequences")
+
+    tar = tarfile.open(
+        "/home/aaronragsdale/Data/1000G/human_ancestor/human_ancestor_GRCh37_e59.tar.bz2",
+        mode="r:bz2",
+    )
+    for member in tar.getmembers():
+        if member.isfile() is False:
+            # print(member)
+            continue
+        if member.get_info()["name"].endswith("fa"):
+            f = tar.extractfile(member)
+            ## can't get SeqIO.parse to read in the extracted file
+            lines = f.readlines()
+            chrom = lines[0].decode().split(":")[2]
+            print("found chrom", chrom)
+            ancestral_seqs[chrom] = "".join([l.decode().strip() for l in lines[1:]])
+            # for record in SeqIO.parse(x, "fasta"):
+            #    chrom = record.id.split(":")[2]
+            #    print(chrom)
+            #    ancestral_seqs[chrom] = record.seq
+    tar.close()
+
+
+def get_annotations(annot_file):
+    annots = {}  # pos: alt: (gene, csq)
+    for line in open(annot_file, "r"):
+        chrom, pos, ref, alt, gene, csq = line.split()
+        annots.setdefault(pos, {})
+        annots[pos][alt] = (gene, csq)
+    return annots
+
+
+def flip_gts(gts):
+    gts_out = []
+    for gt in gts:
+        if gt == "0|0":
+            gts_out.append("1|1")
+        elif gt == "0|1":
+            gts_out.append("1|0")
+        elif gt == "1|0":
+            gts_out.append("0|1")
+        elif gt == "1|1":
+            gts_out.append("0|0")
+        else:
+            raise ValueError("unexpected genotype", gt)
+    return gts_out
+
+
+def to_genotypes(gts):
+    g = np.zeros(len(gts))
+    for ii, gg in enumerate(gts):
[…]

### 2022-03-23 analysis/compute_LD_unphased_domains_freq_matching.py

@@ -1,0 +1,848 @@
+import numpy as np
+import itertools
+import gzip
+import moments.LD
+import sys
+import pickle
+from collections import defaultdict
+
+
+lof_annotations = [
+    "frameshift_variant",
+    "splice_acceptor_variant",
+    "splice_donor_variant",
+    "start_lost",
+    "stop_gained",
+    "stop_lost",
+    "transcript_ablation",
+]
+
+import tarfile
+
+try:
+    print("already loaded ancestral seqs", len(ancestral_seqs))
+except NameError:
+    ancestral_seqs = {}
+    print("reading ancestral sequences")
+
+    tar = tarfile.open(
+        "/home/aaronragsdale/Data/1000G/human_ancestor/human_ancestor_GRCh37_e59.tar.bz2",
+        mode="r:bz2",
+    )
+    for member in tar.getmembers():
+        if member.isfile() is False:
+            # print(member)
+            continue
+        if member.get_info()["name"].endswith("fa"):
+            f = tar.extractfile(member)
+            ## can't get SeqIO.parse to read in the extracted file
+            lines = f.readlines()
+            chrom = lines[0].decode().split(":")[2]
+            print("found chrom", chrom)
+            ancestral_seqs[chrom] = "".join([l.decode().strip() for l in lines[1:]])
+            # for record in SeqIO.parse(x, "fasta"):
+            #    chrom = record.id.split(":")[2]
+            #    print(chrom)
+            #    ancestral_seqs[chrom] = record.seq
+    tar.close()
+
+
+def get_annotations(annot_file):
+    annots = {}  # pos: alt: (gene, csq)
+    for line in open(annot_file, "r"):
+        chrom, pos, ref, alt, gene, csq = line.split()
+        annots.setdefault(pos, {})
+        annots[pos][alt] = (gene, csq)
+    return annots
+
+
+def flip_gts(gts):
+    gts_out = []
+    for gt in gts:
+        if gt == "0|0":
+            gts_out.append("1|1")
+        elif gt == "0|1":
+            gts_out.append("1|0")
+        elif gt == "1|0":
+            gts_out.append("0|1")
+        elif gt == "1|1":
+            gts_out.append("0|0")
+        else:
+            raise ValueError("unexpected genotype", gt)
+    return gts_out
+
+
+def to_genotypes(gts):
+    g = np.zeros(len(gts))
+    for ii, gg in enumerate(gts):
[…]

### 2022-03-23 analysis/compute_LD_unphased_freq_matching.py

@@ -1,0 +1,472 @@
+import numpy as np
+import gzip
+import moments.LD
+import sys
+import pickle
+
+
+lof_annotations = [
+    "frameshift_variant",
+    "splice_acceptor_variant",
+    "splice_donor_variant",
+    "start_lost",
+    "stop_gained",
+    "stop_lost",
+    "transcript_ablation",
+]
+
+import tarfile
+
+try:
+    print("already loaded ancestral seqs", len(ancestral_seqs))
+except NameError:
+    ancestral_seqs = {}
+    print("reading ancestral sequences")
+
+    tar = tarfile.open(
+        "/home/aaronragsdale/Data/1000G/human_ancestor/human_ancestor_GRCh37_e59.tar.bz2",
+        mode="r:bz2",
+    )
+    for member in tar.getmembers():
+        if member.isfile() is False:
+            # print(member)
+            continue
+        if member.get_info()["name"].endswith("fa"):
+            f = tar.extractfile(member)
+            ## can't get SeqIO.parse to read in the extracted file
+            lines = f.readlines()
+            chrom = lines[0].decode().split(":")[2]
+            print("found chrom", chrom)
+            ancestral_seqs[chrom] = "".join([l.decode().strip() for l in lines[1:]])
+            # for record in SeqIO.parse(x, "fasta"):
+            #    chrom = record.id.split(":")[2]
+            #    print(chrom)
+            #    ancestral_seqs[chrom] = record.seq
+    tar.close()
+
+
+def get_annotations(annot_file):
+    annots = {}  # pos: alt: (gene, csq)
+    for line in open(annot_file, "r"):
+        chrom, pos, ref, alt, gene, csq = line.split()
+        annots.setdefault(pos, {})
+        annots[pos][alt] = (gene, csq)
+    return annots
+
+
+def flip_gts(gts):
+    gts_out = []
+    for gt in gts:
+        if gt == "0|0":
+            gts_out.append("1|1")
+        elif gt == "0|1":
+            gts_out.append("1|0")
+        elif gt == "1|0":
+            gts_out.append("0|1")
+        elif gt == "1|1":
+            gts_out.append("0|0")
+        else:
+            raise ValueError("unexpected genotype", gt)
+    return gts_out
+
+
+def to_genotypes(gts):
+    g = np.zeros(len(gts))
+    for ii, gg in enumerate(gts):
+        g[ii] = int(gg[0]) + int(gg[2])
+    return g
+
+
[…]

### 2022-03-23 analysis/mutations_in_domains.py

@@ -1,0 +1,297 @@
+import numpy as np
+import gzip
+import moments.LD
+import sys
+import pickle
+from collections import defaultdict
+import tarfile
+
+try:
+    print("already loaded ancestral seqs", len(ancestral_seqs))
+except NameError:
+    ancestral_seqs = {}
+    print("reading ancestral sequences")
+
+    tar = tarfile.open(
+        "/home/aaronragsdale/Data/1000G/human_ancestor/human_ancestor_GRCh37_e59.tar.bz2",
+        mode="r:bz2",
+    )
+    for member in tar.getmembers():
+        if member.isfile() is False:
+            # print(member)
+            continue
+        if member.get_info()["name"].endswith("fa"):
+            f = tar.extractfile(member)
+            ## can't get SeqIO.parse to read in the extracted file
+            lines = f.readlines()
+            chrom = lines[0].decode().split(":")[2]
+            print("found chrom", chrom)
+            ancestral_seqs[chrom] = "".join([l.decode().strip() for l in lines[1:]])
+            # for record in SeqIO.parse(x, "fasta"):
+            #    chrom = record.id.split(":")[2]
+            #    print(chrom)
+            #    ancestral_seqs[chrom] = record.seq
+    tar.close()
+
+
+def get_annotations(annot_file):
+    annots = {}  # pos: alt: (gene, csq)
+    for line in open(annot_file, "r"):
+        chrom, pos, ref, alt, gene, csq = line.split()
+        annots.setdefault(pos, {})
+        annots[pos][alt] = (gene, csq)
+    return annots
+
+
+def flip_gts(gts):
+    gts_out = []
+    for gt in gts:
+        if gt == "0|0":
+            gts_out.append("1|1")
+        elif gt == "0|1":
+            gts_out.append("1|0")
+        elif gt == "1|0":
+            gts_out.append("0|1")
+        elif gt == "1|1":
+            gts_out.append("0|0")
+        else:
+            raise ValueError("unexpected genotype", gt)
+    return gts_out
+
+
+def to_genotypes(gts):
+    g = np.zeros(len(gts))
+    for ii, gg in enumerate(gts):
+        g[ii] = int(gg[0]) + int(gg[2])
+    return g
+
+
+SNPs = ["A", "C", "G", "T"]
+
+
+lof_annotations = [
+    "frameshift_variant",
+    "splice_acceptor_variant",
+    "splice_donor_variant",
[…]

### 2022-03-23 figures/data_coarse.py

@@ -1,0 +1,426 @@
+# Figure 5: LD within and between genes
+import sys
+import moments.TwoLocus
+import matplotlib.pylab as plt
+import numpy as np
+import pickle
+
+# set font sizes
+import matplotlib
+
+plt.rcParams["legend.title_fontsize"] = "xx-small"
+matplotlib.rc("xtick", labelsize=6)
+matplotlib.rc("ytick", labelsize=6)
+matplotlib.rc("axes", labelsize=7)
+matplotlib.rc("axes", titlesize=7)
+matplotlib.rc("legend", fontsize=6)
+
+from bokeh.palettes import Colorblind
+
+data = {}
+data_dom = {}
+SE = {}
+SE_dom = {}
+
+populations = {
+    "Africa": ["ESN", "GWD", "LWK", "MSL", "YRI"],
+    "Europe": ["CEU", "FIN", "GBR", "IBS", "TSI"],
+    "East Asia": ["CDX", "CHB", "CHS", "JPT", "KHV"],
+}
+
+for continent, pops in populations.items():
+    for pop in pops:
+        ld_pop = pickle.load(
+            open(f"../analysis/parsed_data_v2/{pop}.unphased.all.bp", "rb")
+        )
+        data[pop] = ld_pop
+        se_pop = pickle.load(
+            open(
+                f"../analysis/parsed_data_v2/{pop}.unphased.all.SEs.bp",
+                "rb",
+            )
+        )
+        SE[pop] = se_pop
+        ld_pop = pickle.load(
+            open(f"../analysis/parsed_data_v2/{pop}.unphased.domains.bp", "rb")
+        )
+        data_dom[pop] = ld_pop
+        se_pop = pickle.load(
+            open(
+                f"../analysis/parsed_data_v2/{pop}.unphased.domains.SEs.bp",
+                "rb",
+            )
+        )
+        SE_dom[pop] = se_pop
+
+
+ms = 3
+lw = 0.5
+markers = ["o", "s"]
+colors = Colorblind[8]
+labels = [
+    "Syn. in domains",
+    "Mis. in domains",
+    "Syn. outside",
+    "Mis. outside",
+]
+
+
+def plot_decay(pop, data, bs, ax, legend=False, ylabel=False):
+    bins = data[pop]["bin_edges"][:-1]
+    xx = np.mean(bins, axis=1)
[…]

### 2022-03-23 figures/data_domains.py

@@ -1,0 +1,247 @@
+# Figure 6: LD within and between gene domains
+import sys
+import moments.TwoLocus
+import matplotlib.pylab as plt
+import numpy as np
+import pickle
+
+# set font sizes
+import matplotlib
+
+plt.rcParams["legend.title_fontsize"] = "xx-small"
+matplotlib.rc("xtick", labelsize=6)
+matplotlib.rc("ytick", labelsize=6)
+matplotlib.rc("axes", labelsize=7)
+matplotlib.rc("axes", titlesize=7)
+matplotlib.rc("legend", fontsize=6)
+
+from bokeh.palettes import Colorblind
+
+data = {}
+
+SE = {}
+
+populations = {
+    "Africa": ["ESN", "GWD", "LWK", "MSL", "YRI"],
+    "Europe": ["CEU", "FIN", "GBR", "IBS", "TSI"],
+    "East Asia": ["CDX", "CHB", "CHS", "JPT", "KHV"],
+}
+
+for continent, pops in populations.items():
+    for pop in pops:
+        ld_pop = pickle.load(
+            open(f"../analysis/parsed_data_v2/{pop}.unphased.domains.bp", "rb")
+        )
+        data[pop] = ld_pop
+        se_pop = pickle.load(
+            open(
+                f"../analysis/parsed_data_v2/{pop}.unphased.domains.SEs.bp",
+                "rb",
+            )
+        )
+        SE[pop] = se_pop
+
+ms = 3
+lw = 0.5
+markers = ["o", "s"]
+colors = Colorblind[8]
+labels = [
+    "Synonymous, w/in domains",
+    "Missense, w/in domains",
+    "Syn., outside domains",
+    "Mis., outside domains",
+]
+
+
+def plot_data(pop, categ, ax):
+    # within syn, within non, match syn, match non
+    to_plot = [
+        data[pop]["within"][categ]["synonymous"]["sd1"],
+        data[pop]["within"][categ]["missense"]["sd1"],
+        data[pop]["outside"]["matching"][categ]["synonymous"]["sd1"],
+        data[pop]["outside"]["matching"][categ]["missense"]["sd1"],
+    ]
+    err_plot = [
+        SE[pop]["within"][categ]["synonymous"]["sd1"],
[…]

### 2022-03-23 figures/demog_YRI_CEU.dominance.py

@@ -1,0 +1,440 @@
+import sys, os
+import moments.TwoLocus
+import matplotlib.pylab as plt
+import numpy as np
+import pickle
+import copy
+import gzip
+import time
+
+# set font sizes
+import matplotlib
+
+plt.rcParams["legend.title_fontsize"] = "xx-small"
+matplotlib.rc("xtick", labelsize=6)
+matplotlib.rc("ytick", labelsize=6)
+matplotlib.rc("axes", labelsize=7)
+matplotlib.rc("axes", titlesize=7)
+matplotlib.rc("legend", fontsize=6)
+
+from bokeh.palettes import Colorblind
+
+
+def piecewise_constant(
+    nus,
+    Ts,
+    n,
+    rho=0.0,
+    gamma=0,
+    h=0.5,
+    dom="simple",
+    eps=0.0,
+    spacing=5,
+    name=None,
+):
+    if name is None:
+        raise ValueError("need a name: bottle, expand, CEU, or YRI")
+
+    fname = f"data/time_series/{name}-n_{n}-rho_{rho}-gamma_{gamma}-h_{h}-eps_{eps}-dom_{dom}-spacing_{spacing}.bp.gz"
+    try:
+        spectra = pickle.load(gzip.open(fname, "rb"))
+        print("loaded from cached data")
+    except IOError:
+        spectra = {"t": [], "data": []}
+        # set up selection parameters
+        if gamma == 0:
+            sel_params = None
+            sel_params_general = None
+        elif h != 0.5 and eps != 0.0:
+            raise ValueError("can only have h or eps non-additive")
+        elif h != 0.5:
+            assert eps == 0.0
+            sel_params = None
+            if dom == "simple":
+                sel_params_general = moments.TwoLocus.Util.simple_dominance(gamma, h=h)
+            elif dom == "gene":
+                sel_params_general = moments.TwoLocus.Util.gene_based_dominance(
+                    gamma, h=h
+                )
+        else:
+            sel_params = moments.TwoLocus.Util.additive_epistasis(gamma, epsilon=eps)
+            sel_params_general = None
+        # simulate data
+        print(f"n={n}, rho={rho}, gamma={gamma}, h={h}, eps={eps}")
[…]

### 2022-03-23 figures/demog_YRI_CEU.dominance.sd2.py

@@ -1,0 +1,440 @@
+import sys, os
+import moments.TwoLocus
+import matplotlib.pylab as plt
+import numpy as np
+import pickle
+import copy
+import gzip
+import time
+
+# set font sizes
+import matplotlib
+
+plt.rcParams["legend.title_fontsize"] = "xx-small"
+matplotlib.rc("xtick", labelsize=6)
+matplotlib.rc("ytick", labelsize=6)
+matplotlib.rc("axes", labelsize=7)
+matplotlib.rc("axes", titlesize=7)
+matplotlib.rc("legend", fontsize=6)
+
+from bokeh.palettes import Colorblind
+
+
+def piecewise_constant(
+    nus,
+    Ts,
+    n,
+    rho=0.0,
+    gamma=0,
+    h=0.5,
+    dom="simple",
+    eps=0.0,
+    spacing=5,
+    name=None,
+):
+    if name is None:
+        raise ValueError("need a name: bottle, expand, CEU, or YRI")
+
+    fname = f"data/time_series/{name}-n_{n}-rho_{rho}-gamma_{gamma}-h_{h}-eps_{eps}-dom_{dom}-spacing_{spacing}.bp.gz"
+    try:
+        spectra = pickle.load(gzip.open(fname, "rb"))
+        print("loaded from cached data")
+    except IOError:
+        spectra = {"t": [], "data": []}
+        # set up selection parameters
+        if gamma == 0:
+            sel_params = None
+            sel_params_general = None
+        elif h != 0.5 and eps != 0.0:
+            raise ValueError("can only have h or eps non-additive")
+        elif h != 0.5:
+            assert eps == 0.0
+            sel_params = None
+            if dom == "simple":
+                sel_params_general = moments.TwoLocus.Util.simple_dominance(gamma, h=h)
+            elif dom == "gene":
+                sel_params_general = moments.TwoLocus.Util.gene_based_dominance(
+                    gamma, h=h
+                )
+        else:
+            sel_params = moments.TwoLocus.Util.additive_epistasis(gamma, epsilon=eps)
+            sel_params_general = None
+        # simulate data
+        print(f"n={n}, rho={rho}, gamma={gamma}, h={h}, eps={eps}")
[…]

### 2022-03-23 figures/demog_YRI_CEU.py

@@ -1,0 +1,443 @@
+import sys, os
+import moments.TwoLocus
+import matplotlib.pylab as plt
+import numpy as np
+import pickle
+import copy
+import gzip
+import time
+
+# set font sizes
+import matplotlib
+
+plt.rcParams["legend.title_fontsize"] = "xx-small"
+matplotlib.rc("xtick", labelsize=6)
+matplotlib.rc("ytick", labelsize=6)
+matplotlib.rc("axes", labelsize=7)
+matplotlib.rc("axes", titlesize=7)
+matplotlib.rc("legend", fontsize=6)
+
+from bokeh.palettes import Colorblind
+
+
+def piecewise_constant(
+    nus,
+    Ts,
+    n,
+    rho=0.0,
+    gamma=0,
+    h=0.5,
+    dom="simple",
+    eps=0.0,
+    spacing=5,
+    name=None,
+):
+    if name is None:
+        raise ValueError("need a name: bottle, expand, CEU, or YRI")
+
+    fname = f"data/time_series/{name}-n_{n}-rho_{rho}-gamma_{gamma}-h_{h}-eps_{eps}-dom_{dom}-spacing_{spacing}.bp.gz"
+    try:
+        spectra = pickle.load(gzip.open(fname, "rb"))
+        print("loaded from cached data")
+    except IOError:
+        spectra = {"t": [], "data": []}
+        # set up selection parameters
+        if gamma == 0:
+            sel_params = None
+            sel_params_general = None
+        elif h != 0.5 and eps != 0.0:
+            raise ValueError("can only have h or eps non-additive")
+        elif h != 0.5:
+            assert eps == 0.0
+            sel_params = None
+            if dom == "simple":
+                sel_params_general = moments.TwoLocus.Util.simple_dominance(gamma, h=h)
+            elif dom == "gene":
+                sel_params_general = moments.TwoLocus.Util.gene_based_dominance(
+                    gamma, h=h
+                )
+        else:
+            sel_params = moments.TwoLocus.Util.additive_epistasis(gamma, epsilon=eps)
+            sel_params_general = None
+        # simulate data
+        print(f"n={n}, rho={rho}, gamma={gamma}, h={h}, eps={eps}")
[…]

### 2022-03-23 figures/demog_YRI_CEU.sd2.py

@@ -1,0 +1,443 @@
+import sys, os
+import moments.TwoLocus
+import matplotlib.pylab as plt
+import numpy as np
+import pickle
+import copy
+import gzip
+import time
+
+# set font sizes
+import matplotlib
+
+plt.rcParams["legend.title_fontsize"] = "xx-small"
+matplotlib.rc("xtick", labelsize=6)
+matplotlib.rc("ytick", labelsize=6)
+matplotlib.rc("axes", labelsize=7)
+matplotlib.rc("axes", titlesize=7)
+matplotlib.rc("legend", fontsize=6)
+
+from bokeh.palettes import Colorblind
+
+
+def piecewise_constant(
+    nus,
+    Ts,
+    n,
+    rho=0.0,
+    gamma=0,
+    h=0.5,
+    dom="simple",
+    eps=0.0,
+    spacing=5,
+    name=None,
+):
+    if name is None:
+        raise ValueError("need a name: bottle, expand, CEU, or YRI")
+
+    fname = f"data/time_series/{name}-n_{n}-rho_{rho}-gamma_{gamma}-h_{h}-eps_{eps}-dom_{dom}-spacing_{spacing}.bp.gz"
+    try:
+        spectra = pickle.load(gzip.open(fname, "rb"))
+        print("loaded from cached data")
+    except IOError:
+        spectra = {"t": [], "data": []}
+        # set up selection parameters
+        if gamma == 0:
+            sel_params = None
+            sel_params_general = None
+        elif h != 0.5 and eps != 0.0:
+            raise ValueError("can only have h or eps non-additive")
+        elif h != 0.5:
+            assert eps == 0.0
+            sel_params = None
+            if dom == "simple":
+                sel_params_general = moments.TwoLocus.Util.simple_dominance(gamma, h=h)
+            elif dom == "gene":
+                sel_params_general = moments.TwoLocus.Util.gene_based_dominance(
+                    gamma, h=h
+                )
+        else:
+            sel_params = moments.TwoLocus.Util.additive_epistasis(gamma, epsilon=eps)
+            sel_params_general = None
+        # simulate data
+        print(f"n={n}, rho={rho}, gamma={gamma}, h={h}, eps={eps}")
[…]

### 2022-03-23 figures/demog_bottle_expansion.dominance.sd2.py

@@ -1,0 +1,439 @@
+# Figure 4: Demography and trajectories.
+# Two columns, left is simple size changes: expansion and bottleneck+recovery
+# Right column has human-like domographic history inferred by Relate
+# Rows: all with rho = 1
+# 1: illustration of demography
+# 2: (a) Additive selection w/ gamma=-1, (b) negative epistasis, gamma=-1, (c) negative epistasis, gamma=-10
+# 3: (a) Simple dominance, gamma=-1, (b) simple dominance, gamma=-10
+# 4: (a) Some gene-based selection model
+import sys
+import moments.TwoLocus
+import matplotlib.pylab as plt
+import numpy as np
+import pickle
+import copy
+import gzip
+import time
+
+# set font sizes
+import matplotlib
+
+plt.rcParams["legend.title_fontsize"] = "xx-small"
+matplotlib.rc("xtick", labelsize=6)
+matplotlib.rc("ytick", labelsize=6)
+matplotlib.rc("axes", labelsize=7)
+matplotlib.rc("axes", titlesize=7)
+matplotlib.rc("legend", fontsize=6)
+
+from bokeh.palettes import Colorblind
+
+
+def piecewise_constant(
+    nus,
+    Ts,
+    n,
+    rho=0.0,
+    gamma=0,
+    h=0.5,
+    dom="simple",
+    eps=0.0,
+    spacing=3,
+    name=None,
+):
+    if name is None:
+        raise ValueError("need a name: bottle, expand, CEU, or YRI")
+
+    fname = f"data/time_series/{name}-n_{n}-rho_{rho}-gamma_{gamma}-h_{h}-eps_{eps}-dom_{dom}-spacing_{spacing}.bp.gz"
+    try:
+        spectra = pickle.load(gzip.open(fname, "rb"))
+        print("loaded from cached data")
+    except IOError:
+        spectra = {"t": [], "data": []}
+        # set up selection parameters
+        if gamma == 0:
+            sel_params = None
+            sel_params_general = None
+        elif h != 0.5 and eps != 0.0:
+            raise ValueError("can only have h or eps non-additive")
+        elif h != 0.5:
+            assert eps == 0.0
+            sel_params = None
+            if dom == "simple":
[…]

### 2022-03-23 figures/demog_bottle_expansion.py

@@ -1,0 +1,456 @@
+# Figure 4: Demography and trajectories.
+# Two columns, left is simple size changes: expansion and bottleneck+recovery
+# Right column has human-like domographic history inferred by Relate
+# Rows: all with rho = 1
+# 1: illustration of demography
+# 2: (a) Additive selection w/ gamma=-1, (b) negative epistasis, gamma=-1, (c) negative epistasis, gamma=-10
+# 3: (a) Simple dominance, gamma=-1, (b) simple dominance, gamma=-10
+# 4: (a) Some gene-based selection model
+import sys
+import moments.TwoLocus
+import matplotlib.pylab as plt
+import numpy as np
+import pickle
+import copy
+import gzip
+import time
+
+# set font sizes
+import matplotlib
+
+plt.rcParams["legend.title_fontsize"] = "xx-small"
+matplotlib.rc("xtick", labelsize=6)
+matplotlib.rc("ytick", labelsize=6)
+matplotlib.rc("axes", labelsize=7)
+matplotlib.rc("axes", titlesize=7)
+matplotlib.rc("legend", fontsize=6)
+
+from bokeh.palettes import Colorblind
+
+
+def piecewise_constant(
+    nus,
+    Ts,
+    n,
+    rho=0.0,
+    gamma=0,
+    h=0.5,
+    dom="simple",
+    eps=0.0,
+    spacing=3,
+    name=None,
+):
+    if name is None:
+        raise ValueError("need a name: bottle, expand, CEU, or YRI")
+
+    fname = f"data/time_series/{name}-n_{n}-rho_{rho}-gamma_{gamma}-h_{h}-eps_{eps}-dom_{dom}-spacing_{spacing}.bp.gz"
+    try:
+        spectra = pickle.load(gzip.open(fname, "rb"))
+        print("loaded from cached data")
+    except IOError:
+        spectra = {"t": [], "data": []}
+        # set up selection parameters
+        if gamma == 0:
+            sel_params = None
+            sel_params_general = None
+        elif h != 0.5 and eps != 0.0:
+            raise ValueError("can only have h or eps non-additive")
+        elif h != 0.5:
+            assert eps == 0.0
+            sel_params = None
+            if dom == "simple":
[…]

### 2022-03-23 figures/demog_bottle_expansion.sd2.py

@@ -1,0 +1,460 @@
+# Figure 4: Demography and trajectories.
+# Two columns, left is simple size changes: expansion and bottleneck+recovery
+# Right column has human-like domographic history inferred by Relate
+# Rows: all with rho = 1
+# 1: illustration of demography
+# 2: (a) Additive selection w/ gamma=-1, (b) negative epistasis, gamma=-1, (c) negative epistasis, gamma=-10
+# 3: (a) Simple dominance, gamma=-1, (b) simple dominance, gamma=-10
+# 4: (a) Some gene-based selection model
+import sys
+import moments.TwoLocus
+import matplotlib.pylab as plt
+import numpy as np
+import pickle
+import copy
+import gzip
+import time
+
+# set font sizes
+import matplotlib
+
+plt.rcParams["legend.title_fontsize"] = "xx-small"
+matplotlib.rc("xtick", labelsize=6)
+matplotlib.rc("ytick", labelsize=6)
+matplotlib.rc("axes", labelsize=7)
+matplotlib.rc("axes", titlesize=7)
+matplotlib.rc("legend", fontsize=6)
+
+from bokeh.palettes import Colorblind
+
+
+def piecewise_constant(
+    nus,
+    Ts,
+    n,
+    rho=0.0,
+    gamma=0,
+    h=0.5,
+    dom="simple",
+    eps=0.0,
+    spacing=3,
+    name=None,
+):
+    if name is None:
+        raise ValueError("need a name: bottle, expand, CEU, or YRI")
+
+    fname = f"data/time_series/{name}-n_{n}-rho_{rho}-gamma_{gamma}-h_{h}-eps_{eps}-dom_{dom}-spacing_{spacing}.bp.gz"
+    try:
+        spectra = pickle.load(gzip.open(fname, "rb"))
+        print("loaded from cached data")
+    except IOError:
+        spectra = {"t": [], "data": []}
+        # set up selection parameters
+        if gamma == 0:
+            sel_params = None
+            sel_params_general = None
+        elif h != 0.5 and eps != 0.0:
+            raise ValueError("can only have h or eps non-additive")
+        elif h != 0.5:
+            assert eps == 0.0
+            sel_params = None
+            if dom == "simple":
[…]

### 2022-03-23 figures/demog_relate.py

@@ -11,6 +11,7 @@ import moments.TwoLocus
 import matplotlib.pylab as plt
 import numpy as np
 import pickle
+import copy
 
 # set font sizes
 import matplotlib
@@ -41,10 +42,7 @@ def piecewise_constant(
     conditions=[],
 ):
     if track_stats:
-        stats = {"t": [], "sd1": {"all": [],}, "sd2": {"all": [],}}
-        for cond in conditions:
-            stats["sd1"][cond] = []
-            stats["sd2"][cond] = []
+        spectra = {"t": [], "data": []}
     if gamma == 0:
         sel_params = None
         sel_params_general = None
@@ -64,12 +62,8 @@ def piecewise_constant(
         n, rho=rho, sel_params=sel_params, sel_params_general=sel_params_general
     )
     if track_stats:
-        stats["t"].append(0)
-        stats["sd1"]["all"].append(F.D() / F.pi2())
-        stats["sd2"]["all"].append(F.D2() / F.pi2())
-        for cond in conditions:
-            stats["sd1"][cond].append(F.D(nA=cond, nB=cond) / F.pi2(nA=cond, nB=cond))
-            stats["sd2"][cond].append(F.D2(nA=cond, nB=cond) / F.pi2(nA=cond, nB=cond))
+        spectra["t"].append(0)
+        spectra["data"].append(copy.deepcopy(F))
 
     print("o" * len(nus), flush=True)
     for nu, T in zip(nus, Ts):
@@ -92,18 +86,10 @@ def piecewise_constant(
                     sel_params=sel_params,
                     sel_params_general=sel_params_general,
                 )
-                stats["t"].append(stats["t"][-1] + T_sub)
-                stats["sd1"]["all"].append(F.D() / F.pi2())
-                stats["sd2"]["all"].append(F.D2() / F.pi2())
-                for cond in conditions:
-                    stats["sd1"][cond].append(
-                        F.D(nA=cond, nB=cond) / F.pi2(nA=cond, nB=cond)
-                    )
-                    stats["sd2"][cond].append(
-                        F.D2(nA=cond, nB=cond) / F.pi2(nA=cond, nB=cond)
-                    )
+                spectra["t"].append(spectra["t"][-1] + T_sub)
[…]

### 2022-03-23 figures/epistasis_prediction.py

@@ -29,7 +29,7 @@ n = 30  # the projection size
 nA = 10
 nB = 10
 
-data_fname = f"fig1_data_ns_{n0}_{n}.bp"
+data_fname = f"data/fig1_data_ns_{n0}_{n}.bp"
 try:
     data1 = pickle.load(open(data_fname, "rb"))
 except IOError:
@@ -89,59 +89,13 @@ rhos = data1["C"]["rhos"]
 
 ms = 4
 
-fig = plt.figure(1, figsize=(6.5, 5))
-
-## previous c and d from figure 1
-
-ax1 = plt.subplot(3, 2, 1)
-
-ax1.plot(rhos, data1["C"]["Ohta_Kimura"], "k--", lw=1, label="Ohta-Kimura")
-
-for ii, gamma in enumerate(gammas):
-    ax1.plot(
-        rhos,
-        data1["C"][gamma]["all"]["sd2"],
-        markers[ii] + "-",
-        ms=ms,
-        lw=0.5,
-        label=rf"$\gamma={gamma}$",
-        color=colors[ii],
-    )
-
-ax1.set_yscale("log")
-ax1.set_xscale("log")
-ax1.set_ylabel(r"$\sigma_d^2$")
-ax1.set_title("Normalized variance of $D$ ($\sigma_d^2$)")
-
-ax2 = plt.subplot(3, 2, 2)
-
-ax2.plot(rhos, 0 * data1["C"]["Ohta_Kimura"], "k--", lw=1, label=None)
-
-for ii, gamma in enumerate(gammas):
-    ax2.plot(
-        rhos,
-        data1["C"][gamma]["all"]["sd1"],
-        markers[ii] + "-",
-        ms=ms,
-        lw=0.5,
-        label=rf"$\gamma={gamma}$",
-        color=colors[ii],
-    )
-
-ax2.set_xscale("log")
-ax2.set_ylabel(r"$\sigma_d^1$")
-ax2.set_title("Normalized signed LD ($\sigma_d^1$)")
-ax2.legend()
-
-
-### previous figure 2
 epsilons = [-0.5, 0.0, 0.5, 1.0]
 gammas = [-1, -10]
 conditions = ["all", 2, 3, 4, 10]
 
 n0 = 50
 
-data_fname = f"fig1b_data_ns_{n0}_{n}.bp"
+data_fname = f"data/fig1b_data_ns_{n0}_{n}.bp"
 try:
[…]

### 2022-03-23 figures/fig2.py

@@ -1,159 +1,0 @@
-# Figure 3: Dominance selection model.
-import sys
-import moments.TwoLocus
-import matplotlib.pylab as plt
-import numpy as np
-import pickle
-
-# set font sizes
-import matplotlib
-
-plt.rcParams["legend.title_fontsize"] = "xx-small"
-matplotlib.rc("xtick", labelsize=6)
-matplotlib.rc("ytick", labelsize=6)
-matplotlib.rc("axes", labelsize=7)
-matplotlib.rc("axes", titlesize=7)
-matplotlib.rc("legend", fontsize=6)
-
-from bokeh.palettes import Colorblind
-
-n0 = 70  # larger sample size for improved accuracy
-n = 30  # the projection size
-
-hs = [0.0, 0.1, 0.5, 1.0]
-gammas = [-1, -5]
-conditions = ["all", 2, 3, 4, 10]
-
-data_fname = f"fig2_data_ns_{n0}_{n}.bp"
-try:
-    data = pickle.load(open(data_fname, "rb"))
-except IOError:
-    data = {}
-    data["ns"] = (n0, n)
-    # stores the sigma_d^2 and sigma_d^1 curves for all snps and
-    rhos = np.logspace(-2, np.log10(30), 20)
-    data["rhos"] = rhos
-    data["Ohta_Kimura"] = (5 + rhos / 2) / (11 + 13 * rhos / 2 + rhos ** 2 / 2)
-    for gamma in gammas:
-        for h in hs:
-            print("Running gamma, h = ", gamma, ",", h, end=" ", flush=True)
-            sel_params = moments.TwoLocus.Util.simple_dominance(gamma, h=h)
-            data[(gamma, h)] = {cond: {"sd1": [], "sd2": []} for cond in conditions}
-            for rho in rhos:
-                print(".", end="", flush=True)
-                F0 = moments.TwoLocus.Demographics.equilibrium(
-                    n0, rho=rho, sel_params_general=sel_params
-                )
-                F = F0.project(n)
-                for cond in conditions:
-                    if cond == "all":
-                        data[(gamma, h)][cond]["sd1"].append(F.D() / F.pi2())
[…]

### 2022-03-23 figures/fig4.py

@@ -1,234 +1,0 @@
-# Figure 5: LD within and between genes
-import sys
-import moments.TwoLocus
-import matplotlib.pylab as plt
-import numpy as np
-import pickle
-
-# set font sizes
-import matplotlib
-
-plt.rcParams["legend.title_fontsize"] = "xx-small"
-matplotlib.rc("xtick", labelsize=6)
-matplotlib.rc("ytick", labelsize=6)
-matplotlib.rc("axes", labelsize=7)
-matplotlib.rc("axes", titlesize=7)
-matplotlib.rc("legend", fontsize=6)
-
-from bokeh.palettes import Colorblind
-
-data = {}
-
-bs = {}
-
-populations = {
-    "Africa": ["ESN", "GWD", "LWK", "MSL", "YRI"],
-    "Europe": ["CEU", "FIN", "GBR", "IBS", "TSI"],
-    "East Asia": ["CDX", "CHB", "CHS", "JPT", "KHV"],
-}
-
-for continent, pops in populations.items():
-    for pop in pops:
-        ld_pop = pickle.load(
-            open(f"../analysis/parsed_data/{pop}.unphased.within.between.bp", "rb")
-        )
-        data[pop] = ld_pop
-        bs_pop = pickle.load(
-            open(
-                f"../analysis/parsed_data/{pop}.unphased.within.between.sigmad1.bootstrap_std_err.bp",
-                "rb",
-            )
-        )
-        bs[pop] = bs_pop
-
-ms = 2
-lw = 0.5
-
-fig = plt.figure(5, figsize=(6.5, 4.5))
-fig.clf()
-
-markers = ["x", "+", "1"]
-colors = Colorblind[8]
-
-# plot all stats within genes
-ax1 = plt.subplot2grid((7, 4), (0, 0), rowspan=2, colspan=4)
-ax1b = plt.subplot2grid((7, 4), (2, 0), rowspan=2, colspan=4)
-
-xx = np.concatenate(
-    (np.linspace(1, 15, 15), np.linspace(18, 32, 15), np.linspace(35, 49, 15))
-)
-yy = []
-err = []
-
-for continent in ["Africa", "Europe", "East Asia"]:
-    for pop in populations[continent]:
-        yy.append(
-            data[pop]["ld_within"]["synonymous"][3]
[…]

### 2022-03-23 figures/fig5.py

@@ -1,320 +1,0 @@
-# Figure 6: LD within and between gene domains
-import sys
-import moments.TwoLocus
-import matplotlib.pylab as plt
-import numpy as np
-import pickle
-
-# set font sizes
-import matplotlib
-
-plt.rcParams["legend.title_fontsize"] = "xx-small"
-matplotlib.rc("xtick", labelsize=6)
-matplotlib.rc("ytick", labelsize=6)
-matplotlib.rc("axes", labelsize=7)
-matplotlib.rc("axes", titlesize=7)
-matplotlib.rc("legend", fontsize=6)
-
-from bokeh.palettes import Colorblind
-
-data = {}
-
-bs = {}
-
-populations = {
-    "Africa": ["ESN", "GWD", "LWK", "MSL", "YRI"],
-    "Europe": ["CEU", "FIN", "GBR", "IBS", "TSI"],
-    "East Asia": ["CDX", "CHB", "CHS", "JPT", "KHV"],
-}
-
-for continent, pops in populations.items():
-    for pop in pops:
-        ld_pop = pickle.load(
-            open(f"../analysis/parsed_data/{pop}.unphased.domains.bp", "rb")
-        )
-        data[pop] = ld_pop
-        bs_pop = pickle.load(
-            open(
-                f"../analysis/parsed_data/{pop}.unphased.domains.sigmad1.bootstrap_std_err.bp",
-                "rb",
-            )
-        )
-        bs[pop] = bs_pop
-
-ms = 2
-lw = 0.5
-markers = ["x", "+", "1"]
-colors = Colorblind[8]
-
-def plot_inside_outside(pop, ax, legend=False, ylabel=False, xlabel=True):
-    # plot missense and synonymous for each, in the order:
-    # gene-wide, within domains, between domains, all outside domains, matched to
-    #   distances within, matched to distances between
-    to_plot = []
-    err = []
-    # add gene-wide values
-    data_all = pickle.load(
-        open(f"../analysis/parsed_data/{pop}.unphased.within.between.bp", "rb")
-    )
-    bs_all = pickle.load(
-        open(
-            f"../analysis/parsed_data/{pop}.unphased.within.between.sigmad1.bootstrap_std_err.bp",
-            "rb",
-        )
-    )
-    to_plot.append(
[…]

### 2022-03-23 figures/gene-based-compensatory.py

@@ -1,0 +1,200 @@
+# Unconditioned vs allele count-conditioned LD.  Four panels, first two showing
+# Hudson slices in the with rho=1 and rho=10 for neutrality and with gamma=-5
+# at both loci. Second two panels show \sigma_d^2 and \sigma_d^1 for nuetrality
+# and using all SNPs with gamma=-5, as well as conditioning on doubletons,
+# tripletons, and few other frequencies.
+
+import moments.TwoLocus
+import matplotlib.pylab as plt
+import numpy as np
+import pickle
+import time
+import gzip
+
+# set font sizes
+import matplotlib
+
+plt.rcParams["legend.title_fontsize"] = "xx-small"
+matplotlib.rc("xtick", labelsize=6)
+matplotlib.rc("ytick", labelsize=6)
+matplotlib.rc("axes", labelsize=7)
+matplotlib.rc("axes", titlesize=7)
+matplotlib.rc("legend", fontsize=6)
+
+from bokeh.palettes import Colorblind
+
+colors = Colorblind[8]
+
+gammas = [-1, -20]
+h = 0.05
+eps = -0.95
+
+
+try:
+    spectra_compens = pickle.load(gzip.open("data/compensatory.bp", "rb"))
+except IOError:
+    rhos = np.concatenate((np.logspace(-2, 1, 22), [15, 20, 30]))
+    spectra_compens = {}
+    spectra_compens["rhos"] = rhos
+    n = 50
+    for gamma in gammas:
+        spectra_compens[(gamma, eps)] = {}
+        sel_params = moments.TwoLocus.Util.additive_epistasis(gamma, eps)
+        for rho in rhos:
+            time1 = time.time()
+            print("rho:", rho)
+            F_c = moments.TwoLocus.Demographics.equilibrium(
+                n, rho=rho, sel_params=sel_params
+            )
+            spectra_compens[(gamma, eps)][rho] = F_c
+            print(" finished in", int(time.time() - time1), "seconds")
+            print(" min:", F_c.min())
+            print(" sd1:", F_c.D() / F_c.pi2())
[…]

### 2022-03-23 figures/hill_robertson.py

@@ -1,0 +1,118 @@
+# Unconditioned vs allele count-conditioned LD.  Four panels, first two showing
+# Hudson slices in the with rho=1 and rho=10 for neutrality and with gamma=-5
+# at both loci. Second two panels show \sigma_d^2 and \sigma_d^1 for nuetrality
+# and using all SNPs with gamma=-5, as well as conditioning on doubletons,
+# tripletons, and few other frequencies.
+
+import moments.TwoLocus
+import matplotlib.pylab as plt
+import numpy as np
+import pickle
+import gzip
+
+# set font sizes
+import matplotlib
+
+plt.rcParams["legend.title_fontsize"] = "xx-small"
+matplotlib.rc("xtick", labelsize=6)
+matplotlib.rc("ytick", labelsize=6)
+matplotlib.rc("axes", labelsize=7)
+matplotlib.rc("axes", titlesize=7)
+matplotlib.rc("legend", fontsize=6)
+
+from bokeh.palettes import Colorblind
+
+colors = Colorblind[8]
+
+markers = ["o", "s", "v", "^"]
+ms = 2
+
+fig = plt.figure(1, figsize=(6.5, 2))
+fig.clf()
+
+ax1 = plt.subplot(1, 3, 1)
+
+n0 = 40
+n = 30
+data = pickle.load(open(f"data/fig1_data_ns_{n0}_{n}.bp", "rb"))
+gammas = [-0.1, -1, -5, -10]
+rhos = data["C"]["rhos"]
+
+ax1.plot(rhos, data["C"]["Ohta_Kimura"], "k--", lw=1, label="Ohta & Kimura")
+
+for ii, gamma in enumerate(gammas):
+    ax1.plot(
+        rhos,
+        data["C"][gamma]["all"]["sd2"],
+        #markers[ii] + "-",
+        "-",
+        ms=ms,
+        lw=1,
+        label=rf"$\gamma={gamma}$",
+        color=colors[ii],
+    )
+
+ax1.set_yscale("log")
+ax1.set_xscale("log")
+ax1.set_ylabel(r"$\sigma_d^2$")
+ax1.set_title("Normalized variance of $D$ ($\sigma_d^2$)")
+ax1.set_xlabel(r"$\rho=4Nr$")
+
+ax2 = plt.subplot(1, 3, 2)
+
[…]

### 2022-03-23 figures/ld_decay_gene_wide.py

@@ -1,0 +1,85 @@
+import sys
+import moments.TwoLocus
+import matplotlib.pylab as plt
+import numpy as np
+import pickle
+
+# set font sizes
+import matplotlib
+
+plt.rcParams["legend.title_fontsize"] = "xx-small"
+matplotlib.rc("xtick", labelsize=6)
+matplotlib.rc("ytick", labelsize=6)
+matplotlib.rc("axes", labelsize=7)
+matplotlib.rc("axes", titlesize=7)
+matplotlib.rc("legend", fontsize=6)
+
+from bokeh.palettes import Colorblind
+
+data = {}
+
+populations = {
+    "Africa": ["ESN", "GWD", "LWK", "MSL", "YRI"],
+    "Europe": ["CEU", "FIN", "GBR", "IBS", "TSI"],
+    "East Asia": ["CDX", "CHB", "CHS", "JPT", "KHV"],
+}
+
+for c, ps in populations.items():
+    for p in ps:
+        data[p] = pickle.load(
+            open(f"../analysis/parsed_data_v2/{p}.unphased.all.bp", "rb")
+        )
+
+ms = 2
+lw = 0.5
+markers = ["x", "+", "1"]
+colors = Colorblind[8]
+
+bins = data[p]["bin_edges"]
+bin_mids = np.mean(bins, axis=1)
+
+fig = plt.figure(5, figsize=(6.5, 3.5))
+
+fig.clf()
+
+i = 0
+for continent, pops in populations.items():
+    i += 1
+    ax1 = plt.subplot(2, 3, i)
+    ax2 = plt.subplot(2, 3, i + 3)
+    label_syn = "Synonymous"
+    label_mis = "Missense"
+    ax2.plot(bin_mids[:-1], 0 * bin_mids[:-1], "k--", lw=lw)
+    for pop in pops:
+        sd2_syn = data[pop]["bins"]["all_sites"]["synonymous"]["sd2"]
+        sd2_mis = data[pop]["bins"]["all_sites"]["missense"]["sd2"]
+        sd1_syn = data[pop]["bins"]["all_sites"]["synonymous"]["sd1"]
+        sd1_mis = data[pop]["bins"]["all_sites"]["missense"]["sd1"]
+        ax1.plot(
+                bin_mids[:-1], sd2_syn[:-1], "o--", color=colors[0], ms=ms, lw=lw, label=label_syn
+        )
[…]

### 2022-03-23 manuscript/paper.tex
add tex files

\documentclass[]{article}
\usepackage[round]{natbib}

\usepackage[margin=1in]{geometry}
\usepackage{url} 
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{color}
\usepackage{booktabs}

\usepackage{amsmath,amssymb}
\usepackage{longtable,booktabs,array}

\newcommand{\E}{\mathbb{E}}
\newcommand{\moments}{\texttt{moments}\xspace}
\newcommand{\Relate}{\texttt{Relate}\xspace}
\newcommand{\msprime}{\texttt{msprime}\xspace}

\newcommand{\comment}[1]{{\textcolor{red}{APR: #1}}}

% cross-reference with supplement
\usepackage{xr}
\externaldocument{supplement}

\begin{document}

\title{A heterogeneous landscape of selection and epistasis in protein-coding genes revealed by two-locus statistics}
\author[]{Aaron P. Ragsdale}
\affil[]{Department of Integrative Biology, University of Wisconsin, Madison, WI, USA}
\affil[]{apragsdale@wisc.edu}
\maketitle

\begin{abstract}

Selected mutations interfere and interact with evolutionary processes at nearby
loci, distorting allele frequency trajectories and creating correlations
between pairs of mutations. A number of recent studies have used patterns of
linkage disequilibrium (LD) between selected variants to test for selective
interference and epistatic interactions, with some disagreement over
interpreting observations from data. Interpretation is hindered by the relative
lack of analytic or even numerical expectations for patterns of variation
between pairs of loci under the combined effects of selection, dominance,
epistasis, and demography. Here, I develop a numerical approach to compute the
expected two-locus sampling distribution under diploid selection with arbitrary
epistasis and dominance, recombination, and variable population size. I use
this to explore how epistasis and dominance affect expected signed LD,
including for non-steady-state demography relevant to human populations.
\comment{I then use whole-genome sequencing data from humans to assess how well we
can differentiate modes of selective interactions in practice.}
\comment{this needs updating}
I find that
positive LD between missense mutations within genes is driven by strong
positive allele-frequency correlations between pairs of mutations that fall
within the same conserved domain, pointing to compensatory mutations or
antagonistic epistasis as the prevailing mode of interaction within but not
outside of conserved genic elements. The heterogeneous landscape of both
mutational fitness effects and selective interactions within protein-coding
genes calls for more refined inferences of the joint distribution of fitness
and interactive effects, and the methods presented here should prove useful in
that pursuit.

\end{abstract}

\section{Introduction}\label{sec:introduction}

Most new mutations that affect fitness are deleterious and tend to be
eliminated from a population
\textbf{Citations that most mutations are negatively selected}.
The amount of time that a
deleterious mutation segregates depends on the strength of selection against
genomes that carry it, with very damaging mutations kept at low frequencies and
purged relatively rapidly. In the time between mutation and fixation or loss,
selected variants, both beneficial and damaging, can dramatically impact
patterns of variation in nearby linked regions \textbf{MSH, Charlesworth, Stephan/Kim}.
This distortion away from neutral expectations is now empirically documented using
sequencing data from an ever-growing set of study systems, but questions remain
about the primary mode of interactions between multiple linked variants and
their joint effects on genome-wide patterns of diversity.

In their foundational paper, \citet{Hill1966-gv} recognized that linked
selected variants reciprocally impede the efficacy of selection at each locus,
a process known as selective interference. Linked selection reduces the
fixation probability of advantageous mutations and increases that of
deleterious mutations compared to expectations under single-locus models
\citep{Birky1988-jm}. Allele frequency dynamics and correlations of linked
selected variants are also predicted to deviate from expectations without
interference. Under a multiplicative fitness model, where the fitness reduction
of a genome carrying multiple deleterious variants is equal to the product of
the fitness reduction of each mutation independently, we expect to see net
linkage disequilibrium (LD) equal to zero for unlinked sites
\citep{Kondrashov1995-va}. But for linked loci, those mutations are expected to
segregate on different haplotypes more often than together, leading to
negative, or repulsion, LD, although the extent of LD depends non-trivially on
the strength of selection and probability of recombination separating loci
\citep{Hill1966-gv, McVean2000-ox}.

Non-additive effects, including dominance (i.e., interactions \emph{within} a
locus) and epistasis (interactions \emph{between} loci), further complicate our
evolutionary models. A large fraction of nonsynonymous coding mutations are
thought to be at least partially recessive \citep{Agrawal2011-wb,
Huber2018-cp}, with average levels of dominance correlating with strength of
selection \citep{Kacser1981-nc}, and dominance plays an important role in
shaping expected equilibrium allele frequencies and the mutation load of
strongly damaging disease mutations \citep{Clark1998-kq}. On the other hand,
epistasis differentially impacts the deleterious load in asexually and sexually
reproducing organisms \citep{Kimura1966-gc, Kondrashov1995-va}, has been
invoked as an explanation for the evolutionary advantage of sex
\citep{Kondrashov1982-sf, Charlesworth1990-kw, Barton1998-lu}, and can drive
incompatibilities that lead to postzygotic isolation during the process of
speciation \citep{Turelli2000-kz}. Within populations, epistasis is known to
cause signed LD to deviate dramatically from zero \citep{Charlesworth1990-kw,
Kondrashov1995-va}. However, despite appreciation of the effect of dominance on
linked variation \citep{Turelli2000-kz, Zhao2016-bb} and the evolutionary
importance of epistatic interactions, we currently lack models for predicting
correlation patterns between linked mutations under general selection models.

In this paper, I develop a numerical approach to solve for the two-locus
sampling distribution under a general diploid selection model with variable
recombination and single-population size history. I use this model to describe
how epistasis and dominance shape expected patterns of signed LD, under both
steady-state and non-equilibrium demography, that have been used to test for
interference and epistasis in population genomic data. I then turn to human
sequencing data and compare patterns of LD for synonymous, missense, and
loss-of-function mutations within and between protein-coding genes and
annotated domains. I show that while synonymous and missense variants display
similar slightly positive average LD within genes, for missense mutations this
signal is driven by correlations between pairs of mutations within, but not
between or outside of, protein-coding domains.
This suggests an importance for antagonistic (or ``diminishing-returns'')
epistasis or a prevalence of compensatory
nonsynonymous mutations within conserved elements.
%Finally, I discuss how well
%we can expect to discriminate between modes of selective interactions using
%genome-wide patterns of LD.

\subsection{Empirical observations}\label{sec:empirical-observations}

The most direct way to test for interactions between linked selected variants
is through deep mutation scanning experiments, in which many distinct mutations
are introduced within a target gene and then organismal fitness or some protein
function is experimentally measured \citep{Romero2009-yi, Bank2015-vq,
Puchta2016-rx, Steinberg2016-is}. For example, using the model system of the
TEM-1 \(\beta\)-lactamase gene in \emph{E. coli}, \citet{Bershtein2006-bk}
found evidence for synergistic epistasis, in which the combined effect of
multiple deleterious mutations on individual fitness was larger than would be
expected from multiplying the independent observed effects of each individual
mutation. The scale of mutation scanning experiments continues to improve
dramatically, promising greater resolution of the fitness landscape in such
model systems that can be compared to evolutionary theory
\citep{Otwinowski2018-zk}.

Directed mutational studies are not possible in most natural populations, and
we must turn to population genetic approaches to infer selective interactions
between observed segregating polymorphisms. Motivated by theory that linked
negatively selected mutations will display negative LD due to interference
\citep{Hill1966-gv}, and that epistasis will drive expected LD away from zero,
a number of recent studies have used patterns of LD within classes of
putatively selected variants to infer modes of selective interactions.
\citet{Callahan2011-ac} observed that pairs of tightly linked nonsynonymous
mutations cluster together more than expected along the same lineages in the
\emph{Drosophilid} species complex, and that those clustered mutations tended
to preserve the charge of the protein and were in positive LD compared to pairs
involving synonymous mutations. From this, they proposed that compensatory
nonsynonymous variants are regularly tolerated and maintained. More recently,
\citet{Taverner2020-lk} replicated this finding across a diverse set of genera,
showing that such epistatic interactions are important for protein evolution.

\citet{Sohail2017-zq} observed negative LD between loss-of-function variants in
protein-coding genes (loss-of-function mutations include stop gains and losses,
frameshifts, and other nonsense mutations) in both human and fruit fly
populations. They interpreted this as evidence for widespread synergistic
epistasis between these mutations, in which the fitness reduction of multiple
mutations is greater than the product of that of each individual mutations
independently. Both \citet{Sandler2021-of} and \citet{Garcia2021-zn} have
recently reevaluated patterns of LD between coding variants in humans, fruit
flies, and \emph{Capsella grandiflora}, and suggested interference and
dominance may instead be driving patterns of LD \citep{Garcia2021-zn} or
questioned whether LD between loss-of-function variants is significantly
different from zero \citep{Sandler2021-of}.

A number of factors impede our interpretation of patterns of signed LD between
coding variants. First, for strongly deleterious or loss-of-function mutations,
their low allele frequencies mean that statistical measurement of LD
and other diversity measures are very noisy. Second, comparisons are based on
theory with limiting assumptions, such as steady-state demography, simple
selection and interaction models, or unlinked loci. To generate predictions
under more complex models, we rely on expensive forward simulations. Such
simulations can help build intuition and be used to test inference methods, but
they do not efficiently provide expectations for quantities of interest across
the range of relevant parameters. Analytical and numerical methods for expected
haplotype frequencies and LD under general selective interaction models are
thus crucial for interpreting patterns of variation observed in data.

\section{Results}\label{sec:results}

\subsection{Expected signed LD under steady-state demography}
\label{sec:signed-ld}

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/two-locus-spectrum/stats_summaries}
    \caption{
        \textbf{Sampling distributions and their summaries.}
        Low-order summaries of sampling distributions are commonly computed
        for allele allele frequencies (A, the site-frequency spectrum, SFS)
        and two-locus haplotype distributions (B, linkage disequilibrium, LD).
        Demographic and selective processes affect both the SFS and LD,
        and observations of nonzero values of Tajima's $D$ or signed LD
        ($\sigma_d^1$) are often taken as evidence for selection or interations
        between loci, respectively.
        (C) The full two-locus haplotype sampling distribution is a three-dimensional
        object, making it difficult to visualize. We can instead visualize conditional
        distributions of the full sampling distribution, e.g. conditioned on observing
        five copies of the $A$ allele at the left locus and five copies of $B$ at the
        right locus \citep[e.g.][]{Hudson2001-sg}.
        (D) $\sigma_d^2$, which is closely related to $r^2$, is expected to decay
        with increasing recombination distance between loci \citep{Ohta1969-ie}.
        Selection distorts the level of squared LD away from neutral expectations.
        (E) $\sigma_d^1$, refered to as signed LD here, is expected to be zero for
        pairs of neutral mutations \citep{Hill1968-vu}. Interference between linked
        selected mutation causes negative LD \citep{Hill1966-gv}, and other
        forms of interactions between selected mutations can cause large negative
        or positive signed LD.
        \comment{selected in blue to match all panels, panel labels, and panel A
        should include neutral expectation}
    }
    \label{fig:TwoLocusFS}
\end{figure}

In the Methods, I expand on the moments system developed in
\citet{Ragsdale2019-nt} to compute the expected sampling distribution
\(\Psi_n\) of two-locus haplotypes under a general model of selective
interactions (Figure~\ref{fig:TwoLocusFS}). This sampling distribution stores the
expected density or observed counts of pairs of biallelic loci with each
possible haplotype frequency configuration in a sample of size \(n\). Below, we
compute expectations for \(\Psi_n\) under varying scenarios of selection and
interaction between pairs of loci. It is not possible in this framework to
include the effects of additional linnked selected mutations, such as background
selection due to many linked variants. Individual-based forward simulations are
still needed for such scenarios (e.g., Figures~\ref{fig:bgs1}--\ref{fig:bgs4}).

In many cases it is simpler to visualize summaries such as the expectation or
variance of \(D\) (e.g., Figure~\ref{fig:TwoLocusFS}D,E) or slices of the
distribution (e.g., Figure~\ref{fig:TwoLocusFS}C) instead of the full
three-dimensional distribution. Here I focus on low-order LD statistics
including \(\E[D]\) and \(\E[D^2]\) and their decay with recombination
distance, as these are statistics that are commonly used to test for
interactions between loci. For pairs of biallelic loci, with alleles labeled
\(A/a\) at the left locus and \(B/b\) at the right locus, \(D=f_{AB}-f_A f_B\)
is the standard covariance measure of LD, where \(f_{AB}\) is the frequency of
haplotypes carrying both \(A\) and \(B\), and \(f_A\) and \(f_B\) are the
marginal allele frequencies of \(A\) and \(B\).

Instead of unnormalized \(\E[D^2]\) and \(\E[D]\) I consider expectations for
\(\sigma_d^2 = \E[D^2]/\E[f_A(1-f_A)f_B(1-f_B)]\) and
\(\sigma_d^1 = \E[D]/\E[f_A(1-f_A)f_B(1-f_B)]\).
Such normalized statistics have two benefits:
first, the mutation rate cancels so that expectations are robust to assumptions
about the per-base mutation rate, and second, analytic expectations for these
quantities are available under neutrality at constant population size for
comparison \citep{Ohta1971-yd}.

Below, I first consider the case of additive selection, Hill-Robertson
interference, and epistasis. I then explore the effect of dominance acting
within loci but without epistatis, and then describe a general diploid
selection model and consider gene-based dominance effects. In what follows I
focus on parameters where the strengths of selection and dominance at each
locus are equal, but note that the methods presented here allow for arbitrary
and unequal selection and dominance at the two loci. I also focus primarily on
weak to moderate negative selection (\(|2Ns| \approx 1 - 20\)), since this
range of selection leads to the strongest signals of interference
(Figure~\ref{fig:HillRobertson}) and is the parameter regime for which the
numerical appraoch is most accurate.

\subsubsection{Additive selection and epistasis}
\label{sec:additive-selection}

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/hill_robertson}
    \caption{
        \textbf{Hill-Robertson interference.}
        Interference between pairs of selected mutations can cause deviations
        from expected patterns of LD under neutrality, without dominance
        or epistatic effects.
        (A) The expected normalized variance of $D$ decreases as the strength
        of negative selection increases.
        (B, C) For tightly linked loci (\(4Nr \lesssim 1\)), interference is most
        noticeable for pairs of mutations with \(s \approx 1/N\) as measured by
        signed LD.
        At larger recombination distances (\(4Nr > 1\)), signed LD is most
        negative for stronger selection coefficients. Dashed lines
        show neutral expectations.
    }
    \label{fig:HillRobertson}
\end{figure}

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/epistasis_prediction}
    \caption{
        \textbf{Additive selection and epistasis.}
        Left panels (A and C) show expectations for the decay of
        \(\sigma_d^2 = \E[D^2] / \E[p(1-p)q(1-q)]\) with recombination
        distance, and right panesl (B and D) show expectations for the
        decay of \(\sigma_d^1 = \E[D] / \E[p(1-p)q(1-q)]\). Dashed lines
        show neutral expectations.
        For both weak selection (\(s=-1/2N\), A and B) and stronger selection
        (\(s=-10/2N\), C and D), antagonistic epistasis (\(\epsilon < 0\))
        gives rise to positive signed LD and increased $\sigma_d^2$ over
        a multiplicative model (\(\epsilon = 0\)),
        and synergistic epistasis (\(\epsilon > 0\)) results in negative signed
        LD beyond that of Hill-Robertson interference and decreased \(\sigma_d^2\).
        With epistasis between pairs of selected mutations, nonzero signed LD
    }
    \label{fig:epistasis}
\end{figure}

For mutations under additive selection (\(h=1/2\)) and no epistasis, we
recover the well known \citet{Hill1966-gv} interference result of negative LD
between selected mutations, which is strongest for pairs of mutations that have
selection coefficients \(\gamma = O(1)\), or \(s \approx 1/2N_e\) (Figure
\ref{fig:epistasis}B).
For strongly deleterious mutations, LD is close to zero even
with tight linkage, as they almost always segregate at low enough frequencies
that they are unlikely to interfere with each other \citep{McVean2000-ox}.

With epistasis, mean signed LD is large for both weakly and strongly selected
variants, with sign depending on the direction of epistatic interactions
(Figure \ref{fig:epistasis}).
Synergistic epistasis (in which the effect of two mutations together is larger than
the product of each individual mutation's effect) results in negative LD while
antagonistic epistasis (in which the combined effect is less than the product of
independent effects) results in positive LD, and large positive LD can occur
even when the epistatic effect is relatively weak. Epistasis-induced LD can
extend over long distances, especially for strongly deleterious mutations. For
example, in Figure \ref{fig:epistasis}F even moderately deleterious mutations with
population-size-scaled selection coefficients of \(\gamma=-10\) show large mean
LD that extends to values of \(\rho\) much greater than 1 (for humans, assuming
roughly 1 cM/Mb, this is on the order of 100 kilobases or more).
More strongly deleterious interacting mutations are expected to show large
signed LD over much larger recombination distances.

\subsubsection{Dominance}\label{sec:dominance}

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/dominance_roze}
    \caption{
        \textbf{The effect of dominance on LD.}
        (A, B) The strengths of selection and dominance interact in a nonlinear way
        to shape expected signed LD. For weakly to moderately selected mutations,
        as shown here, signed LD can be large and negative for tightly linked loci
        (e.g., \(\gamma=-5\), \(h=0\), and \(\rho < 1\)). However, this large signed
        LD decays with recombination distance faster under a model of recessivity
        than does signed LD under a model of additive selection and
        epistasis (Figure~\ref{fig:epistasis}).
        (C, D) Recessive strongly deleterious mutations can have positive signed
        LD, as recently shown by \citet{Roze2021-cf}.
        However the dominance threshold at which \(\sigma_d^1\) switches from
        positive to negative depends on the strength of selection, and weakly
        selected mutations can show non-monotonic behavior as $h$ varies.
        \comment{C and D labels misplaced}
    }
    \label{fig:dominance}
\end{figure}

%The effects of non-additive selection within a locus and interference has received
%less attention. \citet{Garcia2020-zn} used large-scale forward simulations to explore
%the effect of dominance on patterns of LD, showing that LD depends on both the
%magnitude of the selection and dominance coefficients in a nonlinear way, but they
%note that expensive large-scale simulations are needed to explore these effects
%as no closed form approach to computing those expectations exist.
%The numerical method presented here provides such a closed form approach, so that expectations of signed LD and other two-locus statistics can be computed more
%efficiently and accurately.
\begin{itemize}
    \item Previous work on dominance and LD between linked loci
        \begin{itemize}
            \item \citep{Garcia2021-zn} - large-scale forward simulations to explore the
                effect of dominance on patterns of LD, showing that LD depends on both 
                the magnitude of the selection and dominance coefficients in a nonlinear way
            \item \citep{Roze2021-cf} -
                analytic expression for LD between pairs of recessive
                strongly deleterious mutations under steady-state demography, showing that
                for $h<0.25$, you can get slightly positive signed LD
        \end{itemize}
    \item confirming findings from \citet{Garcia2021-zn} that LD depends in a
        complicated way on the strength of both dominance and selection
    \item confirmed through discrete simulations and numerical approach that positive LD
        can occur for strong negative selection with small $h$
    \item however, Roze's result seems to hold only for strongly negative selection,
        and the $h=0.25$ threshold appears to be a limiting value for that threshold,
        as weak and moderate selection show strikingly different patterns
    \item While the model derived by \citeauthor{Roze2021-cf} showed signed LD is
        positive when $h<0.25$ and negative when $h>0.25$, this appears to 
\end{itemize}

The combined effect of the strength of selection and dominance on interference
is indeed nontrivial, and some parameter regimes can cause strong negative LD
between pairs of negatively selected variants. For strong selection, pairs of
recessive mutations (\(h<1/2\)) can have large negative signed LD (Figure
\ref{fig:dominance}B), although it decays rapidly with increasing distance between
loci and is roughly zero for \(\rho \gg 1\). For weakly selected mutations with
\(\gamma=-1\), we see that there is no monotonic effect of the level of dominance
on negative LD, with both recessive and dominant pairs of mutations having more
negative LD than the additive case.

\subsubsection{General selection and gene-based dominance}
\label{sec:general-selection}

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/gene-based-compensatory}
    \caption{
        \textbf{Multiple modes of interactions can lead to large positive signed LD.}
        Both antagonistic epistasis (A and C, and which includes compensatory mutation
        models) and gene-based dominance (B and D) lead to large positive signed LD.
        Compensatory mutations (\(\epsilon \approx -1\)) also cause increased
        \(\sigma_d^2\) compared to neutral expectations (dashed black lines),
        while weaker antagonistic epistasis does not increase \(\sigma_d^2\) above
        neutral expectations (compare to Figure~\ref{fig:epistasis}C).
        Gene-based dominance instead causes lower \(\sigma_d^2\) than neutral
        expectations.
        While signed LD may be similar between different interaction models,
        other two-locus summaries of the data may help to distinguish between
        such models.
        }
    \label{fig:gene-based-compensatory}
\end{figure}

Beyond the standard models of epistasis and dominance shown above, a large
family of selection models can be specified by assigning unique fitness effects
to each possible diploid pair of haplotypes. If we assume the diploid genotype
that is homozygous for the ancestral alleles (\(ab/ab\)) has fitness 1, then
there are nine other possible diploid two-locus genotypes that could be given
unique fitnesses (Table \ref{tab:selmodels}), noting that \(AB/ab\) and
\(Ab/aB\) genotypes can have differing selection coefficients.

The case with \(s_{AB/ab} \not= s_{Ab/aB}\) arises in a scenario where a
mutation at either locus within a haplotype impacts some functional region or
element, but a diploid individual carrying at least one copy that is free of
mutations has minimal fitness loss. In this ``gene-based dominance'' scenario
\citep[e.g.,][]{Sanjak2017-zw}, an \(AB/ab\) genotype has higher fitness than
an \(Ab/aB\) type (a simple implementation of this model is given in the final
column in Table \ref{tab:selmodels}). Such a gene-based fitness model gives
similar expected positive signed LD to the model of antagonistic epistasis
(Figure~\ref{fig:gene-based-compensatory}), although the interpretation of
those two models can differ. With a highly parameterized space of possible
general diploid selection models, multiple models with different biological
interpretations can give similar patterns of expected signed LD.

\subsection{The effect of population size changes on signed LD}
\label{sec:demography}

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/demog_bottle_expand}
    \caption{
        \textbf{The effects of demography on signed LD.}
        Simulations under two toy models, one with an instantaneous expansion in
        the past and the other with a five-fold reduction and recovery illustrated
        in (A), show the effects of population size changes on expected patterns of LD. 
        For two selection strengths (\(\gamma=-1\) and \(\gamma=-10\)) and three
        cases of interactions (non-epistatic interference (\(\epsilon=0\), B),
        synergistic epistasis (\(\epsilon=1/2\), C), and antagonistic epistasis
        (\(\epsilon=-1/2\), D), each with \(\rho=0\)),
        sudden decreases in population size can cause
        large changes in signed LD, often in the opposite direction than more
        subtle shifts due to instantaneous expansion events.
        (E) Signed LD between pairs of rare alleles (both \(n_A, n_B \leq 4\), with
        sample size \(n=50\)) is more sensitive to population size changes.
        (F) Signed LD between pairs of common alleles (both \(n_A, n_B \geq 5/50\))
        is comparatively more stable over time.
        Additional comparisons, including for dominance models and showing
        \(\sigma_d^2\), are shown in Figures~\ref{fig:toy_sd2}--\ref{fig:toy_dom_sd2}.
        Dashed lines indicate neutral expectations.
        \comment{need panel labels}
    }
    \label{fig:toy}
\end{figure}

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/demog_YRI_CEU.dominance}
    \caption{
        \textbf{Signed LD under inferred models of human population-size history.}
        Piecewise-constant population size histories inferred by Relate applied to
        \citet{1000_Genomes_Project_Consortium2015-zq} phase 3 data were used to
        simulate time series of two-locus statistics, as in Figure~\ref{fig:toy}.
        (A) The CEU (Utah residents with Northern and Western European ancestry)
        are inferred to have a stronger bottleneck than the YRI
        (Yoruba in Ibadan, Nigeria) 10--100ka, reflecting the out-of-Africa
        event.
        Here, we highlight the effect of size changes on \(\sigma_d^1\) under
        non-additive selection models with \(\rho=0\) and \(\gamma=-2\) at both loci,
        comparing standard interference (B) to site-wise dominance (C) and gene-based
        dominance (C).
        As with epistasis, more severe bottlenecks have larger effects on signed LD,
        and LD among common variants is more stable than among pairs of rare variants.
        Additional comparisons with epistasis and showing \(\sigma_d^2\) are
        in Figures~\ref{fig:relate}--\ref{fig:relate_dom_sd2}.
        Dashed lines indicate neutral expectations.
        \comment{Need panel labels}
    }
    \label{fig:relate_dom}
\end{figure}

The moment system for \(\Psi_n\) readily incorporates variable population size as
well as mutation and recombination rates and selection parameters that change
over time. Here I focus on non-equilibrium population size history and consider
scenarios relevant to human demographic history, including bottlenecks and
expansions. I explore two simple models (Figure \ref{fig:toy}A), one with an
instantaneous expansion and another with a bottleneck followed by recovery. I
also consider two demographic histories inferred using genome-wide gene
genealogy reconstruction \citep{Speidel2019-nj} applied to the
\citet{1000_Genomes_Project_Consortium2015-zq} dataset, and focus on size histories
for the Yoruba from Ibidan, Nigeria (YRI) and Utahns of North and West European
ancestry (CEU) (Figure \ref{fig:relate_dom}B).

For each of the four histories,
Figures~\ref{fig:toy},~\ref{fig:relate_dom},~and~\ref{fig:toy_sd2}--\ref{fig:relate_dom_sd2}
show the dynamics of \(\sigma_d^1\) and \(\sigma_d^2\) for a given
parameterization of two-locus selection, including synergistic and antagonistic
epistasis, dominance within loci, and gene-based dominance. In general across
each selection model, population size expansions are not expected to strongly
affect \(\sigma_d^1\), whether that expansion occurs deeper in the past as in
the simple expansion model or rapid expansion more recently, as in the YRI. On
the other hand, population size reductions tend to push signed LD to more
extreme values and subsequent recoveries or expansion again reduce the
magnitude of LD. Under no selection condition tested here do population size
changes cause expected LD to change sign, showing that while the magnitude of
deviation of LD from zero is sensitive to population size history, interpreting
general patterns of the observed sign of LD in data should not be strongly
affected by population size history.

\subsection{Signed LD within protein-coding genes}

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/data_compact}
    \caption{
        \textbf{LD in human protein-coding genes and annotated domains.}
        \comment{Needs panel labels}
    }
    \label{fig:data}
\end{figure}

%\caption{\textbf{LD within and between protein-coding genes.} FIXME (A-B) Pairs of
%variants in coding regions partitioned by classes of mutations.
%Loss-of-function mutations are defined as stop and start gains and losses and
%other nonsense mutations of high impact. In all populations, we observe
%positive mean signed LD for missense and synonymous variants, although the
%confidence interval of each measurement overlaps with zero. Similarly, we
%observe mean negative LD for pairs of loss-of-function variants within genes.
%There are only \(\sim200-300\) pairs of loss-of-function variants that fall
%within the same gene across the populations tested here, causing large
%uncertainty in each measurement. Gray lines indicate 2 standard errors of the
%measurement, computed by bootstrapping of sets of genes. (C-F) While pairs of
%synonymous and missense mutations within genes are consistently positive, pairs
%between genes are not consistent in their sign across populations. LD for all
%populations analyzed here is shown in Figure \ref{fig:genespops}.
%\textbf{LD within and between annotated domains in protein-coding
%genes.} (A) Considering only variants that fall within annotated domains in
%protein-coding genes, pairs of mutations that fall within the same gene display
%larger positive LD than pairs that lie in different domains. This pattern is
%most striking for missense mutations, for which pairs within domains have
%average signed LD that is consistently larger than for pairs that span two
%domains, or for pairs of synonymous mutations that are within the same domain.
%(B-D) The slightly positive LD that we observe for missense and synonymous
%mutations gene-wide (Figure \ref{fig:fig3}A) is largely driven by the subset
%of pairs mutations that fall within the same annotated domain. LD for all
%populations analyzed here isshown in Figure \ref{fig:domainspops}.}\label{fig:LDdata}
%\end{figure}

Synonymous and missense mutations show similar levels of positive signed LD
when considering pairs of mutations within the same gene. Loss-of-function
mutations have more negative LD, possibly suggesting differing modes of
selective interactions for loss-of-function and missense mutations (Figure
\ref{fig:data}A--B). Within each population, measurement noise gives 95\%
confidence intervals that overlap with zero in each mutation class, although
the pattern is remarkably consistent across African, European, and East Asian
populations in the Thousand Genomes dataset. Comparing mean LD within genes in
African and Eurasian populations, LD in Eurasian populations is somewhat larger
on average, that is, more positive for missense mutations and more negative for
loss-of-function mutations. This is in agreement with differences in
expectations between populations that have or have not gone through a
bottleneck in their recent past
(Figures~\ref{fig:toy}~and~\ref{fig:relate_dom}).

\subsubsection{Positive LD between pairs of missense mutations in conserved domains}

The similarity in signed LD between missense and synonymous mutations would
suggest that interference between missense mutations is minimal. However,
interactive effects differ dramatically between pairs of mutations found in
intra-genic regions, with these opposing effects canceling out when taking
gene-wide averages. Due to the rarity of loss-of-function mutations, I only
compare synonymous and missense mutations when looking at finer partitions of
mutations within genes.

Here, I examine the role that annotated conserved domains in protein-coding
genes play in driving signals of positive LD within genes. Protein-coding
domains are conserved elements of genes, often associated with some known
functional or structural feature of a protein \citep{Stanek2020-pa}. Purifying
selection is expected to be stronger within conserved domains than within the
same gene but outside of those domains. Indeed, the site-frequency spectrum
(SFS) is skewed to lower frequencies for both missense and loss-of-function
mutations within domains when compared to the same classes of mutations outside
of domains, with much more negative values of Tajima's D within domains (Table
\ref{tab:tajimasD}). On the other hand, no difference is observed for
synonymous mutations whether within or outside domains, suggesting roughly
equivalent effects of selection (either direct or linked) on synonymous
variation.

Pairs of missense mutations that both fall within the same functional domain
have large positive LD that is much larger than pairs of synonymous mutations
that both fall within the same domain
(Figures~\ref{fig:data}C--F~and~\ref{fig:LDwithin}--\ref{fig:domainsEAS}). This
difference between pairs of missense and synonymous variants within domains is
observed for both tightly linked (\(\lesssim 100\)bps) and more distant
(\(\gtrsim 1\)kb) pairs (Figure \ref{fig:LDwithin}). Assuming most
synonymous mutations are neutral, we would expect signed LD between missense
mutations to be less than that between synonymous mutations under predictions
from models of either Hill-Robertson interference, synergistic epistasis, or
dominance (Figures~\ref{fig:epistasis}~and~\ref{fig:dominance}). This
observation is opposite to those expectations, suggesting a different
prevailing interactive effect between nonsynonymous mutations within domains
than those that are commonly tested or modeled.
\comment{Due to the observed differences in levels of selection inside and outside domains, frequency matching - result holds}

The large positive signal of LD for missense mutations within the same
domain does not extend to pairs of missense mutations that are in
different domains. Pairs of missense and synonymous mutations show
nearly equal levels of LD close to zero across domains, with missense mutations
even slightly more negative than synonymous mutations (Figure \ref{fig:XXX}).
\comment{need to do this analysis again}
The interactive effect driving large LD within domains is therefore likely
domain-specific. However, the average distance between pairs of mutations
within domains is much smaller than between domains, so this observation may be
primarily driven by the higher recombination distances between pairs of mutations
across distinct domains.

\subsubsection{Reduced LD between pairs of missense mutations outside of conserved domains}

Pairs of mutations that both fall outside of annotated domains have the
opposite pattern of signed LD to pairs of mutations falling within the same
domain. For pairs of mutations outside of domains but with distances matched to
those within domains, synonymous mutations have larger positive LD than
missense mutations. More distant pairs of mutations outside of domains, matched
to the same distances as the between-domain comparison, each have LD roughly
equal to zero. \comment{Todo}

The role that tightly linked variants have in driving these opposing signals
can be seen in the decay of signed LD with distance between mutations
(Figures~\ref{fig:LDgene}--\ref{fig:LDoutside}). Both synonymous and missense
mutation pairs at distances greater than a few hundred base pairs have average
LD that fluctuates around zero. However, for mutations outside domains,
synonymous variants separated by short distances have large positive LD, while
missense mutations have much lower LD (Figure~\ref{fig:LDoutside}). In contrast,
for mutations within the same domains, missense mutations have more positive
LD at short distances than synonymous mutations (Figure~\ref{fig:LDwithin}).

\section{Discussion}\label{sec:discussion}

Previous theoretical and simulation studies have shown that interference and
interactions between selected mutations reduce the efficacy of selection at
linked loci, impacting substitution rates, the deleterious mutation load, and
dynamics of segregating mutations
\citep{Hill1968-vu,Birky1988-jm,Barton1995-mj,McVean2000-ox}. Interference and
synergistic epistasis between moderately deleterious mutations are both
expected to cause negative LD between selected mutations, which can be readily
tested using population genetic data
\citep{Sohail2017-zq,Sandler2021-of,Garcia2021-zn}. Here, I used a closed
numerical approach to generate expectations for LD under a wide range of
selective scenarios, and then compared patterns of LD in human populations
between classes of coding mutations using unbiased estimators for LD from
unphased genotypes.

By taking broad genome- or gene-wide surveys of LD across functional classes of
mutations, heterogeneous patterns of interactive effects that occur within
genes can be missed. From gene-wide averages, LD between pairs of missense
mutations does not appear to be different from pairs of synonymous variants,
although LD between loss-of-function variants is more negative. Hill-Robertson
effects under additive selection are expected to be strongest for slightly to
moderately deleterious variants (with \(|s|\sim 1/N_e\)), as strongly deleterious
mutations are not expected to interfere with one another \citep{McVean2000-ox}.
However, loss-of-function mutations do not typically fall within this regime,
and inference of the distribution of fitness effects (DFE) for new
loss-of-function variants shows that a large majority are strongly deleterious
(Supporting Information). Instead, negative synergistic epistasis between
strongly deleterious mutations does produce large negative deviations of mean
LD. Strongly deleterious \emph{recessive} mutations can also produce this
pattern, making it difficult to distinguish these effects with certainty. This
difficulty is compounded by the large measurement noise for \(\E[D]\),
especially for loss-of-function variants for which only a few hundred
within-gene pairs exist in the human population data analyzed here.

\subsection{Non-uniform interactions between selected mutations within genes}
\label{non-uniform-interactions-between-selected-mutations-within-genes}

Positive average LD between both missense and synonymous mutations has
previously been reported in humans and other species \citep{Sohail2017-zq,
Garcia2020-zn, Sandler2021-of}. The similarity of their gene-wide LD
(Figure~\ref{fig:data}) might suggest that interference or interactions between
missense mutations are minimal, or at least no stronger than those between
synonymous mutations. However, averaging over all observed pairs of mutations
within a gene masks element-specific interactive effects that drive LD in
opposite directions. Nonsynonymous mutations found within conserved protein
domains, identified as conserved subunits of a gene with some structural or
functional role \citep{Stanek2020-pa}, are more strongly selected against on
average, but also have increased signed LD over synonymous mutations at the
same distances within domains. Missense mutation outside of domains but at the
same distances as those within domains have more negative LD, both compared to
distance-matched synonymous mutations outside of domains and to mutations
within domains. Neither dominance effects, synergistic epistasis, nor
Hill-Robertson interference are expected to result in positive LD, so some
other interactive effect should be driving this positive LD signal within
conserved domains.

There are a number of possible interaction scenarios that can result in
positive signed LD between tightly linked loci. One explanation is a prevalence
of pairs of compensatory mutations that are tolerated to co-segregate at high
frequencies within conserved domains \citep{Yeang2007-gj,Ivankov2014-tn}.
\citet{Callahan2011-ac} and \citet{Taverner2020-lk} have proposed such a
mechanism to explain observed clusters of nonsynonymous mutations in lineages
of \emph{Drosophila} and other species. Another possibility is a model of
antagonistic, or diminishing returns epistasis, in which a single amino
acid-changing mutation within a domain damages the functionality of that
subunit, but additional mutations within that same domain reduce fitness by a
factor less than the first mutation. A third possibility, related to the
antagonistic epistasis case, is that selection acts on the functional domain as
a unit instead of at loci within the domain individually (such as under a model
of gene-based dominance). In this scenario, double heterozygotes have different
fitnesses depending on whether the mutations are found on same haplotype or on
different haplotypes.

We do not see an increase in LD between mutations that are found in different
annotated domains within the same gene, which are on average at much larger
recombination distances. Nor do we find increased LD outside of domains.
Rather, for pairs of mutations outside of domains but matched to the same
distance as those within the same domain, missense variants have considerably
lower LD than synonymous variants. This difference between synonymous and
missense pairs of variants largely disappears for SNPs separated by more than a
few hundred base pairs (Figure \ref{fig:lddecay}). This suggests that
Hill-Robertson interference is the primary mode of interaction between
mutations falling outside of domains, as epistasis is expected to impact LD
over larger distances than what is observed. Importantly, however, the strength
of epistasis is also likely to be a function of distance between mutations.

Taken together, selection on segregating variants within protein-coding genes
is nonuniform, with both the overall strength of selection and interactions
between variants differing between annotated elements. Missense and
loss-of-function mutations within conserved domains are subject to stronger
selection, skewing the SFS to lower frequencies. Typical approaches for
inferring the distribution of fitness effects from population genomic data
average over these differences by considering all nonsynonymous mutations
together \citep{Boyko2008-zk,Kim2017-xo}, but if loss-of-function variants are in
fact recessive, even partially, this can impact inferences of the DFE
(Supporting Information). It would be straightforward to adapt DFE-inference
methods to infer more detailed and accurate representation of the heterogeneous
effects of new mutations within genes by partitioning by missense and
loss-of-function classes as well as by annotated domains. Similarly, the
different modes of interaction between mutations across annotated domains mean
that our standard models and simulation approaches may be too simple and will
not capture the evolutionary trajectories and patterns of diversity that differ
at finer scales.

\subsection{Clustered mutations account for positive LD among synonymous mutations}
\label{clustered-mutations-account-for-positive-ld-among-synonymous-mutations}

In a non-structured randomly mating population, neutral mutations are expected to have
average signed LD of zero, but across all populations analyzed here, LD between
synonymous mutations is positive. While selection on some subset of synonymous
variants is possible, it should be much weaker on average than between missense
mutations, and any interference between selected synonymous variants should
lead to negative LD.

Spatial population structure may be responsible for the increase of LD observed
between synonymous variants, as every natural population lives in some
geography that will necessarily induce some structure. Neutral evolution in
structured populations alone cannot create this effect. Under standard
population genetic assumptions of constant mutation and recombination rates and
independent mutation events, migration, admixture, and spatial structure do not
result in nonzero \(\E[D]\). Rather, for a set of mutations with \emph{given}
differences in allele frequencies between source populations, \(D\) can be
nonzero in an admixed population between those particular mutations even when
\(D\) is zero in the source populations \citep{Cavalli-Sforza1971-jv}. This
result relies on conditioning on directional differences in allele frequencies
between populations. The expected difference in mean allele frequency between
source populations is zero (\(\E[(p_1-p_2)(q_1-q_2)] = 0\)), as differences are
equally likely to be positive or negative, and resulting \(\E[D]\) is still
zero when taking genome-wide averages. Instead, \citet{Sohail2017-zq} and
\citet{Sandler2021-of} used forward simulations under multi-population or
explicitly spatial models and a multiplicative fitness function to explore the
joint effect of structure, assortative mating, and interference, scenarios that
can lead to nonzero LD. For example, \citet{Sandler2021-of} found positive LD
among neutral mutations in a simulation with admixture and linked selected
mutations, showing that the combined effects of spatial structure and background
selection can lead to positive average LD between neutral mutations.
\comment{Because the observed positive LD decays rapidly in the data, this
would be worth exploring in more detail with forward simulation approaches}
\comment{split into two paragraphs}

Nonrandom mutational processes provide an alternative explanation for positive
LD between neutral variants separated by short distances, in which clusters of
mutations occur simultaneously in the same mutational event. Such
multinucleotide mutations have been shown to affect patterns of LD over short
distances, on the order of 10s to 100s of base pairs \citep{Harris2014-zg}, and
clustered mutational events may be common in humans \citep{Besenbacher2016-ac}.
Indeed, a simple exponential model in which the fraction of pairs of mutations
arising from a multinucleotide mutation event decays with distance fits the
observed decay of \(\sigma_d^1\) between synonymous variants (Supporting
Information). \comment{results of this here?}
We may therefore treat positive
LD as the baseline expectation for tightly linked variants primarily due to
clustered mutations, so that subsequent selective processes and interactions
cause LD to deviate from that expectation, as seen for missense mutations
across different sub-elements of genes. It may also be more appropriate to
compare the negative LD observed between linked loss-of-function variants to
that positive baseline expectation instead of zero, which would imply that they
are more recessive or have stronger synergistic epistasis than from inferences
assuming a neutral expectation of zero.

\subsection{Challenges to distinguishing highly parameterized models of selective interactions}

When partitioning measurements of LD by mutation classes or regions within
genes, the decreasing number of pairwise comparisons leads to large estimated
measurement noise. Within each population, confidence intervals of observed
\(\sigma_d^1\) often overlap with zero or overlap with that of other classes of
mutations. While observed patterns are remarkably consistent across the 15
populations considered here, their joint evolutionary histories make formal
testing of significance difficult due to shared variation, so they cannot be
treated as independent measurements. Extensive forward simulations will likely
be required to more thoroughly assess significance.

From a modeling perspective, in exploring or inferring multi-locus selective
models, the number of plausible selection scenarios becomes quite large as we
relax the strict asumptions of additivity and multiplicative fitness effects.
The inclusion of dominance effects, epistasis, or other interactions leads to a
rapid increase in the number of parameters to consider. This makes performing
forward simulations that span the range of all such selective interaction
scenarios burdensome.

Instead, closed, numerical approaches to compute expectations for two-locus
statistics under a wide range of selection models, as presented here or in
\citet{Ragsdale2019-nt} and \citet{Friedlander2022-bs}, allow us to explore
this highly parameterized space of models far more efficiently, and it opens
the possibility for performing likelihood-based inference using signed LD or
other two-locus summaries of the data. For example, inferring the joint
distribution of dominance and selection is underpowered using the SFS alone,
but because signed LD is sensitive to the levels of dominance (Figure
\ref{fig:dominance}), inferring the DFE with dominance may be feasible using
the joint distribution of allele frequencies and LD.

Using a single low-order summary of signed LD, such as \(\E[D]\) or
\(\sigma_d^1\), is likely insufficient to confidently discriminate modes of
selective interactions between linked mutations. Both dominance and epistasis
can result in strongly negative LD, although the expected decay of LD under
those scenarios can differ. Similarly, multiple interaction models can lead to
positive signed LD. Among all interaction models, the extent of LD and rate of
its decay also depends on the underlying distribution of selection coefficients
among a class of mutations, which are unknown for a given pair of mutations, so
that we must integrate over a distribution of fitness effects. This DFE,
however, will have been inferred under a simple set of assumptions, such as
additivity and interchangeability between sites within a gene, potentially
biasing any inference using previously inferred DFEs to learn about patterns of
interactions. Again, this underlies the need to jointly infer strengths and
interactions of selected variants.

Finally, exploring additional summaries of the two-locus sampling distribution
could provide additional power to distinguish between interaction models. The
method developed here allows us to compute the full sampling distribution for a
given sample size, and expectations for a large family of informative two-locus
statistics can be computed directly from this distribution. These expectations
can then be compared directly to empirical observations, which can be taken
from either phased or unphased data \citep{Ragsdale2020-nz}. Additional work
performing such an analysis is therefore warranted, which should provide a path
forward for distinguishing between modes of selective interactions.

\section{Methods}\label{sec:methods}

\subsection{Existing theory and numerical methods}
\label{existing-theory-and-numerical-methods}

Many well-known properties of two-locus dynamics and equilibrium LD come from
early work on the multi-locus diffusion approximation \citep{Kimura1955-qe, Hill1968-vu, Ohta1969-ie, Ohta1971-yd}.
This includes the result that
genome-wide averages of signed LD are expected to be zero under neutrality.
Under a two-locus biallelic model, where the left locus allows alleles \(A\) and
\(a\) and the right locus allows alleles \(B\) and \(b\), the standard covariance
measure of LD is defined as \(D = f_{AB} - f_{A}f_{B}\), where \(f_{AB}\) is the
haplotype frequency of double-derived types carrying both \(A\) and \(B\), and
\(f_{A}\) and \(f_{B}\) are the marginal frequencies of the derived alleles at each
locus. This covariance decays due to both drift and recombination at a rate
proportional to the inverse of the effective population size and the distance
separating loci \citep{Hill1968-vu}: \[\E[D]_{t+1} = \left(1 - \frac{1}{2N_e(t)} - r
\right)\E[D]_t.\] While \(\E[D] = 0\), the variance of \(D\) is non-zero, and
\citet{Ohta1971-yd} presented their groundbreaking result that the variance of \(D\)
under neutrality and steady-state demography, normalized by the joint
heterozygosity of the two loci, is \begin{equation} \label{eq:ohta} \sigma_d^2 =
\frac{\E[D^2]}{\E[f_A(1-f_A)f_B(1-f_B)]}\approx\frac{5 + \rho / 2}{11 + 13\rho
/ 2 + \rho^2 / 2}, \end{equation} where \(\rho = 4N_e r\).

Analytic progress beyond these results has come haltingly. In the 1980s,
recursions were developed to compute the two-locus sampling distribution under
neutrality, that is, the probability of observing given counts of two-locus
haplotypes in a sample of size \(n\) \citep{Golding1984-pu}. This approach would
continue to be developed and later form the foundation for the inference of
local recombination rates from population genetic data \citep{Hudson2001-sg, McVean2004-gj}. More recently, \citet{Song2007-qk} computed \(\E[r^2]\) using a
diffusion approximation approach, although their solution involves the
summation of infinitely many terms and is restricted to neutrality and
steady-state demography.

To include selection, however, there have been relatively few advances beyond
the Monte Carlo simulation approach taken by \citet{Hill1966-gv}, albeit now with more
powerful computational resources and sophisticated software for performing
flexible forward simulation (e.g., \citet{Thornton2014-pn}, \citet{Haller2019-vm}). Analytic
results for two-locus distributions under selection are notoriously difficult,
with a few notable flashes of progress. For example, \citet{McVean2007-xd} considered
the effect of a recent sweep on patterns of LD between neutral loci near the
locus under selection, and in a recent paper, \citet{Good2020-wj} presented analytic
solutions for patterns of LD between rare mutations under additive selection
with epistasis. Nonetheless, such approaches are typically confined to
steady-state demography and constrained selection models.

Numerical methods inhabit the space in between expensive discrete simulations
and limited analytic solutions, providing a more efficient and practical method
to compute expectations of two-locus diversity measures under a wider range of
parameters and demographic scenarios. \citet{Ragsdale2017-gg} used a finite
differencing approach to numerically solve the two-locus diffusion equation
with additive selection at either locus, although that paper focused on the
applicability of two-locus statistics to demographic inference. More recently,
\citet{Ragsdale2019-nt} extended the \citet{Hill1968-vu} system for \(\E[D^2]\) to compute
arbitrary moments of the distribution of \(D\) for any number of populations
connected by migration and admixture. They also showed that such a
moments-construction can be used to solve for the two-locus sampling
distribution within a single population, though it requires a moment-closure
approximation for nonzero recombination and selection. Below, I extend this
approach to model arbitrary diploid selection, which encompasses dominance,
epistasis, and other forms of selective interactions between two loci. In a
concurrent study to this paper, \citet{Friedlander2021-bs} developed numerical
solutions to the same moments system, which they used to describe selected
haplotype trajectories and the distortion of neutral diversity at loci variably
linked to beneficial alleles that sweep to high frequencies under
non-equilibrium demography.

\subsection{The two-locus sampling distribution with arbitrary selection}
\label{the-two-locus-sampling-distribution-with-arbitrary-selection}

The two-locus sampling distribution is the direct analog to the single-locus
site frequency spectrum (SFS) of a given sample size. Instead of describing the
density or number of mutations with a given allele count out of \(n\) samples,
the two-locus distribution \(\Psi_n\) stores the density or number of pairs of
loci with observed haplotype counts, so that \(\Psi_n(i, j, k)\) is the number of
pairs for which we observe \(i\) copies of the \(AB\) haplotype, \(j\) of type \(Ab\),
\(k\) of type \(aB\), and \(n-i-j-k\) of type \(ab\). The size of \(\Psi_n\) grows
rapidly, with \(O(n^3)\) entries, which practically limits computational
approaches to moderate sample sizes and a single population.

Under neutrality, a number of approaches exist to compute \(\Psi_n\), including
the recursion due to \citet{Golding1984-pu} and \citet{Ethier1990-tb}, or more recent
numerical approaches to the two-locus coalescent \citep{Kamm2016-ag} or diffusion
approximation \citep{Ragsdale2017-gg}. Selection is most easily included using the
forward-in-time diffusion equation \citep{Kimura1955-qe, Hill1966-gv}, where a
standard approach is to first solve for the continuous distribution \(\psi\) of
the density of two-locus haplotype configurations in the full population, and
then integrate \(\psi\) against the multinomial sampling function to obtain
\(\Psi_n\).

Alternatively, \citet{Ragsdale2019-nt} showed that there exists a system of ordinary
differential equations directly on the entries of \(\Psi_n\), although that paper
focused primarily on neutral dynamics and considered a model with selection at
only one of the two loci. I briefly summarize this general approach below, but
refer readers to that previous paper for detailed derivations of the drift,
recombination, and mutation terms and the moment-closure approximation.
Instead, here I focus on generalizing the selection operator to include
epistasis, dominance, and other forms of two-locus interactions.

\subsubsection{Moment equation for \(\Psi_n\)}\label{moment-equation-for-psi_n}

The system of linear ordinary differential equations for the entries of
\(\Psi_n\) takes the form
\begin{equation}
\label{eq:system}
{\Psi}_n^{t+1}(i, j, k; t) =
\mathcal{D}_{N(t)}\Psi_n^t
+ \mathcal{R}_{r}\Psi_{n+1}^t
+ \mathcal{U}_{u}\Psi_n^t
+ \mathcal{S}_{s_A, s_B, \ldots, h_A, h_B, \ldots}\Psi_{n+2}^t.
\end{equation}
Here, \(\mathcal{D}_{N(t)}\) is a sparse linear operator accounting for drift
with population size \(N(t)\), \(\mathcal{R}\) accounts for recombination
with per-generation recombination probability \(r\) between the two loci,
\(\mathcal{U}\) accounts for mutation, either under an infinite sites or
biallelic reversible mutation model, and \(\mathcal{S}\) accounts for
selection.

The moment system for \(\Psi_n\) can be derived directly from the diffusion
approximation, or it can be found through a more intuitive process of tracking
the dynamics of allelic states of a sample of size of \(n\) from the full
population. We assume \(n \ll N_e\), and \(r\) and \(s\) are \(O(1/N_e)\) so that
multiple coalescence, recombination, or selective events within the \(n\)
lineages are rare in any given generation. For example, for drift events, we
account for the probability that two lineages within the \(n\) tracked lineages
at time \(t+1\) coalesced in the previous generation and, given a coalescence
event, the probability that haplotype counts changed due to this coalescence
event {[}see Supporting Information; \citet{Jouganous2017-pq}; \citet{Ragsdale2019-nt}{]}. In
typical diffusion approximation fashion, we multiple through by \(2N_{ref}\) so
that time is measured in \(2N_{ref}\) generations, and we consider scaled
parameters \(\rho = 4Nr\), \(\theta = 4Nu\), and \(\gamma=2Ns\).

\subsubsection{Moment closure}\label{moment-closure}

In the absence of selection and for fully linked loci (i.e., \(\rho=0\)), the
system is closed and can be solved exactly. However, for nonzero recombination
or selection, the entries of \(\Psi_n\) rely on the slightly larger sampling
distributions with sample sizes \(n+1\) (for recombination and additive
selection) or \(n+2\) (for non-additive selection). This is because if a
recombination event occurs within one of \(n\) lineages being tracked by
\(\Psi_n\), we need to draw an additional lineage from the full population to
recombine with that chosen lineage, thus requiring \(\Psi_{n+1}^t\) to find
\(\Psi_n^{t+1}\). Selection events similarly require extra lineages from the full
population, which replace a chosen lineage that fails to reproduce with
probability proportional to its selection coefficient, assuming negative
selection.

This requirement of extra lineages for nonzero recombination and selection
means that the system in \eqref{eq:system} is not closed, so that we need a
moment-closure approximation to solve for \(\Psi_n\). As in \citet{Ragsdale2019-nt}, a
jackknife approximation is used to estimate \(\Psi_{n+l}\), for \(l=1\) or \(2\),
from \(\Psi_n\) \citep[following the single-locus closure introduced in][]{Jouganous2017-pq}, so that \(\hat{\Psi}_{n+l}(i, j, k) = \mathcal{J}_{n, l}\Psi_n\), although other accurate closure approximations are possible
\citep{Friedlander2021-bs}. This emits a closed approximate system,
\begin{equation}
\label{eq:closed-system}
\dot{\hat\Psi}_n(i, j, k; t) =
\mathcal{D}_{\nu(t)}\hat\Psi_n(t)
+ \mathcal{R}_{\rho}\mathcal{J}_{n, 1}\hat\Psi_{n}(t)
+ \mathcal{U}_{\theta}\hat\Psi_n(t)
+ \mathcal{S}_{\gamma, h}\mathcal{J}_{n, 2}\hat\Psi_{n}(t).
\end{equation}

The jackknife approximation, which approximates an entry \(\Psi_{n+l}(i,j,k)\)
using nearby entries in \(\Psi_n\), is more accurate for larger sample sizes,
creating a tension between efficiency and accuracy: larger sample sizes result
in more accurate solutions, as error in the jackknife is diminished, but
computational complexity also grows rapidly in the number of entries of
\(\Psi_n\), which is \(O(n^3)\). Accuracy and runtime are shown in Figure
\ref{fig:jackknife}. In the results presented in this paper, sample sizes
between \(n=30\) and \(n=80\) are used. Derivations for the drift, recombination,
and mutation operators and the jackknife moment-closure approximation can be
found in section S1.3 of \citet{Ragsdale2019-nt}, and I repeat the main results in the
Supporting Information of this paper.

\subsubsection{Selection models with epistasis and dominance}
\label{selection-models-with-epistasis-and-dominance}

To include selection, we consider a model where we draw lineages uniformly from
the previous generation, but keep lineages with probability proportional to
their fitness. In the absence of dominance, selection reduces to a haploid
model, with acceptance and rejection probabilities depending on the fitnesses
of each haploid copy, where haplotype \(Ab\) has fitness \(1 + s_{A}\), \(aB\) has
fitness \(1 + s_{B}\), and \(AB\) has fitness \(1 + s_{AB}\). We assume the doubly
ancestral haplotype \(ab\) has fitness \(1\), so fitnesses are relative to that of
\(ab\) haplotypes. The standard multiplicative fitness function assumes that
\(s_{AB} \approx s_{A} + s_{B}\), and a model for epistasis can be written as
\[s_{AB} = (s_{A} + s_{B}) (1 + \epsilon),\] so that \(\epsilon > 0\) implies
synergistic epistasis, while \(\epsilon < 0\) implies antagonistic epistasis.

To obtain the recursion equation under selection we consider drawing \(n\)
lineages from generation \(t\), which has an expected sampling distribution of
haplotype counts given by \(\Psi_n^t\). However, assuming \(s\leq0\) for each
derived haplotype, each of those sampled lineages has probability of being
rejected equal to the selection coefficient assigned to its haplotype state. If
a lineage is rejected, a replacement is drawn from the full population. Under
the assumption that \(ns_{\cdot\cdot} \ll 1\), the probability that more than one
selection event occurs in any given generation is negligibly small, so that the
case of multiple simultaneous rejections can be ignored. Then \(\Psi_n^{t+1}\)
relies only on \(\Psi_n^t\) and \(\Psi_{n+1}^t\) for additive selection. The full
selection operator \(\mathcal{S}\) for additive selection is given in the
Supporting Information.

To account for dominance within loci, or other general forms of two-locus
selection, the selection operator no longer reduces to individual haplotypes,
but instead we need to know the state of two-locus genotypes. For example, the
fitness of an individual carrying an \(Ab\) haplotype depends on whether their
second haplotype is \(ab\), \(Ab\), \(aB\), or \(AB\). We can therefore assign a
selection coefficient to each possible diploid configuration, \(s_{Ab/ab}\),
\(s_{Ab/Ab}\), and so on. Assuming that the doubly homozygous ancestral \(ab/ab\)
genotype has relative fitness 1, this gives nine possible unique selection
coefficients in the most general two-locus selection model. Note that \(AB/ab\)
and \(Ab/aB\) genotypes need not have the same selection coefficient, which
allows for simulation under a gene-based model of dominance (Table
\ref{tab:selmodels}).

The general selection operator follows the same approach as the haploid
selection operator with epistasis described above. Now, in the case of a
selection event rejecting a lineage within our tracked samples, we need to draw
not only the replacement lineage from the full population but also a second
haplotype from the full population to form the diploid genotype, as this
determines the probability that we reject the focal haplotype. We again assume
that \(ns \ll 1\) for all genotype selection coefficients, so that we may assume
at most a single selection event occurs in any given generation. This means
that to find \(\Psi_n^{t+1}\) under a general two-locus selection model, which
encompasses dominance within either locus, gene-based dominance, or a
combination or dominance and epistatic effects, we need \(\Psi_{n+2}^t\). Again,
a full derivation and expressions for the general selection operator are given
in the Supporting Information.

\subsubsection{Low-order summaries of the sampling distribution}
\label{low-order-summaries-of-the-sampling-distribution}

From \(\Psi_n\), expectations for any two-locus statistic can be found by
downsampling to the appropriate sample size. For example, to compute \(\E[D]\),
the sum is taken over all haplotype configurations \(\mathbf{n} = (n_{AB}, n_{Ab}, n_{aB}, n_{ab})\), weighted by the density \(\Psi_n\) for that
configuration:
\begin{equation}
\E[D] = \sum_{\mathbf{n}} \Psi_n(\mathbf{n})
\frac{n_{AB}n_{ab} - n_{Ab}n_{aB}}{n(n-1)}.
\end{equation}
For large sample sizes, this is approximately equal to computing \(D\) by taking
the maximum likelihood estimate for each allele frequency \(f_i = n_i / n\), but
the maximum likelihood-based estimate will be noticeably biased for small to
moderate sample sizes. Other low-order two-locus statistics can be computed
using the same approach, as implemented in \emph{moments} following
\citet{Ragsdale2020-nz}, which can be compared across sample sizes and between
estimates from phased or unphased data. In this paper, I focus on \(\sigma_d^2 = \E[D^2] / \E[p(1-p)q(1-q)]\) and \(\sigma_d^1 = \E[D] / \E[p(1-p)q(1-q)]\),
averaged over pairs of variants at all frequencies. However, allele-frequency
conditioned statistics (such as keeping only loci below some frequency
threshold as in \citet{Good2020-wj}) can be considered using this same approach.

\subsubsection{Simulations of non-steady-state demography}
\label{simulations-of-non-steady-state-demography}

I considered four variable population size histories, two simple toy models and
two inferred from human populations in African and Europe. For each size
history scenario, I tracked the evolution of \(\Psi_n(t)\) for varying selection
models, and Figure \ref{fig:fig3} plots the trajectory of \(\sigma_d^1\) over
time. The selection strength at both loci was fixed at \(\gamma=-2\) for all
scenarios, and I considered additive selection without epistasis for fully
linked (\(\rho=0\)) and more loosely linked (\(\rho=5\)) loci. With \(\rho=1\), I
also considered positive and negative epistasis (\(\epsilon=0.5\) and \(-0.5\)),
recessive loci (\(h=0.1\)), and a gene-based dominance model with \(h=0.1\).

The simple size change models both had ancestral \(N_e=10,000\), with one a
\(3\)-fold population expansion that occurs 3,000 generations ago, and the other
a \(10\)-fold reduction 2,000 generations ago followed by an expansion to double
the ancestral size \(1,000\) generations ago. The size histories for YRI and CEU
were inferred using Relate \citep{Speidel2019-nj} applied to the phased autosomal
data from \citet{1000_Genomes_Project_Consortium2015-zq}, using default parameters as
recommended in the Relate tutorial. Relate returns estimates of coalescence
rates within specified time bins, and population sizes are estimated as their
inverses. Estimates using Relate for population sizes in the very recent past
(\(<3,000\) years, or \(\approx 100\) generations) diverged, so I truncated the
history over this time period and assumed a constant size from the most recent
non-diverged bin.

\subsection{Analysis of human genomic data}
\label{analysis-of-human-genomic-data}

Using the annotated variant call format (VCF) files from the phase 3
\citet{1000_Genomes_Project_Consortium2015-zq} (Thousand Genomes) data release, I
subset the genotype VCFs of the same release to autosomal variants that were
annotated as either synonymous or nonsynonymous (including both missense and
more damaging mutations such as stop gains, stop losses, and start losses). I
further subset to samples within each non-admixed population in the African,
European, and East Asian continental groups (Table \ref{tab:1kgpops}). Signed
LD is sensitive to ancestral state misidentification, so I only kept sites for
which ancestral alleles were estimated with high confidence in both the VCF
info field and the Thousand Genomes human ancestor reconstructed from a
phylogeny of six primates.

In addition to ancestral state misidentification, measured LD is sensitive to
phasing error, so I computed LD statistics using unphased genotypes following
\citet{Ragsdale2020-nz}. This approach provides unbiased estimates for pairwise LD,
under the assumption that individuals are not inbred. I considered pairs of
mutations within the same mutation class (synonymous, missense, and
loss-of-function) either within the same gene, in different genes within
distance bins up to 500kb, and inside or outside of annotated domains within
the same protein-coding genes. I used a dataset of annotated protein domains
mapped to the hg19 human reference build compiled by \citet{Stanek2020-pa} to
determine if a given mutation falls within a domain or not.

\subsection{Data and software availability}
\label{data-and-software-availability}

All data and software used in this paper are publicly available and open source.
I downloaded the Thousand Genomes annotations and genotypes VCFs from the ftp
server at \url{ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/}, and
the Thousand Genomes human ancestor fasta file from
\url{ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase1/analysis_results/supporting/ancestral_alignments/}.
Protein domain information from \citet{Stanek2020-pa} was downloaded from
\url{http://prot2hg.com/dbdownload.php}.

Implementation of moment equations to compute expectations for two-locus and
linkage disequilibrium statistics are implemented in Python using Numpy
\citep{Harris2020-pc} and sparse matrix solvers in Scipy \citep{Virtanen2020-kr}. These
methods are packaged within \emph{moments}, and analyses here were performed using
\emph{moments} version 1.1.0, available from
\url{https://bitbucket.org/simongravel/moments/src/master/} with extensive
documentation at \url{https://moments.readthedocs.org}. Scripts to run all
analyses, recreate figures, and compile this manuscript are available at
\url{https://github.com/apragsdale/two_locus_selection}.

\section{Acknowledgements}\label{acknowledgements}

I thank Alex Diaz-Papkovich, Eric Friedlander, Simon Gravel, Mashaal Sohail,
Matthias Steinrücken, and Kevin Thornton for helpful discussions and feedback
on earlier versions of this manuscript.

\bibliographystyle{genetics}  
\bibliography{paper}

\end{document}

### 2022-03-23 manuscript/supplement.tex
add tex files

\documentclass[]{article}
\usepackage[round]{natbib}

\usepackage[margin=1in]{geometry}
\usepackage{url}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{color}
\usepackage{xspace}
\usepackage{amsmath,amssymb}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths

% cross-reference with main text
\usepackage{xr}
\externaldocument{paper}

\begin{document}

\title{Supporting Information for\\
``A heterogeneous landscape of selection and epistasis in protein-coding genes''}
\author[]{Aaron P. Ragsdale}
\affil[1]{Department of Integrative Biology, University of Wisconsin, Madison, WI, USA}
\affil[*]{apragsdale@wisc.edu}
\date{\today}
\maketitle

\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\theequation}{S\arabic{equation}}
\setcounter{figure}{0}
\setcounter{table}{0}
\setcounter{equation}{0}

\small
\tableofcontents
\normalsize
\newpage

\clearpage

\section{The diffusion equation and moment system for the two-locus sampling distribution}

The two-locus diffusion equation with additive selection was first described by
\citet{Kimura1955-qe} and studied extensively in the 1960s and 70s (e.g.,
\citet{Hill1966-gv,Ohta1969-ie}). The continuous distribution \(\psi(x_1, x_2,
x_3)\) of haplotype frequencies in a population, where \(x_1\) is the frequency
of \(AB\), \(x_2\) of \(Ab\), and \(x_3\) of \(aB\), is governed by the
multi-dimensional Fokker-Planck equation:
\begin{align} \label{eq:diffeq}
\frac{\partial \psi}{\partial \tau} = &
\frac{1}{2}\mathop{\sum\sum}_{1\leq i, j \leq 3}
\frac{\partial^2}{\partial x_i \partial x_j}
\left[\frac{x_i(\delta_{i=j}-x_j)\psi}{\nu(\tau)}\right] \\\nonumber
& -\frac{\rho}{2}\left(-\frac{\partial}{\partial x_1} D\psi
  + \frac{\partial}{\partial x_2} D\psi
  + \frac{\partial}{\partial x_3} D\psi\right) \\\nonumber
& - \frac{\gamma_A}{2}\left[
  \frac{\partial}{\partial x_1} x_1(1-x_1-x_2)\psi
  + \frac{\partial}{\partial x_2} x_2(1-x_1-x_2)\psi
  - \frac{\partial}{\partial x_3} x_3(x_1+x_2)\psi
  \right] \\\nonumber
& -\frac{\gamma_B}{2}\left[
  \frac{\partial}{\partial x_1} x_1(1-x_1-x_3)\psi
  - \frac{\partial}{\partial x_2} x_2(x_1+x_3)\psi
  + \frac{\partial}{\partial x_3} x_3(1-x_1-x_3)\psi
  \right].
\end{align}
\(D\) is the standard covariance measure of linkage disequilibrium, \[D=x_1 -
(x_1+x_2)(x_1+x_3) = x_1 x_4 - x_2 x_3,\] \(\gamma_A\) and \(\gamma_B\) are the
(additive) scaled selection coefficients at the left and right locus, and
\(\rho\) is the scaled recombination rate between the two loci. Time \(\tau\) is
measured in \(2N_e\) generations, and \(\nu(\tau)\) is the population size relative
to the ancestral size or some reference size at time \(\tau\).

Given a function \(\psi\) that solves Equation \ref{eq:diffeq}, the two-locus
sampling distribution for a sample size of \(n\) haploids can be found by
integrating \(\Psi\) against the multinomial sampling function, so that
\begin{align} \label{eq:multinomial}
\Psi_n(i, j, k) =
{n \choose{i, j, k, n-i-j-k}}
\mathop{\mathop{\int\int\int}_{x_1, x_2, x_3 \geq0}}_{x_1+x_2+x_3\leq1}
\psi(x_1, x_2, x_3) x_1^i x_2^j x_3^k (1-x_1-x_2-x_3)^{n-i-j-k}
dx_1 dx_2 dx_3.
\end{align}

In the method-of-moments approach, instead of solving the differential equation
for \(\psi\), we instead integrate both sides of the differential equation
against the multinomial sampling function for a given sampling configuration
\((i, j, k)\). On the left side, we get \(\partial_t \Psi_n(i, j, k)\), and on the
right we obtain, after some simple integration by parts and somewhat tedious
simplification, terms for drift, recombination, and selection that can be
written as sparse linear operators of \(\Psi_n\). Written compactly, this takes
the form
\begin{equation}
\partial_\tau \Psi_n =
\frac{1}{2\nu(\tau)}\mathcal{D}_{n}\Psi_n
+ \frac{\rho}{2}\mathcal{R}_{n}\Psi_n
+ \frac{\theta}{2}\mathcal{U}_{n}\Psi_n
+ \mathcal{S}_{n, \boldsymbol{\gamma}, \mathbf{h}}\Psi_n.
\end{equation}.

Alternatively, we arrive at this same linear system of equations by tracking
the expected sampling distribution over \(n\) lineages within the full population
and how that changes over time by drawing lineages from one generation to the
next in the style of \citet{Wright1931-wy}. Both \citet{Jouganous2017-pq}, for the
single-locus SFS, and \citet{Ragsdale2019-nt} drew this connection in detail, so I
refer readers to those previous papers for a fuller description of those
derivations and discussion. In the next section I simply repeat the results for
\(\mathcal{D}\), \(\mathcal{R}\), and \(\mathcal{U}\), briefly describe the moment
closure approximation (which is the same as presented in \citet{Ragsdale2019-nt}), and
then describe the selection operator \(\mathcal{S}\) for selection with
epistasis, both with and without dominance.

\subsection{Drift, mutation, recombination, and moment closure}\label{drift-mutation-recombination-and-moment-closure}

\subsubsection{Drift}\label{drift}

Drift for an entry \((i, j, k)\) depends only on \(\Psi_n\) and therefore closes.
The entries of \(\mathcal{D}\) are found by considering the possibility of a
coalescence event occurring within a given generation within \(n\) lineages in
the full population. If \(n \ll N\), we can safely assume that at most a single
such event occurs in any given generation.

\begin{align}
\mathcal{D}_{n}(i, j, k)\Psi_n =
& (i - 1) (n - i - j - k + 1) \Psi_n(i - 1, j, k) \\\nonumber
& + (i + 1) (n - i - j - k - 1) \Psi_n(i + 1, j, k) \\\nonumber
& + (i - 1) (k + 1) \Psi_n(i - 1, j, k + 1) \\\nonumber
& + (i + 1) (k - 1) \Psi_n(i + 1, j, k - 1) \\\nonumber
& + (i - 1) (j + 1) \Psi_n(i - 1, j + 1, k) \\\nonumber
& + (i + 1) (j - 1) \Psi_n(i + 1, j - 1, k) \\\nonumber
& + (j - 1) (n - i - j - k + 1) \Psi_n(i, j - 1, k) \\\nonumber
& + (j + 1) (n - i - j - k - 1) \Psi_n(i, j + 1, k) \\\nonumber
& + (j - 1) (k + 1) \Psi_n(i, j - 1, k + 1) \\\nonumber
& + (j + 1) (k - 1) \Psi_n(i, j + 1, k - 1) \\\nonumber
& + (k - 1) (n - i - j - k + 1) \Psi_n(i, j, k - 1) \\\nonumber
& + (k + 1) (n - i - j - k - 1) \Psi_n(i, j, k + 1) \\\nonumber
& - 2\left( i ( n - i - j - k) + i k + i j + j (n - i - j - k) + j k + k(n - i - j - k)\right)\Psi_n(i, j, k)
\end{align}

\subsubsection{Recombination}\label{recombination}

If a lineage in our sample of size \(n\) recombines in a given generation, which
occurs with probability \(nr\), we need to draw an extra lineage from the full
population for it to recombine with. This means we need \(\Psi_{n+1}\) in the
previous generation. After drawing that extra lineage, \(\Psi_n\) changes as we
draw one of the two recombinant types (each with probability \(1/2\)) instead of
the lineage that was chosen to recombine.

\begin{align}
\mathcal{R}_{n}(i, j, k)\Psi_n = &
\frac{(i + 1)(n - i - j - k + 1)}{n + 1}\Psi_{n+1}(i + 1, j - 1, k) \\\nonumber
& + \frac{(i + 1)(n - i - j - k + 1)}{n + 1}\Psi_{n+1}(i + 1, j, k - 1) \\\nonumber
& + \frac{(j + 1)(k + 1}{n + 1)}\Psi_{n+1}(i - 1, j + 1, k + 1) \\\nonumber
& + \frac{(j + 1)(k + 1}{n + 1)}\Psi_{n+1}(i, j + 1, k + 1) \\\nonumber
& - \frac{(i + 1)(n - i - j - k)}{n + 1}\Psi_{n+1}(i + 1, j, k) \\\nonumber
& - \frac{(j + 1) k}{n + 1}\Psi_{n+1}(i, j + 1, k) \\\nonumber
& - \frac{j (k + 1)}{n + 1}\Psi_{n+1}(i, j, k + 1) \\\nonumber
& - \frac{i (n - i - j - k + 1}{n + 1}\Psi_{n+1}(i, j, k)
\end{align}

\subsubsection{Mutation}\label{mutation}

We assume an infinite sites mutation (ISM) model where new mutations occur at
previously unmutated loci. In the two-locus ISM model, two-locus pairs of
variable loci arise when a mutation occurs at one locus when the other locus is
already variable. Thus, new mutations at the \(B/b\) locus occur against the
single-locus allele frequency distribution \(\Phi_{n,A}\), and new mutations at
the \(A/a\) locus occur against \(\Phi_{n, B}\), which are found via the single-locus
system from \citet{Jouganous2017-pq}.

\begin{align}
\mathcal{U}_{n}(i, j, k)\Psi_n = &
(j + 1)\frac{\theta_B}{2} \Phi_{n, A}(j+1) \delta_{i=1, k=0} \\\nonumber
& + (n - j)\frac{\theta_B}{2} \Phi_{n, A}(j) \delta_{i=0, k=1} \\\nonumber
& + (i + 1)\frac{\theta_A}{2} \Phi_{n, B}(i+1) \delta_{i=1, j=0} \\\nonumber
& + (n - i)\frac{\theta_A}{2} \Phi_{n, B}(i) \delta_{i=0, j=1} \\\nonumber
\end{align}

\subsubsection{Jackknife moment closure approximation}\label{jackknife-moment-closure-approximation}

We use a jackknife approximation to write the entries of \(\Psi_{n+1}\) and
\(\Psi_{n+2}\) as linear combinations of entries in \(\Psi_n\). The general
strategy is to assume the underlying continuous distribution \(\psi(x_1, x_2, x_3)\) can be approximated locally as a quadratic, and then use entries in
\(\Psi_n\) that are close in frequency to a given entry in \(\Psi_{n+l}\) to
estimate the coefficients of that quadratic using the multinomial sampling
formula. Then this quadratically local approximation to \(\psi\) can be used to
compute \(\Psi_{n+l}(i, j, k)\) using Eq. \eqref{eq:multinomial}. Readers should
refer to section S1.3.5 in the Supporting material for \citet{Ragsdale2019-nt} for
details.

\subsection{Selection}\label{selection}

First consider the case of no dominance, so that the haplotypes \(Ab\), \(aB\),
and \(AB\) have selection coefficients \(s_{Ab}\), \(s_{aB}\), and \(s_{AB}\),
respectively. Note that the case with \(s_{AB} = s_{Ab} + s_{aB}\) implies no
epistasis between the \(A/a\) and \(B/b\) loci. Here, we assume all selection
coefficients are negative. In a given generation, a selection event could
occur in which a haplotype is rejected (selected against) with probability
proportional to its selection coefficient, \(-s\). We then draw an extra lineage
from the full population to replace that rejected lineage.

For example, the probability that an \(AB\) haplotype is selected against and
replaced by an \(Ab\) haplotype is \[-ns_{AB} \frac{i}{n+1}{j+1}{n}\Psi_{n+1}(i,
j + 1, k),\] where the additional \(j + 1\) lineage in a sample of size \(n + 1\)
accounts drawing that extra \(Ab\) haplotype. Taking all such selective events
together, for additive selection we get
\begin{align}
\mathcal{S}_{n}(i, j, k)\Psi_n = &
\frac{i+1}{n+1}\left(-s_{AB}(n-i) + s_{Ab}j
+ s_{aB}k\right)\Psi_{n+1}(i+1, j, k) \\\nonumber
& +\frac{j+1}{n+1}\left(s_{AB}i - s_{Ab}(n-j)
+ s_{aB}k\right)\Psi_{n+1}(i, j+1, k) \\\nonumber
& +\frac{k+1}{n+1}\left(s_{AB}i + s_{Ab}j
- s_{aB}(n-k)\right)\Psi_{n+1}(i, j, k+1) \\\nonumber
& + \frac{n-i-j-k+1}{n+1}\left(s_{AB}i + s_{Ab}j
+ s_{aB}k\right)\Psi_{n+1}(i, j, k)
\end{align}

For a general diploid selection model, the idea is nearly the same, but we need
to draw an extra lineage to determine the fitness of a diploid individual. For
example, the probability that an \(AB\) haplotype is paired with an additional
lineage \(Ab\) and selected against, and then replaced by an \(aB\) haplotype is
\[-ns_{AB/Ab}\frac{i}{n+2}\frac{j+1}{n+1}\frac{k+1}{n}\Psi_{n+2}(i, j+1,
k+1).\] There are now many more possible selective events to consider, but
after accounting for all possible diploid pairs and replacements (90 in total)
and simplifying, we find

\begin{align}
\mathcal{S}_{n}(i, j, k)\Psi_n = &
\frac{n-i-j-k+2}{n+2}\frac{n-i-j-k+1}{n+1}\left(
  s_{AB/ab}i
  + s_{Ab/ab}j
  + s_{aB/ab}k
\right)\Psi_{n+2}(i,j,k) \\\nonumber
& + \frac{i+1}{n+2}\frac{n-i-j-k+1}{n+1}(
  s_{AB/AB}i
  + s_{AB/Ab}j
  + s_{AB/aB}k
  + s_{Ab/ab}j \\\nonumber & \hspace{20pt}
  + s_{aB/ab}k
  - s_{AB/ab}(n + j + k)
)\Psi_{n+2}(i+1, j, k) \\\nonumber
& + \frac{i+2}{n+2}\frac{i+1}{n+1}(
  s_{AB/Ab}j
  + s_{AB/aB}k
  + s_{AB/ab}(n-i-j-k)
  - s_{AB/AB}(n-i)
)\Psi_{n+2}(i+2, j, k) \\\nonumber
& + \frac{i+1}{n+2}\frac{j+1}{n+2}(
  s_{AB/AB}i
  + s_{AB/aB}k
  + s_{AB/ab}(n-i-j-k)
  + s_{Ab/Ab}j \\\nonumber & \hspace{20pt}
  + s_{Ab/aB}k
  + s_{Ab/ab}(n-i-j-k)
  - s_{AB/Ab}(2n-i-j)
)\Psi_{n+2}(i+1, j+1, k) \\\nonumber
& + \frac{i+1}{n+2}\frac{k+1}{n+1}(
  s_{AB/AB}i
  + s_{AB/Ab}j
  + s_{AB/ab}(n-i-j-k)
  + s_{Ab/aB}j \\\nonumber & \hspace{20pt}
  + s_{aB/aB}k
  + s_{aB/ab}(n-i-j-k)
  - s_{AB/aB}(2n-i-k)
)\Psi_{n+2}(i+1, j, k+1) \\\nonumber
& + \frac{j+1}{n+2}\frac{n-i-j-k+1}{n+1}(
  s_{AB/Ab}i
  + s_{AB/ab}i
  + s_{Ab/Ab}j
  + s_{Ab/aB}k \\\nonumber & \hspace{20pt}
  + s_{aB/ab}k
  -s_{Ab/ab}(n+i+k)
)\Psi_{n+2}(i, j+1, k) \\\nonumber
& + \frac{j+2}{n+2}\frac{j+1}{n+1}(
s_{AB/Ab}i + s_{Ab/aB}k + s_{Ab/ab}(n-i-j-k) - s_{Ab/Ab}(n-j)
)\Psi_{n+2}(i, j+2, k) \\\nonumber
& + \frac{j+1}{n+2}\frac{k+1}{n+1}(
  s_{AB/Ab}i
  + s_{AB/aB}i
  + s_{Ab/Ab}j
  + s_{Ab/ab}(n-i-j-k) \\\nonumber & \hspace{20pt}
  + s_{aB/aB}k
  + s_{aB/ab}(n-i-j-k)
  - s_{Ab/aB}(2n - j - k)
)\Psi_{n+2}(i, j+1, k+1) \\\nonumber
& + \frac{k+1}{n+2}\frac{n-i-j-k+1}{n+1}(
  s_{AB/aB}i
  + s_{AB/ab}i
  + s_{Ab/aB}j
  + s_{Ab/ab}j \\\nonumber & \hspace{20pt}
  + s_{aB/aB}k
  - s_{aB/ab}(n + i + j)
)\Psi_{n+2}(i, j, k+1) \\\nonumber
& + \frac{k+2}{n+2}\frac{k+1}{n+1}(
  s_{AB/aB}i
  + s_{Ab/aB}j
  + s_{aB/ab}(n-i-j-k)
  - s_{aB/aB}(n - k)
)\Psi_{n+2}(i, j, k+2).
\end{align}
Multiplying through by \(2N_{ref}\) gives us selection operators in terms of
\(\gamma\) instead of \(s\).

\subsection{Validation against simulation}\label{validation-against-simulation}

\subsection{Background selection and associative overdominance}\label{background-selection-and-associative-overdominance}

\subsection{Frequency dependence of signed LD}\label{frequency-dependence-of-signed-ld}

\section{Data analysis}\label{data-analysis}

\subsection{DFE for missense and LOF variants}\label{dfe-for-missense-and-lof-variants}

Loss-of-function (LOF) variants show a dramatic skew toward low-frequency
variants across all human populations (Table \ref{tab:tajimasD}). Here, using
the folded SFS for synonymous, missense, and LOF mutations across all autosomal
genes, I inferred DFEs for missense and LOF mutations independently. I
considered a few different dominance coefficients to explore the effect of the
assumed recessivity of the two classes of mutations.

The standard SFS approach to fitting the DFE involves first inferring a
demographic history for the population using putatively neutral variants (here,
synonymous mutations), and then fixing that demography and fitting a
parameterized function for the distribution of selection coefficients for new
mutations for the selected classes. DFE inference also requires an estimate for
the total mutation rate of the different mutation classes, as much of the
signal for strongly selected mutations comes from observing fewer mutations
than expected given a known mutation rate (with the assumption that selection
purges some fraction of strongly deleterious mutations which are unseen in the
sample). Here, I fit demography and DFEs to the folded SFS from the Mende in
Sierra Leone (MSL) using \emph{moments} version 1.1.0 \citep{Jouganous2017-pq}.

I used the mutation model from \citet{Karczewski2020-le} to estimate the total
mutation rate across autosomal genes (\(uL\), where \(u\) is the per-base mutation
rate of a given mutation class, and \(L\) is the total length of the coding
genome). These values were \((0.1442, 0.3426, 0.0256)\) for synonymous, missense,
and LOF mutations, respectively. Roughly two thirds of new mutations in coding
regions are expected to be missense mutations, while only 5\% of new mutations
are LOF. I fit a demographic model to the synonymous variants, which included a
population expansion in the deeper past and exponential growth in the recent
past (Figure \ref{fig:msldemogdfe}A). Using the inferred optimal scaled mutation
rate, \(\theta=4N_euL\), I estimated \(Ne\approx12,300\), and assuming an average
generation time of 29 years I converted the inferred genetic units to physical
units. The best-fit model had a roughly two-fold expansion 400 thousand years
ago, and then exponential growth over the past 20-30 thousand years, with a
current effective size of \(\approx 63,000\).

Under this demographic model, I fit a gamma distribution for the distribution
of fitness effects to missense and LOF mutations (Table \ref{tab:msldfe}). For
each fit, I fixed the scaled mutation rate for each mutation class, so that
\(\theta_{mis} = \frac{u_{mis}}{u_{syn}} \hat\theta_{syn}\) and \(\theta_{lof} = \frac{u_{lof}}{u_{syn}} \hat\theta_{syn}\), where values of \(u\) were found using
the GNOMAD mutation model \citep{Karczewski2020-le}. I tested three values for the
dominance coefficient \(h\): \(0\), \(0.2\) and \(0.5\). For missense mutations, \(h=0\)
gave a poor fit to the data, and \(h=0.5\) fit best among the three tested
dominance coefficients. For LOF variants, \(h=0\) also fit poorly, but \(h=0.2\)
and \(h=0.5\) gave similar likelihoods, highlighting that inferring dominance
using the SFS is poorly constrained. Regardless of the dominance coefficient
assumed, however, the vast majority of LOF variants were inferred to be
strongly deleterious, with only \(\sim10\%\) of new mutations having selection
coefficients on the order \(1/N_e\) or less.

\subsection{Multinucleotide mutations and positive LD between linked synonymous variants}\label{multinucleotide-mutations-and-positive-ld-between-linked-synonymous-variants}

Multinucleotide mutations (MNMs) are complex mutational events that result in
multiple mutations occurring on the same haplotype background in a single
generation. Because MNMs fall on the same haplotype, those mutations will be in
positive LD, and LD between those pairs that are very tightly linked will not
be broken down all that rapidly. MNMs are expected to occur over relatively
short distances, on the order of 10s or 100s of base pairs, making them a
likely culprit of the observed positive LD among synonymous mutations at short
distances.

MNM events can be easily incorporated into the moment system with a simple
adjustment to the mutation operator. Instead of all mutations occurring
independently in haplotypes with mutations already segregating at the other
locus, some fraction of new mutations could instead occur spontaneously and
create a new pair of mutations with initial counts \(n_{AB} = 1\) and \(n_{ab} = n-1\).

Here, I fit a simple exponential model for the fraction of new mutations at a
given distance that arose through a MNM event, so that \(P(MNM | d) = Ae^{-\lambda d}\), where \(d\) is the distance separating pairs of mutations. I
considered all synonymous mutations within genes in the MSL population and used
the same population size history model as inferred in the DFE section above for
a demographic control. This left two parameters to be fit, \(A\) and \(\lambda\),
which I fit to the binned decay curve of \(\sigma_d^1\). I needed to assume an
average per-base recombination rate \(r\) across gene regions, and tested a
number of values between \(10^{-9}\) and \(2\times 10^{-8}\). The optimization was
insensitive to the chosen value of \(r\), because the decay of positive LD occurs
rapidly. For any plausible value of \(r\), \(\sigma_d^1\) decays to zero well
before distances between pairs have scaled recombination rates \(\rho=4 N_e r d \sim 1\), and expected statistics for \(\rho \ll 1\) vary only negligibly.

In fitting the LD decay of \(\sigma_d^1\), the best fit parameters were
\(A=0.132\), and \(\lambda=0.0103\). An exponential scaling of \(0.01\) implies that
the vast majority of new mutations \emph{pairs} do not occur via MNMs for distances
greater than 200 bp, though a substantial fraction (\(10-15\%\)) occur via MNMs
for very tightly linked loci with distances on the order \(0-50\) base pairs. It
is important to note that this does not mean that \(10-15\%\) of new mutations
occur via MNMs, since this fraction is conditioned on two mutations occurring
at short distances.

\subsection{Grouping Thousand Genomes populations based on clustering}
\label{grouping-thousand-genomes-populations-based-on-clustering}

The large confidence intervals for measurements of signed LD could be driven by
either averaging over relatively few observed pairs of mutations, or due to
small sample sizes that make each individual measurement a noisy estimate of
the LD for that pair of mutations in the full population. To explore the
underlying cause of measurement uncertainty in the
\citet{1000_Genomes_Project_Consortium2015-zq} data, I considered larger sets of
samples by combining populations that consistently cluster together in PCA and
UMAP space and have low differentiation \citep{Diaz-Papkovich2020-ee}. I took
combinations of CEU/GBR, CHB/CHS, CDX/KHV, and MSL/GWD. While recognizing that
residual population structure in these population combinations could alter
expected LD statistics compared to the respective single-population estimates,
I was more interested in the effect that increasing the sample sizes would have
on estimated measurement error.

Across each of the four combinations tested, confidence intervals were roughly
equivalent to those of each of the individual populations. This suggests that
the limiting factor to accurate LD measurement is not sample size but rather
the overall levels of diversity and number of pairs of mutations that we
compare. \(\mathbb{E}[D]\) is most affected by common variants, and the sample
sizes of the Thousand Genomes Project data are likely sufficient to accurately
estimate common allele frequencies. Adding additional samples will increase the
number of rare variants that we observe, but rare variants have minimal impact
on \(\sigma_d^1\). Thus, the accuracy of estimates of \(\sigma_d^1\) is more
fundamentally limited by evolutionary history and genome biology (i.e.~past
population sizes, mutation and recombination rates) than by sample sizes.

\addcontentsline{toc}{section}{References}
\bibliographystyle{genetics}
\bibliography{paper}

\clearpage
\newpage

\addcontentsline{toc}{section}{Supporting tables}
\section*{Supporting tables}

\begin{table}[ht!]
    \centering
    \begin{tabular}{llll}
        \toprule
        Diploid genotype & General model & Simple dominance & Gene-based dominance\\
        \midrule
        $AB$ / $AB$ & $1 + s_{AB/AB}$ & $1 + 2s_A + 2s_B$ & $1 + 2s$\\
        $AB$ / $Ab$ & $1 + s_{AB/Ab}$ & $1 + 2s_A + 2s_Bh_B$ & $1 + 2s$\\
        $AB$ / $aB$ & $1 + s_{AB/aB}$ & $1 + 2s_Ah_A + 2s_B$ & $1 + 2s$\\
        $AB$ / $ab$ & $1 + s_{AB/ab}$ & $1 + 2s_Ah_A + 2s_Bh_B$ & $1 + 2sh$\\
        $Ab$ / $Ab$ & $1 + s_{Ab/Ab}$ & $1 + 2s_A$ & $1 + 2s$\\
        \addlinespace
        $Ab$ / $aB$ & $1 + s_{Ab/aB}$ & $1 + 2s_Ah_A + 2s_Bh_B$ & $1 + 2s$\\
        $Ab$ / $ab$ & $1 + s_{Ab/ab}$ & $1 + 2s_Ah_A$ & $1 + 2sh$\\
        $aB$ / $aB$ & $1 + s_{aB/aB}$ & $1 + 2s_B$ & $1 + 2s$\\
        $aB$ / $ab$ & $1 + s_{aB/ab}$ & $1 + 2s_Bh_B$ & $1 + 2sh$\\
        $ab$ / $ab$ & $1$ & $1$ & $1$\\
        \bottomrule
    \end{tabular}
    \caption{General selection model for diploids and dominance models.}
    \label{tab:selmodels}
\end{table}

\begin{table}[ht!]
    \centering
    \begin{tabular}{ll}
        \toprule
        Haplotype & Fitness\\
        \midrule
        $AB$ & $(1 + s_A + s_B)(1+\epsilon)$\\
        $Ab$ & $1 + s_A$\\
        $aB$ & $1 + s_B$\\
        $ab$ & $1$\\
        \bottomrule
    \end{tabular}
    \caption{Haploid epistasis model.}
    \label{tab:epistasis}
\end{table}

\begin{table}[ht!]
    \centering
    \begin{tabular}{lll}
        \toprule
        Code & Description & Region\\
        \midrule
        ESN & Esan in Nigeria & Africa\\
        GWD & Gambian in Western Divisions in the Gambia & Africa\\
        LWK & Luhya in Webuye, Kenya & Africa\\
        MSL & Mende in Sierra Leone & Africa\\
        YRI & Yoruba in Ibadan, Nigeria & Africa\\
        \addlinespace
        CEU & Utah Residents (CEPH) with Northern and Western European Ancestry & Europe\\
        GBR & British in England and Scotland & Europe\\
        FIN & Finnish in Finland & Europe\\
        IBS & Iberian Population in Spain & Europe\\
        TSI & Toscani in Italia & Europe\\
        \addlinespace
        CDX & Chinese Dai in Xishuangbanna, China & East Asia\\
        CHB & Han Chinese in Beijing, China & East Asia\\
        CHS & Southern Han Chinese & East Asia\\
        JPT & Japanese in Tokyo, Japan & East Asia\\
        KHV & Kinh in Ho Chi Minh City, Vietnam & East Asia\\
        \bottomrule
    \end{tabular}
    \caption{Thousand Genomes Project population descriptions for populations used in this study.}
    \label{tab:1kgpops}
\end{table}

\begin{longtable}[t]{lllr}
\caption{\label{tab:tajimasD}Tamija's $D$ for classes of coding mutations, both within annotated domains and outside of domains.}\\
\toprule
Population & Mutation type & Region & Tajima's D\\
\midrule
\endfirsthead
\caption[]{\label{tab:tajimasD}Tamija's $D$ for classes of coding mutations, both within annotated domains and outside of domains. \textit{(continued)}}\\
\toprule
Population & Mutation type & Region & Tajima's D\\
\midrule
\endhead

\endfoot
\bottomrule
\endlastfoot
ESN & Synonymous & All & -0.882\\
 &  & In domain & -0.854\\
 &  & Not in domain & -0.921\\
 & Missense & All & -1.414\\
 &  & In domain & -1.535\\
 &  & Not in domain & -1.293\\
 & Loss of function & All & -1.483\\
 &  & In domain & -2.156\\
 &  & Not in domain & -1.282\\
\addlinespace
GWD & Synonymous & All & -1.011\\
 &  & In domain & -0.981\\
 &  & Not in domain & -1.052\\
 & Missense & All & -1.566\\
 &  & In domain & -1.678\\
 &  & Not in domain & -1.452\\
 & Loss of function & All & -1.697\\
 &  & In domain & -2.328\\
 &  & Not in domain & -1.501\\
\addlinespace
LWK & Synonymous & All & -1.109\\
 &  & In domain & -1.088\\
 &  & Not in domain & -1.139\\
 & Missense & All & -1.589\\
 &  & In domain & -1.700\\
 &  & Not in domain & -1.477\\
 & Loss of function & All & -1.666\\
 &  & In domain & -2.278\\
 &  & Not in domain & -1.477\\
\addlinespace
MSL & Synonymous & All & -0.983\\
 &  & In domain & -0.959\\
 &  & Not in domain & -1.017\\
 & Missense & All & -1.501\\
 &  & In domain & -1.603\\
 &  & Not in domain & -1.400\\
 & Loss of function & All & -1.559\\
 &  & In domain & -2.303\\
 &  & Not in domain & -1.332\\
\addlinespace
YRI & Synonymous & All & -0.928\\
 &  & In domain & -0.898\\
 &  & Not in domain & -0.971\\
 & Missense & All & -1.467\\
 &  & In domain & -1.586\\
 &  & Not in domain & -1.348\\
 & Loss of function & All & -1.624\\
 &  & In domain & -2.237\\
 &  & Not in domain & -1.424\\
\addlinespace
CEU & Synonymous & All & -0.417\\
 &  & In domain & -0.392\\
 &  & Not in domain & -0.452\\
 & Missense & All & -1.248\\
 &  & In domain & -1.404\\
 &  & Not in domain & -1.082\\
 & Loss of function & All & -1.501\\
 &  & In domain & -2.196\\
 &  & Not in domain & -1.280\\
\addlinespace
FIN & Synonymous & All & -0.058\\
 &  & In domain & -0.047\\
 &  & Not in domain & -0.075\\
 & Missense & All & -0.883\\
 &  & In domain & -1.048\\
 &  & Not in domain & -0.710\\
 & Loss of function & All & -1.200\\
 &  & In domain & -2.034\\
 &  & Not in domain & -0.906\\
\addlinespace
GBR & Synonymous & All & -0.319\\
 &  & In domain & -0.300\\
 &  & Not in domain & -0.345\\
 & Missense & All & -1.120\\
 &  & In domain & -1.276\\
 &  & Not in domain & -0.954\\
 & Loss of function & All & -1.313\\
 &  & In domain & -2.178\\
 &  & Not in domain & -0.997\\
\addlinespace
IBS & Synonymous & All & -0.689\\
 &  & In domain & -0.664\\
 &  & Not in domain & -0.724\\
 & Missense & All & -1.424\\
 &  & In domain & -1.560\\
 &  & Not in domain & -1.279\\
 & Loss of function & All & -1.636\\
 &  & In domain & -2.349\\
 &  & Not in domain & -1.378\\
\addlinespace
TSI & Synonymous & All & -0.650\\
 &  & In domain & -0.625\\
 &  & Not in domain & -0.685\\
 & Missense & All & -1.422\\
 &  & In domain & -1.568\\
 &  & Not in domain & -1.266\\
 & Loss of function & All & -1.655\\
 &  & In domain & -2.349\\
 &  & Not in domain & -1.397\\
\addlinespace
CDX & Synonymous & All & -0.374\\
 &  & In domain & -0.366\\
 &  & Not in domain & -0.385\\
 & Missense & All & -1.179\\
 &  & In domain & -1.323\\
 &  & Not in domain & -1.026\\
 & Loss of function & All & -1.360\\
 &  & In domain & -2.194\\
 &  & Not in domain & -1.062\\
\addlinespace
CHB & Synonymous & All & -0.598\\
 &  & In domain & -0.593\\
 &  & Not in domain & -0.606\\
 & Missense & All & -1.389\\
 &  & In domain & -1.528\\
 &  & Not in domain & -1.239\\
 & Loss of function & All & -1.586\\
 &  & In domain & -2.344\\
 &  & Not in domain & -1.298\\
\addlinespace
CHS & Synonymous & All & -0.544\\
 &  & In domain & -0.545\\
 &  & Not in domain & -0.544\\
 & Missense & All & -1.334\\
 &  & In domain & -1.499\\
 &  & Not in domain & -1.150\\
 & Loss of function & All & -1.559\\
 &  & In domain & -2.290\\
 &  & Not in domain & -1.292\\
\addlinespace
JPT & Synonymous & All & -0.371\\
 &  & In domain & -0.368\\
 &  & Not in domain & -0.376\\
 & Missense & All & -1.194\\
 &  & In domain & -1.355\\
 &  & Not in domain & -1.019\\
 & Loss of function & All & -1.410\\
 &  & In domain & -2.272\\
 &  & Not in domain & -1.086\\
\addlinespace
KHV & Synonymous & All & -0.576\\
 &  & In domain & -0.562\\
 &  & Not in domain & -0.596\\
 & Missense & All & -1.346\\
 &  & In domain & -1.473\\
 &  & Not in domain & -1.210\\
 & Loss of function & All & -1.535\\
 &  & In domain & -2.294\\
 &  & Not in domain & -1.269\\*    
\end{longtable}

\begin{table}
    \centering
    \small
    \begin{tabular}{lrrrrrrrrr}
        \toprule
        Class & $h$ & shape & scale & LL &
        $[0,10^{-5})$ & $[10^{-5},10^{-4})$ & $[10^{-4},10^{-3})$ &
        $[10^{-3},10^{-2})$ & $[10^{-2},\infty)$\\
        \midrule
        Missense & 0.0 & 0.093 & 768505 & -678.2 & 0.260 & 0.062 & 0.077 & 0.096 & 0.505\\
         & 0.2 & 0.138 & 6660 & -416.7 & 0.260 & 0.098 & 0.134 & 0.182 & 0.327\\
         & 0.5 & 0.147 & 2117 & -392.0 & 0.282 & 0.114 & 0.159 & 0.214 & 0.231\\
        \addlinespace
        LOF & 0.0 & 0.132 & 99999054 & -248.3 & 0.077 & 0.028 & 0.037 & 0.051 & 0.807\\
         & 0.2 & 0.177 & 477994 & -226.7 & 0.083 & 0.042 & 0.063 & 0.095 & 0.717\\
         & 0.5 & 0.188 & 121419 & -224.2 & 0.092 & 0.050 & 0.077 & 0.119 & 0.662\\
        \bottomrule
    \end{tabular}
    \caption{
        DFEs inferred for missense and loss-of-function variants in MSL for varying
        values of $h$. General patterns are consistent across different chosen values
        of $h$, although $h=0$ results in poorer fits for both missense and LOF variants.
        Columns to the right of the log-likelihood (LL) column show proportions of
        new mutations with $|s|$ in each given bin.
    }
    \label{tab:msldfe}
\end{table}

\clearpage
\newpage

\addcontentsline{toc}{section}{Supporting Figures}
\section*{Supporting figures}

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/jackknife}
    \caption{
        \textbf{Accuracy of the jackknife approximation and runtime} Small
        sample sizes can lead to large error in the moment-closure
        approximation for larger recombination distances or selection
        coefficients. Generally, the jackknife approximation breaks down for
        recombination rates greater than \(\rho\approx30\). While increasing
        sample size leads to more accurate solutions, it comes at the cost of
        both increased runtime and memory usage. Most analyses in this paper
        used sample sizes between 40 and 80.
    }
    \label{fig:jackknife}
\end{figure}

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/bgs_gamma_-1.0_h_0.5_n_50}
    \caption{
        \textbf{Background selection with slightly deleterious mutations.}
        Using an individual based forward simulator
    }
    \label{fig:bgs1}
\end{figure}

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/bgs_gamma_-20.0_h_0.5_n_50}
    \caption{
        \textbf{Background selection with moderately deleterious mutations.}
    }
    \label{fig:bgs2}
\end{figure}

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/bgs_gamma_-1.0_h_0.0_n_50}
    \caption{
        \textbf{Associative overdominance due to slightly deleterious recessive mutations.}
    }
    \label{fig:bgs3}
\end{figure}

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/bgs_gamma_-20.0_h_0.1_n_80}
    \caption{
        \textbf{Background selection with moderately deleterious recessive mutations.}
    }
    \label{fig:bgs4}
\end{figure}

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/demog_bottle_expand.sd2}
    \caption{
        \textbf{.}
    }
    \label{fig:toy_sd2}
\end{figure}

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/demog_bottle_expand.dominance}
    \caption{
        \textbf{.}
    }
    \label{fig:toy_dom}
\end{figure}

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/demog_bottle_expand.dominance.sd2}
    \caption{
        \textbf{.}
    }
    \label{fig:toy_dom_sd2}
\end{figure}

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/demog_YRI_CEU}
    \caption{
        \textbf{.}
    }
    \label{fig:relate}
\end{figure}

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/demog_YRI_CEU.sd2}
    \caption{
        \textbf{.}
    }
    \label{fig:relate_sd2}
\end{figure}

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/demog_YRI_CEU.dominance.sd2}
    \caption{
        \textbf{.}
    }
    \label{fig:relate_dom_sd2}
\end{figure}

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/ld_decay_gene_wide}
    \caption{
        \textbf{LD decay for pairs of synonymous and missense mutations
        in the same gene.}
    }
    \label{fig:LDgene}
\end{figure}

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/ld_decay_in_domain}
    \caption{
        \textbf{LD decay for pairs of synonymous and missense mutations
        that fall inside the same domain.}
    }
    \label{fig:LDwithin}
\end{figure}

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/ld_decay_outside_domains}
    \caption{
        \textbf{LD decay for pairs of synonymous and missense mutations
        outside of domains.}
    }
    \label{fig:LDoutside}
\end{figure}

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/data_domains_afr}
    \caption{
        \textbf{LD for pairs of synonymous and missense mutations within the
        same domain and outside domains at matched distances.}
    }
    \label{fig:domainsAFR}
\end{figure}

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/data_domains_eur}
    \caption{
        \textbf{LD for pairs of synonymous and missense mutations within the
        same domain and outside domains at matched distances.}
    }
    \label{fig:domainsEUR}
\end{figure}

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/data_domains_eas}
    \caption{
        \textbf{LD for pairs of synonymous and missense mutations within the
        same domain and outside domains at matched distances.}
    }
    \label{fig:domainsEAS}
\end{figure}

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/mutations_in_genes}
    \caption{
        \textbf{Correlation between numbers of missense and synonymous mutations
        within each gene.}
    }
    \label{fig:mutCorrGene}
\end{figure}

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/mutations_in_domains}
    \caption{
        \textbf{Correlation between numbers of missense and synonymous mutations
        within each annotated domain.}
    }
    \label{fig:mutCorrDomain}
\end{figure}

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/msl_demography_dfes}
    \caption{
        \textbf{Demography and DFE for MSL.} A demographic model was fit to
        the folded synonymous SFS, and DFEs were fit to missense and loss-of-function
        SFS. Shown here are DFEs fit with \(h=0.5\). See Table \ref{tab:msldfe} for
        inferred best-fit parameters.
    }
    \label{fig:msldemogdfe}
\end{figure}

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/msl_mnms}
    \caption{
        \textbf{Optimization of fraction of new mutations arising via
        multinucleotide mutations by distance.} A simple exponential function was fit
        to describe the probability that a pair of mutations arose through a MSM event
        at a given distance \(d\), as \(Ae^{-\lambda d}\). Across all recombination rates
        tested, the best fit parameters were \(A=0.13\) and \(\lambda=0.010\).
    }
    \label{fig:mslmnms}
\end{figure}

\end{document}

### 2022-03-25 manuscript/paper.tex
getting closer..

@@
 \usepackage{booktabs}
+\usepackage{xspace}
 
@@
 interference and epistatic interactions, with some disagreement over
-interpreting observations from data. Interpretation is hindered by the relative
-lack of analytic or even numerical expectations for patterns of variation
-between pairs of loci under the combined effects of selection, dominance,
-epistasis, and demography. Here, I develop a numerical approach to compute the
-expected two-locus sampling distribution under diploid selection with arbitrary
+interpreting observations from data. Interpretation is hindered by a lack of
+analytic or even numerical expectations for patterns of variation between pairs
+of loci under the combined effects of selection, dominance, epistasis, and
+demography. Here, I develop a numerical approach to compute the expected
+two-locus sampling distribution under diploid selection with arbitrary
 epistasis and dominance, recombination, and variable population size. I use
 this to explore how epistasis and dominance affect expected signed LD,
-including for non-steady-state demography relevant to human populations.
-\comment{I then use whole-genome sequencing data from humans to assess how well we
-can differentiate modes of selective interactions in practice.}
-\comment{this needs updating}
-I find that
-positive LD between missense mutations within genes is driven by strong
-positive allele-frequency correlations between pairs of mutations that fall
-within the same conserved domain, pointing to compensatory mutations or
-antagonistic epistasis as the prevailing mode of interaction within but not
-outside of conserved genic elements. The heterogeneous landscape of both
-mutational fitness effects and selective interactions within protein-coding
-genes calls for more refined inferences of the joint distribution of fitness
-and interactive effects, and the methods presented here should prove useful in
-that pursuit.
+including for non-steady-state demography relevant to human populations. Using
+whole-genome sequencing data from humans, I explore genome-wide patterns of LD
+within protein-coding genes. I show that positive LD between missense mutations
+within genes is driven by strong positive allele-frequency correlations between
+pairs of mutations that fall within the same annotated conserved domain,
+pointing to compensatory mutations or antagonistic epistasis as the prevailing
+mode of interaction within conserved genic elements. LD between missense
+mutations is reduced outside of conserved domains, as would expected under
+Hill-Robertson interference. The heterogeneous landscape of both mutational
+fitness effects and selective interactions within protein-coding genes calls
+for more refined inferences of the joint distribution of fitness and
+interactive effects, and the methods presented here should prove useful in that
+pursuit.
 
@@
 Most new mutations that affect fitness are deleterious and tend to be
-eliminated from a population
-\textbf{Citations that most mutations are negatively selected}.
-The amount of time that a
-deleterious mutation segregates depends on the strength of selection against
-genomes that carry it, with very damaging mutations kept at low frequencies and
-purged relatively rapidly. In the time between mutation and fixation or loss,
-selected variants, both beneficial and damaging, can dramatically impact
-patterns of variation in nearby linked regions \textbf{MSH, Charlesworth, Stephan/Kim}.
-This distortion away from neutral expectations is now empirically documented using
-sequencing data from an ever-growing set of study systems, but questions remain
-about the primary mode of interactions between multiple linked variants and
-their joint effects on genome-wide patterns of diversity.
+eliminated from a population. The amount of time that a deleterious mutation
+segregates depends on the strength of selection against genomes that carry it,
+with very damaging mutations kept at low frequencies and purged relatively
+rapidly. In the time between mutation and fixation or loss, selected variants,
+both beneficial and damaging, can dramatically impact patterns of variation in
+nearby linked regions \citep[e.g.][]{Smith1974-em,Charlesworth1995-dq,Kim2000-on}.
+This distortion away from neutral expectations has been empirically documented
+using sequencing data from an ever-growing set of study systems, but questions
+remain about the primary mode of interactions between multiple linked variants
+and their joint effects on genome-wide patterns of diversity.
 
@@
 \citep{Birky1988-jm}. Allele frequency dynamics and correlations of linked
-selected variants are also predicted to deviate from expectations without
-interference. Under a multiplicative fitness model, where the fitness reduction
-of a genome carrying multiple deleterious variants is equal to the product of
-the fitness reduction of each mutation independently, we expect to see net
-linkage disequilibrium (LD) equal to zero for unlinked sites
-\citep{Kondrashov1995-va}. But for linked loci, those mutations are expected to
-segregate on different haplotypes more often than together, leading to
-negative, or repulsion, LD, although the extent of LD depends non-trivially on
-the strength of selection and probability of recombination separating loci
-\citep{Hill1966-gv, McVean2000-ox}.
+selected variants are also predicted to deviate from single-locus expectations.
+Under a multiplicative fitness model, where the fitness reduction of a genome
+carrying multiple deleterious variants is equal to the product of the fitness
+reduction of each independent mutation, we expect to see net linkage
+disequilibrium (LD) equal to zero for unlinked sites \citep{Kondrashov1995-va}.
+But for linked loci, those mutations are expected to segregate on different
+haplotypes more often than together, leading to negative, or repulsion, LD,
+although the extent of LD depends non-trivially on the strength of selection
+and probability of recombination separating loci
+\citep{Hill1966-gv,McVean2000-ox}.
 
@@
 epistasis differentially impacts the deleterious load in asexually and sexually
-reproducing organisms \citep{Kimura1966-gc, Kondrashov1995-va}, has been
-invoked as an explanation for the evolutionary advantage of sex
-\citep{Kondrashov1982-sf, Charlesworth1990-kw, Barton1998-lu}, and can drive
+reproducing organisms \citep{Kimura1966-gc,Kondrashov1995-va}, has been invoked
+as an explanation for the evolutionary advantage of sex
+\citep{Kondrashov1982-sf,Charlesworth1990-kw,Barton1998-lu}, and can drive
 incompatibilities that lead to postzygotic isolation during the process of
@@
 importance of epistatic interactions, we currently lack models for predicting
-correlation patterns between linked mutations under general selection models.
+patterns of correlations between linked mutations under general selection
+models.
 
@@
 sequencing data and compare patterns of LD for synonymous, missense, and
-loss-of-function mutations within and between protein-coding genes and
-annotated domains. I show that while synonymous and missense variants display
-similar slightly positive average LD within genes, for missense mutations this
-signal is driven by correlations between pairs of mutations within, but not
-between or outside of, protein-coding domains.
-This suggests an importance for antagonistic (or ``diminishing-returns'')
-epistasis or a prevalence of compensatory
-nonsynonymous mutations within conserved elements.
-%Finally, I discuss how well
-%we can expect to discriminate between modes of selective interactions using
-%genome-wide patterns of LD.
+loss-of-function mutations in protein-coding genes and annotated conserved
+domains. I show that while synonymous and missense variants display similar
+slightly positive average LD within genes, for missense mutations this signal
+is driven by correlations between pairs of mutations within, but not between or
+outside of, protein-coding domains. This suggests an importance for
+antagonistic (or ``diminishing-returns'') epistasis or a prevalence of
+compensatory nonsynonymous mutations within conserved elements.
 
@@
 expected from multiplying the independent observed effects of each individual
-mutation. The scale of mutation scanning experiments continues to improve
-dramatically, promising greater resolution of the fitness landscape in such
-model systems that can be compared to evolutionary theory
-\citep{Otwinowski2018-zk}.
+mutation. The scale of mutation scanning experiments continues to increase,
+promising greater resolution of the fitness landscape in such model systems
+that can be compared to evolutionary theory \citep{Otwinowski2018-zk}.
 
@@
 mutations cluster together more than expected along the same lineages in the
-\emph{Drosophilid} species complex, and that those clustered mutations tended
-to preserve the charge of the protein and were in positive LD compared to pairs
-involving synonymous mutations. From this, they proposed that compensatory
-nonsynonymous variants are regularly tolerated and maintained. More recently,
-\citet{Taverner2020-lk} replicated this finding across a diverse set of genera,
-showing that such epistatic interactions are important for protein evolution.
+\emph{Drosophilid} species complex, and that those clustered mutations tend to
+preserve the charge of the protein and were in positive LD compared to pairs of
+synonymous mutations at the same distance. From this, they proposed that
+compensatory nonsynonymous variants are regularly tolerated and maintained.
+More recently, \citet{Taverner2020-lk} replicated this finding across a diverse
+set of genera, showing that such epistatic interactions are important for
+protein evolution.
 
@@
 frameshifts, and other nonsense mutations) in both human and fruit fly
-populations. They interpreted this as evidence for widespread synergistic
+populations. This was interpreted as evidence for widespread synergistic
 epistasis between these mutations, in which the fitness reduction of multiple
@@
         Low-order summaries of sampling distributions are commonly computed
-        for allele allele frequencies (A, the site-frequency spectrum, SFS)
+        for allele frequencies (A, the site-frequency spectrum, SFS)
         and two-locus haplotype distributions (B, linkage disequilibrium, LD).
@@
         and observations of nonzero values of Tajima's $D$ or signed LD
-        ($\sigma_d^1$) are often taken as evidence for selection or interations
+        ($\sigma_d^1$) are often taken as evidence for selection or interactions
         between loci, respectively.
@@
         pairs of neutral mutations \citep{Hill1968-vu}. Interference between linked
-        selected mutation causes negative LD \citep{Hill1966-gv}, and other
+        selected mutation causes negative signed LD \citep{Hill1966-gv}, and other
         forms of interactions between selected mutations can cause large negative
         or positive signed LD.
-        \comment{selected in blue to match all panels, panel labels, and panel A
-        should include neutral expectation}
     }
@@
 \citet{Ragsdale2019-nt} to compute the expected sampling distribution
-\(\Psi_n\) of two-locus haplotypes under a general model of selective
-interactions (Figure~\ref{fig:TwoLocusFS}). This sampling distribution stores the
+\(\Psi_n\) (Figure~\ref{fig:TwoLocusFS}) of two-locus haplotypes under a
+general model of selective interactions. This sampling distribution stores the
 expected density or observed counts of pairs of biallelic loci with each
@@
 interaction between pairs of loci. It is not possible in this framework to
-include the effects of additional linnked selected mutations, such as background
-selection due to many linked variants. Individual-based forward simulations are
-still needed for such scenarios (e.g., Figures~\ref{fig:bgs1}--\ref{fig:bgs4}).
+include the effects of additional linked selected mutations, such as background
+selection due to many linked variants, and individual-based forward simulations
+are still needed for such scenarios (e.g.,
+Figures~\ref{fig:bgs1}--\ref{fig:bgs4}).
 
 In many cases it is simpler to visualize summaries such as the expectation or
-variance of \(D\) (e.g., Figure~\ref{fig:TwoLocusFS}D,E) or slices of the
-distribution (e.g., Figure~\ref{fig:TwoLocusFS}C) instead of the full
+variance of \(D\) (Figure~\ref{fig:TwoLocusFS}D, E) or conditional slices of the
+distribution (Figure~\ref{fig:TwoLocusFS}C) instead of the full
 three-dimensional distribution. Here I focus on low-order LD statistics
@@
 Instead of unnormalized \(\E[D^2]\) and \(\E[D]\) I consider expectations for
-\(\sigma_d^2 = \E[D^2]/\E[f_A(1-f_A)f_B(1-f_B)]\) and
-\(\sigma_d^1 = \E[D]/\E[f_A(1-f_A)f_B(1-f_B)]\).
-Such normalized statistics have two benefits:
-first, the mutation rate cancels so that expectations are robust to assumptions
-about the per-base mutation rate, and second, analytic expectations for these
-quantities are available under neutrality at constant population size for
-comparison \citep{Ohta1971-yd}.
+\(\sigma_d^2 = \E[D^2]/\E[f_A(1-f_A)f_B(1-f_B)]\) and \(\sigma_d^1 =
+\E[D]/\E[f_A(1-f_A)f_B(1-f_B)]\).  Such normalized statistics have two
+benefits: first, the mutation rate cancels so that expectations are robust to
+assumptions about the per-base mutation rate, and second, we can compare to
+analytic expectations for these quantities under neutrality and constant
+population size \citep{Ohta1971-yd}.
 
@@
         or epistatic effects.
-        (A) The expected normalized variance of $D$ decreases as the strength
-        of negative selection increases.
+        (A) The expected normalized variance of $D$ (\(\sigma_d^2\))
+        decreases below neutral expectations
+        as the strength of negative selection increases.
         (B, C) For tightly linked loci (\(4Nr \lesssim 1\)), interference is most
-        noticeable for pairs of mutations with \(s \approx 1/N\) as measured by
-        signed LD.
+        noticeable as measured by signed LD for pairs of mutations with
+        \(s \approx 1/N\).
         At larger recombination distances (\(4Nr > 1\)), signed LD is most
-        negative for stronger selection coefficients. Dashed lines
-        show neutral expectations.
+        negative for somewhat stronger selection coefficients.
+        Dashed lines show neutral expectations.
     }
@@
         show neutral expectations.
-        For both weak selection (\(s=-1/2N\), A and B) and stronger selection
-        (\(s=-10/2N\), C and D), antagonistic epistasis (\(\epsilon < 0\))
+        For both weak (\(s=-1/2N\), A and B) and moderate
+        (\(s=-10/2N\), C and D) selection,
+        antagonistic epistasis (\(\epsilon < 0\))
         gives rise to positive signed LD and increased $\sigma_d^2$ over
@@
         LD beyond that of Hill-Robertson interference and decreased \(\sigma_d^2\).
-        With epistasis between pairs of selected mutations, nonzero signed LD
     }
@@
 selection coefficients \(\gamma = O(1)\), or \(s \approx 1/2N_e\) (Figure
-\ref{fig:epistasis}B).
+\ref{fig:HillRobertson}).
 For strongly deleterious mutations, LD is close to zero even
@@
         epistasis (Figure~\ref{fig:epistasis}).
-        (C, D) Recessive strongly deleterious mutations can have positive signed
+        (C) Interference effects are most pronounced for recessive deleterious
+        variants.
+        (D, E) Recessive strongly deleterious mutations can have positive signed
         LD, as recently shown by \citet{Roze2021-cf}.
@@
         selected mutations can show non-monotonic behavior as $h$ varies.
-        \comment{C and D labels misplaced}
+        Selection parameters of \(sh=-0.1\) imply extremely strong selection
+        (\(h=0.05\) results in \(s=-0.2\) and \(\gamma=-400\) for homozygous
+        diploids at a single locus). The numerical appraoch for \(\Psi_n\)
+        cannot handle such strong selection.
     }
@@
 
-%The effects of non-additive selection within a locus and interference has received
-%less attention. \citet{Garcia2020-zn} used large-scale forward simulations to explore
-%the effect of dominance on patterns of LD, showing that LD depends on both the
-%magnitude of the selection and dominance coefficients in a nonlinear way, but they
-%note that expensive large-scale simulations are needed to explore these effects
-%as no closed form approach to computing those expectations exist.
-%The numerical method presented here provides such a closed form approach, so that expectations of signed LD and other two-locus statistics can be computed more
-%efficiently and accurately.
-\begin{itemize}
-    \item Previous work on dominance and LD between linked loci
-        \begin{itemize}
-            \item \citep{Garcia2021-zn} - large-scale forward simulations to explore the
-                effect of dominance on patterns of LD, showing that LD depends on both 
-                the magnitude of the selection and dominance coefficients in a nonlinear way
-            \item \citep{Roze2021-cf} -
-                analytic expression for LD between pairs of recessive
-                strongly deleterious mutations under steady-state demography, showing that
-                for $h<0.25$, you can get slightly positive signed LD
-        \end{itemize}
-    \item confirming findings from \citet{Garcia2021-zn} that LD depends in a
-        complicated way on the strength of both dominance and selection
-    \item confirmed through discrete simulations and numerical approach that positive LD
-        can occur for strong negative selection with small $h$
-    \item however, Roze's result seems to hold only for strongly negative selection,
-        and the $h=0.25$ threshold appears to be a limiting value for that threshold,
-        as weak and moderate selection show strikingly different patterns
-    \item While the model derived by \citeauthor{Roze2021-cf} showed signed LD is
-        positive when $h<0.25$ and negative when $h>0.25$, this appears to 
-\end{itemize}
+The effect of non-additive selection on correlations between mutations has
+received increased attention recently. For example, \citet{Garcia2021-zn} used
+large-scale forward simulations to explore how dominance impacts patterns of
+LD, showing that LD depends on the magnitudes of both the selection and
+dominance coefficients in a nonlinear way. \citet{Roze2021-cf} found an
+analytic expression for LD between pairs of strongly deleterious mutations
+under steady-state demography, showing that LD can be either positive or
+negative depending on the strength of dominance.
 
 The combined effect of the strength of selection and dominance on interference
-is indeed nontrivial, and some parameter regimes can cause strong negative LD
-between pairs of negatively selected variants. For strong selection, pairs of
-recessive mutations (\(h<1/2\)) can have large negative signed LD (Figure
-\ref{fig:dominance}B), although it decays rapidly with increasing distance between
-loci and is roughly zero for \(\rho \gg 1\). For weakly selected mutations with
-\(\gamma=-1\), we see that there is no monotonic effect of the level of dominance
-on negative LD, with both recessive and dominant pairs of mutations having more
-negative LD than the additive case.
+is indeed nontrivial (Figure~\ref{fig:dominance}A), as observed by
+\citet{Garcia2021-zn}. Some parameter regimes can cause strong negative LD
+between pairs of negatively selected variants, with moderately selected
+recessive variants having stronger signals of interference than additive
+selection (Figure~\ref{fig:dominance}B). Unlike epistatic interactions, signed
+LD decays rapidly with increasing distance between loci and is roughly zero for
+\(\rho \gg 1\). For weakly selected mutations with \(\gamma=-1\), there is no
+monotonic effect of the level of dominance on negative LD, with both recessive
+and dominant pairs of mutations having more negative LD than the additive case.
+
+Discrete simulations and the moments approach confirm the result from
+\citet{Roze2021-cf}, that positive LD can occur for strong negative selection
+and small values of \(h\) (Figure~\ref{fig:dominance}D, E). However, while the
+analytic formula in \citet{Roze2021-cf} predicts that LD should be positive for
+\(h<0.25\) and negative for \(h>0.25\), this appears to only hold in the limit
+of strong selection (compare to Figure 1A in \citet{Roze2021-cf}). For moderate
+to moderately strong selection, this threshold of \(h\) can be less than
+\(0.25\), and LD is negative for all \(0\leq h \leq 1\) for weakly deleterious
+mutations with interference strongest between pairs of recessive mutations
+(Figure~\ref{fig:dominance}C).
 
@@
         Dashed lines indicate neutral expectations.
-        \comment{need panel labels}
     }
@@
         Dashed lines indicate neutral expectations.
-        \comment{Need panel labels}
     }
@@
         \textbf{LD in human protein-coding genes and annotated domains.}
-        \comment{Needs panel labels}
+        (A) Gene-wide averages of signed LD are slightly positive between pairs
+        of both missense and synonymous mutations. This positive, equal
+        \(\sigma_d^1\) is also observed when conditioning on common variants
+        (minor allele frequencies \(\gtrsim 0.05\)).
+        (B) While there are relatively fewer pairs of loss-of-function mutations
+        within genes, causing larger measurement uncertainty, such pairs have
+        negative average LD. Note that measurement noises for each class of
+        mutations overlap with zero and with each other.
+        (C) Partitioning pairs of mutations as falling within or outside
+        of conserved domains reveals opposing patterns of signed LD, with
+        \(\sigma_d^1\) between missense mutations larger than that of synonymous
+        mutations within domains. Outside of conserved domains, missense mutations
+        have reduced LD compared to synonymous mutations. Distances of pairs outside
+        of domains were matched to within-domain mutation pair distances.
+        (D) This pattern again holds for common variants.
+        (E, F) The signal of increased LD between missense variants within domains
+        and decreased LD outside of domains is driven largely by tightly linked
+        mutations (distances \(\lesssim 400\) bp), here showing African populations
+        in the \citet{1000_Genomes_Project_Consortium2015-zq}.
+        Additional comparisons are shown in the Supporing Information, and Thousand
+        Genomes population labels are described in Table~\ref{tab:1kgpops}.
     }
@@
 
-%\caption{\textbf{LD within and between protein-coding genes.} FIXME (A-B) Pairs of
-%variants in coding regions partitioned by classes of mutations.
-%Loss-of-function mutations are defined as stop and start gains and losses and
-%other nonsense mutations of high impact. In all populations, we observe
-%positive mean signed LD for missense and synonymous variants, although the
-%confidence interval of each measurement overlaps with zero. Similarly, we
-%observe mean negative LD for pairs of loss-of-function variants within genes.
-%There are only \(\sim200-300\) pairs of loss-of-function variants that fall
-%within the same gene across the populations tested here, causing large
-%uncertainty in each measurement. Gray lines indicate 2 standard errors of the
-%measurement, computed by bootstrapping of sets of genes. (C-F) While pairs of
-%synonymous and missense mutations within genes are consistently positive, pairs
-%between genes are not consistent in their sign across populations. LD for all
-%populations analyzed here is shown in Figure \ref{fig:genespops}.
-%\textbf{LD within and between annotated domains in protein-coding
-%genes.} (A) Considering only variants that fall within annotated domains in
-%protein-coding genes, pairs of mutations that fall within the same gene display
-%larger positive LD than pairs that lie in different domains. This pattern is
-%most striking for missense mutations, for which pairs within domains have
-%average signed LD that is consistently larger than for pairs that span two
-%domains, or for pairs of synonymous mutations that are within the same domain.
-%(B-D) The slightly positive LD that we observe for missense and synonymous
-%mutations gene-wide (Figure \ref{fig:fig3}A) is largely driven by the subset
-%of pairs mutations that fall within the same annotated domain. LD for all
-%populations analyzed here isshown in Figure \ref{fig:domainspops}.}\label{fig:LDdata}
-%\end{figure}
-
-Synonymous and missense mutations show similar levels of positive signed LD
-when considering pairs of mutations within the same gene. Loss-of-function
-mutations have more negative LD, possibly suggesting differing modes of
-selective interactions for loss-of-function and missense mutations (Figure
-\ref{fig:data}A--B). Within each population, measurement noise gives 95\%
-confidence intervals that overlap with zero in each mutation class, although
-the pattern is remarkably consistent across African, European, and East Asian
-populations in the Thousand Genomes dataset. Comparing mean LD within genes in
-African and Eurasian populations, LD in Eurasian populations is somewhat larger
-on average, that is, more positive for missense mutations and more negative for
+Here, I examine patterns of signed LD between mutations in human protein-coding
+genes partitioned by functional annotations. Synonymous and missense mutations
+show similar levels of slighly positive signed LD when considering pairs of
+mutations within the same gene averaged over all autosomal chromosomes.
+Loss-of-function mutations have more negative LD, possibly suggesting differing
+modes of selective interactions for loss-of-function and missense mutations
+(Figure \ref{fig:data}A--B). Within each population, measurement noise gives
+95\% confidence intervals that overlap with zero in each mutation class,
+although the observed patterns are remarkably consistent across African,
+European, and East Asian populations in the Thousand Genomes dataset. Comparing
+mean LD across populations, LD in Eurasian populations is somewhat larger on
+average, that is, more positive for missense mutations and more negative for
 loss-of-function mutations. This is in agreement with differences in
 expectations between populations that have or have not gone through a
-bottleneck in their recent past
+bottleneck in their recent history
 (Figures~\ref{fig:toy}~and~\ref{fig:relate_dom}).
@@
 The similarity in signed LD between missense and synonymous mutations would
-suggest that interference between missense mutations is minimal. However,
-interactive effects differ dramatically between pairs of mutations found in
-intra-genic regions, with these opposing effects canceling out when taking
-gene-wide averages. Due to the rarity of loss-of-function mutations, I only
-compare synonymous and missense mutations when looking at finer partitions of
-mutations within genes.
-
-Here, I examine the role that annotated conserved domains in protein-coding
-genes play in driving signals of positive LD within genes. Protein-coding
-domains are conserved elements of genes, often associated with some known
-functional or structural feature of a protein \citep{Stanek2020-pa}. Purifying
-selection is expected to be stronger within conserved domains than within the
-same gene but outside of those domains. Indeed, the site-frequency spectrum
-(SFS) is skewed to lower frequencies for both missense and loss-of-function
-mutations within domains when compared to the same classes of mutations outside
-of domains, with much more negative values of Tajima's D within domains (Table
+suggest that interference between missense mutations is minimal, or at least no
+stronger than interference between synonymous mutations. However, interactive
+effects differ dramatically between pairs of mutations found in different
+intra-genic regions, with opposing effects canceling out when taking gene-wide
+averages. Due to the rarity of loss-of-function mutations, I only compare
+synonymous and missense mutations when looking at finer partitions of mutations
+within genes.
+
+Annotated conserved domains in protein-coding genes play a significant role in
+driving signals of positive LD within genes. Such protein-coding domains are
+conserved elements of genes, often associated with some known functional or
+structural feature of a protein \citep{Stanek2020-pa}. Purifying selection is
+expected to be stronger within conserved domains than within the same gene but
+outside of those domains. Indeed, the site-frequency spectrum (SFS) is skewed
+to lower frequencies for both missense and loss-of-function mutations within
+domains when compared to the same classes of mutations outside of domains, with
+much more negative values of Tajima's D within domains (Table
 \ref{tab:tajimasD}). On the other hand, no difference is observed for
@@
 Pairs of missense mutations that both fall within the same functional domain
-have large positive LD that is much larger than pairs of synonymous mutations
-that both fall within the same domain
+have large positive LD that is elevated above that of pairs of synonymous
+mutations that both fall within the same domain
 (Figures~\ref{fig:data}C--F~and~\ref{fig:LDwithin}--\ref{fig:domainsEAS}). This
 difference between pairs of missense and synonymous variants within domains is
-observed for both tightly linked (\(\lesssim 100\)bps) and more distant
-(\(\gtrsim 1\)kb) pairs (Figure \ref{fig:LDwithin}). Assuming most
+especially pronounced for both linked pairs (\(\lesssim\) a few hundred base
+pairs) (Figures~\ref{fig:data}E~and~\ref{fig:LDwithin}). Assuming most
 synonymous mutations are neutral, we would expect signed LD between missense
@@
 than those that are commonly tested or modeled.
-\comment{Due to the observed differences in levels of selection inside and outside domains, frequency matching - result holds}
-
-The large positive signal of LD for missense mutations within the same
-domain does not extend to pairs of missense mutations that are in
-different domains. Pairs of missense and synonymous mutations show
-nearly equal levels of LD close to zero across domains, with missense mutations
-even slightly more negative than synonymous mutations (Figure \ref{fig:XXX}).
-\comment{need to do this analysis again}
-The interactive effect driving large LD within domains is therefore likely
-domain-specific. However, the average distance between pairs of mutations
-within domains is much smaller than between domains, so this observation may be
-primarily driven by the higher recombination distances between pairs of mutations
-across distinct domains.
+
+The strength of selection on missense mutations within and outside of domains
+is observed to differ, leading to an excess of rare missense mutations within
+conserved domains (Table~\ref{tab:tajimasD}). LD is known to be sensitive to
+allele frequencies, with rare mutations showing large positive signed LD
+\citep{Good2022-ot}. To test whether the signal of increased LD between
+missense mutations within domains is driven by rare variants, I considered
+subsets of pairs of mutations based on their derived allele frequences
+(Figures~\ref{fig:data}D~and~\ref{fig:domainsAFR}--\ref{fig:domainsEAS}). Rare
+and uncommon variants show large average LD for each class of mutations, but
+common variants recapitulate the opposing patterns of LD that is seen when
+considering pairs at all frequencies. This is because unconditioned statistics
+in the form of \(\sigma_d^1\) and \(\sigma_d^2\) are already dominated by
+common variants.
 
 \subsubsection{Reduced LD between pairs of missense mutations outside of conserved domains}
+
+The large positive signal of LD for missense mutations within the same domain
+does not extend to pairs of missense mutations that are in different domains.
+Pairs of missense and synonymous mutations show nearly equal levels of LD close
+to zero across domains, with missense mutations slightly more negative than
+synonymous mutations (Figure \ref{fig:between_domains}). The interactive effect
+driving large LD within domains is therefore likely domain-specific. However,
+the average distance between pairs of mutations within domains is much smaller
+than between domains, so this observation may be primarily driven by the higher
+recombination distances between pairs of mutations across distinct domains.
 
@@
 domain. For pairs of mutations outside of domains but with distances matched to
-those within domains, synonymous mutations have larger positive LD than
-missense mutations. More distant pairs of mutations outside of domains, matched
-to the same distances as the between-domain comparison, each have LD roughly
-equal to zero. \comment{Todo}
+those within domains, pairs of synonymous mutations have larger positive LD
+than missense mutations. More distant pairs of mutations outside of domains,
+matched to the same distances as the between-domain comparison, each have LD
+roughly equal to zero.
 
@@
 
-By taking broad genome- or gene-wide surveys of LD across functional classes of
-mutations, heterogeneous patterns of interactive effects that occur within
+By taking broad genome- and gene-wide surveys of LD across functional classes
+of mutations, heterogeneous patterns of interactive effects that occur within
 genes can be missed. From gene-wide averages, LD between pairs of missense
@@
 although LD between loss-of-function variants is more negative. Hill-Robertson
-effects under additive selection are expected to be strongest for slightly to
-moderately deleterious variants (with \(|s|\sim 1/N_e\)), as strongly deleterious
-mutations are not expected to interfere with one another \citep{McVean2000-ox}.
-However, loss-of-function mutations do not typically fall within this regime,
-and inference of the distribution of fitness effects (DFE) for new
-loss-of-function variants shows that a large majority are strongly deleterious
-(Supporting Information). Instead, negative synergistic epistasis between
-strongly deleterious mutations does produce large negative deviations of mean
-LD. Strongly deleterious \emph{recessive} mutations can also produce this
-pattern, making it difficult to distinguish these effects with certainty. This
+effects are expected to be strongest for slightly to moderately deleterious
+variants (with \(|s|\sim 1/N_e\)), as strongly deleterious mutations are not
+expected to interfere with one another \citep{McVean2000-ox}. However,
+loss-of-function mutations do not typically fall within this regime, and
+inference of the distribution of fitness effects (DFE) for new loss-of-function
+variants shows that a large majority are strongly deleterious (Supporting
+Information). Instead, negative synergistic epistasis between strongly
+deleterious mutations does produce large negative deviations of mean LD.
+Strongly deleterious \emph{recessive} mutations can also produce this pattern,
+making it difficult to distinguish these effects with certainty. This
 difficulty is compounded by the large measurement noise for \(\E[D]\),
@@
 \subsection{Non-uniform interactions between selected mutations within genes}
-\label{non-uniform-interactions-between-selected-mutations-within-genes}
 
 Positive average LD between both missense and synonymous mutations has
-previously been reported in humans and other species \citep{Sohail2017-zq,
-Garcia2020-zn, Sandler2021-of}. The similarity of their gene-wide LD
-(Figure~\ref{fig:data}) might suggest that interference or interactions between
-missense mutations are minimal, or at least no stronger than those between
-synonymous mutations. However, averaging over all observed pairs of mutations
-within a gene masks element-specific interactive effects that drive LD in
-opposite directions. Nonsynonymous mutations found within conserved protein
-domains, identified as conserved subunits of a gene with some structural or
-functional role \citep{Stanek2020-pa}, are more strongly selected against on
-average, but also have increased signed LD over synonymous mutations at the
-same distances within domains. Missense mutation outside of domains but at the
-same distances as those within domains have more negative LD, both compared to
-distance-matched synonymous mutations outside of domains and to mutations
-within domains. Neither dominance effects, synergistic epistasis, nor
+previously been reported in humans, drosophila, and other species
+\citep{Sohail2017-zq,Garcia2021-zn,Sandler2021-of}. The similarity of their
+gene-wide LD (Figure~\ref{fig:data}) might suggest that interference or
+interactions between missense mutations are minimal, or at least no stronger
+than those between synonymous mutations. However, averaging over all observed
+pairs of mutations within a gene masks element-specific interactive effects
+that drive LD in opposite directions. Nonsynonymous mutations found within
+conserved protein domains, identified as conserved subunits of a gene with some
+structural or functional role \citep{Stanek2020-pa}, are more strongly selected
+against on average, but also have increased signed LD over synonymous mutations
+at the same distances within domains. Missense mutation outside of domains but
+at the same distances as those within domains have more negative LD, both
+compared to distance-matched synonymous mutations outside of domains and to
+mutations within domains. Neither dominance effects, synergistic epistasis, nor
 Hill-Robertson interference are expected to result in positive LD, so some
@@
 subunit, but additional mutations within that same domain reduce fitness by a
-factor less than the first mutation. A third possibility, related to the
-antagonistic epistasis case, is that selection acts on the functional domain as
-a unit instead of at loci within the domain individually (such as under a model
-of gene-based dominance). In this scenario, double heterozygotes have different
-fitnesses depending on whether the mutations are found on same haplotype or on
-different haplotypes.
+factor less than the first mutation. A third possibility, related to
+antagonistic epistasis, is that selection acts on the functional domain as a
+unit instead of on mutations within the domain individually (such as under a
+model of gene-based dominance). In this scenario, double heterozygotes have
+different fitnesses depending on whether the mutations are found on same
+haplotype or on different haplotypes.
 
@@
 missense pairs of variants largely disappears for SNPs separated by more than a
-few hundred base pairs (Figure \ref{fig:lddecay}). This suggests that
+few hundred base pairs (Figure~\ref{fig:data}E, F). This suggests that
 Hill-Robertson interference is the primary mode of interaction between
@@
 average over these differences by considering all nonsynonymous mutations
-together \citep{Boyko2008-zk,Kim2017-xo}, but if loss-of-function variants are in
-fact recessive, even partially, this can impact inferences of the DFE
-(Supporting Information). It would be straightforward to adapt DFE-inference
-methods to infer more detailed and accurate representation of the heterogeneous
-effects of new mutations within genes by partitioning by missense and
-loss-of-function classes as well as by annotated domains. Similarly, the
-different modes of interaction between mutations across annotated domains mean
-that our standard models and simulation approaches may be too simple and will
-not capture the evolutionary trajectories and patterns of diversity that differ
-at finer scales.
-
-\subsection{Clustered mutations account for positive LD among synonymous mutations}
-\label{clustered-mutations-account-for-positive-ld-among-synonymous-mutations}
+together \citep{Boyko2008-zk,Kim2017-xo}. If patterns of selective differ by
+both mutational class and location or if loss-of-function variants are in fact
+recessive, this can impact inferences of the DFE (Supporting Information). It
+would be straightforward to adapt DFE-inference methods to infer more detailed
+representation of the heterogeneous effects of new mutations within genes by
+partitioning by missense and loss-of-function classes as well as by annotated
+domains. Similarly, the different modes of interaction between mutations across
+annotated domains mean that our standard models and simulation approaches may
+be too simple to capture the evolutionary trajectories and patterns of
+diversity that differ at finer scales.
+
+\subsection{Clustered mutations may cause positive LD among synonymous mutations}
 
@@
 equally likely to be positive or negative, and resulting \(\E[D]\) is still
-zero when taking genome-wide averages. Instead, \citet{Sohail2017-zq} and
-\citet{Sandler2021-of} used forward simulations under multi-population or
-explicitly spatial models and a multiplicative fitness function to explore the
-joint effect of structure, assortative mating, and interference, scenarios that
-can lead to nonzero LD. For example, \citet{Sandler2021-of} found positive LD
-among neutral mutations in a simulation with admixture and linked selected
-mutations, showing that the combined effects of spatial structure and background
-selection can lead to positive average LD between neutral mutations.
-\comment{Because the observed positive LD decays rapidly in the data, this
-would be worth exploring in more detail with forward simulation approaches}
-\comment{split into two paragraphs}
+zero when taking genome-wide averages.
+
+Instead, \citet{Sohail2017-zq} and \citet{Sandler2021-of} used forward
+simulations under multi-population or explicit spatial models and a
+multiplicative fitness function to explore the joint effects of structure,
+assortative mating, and interference, finding that together these can lead to
+nonzero LD. For example, \citet{Sandler2021-of} found positive LD among neutral
+mutations in a simulation with admixture and linked selected mutations, showing
+that the combined effects of spatial structure and background selection can
+lead to positive average LD between neutral mutations. Positive LD between
+synonymous mutations is only observed at short distances in human data, which
+could be used to constrain such models that predict positive LD over varying
+distances.
 
@@
 observed decay of \(\sigma_d^1\) between synonymous variants (Supporting
-Information). \comment{results of this here?}
-We may therefore treat positive
-LD as the baseline expectation for tightly linked variants primarily due to
-clustered mutations, so that subsequent selective processes and interactions
-cause LD to deviate from that expectation, as seen for missense mutations
-across different sub-elements of genes. It may also be more appropriate to
-compare the negative LD observed between linked loss-of-function variants to
-that positive baseline expectation instead of zero, which would imply that they
-are more recessive or have stronger synergistic epistasis than from inferences
-assuming a neutral expectation of zero.
+Information, Figure~\ref{fig:mslmnms}), requiring only a small fraction of
+mutations to cause multinucleotide mutation events. We may therefore treat
+positive LD as the baseline expectation for tightly linked variants primarily
+due to clustered mutations, so that subsequent selective processes and
+interactions cause LD to deviate from that expectation, as seen for missense
+mutations across different sub-elements of genes. It may also be more
+appropriate to compare the negative LD observed between linked loss-of-function
+variants to that positive baseline expectation instead of zero, which would
+imply that they are more recessive or have stronger synergistic epistasis than
+from inferences assuming a neutral expectation of zero.
 
@@
 models, the number of plausible selection scenarios becomes quite large as we
-relax the strict asumptions of additivity and multiplicative fitness effects.
+relax the strict assumptions of additivity and multiplicative fitness effects.
 The inclusion of dominance effects, epistasis, or other interactions leads to a
@@
 Instead, closed, numerical approaches to compute expectations for two-locus
-statistics under a wide range of selection models, as presented here or in
-\citet{Ragsdale2019-nt} and \citet{Friedlander2022-bs}, allow us to explore
-this highly parameterized space of models far more efficiently, and it opens
-the possibility for performing likelihood-based inference using signed LD or
-other two-locus summaries of the data. For example, inferring the joint
-distribution of dominance and selection is underpowered using the SFS alone,
-but because signed LD is sensitive to the levels of dominance (Figure
-\ref{fig:dominance}), inferring the DFE with dominance may be feasible using
-the joint distribution of allele frequencies and LD.
+statistics under a wide range of selection models, as presented here and in
+\citet{Friedlander2022-bs}, allow us to explore this highly parameterized space
+of models far more efficiently, and it opens the possibility for performing
+likelihood-based inference using signed LD or other two-locus summaries of the
+data. For example, inferring the joint distribution of dominance and selection
+is underpowered using the SFS alone, but because signed LD is sensitive to the
+levels of dominance (Figure~\ref{fig:dominance}), inferring the DFE with
+dominance may be feasible using the joint distribution of allele frequencies
+and LD.
 
@@
 biasing any inference using previously inferred DFEs to learn about patterns of
-interactions. Again, this underlies the need to jointly infer strengths and
+interactions. Again, this underlines the need to jointly infer strengths and
 interactions of selected variants.
@@
 Finally, exploring additional summaries of the two-locus sampling distribution
-could provide additional power to distinguish between interaction models. The
-method developed here allows us to compute the full sampling distribution for a
-given sample size, and expectations for a large family of informative two-locus
+should provide additional power to distinguish between interaction models. The
+numerical methods developed here provide the full two-locus sampling
+distribution, and expectations for a large family of informative two-locus
 statistics can be computed directly from this distribution. These expectations
@@
 \subsection{Existing theory and numerical methods}
-\label{existing-theory-and-numerical-methods}
 
 Many well-known properties of two-locus dynamics and equilibrium LD come from
-early work on the multi-locus diffusion approximation \citep{Kimura1955-qe, Hill1968-vu, Ohta1969-ie, Ohta1971-yd}.
-This includes the result that
+early work on the multi-locus diffusion approximation \citep{Kimura1955-qe,
+Hill1968-vu, Ohta1969-ie, Ohta1971-yd}.  This includes the result that
 genome-wide averages of signed LD are expected to be zero under neutrality.
-Under a two-locus biallelic model, where the left locus allows alleles \(A\) and
-\(a\) and the right locus allows alleles \(B\) and \(b\), the standard covariance
-measure of LD is defined as \(D = f_{AB} - f_{A}f_{B}\), where \(f_{AB}\) is the
-haplotype frequency of double-derived types carrying both \(A\) and \(B\), and
-\(f_{A}\) and \(f_{B}\) are the marginal frequencies of the derived alleles at each
-locus. This covariance decays due to both drift and recombination at a rate
-proportional to the inverse of the effective population size and the distance
-separating loci \citep{Hill1968-vu}: \[\E[D]_{t+1} = \left(1 - \frac{1}{2N_e(t)} - r
-\right)\E[D]_t.\] While \(\E[D] = 0\), the variance of \(D\) is non-zero, and
-\citet{Ohta1971-yd} presented their groundbreaking result that the variance of \(D\)
-under neutrality and steady-state demography, normalized by the joint
-heterozygosity of the two loci, is \begin{equation} \label{eq:ohta} \sigma_d^2 =
-\frac{\E[D^2]}{\E[f_A(1-f_A)f_B(1-f_B)]}\approx\frac{5 + \rho / 2}{11 + 13\rho
-/ 2 + \rho^2 / 2}, \end{equation} where \(\rho = 4N_e r\).
+Under a two-locus biallelic model, where the left locus allows alleles \(A\)
+and \(a\) and the right locus allows alleles \(B\) and \(b\), the standard
+covariance measure of LD is defined as \(D = f_{AB} - f_{A}f_{B}\), where
+\(f_{AB}\) is the haplotype frequency of double-derived types carrying both
+\(A\) and \(B\), and \(f_{A}\) and \(f_{B}\) are the marginal frequencies of
+the derived alleles at each locus. This covariance decays due to both drift and
+recombination at a rate proportional to the inverse of the effective population
+size and the distance separating loci \citep{Hill1968-vu}:
+\[\E[D]_{t+1} = \left(1 - \frac{1}{2N_e(t)} - r \right)\E[D]_t.\]
+While \(\E[D] = 0\), the variance of \(D\) is non-zero, and \citet{Ohta1971-yd}
+presented their groundbreaking result that the variance of \(D\) under
+neutrality and steady-state demography, normalized by the joint heterozygosity
+of the two loci, is
+\begin{equation}
+    \label{eq:ohta}
+    \sigma_d^2 = \frac{\E[D^2]}{\E[f_A(1-f_A)f_B(1-f_B)]}
+    \approx\frac{5 + \rho / 2}{11 + 13\rho/ 2 + \rho^2 / 2},
+\end{equation}
+where \(\rho = 4N_e r\).
 
@@
 neutrality, that is, the probability of observing given counts of two-locus
-haplotypes in a sample of size \(n\) \citep{Golding1984-pu}. This approach would
-continue to be developed and later form the foundation for the inference of
-local recombination rates from population genetic data \citep{Hudson2001-sg, McVean2004-gj}. More recently, \citet{Song2007-qk} computed \(\E[r^2]\) using a
-diffusion approximation approach, although their solution involves the
-summation of infinitely many terms and is restricted to neutrality and
-steady-state demography.
+haplotypes in a sample of size \(n\) \citep{Golding1984-pu}. This approach
+would continue to be developed and later form the foundation for the inference
+of local recombination rates from population genetic data
+\citep{Hudson2001-sg,McVean2004-gj}. More recently, \citet{Song2007-qk}
+computed \(\E[r^2]\) using a diffusion approximation approach, although their
+solution involves the summation of infinitely many terms and is restricted to
+neutrality and steady-state demography.
 
 To include selection, however, there have been relatively few advances beyond
-the Monte Carlo simulation approach taken by \citet{Hill1966-gv}, albeit now with more
-powerful computational resources and sophisticated software for performing
-flexible forward simulation (e.g., \citet{Thornton2014-pn}, \citet{Haller2019-vm}). Analytic
-results for two-locus distributions under selection are notoriously difficult,
-with a few notable flashes of progress. For example, \citet{McVean2007-xd} considered
-the effect of a recent sweep on patterns of LD between neutral loci near the
-locus under selection, and in a recent paper, \citet{Good2020-wj} presented analytic
+the Monte Carlo simulation approach taken by \citet{Hill1966-gv}, albeit now
+with more powerful computational resources and sophisticated software for
+performing flexible forward simulation
+\citep[e.g.][0]{Thornton2019-qc,Haller2019-vm}. Analytic results for two-locus
+distributions under selection are notoriously difficult, with a few notable
+flashes of progress. For example, \citet{McVean2007-xd} considered the effect
+of a recent sweep on patterns of LD between neutral loci near the locus under
+selection, and in a recent paper, \citet{Good2022-ot} presented analytic
 solutions for patterns of LD between rare mutations under additive selection
@@
 applicability of two-locus statistics to demographic inference. More recently,
-\citet{Ragsdale2019-nt} extended the \citet{Hill1968-vu} system for \(\E[D^2]\) to compute
-arbitrary moments of the distribution of \(D\) for any number of populations
-connected by migration and admixture. They also showed that such a
+\citet{Ragsdale2019-nt} extended the \citet{Hill1968-vu} system for \(\E[D^2]\)
+to compute arbitrary moments of the distribution of \(D\) for any number of
+populations connected by migration and admixture. They also showed that such a
 moments-construction can be used to solve for the two-locus sampling
@@
 epistasis, and other forms of selective interactions between two loci. In a
-concurrent study to this paper, \citet{Friedlander2021-bs} developed numerical
+concurrent study to this paper, \citet{Friedlander2022-bs} developed numerical
 solutions to the same moments system, which they used to describe selected
@@
 \subsection{The two-locus sampling distribution with arbitrary selection}
-\label{the-two-locus-sampling-distribution-with-arbitrary-selection}
 
@@
 the two-locus distribution \(\Psi_n\) stores the density or number of pairs of
-loci with observed haplotype counts, so that \(\Psi_n(i, j, k)\) is the number of
-pairs for which we observe \(i\) copies of the \(AB\) haplotype, \(j\) of type \(Ab\),
-\(k\) of type \(aB\), and \(n-i-j-k\) of type \(ab\). The size of \(\Psi_n\) grows
-rapidly, with \(O(n^3)\) entries, which practically limits computational
-approaches to moderate sample sizes and a single population.
+loci with observed haplotype counts, so that \(\Psi_n(i, j, k)\) is the number
+of pairs for which we observe \(i\) copies of the \(AB\) haplotype, \(j\) of
+type \(Ab\), \(k\) of type \(aB\), and \(n-i-j-k\) of type \(ab\). The size of
+\(\Psi_n\) grows rapidly, with \(O(n^3)\) entries, which practically limits
+computational approaches to moderate sample sizes \(n\lesssim 100\) and a
+single population.
 
 Under neutrality, a number of approaches exist to compute \(\Psi_n\), including
-the recursion due to \citet{Golding1984-pu} and \citet{Ethier1990-tb}, or more recent
-numerical approaches to the two-locus coalescent \citep{Kamm2016-ag} or diffusion
-approximation \citep{Ragsdale2017-gg}. Selection is most easily included using the
-forward-in-time diffusion equation \citep{Kimura1955-qe, Hill1966-gv}, where a
-standard approach is to first solve for the continuous distribution \(\psi\) of
-the density of two-locus haplotype configurations in the full population, and
-then integrate \(\psi\) against the multinomial sampling function to obtain
-\(\Psi_n\).
-
-Alternatively, \citet{Ragsdale2019-nt} showed that there exists a system of ordinary
-differential equations directly on the entries of \(\Psi_n\), although that paper
-focused primarily on neutral dynamics and considered a model with selection at
-only one of the two loci. I briefly summarize this general approach below, but
-refer readers to that previous paper for detailed derivations of the drift,
-recombination, and mutation terms and the moment-closure approximation.
-Instead, here I focus on generalizing the selection operator to include
-epistasis, dominance, and other forms of two-locus interactions.
-
-\subsubsection{Moment equation for \(\Psi_n\)}\label{moment-equation-for-psi_n}
+the recursion due to \citet{Golding1984-pu} and \citet{Ethier1990-tb}, or more
+recent numerical approaches to the two-locus coalescent \citep{Kamm2016-ag} or
+diffusion approximation \citep{Ragsdale2017-gg}. Selection is most easily
+included using the forward-in-time diffusion equation
+\citep{Kimura1955-qe,Hill1966-gv}, where a standard approach is to first solve
+for the continuous distribution \(\psi\) of the density of two-locus haplotype
+configurations in the full population, and then integrate \(\psi\) against the
+multinomial sampling function to obtain \(\Psi_n\).
+
+Alternatively, \citet{Ragsdale2019-nt} showed that there exists a system of
+ordinary differential equations directly on the entries of \(\Psi_n\). I
+briefly summarize this general approach below, but refer readers to that
+previous paper for detailed derivations of the drift, recombination, and
+mutation terms and the moment-closure approximation. Instead, here I focus on
+generalizing the selection operator to include epistasis, dominance, and other
+forms of two-locus interactions.
+
+\subsubsection{Moment equation for \(\Psi_n\)}
 
@@
 from \(\Psi_n\) \citep[following the single-locus closure introduced in][]{Jouganous2017-pq}, so that \(\hat{\Psi}_{n+l}(i, j, k) = \mathcal{J}_{n, l}\Psi_n\), although other accurate closure approximations are possible
-\citep{Friedlander2021-bs}. This emits a closed approximate system,
+\citep{Friedlander2022-bs}. This emits a closed approximate system,
 \begin{equation}
@@
 \subsubsection{Selection models with epistasis and dominance}
-\label{selection-models-with-epistasis-and-dominance}
 
@@
 model, with acceptance and rejection probabilities depending on the fitnesses
-of each haploid copy, where haplotype \(Ab\) has fitness \(1 + s_{A}\), \(aB\) has
-fitness \(1 + s_{B}\), and \(AB\) has fitness \(1 + s_{AB}\). We assume the doubly
-ancestral haplotype \(ab\) has fitness \(1\), so fitnesses are relative to that of
-\(ab\) haplotypes. The standard multiplicative fitness function assumes that
-\(s_{AB} \approx s_{A} + s_{B}\), and a model for epistasis can be written as
-\[s_{AB} = (s_{A} + s_{B}) (1 + \epsilon),\] so that \(\epsilon > 0\) implies
-synergistic epistasis, while \(\epsilon < 0\) implies antagonistic epistasis.
+of each haploid copy, where haplotype \(Ab\) has fitness \(1 + s_{A}\), \(aB\)
+has fitness \(1 + s_{B}\), and \(AB\) has fitness \(1 + s_{AB}\). We assume the
+doubly ancestral haplotype \(ab\) has fitness \(1\), so fitnesses are relative
+to that of \(ab\) haplotypes. The standard multiplicative fitness function
+assumes that \(s_{AB} \approx s_{A} + s_{B}\) (\(s^2\approx0\)), and a model
+for epistasis can be written as \[s_{AB} = (s_{A} + s_{B}) (1 + \epsilon),\] so
+that \(\epsilon > 0\) implies synergistic epistasis, while \(\epsilon < 0\)
+implies antagonistic epistasis.
 
@@
 a lineage is rejected, a replacement is drawn from the full population. Under
-the assumption that \(ns_{\cdot\cdot} \ll 1\), the probability that more than one
-selection event occurs in any given generation is negligibly small, so that the
-case of multiple simultaneous rejections can be ignored. Then \(\Psi_n^{t+1}\)
-relies only on \(\Psi_n^t\) and \(\Psi_{n+1}^t\) for additive selection. The full
+the assumption that \(ns \ll 1\), the probability that more than one selection
+event occurs in any given generation is negligibly small, so that the case of
+multiple simultaneous rejections can be ignored. Then \(\Psi_n^{t+1}\) relies
+only on \(\Psi_n^t\) and \(\Psi_{n+1}^t\) for additive selection. The full
 selection operator \(\mathcal{S}\) for additive selection is given in the
@@
 fitness of an individual carrying an \(Ab\) haplotype depends on whether their
-second haplotype is \(ab\), \(Ab\), \(aB\), or \(AB\). We can therefore assign a
-selection coefficient to each possible diploid configuration, \(s_{Ab/ab}\),
-\(s_{Ab/Ab}\), and so on. Assuming that the doubly homozygous ancestral \(ab/ab\)
-genotype has relative fitness 1, this gives nine possible unique selection
-coefficients in the most general two-locus selection model. Note that \(AB/ab\)
-and \(Ab/aB\) genotypes need not have the same selection coefficient, which
-allows for simulation under a gene-based model of dominance (Table
+second haplotype is \(ab\), \(Ab\), \(aB\), or \(AB\). We can therefore assign
+a selection coefficient to each possible diploid configuration, \(s_{Ab/ab}\),
+\(s_{Ab/Ab}\), and so on. Assuming that the doubly homozygous ancestral
+\(ab/ab\) genotype has relative fitness 1, this gives nine possible unique
+selection coefficients in the most general two-locus selection model. Note that
+\(AB/ab\) and \(Ab/aB\) genotypes need not have the same selection coefficient,
+which allows for simulation under a gene-based model of dominance (Table
 \ref{tab:selmodels}).
@@
 determines the probability that we reject the focal haplotype. We again assume
-that \(ns \ll 1\) for all genotype selection coefficients, so that we may assume
-at most a single selection event occurs in any given generation. This means
-that to find \(\Psi_n^{t+1}\) under a general two-locus selection model, which
-encompasses dominance within either locus, gene-based dominance, or a
-combination or dominance and epistatic effects, we need \(\Psi_{n+2}^t\). Again,
-a full derivation and expressions for the general selection operator are given
-in the Supporting Information.
+that \(ns \ll 1\) for all genotype selection coefficients, so that we may
+assume at most a single selection event occurs in any given generation. This
+means that to find \(\Psi_n^{t+1}\) under a general two-locus selection model,
+which encompasses dominance within either locus, gene-based dominance, or a
+combination or dominance and epistatic effects, we need \(\Psi_{n+2}^t\).
+Again, a full derivation and expressions for the general selection operator are
+given in the Supporting Information.
 
 \subsubsection{Low-order summaries of the sampling distribution}
-\label{low-order-summaries-of-the-sampling-distribution}
 
@@
 \end{equation}
-For large sample sizes, this is approximately equal to computing \(D\) by taking
-the maximum likelihood estimate for each allele frequency \(f_i = n_i / n\), but
-the maximum likelihood-based estimate will be noticeably biased for small to
-moderate sample sizes. Other low-order two-locus statistics can be computed
-using the same approach, as implemented in \emph{moments} following
+For large sample sizes, this is approximately equal to computing \(D\) by
+taking the maximum likelihood estimate for each allele frequency \(f_i = n_i /
+n\), but the maximum likelihood-based estimate will be noticeably biased for
+small to moderate sample sizes. Other low-order two-locus statistics can be
+computed using the same approach, as implemented in \texttt{moments} following
 \citet{Ragsdale2020-nz}, which can be compared across sample sizes and between
-estimates from phased or unphased data. In this paper, I focus on \(\sigma_d^2 = \E[D^2] / \E[p(1-p)q(1-q)]\) and \(\sigma_d^1 = \E[D] / \E[p(1-p)q(1-q)]\),
+estimates from phased or unphased data. In this paper, I focus on \(\sigma_d^2
+= \E[D^2] / \E[p(1-p)q(1-q)]\) and \(\sigma_d^1 = \E[D] / \E[p(1-p)q(1-q)]\),
 averaged over pairs of variants at all frequencies. However, allele-frequency
 conditioned statistics (such as keeping only loci below some frequency
-threshold as in \citet{Good2020-wj}) can be considered using this same approach.
+threshold as in \citet{Good2022-ot}) can be considered using this same
+approach.
 
 \subsubsection{Simulations of non-steady-state demography}
-\label{simulations-of-non-steady-state-demography}
 
@@
 two inferred from human populations in African and Europe. For each size
-history scenario, I tracked the evolution of \(\Psi_n(t)\) for varying selection
-models, and Figure \ref{fig:fig3} plots the trajectory of \(\sigma_d^1\) over
-time. The selection strength at both loci was fixed at \(\gamma=-2\) for all
-scenarios, and I considered additive selection without epistasis for fully
-linked (\(\rho=0\)) and more loosely linked (\(\rho=5\)) loci. With \(\rho=1\), I
-also considered positive and negative epistasis (\(\epsilon=0.5\) and \(-0.5\)),
-recessive loci (\(h=0.1\)), and a gene-based dominance model with \(h=0.1\).
+history scenario, I tracked the evolution of \(\Psi_n(t)\) for varying
+selection models, plotting the trajectories of \(\sigma_d^1\) and
+\(\sigma_d^2\) over time (Figures~\ref{fig:toy}, \ref{fig:relate_dom}, and
+\ref{fig:toy_sd2}--\ref{fig:relate_dom_sd2}). The selection strength at both
+loci was fixed at either \(\gamma=-1, -2, and -10\), and recombination was set
+to zero.
 
 The simple size change models both had ancestral \(N_e=10,000\), with one a
-\(3\)-fold population expansion that occurs 3,000 generations ago, and the other
-a \(10\)-fold reduction 2,000 generations ago followed by an expansion to double
-the ancestral size \(1,000\) generations ago. The size histories for YRI and CEU
-were inferred using Relate \citep{Speidel2019-nj} applied to the phased autosomal
-data from \citet{1000_Genomes_Project_Consortium2015-zq}, using default parameters as
-recommended in the Relate tutorial. Relate returns estimates of coalescence
-rates within specified time bins, and population sizes are estimated as their
-inverses. Estimates using Relate for population sizes in the very recent past
-(\(<3,000\) years, or \(\approx 100\) generations) diverged, so I truncated the
-history over this time period and assumed a constant size from the most recent
-non-diverged bin.
+\(3\)-fold population expansion that occurs 3,000 generations ago, and the
+other a \(5\)-fold reduction 2,000 generations ago followed by a recovery to
+its initial size \(1,000\) generations ago. The size histories for YRI and CEU
+were inferred using \texttt{Relate} \citep{Speidel2019-nj} applied to the
+phased autosomal data from \citet{1000_Genomes_Project_Consortium2015-zq},
+using default parameters as recommended in the \texttt{Relate} tutorial.
+\texttt{Relate} returns estimates of coalescence rates within specified time
+bins, and population sizes are estimated as their inverses. Estimates using
+\texttt{Relate} for population sizes in the very recent past (\(<3,000\) years,
+or \(\approx 100\) generations) diverged, so I truncated the history over this
+time period and assumed a constant size from the most recent non-diverged bin.
 
@@
 Using the annotated variant call format (VCF) files from the phase 3
-\citet{1000_Genomes_Project_Consortium2015-zq} (Thousand Genomes) data release, I
-subset the genotype VCFs of the same release to autosomal variants that were
+\citet{1000_Genomes_Project_Consortium2015-zq} (Thousand Genomes) data release,
+I subset the genotype VCFs of the same release to autosomal variants that were
 annotated as either synonymous or nonsynonymous (including both missense and
@@
 phasing error, so I computed LD statistics using unphased genotypes following
-\citet{Ragsdale2020-nz}. This approach provides unbiased estimates for pairwise LD,
-under the assumption that individuals are not inbred. I considered pairs of
+\citet{Ragsdale2020-nz}. This approach provides unbiased estimates for pairwise
+LD, under the assumption that individuals are not inbred. I considered pairs of
 mutations within the same mutation class (synonymous, missense, and
-loss-of-function) either within the same gene, in different genes within
-distance bins up to 500kb, and inside or outside of annotated domains within
-the same protein-coding genes. I used a dataset of annotated protein domains
-mapped to the hg19 human reference build compiled by \citet{Stanek2020-pa} to
-determine if a given mutation falls within a domain or not.
+loss-of-function) either within the same gene and inside or outside of
+annotated domains within the same protein-coding genes. I used a dataset of
+annotated protein domains mapped to the hg19 human reference build compiled by
+\citet{Stanek2020-pa} to determine if a given mutation falls within a domain or
+not.
 
@@
 
-All data and software used in this paper are publicly available and open source.
-I downloaded the Thousand Genomes annotations and genotypes VCFs from the ftp
-server at \url{ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/}, and
-the Thousand Genomes human ancestor fasta file from
+All data and software used in this paper are publicly available and open
+source. I downloaded the Thousand Genomes annotations and genotypes VCFs from
+the ftp server at
+\url{ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/}, and the
+Thousand Genomes human ancestor fasta file from
 \url{ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase1/analysis_results/supporting/ancestral_alignments/}.
@@
 linkage disequilibrium statistics are implemented in Python using Numpy
-\citep{Harris2020-pc} and sparse matrix solvers in Scipy \citep{Virtanen2020-kr}. These
-methods are packaged within \emph{moments}, and analyses here were performed using
-\emph{moments} version 1.1.0, available from
-\url{https://bitbucket.org/simongravel/moments/src/master/} with extensive
-documentation at \url{https://moments.readthedocs.org}. Scripts to run all
-analyses, recreate figures, and compile this manuscript are available at
-\url{https://github.com/apragsdale/two_locus_selection}.
+\citep{Harris2020-pc} and sparse matrix solvers in Scipy
+\citep{Virtanen2020-kr}. These methods are packaged within \texttt{moments},
+and analyses here were performed using \texttt{moments} version 1.1.10,
+available from \url{https://bitbucket.org/simongravel/moments/src/master/} with
+extensive documentation at \url{https://moments.readthedocs.org}. Scripts to
+run all analyses, recreate figures, and compile this manuscript are available
+at \url{https://github.com/apragsdale/two_locus_selection}.
 

### 2022-03-25 manuscript/supplement.tex
getting closer..

@@
 \usepackage{calc} % for calculating minipage widths
+\usepackage[numbib]{tocbibind} % makes the references have a section number
+\renewcommand{\bibname}{Supporting references}
 
@@
 \renewcommand{\theequation}{S\arabic{equation}}
+\renewcommand{\thesection}{S\arabic{section}}
 \setcounter{figure}{0}
@@
 \setcounter{equation}{0}
+\setcounter{section}{0}
 
-\small
 \tableofcontents
-\normalsize
 \newpage
@@
 
-\subsection{Drift, mutation, recombination, and moment closure}\label{drift-mutation-recombination-and-moment-closure}
+\subsection{Drift, mutation, recombination, and moment closure}
 
-\subsubsection{Drift}\label{drift}
+\subsubsection{Drift}
 
@@
 
-\subsubsection{Recombination}\label{recombination}
+\subsubsection{Recombination}
 
@@
 
-\subsubsection{Mutation}\label{mutation}
+\subsubsection{Mutation}
 
@@
 
-\subsection{Selection}\label{selection}
+\subsection{Selection}
 
@@
 
-\subsection{Validation against simulation}\label{validation-against-simulation}
+\subsection{Validation against simulation}
 
-\subsection{Background selection and associative overdominance}\label{background-selection-and-associative-overdominance}
+To assess the accuracy of the numerical approximation, I carried out discrete
+two-locus simulations over a range of recombination rates and selection
+parameterizations. Such discrete simulations track allele and haplotype
+frequencies in a population of a given size, with offspring generations
+constructed via multinomial sampling from the haplotype frequencies in the
+parental generation. Because we only track the two focal loci, these
+simulations do not contain any additional effects of linked selection (see
+Section~\ref{sec:bgs} for simulations with many linked loci).
 
-\subsection{Frequency dependence of signed LD}\label{frequency-dependence-of-signed-ld}
+Simulations had diploid population sizes of 5000 and I drew haploid sample
+sizes of 50 (25 diploids), with recombination rates varying from \(0\) to
+\(0.001\) (or \(\rho=0-10\)). I tested selection coefficients of \(s\) of 0,
+\(-2\times10^{-5}\), \(-2\times10^{-4}\), and \(-2\times10^{-3}\) (or
+\(\gamma=0\), \(-0.2\), \(-2\), and \(-10\)). Epistasis was set to
+\(\epsilon=0\), \(-0.5\), and \(0.5\). Conditional distributions of the
+two-locus sampling distribution show good agreement between numerical solutions
+and simulations, and residuals do not show any directional bias
+(Figures~\ref{fig:validation}--\ref{fig:last_validation}).
 
-\section{Data analysis}\label{data-analysis}
+To assess the accuracy under dominance, I simulated under the
+\citet{Roze2021-cf} scenarios (\(N=1000\), \(r=0.0001\) and \(0.001\), and
+\(sh=-0.0001\), \(-0.0004\), \(-0.001\), and \(-0.01\)). Note that \(sh=-0.01\)
+is extremely strong selection, especially for small \(h\), and \texttt{moments}
+was unstable in that parameter regime. For weak and moderate selection, there
+good agreement between simulated data and \texttt{moments} solutions
+(Figure~\ref{fig:dominance}D, E).
 
-\subsection{DFE for missense and LOF variants}\label{dfe-for-missense-and-lof-variants}
+\subsection{Background selection and associative overdominance}\label{sec:bgs}
+
+To explore the effects of many linked selected loci on patterns of LD, I used
+\texttt{fwdpy11} \citep{Thornton2014-pn,Thornton2019-qc} to simulate large
+genomic segments with many selected mutations. Each simulation had population
+size of \(N=1000\), a sequence length of 1Mb, with mutation and recombination
+rates both \(u=r=10^{-8}\). I considered four scenarios for selected mutaions:
+additive selection \(h=0.5\) with \(s=-0.0005\) and \(s=-0.01\) (\(\gamma=-1\)
+and \(-20\)), and recessivity \(h=0.0\) or \(h=0.1\) with the same selection
+coefficients. These sets of parameters should give rise to strong effects of
+linked selection, as there are many cosegregating, tightly linked selected
+variants. For each setting, I examined site-frequency spectra and LD decay for
+both selected and neutral mutations across the region.
+
+For each simulation, I ran a large number of replicates (100 independent
+replicates, and within each replicate, I sampled 25 diploid individuals every
+100 generations for a total of 20,000 samples from each replicate, after
+burn-in of \(100N\) generations). Additive selection caused minor distortions
+to the SFS, and reduced observed \(\sigma_d^1\) and \(\sigma_d^2\) in the case
+of weak selection (\(\gamma=-1\)) (Figure~\ref{fig:bgs1}). LD was largely
+unaffected in the stronger selection scenario, consistent with intuition that
+interference (including interference from additional linked sites) is most
+noticeable for \(s\approx1/N\) (Figure~\ref{fig:bgs2}).
+
+Larger deviations from two-locus expectations were observed for recessive
+mutations. For recessive weakly deleterious mutations (Figure~\ref{fig:bgs3}),
+there is a large excess of neutral and selected mutations at common frequencies
+(due to associative overdominance \citep{Zhao2016-bb}). \(\sigma_d^2\) for
+neutral and selected mutations is also increased over two-locus expectations.
+Stronger recessive mutations (Figure~\ref{fig:bgs4}) also show distorted allele
+frequencies, though the distortion to LD is not as strong as observed for weak
+recessive variants.
+
+\subsection{Frequency dependence of signed LD}
+
+\begin{itemize}
+    \item Why did we do it
+    \item How did we do it.. quite simple
+\end{itemize}
+
+\section{Data analysis}
+
+\subsection{DFE for missense and LOF variants}
 
@@
 I used the mutation model from \citet{Karczewski2020-le} to estimate the total
-mutation rate across autosomal genes (\(uL\), where \(u\) is the per-base mutation
-rate of a given mutation class, and \(L\) is the total length of the coding
-genome). These values were \((0.1442, 0.3426, 0.0256)\) for synonymous, missense,
-and LOF mutations, respectively. Roughly two thirds of new mutations in coding
-regions are expected to be missense mutations, while only 5\% of new mutations
-are LOF. I fit a demographic model to the synonymous variants, which included a
-population expansion in the deeper past and exponential growth in the recent
-past (Figure \ref{fig:msldemogdfe}A). Using the inferred optimal scaled mutation
-rate, \(\theta=4N_euL\), I estimated \(Ne\approx12,300\), and assuming an average
-generation time of 29 years I converted the inferred genetic units to physical
-units. The best-fit model had a roughly two-fold expansion 400 thousand years
-ago, and then exponential growth over the past 20-30 thousand years, with a
-current effective size of \(\approx 63,000\).
+mutation rate across autosomal genes (\(uL\), where \(u\) is the per-base
+mutation rate of a given mutation class, and \(L\) is the total length of the
+coding genome). These values were \((0.1442, 0.3426, 0.0256)\) for synonymous,
+missense, and LOF mutations, respectively. Roughly two thirds of new mutations
+in coding regions are expected to be missense mutations, while only 5\% of new
+mutations are LOF. I fit a demographic model to the synonymous variants, which
+included a population expansion in the deeper past and exponential growth in
+the recent past (Figure \ref{fig:msldemogdfe}A). Using the inferred optimal
+scaled mutation rate, \(\theta=4N_euL\), I estimated \(Ne\approx12,300\), and
+assuming an average generation time of 29 years I converted the inferred
+genetic units to physical units. The best-fit model had a roughly two-fold
+expansion 400 thousand years ago, and then exponential growth over the past
+20-30 thousand years, with a current effective size of \(\approx 63,000\).
 
@@
 each fit, I fixed the scaled mutation rate for each mutation class, so that
-\(\theta_{mis} = \frac{u_{mis}}{u_{syn}} \hat\theta_{syn}\) and \(\theta_{lof} = \frac{u_{lof}}{u_{syn}} \hat\theta_{syn}\), where values of \(u\) were found using
-the GNOMAD mutation model \citep{Karczewski2020-le}. I tested three values for the
-dominance coefficient \(h\): \(0\), \(0.2\) and \(0.5\). For missense mutations, \(h=0\)
-gave a poor fit to the data, and \(h=0.5\) fit best among the three tested
-dominance coefficients. For LOF variants, \(h=0\) also fit poorly, but \(h=0.2\)
-and \(h=0.5\) gave similar likelihoods, highlighting that inferring dominance
-using the SFS is poorly constrained. Regardless of the dominance coefficient
-assumed, however, the vast majority of LOF variants were inferred to be
-strongly deleterious, with only \(\sim10\%\) of new mutations having selection
-coefficients on the order \(1/N_e\) or less.
+\(\theta_{mis} = \frac{u_{mis}}{u_{syn}} \hat\theta_{syn}\) and \(\theta_{lof}
+= \frac{u_{lof}}{u_{syn}} \hat\theta_{syn}\), where values of \(u\) were found
+using the GNOMAD mutation model \citep{Karczewski2020-le}. I tested three
+values for the dominance coefficient \(h\): \(0\), \(0.2\) and \(0.5\). For
+missense mutations, \(h=0\) gave a poor fit to the data, and \(h=0.5\) fit best
+among the three tested dominance coefficients. For LOF variants, \(h=0\) also
+fit poorly, but \(h=0.2\) and \(h=0.5\) gave similar likelihoods, highlighting
+that inferring dominance using the SFS is poorly constrained. Regardless of the
+dominance coefficient assumed, however, the vast majority of LOF variants were
+inferred to be strongly deleterious, with only \(\sim10\%\) of new mutations
+having selection coefficients on the order \(1/N_e\) or less.
 
-\subsection{Multinucleotide mutations and positive LD between linked synonymous variants}\label{multinucleotide-mutations-and-positive-ld-between-linked-synonymous-variants}
+\subsection{Multinucleotide mutations and positive LD between linked synonymous variants}
 
@@
 locus, some fraction of new mutations could instead occur spontaneously and
-create a new pair of mutations with initial counts \(n_{AB} = 1\) and \(n_{ab} = n-1\).
+create a new pair of mutations with initial counts \(n_{AB}=1\) and
+\(n_{ab}=n-1\). We can partition $\Psi_n$ into pairs of mutations due to MNM
+events (\(\Psi_n^{MNM}\)) and those due to sequential mutation events
+(\(\Psi_n^{ISM}\)), so that \(\Psi_n=\Psi_n^{MNM}+\Psi_n^{ISM}\). For given
+scaled mutation rates \(\theta=4N_e\mu\), \(\Psi_n^{ISM}\propto\theta^2\) and
+\(\Psi_n^{MNM}\propto p_{MNM}\theta\), where \(p_{MNM}\) is the probability
+that a mutational event causes a multinucleotide mutation. This difference in
+scaling means that \(\Psi_n\) is sensitive to both the overall mutation rate
+\(\theta\) as well as the probability that a mutation is a MNM at a given
+distance.
+
+\subsubsection{Optimization details}
 
 Here, I fit a simple exponential model for the fraction of new mutations at a
-given distance that arose through a MNM event, so that \(P(MNM | d) = Ae^{-\lambda d}\), where \(d\) is the distance separating pairs of mutations. I
-considered all synonymous mutations within genes in the MSL population and used
-the same population size history model as inferred in the DFE section above for
-a demographic control. This left two parameters to be fit, \(A\) and \(\lambda\),
-which I fit to the binned decay curve of \(\sigma_d^1\). I needed to assume an
-average per-base recombination rate \(r\) across gene regions, and tested a
-number of values between \(10^{-9}\) and \(2\times 10^{-8}\). The optimization was
-insensitive to the chosen value of \(r\), because the decay of positive LD occurs
-rapidly. For any plausible value of \(r\), \(\sigma_d^1\) decays to zero well
-before distances between pairs have scaled recombination rates \(\rho=4 N_e r d \sim 1\), and expected statistics for \(\rho \ll 1\) vary only negligibly.
+given distance that give rise to a MNM event, so that \(p_{MNM,d} = P(MNM | d)
+= Ae^{-\lambda d}\), where \(d\) is the distance separating pairs of mutations
+in base pairs. I considered all synonymous mutations within genes in the MSL
+data and used the same population size history model as inferred in the DFE
+section above for a demographic control. This left two parameters to be fit,
+\(A\) and \(\lambda\), which I fit to the binned decay curve of \(\sigma_d^1\)
+using the midpoint of each bin as the distance for that bin. I needed to assume
+an average per-base recombination rate \(r\) across gene regions, and tested a
+number of values between \(10^{-9}\) and \(2\times 10^{-8}\). Optimization was
+insensitive to the chosen value of \(r\), because the decay of positive LD
+occurs rapidly. For any plausible value of \(r\), \(\sigma_d^1\) decays to zero
+well before distances between pairs have scaled recombination rates \(\rho=4
+N_e r d \sim 1\), and expected statistics for \(\rho \ll 1\) vary negligibly.
+
+To perform optimization, I assumed a Guassian likelihood function within each
+bin using variances estimated via bootstrap replicates from sampling genes with
+replacement. A composite likelihood was computed by taking the product of
+likelihoods across bins. I used the estimate of \(Ne\approx12,300\) from the
+DFE optimization using MSL data, a coding mutation rate of
+\(\mu=1.33\times10^{-8}\) \citep{Karczewski2020-le}, and the fraction of coding
+mutations that result in synonymous mutations was assumed to be (assuming a
+ration of nonsynonymous mutations to synonymous mutations of 2.5:1). Thus, the
+scaled mutation rate was \(\theta=4\times
+N_e\times\mu\times\frac{1}{3.5}\approx1.87\times10^{-4}\).
+
+\subsubsection{MNM optimization results and discussion}
 
 In fitting the LD decay of \(\sigma_d^1\), the best fit parameters were
-\(A=0.132\), and \(\lambda=0.0103\). An exponential scaling of \(0.01\) implies that
-the vast majority of new mutations \emph{pairs} do not occur via MNMs for distances
-greater than 200 bp, though a substantial fraction (\(10-15\%\)) occur via MNMs
-for very tightly linked loci with distances on the order \(0-50\) base pairs. It
-is important to note that this does not mean that \(10-15\%\) of new mutations
-occur via MNMs, since this fraction is conditioned on two mutations occurring
-at short distances.
+\(A=2.44\times10^{-5}\), and \(\lambda=0.00919\). Thus, only a small fraction
+of new mutations were inferred to cause MNM events (\(\ll1\%\)), and
+\(1/\lambda\approx 100\) means that the proportion of mutation events that
+cause MNM decays quickly within a few hundred base pairs.
+
+However, because of the differences in scaling of \(\Psi_n^{ISM}\) and
+\(\Psi_n^{MNM}\) (with the square of \(\theta\) vs linearly as
+\(p_{MNM}\theta\)), the number of \emph{observed} pairs of mutations at a given
+distance that arose due to a MNM event is much larger. A rough estimate given
+the assumed mutation rate and inferred parameters is that \(p_{MNM}\theta /
+\theta^2 \approx 10\%\) of pairs of synonymous mutations observed at
+\(d\approx0\) are due to MNM events, roughly \(p_{MNM}e^{-\lambda 100}\theta /
+\theta^2 \approx 4\%\) at \(d=100\) base pairs, and only about
+\(p_{MNM}e^{-\lambda 400}\theta / \theta^2 \approx 0.25\%\) at \(d=400\) base
+pairs. This inference is very sensitive to the assumed mutation rate and the
+proportion of mutations that cause synonymous vs nonsynonymous variants, and
+future work will be needed to refine these estimates.
 
 \subsection{Grouping Thousand Genomes populations based on clustering}
-\label{grouping-thousand-genomes-populations-based-on-clustering}
 
@@
 
-\addcontentsline{toc}{section}{References}
 \bibliographystyle{genetics}
@@
 
-\addcontentsline{toc}{section}{Supporting tables}
-\section*{Supporting tables}
+\section{Supporting tables}
 
@@
 
-\addcontentsline{toc}{section}{Supporting Figures}
-\section*{Supporting figures}
+\section{Supporting figures}
+
+\textbf{all supporting figures need captions}
 
@@
     \caption{
-        \textbf{Accuracy of the jackknife approximation and runtime} Small
-        sample sizes can lead to large error in the moment-closure
+        \textbf{Accuracy of the jackknife approximation and runtime.}
+        Small sample sizes can lead to large error in the moment-closure
         approximation for larger recombination distances or selection
@@
     \label{fig:jackknife}
+\end{figure}
+
+\begin{figure}[ht!]
+    \centering
+    %\includegraphics{}
+    \caption{\textbf{Validation figures.}}
+    \label{fig:validation}
+\end{figure}
+
+\begin{figure}[ht!]
+    \centering
+    %\includegraphics{}
+    \caption{\textbf{Validation figures.}}
+    \label{fig:last_validation}
 \end{figure}
@@
     \centering
+    \includegraphics[width=\textwidth]{../figures/data_within_between}
+    \caption{
+        \textbf{Within vs between.}
+    }
+    \label{fig:between_domains}
+\end{figure}
+
+\begin{figure}[ht!]
+    \centering
     \includegraphics{../figures/mutations_in_genes}
@@
     \centering
-    \includegraphics{../figures/msl_mnms}
+    \includegraphics{../figures/msl_mnms.v2}
     \caption{
@@
         at a given distance \(d\), as \(Ae^{-\lambda d}\). Across all recombination rates
-        tested, the best fit parameters were \(A=0.13\) and \(\lambda=0.010\).
+        tested, the best fit parameters were \(A=0.11\) and \(\lambda=0.0085\).
     }

### 2022-03-25 analysis/fit_MSL_clustered_mutations.v2.py

@@ -13,8 +13,12 @@ bin_mids = np.mean(bin_edges, axis=1)[:-2]
 
 sd1 = np.array(data["data"]["bins"]["all_sites"]["synonymous"]["sd1"])[:-2]
 sd2 = np.array(data["data"]["bins"]["all_sites"]["synonymous"]["sd2"])[:-2]
-v1 = np.array([data["SEs"]["bins"]["all_sites"]["synonymous"][b]["sd1"] for b in bin_edges])[:-2]
-v2 = np.array([data["SEs"]["bins"]["all_sites"]["synonymous"][b]["sd2"] for b in bin_edges])[:-2]
+v1 = np.array(
+    [data["SEs"]["bins"]["all_sites"]["synonymous"][b]["sd1"] for b in bin_edges]
+)[:-2]
+v2 = np.array(
+    [data["SEs"]["bins"]["all_sites"]["synonymous"][b]["sd2"] for b in bin_edges]
+)[:-2]
 
 demo_params = [2.194, 5.145, 0.579, 0.0400]
 Ne = 12314.8
@@ -81,12 +85,12 @@ def ll_normal(data, model, variances):
         - 1 / 2 * np.log(2 * np.pi * len(sd1))
         - 1 / 2 * np.log(v1)
     )
-    #ll_sd2 = np.sum(
+    # ll_sd2 = np.sum(
     #    -1 / 2 * (sd2 - Esd2) ** 2 / v2
     #    - 1 / 2 * np.log(2 * np.pi * len(sd2))
     #    - 1 / 2 * np.log(v2)
-    #)
-    return ll_sd1 #+ ll_sd2
+    # )
+    return ll_sd1  # + ll_sd2
 
 
 def object_func_log(log_params, *args, **kwargs):
@@ -115,7 +119,13 @@ _counter = 0
 
 
 def object_func(
-    params, model_func, data, variances, verbose=0, lower_bound=None, upper_bound=None,
+    params,
+    model_func,
+    data,
+    variances,
+    verbose=0,
+    lower_bound=None,
+    upper_bound=None,
 ):
     global _counter
     _counter += 1
@@ -165,17 +175,19 @@ for r in rs:
         Esd1 = []
         Esd2 = []
         for w, b in zip(weights, bin_mids):
+            w_ism = theta ** 2
+            w_mnm = w * theta
             Esd1.append(
[…]

### 2022-03-25 figures/data_coarse.py

@@ -54,7 +54,7 @@ for continent, pops in populations.items():
         SE_dom[pop] = se_pop
 
 
-ms = 3
+ms = 2
 lw = 0.5
 markers = ["o", "s"]
 colors = Colorblind[8]
@@ -152,7 +152,7 @@ def plot_domain_data(pops, categ, ax, legend=False):
     ax.set_xticks(np.mean(xx.reshape(len(pops), 4), axis=1))
     ax.set_xticklabels(pops)
     if legend:
-        ax.legend(fontsize=5, frameon=False)
+        ax.legend(fontsize=5, frameon=False, loc="upper left")
 
 
 fig = plt.figure(5, figsize=(6.5, 4.5))
@@ -309,117 +309,208 @@ plot_domain_data(pops, "all", ax2, legend=True)
 ax2.set_ylabel("$\sigma_d^1$")
 ax2.set_xlabel("Populations")
 ax2.set_title("All pairs")
+ax2.set_ylim(-0.2, 1)
+
+#ax3 = plt.subplot2grid(dims, (5, 1), rowspan=3)
+#plot_domain_data(pops, "leq2", ax3)
+#ax3.set_xlabel("Populations")
+#ax3.set_title("$n_A, n_B \leq 2$")
+#
+#ax4 = plt.subplot2grid(dims, (5, 2), rowspan=3)
+#plot_domain_data(pops, "3to8", ax4)
+#ax4.set_xlabel("Populations")
+#ax4.set_title("$3 \leq n_A, n_B \leq 8$")
 
 ax3 = plt.subplot2grid(dims, (5, 1), rowspan=3)
-plot_domain_data(pops, "leq2", ax3)
+plot_domain_data(pops, "geq9", ax3)
 ax3.set_xlabel("Populations")
-ax3.set_title("$n_A, n_B \leq 2$")
+ax3.set_title("$n_A, n_B \geq 9$")
+ax3.sharey(ax2)
 
+## plots of LD decay within and outside of domains for AFR populations
 ax4 = plt.subplot2grid(dims, (5, 2), rowspan=3)
-plot_domain_data(pops, "3to8", ax4)
-ax4.set_xlabel("Populations")
-ax4.set_title("$3 \leq n_A, n_B \leq 8$")
+
+decay_data = {}
+for c, ps in populations.items():
+    for p in ps:
+        decay_data[p] = pickle.load(
[…]

### 2022-03-25 figures/data_domains.py

@@ -160,13 +160,13 @@ plot_data(pop_eas, categ, ax12)
 plt.tight_layout()
 plt.savefig("data_domains.pdf")
 
-fig2 = plt.figure(98765, figsize=(6.5, 6))
+fig2 = plt.figure(98765, figsize=(6.5, 7))
 fig2.clf()
 i = 0
 for pop in populations["Africa"]:
-    if pop != pop_afr:
+    #if pop != pop_afr:
         for categ in ["all", "leq2", "3to8", "geq9"]:
-            ax = plt.subplot(4, 4, i + 1)
+            ax = plt.subplot(5, 4, i + 1)
             if i == 0:
                 ax.set_title("All pairs")
             elif i == 1:
@@ -189,11 +189,11 @@ for pop in populations["Africa"]:
 plt.tight_layout()
 plt.savefig("data_domains_afr.pdf")
 
-fig3 = plt.figure(98764, figsize=(6.5, 6))
+fig3 = plt.figure(98764, figsize=(6.5, 7))
 fig3.clf()
 i = 0
 for pop in populations["Europe"]:
-    if pop != pop_eur:
+    #if pop != pop_eur:
         for categ in ["all", "leq2", "3to8", "geq9"]:
             ax = plt.subplot(4, 4, i + 1)
             if i == 0:
@@ -217,11 +217,11 @@ for pop in populations["Europe"]:
 plt.tight_layout()
 plt.savefig("data_domains_eur.pdf")
 
-fig4 = plt.figure(98763, figsize=(6.5, 6))
+fig4 = plt.figure(98763, figsize=(6.5, 7))
 fig4.clf()
 i = 0
 for pop in populations["East Asia"]:
-    if pop != pop_eas:
+    #if pop != pop_eas:
         for categ in ["all", "leq2", "3to8", "geq9"]:
             ax = plt.subplot(4, 4, i + 1)
             if i == 0:

### 2022-03-25 figures/demog_YRI_CEU.dominance.py

@@ -437,4 +437,10 @@ for i in range(1, 6):
     axes[i].set_xticks([1e6, 1e5, 1e4, 1e3])
 
 fig.tight_layout()
+fig.text(0.02, 0.95, "A", fontsize=8, va="center", ha="center")
+fig.text(0.37, 0.95, "B", fontsize=8, va="center", ha="center")
+fig.text(0.70, 0.95, "C", fontsize=8, va="center", ha="center")
+fig.text(0.02, 0.49, "D", fontsize=8, va="center", ha="center")
+fig.text(0.37, 0.49, "E", fontsize=8, va="center", ha="center")
+fig.text(0.70, 0.49, "F", fontsize=8, va="center", ha="center")
 plt.savefig("demog_YRI_CEU.dominance.pdf")

### 2022-03-25 figures/demog_YRI_CEU.dominance.sd2.py

@@ -437,4 +437,10 @@ for i in range(1, 6):
     axes[i].set_xticks([1e6, 1e5, 1e4, 1e3])
 
 fig.tight_layout()
+fig.text(0.02, 0.95, "A", fontsize=8, va="center", ha="center")
+fig.text(0.37, 0.95, "B", fontsize=8, va="center", ha="center")
+fig.text(0.70, 0.95, "C", fontsize=8, va="center", ha="center")
+fig.text(0.02, 0.49, "D", fontsize=8, va="center", ha="center")
+fig.text(0.37, 0.49, "E", fontsize=8, va="center", ha="center")
+fig.text(0.70, 0.49, "F", fontsize=8, va="center", ha="center")
 plt.savefig("demog_YRI_CEU.dominance.sd2.pdf")

### 2022-03-25 figures/demog_YRI_CEU.py

@@ -440,4 +440,10 @@ for i in range(1, 6):
     axes[i].set_xticks([1e6, 1e5, 1e4, 1e3])
 
 fig.tight_layout()
+fig.text(0.02, 0.95, "A", fontsize=8, va="center", ha="center")
+fig.text(0.37, 0.95, "B", fontsize=8, va="center", ha="center")
+fig.text(0.70, 0.95, "C", fontsize=8, va="center", ha="center")
+fig.text(0.02, 0.49, "D", fontsize=8, va="center", ha="center")
+fig.text(0.37, 0.49, "E", fontsize=8, va="center", ha="center")
+fig.text(0.70, 0.49, "F", fontsize=8, va="center", ha="center")
 plt.savefig("demog_YRI_CEU.pdf")

### 2022-03-25 figures/demog_YRI_CEU.sd2.py

@@ -440,4 +440,10 @@ for i in range(1, 6):
     axes[i].set_xticks([1e6, 1e5, 1e4, 1e3])
 
 fig.tight_layout()
+fig.text(0.02, 0.95, "A", fontsize=8, va="center", ha="center")
+fig.text(0.37, 0.95, "B", fontsize=8, va="center", ha="center")
+fig.text(0.70, 0.95, "C", fontsize=8, va="center", ha="center")
+fig.text(0.02, 0.49, "D", fontsize=8, va="center", ha="center")
+fig.text(0.37, 0.49, "E", fontsize=8, va="center", ha="center")
+fig.text(0.70, 0.49, "F", fontsize=8, va="center", ha="center")
 plt.savefig("demog_YRI_CEU.sd2.pdf")

### 2022-03-25 figures/demog_bottle_expansion.dominance.py

@@ -433,4 +433,10 @@ for i in range(1, 6):
     axes[i].set_xticklabels(axes[0].get_xticklabels())
 
 fig.tight_layout()
+fig.text(0.03, 0.95, "A", fontsize=8, va="center", ha="center")
+fig.text(0.37, 0.95, "B", fontsize=8, va="center", ha="center")
+fig.text(0.70, 0.95, "C", fontsize=8, va="center", ha="center")
+fig.text(0.03, 0.49, "D", fontsize=8, va="center", ha="center")
+fig.text(0.37, 0.49, "E", fontsize=8, va="center", ha="center")
+fig.text(0.70, 0.49, "F", fontsize=8, va="center", ha="center")
 plt.savefig("demog_bottle_expand.dominance.pdf")

### 2022-03-25 figures/demog_bottle_expansion.dominance.sd2.py

@@ -436,4 +436,10 @@ for i in range(1, 6):
     axes[i].set_xticklabels(axes[0].get_xticklabels())
 
 fig.tight_layout()
+fig.text(0.03, 0.95, "A", fontsize=8, va="center", ha="center")
+fig.text(0.37, 0.95, "B", fontsize=8, va="center", ha="center")
+fig.text(0.70, 0.95, "C", fontsize=8, va="center", ha="center")
+fig.text(0.03, 0.49, "D", fontsize=8, va="center", ha="center")
+fig.text(0.37, 0.49, "E", fontsize=8, va="center", ha="center")
+fig.text(0.70, 0.49, "F", fontsize=8, va="center", ha="center")
 plt.savefig("demog_bottle_expand.dominance.sd2.pdf")

### 2022-03-25 figures/demog_bottle_expansion.py

@@ -453,4 +453,11 @@ for i in range(1, 6):
     axes[i].set_xticklabels(axes[0].get_xticklabels())
 
 fig.tight_layout()
+
+fig.text(0.03, 0.95, "A", fontsize=8, va="center", ha="center")
+fig.text(0.37, 0.95, "B", fontsize=8, va="center", ha="center")
+fig.text(0.70, 0.95, "C", fontsize=8, va="center", ha="center")
+fig.text(0.03, 0.49, "D", fontsize=8, va="center", ha="center")
+fig.text(0.37, 0.49, "E", fontsize=8, va="center", ha="center")
+fig.text(0.70, 0.49, "F", fontsize=8, va="center", ha="center")
 plt.savefig("demog_bottle_expand.pdf")

### 2022-03-25 figures/demog_bottle_expansion.sd2.py

@@ -457,4 +457,10 @@ for i in range(1, 6):
     axes[i].set_xticklabels(axes[0].get_xticklabels())
 
 fig.tight_layout()
+fig.text(0.03, 0.95, "A", fontsize=8, va="center", ha="center")
+fig.text(0.37, 0.95, "B", fontsize=8, va="center", ha="center")
+fig.text(0.70, 0.95, "C", fontsize=8, va="center", ha="center")
+fig.text(0.03, 0.49, "D", fontsize=8, va="center", ha="center")
+fig.text(0.37, 0.49, "E", fontsize=8, va="center", ha="center")
+fig.text(0.70, 0.49, "F", fontsize=8, va="center", ha="center")
 plt.savefig("demog_bottle_expand.sd2.pdf")

### 2022-03-25 figures/dominance_and_roze.py

@@ -7,7 +7,7 @@
 import moments.TwoLocus
 import matplotlib.pylab as plt
 import numpy as np
-import pickle
+import pickle, gzip
 
 # set font sizes
 import matplotlib
@@ -75,13 +75,15 @@ colors = Colorblind[8]
 fig = plt.figure(19382, figsize=(6.5, 4.5))
 fig.clf()
 
-grid = (5, 2)
+grid = (5, 6)
 # sigma_d^1 for gamma = -1 and -5
-ax1 = plt.subplot2grid(grid, (0, 0), rowspan=2)
-ax2 = plt.subplot2grid(grid, (0, 1), rowspan=2)
+ax1 = plt.subplot2grid(grid, (0, 0), rowspan=2, colspan=2)
+ax2 = plt.subplot2grid(grid, (0, 2), rowspan=2, colspan=2)
 # roze comparisons
-ax3 = plt.subplot2grid(grid, (2, 0), rowspan=3)
-ax4 = plt.subplot2grid(grid, (2, 1), rowspan=3)
+ax3 = plt.subplot2grid(grid, (2, 0), rowspan=3, colspan=3)
+ax4 = plt.subplot2grid(grid, (2, 3), rowspan=3, colspan=3)
+# varying gamma (upper right)
+ax5 = plt.subplot2grid(grid, (0, 4), rowspan=2, colspan=2)
 
 ax1.plot(rhos, 0 * ok, "k--", lw=1, label=None)
 ax2.plot(rhos, 0 * ok, "k--", lw=1, label=None)
@@ -104,6 +106,21 @@ for ii, h in enumerate(hs):
         label=rf"$h={h}$",
     )
 
+
+data_vary = pickle.load(gzip.open("data/dominance_vary_gamma.bp.gz", "rb"))
+gammas = data_vary["gammas"]
+hs = sorted(data_vary["sd1"].keys())
+
+ax5.plot(-gammas, 0*gammas, "k--", lw=1, label=None)
+for h in hs:
+    ax5.plot(-gammas, data_vary["sd1"][h], lw=1, label=rf"$h={h}$")
+
+ax5.set_title(r"$\rho=0$")
+ax5.set_xlabel(r"$\gamma=2Ns$")
+ax5.set_xscale("log")
+ax5.legend(fontsize=6, loc="lower left")
+ax5.set_xticks([0.1, 1, 10])
[…]

### 2022-03-27 manuscript/supplement.tex
finish up supplement

@@
 sizes of 50 (25 diploids), with recombination rates varying from \(0\) to
-\(0.001\) (or \(\rho=0-10\)). I tested selection coefficients of \(s\) of 0,
-\(-2\times10^{-5}\), \(-2\times10^{-4}\), and \(-2\times10^{-3}\) (or
-\(\gamma=0\), \(-0.2\), \(-2\), and \(-10\)). Epistasis was set to
-\(\epsilon=0\), \(-0.5\), and \(0.5\). Conditional distributions of the
+\(0.001\) (or \(\rho=0-20\)). I tested selection coefficients of \(s\) of 0,
+\(-1\times10^{-5}\), \(-2\times10^{-4}\), and \(-2\times10^{-3}\) (or
+\(\gamma=0\), \(-0.1\), \(-2\), and \(-20\)). Epistasis was set to
+\(\epsilon=0\), \(-0.5\), and \(0.5\), and dominance was set to either
+\(h=0.5\) (additive) or \(0.1\) (recessive). Conditional distributions of the
 two-locus sampling distribution show good agreement between numerical solutions
 and simulations, and residuals do not show any directional bias
-(Figures~\ref{fig:validation}--\ref{fig:last_validation}).
+(Figures~\ref{fig:validation}--\ref{fig:last_validation}). The Support Figures
+includes a handful of these comparisons, and the full set of comparisons (over
+40 in total across the range of parameters listed above), simulation, and
+plotting scripts in Python are available at
+\url{https://github.com/apragsdale/two_locus_selection}.
 
@@
 
-\begin{itemize}
-    \item Why did we do it
-    \item How did we do it.. quite simple
-\end{itemize}
+Mutations at different frequencies are expected to have different signs and
+magnitudes of LD. For example, rare mutations show highly positive LD, even for
+pairs of neutral mutations \citep{Good2022-ot}. Mutations of different
+functional classes (synonymous, missense, and loss-of-function, for example)
+are expected to segregate at different average frequencies, and average
+selection pressures are observed to be greater within annotated conserved
+domains, with stronger purifying selection driving allele frqeuencies to
+relatively lower levels among missense and loss-of-function variants. Since
+mutations of difference classes and locations have different average
+frequencies, this may drive differences in patterns of LD when comparing
+between them.
+
+One approach to lessen the impact of differing allele frequencies when
+comparing between classes of mutations is to condition on allele frequencies.
+From sampling data with observed derived allele counts, it is simplest to
+condition on those allele counts \citep[e.g.,\(n_A,n_B=2\),][]{Garcia2021-zn},
+and to compute \(\sigma_d^1\) and \(\sigma_d^2\) from pairs of mutations that
+satisfy that condition. In the simulations with haploid sample sizes \(n=50\)
+(Figures~\ref{fig:toy},~\ref{fig:relate_dom},~\ref{fig:toy_sd2}--\ref{fig:relate_dom_sd2}),
+I partitioned allele counts as uncommon (\(n_A, n_B \leq 4\), or \(\hat{f_A},
+\hat{f_B} \leq 0.08\)), or common (\(n_A, n_B \geq 5\)). In the
+\citet{1000_Genomes_Project_Consortium2015-zq} data, in which most populations
+have diploid sample sizes between 80 and 100, I considered \(n_A, n_B \geq 2\),
+\(3\leq n_A, n_B \leq 8\), and \(n_A, n_B \geq 9\) (or rare, uncommon, and
+common (\(\gtrsim 0.05\))). Functions to compute allele-count-conditioned
+statistics from \(\Psi_n\) are packaged within \texttt{moments}.
 
@@
 \subsection{Multinucleotide mutations and positive LD between linked synonymous variants}
+\label{sec:mnm}
 
@@
     \end{tabular}
-    \caption{General selection model for diploids and dominance models.}
+    \caption{
+        \textbf{General selection model for diploids and dominance models.}
+        Different dominance and interactive effects can be implemented by
+        assigning the appropriate values of $s$.
+    }
     \label{tab:selmodels}
@@
     \end{tabular}
-    \caption{Haploid epistasis model.}
+    \caption{\textbf{Haploid epistasis model.}}
     \label{tab:epistasis}
@@
     \end{tabular}
-    \caption{Thousand Genomes Project population descriptions for populations used in this study.}
+    \caption{
+        \textbf{
+            Thousand Genomes Project population descriptions
+            for populations used in this study.
+        }
+    }
     \label{tab:1kgpops}
@@
 \begin{longtable}[t]{lllr}
-\caption{\label{tab:tajimasD}Tamija's $D$ for classes of coding mutations, both within annotated domains and outside of domains.}\\
+\caption{\label{tab:tajimasD}
+\textbf{Tamija's $D$ for classes of coding mutations.} Here, we partitioned by
+synonymous, missense, and nonsense mutationsi, and considered all mutations gene-wide,
+mutations falling within annotated conserved elements,
+and mutations falling outside of those
+elements.}\\
 \toprule
@@
 \endfirsthead
-\caption[]{\label{tab:tajimasD}Tamija's $D$ for classes of coding mutations, both within annotated domains and outside of domains. \textit{(continued)}}\\
+\caption[]{\label{tab:tajimasD}Tamija's $D$ for classes of coding mutations.
+\textit{(continued)}}\\
 \toprule
@@
     \caption{
+        \textbf{Distribution of fitness effects inferred for nonsynonymous mutations.}
         DFEs inferred for missense and loss-of-function variants in MSL for varying
@@
 
-\textbf{all supporting figures need captions}
-
 \begin{figure}[ht!]
@@
         coefficients. Generally, the jackknife approximation breaks down for
-        recombination rates greater than \(\rho\approx30\). While increasing
+        recombination rates greater than \(\rho\gtrsim50\), depending on
+        the sample size. While increasing
         sample size leads to more accurate solutions, it comes at the cost of
@@
     \centering
-    %\includegraphics{}
-    \caption{\textbf{Validation figures.}}
-    \label{fig:validation}
-\end{figure}
-
-\begin{figure}[ht!]
-    \centering
-    %\includegraphics{}
-    \caption{\textbf{Validation figures.}}
-    \label{fig:last_validation}
+    \includegraphics[width=\textwidth]{../simulations/discrete/plots/comp_Ne_5000_n_50_r_0.0_s_0.0_e_0.0.pdf}
+    \caption{
+        \textbf{Comparison between numerics and simulations for fully linked neutral loci.}
+        With no recombination (\(r=0\)) and no selection, the moments solution is
+        exact, and computing \(sigma_d^2\) from \(\Psi_n\) gives 5/11, the analytic
+        solution from \citet{Ohta1969-ie}.
+        Discrepancies between numerical expectations and the simulations are
+        due to simulation noise, with Anscombe Poisson residuals roughly normally
+        distributed around zero \citep{Pierce1986-um}.
+        In the bar plots, ``E'' (blue) are expectations from \texttt{moments} and
+        ``D'' are simulated data.
+    }
+    \label{fig:validation1}
+\end{figure}
+
+\begin{figure}[ht!]
+    \centering
+    \includegraphics[width=\textwidth]{../simulations/discrete/plots/comp_Ne_5000_n_50_r_0.001_s_0.002_e_0.pdf}
+    \caption{
+        \textbf{Comparison to simulations for loosely linked additively selected loci.}
+        \(Ne=5000\), \(r=0.001\), and \(s=-0.002\) (\(\rho=20\) and \(\gamma=-20\)
+        at both loci), with no epistatic interactions (\(\epsilon=0\)).
+        E: \texttt{moments} expectations, D: simulated data.
+    }
+    \label{fig:validation2}
+\end{figure}
+
+\begin{figure}[ht!]
+    \centering
+    \includegraphics[width=\textwidth]{../simulations/discrete/plots/comp_Ne_5000_n_50_r_5e-05_s_0.0002_e_-0.5.pdf}
+    \caption{
+        \textbf{Comparison to simulations with antagonistic epistasis.}
+        \(Ne=5000\), \(r=0.00005\), \(s=-0.0002\) (\(\rho=1\) and \(\gamma=-2\)
+        at both loci), and \(\epsilon=-0.5\).
+        E: \texttt{moments} expectations, D: simulated data.
+    }
+    \label{fig:validation3}
+\end{figure}
+
+\begin{figure}[ht!]
+    \centering
+    \includegraphics[width=\textwidth]{../simulations/discrete/plots/comp_Ne_5000_n_50_r_5e-05_s_0.0002_e_-0.5.pdf}
+    \caption{
+        \textbf{Comparison to simulations with synergistic epistasis.}
+        \(Ne=5000\), \(r=0.00005\), \(s=-0.0002\) (\(\rho=1\) and \(\gamma=-2\)
+        at both loci), and \(\epsilon=0.5\).
+        E: \texttt{moments} expectations, D: simulated data.
+    }
+    \label{fig:validation4}
+\end{figure}
+
+\begin{figure}[ht!]
+    \centering
+    \includegraphics[width=\textwidth]{../simulations/discrete/plots/comp_Ne_5000_n_50_r_5e-06_s_0.0002_h_0.1.pdf}
+    \caption{
+        \textbf{Comparison to simulations with site-wise dominance.}
+        \(Ne=5000\), \(r=5\times10^{-6}\), and \(s=-0.0002\) (\(\rho=0.1\) and
+        \(\gamma=-2\)), with \(h=0.1\) at both loci and no
+        epistatic interactions (\(\epsilon=0\)).
+        E: \texttt{moments} expectations, D: simulated data.
+        Additional comparisons to simulated data are availabe at
+        \texttt{https://github.com/apragsdale/two\_locus\_selection}.
+    }
+    \label{fig:validation5}
 \end{figure}
@@
         \textbf{Background selection with slightly deleterious mutations.}
-        Using an individual based forward simulator
+        Using an individual based forward simulator \citep{Thornton2019-qc}, I
+        simulated many replicates of 1Mb regions with high mutation rates of
+        selected mutations (see Section~\ref{sec:bgs} for details). Here,
+        many linked additive weakly selected mutations slightly distort the
+        SFS for both selected and neutral mutations. \(\sigma_d^2\) is reduced
+        relative to two-locus expectations within background selection, and
+        Hill-Robertson interference effects are slightly diminished with
+        \(\sigma_d^1\) not as strongly negative as expected.
     }
@@
         \textbf{Background selection with moderately deleterious mutations.}
+        In the case of many linked strongly selected additive mutations, neither
+        selected nor neutral statistics deviate dramatically from two-locus expectations.
+        The SFS from the simulated data matches expectations, with a slight excess
+        of selected mutations. LD is also largely unaffected, with \(\sigma_d^2\)
+        matching expectations and \(\sigma_d^1\) fluctuating near zero as expected.
     }
@@
         \textbf{Associative overdominance due to slightly deleterious recessive mutations.}
+        The presence of many linked weakly selected recessive mutations can cause
+        measues of diversity to have large increases over single- and two-locus
+        expectations. The SFS for both neutral and selected mutations both show
+        a very large excess of common variants (note the log scale), and \(\sigma_d^2\)
+        are also increased over two-locus expectations for both classes of mutations.
+        Still, \(\sigma_d^1\) remains 0 for neutral mutations and is not as strongly
+        negative for pairs of selected mutations.
     }
@@
         \textbf{Background selection with moderately deleterious recessive mutations.}
+        More strongly deleterious recessive mutations do not show as dramatic of
+        deviations from expectations as the case of weakly deleterious recessive
+        mutations. The neutral SFS is reduced, and we observe more selected mutations
+        at higher frequencies than expected. LD is also only weakly distorted from
+        two-locus expectations. Notably, we see positive LD (matching between
+        simulations and expectations) for strongly deleterious recessive mutations,
+        as predicted in \citet{Roze2021-cf}.
     }
@@
     \caption{
-        \textbf{.}
+        \textbf{Time series of squared LD under bottleneck 
+        and expansion histories with epistasis.}
+        Bottlenecks increase \(\sigma_d^2\) (squared LD),
+        while expansions tend to decrease squared LD.
+        Dashed lines show neutral expectations.
     }
@@
     \caption{
-        \textbf{.}
+        \textbf{Time series of signed LD under bottleneck 
+        and expansion histories with dominance.}
+        Gene-based dominance, unlike site-wise dominance effects, can cause
+        large positive signed LD, similar to compensatory mutations or
+        antagonistic epistasis. In general, common variants are more
+        stable under population size changes than rare variants.
+        Dashed lines show neutral expectations.
     }
@@
     \caption{
-        \textbf{.}
+        \textbf{Time series of squared LD under bottleneck 
+        and expansion histories with dominance.}
+        Similar to the cases with epistasis, bottlenecks tend to increase
+        \(\sigma_d^2\) and expansions decrease squared LD.
+        Dashed lines show neutral expectations.
     }
@@
     \caption{
-        \textbf{.}
+        \textbf{Time series of signed LD under inferred models of human population
+        size history with with epistasis.}
+        Similar to the instantaneous size changes in the toy models
+        (Figure~\ref{fig:toy}), bottlenecks tend to make signed LD more extreme.
+        However, in the case of strong selection and antagonistic epistasis,
+        the CEU \(\sigma_d^1\) trajector decreases relative to the YRI
+        trajectory. Again, statistics for rare variants fluctuate more rapidly
+        than common variants.
+        Dashed lines show neutral expectations.
     }
@@
     \caption{
-        \textbf{.}
+        \textbf{Time series of squared LD under inferred models of human population
+        size history with with epistasis.}
+        Size reductions, such as bottlenecks in the recent past of Eurasian populations,
+        increase squared LD.
+        Dashed lines show neutral expectations.
     }
@@
     \caption{
-        \textbf{.}
+        \textbf{Time series of squared LD under inferred models of human population
+        size history with with dominance.}
+        Again, bottlenecks increase squared LD relative to populations that did
+        not experience such strong size reductions in their history.
+        Dashed lines show neutral expectations.
     }
@@
         in the same gene.}
+        When taking gene-wide averages of LD, there are no obvious differences
+        between the LD decay curves of synonymous and missense mutations.
+        Each panel contains five LD decay curves for each mutation class, one for
+        each \citet{1000_Genomes_Project_Consortium2015-zq} population in the
+        continental groupings. Shared demographic history leads to strong
+        correlation in statistics between populations in the same continental
+        groups.
     }
@@
         that fall inside the same domain.}
+        At short distances between SNPs that both fall within sthe same conserved
+        annotated domain, missense mtuations show consistently
+        larger signed LD than synonymous variants. This may be caused by
+        antagonistic epistasis or compensatory effects between pairs of
+        nearby missense mutations.
     }
@@
         outside of domains.}
+        Outside of conserved domains,
+        missense mutations have reduced signed LD at short distances
+        compared to synonymous mutations,
+        opposite the pattern observed within domains (Figure~\ref{fig:LDwithin}).
+        This pattern is consistent with Hill-Robertson interference between
+        missense mutations outside of conserved elements.
     }
@@
         same domain and outside domains at matched distances.}
+        Showing the five populations in the Thousand Genomes African group.
+        We observe increased LD between missense mutations within domains,
+        but decrased LD outside of domains at matched distances. This pattern
+        holds when conditioning on non-rare variation (\(n_A, n_B \geq 9\),
+        MAF \(\gtrsim 0.05\)).
     }
@@
         same domain and outside domains at matched distances.}
+        Showing the five populations in the Thousand Genomes European group.
+        We again observe increased LD between missense mutations within domains,
+        but decrased LD outside of domains at matched distances. And again,
+        this pattern holds when conditioning on non-rare variation
+        (\(n_A, n_B \geq 9\), MAF \(\gtrsim 0.05\)).
     }
@@
         same domain and outside domains at matched distances.}
+        Showing the five populations in the Thousand Genomes East Asian group.
+        As with the African and European data,
+        wobserve increased LD between missense mutations within domains,
+        but decrased LD outside of domains at matched distances. And again,
+        this pattern holds when conditioning on non-rare variation
+        (\(n_A, n_B \geq 9\), MAF \(\gtrsim 0.05\)).
     }
@@
     \caption{
-        \textbf{Within vs between.}
+        \textbf{LD for pairs of synonymous and missense mutations both within the same
+        domain and in differing domains.}
+        In each panel (one for each population), the left two data points show signed
+        LD between mutations that fall within the same domain, and the right two
+        data points show signed LD between mutations falling within domains but
+        in \emph{different} domains within the same protein-coding gene. Distances
+        are not matched between these data classes (as between-domain pairs are
+        typically more distant than within-domain pairs). Increased LD between
+        missense mutations is only observed for pairs of mutations within the same
+        domain, but not between domains.
     }
@@
         within each gene.}
+        Human genes have different sizes and sequence content, so they may differ in the
+        numbers of missense and synonymous mutations observed within them.
+        The number of missense mutations within a gene is predicted well by
+        the number of synonymous mutations (\(r=0.86\)).
     }
@@
         within each annotated domain.}
+        Domains are typically much smaller than the average gene-size, and while
+        the number of synonymous mutations within a domain is moderately predictive
+        of the number of missense mutations within the same domain (\(r=0.57\)),
+        this correlation is weaker than the correlation in gene-wide mutation
+        counts (Figure~\ref{fig:mutCorrGene}).
     }
@@
         \textbf{Optimization of fraction of new mutations arising via
-        multinucleotide mutations by distance.} A simple exponential function was fit
-        to describe the probability that a pair of mutations arose through a MSM event
-        at a given distance \(d\), as \(Ae^{-\lambda d}\). Across all recombination rates
-        tested, the best fit parameters were \(A=0.11\) and \(\lambda=0.0085\).
+        multinucleotide mutations by distance.} 
+        A simple exponential function was fit to LD decay of synonymous mutations,
+        to describe the probability that a given mutation event is a multinucleotide
+        mutation event that spontaneously gives rise to a pair of perfectly linked
+        mutations at count \(1/2N\).
+        Fitting \(Ae^{-\lambda d}\) to mutations (see Section~\ref{sec:mnm}),
+        the best fit parameters across all recombination rates
+        tested were \(A=2.4\times10^{-5}\) and \(\lambda=0.0092\), showing that even
+        small rates of multinucleotide mutation events can cause increased signed
+        LD at short distances.
     }

### 2022-03-29 manuscript/paper.tex
getting there

@@
 This distortion away from neutral expectations has been empirically documented
-using sequencing data from an ever-growing set of study systems, but questions
+using sequencing data from an ever-growing set of study systems
+\citep{Novembre2009-kc,Cutter2013-mm,Comeron2014-oy}, but questions
 remain about the primary mode of interactions between multiple linked variants

### 2022-03-29 figures/data_coarse.py

@@ -31,23 +31,23 @@ populations = {
 for continent, pops in populations.items():
     for pop in pops:
         ld_pop = pickle.load(
-            open(f"../analysis/parsed_data_v2/{pop}.unphased.all.bp", "rb")
+            open(f"../analysis/parsed_data_v2/{pop}.unphased.all.0.02_to_0.1.bp", "rb")
         )
         data[pop] = ld_pop
         se_pop = pickle.load(
             open(
-                f"../analysis/parsed_data_v2/{pop}.unphased.all.SEs.bp",
+                f"../analysis/parsed_data_v2/{pop}.unphased.all.SEs.0.02_to_0.1.bp",
                 "rb",
             )
         )
         SE[pop] = se_pop
         ld_pop = pickle.load(
-            open(f"../analysis/parsed_data_v2/{pop}.unphased.domains.bp", "rb")
+            open(f"../analysis/parsed_data_v2/{pop}.unphased.domains.0.02_to_0.1.bp", "rb")
         )
         data_dom[pop] = ld_pop
         se_pop = pickle.load(
             open(
-                f"../analysis/parsed_data_v2/{pop}.unphased.domains.SEs.bp",
+                f"../analysis/parsed_data_v2/{pop}.unphased.domains.0.02_to_0.1.SEs.bp",
                 "rb",
             )
         )
@@ -322,9 +322,9 @@ ax2.set_ylim(-0.2, 1)
 #ax4.set_title("$3 \leq n_A, n_B \leq 8$")
 
 ax3 = plt.subplot2grid(dims, (5, 1), rowspan=3)
-plot_domain_data(pops, "geq9", ax3)
+plot_domain_data(pops, "common", ax3)
 ax3.set_xlabel("Populations")
-ax3.set_title("$n_A, n_B \geq 9$")
+ax3.set_title("$p_A, p_B \gtrsim 0.1$")
 ax3.sharey(ax2)
 
 ## plots of LD decay within and outside of domains for AFR populations
@@ -437,16 +437,16 @@ fig2 = plt.figure(98765, figsize=(6.5, 7))
 fig2.clf()
 i = 0
 for pop in populations["Africa"]:
-    for categ in ["all", "leq2", "3to8", "geq9"]:
+    for categ in ["all", "rare", "uncommon", "common"]:
         ax = plt.subplot(5, 4, i + 1)
         if i == 0:
[…]

### 2022-03-30 manuscript/paper.tex
super close

@@
 difference between pairs of missense and synonymous variants within domains is
-especially pronounced for both linked pairs (\(\lesssim\) a few hundred base
-pairs) (Figures~\ref{fig:data}E~and~\ref{fig:LDwithin}). Assuming most
+especially pronounced for linked pairs within a few hundred base pairs of each
+other (Figures~\ref{fig:data}E~and~\ref{fig:LDwithin}). Assuming most
 synonymous mutations are neutral, we would expect signed LD between missense
 mutations to be less than that between synonymous mutations under predictions
-from models of either Hill-Robertson interference, synergistic epistasis, or
-dominance (Figures~\ref{fig:epistasis}~and~\ref{fig:dominance}). This
-observation is opposite to those expectations, suggesting a different
-prevailing interactive effect between nonsynonymous mutations within domains
-than those that are commonly tested or modeled.
+from models of either Hill-Robertson interference, synergistic epistasis, and
+some models of dominance (Figures~\ref{fig:epistasis}~and~\ref{fig:dominance}).
+This observation is opposite to those expectations, suggesting a different
+prevailing interactive effect between nonsynonymous mutations within domains.
 
@@
 common variants recapitulate the opposing patterns of LD that is seen when
-considering pairs at all frequencies. This is because unconditioned statistics
-in the form of \(\sigma_d^1\) and \(\sigma_d^2\) are already dominated by
-common variants.
+averaging over pairs at all frequencies, as unconditioned statistics in the
+form of \(\sigma_d^1\) and \(\sigma_d^2\) are dominated by common variants.
 
@@
 synonymous mutations (Figure \ref{fig:between_domains}). The interactive effect
-driving large LD within domains is therefore likely domain-specific. However,
-the average distance between pairs of mutations within domains is much smaller
-than between domains, so this observation may be primarily driven by the higher
+driving large LD in domains is therefore likely domain-specific. However, the
+average distance between pairs of mutations within domains is much smaller than
+between domains, and this observation may be primarily driven by the higher
 recombination distances between pairs of mutations across distinct domains.
@@
 matched to the same distances as the between-domain comparison, each have LD
-roughly equal to zero.
+roughly equal to zero (Figure~\ref{fig:between_domains}).
 
@@
 synonymous variants separated by short distances have large positive LD, while
-missense mutations have much lower LD (Figure~\ref{fig:LDoutside}). In contrast,
-for mutations within the same domains, missense mutations have more positive
-LD at short distances than synonymous mutations (Figure~\ref{fig:LDwithin}).
+missense mutations have lower LD (Figure~\ref{fig:LDoutside}). In contrast, for
+mutations within the same domains, missense mutations have more positive LD at
+short distances than synonymous mutations (Figure~\ref{fig:LDwithin}).
 
@@
 variants shows that a large majority are strongly deleterious (Supporting
-Information). Instead, negative synergistic epistasis between strongly
-deleterious mutations does produce large negative deviations of mean LD.
-Strongly deleterious \emph{recessive} mutations can also produce this pattern,
-making it difficult to distinguish these effects with certainty. This
-difficulty is compounded by the large measurement noise for \(\E[D]\),
-especially for loss-of-function variants for which only a few hundred
-within-gene pairs exist in the human population data analyzed here.
+Information).
+
+Instead, negative synergistic epistasis between strongly deleterious mutations
+does produce large negative deviations of mean LD. Weakly deleterious
+recessive mutations can also produce this pattern, but strongly deleterious
+recessive mutations lead to slightly positive LD \citep{Roze2021-cf}. While
+most loss-of-function mutations are strongly deleterious, those that rise to
+appreciable frequency are likely more benign and \(\sigma_d^1\) may be driven
+by patterns of weakly deleterious loss-of-function mutations. The difficulty in
+distinguishing these effects is compounded by the large measurement noise for
+\(\E[D]\), especially for loss-of-function variants for which only a few
+hundred within-gene pairs exist in the human population data analyzed here and
+which are separated by larger distances on average than neighboring missense
+and synonymous mutations.
+
+In addition to LD varying by distance, LD can also vary due to differences in
+allele frequencies among classes of mutations (as selection drives mutations to
+higher or lower frequencies). Matching distances between pairs as well as
+allele frequecies between classes of mutations helps to reduce these concerns.
+It is possible that recombination rates can vary between annotated regions,
+resulting in differing patterns of background selection, which can affect
+allele frequencies and LD (e.g., Figures~\ref{fig:bgs1}--\ref{fig:bgs4}). I did
+not condition on local recombination rates or inferred levels of background
+selection here, which is left to future work.
 
@@
 
-Positive average LD between both missense and synonymous mutations has
-previously been reported in humans, drosophila, and other species
-\citep{Sohail2017-zq,Garcia2021-zn,Sandler2021-of}. The similarity of their
-gene-wide LD (Figure~\ref{fig:data}) might suggest that interference or
-interactions between missense mutations are minimal, or at least no stronger
-than those between synonymous mutations. However, averaging over all observed
-pairs of mutations within a gene masks element-specific interactive effects
-that drive LD in opposite directions. Nonsynonymous mutations found within
-conserved protein domains, identified as conserved subunits of a gene with some
+Positive average LD between both missense and synonymous mutations has been
+reported in humans, \emph{Drosophila}, and other species
+\citep{Sohail2017-zq,Sandler2021-of}, while others have found that
+nonsynonymous mutations show lower LD than synonymous mutations
+\citep{Garcia2021-zn}. The similarity of their gene-wide LD observed in this
+study (Figure~\ref{fig:data}) might suggest that interference or interactions
+between missense mutations are minimal, or at least no stronger than those
+between synonymous mutations. However, averaging over all observed pairs of
+mutations within a gene masks element-specific interactive effects that drive
+LD in opposite directions. Nonsynonymous mutations found within conserved
+protein domains, identified as conserved subunits of a gene with some
 structural or functional role \citep{Stanek2020-pa}, are more strongly selected
@@
 compared to distance-matched synonymous mutations outside of domains and to
-mutations within domains. Neither dominance effects, synergistic epistasis, nor
+mutations within domains.  Neither dominance effects (aside from very strongly
+selected recessive mutations \citep{Roze2021-cf}), synergistic epistasis, nor
 Hill-Robertson interference are expected to result in positive LD, so some
-other interactive effect should be driving this positive LD signal within
+other interactive effect should be driving this signal of positive LD within
 conserved domains.
@@
 few hundred base pairs (Figure~\ref{fig:data}E, F). This suggests that
-Hill-Robertson interference is the primary mode of interaction between
-mutations falling outside of domains, as epistasis is expected to impact LD
-over larger distances than what is observed. Importantly, however, the strength
-of epistasis is also likely to be a function of distance between mutations.
+Hill-Robertson interference is the primary mode of interaction between missense
+mutations falling outside of domains, in agreement with \citet{Garcia2021-zn},
+as epistasis is expected to impact LD over larger distances than what is
+observed. Importantly, however, the strength of epistasis is also likely to be
+a function of distance between mutations, complicating this interpretation.
 
@@
 average over these differences by considering all nonsynonymous mutations
-together \citep{Boyko2008-zk,Kim2017-xo}. If patterns of selective differ by
+together \citep{Boyko2008-zk,Kim2017-xo}. If patterns of selection differ by
 both mutational class and location or if loss-of-function variants are in fact
@@
 Spatial population structure may be responsible for the increase of LD observed
-between synonymous variants, as every natural population lives in some
-geography that will necessarily induce some structure. Neutral evolution in
-structured populations alone cannot create this effect. Under standard
-population genetic assumptions of constant mutation and recombination rates and
-independent mutation events, migration, admixture, and spatial structure do not
-result in nonzero \(\E[D]\). Rather, for a set of mutations with \emph{given}
-differences in allele frequencies between source populations, \(D\) can be
-nonzero in an admixed population between those particular mutations even when
-\(D\) is zero in the source populations \citep{Cavalli-Sforza1971-jv}. This
-result relies on conditioning on directional differences in allele frequencies
-between populations. The expected difference in mean allele frequency between
-source populations is zero (\(\E[(p_1-p_2)(q_1-q_2)] = 0\)), as differences are
-equally likely to be positive or negative, and resulting \(\E[D]\) is still
-zero when taking genome-wide averages.
+between synonymous variants. Neutral evolution in structured populations alone
+cannot create this effect. Under standard population genetic assumptions of
+constant mutation and recombination rates and independent mutation events,
+migration, admixture, and spatial structure do not result in nonzero \(\E[D]\).
+Rather, for a set of mutations with \emph{given} differences in allele
+frequencies between source populations, \(D\) can be nonzero in an admixed
+population between those particular mutations even when \(D\) is zero in the
+source populations \citep{Cavalli-Sforza1971-jv}. This result relies on
+conditioning on directional differences in allele frequencies between
+populations. Without conditioning, the expected difference in mean allele
+frequency between source populations is zero (\(\E[(p_1-p_2)(q_1-q_2)] = 0\)),
+as differences are equally likely to be positive or negative, and thus
+resulting \(\E[D]\) is still zero when taking genome-wide averages.
 
@@
 observed decay of \(\sigma_d^1\) between synonymous variants (Supporting
-Information, Figure~\ref{fig:mslmnms}), requiring only a small fraction of
-mutations to cause multinucleotide mutation events. We may therefore treat
-positive LD as the baseline expectation for tightly linked variants primarily
-due to clustered mutations, so that subsequent selective processes and
-interactions cause LD to deviate from that expectation, as seen for missense
-mutations across different sub-elements of genes. It may also be more
-appropriate to compare the negative LD observed between linked loss-of-function
-variants to that positive baseline expectation instead of zero, which would
-imply that they are more recessive or have stronger synergistic epistasis than
-from inferences assuming a neutral expectation of zero.
+Information, Figure~\ref{fig:mslmnms}), with the best-fit model requiring only
+a small fraction of mutations to cause multinucleotide mutation events. We may
+therefore treat positive LD as the baseline expectation for tightly linked
+variants primarily due to clustered mutations, so that subsequent selective
+processes and interactions cause LD to deviate from that expectation, as seen
+for missense mutations across different sub-elements of genes. It may also be
+more appropriate to compare the negative LD observed between linked
+loss-of-function variants to that positive baseline expectation instead of
+zero, which would imply that they are more recessive or have stronger
+synergistic epistasis than from inferences assuming a neutral expectation of
+zero. Additional analyses and extensive simulations will be required to tease
+apart the effects of population structure, selective interference at linked
+sites, and clustered mutational events on patterns of LD.
 
@@
 populations considered here, their joint evolutionary histories make formal
-testing of significance difficult due to shared variation, so they cannot be
+testing of significance difficult due to shared variation, as they cannot be
 treated as independent measurements. Extensive forward simulations will likely
-be required to more thoroughly assess significance.
+be needed to more thoroughly assess significance.
 
@@
 dominance may be feasible using the joint distribution of allele frequencies
-and LD.
+and LD. The results presented in this paper likely do not cover the space of
+all possible two-locus models, and other unexplored models may result in
+similar patterns of signed LD. Comparisons to empirical data should therefore
+be treated with caution, and additional careful modeling will allow a more firm
+determination of the interactive effects causing observed patterns of
+variation.
 
@@
 The two-locus sampling distribution is the direct analog to the single-locus
-site frequency spectrum (SFS) of a given sample size. Instead of describing the
-density or number of mutations with a given allele count out of \(n\) samples,
-the two-locus distribution \(\Psi_n\) stores the density or number of pairs of
-loci with observed haplotype counts, so that \(\Psi_n(i, j, k)\) is the number
-of pairs for which we observe \(i\) copies of the \(AB\) haplotype, \(j\) of
-type \(Ab\), \(k\) of type \(aB\), and \(n-i-j-k\) of type \(ab\). The size of
+site frequency spectrum (SFS) of a given sample size
+(Figure~\ref{fig:TwoLocusFS}). Instead of describing the density or number of
+mutations with a given allele count out of \(n\) samples, the two-locus
+distribution \(\Psi_n\) stores the density or number of pairs of loci with
+observed haplotype counts, so that \(\Psi_n(i, j, k)\) is the number of pairs
+for which we observe \(i\) copies of the \(AB\) haplotype, \(j\) of type
+\(Ab\), \(k\) of type \(aB\), and \(n-i-j-k\) of type \(ab\). The size of
 \(\Psi_n\) grows rapidly, with \(O(n^3)\) entries, which practically limits
@@
 ordinary differential equations directly on the entries of \(\Psi_n\). I
-briefly summarize this general approach below, but refer readers to that
-previous paper for detailed derivations of the drift, recombination, and
-mutation terms and the moment-closure approximation. Instead, here I focus on
-generalizing the selection operator to include epistasis, dominance, and other
-forms of two-locus interactions.
+briefly summarize this general approach below, but refer readers to that paper
+for detailed derivations of the drift, recombination, and mutation terms and
+the moment-closure approximation. Instead, here I focus on generalizing the
+selection operator to include epistasis, dominance, and other forms of
+two-locus interactions.
 
@@
 the dynamics of allelic states of a sample of size of \(n\) from the full
-population. We assume \(n \ll N_e\), and \(r\) and \(s\) are \(O(1/N_e)\) so that
-multiple coalescence, recombination, or selective events within the \(n\)
+population. We assume \(n \ll N_e\), and \(r\) and \(s\) are \(O(1/N_e)\) so
+that multiple coalescence, recombination, or selective events within the \(n\)
 lineages are rare in any given generation. For example, for drift events, we
@@
 event, the probability that haplotype counts changed due to this coalescence
-event {[}see Supporting Information; \citet{Jouganous2017-pq}; \citet{Ragsdale2019-nt}{]}. In
+event \citep[Supporting Information;][]{Jouganous2017-pq,Ragsdale2019-nt}. In
 typical diffusion approximation fashion, we multiple through by \(2N_{ref}\) so
@@
 population, which replace a chosen lineage that fails to reproduce with
-probability proportional to its selection coefficient, assuming negative
-selection.
+probability proportional to its relative fitness.
 
@@
 to that of \(ab\) haplotypes. The standard multiplicative fitness function
-assumes that \(s_{AB} \approx s_{A} + s_{B}\) (\(s^2\approx0\)), and a model
-for epistasis can be written as \[s_{AB} = (s_{A} + s_{B}) (1 + \epsilon),\] so
-that \(\epsilon > 0\) implies synergistic epistasis, while \(\epsilon < 0\)
-implies antagonistic epistasis.
+assumes that \(s_{AB} \approx s_{A} + s_{B}\) (assuming \(s^2\approx0\)), and a
+model for epistasis can be written as \[s_{AB} = (s_{A} + s_{B}) (1 +
+\epsilon),\] so that \(\epsilon > 0\) implies synergistic epistasis, while
+\(\epsilon < 0\) implies antagonistic epistasis.
 
@@
 derived haplotype, each of those sampled lineages has probability of being
-rejected equal to the selection coefficient assigned to its haplotype state. If
-a lineage is rejected, a replacement is drawn from the full population. Under
-the assumption that \(ns \ll 1\), the probability that more than one selection
-event occurs in any given generation is negligibly small, so that the case of
-multiple simultaneous rejections can be ignored. Then \(\Psi_n^{t+1}\) relies
-only on \(\Psi_n^t\) and \(\Psi_{n+1}^t\) for additive selection. The full
-selection operator \(\mathcal{S}\) for additive selection is given in the
-Supporting Information.
-
-To account for dominance within loci, or other general forms of two-locus
-selection, the selection operator no longer reduces to individual haplotypes,
-but instead we need to know the state of two-locus genotypes. For example, the
-fitness of an individual carrying an \(Ab\) haplotype depends on whether their
-second haplotype is \(ab\), \(Ab\), \(aB\), or \(AB\). We can therefore assign
-a selection coefficient to each possible diploid configuration, \(s_{Ab/ab}\),
+rejected equal to the absolute value of the selection coefficient assigned to
+its haplotype state. If a lineage is rejected, a replacement is drawn from the
+full population. Under the assumption that \(ns \ll 1\), the probability that
+more than one selection event occurs in any given generation is negligibly
+small, so that the case of multiple simultaneous rejections can be ignored.
+Then \(\Psi_n^{t+1}\) relies only on \(\Psi_n^t\) and \(\Psi_{n+1}^t\) for
+additive selection. The full selection operator \(\mathcal{S}\) for additive
+selection is given in the Supporting Information.
+
+To account for dominance, or other general forms of two-locus selection, the
+selection operator no longer reduces to individual haplotypes, but instead we
+need to know the state of two-locus genotypes. For example, the fitness of an
+individual carrying an \(Ab\) haplotype depends on whether their second
+haplotype is \(ab\), \(Ab\), \(aB\), or \(AB\). We can therefore assign a
+selection coefficient to each possible diploid configuration, \(s_{Ab/ab}\),
 \(s_{Ab/Ab}\), and so on. Assuming that the doubly homozygous ancestral
@@
 which encompasses dominance within either locus, gene-based dominance, or a
-combination or dominance and epistatic effects, we need \(\Psi_{n+2}^t\).
+combination of dominance and epistatic effects, we need \(\Psi_{n+2}^t\).
 Again, a full derivation and expressions for the general selection operator are
@@
 downsampling to the appropriate sample size. For example, to compute \(\E[D]\),
-the sum is taken over all haplotype configurations \(\mathbf{n} = (n_{AB}, n_{Ab}, n_{aB}, n_{ab})\), weighted by the density \(\Psi_n\) for that
+the sum is taken over all haplotype configurations \(\mathbf{n} = (n_{AB},
+n_{Ab}, n_{aB}, n_{ab})\), weighted by the density \(\Psi_n\) for that
 configuration:
@@
 I considered four variable population size histories, two simple toy models and
-two inferred from human populations in African and Europe. For each size
-history scenario, I tracked the evolution of \(\Psi_n(t)\) for varying
-selection models, plotting the trajectories of \(\sigma_d^1\) and
-\(\sigma_d^2\) over time (Figures~\ref{fig:toy}, \ref{fig:relate_dom}, and
-\ref{fig:toy_sd2}--\ref{fig:relate_dom_sd2}). The selection strength at both
-loci was fixed at either \(\gamma=-1, -2, and -10\), and recombination was set
-to zero.
+two inferred from human populations in African and Europe using \texttt{Relate}
+\citep{Speidel2019-nj}. For each size history scenario, I tracked the evolution
+of \(\Psi_n(t)\) for varying selection models, plotting the trajectories of
+\(\sigma_d^1\) and \(\sigma_d^2\) over time (Figures~\ref{fig:toy},
+\ref{fig:relate_dom}, and \ref{fig:toy_sd2}--\ref{fig:relate_dom_sd2}). The
+selection strength at both loci was fixed at either \(\gamma=-1\) or \(-10\)
+for the models with epistasis, or \(\gamma=-2\) for models with dominance, and
+recombination was set to zero.
 
@@
 its initial size \(1,000\) generations ago. The size histories for YRI and CEU
-were inferred using \texttt{Relate} \citep{Speidel2019-nj} applied to the
-phased autosomal data from \citet{1000_Genomes_Project_Consortium2015-zq},
-using default parameters as recommended in the \texttt{Relate} tutorial.
-\texttt{Relate} returns estimates of coalescence rates within specified time
-bins, and population sizes are estimated as their inverses. Estimates using
-\texttt{Relate} for population sizes in the very recent past (\(<3,000\) years,
-or \(\approx 100\) generations) diverged, so I truncated the history over this
-time period and assumed a constant size from the most recent non-diverged bin.
+were inferred using \texttt{Relate} \citep{Speidel2019-nj} applied to the phase
+3 haplotype-phased autosomal data from
+\citet{1000_Genomes_Project_Consortium2015-zq}, using default parameters as
+recommended in the \texttt{Relate} online tutorial, assuming a mutation rate of
+\(1.25\times10^{-8}\) per-bp per-meiosis and a human generation time of 29
+years. \texttt{Relate} returns estimates of coalescence rates within specified
+time bins, and population sizes are estimated as their inverses. Estimates
+using \texttt{Relate} for population sizes in the very recent past (\(<3,000\)
+years, or \(\approx 100\) generations) diverged, so I truncated the history
+over this time period and assumed a constant size from the most recent
+non-diverged bin.
 
@@
 \citet{1000_Genomes_Project_Consortium2015-zq} (Thousand Genomes) data release,
-I subset the genotype VCFs of the same release to autosomal variants that were
-annotated as either synonymous or nonsynonymous (including both missense and
-more damaging mutations such as stop gains, stop losses, and start losses). I
-further subset to samples within each non-admixed population in the African,
-European, and East Asian continental groups (Table \ref{tab:1kgpops}). Signed
-LD is sensitive to ancestral state misidentification, so I only kept sites for
-which ancestral alleles were estimated with high confidence in both the VCF
-info field and the Thousand Genomes human ancestor reconstructed from a
+I subset the genotype VCFs to autosomal variants that were annotated as either
+synonymous or nonsynonymous, including both missense mutations and more
+damaging ``high impact'' loss-of-function mutations. Loss-of-function
+annotations include frameshifts, splice acceptor, splice donor, start loss,
+stop gain, stop loss, and transcript ablation variants. I further subset to
+samples within each non-admixed population in the African, European, and East
+Asian continental groups (five populations each, Table \ref{tab:1kgpops}).
+Signed LD is sensitive to ancestral state misidentification, so I only kept
+sites for which ancestral alleles were estimated with high confidence in both
+the VCF info field and the Thousand Genomes human ancestor reconstructed from a
 phylogeny of six primates.
@@
 
-I thank Alex Diaz-Papkovich, Eric Friedlander, Simon Gravel, Mashaal Sohail,
-Matthias Steinrücken, and Kevin Thornton for helpful discussions and feedback
-on earlier versions of this manuscript.
+I thank Alex Diaz-Papkovich, Eric Friedlander, Simon Gravel, Carol Lee, Mashaal
+Sohail, Matthias Steinrücken, and Kevin Thornton for helpful discussions and
+feedback on earlier versions of this manuscript.
 

### 2022-03-30 response_to_reviewers/response.tex
responses

\documentclass{article}
\usepackage[round]{natbib}

\usepackage{fullpage}
\usepackage{listings}
\usepackage{url}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{color}
\usepackage{hyperref}
\usepackage{amsmath,amssymb}

\setlength{\parindent}{0pt}
\setlength{\parskip}{1em}

\lstset{language=Python}

% local definitions
\newcommand{\comment}[1]{{\textcolor{red}{Comment: #1}}}

\newenvironment{response}%
  {\list{}{\leftmargin=0.5in\rightmargin=0.5in\color{blue}}\item[]}%
  {\endlist}

\usepackage{lipsum}

\begin{document}

May 27, 2021

GENETICS-2021-304236
Can we distinguish modes of selective interactions using linkage disequilibrium?

Dear Dr. Ragsdale:

Your manuscript has been evaluated by two reviewers.  I agree with the
assessment of the reviewers of that this has the potential to be an important
contribution but that it is not there yet.  Your manuscript would need to be
revised substantially, including conducting additional analysis, to be
acceptable for publication in GENETICS.  Please understand that incremental
changes will not be sufficient.  If you intend to resubmit, please let the
editorial office know approximately how long you expect to need for revisions.
Upon resubmission, please include a detailed response to the reviewers'
comments and to the concerns listed above; it will likely be sent out for
review.  I am hopeful you will be willing to do these revisions as I am excited
about the potential of your manuscript.

\begin{response}
    Dear Dr. Agrawal,

    Thank you for the opportunity to revise this manuscript for consideration
    in Genetics and your patience in completing the revisions. I appreciate
    the careful reading and detailed comments that you and each reviewer have
    made, and I believe that these suggestions have substantially improved the
    paper. Below, my responses to each comment are indented and in blue.

    Some major changes and additions I'd like to highlight:
    \begin{itemize}
        \item The addition of a large suite of numerical validations against
            individual-based simulations, as well as a handful of forward
            simulations to demonstrate the effects of many linked loci on diversity
            statistics
        \item A new Figure 1 that provides some intuition for the two-locus
            sampling distribution and its summaries (contrasted with the SFS
            and its summaries)
        \item Extended numerical analyses, in particular for patterns of LD under
            dominance models
        \item Partitioning LD by allele-frequency binning -- in particular, the
            result of increased LD among missense mutations within conserved
            domains is driven by common variants above $\approx 10\%$
            (I expand on this below)
    \end{itemize}

    \noindent
    Sincerely,\\
    Aaron Ragsdale
\end{response}

The manuscript has a theoretical and empirical component and each needs
revision.  The reviewers have each made a number of comments that should be
addressed and I will summarize the main ones from my perspective.

Theoretical Section: I agree with the reviewers that there is value in your
numerical approach to two-locus sampling distribution. This comment from
Reviewer 2 captures my sentiment: ``the first part of the manuscript does not
seem to fully capitalize on the potential that these new numerical approaches
offer to build quantitative in the extremely large space of models that is
explored.'' Here are what I view as the most important ways the theory section
of the ms should be developed:

1) The numerical approach is based on some approximations but there is no
testing of how well these approximations hold. I agree with both reviewers that
there should be some level of testing (via comparison to forward simulations)
to show how well these approximations work and, when they fail, what kind of
biases they cause.

\begin{response}
    The Supporting Material now includes validation using discrete
    forward simulation. The forward simulations are individual based
    Wright-Fisher simulations, with implementation details in the Supporting
    Material. These comparisons show that the numerical approach introduced in
    this paper is accurate for weak to moderate recombination and selection
    strengths ($\rho, \gamma \lesssim 30$).
    Most simulations only tracked the allele frequencies at the two loci,
    neglecting the background selection-like effects,
    although I also ran a small number of simulations of larger genomic
    regions that allowed many mutations to segregate at once. These are included
    in the Supporting Information.
\end{response}

2) There should be some theoretical results presented for using allele
frequency thresholds or allele frequency bins. LD patterns may differ between
rare versus common variants and, further, this would allow comparison to the
manuscript by Good cited in the current ms (but not very meaningfully
discussed) as raised by Reviewer 2. More generally, to extent to which the
patterns you report in Figures 1-3 were driven by rare or common variants. For
example, I found myself wondering whether the effects of the population size
changes you reported may be different for LD among rare variants vs LD among
common variants. (Further, empirical data may be filtered by frequency for
other reasons; see below.)

\begin{response}
    This is an important point, and I agree that describing expectations of LD
    for different frequency bins and thresholds will be useful for interpreting
    the theoretical and empirical results presented here. From the two-locus
    sampling distribution, it is straightforward to compute LD based on
    arbitrary observed allele count conditions, and I now include figures showing
    expected LD for varying conditions for the same selection scenarios I had
    highlighted before.
\end{response}

3) Both reviewers pointed out that selection occurs at more than two sites and
that pairwise LD could be affected by selection at other loci. That seems an
important caveat that should be stated and ideally tested.

\begin{response}
    Thank you for the suggestion. I now include this caveat at the beginning
    of the results. I also include a some simulations of large regions with
    many selected loci to give a sense of the bias caused by selection at other
    loci than the focal pair.
\end{response}

4) I did not look at the scripts you made available but I would encourage you
to make them as user-friendly as possible. This ms has the potential to make a
bigger impact if it is easy for empirical (not just theoretical!) population
geneticists to use your scripts.

\begin{response}
    I fully agree that scripts should be made public, and I have tried to keep
    and organized an documented/commented repository of all scripts used in the
    analysis here (\url{https://www.github.com/apragsdale/two_locus_selection.git}).
    Additionally, the numerical methods are implemented, documented, and tested
    within the \emph{moments} software (which includes single-locus SFS and
    multi-population LD methods, documentation at \url{https://moments.readthedocs.io/}),
    which I and other developers actively maintain.
\end{response}

Empirical Section: The analysis of the human data builds up the original
analysis of Sohail et al. by focusing primarily on intragenic LD patterns. The
results are presented in a manner to contrast different mutation classes (e.g.,
synonymous, missense, LOFs) with the implication being that a difference in LD
among classes is due to differences in the nature of selection among these site
types. However, there are several key concerns:

5) Allele frequencies. The LD patterns reported using the full spectrum of
allele frequencies. However, the allele frequency spectrum will be different
among these mutational classes, which is an important confounding effect that
should be considered. Reviewer 1 suggests allele frequency matching. In
addition to allele frequency matching, there is good reason to consider
different allele frequency classes separately if you are interested in making
inferences about selection because a substantial fraction of missense mutations
are likely to be effectively neutral (and some might even experience balancing
selection); low frequency missense variants are likely to be in more heavily
enriched for deleterious vs. effectively neutral variants compared to
intermediate or high frequency variants. For this reason, I would like to see
an analysis using only rare variants or, even better, to consider multiple
different allele frequency categories separately (e.g., $p < 0.02$, $0.02 \leq
p \leq 0.1$, $p > 0.1$). Consider your striking pattern for within-domain LD
for missense mutations. My interpretation of what that pattern might indicate
is quite different if it is driven entirely by intermediate frequency variants
rather than by rare variants. Further, examining LD patterns for different
allele frequency bins could dovetail nicely with suggestions made for the
theoretical section.

\begin{response}
    Thank you for this suggestions.
    Previously I had matched synonymous and missense mutations only for distances,
    and not frequency. I now also partition mutations by allele frequencies
    (rare: $p\lesssim 0.02$, uncommon: $0.02\lesssim p \lesssim 0.1$, and common).
    From this, the opposing patterns of increased or decreased LD between missense
    mutations within vs outside of conserved domains is predominantly driven by
    pairs of common variants. The unconditioned statistic \(\sigma_d^1\) is
    already dominated by common variants, so I don't think this result is too
    surprising.

    In interpreting this result, this suggests to me that there is something
    more interesting going on than the explanation that common missense
    mutations are simply effectively neutral. Assuming synonymous mutations are
    effectively neutral (this is of course not always the case, although common
    synonymous mutations would be neutral by the same argument, and we don't
    see a difference in Tajima's D for syn variation within vs outside of domains),
    we still observe increased LD of missense mutations over that of synonymous
    mutations. If missense mutations at these common frequencies were entirely
    neutral, we'd expect their levels of LD
    to be roughly equal to those of synonymous mutations.

    Background selection also appears to have minimal effects on signed LD between
    neutral mutations (Supp Figures 7--10), so I lean toward the explanation that
    some form of direct selection/interactions on those missense mutations (or at
    least a subset of them) is driving this signal. More careful analysis is needed
    to say anything definitively, which I have left for future work.
\end{response}

6) Distances. The average pairwise distance will be lower for mutational
classes that are more abundant. Presumably LOFs are much less abundant than
synonymous or missense mutations so I suspect that the average pairwise
distance between LOFs is larger than for other two types (which probably differ
from one another). Though this won't affect the sign of LD it is obviously
expected to affect the magnitude.

\begin{response}
    In the previous version of the manuscript, I had considered average LD for
    pairs of mutations outside of domains weighted to match the distances between
    mutations observed inside domains. I repeated that analysis here and placed
    more emphasis on comparing pairs at matching distances. For the LOFs, that
    is a good point, although how few LOF pairs exist within the same gene makes
    further subsetting of LOFs difficult. I now include this caveat in the discussion.
\end{response}

Some other comments of my own.

7) I don't like the title as I don't believe it reflects the value of your
contributions. (I like the work presented in this ms but I do not think this ms
does a particularly strong job of carefully considering or answering the
question posed by the title.)

\begin{response}
    I have changed the title, and agree that the previous title wasn't a great
    fit. Now:
    ``A heterogeneous landscape of selection and epistasis  in
    protein-coding genes revealed by two-locus statistics''
\end{response}

8) See the recent PNAS paper from Denis Roze (2021), which has analytical
results showing the effect of dominance on LD. He shows positive LD if $h < 0.25$
and negative LD otherwise.

\begin{response}
    Thank you for bringing this paper to my attention, which I had not seen
    previously. I now include a direct comparison to Denis Roze's paper, shown
    in Figure 4. That result of positive LD when $h<0.25$ appears to only hold
    for very strongly selected mutations. For weak to moderate selection,
    this threshold can change, and some parameter values give LD less than zero
    for all $0 \leq h \leq 1$. I now show these comparisons and results and discuss
    them in the main text.
\end{response}

9) The legend of Figure 4 reads, ``Loss-of-function mutations are defined as
stop and start gains and losses and other nonsense mutations of high impact.''
Please explicitly specify what ``other nonsense mutations of high impact'' means
in the Methods; I couldn't find anything explicit.

\begin{response}
    This is now clarified in the Methods.
\end{response}

10) Line 340-345: You are of course correct when averaging over all allele
frequencies. However, migration generate positive LD among derived variants
when conditioning on those variants being rare (see Sandler et al. 2021, Fig
S3). (An example of how LD patterns can be sensitive to allele frequency
class!)

\begin{response}
    Thanks for this clarification, and I have updated the Discussion to
    make this point clearer.
\end{response}

11) Line 350: The comment about Sandler et al is misleading from my
perspective. The gene flow simulations were not intended to quantitatively
mimic the data (the intention was to simply show a qualitative match, i.e.,
positive LD); it is easy to make LD from gene flow go up or down in magnitude
by tweaking the magnitude and nature of gene flow in the simulation. (Moreover,
that figure was removed for other reasons from the revised version of that ms.)

\begin{response}
    My apologies for potentially misrepresenting the results from those simulations
    in Sandler et al -- that certainly wasn't my intention. I have removed these
    lines in this version of the manuscript, though I do think the observation is
    important that positive LD between synonymous mutations only extends over short
    distances (a few hundred base pairs). I have edited the Discussion about this point.
\end{response}

12) Line 350-360. The point about multinucleotide mutations (MNMs) is
important. I would have liked to see more detail in the legend of Figure S8 of
what was actually done to produce this figure (i.e., enough of a methods
description that one could reproduce this). While the model fits reasonably
well, it implicates what seems like a very proportion of multinucleotide
mutation events. Is there any evidence to support this, perhaps from Icelandic
trio dataset? I am not suggesting you need to do a formal analysis (though it
would be great if you did) but you should comment on whether there is any
evidence that the estimated fraction of MNMs is realistic. (A personal note: I
think you are right to point out that MNMs make the ``null'' expectation to be
positive LD because MNMs are a biological reality. I take the same attitude to
population structure / gene flow. The question then becomes about the relative
importance of the two as the source of positive LD for neutral variants. I do
not claim to know with certainty the answer to that question but I am highly
skeptical that MNMs are sufficient to explain long range positive LD. To be
clear, I am not requiring you to convince me otherwise, but I will be impressed
if you can!)

\begin{response}
    I have extended the methods description for this analysis in the Supplement, as
    well as refined the inference to directy estimate the proportion of mutations
    that result in MNM events instead of simply fitting the probability that a pair
    of mutations were caused by a MNM event.

    In the previous version of the manuscript, I pointed out (though I think not too
    clearly) that the proportion of pairs of mutations observed at a given distance
    that arose through a MNM event is different than the proportion of new mutations
    that are single- vs multi-nucleotide mutation events. The probability that a new
    mutation results in a MNM can be very small (much less than 1\%), but the proportion
    of mutation pairs we observe that were initially MNM events can be quite high
    (e.g., 5--10\%). I expand on this point in the Supplement.

    In fitting the MNM model to synonymous data, I infer that new mutations affect
    multiple nucleotides at relatively small rates ($\ll 1\%$). However, because of
    the scaling of the non-MNM two-locus sampling distribution in $\theta^2$ and
    the scaling of the MNM two-locus sampling distribution proportional to
    $\theta \cdot p_{MNM}$, up to $10\%$ of observed mutations at short distances may
    have arisen from MNM events. I expand on this in the Supplement.
\end{response}

Sincerely,

Aneil Agrawal\\
Associate Editor\\
GENETICS\\

\textbf{Reviewer \#1:}

The manuscript by Ragsdale examines two-locus sampling distributions with
arbitrary natural selection and complex demographies. In particular, the author
uses a numerical approach of moment closures to compute the expected number of
two-locus haplotype counts under complex evolutionary models. Previously,
two-locus sampling distributions were only available for neutral models. Models
including selection were only accessible via forward simulations or assuming
selected variants are independent of one another. As such, this paper addresses
an important problem in the field, providing a much-needed solution.

Further, these two-locus models provide the opportunity to study patterns of
epistasis, which has not received as much attention in the empirical population
genetics literature. Thus, I predict this work will be foundational for future
theoretical and empirical studies.

While I am very positive about the paper as a whole and the potential of the
methods, I have a number of major comments that should be addressed prior to
publication:

1) In the Results and Discussion, it is important to emphasize that these
predictions are for the LD between 2 loci that could be in LD with each other,
but are not linked to other selected loci. In real genomes, the pair of
selected loci could be linked to other nearby selected variants that could have
an effect on the LD of the focal pair. As studying a region with many
deleterious mutations would be even bigger undertaking than that presented
here, I do not think that this problem needs to be solved in this manuscript.
Rather, adding a simple caveat to the paper should suffice.

\begin{response}
    I agree this is an important point, and also think that fully accounting
    for many linked selected mutations is beyond the scope of this paper. I have
    included a caveat when introducing numerical methods.

    I also now include a few forward-in-time individual-based simulations
    of large regions with high rates of selected mutations, to get an idea of the
    magnitude of distortion that can occur due to background selection, associative
    overdominance, and linked purifying selection. These are included in the Supplement.
\end{response}

2) It would be good if some of the predictions for the behavior of LD
statistics in complex models could be verified through forward in time
simulations (using SLiM or something similar). This isn't necessary for all
models (because that is the point of the numerical methods to avoid lots of
simulations), but I think it would add rigor to this work to show that at least
for some models, the predictions match up to the simulations.

\begin{response}
    I have now included a simulation of a region allowing many simultaneous
    deleterious mutations (with only Hill-Robertson interference, that is,
    no epistasis), for additive ($h=0.5$) and recessive ($h=0, 0.1$) mutations.
    The goals of these larger simulations were to explore the effects of multiple
    linked selected loci.

    In addition to these simulation, I ran a large suite of discrete simulations
    tracking haplotype frequencies for pairs of loci under range of epistatic
    and dominance interaction scenarios, comparing the resulting two-locus
    sampling distribution. These are shown in the Supplement and online
    documentation of moments and this manuscript, and show good agreement between
    simulations and the numerical approach introduced here.
\end{response}

3) Analysis of empirical data (Fig 4 and Fig 5): Missense and variants within
protein domains seem to be in more positive LD than those not within protein
domains or synonymous SNPs. This is an interesting and potentially important
result. It does not appear that the different types of variants were matched
for allele frequency. Missense and synonymous SNPs have very different allele
frequency distributions, with the former having more rare variants. It is
well-known that allele frequencies can influence LD statistics (see for
example, vanLiere \& Rosenberg (2008 TPB); Hedrick (1987 Genetics); Eberle et
al. (2006 PLoS Genetics)). Further, the fact that Table S4 shows that Tajima's
D differs among these different types of variants further supports the notion
that the allele frequencies inside protein domains differ from those outside.
Thus, I'm concerned that the differences in LD patterns can be simply driven by
differences in allele frequency, rather than differences in LD driven by
epistasis or compensatory mutations. Matching on allele frequency when
computing the LD statistics would alleviate this problem.

\begin{response}
    As described above, I now condition on both alleles being rare, uncommon,
    or common. From this, we can see that the opposing signals of increased/decreased
    LD between missense mutations within/outside of domains is driven by
    mutations with frequencies greater than $0.1$. Possible implications of this
    observation are now expanded on in the Discussion.
\end{response}

4) Related to point 3 above, when analyzing the empirical data, are the
synonymous and missense variants from the same protein domains? Or, are there
some domains that may only have synonymous variants and others that have only
missense variants? In the latter scenario, the synonymous variants would not be
as good of a control for the missense variants because they are in different
genomic regions. While I think it's great that Tajima's D was compared for
variants within and outside of protein domains, I wonder about other
properties, such as recombination rate or strength of background selection
(maybe quantified by the McVicker B- value). If domains have different
recombination rates than do the coding regions outside of domains, then this
could lead to LD patterns appearing different.

\begin{response}
    I agree that there are many potentially confounding factors when making comparisons
    between different genomic regions (even within the same gene), and even when
    comparing classes of mutations within the same region. I now include
    these caveats in the Discussion.

    To your point about genes and domains containing different numbers of synonymous
    vs missense mutations, I tested the correlation between those numbers at both
    the gene and the domain level (Supp Figures 24 and 25). At the gene level (which
    are larger regions that contain more mutations overall), there is a fairly strong
    correlation between numbers of missense and synonymous mutations ($r=0.86$).
    This correlation is somewhat weaker at the domain level,
    as these are smaller regions ($r=0.57$). Because there are already relatively few
    mutations within domains overall, leading to larger measurement noise, I chose not
    to reduce the total number of compared pairs of mutations even further.
\end{response}

Minor comments:

1) Abstract, line 11: Maybe say, ``distorting allele frequency trajectories and
creating correlations between pairs of mutations''.

\begin{response}
    Thank you for the suggestion.
\end{response}

2) Introduction, line 28: It would be good to provide some references in the
first paragraph about deleterious mutations and negative selection, how they
affect neutral regions. Further, references should be provided when saying that
something is ``empirically documented''.

\begin{response}
    Thank you for the suggestions. I have added references in the first paragraph
    of the Introduction as suggested.
\end{response}

3) The terms ``synergistic'' and ``antagonistic'' epistasis are used extensively
throughout. It would be good to provide a more clear definition of these terms
early on in the manuscript and how they are used here for describing
deleterious mutations.

\begin{response}
    Thank you for the suggestion. In the first paragraph of the Empirical observations
    section, where the term ``synergistic'' epistasis is first used, I attempted
    to make the definition a bit more clear. I do the same for both synergistic and
    antagonistic epistasis when introducing the epistasis models in the results.
\end{response}

4) Caption of Figure 1: I believe that the signs of the synergistic and
antagonistic epistasis are reversed. Given the definition in Methods, shouldn't
epsilon be >0 for synergistic epistasis as more deleterious mutations have
lower fitness than predicted under multiplicative models? Then, epsilon would
be <0 for antagonistic epistasis.

\begin{response}
    Thank you for catching this, and sorry if it caused any confusion. Caption fixed.
\end{response}

5) Figures 1-2: I think it would be good to highlight that these plots show the
expected values of these LD statistics for the given models.

\begin{response}
    I think the Figure 1 caption was already clear that what is shown are
    expectations. I have updated the Figure 2 caption to also clarify that
    it is expectations of LD that are being plotted.
\end{response}

6) Line 142: When mentioning that mean LD can extend over 100kb, I think it
would be good to also put this more in terms of human-like parameters, as the
physical distance over which LD is expected to extend under these models would
depend on the recombination rate.

\begin{response}
    Thank you for the suggestion. I have updated this section to be more explicit
    in this regard.
\end{response}

7) Line 146: Cite some references if saying that the effect of epistasis on LD
is widely studied.

\begin{response}
    With the rearrangement of the text, this line has been removed.
\end{response}

8) Supplementary Figure 1: I think it would be instructive to move this to the
main text to show a more intuitive depiction of the 2-locus sampling
distribution.

\begin{response}
    A new Figure 1 in this version hopefully provides a more intuitive depiction
    of the two-locus sampling distribution and its summaries (with direct comparison
    to the SFS and some of its summaries). Thank for you this suggestion, and I hope
    it is helpful in introducing these statistics.
\end{response}

9) Lines 285-289: These lines seem at odds with what Garcia \& Lohmueller
reported. They found that interference interactions are likely common among
nonsynonymous mutations and that nonsynonymous variants have more LD than
synonymous ones.

\begin{response}
    Thanks for pointing this out, and this is now fixed in the Discussion.
\end{response}

10) Line 417: Yun Song has done some theory on expectations of r2 in Song \&
Song (2007, TPB). This work should be cited here.

\begin{response}
    Thank you for the suggestion. Song and Song, 2007 is now commented on here.
\end{response}

11) Line 548: More detail should be given in this section about how the
analysis of non-steady state demography was done here.

\begin{response}
    This is now expanded on in the Methods.
\end{response}

\textbf{Reviewer \#2:}

In this manuscript, the author investigates expected patterns of linkage
disequilibrium (LD) between two genetic loci under a range of evolutionary
models. In the first part of the manuscript, the author extends a previous
numerical approach by Ragsdale and Gravel (2019) to calculate the expected
sampling distribution of two-locus haplotypes under any selection model. The
author uses this approach to investigate the effects of Hill-Robertson
interference, epistasis, dominance, and changing demography on patterns of LD,
and then uses these insights to analyze and interpret patterns of excess
linkage between non synonymous variants within conserved elements.

In general, I think this is a very interesting and important topic, and I think
there are many interesting observations made by the author. I struggle however
with the manuscript overall, for two reasons. First, while I generally strongly
agree that numerical approaches to estimating population genetic quantities in
more computationally tractable ways than whole-population whole-genome
simulations, the first part of the manuscript does not seem to fully capitalize
on the potential that these new numerical approaches offer to build
quantitative in the extremely large space of models that is explored. Second, I
am not sure that this two-locus intuition can be directly translated to
interpret patterns of human genetic diversity, given that human genomes may be
experiencing selection at a large number of linked loci, which may lead to
behaviors that are not seen in the two-locus model alone. Finally, I have a
technical concern about the limitations of the moment closure approach and
wonder whether these technical limitations force the conclusion about epistasis
made in the last part of the manuscript. I explain my concerns in detail below.

Major comments:

1) As far as I can tell, the calculations in the SI are correct under the
approximations made, but there is very little discussion of the importance of
the approximations made and the limitations these impose for interpreting data.
Specifically, the condition that $r$ and $s$ are order $1/N_e$ and that $n \ll
N_e$ implies that $ns \ll 1$. This means that this approach is unable to
accurately capture the statistics of mutations segregating at frequencies of
order $1/Ns$ or larger. I find this both broadly and specifically concerning.
Specifically, Good (2020) bioRxiv [cited within the discussion] has
demonstrated that it is precisely these mutations that show an excess of LD
that resembles the effects of antagonistic epistases, even in purely additive
models, which is not seen here. This concern is also directly related to the
finding that antagonistic epistasis may be responsible for unusually high LD
within genes -- it is unclear to me whether negative selection alone may be
responsible for these patterns.

\begin{response}
    I appreciate the concern about accuracy of the numerical methods, and I
    have added extensive validations against simulations for the cases
    of additivity, synergistic epistasis, antagonistic epistasis, and recessive
    mutations. For the parameter values considered here
    ($\rho, |\gamma| = 0 - 30$), we find very good agreement between simulation
    and numerical solutions. The condition that $r$ and $s$ are of order $1/N_e$
    is standard in the diffusion approximation, allowing us to neglect terms
    of order $s^2$, $rs$, $s/N_e$, etc. In practice, diffusion methods still
    provide reasonably accurate results and only break down for very strong
    selection beyond the parameter ranges I considered here
    (see, e.g., Krukov and Gravel, bioRxiv 2021)

    I don't exactly follow the conclusion, that ``This means that this approach
    is unable to accurately capture the statistics of mutations segregating at
    frequencies of order $1/Ns$ or larger.'' The condition that $ns\ll 1$ is only
    violated for extremely strong selection (say, $|s| \gtrsim 0.01$, as sample sizes
    that I consider here are not too large). But that strength of selection is beyond
    the results presented in this paper.

    I have expanded in the Methods on the parameter ranges that this method is
    applicable for and point to simulation results showing accuracy over this range
    of parameters. 
\end{response}

2) More broadly, without more detailed exploration of the parameter space of
the models considered and without more work to interpret quanitatively at which
frequencies different behaviors will occur as a function of the parameters, it
is difficult to interpret the generalizability of the results quoted here. Note
that this is especially a concern when the results are used to shape
qualitative insights about the patterns seen in genomes. It would be important
to describe the theoretical reasoning behind the choice of specific values of
parameters the simulations, and why the author believes they tile the space of
important behaviors of a two locus model.

\begin{response}
    I agree that it is difficult to make more definitive claims about the
    causes of variation in patterns of LD seen in the data without a more
    detailed exploration of the relevant parameter space. Ideally, this would
    come in the form of likelihood-based inference of the joint effects of 
    selection and interactions, which I comment on in the Discussion but leave
    to future work.

    I have expanded the theoretical/numerical results to consider additional
    parameters and the dependence on frequency conditioning, though this still
    falls short of a complete exploration of parameter space. I focused primarily
    on selection strengths in the range $\gamma\approx -1$ to $-10$, as these
    selection strengths show the largest effects of interference without 
    epistasis, including for dominance. The very large space of possible selection
    models makes it difficult to fully tile this space. Nonetheless, I
    believe that the results presented in the main text and the expanded results
    in the Supplement are useful for building intuition about the behavior or
    two-locus statistics under different interaction models.
    I have also edited the Discussion to include these caveats and
    point to future work in these directions.
\end{response}

3) On a related note, given the restriction that $n \ll N$, many of these
results should be directly comparable to the recent analytical results from
Good (2020) bioRxiv on expected patterns of LD in a two-locus model. I think
that using the rationale developed by that and similar studies would really
make it possible to capitalize on the ability to rapidly calculate LD in a
broader range of contexts.

\begin{response}
    This is a good suggestion, I spent quite a bit of time trying to implement
    the analytical results from Good (2020). I was unable to make much
    progress in such a comparison. For the most part, I was able to replicate
    the results in Figures 3 and 4 in that paper (e.g., see below, though there
    are qualitative differences between the written formulae and the plots in
    Figure 3 in Good).
    But to compare the analytical results in Good (2020) directly
    to the numerical approximations
    here, we need to make use of the finite sample estimators (Appendix G in Good).
    Unfortunately, I was unable to obtain sensical results implementing the
    formulae in Appendix G, and there may be one (or more) typos in Appendix G
    that make a direct comparison impossible without rederiving those results.

    I did not include this figure in my revision here, because I could not
    faithfully recreate the results presented in Figure 3 in Good 2020.
\end{response}

\begin{figure}[h!]
    \centering
    \includegraphics{./good_comparison}
    \caption{
        Left: Attempt to recreate analytical curves shown in Good (2020) Figure 3,
        using equation C9 and Scipy's \texttt{scipy.special.hyp2f1} function in Python.
        Behavior is qualitatively similar, but differs from Figure 3 in Good. 
        Right: Approximate frequency-conditioned signed LD, where
        $\hat{f}_0 = n_A/n$ and $n_B/n$, so these
        are not unbiased estimates of frequency conditioned $\sigma_d^1$ and cannot
        be compared directly.
    }
    \label{fig:good}
\end{figure}

4) I find many of the results of these simulations really intriguing: it is
notable that population expansions do not appear to change patterns of LD, and
the dominance can have such complex effects on LD (e.g Fig 2B). It would be a
valuable contribution to the literature to expand on these analyses and
describe the intuition behind these observations, though I understand it may be
beyond the scope of this study.

\begin{response}
    I have expanded the numerical and simulation results for both dominance
    and non-steady-state demography. I agree that the complex effects of
    dominance on LD are worth exploring further, especially contrasting the
    results in this paper with Roze (2021). I do think that fully diving into
    this is a bit beyond the scope of this study, but hopefully the expanded
    results and discussion on these points are still helpful.
\end{response}

5) The results on the importance of classifying mutations by the functional
annotations of the regions in which they occur is intriguing and important and
overall I think that despite these patterns being weak, they are interesting
and the section is well-caveated. However, I struggle to make a direct
connection between the intuition derived from previous part of the manuscript
and the conclusions made here. In addition to the concerns raised above to do
with comparing the results of a limited number of two-locus numerical results
to a fundamentally multi-locus system, and the specific concern about the
finding of likely epistasis between mutations within genes, I find that
statements like the one on line 313-316 require further substantiation or
should be weakened: they rely on a quantitative extrapolation that clearly
depends on the parameters and model assumptions.

\begin{response}
    I have attempted to better caveat some of these interpretations, like
    those lines that you highlight. I also now include a statement that more detailed
    modeling and comparisons in a statistical framework will be needed to
    make firm conclusions.
\end{response}

Minor comments:

1) Line 117: Although a well known quantity, it would be good to define $D$
here, since it's the first time it appears.

\begin{response}
    Thank you for the suggestion. $D$ is now defined here.
\end{response}

2) Figure 1: I suggest considering using different colors for panels C-F
compared to A-B to highlight that a different parameter is being varied in
these panels.

\begin{response}
    I have done quite a bit of figure rearrangement, careful to make color schemes
    more distinguishable.
\end{response}

3) Line 151-152: The closed form approached used here should be contrasted with
the fact that it is in principle not possible to include the effects of other
loci, unlike in these other approaches.

\begin{response}
    Thank you for the suggestion. I now point out that it not possible to include
    the effects of other linked loci above this section, when introducing the
    method. That seemed a more natural place to discuss that limitation.
\end{response}

4) Figure 4: consider making the y axis consistent between panels.

\begin{response}
    There are new figures in this version of the paper, with hopefully cleaner figures.
\end{response}

\end{document}

### 2022-03-31 manuscript/paper.tex
so close

@@
         (A) Gene-wide averages of signed LD are slightly positive between pairs
-        of both missense and synonymous mutations. This positive, equal
-        \(\sigma_d^1\) is also observed when conditioning on common variants
-        (minor allele frequencies \(\gtrsim 0.05\)).
+        of both missense and synonymous mutations, considering pairs of mutations
+        at matching distances. This positive, equal \(\sigma_d^1\) is also observed
+        when conditioning on allele frequencies or considering only common variants
+        with minor allele frequencies \(\gtrsim 0.1\)
+        (Tables~\ref{tab:afrsd1}--\ref{tab:eassd2}).
         (B) While there are relatively fewer pairs of loss-of-function mutations
-        within genes, causing larger measurement uncertainty, such pairs have
+        within genes, causing larger measurement uncertainty, such pairs tend to have
         negative average LD. Note that measurement noises for each class of
-        mutations overlap with zero and with each other.
+        mutations overlap with zero and with each other, making it difficult to
+        draw firm conclusions on the patterns of interactions occurring gene-wide.
         (C) Partitioning pairs of mutations as falling within or outside

### 2022-03-31 manuscript/supplement.tex
so close

@@
 \externaldocument{paper}
+
+% external latex tables
+\usepackage{catchfile}
+\begin{filecontents*}{AFR.sd1.tex}
+\end{filecontents*}
 
@@
 
+\CatchFileDef{\afrsigned}{AFR.sd1.tex}{}
+\begin{table}
+    \begin{center}
+        \begin{tabular}{lccc}
+            \afrsigned
+        \end{tabular}
+    \end{center}
+    \caption{
+        \label{tab:afrsd1}
+        \textbf{Signed LD in AFR-labeled populations matched to distances
+        and allele frequencies of synonymous mutations.}
+        Gene-wide signed LD is observed to be at similar slightly positive levels between
+        pairs of synonymous and pairs of missense variants. This observation is
+        consistest when weighting missense variants to match distances between
+        pairs of synonymous variants and when weighting to match synonymous allele
+        frequencies, with only a slight reduction in missense \(\sigma_d^1\).
+        The distribution of distances between synonymous and missense mutations
+        within genes are very similar (Figure~\ref{fig:mutDistances}),
+        so matching by distances separating mutations
+        has a minor effect on measured average LD.
+        The large measurement uncertainty of loss-of-function mutation pairs
+        due to few observed pairs of such mutations within the same gene causes
+        signed LD to fluctuate when conditioning on distances and frequencies
+        matched to synonymous mutations.
+    }
+\end{table}
+
+\CatchFileDef{\afrsquared}{AFR.sd2.tex}{}
+\begin{table}
+    \begin{center}
+        \begin{tabular}{lccc}
+            \afrsquared
+        \end{tabular}
+    \end{center}
+    \caption{
+        \label{tab:afrsd2}
+        \textbf{Squared LD in AFR-labeled populations matched to distances
+        and allele frequencies of synonymous mutations.}
+    }
+\end{table}
+
+\CatchFileDef{\eursigned}{EUR.sd1.tex}{}
+\begin{table}
+    \begin{center}
+        \begin{tabular}{lccc}
+            \eursigned
+        \end{tabular}
+    \end{center}
+    \caption{
+        \label{tab:eursd1}
+        \textbf{Signed LD in EUR-labeled populations matched to distances
+        and allele frequencies of synonymous mutations.}
+        Similar to the pattern observed in AFR-labeled populations
+        (Table~\ref{tab:afrsd1}), \(sigma_d^1\) between missense mutations matched
+        to distances and allele frequencies observed in synonymous mutation pairs
+        has only a small effect on average signed LD, and gene-wide measurements
+        are roughy equal between these two mutation classes.
+        Again, measurement noise for loss-of-function mutation pairs is high.
+    }
+\end{table}
+
+\CatchFileDef{\eursquared}{EUR.sd2.tex}{}
+\begin{table}
+    \begin{center}
+        \begin{tabular}{lccc}
+            \eursquared
+        \end{tabular}
+    \end{center}
+    \caption{
+        \label{tab:eursd2}
+        \textbf{Squared LD in EUR-labeled populations matched to distances
+        and allele frequencies of synonymous mutations.}
+    }
+\end{table}
+
+\CatchFileDef{\eassigned}{EAS.sd1.tex}{}
+\begin{table}
+    \begin{center}
+        \begin{tabular}{lccc}
+            \eassigned
+        \end{tabular}
+    \end{center}
+    \caption{
+        \label{tab:eassd1}
+        \textbf{Signed LD in EUR-labeled populations matched to distances
+        and allele frequencies of synonymous mutations.}
+        Similar to the pattern observed in AFR-labeled populations
+        (Table~\ref{tab:afrsd1}), \(sigma_d^1\) between missense mutations matched
+        to distances and allele frequencies observed in synonymous mutation pairs
+        has only a small effect on average signed LD, and gene-wide measurements
+        are roughly equal between these two mutation classes. 
+        Again, measurement noise for loss-of-function mutation pairs is high.
+    }
+\end{table}
+
+\CatchFileDef{\eassquared}{EAS.sd2.tex}{}
+\begin{table}
+    \begin{center}
+        \begin{tabular}{lccc}
+            \eassquared
+        \end{tabular}
+    \end{center}
+    \caption{
+        \label{tab:eassd2}
+        \textbf{Squared LD in EAS-labeled populations matched to distances
+        and allele frequencies of synonymous mutations.}
+    }
+\end{table}
+
 \clearpage
@@
     \label{fig:mutCorrDomain}
+\end{figure}
+
+\begin{figure}[ht!]
+    \centering
+    \includegraphics{../figures/mutation_class_distances}
+    \caption{
+        \textbf{Distances between mutation pairs for different classes of variants.}
+        Despite different mutation rates between classes of mutations, the
+        proportions of pairs of mutations at given distances are similar.
+        Missense mutations, which have the highest mutation rate within genes,
+        have slightly more pairs of mutations at close distances than synonymous,
+        with loss of function mutations, which have the lowest mutation rate within
+        genes, have the fewest pairs at short distances.
+    }
+    \label{fig:mutDistances}
 \end{figure}

### 2022-03-31 analysis/compute_LD_unphased_domains_freq_matching_0.02_0.1.py

@@ -1,0 +1,878 @@
+import numpy as np
+import itertools
+import gzip
+import moments.LD
+import sys
+import pickle
+from collections import defaultdict
+
+
+lof_annotations = [
+    "frameshift_variant",
+    "splice_acceptor_variant",
+    "splice_donor_variant",
+    "start_lost",
+    "stop_gained",
+    "stop_lost",
+    "transcript_ablation",
+]
+
+import tarfile
+
+try:
+    print("already loaded ancestral seqs", len(ancestral_seqs))
+except NameError:
+    ancestral_seqs = {}
+    print("reading ancestral sequences")
+
+    tar = tarfile.open(
+        "/home/aaronragsdale/Data/1000G/human_ancestor/human_ancestor_GRCh37_e59.tar.bz2",
+        mode="r:bz2",
+    )
+    for member in tar.getmembers():
+        if member.isfile() is False:
+            # print(member)
+            continue
+        if member.get_info()["name"].endswith("fa"):
+            f = tar.extractfile(member)
+            ## can't get SeqIO.parse to read in the extracted file
+            lines = f.readlines()
+            chrom = lines[0].decode().split(":")[2]
+            print("found chrom", chrom)
+            ancestral_seqs[chrom] = "".join([l.decode().strip() for l in lines[1:]])
+            # for record in SeqIO.parse(x, "fasta"):
+            #    chrom = record.id.split(":")[2]
+            #    print(chrom)
+            #    ancestral_seqs[chrom] = record.seq
+    tar.close()
+
+
+def get_annotations(annot_file):
+    annots = {}  # pos: alt: (gene, csq)
+    for line in open(annot_file, "r"):
+        chrom, pos, ref, alt, gene, csq = line.split()
+        annots.setdefault(pos, {})
+        annots[pos][alt] = (gene, csq)
+    return annots
+
+
+def flip_gts(gts):
+    gts_out = []
+    for gt in gts:
+        if gt == "0|0":
+            gts_out.append("1|1")
+        elif gt == "0|1":
+            gts_out.append("1|0")
+        elif gt == "1|0":
+            gts_out.append("0|1")
+        elif gt == "1|1":
+            gts_out.append("0|0")
+        else:
+            raise ValueError("unexpected genotype", gt)
+    return gts_out
+
+
+def to_genotypes(gts):
+    g = np.zeros(len(gts))
+    for ii, gg in enumerate(gts):
[…]

### 2022-03-31 analysis/compute_LD_unphased_freq_and_distance_matching.py

@@ -1,0 +1,901 @@
+"""
+Compute LD with:
+    - no frequency or distance matching
+    - matching distances based on weights of synonymous mutations in bins
+    - matching frequencies based on weights of syn muts as rare/uncommon/common
+        (R/R, R/U, R/C, U/U, U/C, C/C, since R/U and U/R combined)
+    - matching both
+"""
+
+
+import numpy as np
+import gzip
+import moments.LD
+import sys
+import pickle
+
+
+lof_annotations = [
+    "frameshift_variant",
+    "splice_acceptor_variant",
+    "splice_donor_variant",
+    "start_lost",
+    "stop_gained",
+    "stop_lost",
+    "transcript_ablation",
+]
+
+import tarfile
+
+try:
+    print("already loaded ancestral seqs", len(ancestral_seqs))
+except NameError:
+    ancestral_seqs = {}
+    print("reading ancestral sequences")
+
+    tar = tarfile.open(
+        "/home/aaronragsdale/Data/1000G/human_ancestor/human_ancestor_GRCh37_e59.tar.bz2",
+        mode="r:bz2",
+    )
+    for member in tar.getmembers():
+        if member.isfile() is False:
+            # print(member)
+            continue
+        if member.get_info()["name"].endswith("fa"):
+            f = tar.extractfile(member)
+            ## can't get SeqIO.parse to read in the extracted file
+            lines = f.readlines()
+            chrom = lines[0].decode().split(":")[2]
+            print("found chrom", chrom)
+            ancestral_seqs[chrom] = "".join([l.decode().strip() for l in lines[1:]])
+            # for record in SeqIO.parse(x, "fasta"):
+            #    chrom = record.id.split(":")[2]
+            #    print(chrom)
+            #    ancestral_seqs[chrom] = record.seq
+    tar.close()
+
+
+def get_annotations(annot_file):
+    annots = {}  # pos: alt: (gene, csq)
+    for line in open(annot_file, "r"):
+        chrom, pos, ref, alt, gene, csq = line.split()
+        annots.setdefault(pos, {})
+        annots[pos][alt] = (gene, csq)
+    return annots
+
+
+def flip_gts(gts):
+    gts_out = []
+    for gt in gts:
+        if gt == "0|0":
+            gts_out.append("1|1")
+        elif gt == "0|1":
+            gts_out.append("1|0")
+        elif gt == "1|0":
[…]

### 2022-03-31 analysis/compute_LD_unphased_freq_matching_0.02_0.1.py

@@ -1,0 +1,483 @@
+import numpy as np
+import gzip
+import moments.LD
+import sys
+import pickle
+
+
+lof_annotations = [
+    "frameshift_variant",
+    "splice_acceptor_variant",
+    "splice_donor_variant",
+    "start_lost",
+    "stop_gained",
+    "stop_lost",
+    "transcript_ablation",
+]
+
+import tarfile
+
+try:
+    print("already loaded ancestral seqs", len(ancestral_seqs))
+except NameError:
+    ancestral_seqs = {}
+    print("reading ancestral sequences")
+
+    tar = tarfile.open(
+        "/home/aaronragsdale/Data/1000G/human_ancestor/human_ancestor_GRCh37_e59.tar.bz2",
+        mode="r:bz2",
+    )
+    for member in tar.getmembers():
+        if member.isfile() is False:
+            # print(member)
+            continue
+        if member.get_info()["name"].endswith("fa"):
+            f = tar.extractfile(member)
+            ## can't get SeqIO.parse to read in the extracted file
+            lines = f.readlines()
+            chrom = lines[0].decode().split(":")[2]
+            print("found chrom", chrom)
+            ancestral_seqs[chrom] = "".join([l.decode().strip() for l in lines[1:]])
+            # for record in SeqIO.parse(x, "fasta"):
+            #    chrom = record.id.split(":")[2]
+            #    print(chrom)
+            #    ancestral_seqs[chrom] = record.seq
+    tar.close()
+
+
+def get_annotations(annot_file):
+    annots = {}  # pos: alt: (gene, csq)
+    for line in open(annot_file, "r"):
+        chrom, pos, ref, alt, gene, csq = line.split()
+        annots.setdefault(pos, {})
+        annots[pos][alt] = (gene, csq)
+    return annots
+
+
+def flip_gts(gts):
+    gts_out = []
+    for gt in gts:
+        if gt == "0|0":
+            gts_out.append("1|1")
+        elif gt == "0|1":
+            gts_out.append("1|0")
+        elif gt == "1|0":
+            gts_out.append("0|1")
+        elif gt == "1|1":
+            gts_out.append("0|0")
+        else:
+            raise ValueError("unexpected genotype", gt)
+    return gts_out
+
+
+def to_genotypes(gts):
+    g = np.zeros(len(gts))
+    for ii, gg in enumerate(gts):
+        g[ii] = int(gg[0]) + int(gg[2])
+    return g
+
+
[…]

### 2022-03-31 analysis/compute_LD_unphased_freq_matching_nonsynonymous.py

@@ -1,0 +1,472 @@
+import numpy as np
+import gzip
+import moments.LD
+import sys
+import pickle
+
+"""
+Combine missense and lof into "nonsynonymous"
+"""
+
+lof_annotations = [
+    "frameshift_variant",
+    "splice_acceptor_variant",
+    "splice_donor_variant",
+    "start_lost",
+    "stop_gained",
+    "stop_lost",
+    "transcript_ablation",
+]
+
+import tarfile
+
+try:
+    print("already loaded ancestral seqs", len(ancestral_seqs))
+except NameError:
+    ancestral_seqs = {}
+    print("reading ancestral sequences")
+
+    tar = tarfile.open(
+        "/home/aaronragsdale/Data/1000G/human_ancestor/human_ancestor_GRCh37_e59.tar.bz2",
+        mode="r:bz2",
+    )
+    for member in tar.getmembers():
+        if member.isfile() is False:
+            # print(member)
+            continue
+        if member.get_info()["name"].endswith("fa"):
+            f = tar.extractfile(member)
+            ## can't get SeqIO.parse to read in the extracted file
+            lines = f.readlines()
+            chrom = lines[0].decode().split(":")[2]
+            print("found chrom", chrom)
+            ancestral_seqs[chrom] = "".join([l.decode().strip() for l in lines[1:]])
+            # for record in SeqIO.parse(x, "fasta"):
+            #    chrom = record.id.split(":")[2]
+            #    print(chrom)
+            #    ancestral_seqs[chrom] = record.seq
+    tar.close()
+
+
+def get_annotations(annot_file):
+    annots = {}  # pos: alt: (gene, csq)
+    for line in open(annot_file, "r"):
+        chrom, pos, ref, alt, gene, csq = line.split()
+        annots.setdefault(pos, {})
+        annots[pos][alt] = (gene, csq)
+    return annots
+
+
+def flip_gts(gts):
+    gts_out = []
+    for gt in gts:
+        if gt == "0|0":
+            gts_out.append("1|1")
+        elif gt == "0|1":
+            gts_out.append("1|0")
+        elif gt == "1|0":
+            gts_out.append("0|1")
+        elif gt == "1|1":
+            gts_out.append("0|0")
+        else:
+            raise ValueError("unexpected genotype", gt)
+    return gts_out
+
+
+def to_genotypes(gts):
+    g = np.zeros(len(gts))
[…]

### 2022-03-31 figures/data_coarse.py

@@ -31,12 +31,14 @@ populations = {
 for continent, pops in populations.items():
     for pop in pops:
         ld_pop = pickle.load(
-            open(f"../analysis/parsed_data_v2/{pop}.unphased.all.0.02_to_0.1.bp", "rb")
+            #open(f"../analysis/parsed_data_v2/{pop}.unphased.all.0.02_to_0.1.bp", "rb")
+            open(f"../analysis/parsed_data_v2/{pop}.unphased.dist_freq.all.bp", "rb")
         )
         data[pop] = ld_pop
         se_pop = pickle.load(
             open(
-                f"../analysis/parsed_data_v2/{pop}.unphased.all.SEs.0.02_to_0.1.bp",
+                #f"../analysis/parsed_data_v2/{pop}.unphased.all.SEs.0.02_to_0.1.bp",
+                f"../analysis/parsed_data_v2/{pop}.unphased.all.dist_freq.SEs.bp",
                 "rb",
             )
         )
@@ -171,12 +173,12 @@ err = []
 
 for continent in ["Africa", "Europe", "East Asia"]:
     for pop in populations[continent]:
-        yy.append(data[pop]["all_sites"]["all"]["synonymous"]["sd1"])
-        yy.append(data[pop]["all_sites"]["all"]["missense"]["sd1"])
-        yy.append(data[pop]["all_sites"]["all"]["loss_of_function"]["sd1"])
-        err.append(SE[pop]["all_sites"]["all"]["synonymous"]["sd1"])
-        err.append(SE[pop]["all_sites"]["all"]["missense"]["sd1"])
-        err.append(SE[pop]["all_sites"]["all"]["loss_of_function"]["sd1"])
+        yy.append(data[pop]["distances"]["synonymous"]["sd1"])
+        yy.append(data[pop]["distances"]["missense"]["sd1"])
+        yy.append(data[pop]["distances"]["loss_of_function"]["sd1"])
+        err.append(SE[pop]["distances"]["synonymous"]["sd1"])
+        err.append(SE[pop]["distances"]["missense"]["sd1"])
+        err.append(SE[pop]["distances"]["loss_of_function"]["sd1"])
 
 yy = np.array(yy)
 err = np.array(err)

### 2022-03-31 response_to_reviewers/response.tex
ready with the main text

@@
 
-    Some major changes and additions I'd like to highlight:
+    I'd first like to highlight a few of the more substantial changes:
     \begin{itemize}
@@
     Material. These comparisons show that the numerical approach introduced in
-    this paper is accurate for weak to moderate recombination and selection
-    strengths ($\rho, \gamma \lesssim 30$).
+    this paper is accurate for the weak to moderate recombination and selection
+    strengths studied in this paper ($\rho, \gamma \lesssim 30$).
     Most simulations only tracked the allele frequencies at the two loci,
@@
 \begin{response}
-    I fully agree that scripts should be made public, and I have tried to keep
-    and organized an documented/commented repository of all scripts used in the
+    I am happy to make all scripts public, and I have tried to keep
+    an organized an documented/commented repository of all scripts used in the
     analysis here (\url{https://www.github.com/apragsdale/two_locus_selection.git}).
@@
     mutations observed inside domains. I repeated that analysis here and placed
-    more emphasis on comparing pairs at matching distances. For the LOFs, that
-    is a good point, although how few LOF pairs exist within the same gene makes
-    further subsetting of LOFs difficult. I now include this caveat in the discussion.
+    more emphasis on comparing pairs at matching distances (see Supp Fig 26 for a
+    comparison of density of pairs in differnece classes at given distances).
+    For the LOFs, that is a good point, although how few LOF pairs exist within
+    the same gene makes further subsetting of LOFs difficult.
+    I now include this caveat in the Discussion.
 \end{response}
@@
     well as refined the inference to directy estimate the proportion of mutations
-    that result in MNM events instead of simply fitting the probability that a pair
-    of mutations were caused by a MNM event.
+    that result in MNM events instead of simply fitting the probability that an
+    \emph{observed} pair of mutations at a given distance were caused by a MNM event.
 
@@
     clearly) that the proportion of pairs of mutations observed at a given distance
-    that arose through a MNM event is different than the proportion of new mutations
-    that are single- vs multi-nucleotide mutation events. The probability that a new
+    that arose through a MNM event is different than the proportion of new mutation events
+    that are single- vs multi-nucleotide mutations. The probability that a new
     mutation results in a MNM can be very small (much less than 1\%), but the proportion
@@
     documentation of moments and this manuscript, and show good agreement between
-    simulations and the numerical approach introduced here.
+    simulations (in the absense of BGS) and the numerical approach introduced here.
 \end{response}
@@
 \begin{response}
-    As described above, I now condition on both alleles being rare, uncommon,
-    or common. From this, we can see that the opposing signals of increased/decreased
+    Thank you for these suggestions. While I had matched pairs of mutations based
+    on distances, I had not matched by allele frequences. As mentioned above,
+    I now condition on both alleles being rare, uncommon, or common and compare
+    LD across these categories.
+    From this, the opposing signals of increased/decreased
     LD between missense mutations within/outside of domains is driven by
-    mutations with frequencies greater than $0.1$. Possible implications of this
-    observation are now expanded on in the Discussion.
+    mutations with frequencies greater than $0.1$. The agreement between
+    high-frequency-conditioned and unconditioned $\sigma_d^1$ is unsurprising,
+    since $\sigma_d^1$ is dominated by mutations at common frequencies (while
+    $r$ and $r^2$ are more sensitive to rare variants).
+    I now include a much larger set of conditioned LD comparisons (in the
+    Supplement, Tables S6--S11 and Figures S20-S24).
 \end{response}
@@
 antagonistic epistasis are reversed. Given the definition in Methods, shouldn't
-epsilon be >0 for synergistic epistasis as more deleterious mutations have
+epsilon be $>0$ for synergistic epistasis as more deleterious mutations have
 lower fitness than predicted under multiplicative models? Then, epsilon would
-be <0 for antagonistic epistasis.
+be $<0$ for antagonistic epistasis.
 
 \begin{response}
-    Thank you for catching this, and sorry if it caused any confusion. Caption fixed.
+    Thank you for catching this error, and I'm sorry if it caused any confusion.
+    Caption fixed.
 \end{response}
@@
 \begin{response}
-    I think the Figure 1 caption was already clear that what is shown are
-    expectations. I have updated the Figure 2 caption to also clarify that
-    it is expectations of LD that are being plotted.
+    I have updated the figure captions to clarify that
+    expectations of LD that are being plotted.
 \end{response}
@@
     to the SFS and some of its summaries). Thank for you this suggestion, and I hope
-    it is helpful in introducing these statistics.
+    the new figure is helpful in introducing these statistics.
 \end{response}
@@
 \begin{response}
-    Thanks for pointing this out, and this is now fixed in the Discussion.
+    Thanks for pointing this out, and this is now clarified in the Discussion.
 \end{response}
@@
     mutations. For the parameter values considered here
-    ($\rho, |\gamma| = 0 - 30$), we find very good agreement between simulation
+    ($\rho, |\gamma| \approx 0 - 30$), we find very good agreement between simulation
     and numerical solutions. The condition that $r$ and $s$ are of order $1/N_e$
@@
     of order $s^2$, $rs$, $s/N_e$, etc. In practice, diffusion methods still
-    provide reasonably accurate results and only break down for very strong
+    provide reasonably accurate results for larger selection and recombination
+    values, and only break down for very strong
     selection beyond the parameter ranges I considered here
@@
 
-    I have expanded in the Methods on the parameter ranges that this method is
-    applicable for and point to simulation results showing accuracy over this range
-    of parameters. 
+    I have expanded on the parameter ranges that this method is applicable for
+    and point to simulation results showing accuracy over this range of parameters.
 \end{response}
@@
     come in the form of likelihood-based inference of the joint effects of 
-    selection and interactions, which I comment on in the Discussion but leave
-    to future work.
+    selection and interactions, as well as validation against models of BGS.
+    I comment on this in the Discussion but leave it to future work.
 
@@
     the results in Figures 3 and 4 in that paper (e.g., see below, though there
-    are qualitative differences between the written formulae and the plots in
+    are noticeable differences between the written formulae and the plots in
     Figure 3 in Good).
@@
     the effects of other linked loci above this section, when introducing the
-    method. That seemed a more natural place to discuss that limitation.
+    methods.
 \end{response}
@@
 \begin{response}
-    There are new figures in this version of the paper, with hopefully cleaner figures.
+    There are new figures in this version of the paper, with hopefully cleaner presentation.
 \end{response}

### 2022-03-31 manuscript/paper.tex
ready with the main text

@@
 mutations is reduced outside of conserved domains, as would expected under
-Hill-Robertson interference. The heterogeneous landscape of both mutational
-fitness effects and selective interactions within protein-coding genes calls
-for more refined inferences of the joint distribution of fitness and
-interactive effects, and the methods presented here should prove useful in that
-pursuit.
+Hill-Robertson interference. This variation in both mutational fitness effects
+and selective interactions within protein-coding genes calls for more refined
+inferences of the joint distribution of fitness and interactive effects, and
+the methods presented here should prove useful in that pursuit.
 
@@
 both beneficial and damaging, can dramatically impact patterns of variation in
-nearby linked regions \citep[e.g.][]{Smith1974-em,Charlesworth1995-dq,Kim2000-on}.
+nearby linked regions \citep[e.g.,][]{Smith1974-em,Charlesworth1995-dq,Kim2000-on}.
 This distortion away from neutral expectations has been empirically documented
@@
 although the extent of LD depends non-trivially on the strength of selection
-and probability of recombination separating loci
+and the probability of recombination separating loci
 \citep{Hill1966-gv,McVean2000-ox}.
@@
 outside of, protein-coding domains. This suggests an importance for
-antagonistic (or ``diminishing-returns'') epistasis or a prevalence of
-compensatory nonsynonymous mutations within conserved elements.
+antagonistic epistasis or a prevalence of compensatory nonsynonymous mutations
+within conserved elements.
 
@@
 we must turn to population genetic approaches to infer selective interactions
-between observed segregating polymorphisms. Motivated by theory that linked
-negatively selected mutations will display negative LD due to interference
-\citep{Hill1966-gv}, and that epistasis will drive expected LD away from zero,
-a number of recent studies have used patterns of LD within classes of
-putatively selected variants to infer modes of selective interactions.
-\citet{Callahan2011-ac} observed that pairs of tightly linked nonsynonymous
-mutations cluster together more than expected along the same lineages in the
-\emph{Drosophilid} species complex, and that those clustered mutations tend to
-preserve the charge of the protein and were in positive LD compared to pairs of
-synonymous mutations at the same distance. From this, they proposed that
-compensatory nonsynonymous variants are regularly tolerated and maintained.
-More recently, \citet{Taverner2020-lk} replicated this finding across a diverse
-set of genera, showing that such epistatic interactions are important for
-protein evolution.
+between observed segregating polymorphisms. Motivated by theoretical
+predictions that linked negatively selected mutations will display negative LD
+due to interference \citep{Hill1966-gv} and that epistasis will drive expected
+LD away from zero, a number of recent studies have used patterns of LD within
+classes of putatively selected variants to infer modes of selective
+interactions.  \citet{Callahan2011-ac} observed that pairs of tightly linked
+nonsynonymous mutations cluster together more than expected along the same
+lineages in the \emph{Drosophilid} species complex, and that those clustered
+mutations tend to preserve the charge of the protein and were in positive LD
+compared to pairs of synonymous mutations at the same distance. From this, they
+proposed that compensatory nonsynonymous variants are regularly tolerated and
+maintained.  More recently, \citet{Taverner2020-lk} replicated this finding
+across a diverse set of genera, showing that such epistatic interactions are
+important for protein evolution.
 
@@
 epistasis between these mutations, in which the fitness reduction of multiple
-mutations is greater than the product of that of each individual mutations
+mutations is greater than the product of that of each individual mutation
 independently. Both \citet{Sandler2021-of} and \citet{Garcia2021-zn} have
@@
 coding variants. First, for strongly deleterious or loss-of-function mutations,
-their low allele frequencies mean that statistical measurement of LD
-and other diversity measures are very noisy. Second, comparisons are based on
-theory with limiting assumptions, such as steady-state demography, simple
-selection and interaction models, or unlinked loci. To generate predictions
-under more complex models, we rely on expensive forward simulations. Such
-simulations can help build intuition and be used to test inference methods, but
-they do not efficiently provide expectations for quantities of interest across
-the range of relevant parameters. Analytical and numerical methods for expected
-haplotype frequencies and LD under general selective interaction models are
-thus crucial for interpreting patterns of variation observed in data.
+their low allele frequencies mean that measurements of LD and other diversity
+statistics are very noisy. Second, comparisons are based on theory with
+limiting assumptions, such as steady-state demography, simple selection and
+interaction models, or unlinked loci. To generate predictions under more
+complex models, we rely on expensive forward simulations. Such simulations can
+help build intuition and be used to test inference methods, but they do not
+efficiently provide expectations for quantities of interest across the range of
+relevant parameters. Analytical and numerical methods for expected haplotype
+frequencies and LD under general selective interaction models are thus crucial
+for interpreting patterns of variation observed in data.
 
@@
         object, making it difficult to visualize. We can instead visualize conditional
-        distributions of the full sampling distribution, e.g. conditioned on observing
-        five copies of the $A$ allele at the left locus and five copies of $B$ at the
-        right locus \citep[e.g.][]{Hudson2001-sg}.
+        distributions of the full sampling distribution, e.g., conditioned on observing
+        \(n_A\) copies of the \(A\) allele at the left locus and \(n_B\) copies of
+        \(B\) at the right locus \citep[e.g.,][]{Hudson2001-sg}.
         (D) $\sigma_d^2$, which is closely related to $r^2$, is expected to decay
         with increasing recombination distance between loci \citep{Ohta1969-ie}.
-        Selection distorts the level of squared LD away from neutral expectations.
+        Selection distorts squared LD away from neutral expectations.
         (E) $\sigma_d^1$, refered to as signed LD here, is expected to be zero for
@@
 In the Methods, I expand on the moments system developed in
-\citet{Ragsdale2019-nt} to compute the expected sampling distribution
-\(\Psi_n\) (Figure~\ref{fig:TwoLocusFS}) of two-locus haplotypes under a
-general model of selective interactions. This sampling distribution stores the
-expected density or observed counts of pairs of biallelic loci with each
-possible haplotype frequency configuration in a sample of size \(n\). Below, we
-compute expectations for \(\Psi_n\) under varying scenarios of selection and
+\citet{Ragsdale2019-nt} to compute the expected sampling distribution of
+two-locus haplotypes (\(\Psi_n\), Figure~\ref{fig:TwoLocusFS}) under a general
+model of selective interactions. This sampling distribution stores the expected
+density or observed counts of pairs of biallelic loci with each possible
+haplotype frequency configuration in a sample of size \(n\). Below, we compute
+expectations for \(\Psi_n\) under varying scenarios of selection and
 interaction between pairs of loci. It is not possible in this framework to
@@
 In many cases it is simpler to visualize summaries such as the expectation or
-variance of \(D\) (Figure~\ref{fig:TwoLocusFS}D, E) or conditional slices of the
-distribution (Figure~\ref{fig:TwoLocusFS}C) instead of the full
-three-dimensional distribution. Here I focus on low-order LD statistics
-including \(\E[D]\) and \(\E[D^2]\) and their decay with recombination
-distance, as these are statistics that are commonly used to test for
-interactions between loci. For pairs of biallelic loci, with alleles labeled
-\(A/a\) at the left locus and \(B/b\) at the right locus, \(D=f_{AB}-f_A f_B\)
-is the standard covariance measure of LD, where \(f_{AB}\) is the frequency of
-haplotypes carrying both \(A\) and \(B\), and \(f_A\) and \(f_B\) are the
-marginal allele frequencies of \(A\) and \(B\).
+variance of \(D\) (Figure~\ref{fig:TwoLocusFS}D, E) or conditional slices of
+the distribution (Figure~\ref{fig:TwoLocusFS}C) instead of the full
+three-dimensional distribution \(\Psi_n\). Here I focus on low-order LD
+statistics including \(\E[D]\) and \(\E[D^2]\) and their decay with
+recombination distance, as these are statistics that are commonly used to test
+for interactions between loci. For pairs of biallelic loci, with alleles
+labeled \(A/a\) at the left locus and \(B/b\) at the right locus,
+\(D=f_{AB}-f_A f_B\) is the standard covariance measure of LD, where \(f_{AB}\)
+is the frequency of haplotypes carrying both \(A\) and \(B\), and \(f_A\) and
+\(f_B\) are the marginal allele frequencies of \(A\) and \(B\).
 
@@
         \textbf{Hill-Robertson interference.}
-        Interference between pairs of selected mutations can cause deviations
-        from expected patterns of LD under neutrality, without dominance
-        or epistatic effects.
+        Interference between pairs of selected mutations can cause negative
+        signed LD in the absence of dominance and epistatic effects.
         (A) The expected normalized variance of $D$ (\(\sigma_d^2\))
@@
         (B, C) For tightly linked loci (\(4Nr \lesssim 1\)), interference is most
-        noticeable as measured by signed LD for pairs of mutations with
-        \(s \approx 1/N\).
+        noticeable for pairs of mutations with \(s \approx 1/N\).
         At larger recombination distances (\(4Nr > 1\)), signed LD is most
@@
         \(\sigma_d^2 = \E[D^2] / \E[p(1-p)q(1-q)]\) with recombination
-        distance, and right panesl (B and D) show expectations for the
+        distance, and right panels (B and D) show expectations for the
         decay of \(\sigma_d^1 = \E[D] / \E[p(1-p)q(1-q)]\). Dashed lines
@@
 antagonistic epistasis (in which the combined effect is less than the product of
-independent effects) results in positive LD, and large positive LD can occur
-even when the epistatic effect is relatively weak. Epistasis-induced LD can
+independent effects) results in positive LD, and large nonzero LD can occur
+even when epistasis is relatively weak. Epistasis-induced LD can
 extend over long distances, especially for strongly deleterious mutations. For
@@
 \(0.25\), and LD is negative for all \(0\leq h \leq 1\) for weakly deleterious
-mutations with interference strongest between pairs of recessive mutations
-(Figure~\ref{fig:dominance}C).
+mutations with interference strongest between pairs of partially recessive
+mutations (Figure~\ref{fig:dominance}C).
 
@@
         \textbf{Signed LD under inferred models of human population-size history.}
-        Piecewise-constant population size histories inferred by Relate applied to
+        Piecewise-constant population size histories inferred by \texttt{Relate} applied to
         \citet{1000_Genomes_Project_Consortium2015-zq} phase 3 data were used to
@@
 for the Yoruba from Ibidan, Nigeria (YRI) and Utahns of North and West European
-ancestry (CEU) (Figure \ref{fig:relate_dom}B).
+ancestry (CEU) (Figure \ref{fig:relate_dom}A).
 
@@
 mutations within the same gene averaged over all autosomal chromosomes.
-Loss-of-function mutations have more negative LD, possibly suggesting differing
+Loss-of-function mutations have more negative LD, possibly due to differing
 modes of selective interactions for loss-of-function and missense mutations
@@
 \citet{Callahan2011-ac} and \citet{Taverner2020-lk} have proposed such a
-mechanism to explain observed clusters of nonsynonymous mutations in lineages
-of \emph{Drosophila} and other species. Another possibility is a model of
+mechanism to explain observed clusters of nonsynonymous substitutions in
+\emph{Drosophila} and other species. Another possibility is a model of
 antagonistic, or diminishing returns epistasis, in which a single amino
@@
 
-\subsection{Clustered mutations may cause positive LD among synonymous mutations}
+\subsection{Causes of positive LD between synonymous mutations}
 
@@
 clustered mutational events may be common in humans \citep{Besenbacher2016-ac}.
-Indeed, a simple exponential model in which the fraction of pairs of mutations
-arising from a multinucleotide mutation event decays with distance fits the
-observed decay of \(\sigma_d^1\) between synonymous variants (Supporting
-Information, Figure~\ref{fig:mslmnms}), with the best-fit model requiring only
-a small fraction of mutations to cause multinucleotide mutation events. We may
+Indeed, a simple exponential model in which the fraction of mutations causing a
+multinucleotide mutation event decays with distance fits the observed patterns
+of \(\sigma_d^1\) between synonymous variants (Supporting Information,
+Figure~\ref{fig:mslmnms}), with the best-fit model requiring only a small
+fraction of mutations to cause multinucleotide mutation events. We may
 therefore treat positive LD as the baseline expectation for tightly linked
@@
 models, the number of plausible selection scenarios becomes quite large as we
-relax the strict assumptions of additivity and multiplicative fitness effects.
+relax the strict assumptions of additivity and multiplicative interactions.
 The inclusion of dominance effects, epistasis, or other interactions leads to a
@@
 interactions. Again, this underlines the need to jointly infer strengths and
-interactions of selected variants.
+interactions of selected variants and to consider patterns of variation at
+finer genomic scales.
 
 Finally, exploring additional summaries of the two-locus sampling distribution
-should provide additional power to distinguish between interaction models. The
+should increase power to distinguish between interaction models. The
 numerical methods developed here provide the full two-locus sampling
@@
 performing flexible forward simulation
-\citep[e.g.][0]{Thornton2019-qc,Haller2019-vm}. Analytic results for two-locus
+\citep[e.g.,][]{Thornton2019-qc,Haller2019-vm}. Analytic results for two-locus
 distributions under selection are notoriously difficult, with a few notable
@@
 differencing approach to numerically solve the two-locus diffusion equation
-with additive selection at either locus, although that paper focused on the
+with additive selection at either locus, although they focused on the
 applicability of two-locus statistics to demographic inference. More recently,
@@
 that multiple coalescence, recombination, or selective events within the \(n\)
-lineages are rare in any given generation. For example, for drift events, we
-account for the probability that two lineages within the \(n\) tracked lineages
-at time \(t+1\) coalesced in the previous generation and, given a coalescence
-event, the probability that haplotype counts changed due to this coalescence
-event \citep[Supporting Information;][]{Jouganous2017-pq,Ragsdale2019-nt}. In
-typical diffusion approximation fashion, we multiple through by \(2N_{ref}\) so
-that time is measured in \(2N_{ref}\) generations, and we consider scaled
-parameters \(\rho = 4Nr\), \(\theta = 4Nu\), and \(\gamma=2Ns\).
+lineages are rare in any given generation \citep[Supporting
+Information;][]{Jouganous2017-pq,Ragsdale2019-nt}. In typical diffusion
+approximation fashion, we multiple through by \(2N_{ref}\) so that time is
+measured in \(2N_{ref}\) generations, and we consider scaled parameters \(\rho
+= 4Nr\), \(\theta = 4Nu\), and \(\gamma=2Ns\).
 
@@
 computational complexity also grows rapidly in the number of entries of
-\(\Psi_n\), which is \(O(n^3)\). Accuracy and runtime are shown in Figure
-\ref{fig:jackknife}. In the results presented in this paper, sample sizes
-between \(n=30\) and \(n=80\) are used. Derivations for the drift, recombination,
-and mutation operators and the jackknife moment-closure approximation can be
-found in section S1.3 of \citet{Ragsdale2019-nt}, and I repeat the main results in the
-Supporting Information of this paper.
+\(\Psi_n\), which is \(O(n^3)\) (Figure~\ref{fig:jackknife}). In the results
+presented in this paper, sample sizes between \(n=30\) and \(n=80\) are used.
+Derivations for the drift, recombination, and mutation operators and the
+jackknife moment-closure approximation can be found in section S1.3 of
+\citet{Ragsdale2019-nt}, and I repeat the main results in the Supporting
+Information of this paper.
 
@@
 = \E[D^2] / \E[p(1-p)q(1-q)]\) and \(\sigma_d^1 = \E[D] / \E[p(1-p)q(1-q)]\),
-averaged over pairs of variants at all frequencies. However, allele-frequency
-conditioned statistics (such as keeping only loci below some frequency
-threshold as in \citet{Good2022-ot}) can be considered using this same
-approach.
+which can be averaged over pairs of variants at all frequencies.
+Allele-frequency conditioned statistics (such as keeping only loci below some
+frequency threshold as in \citet{Good2022-ot}) can be considered using this
+same approach.
 
@@
 and analyses here were performed using \texttt{moments} version 1.1.10,
-available from \url{https://bitbucket.org/simongravel/moments/src/master/} with
-extensive documentation at \url{https://moments.readthedocs.org}. Scripts to
-run all analyses, recreate figures, and compile this manuscript are available
-at \url{https://github.com/apragsdale/two_locus_selection}.
+available from \url{https://bitbucket.org/simongravel/moments/src/master/} and
+via \texttt{conda}, with extensive documentation at
+\url{https://moments.readthedocs.org}. Scripts to run all analyses, recreate
+figures, and compile this manuscript are available at
+\url{https://github.com/apragsdale/two_locus_selection}.
 

### 2022-03-31 figures/dominance_and_roze.py

@@ -112,8 +112,8 @@ gammas = data_vary["gammas"]
 hs = sorted(data_vary["sd1"].keys())
 
 ax5.plot(-gammas, 0*gammas, "k--", lw=1, label=None)
-for h in hs:
-    ax5.plot(-gammas, data_vary["sd1"][h], lw=1, label=rf"$h={h}$")
+for ii, h in enumerate(hs):
+    ax5.plot(-gammas, data_vary["sd1"][h], color=colors[ii + 4], lw=1, label=rf"$h={h}$")
 
 ax5.set_title(r"$\rho=0$")
 ax5.set_xlabel(r"$\gamma=2Ns$")

### 2022-03-31 figures/hill_robertson.py

@@ -44,7 +44,7 @@ for ii, gamma in enumerate(gammas):
     ax1.plot(
         rhos,
         data["C"][gamma]["all"]["sd2"],
-        #markers[ii] + "-",
+        # markers[ii] + "-",
         "-",
         ms=ms,
         lw=1,
@@ -60,12 +60,12 @@ ax1.set_xlabel(r"$\rho=4Nr$")
 
 ax2 = plt.subplot(1, 3, 2)
 
-ax2.plot(rhos, 0*rhos, "k--", lw=1, label=None)
+ax2.plot(rhos, 0 * rhos, "k--", lw=1, label=None)
 for ii, gamma in enumerate(gammas):
     ax2.plot(
         rhos,
         data["C"][gamma]["all"]["sd1"],
-        #markers[ii] + "-",
+        # markers[ii] + "-",
         "-",
         ms=ms,
         lw=1,
@@ -97,9 +97,9 @@ except IOError:
             print(rho, gamma)
     pickle.dump(sd1s, open(fname, "wb+"))
 
-ax3.plot(gammas, 0*gammas, "k--", lw=1, label=None)
-for rho in rhos:
-    ax3.plot(gammas, sd1s[rho], lw=1, label=rf"$\rho={rho}$")
+ax3.plot(gammas, 0 * gammas, "k--", lw=1, label=None)
+for ii, rho in enumerate(rhos):
+    ax3.plot(gammas, sd1s[rho], color=colors[ii + 4], lw=1, label=rf"$\rho={rho}$")
 
 ax3.set_title("Normalized signed LD ($\sigma_d^1$)")
 ax3.set_ylabel(r"$\sigma_d^1$")
@@ -115,4 +115,4 @@ fig.text(0.36, 0.93, "B", fontsize=8, ha="center", va="center")
 fig.text(0.69, 0.93, "C", fontsize=8, ha="center", va="center")
 
 plt.savefig(f"hill_robertson.pdf")
-plt.savefig(f"hill_robertsonn.png", dpi=300)
+plt.savefig(f"hill_robertson.png", dpi=300)

### 2022-03-31 figures/two-locus-spectrum/plot_stats_explained.py

@@ -235,7 +235,7 @@ fig.clf()
 
 grids = (2, 10)
 
-ax1 = plt.subplot2grid(grids, (0, 0), colspan=4)
+ax1 = plt.subplot2grid(grids, (0, 0), colspan=5)
 
 fs_neu = moments.Demographics1D.snm([10]) * theta
 ax1.plot(fs_neu, "o--", color=colors[0], ms=3, lw=0.5, label="Neutral")
@@ -268,7 +268,7 @@ ax1.text(
 
 ax1.legend(loc="center right")
 
-ax2 = plt.subplot2grid(grids, (0, 4), colspan=4, projection="3d")
+ax2 = plt.subplot2grid(grids, (0, 5), colspan=3, projection="3d")
 
 plot_3d(ax2, F)
 
@@ -283,7 +283,7 @@ ax2.text(10, -2.5, -1, "10", fontsize=6)
 ax2.text(0, 12, 0, "$n_{aB}$")
 
 fig.text(
-    0.70,
+    0.75,
     0.95,
     "Two-locus haplotype sampling distribution",
     fontsize=7,
@@ -291,7 +291,7 @@ fig.text(
     ha="center",
 )
 fig.text(
-    0.70,
+    0.75,
     0.91,
     r"($2Ns_A = -2, 2Ns_B=-2, \epsilon=0$)",
     fontsize=7,
@@ -299,19 +299,19 @@ fig.text(
     ha="center",
 )
 
-fig.text(0.75, 0.80, "\\underline{Summaries:}", fontsize=6, ha="left", va="center")
-fig.text(0.75, 0.75, rf"$r^2 = {get_r2(F):.3f}$", fontsize=6, ha="left", va="center")
+fig.text(0.8, 0.85, "\\underline{Summaries:}", fontsize=6, ha="left", va="center")
+fig.text(0.8, 0.81, rf"$r^2 = {get_r2(F):.3f}$", fontsize=6, ha="left", va="center")
 fig.text(
-    0.75,
-    0.70,
+    0.8,
+    0.77,
     rf"$\sigma_d^2 = {F.D2() / F.pi2():.3f}$",
     fontsize=6,
     ha="left",
     va="center",
 )
 fig.text(
-    0.75,
-    0.65,
+    0.8,
+    0.73,
     rf"$\sigma_d^1 = {F.D() / F.pi2():.3f}$",
     fontsize=6,
     ha="left",
@@ -355,7 +355,7 @@ ax5.legend(fontsize=5)
 
[…]

### 2022-04-01 manuscript/paper.tex
ready to resubmit

@@
 \externaldocument{supplement}
+
+\usepackage{lineno}
+\linenumbers
 
@@
 \begin{figure}[tb!]
+    \internallinenumbers
     \centering
@@
 
-I thank Alex Diaz-Papkovich, Eric Friedlander, Simon Gravel, Carol Lee, Mashaal
-Sohail, Matthias Steinrücken, and Kevin Thornton for helpful discussions and
-feedback on earlier versions of this manuscript.
+I thank Alex Diaz-Papkovich, Eric Friedlander, Simon Gravel, Mashaal Sohail,
+Matthias Steinrücken, and Kevin Thornton for helpful discussions and feedback
+on earlier versions of this manuscript.
 

### 2022-04-01 manuscript/supplement.tex
ready to resubmit

@@
 \title{Supporting Information for\\
-``A heterogeneous landscape of selection and epistasis in protein-coding genes''}
+``A heterogeneous landscape of selection and epistasis in protein-coding genes revealed by two-locus statistics''}
 \author[]{Aaron P. Ragsdale}
-\affil[1]{Department of Integrative Biology, University of Wisconsin, Madison, WI, USA}
-\affil[*]{apragsdale@wisc.edu}
+\affil[]{Department of Integrative Biology, University of Wisconsin, Madison, WI, USA}
+\affil[]{apragsdale@wisc.edu}
 \date{\today}
@@
 and simulations, and residuals do not show any directional bias
-(Figures~\ref{fig:validation}--\ref{fig:last_validation}). The Support Figures
+(Figures~\ref{fig:validation1}--\ref{fig:validation5}). The Supporting Figures
 includes a handful of these comparisons, and the full set of comparisons (over

### 2022-04-01 manuscript/paper.tex
ok now i'm ready

@@
 
-\title{A heterogeneous landscape of selection and epistasis in protein-coding genes revealed by two-locus statistics}
+%\title{A heterogeneous landscape of selection and epistasis in protein-coding genes revealed by two-locus statistics}
+\title{Local fitness and epistatic effects lead to distinct patterns of linkage disequilibrium in protein-coding genes}
 \author[]{Aaron P. Ragsdale}
-\affil[]{Department of Integrative Biology, University of Wisconsin, Madison, WI, USA}
+\affil[]{Department of Integrative Biology, University of Wisconsin--Madison, WI, USA}
 \affil[]{apragsdale@wisc.edu}

### 2022-04-01 manuscript/supplement.tex
ok now i'm ready

@@
 \title{Supporting Information for\\
-``A heterogeneous landscape of selection and epistasis in protein-coding genes revealed by two-locus statistics''}
+``Local fitness and epistatic effects lead to distinct patterns of linkage disequilibrium in protein-coding genes''}
 \author[]{Aaron P. Ragsdale}
-\affil[]{Department of Integrative Biology, University of Wisconsin, Madison, WI, USA}
+\affil[]{Department of Integrative Biology, University of Wisconsin--Madison, WI, USA}
 \affil[]{apragsdale@wisc.edu}

### 2022-04-01 figures/two-locus-spectrum/plot_stats_explained.py

@@ -258,7 +258,7 @@ ax1.set_title("Single-locus site frequency spectrum ($2Ns = -2$)", fontsize=7)
 # )
 
 # inset panel
-ax1.text(3, 50000, "\\underline{Summaries (selected):}", fontsize=6, ha="left", va="center")
+ax1.text(3, 50000, "\\underline{Summaries:}", fontsize=6, ha="left", va="center")
 ax1.text(
     3, 44000, rf"$\pi$ per bp = ${fs.pi() / L:.5f}$", fontsize=6, ha="left", va="center"
 )

### 2022-06-03 manuscript/paper.tex
release 0.2
updates for minor revisions

@@
 \usepackage{lineno}
-\linenumbers
+%\linenumbers
 
@@
 loci, distorting allele frequency trajectories and creating correlations
-between pairs of mutations. A number of recent studies have used patterns of
-linkage disequilibrium (LD) between selected variants to test for selective
+between pairs of mutations. Recent studies have used patterns of linkage
+disequilibrium (LD) between selected variants to test for selective
 interference and epistatic interactions, with some disagreement over
@@
 within genes is driven by strong positive allele-frequency correlations between
-pairs of mutations that fall within the same annotated conserved domain,
-pointing to compensatory mutations or antagonistic epistasis as the prevailing
-mode of interaction within conserved genic elements. LD between missense
-mutations is reduced outside of conserved domains, as would expected under
-Hill-Robertson interference. This variation in both mutational fitness effects
-and selective interactions within protein-coding genes calls for more refined
-inferences of the joint distribution of fitness and interactive effects, and
-the methods presented here should prove useful in that pursuit.
+mutations that fall within the same annotated conserved domain, pointing to
+compensatory mutations or antagonistic epistasis as the prevailing mode of
+interaction within conserved genic elements. LD between missense mutations is
+reduced outside of conserved domains, as expected under Hill-Robertson
+interference. This variation in both mutational fitness effects and selective
+interactions within protein-coding genes calls for more refined inferences of
+the joint distribution of fitness and interactive effects, and the methods
+presented here should prove useful in that pursuit.
 
@@
 
-Most new mutations that affect fitness are deleterious and tend to be
-eliminated from a population. The amount of time that a deleterious mutation
-segregates depends on the strength of selection against genomes that carry it,
-with very damaging mutations kept at low frequencies and purged relatively
-rapidly. In the time between mutation and fixation or loss, selected variants,
-both beneficial and damaging, can dramatically impact patterns of variation in
-nearby linked regions \citep[e.g.,][]{Smith1974-em,Charlesworth1995-dq,Kim2000-on}.
-This distortion away from neutral expectations has been empirically documented
-using sequencing data from an ever-growing set of study systems
-\citep{Novembre2009-kc,Cutter2013-mm,Comeron2014-oy}, but questions
-remain about the primary mode of interactions between multiple linked variants
-and their joint effects on genome-wide patterns of diversity.
+Most new mutations that affect fitness are deleterious and will be eliminated
+from a population. The amount of time that a deleterious mutation segregates
+depends on the strength of selection against genomes that carry it, with very
+damaging mutations kept at low frequencies and purged relatively rapidly. In
+the time between mutation and fixation or loss, selected variants can
+dramatically impact patterns of variation in nearby regions
+\citep{Smith1974-em,Charlesworth1995-dq,Kim2000-on}. This distortion
+away from neutral expectations has been empirically documented using sequencing
+data from an ever-growing set of study systems
+\citep[e.g.,][]{Novembre2009-kc,Cutter2013-mm,Comeron2014-oy}, but questions remain
+about the primary mode of interactions between selected variants and their
+joint effects on genome-wide patterns of diversity.
 
@@
 carrying multiple deleterious variants is equal to the product of the fitness
-reduction of each independent mutation, we expect to see net linkage
-disequilibrium (LD) equal to zero for unlinked sites \citep{Kondrashov1995-va}.
-But for linked loci, those mutations are expected to segregate on different
-haplotypes more often than together, leading to negative, or repulsion, LD,
-although the extent of LD depends non-trivially on the strength of selection
-and the probability of recombination separating loci
-\citep{Hill1966-gv,McVean2000-ox}.
+reduction of each independent mutation, net linkage disequilibrium (LD) is
+expected to be zero for unlinked sites \citep{Kondrashov1995-va}.  However,
+mutations at linked loci are expected to segregate on different haplotypes more
+often than together, leading to negative, or repulsion, LD, and the magnitude
+of LD depends non-trivially on the strength of selection and the probability of
+recombination separating loci \citep{Hill1966-gv,McVean2000-ox}.
 
@@
 
-In this paper, I develop a numerical approach to solve for the two-locus
+Here I develop a numerical approach to solve for the two-locus
 sampling distribution under a general diploid selection model with variable
@@
 is driven by correlations between pairs of mutations within, but not between or
-outside of, protein-coding domains. This suggests an importance for
+outside of, conserved annotated elements. This suggests an importance for
 antagonistic epistasis or a prevalence of compensatory nonsynonymous mutations
-within conserved elements.
+within conserved gene domains.
 
@@
 is through deep mutation scanning experiments, in which many distinct mutations
-are introduced within a target gene and then organismal fitness or some protein
+are introduced within a target gene and then organismal fitness or protein
 function is experimentally measured \citep{Romero2009-yi, Bank2015-vq,
-Puchta2016-rx, Steinberg2016-is}. For example, using the model system of the
+Puchta2016-rx, Steinberg2016-is}. Using the model system of the
 TEM-1 \(\beta\)-lactamase gene in \emph{E. coli}, \citet{Bershtein2006-bk}
@@
 due to interference \citep{Hill1966-gv} and that epistasis will drive expected
-LD away from zero, a number of recent studies have used patterns of LD within
+LD away from zero, multiple recent studies have used patterns of LD within
 classes of putatively selected variants to infer modes of selective
 interactions. \citet{Callahan2011-ac} observed that pairs of tightly linked
-nonsynonymous mutations cluster together more than expected along the same
-lineages in the \emph{Drosophilid} species complex, and that those clustered
-mutations tend to preserve the charge of the protein and were in positive LD
-compared to pairs of synonymous mutations at the same distance. From this, they
-proposed that compensatory nonsynonymous variants are regularly tolerated and
-maintained.  More recently, \citet{Taverner2020-lk} replicated this finding
-across a diverse set of genera, showing that such epistatic interactions are
-important for protein evolution.
+nonsynonymous mutations cluster together more than expected along lineages in
+the \emph{Drosophilid} species complex, and that those clustered mutations tend
+to preserve the charge of the protein and were in positive LD compared to pairs
+of synonymous mutations at the same distance. From this, they proposed that
+compensatory nonsynonymous variants regularly arise and are maintained. More
+recently, \citet{Taverner2020-lk} replicated this finding across a diverse set
+of genera, showing that such epistatic interactions are important in protein
+evolution.
 
@@
 mutations is greater than the product of that of each individual mutation
-independently. Both \citet{Sandler2021-of} and \citet{Garcia2021-zn} have
+independently. Both \citet{Sandler2021-of} and \citet{Garcia2021-zn}
 recently reevaluated patterns of LD between coding variants in humans, fruit
-flies, and \emph{Capsella grandiflora}, and suggested interference and
-dominance may instead be driving patterns of LD \citep{Garcia2021-zn} or
-questioned whether LD between loss-of-function variants is significantly
+flies, and \emph{Capsella grandiflora}, suggesting interference and
+dominance could be driving patterns of LD \citep{Garcia2021-zn} and
+questioning whether LD between loss-of-function variants is significantly
 different from zero \citep{Sandler2021-of}.
 
-A number of factors impede our interpretation of patterns of signed LD between
+Many factors impede our interpretation of patterns of signed LD between
 coding variants. First, for strongly deleterious or loss-of-function mutations,
@@
 \begin{figure}[tb!]
-    \internallinenumbers
+    %\internallinenumbers
     \centering
@@
         \(B\) at the right locus \citep[e.g.,][]{Hudson2001-sg}.
-        (D) $\sigma_d^2$, which is closely related to $r^2$, is expected to decay
+        (D) $\sigma_d^2$, which is closely related to $r^2$, decays
         with increasing recombination distance between loci \citep{Ohta1969-ie}.
         Selection distorts squared LD away from neutral expectations.
-        (E) $\sigma_d^1$, refered to as signed LD here, is expected to be zero for
+        (E) $\sigma_d^1$ (signed LD) is zero for
         pairs of neutral mutations \citep{Hill1968-vu}. Interference between linked
@@
 
-In the Methods, I expand on the moments system developed in
+In the Methods, I expand on the moment system developed in
 \citet{Ragsdale2019-nt} to compute the expected sampling distribution of
@@
 density or observed counts of pairs of biallelic loci with each possible
-haplotype frequency configuration in a sample of size \(n\). Below, we compute
+haplotype configuration in a sample of size \(n\). Below, we compute
 expectations for \(\Psi_n\) under varying scenarios of selection and
-interaction between pairs of loci. It is not possible in this framework to
+interaction between pairs of mutations. It is not possible in this framework to
 include the effects of additional linked selected mutations, such as background
-selection due to many linked variants, and individual-based forward simulations
-are still needed for such scenarios (e.g.,
-Figures~\ref{fig:bgs1}--\ref{fig:bgs4}).
+selection, and individual-based forward simulations are still needed for such
+scenarios (e.g., Figures~\ref{fig:bgs1}--\ref{fig:bgs4}).
 
@@
 the distribution (Figure~\ref{fig:TwoLocusFS}C) instead of the full
-three-dimensional distribution \(\Psi_n\). Here I focus on low-order LD
-statistics including \(\E[D]\) and \(\E[D^2]\) and their decay with
-recombination distance, as these are statistics that are commonly used to test
-for interactions between loci. For pairs of biallelic loci, with alleles
-labeled \(A/a\) at the left locus and \(B/b\) at the right locus,
+three-dimensional distribution \(\Psi_n\). For pairs of biallelic loci, with
+alleles labeled \(A/a\) at the left locus and \(B/b\) at the right locus,
 \(D=f_{AB}-f_A f_B\) is the standard covariance measure of LD, where \(f_{AB}\)
 is the frequency of haplotypes carrying both \(A\) and \(B\), and \(f_A\) and
-\(f_B\) are the marginal allele frequencies of \(A\) and \(B\).
-
-Instead of unnormalized \(\E[D^2]\) and \(\E[D]\) I consider expectations for
-\(\sigma_d^2 = \E[D^2]/\E[f_A(1-f_A)f_B(1-f_B)]\) and \(\sigma_d^1 =
-\E[D]/\E[f_A(1-f_A)f_B(1-f_B)]\).  Such normalized statistics have two
-benefits: first, the mutation rate cancels so that expectations are robust to
-assumptions about the per-base mutation rate, and second, we can compare to
-analytic expectations for these quantities under neutrality and constant
-population size \citep{Ohta1971-yd}.
+\(f_B\) are the marginal allele frequencies of \(A\) and \(B\). Here I focus on
+low-order LD statistics and their decay with recombination distance, as these
+are statistics that are commonly used to test for interactions between loci.
+Instead of \(\E[D^2]\) and \(\E[D]\) I consider expectations for \(\sigma_d^2 =
+\E[D^2]/\E[f_A(1-f_A)f_B(1-f_B)]\) and \(\sigma_d^1 =
+\E[D]/\E[f_A(1-f_A)f_B(1-f_B)]\). Normalized statistics have two benefits: (1)
+the mutation rate cancels so that expectations are robust to assumptions about
+the per-base mutation rate, and (2) we can compare to analytic expectations for
+these quantities under neutrality and constant population size
+\citep{Ohta1971-yd}.
 
@@
 within loci but without epistatis, and then describe a general diploid
-selection model and consider gene-based dominance effects. In what follows I
-focus on parameters where the strengths of selection and dominance at each
-locus are equal, but note that the methods presented here allow for arbitrary
-and unequal selection and dominance at the two loci. I also focus primarily on
-weak to moderate negative selection (\(|2Ns| \approx 1 - 20\)), since this
-range of selection leads to the strongest signals of interference
-(Figure~\ref{fig:HillRobertson}) and is the parameter regime for which the
-numerical appraoch is most accurate.
+selection model and consider gene-based dominance effects. I present results
+for equal strengths of selection and dominance at each locus, but note that the
+methods presented here allow for arbitrary and unequal selection and dominance
+at the two loci. I also focus primarily on weak to moderate negative selection
+(\(|2Ns| \approx 1 - 20\)), since this range of selection leads to the
+strongest signals of interference (Figure~\ref{fig:HillRobertson}) and is the
+parameter regime for which the numerical approach is most accurate.
 
@@
         \textbf{Hill-Robertson interference.}
-        Interference between pairs of selected mutations can cause negative
-        signed LD in the absence of dominance and epistatic effects.
+        Interference between pairs of selected mutations causes negative
+        signed LD.
         (A) The expected normalized variance of $D$ (\(\sigma_d^2\))
@@
         Left panels (A and C) show expectations for the decay of
-        \(\sigma_d^2 = \E[D^2] / \E[p(1-p)q(1-q)]\) with recombination
+        \(\sigma_d^2\) with recombination
         distance, and right panels (B and D) show expectations for the
-        decay of \(\sigma_d^1 = \E[D] / \E[p(1-p)q(1-q)]\). Dashed lines
+        decay of \(\sigma_d^1\). Dashed lines
         show neutral expectations.
@@
         antagonistic epistasis (\(\epsilon < 0\))
-        gives rise to positive signed LD and increased $\sigma_d^2$ over
+        causes positive signed LD and increased $\sigma_d^2$ over
         a multiplicative model (\(\epsilon = 0\)),
@@
 between selected mutations, which is strongest for pairs of mutations that have
-selection coefficients \(\gamma = O(1)\), or \(s \approx 1/2N_e\) (Figure
+selection coefficients \(\gamma = 2N_es = O(1)\), or \(s \approx 1/2N_e\) (Figure
 \ref{fig:HillRobertson}).
@@
 even when epistasis is relatively weak. Epistasis-induced LD can
-extend over long distances, especially for strongly deleterious mutations. For
-example, in Figure \ref{fig:epistasis}F even moderately deleterious mutations with
+extend over long distances, especially for strongly deleterious mutations.
+Even moderately deleterious mutations with
 population-size-scaled selection coefficients of \(\gamma=-10\) show large mean
-LD that extends to values of \(\rho\) much greater than 1 (for humans, assuming
+LD that extends to values of \(\rho\) much greater than 1
+(Figure \ref{fig:epistasis}F; for humans, assuming
 roughly 1 cM/Mb, this is on the order of 100 kilobases or more).
@@
         (\(h=0.05\) results in \(s=-0.2\) and \(\gamma=-400\) for homozygous
-        diploids at a single locus). The numerical appraoch for \(\Psi_n\)
+        diploids at a single locus). The numerical approach for \(\Psi_n\)
         cannot handle such strong selection.
@@
 large-scale forward simulations to explore how dominance impacts patterns of
-LD, showing that LD depends on the magnitudes of both the selection and
-dominance coefficients in a nonlinear way. \citet{Roze2021-cf} found an
-analytic expression for LD between pairs of strongly deleterious mutations
-under steady-state demography, showing that LD can be either positive or
-negative depending on the strength of dominance.
-
-The combined effect of the strength of selection and dominance on interference
-is indeed nontrivial (Figure~\ref{fig:dominance}A), as observed by
-\citet{Garcia2021-zn}. Some parameter regimes can cause strong negative LD
-between pairs of negatively selected variants, with moderately selected
-recessive variants having stronger signals of interference than additive
-selection (Figure~\ref{fig:dominance}B). Unlike epistatic interactions, signed
-LD decays rapidly with increasing distance between loci and is roughly zero for
-\(\rho \gg 1\). For weakly selected mutations with \(\gamma=-1\), there is no
-monotonic effect of the level of dominance on negative LD, with both recessive
-and dominant pairs of mutations having more negative LD than the additive case.
-
-Discrete simulations and the moments approach confirm the result from
+LD, showing that LD depends nonlinearly on the magnitudes of both selection
+and dominance. \citet{Roze2021-cf} found an analytic expression
+for LD between pairs of strongly deleterious mutations under steady-state
+demography, showing that LD can be either positive or negative depending on the
+strength of dominance.
+
+The combined effect of the strengths of selection and dominance on interference
+is indeed nontrivial (Figure~\ref{fig:dominance}A). Some parameter regimes can
+cause large negative LD between negatively selected variants, with
+moderately selected recessive variants showing stronger signals of interference
+than additive selection (Figure~\ref{fig:dominance}B). Unlike epistatic
+interactions, signed LD decays rapidly with increasing distance between loci
+and is roughly zero for \(\rho \gg 1\). For weakly selected mutations
+(\(|\gamma|\approx 1\)), there is no monotonic effect of the level of dominance
+on negative LD, with both recessive and dominant pairs of mutations having more
+negative LD than additive mutations.
+
+Discrete simulations and the moment approach confirm the result from
 \citet{Roze2021-cf}, that positive LD can occur for strong negative selection
@@
 \(0.25\), and LD is negative for all \(0\leq h \leq 1\) for weakly deleterious
-mutations with interference strongest between pairs of partially recessive
 mutations (Figure~\ref{fig:dominance}C).
 
-\subsubsection{General selection and gene-based dominance}
+\subsubsection{Arbitrary two-locus selection models}
 \label{sec:general-selection}
@@
         other two-locus summaries of the data may help to distinguish between
-        such models.
+        interaction models.
         }
@@
 
-Beyond the standard models of epistasis and dominance shown above, a large
-family of selection models can be specified by assigning unique fitness effects
-to each possible diploid pair of haplotypes. If we assume the diploid genotype
-that is homozygous for the ancestral alleles (\(ab/ab\)) has fitness 1, then
-there are nine other possible diploid two-locus genotypes that could be given
-unique fitnesses (Table \ref{tab:selmodels}), noting that \(AB/ab\) and
-\(Ab/aB\) genotypes can have differing selection coefficients.
-
-The case with \(s_{AB/ab} \not= s_{Ab/aB}\) arises in a scenario where a
-mutation at either locus within a haplotype impacts some functional region or
-element, but a diploid individual carrying at least one copy that is free of
-mutations has minimal fitness loss. In this ``gene-based dominance'' scenario
+Beyond standard models of epistasis and dominance, a large family of selection
+models can be specified by assigning unique fitness effects to each possible
+diploid pair of haplotypes. Assuming the diploid genotype homozygous for the
+ancestral alleles (\(ab/ab\)) has relative fitness 1, there are nine other
+diploid two-locus genotypes that could be given unique fitnesses (Table
+\ref{tab:selmodels}), noting that \(AB/ab\) and \(Ab/aB\) genotypes can have
+differing selection coefficients.
+
+The case with \(s_{AB/ab} \not= s_{Ab/aB}\) may arise in a scenario where a
+mutation at either locus within a haplotype impacts some functional element,
+but a diploid individual carrying at least one copy that is free of mutations
+has minimal fitness loss. In this ``gene-based dominance'' scenario
 \citep[e.g.,][]{Sanjak2017-zw}, an \(AB/ab\) genotype has higher fitness than
-an \(Ab/aB\) type (a simple implementation of this model is given in the final
-column in Table \ref{tab:selmodels}). Such a gene-based fitness model gives
-similar expected positive signed LD to the model of antagonistic epistasis
+an \(Ab/aB\) type (Table~\ref{tab:selmodels}). Such a gene-based fitness model gives
+positive signed LD, similar to the model of antagonistic epistasis
 (Figure~\ref{fig:gene-based-compensatory}), although the interpretation of
@@
         \textbf{The effects of demography on signed LD.}
-        Simulations under two toy models, one with an instantaneous expansion in
-        the past and the other with a five-fold reduction and recovery illustrated
-        in (A), show the effects of population size changes on expected patterns of LD. 
+        (A) Simulations under histories of an instantaneous size expansion
+        and a five-fold reduction and recovery.
         For two selection strengths (\(\gamma=-1\) and \(\gamma=-10\)) and three
@@
         synergistic epistasis (\(\epsilon=1/2\), C), and antagonistic epistasis
-        (\(\epsilon=-1/2\), D), each with \(\rho=0\)),
-        sudden decreases in population size can cause
+        (\(\epsilon=-1/2\), D), each with \(\rho=0\)), a
+        sudden decrease in population size can cause
         large changes in signed LD, often in the opposite direction than more
@@
         sample size \(n=50\)) is more sensitive to population size changes.
-        (F) Signed LD between pairs of common alleles (both \(n_A, n_B \geq 5/50\))
+        (F) Signed LD between pairs of common alleles (both \(n_A, n_B \geq 5\))
         is comparatively more stable over time.
+        Colors are matched to conditions shown in B--D.
         Additional comparisons, including for dominance models and showing
@@
         simulate time series of two-locus statistics, as in Figure~\ref{fig:toy}.
-        (A) The CEU (Utah residents with Northern and Western European ancestry)
-        are inferred to have a stronger bottleneck than the YRI
-        (Yoruba in Ibadan, Nigeria) 10--100ka, reflecting the out-of-Africa
-        event.
-        Here, we highlight the effect of size changes on \(\sigma_d^1\) under
-        non-additive selection models with \(\rho=0\) and \(\gamma=-2\) at both loci,
-        comparing standard interference (B) to site-wise dominance (C) and gene-based
-        dominance (C).
-        As with epistasis, more severe bottlenecks have larger effects on signed LD,
-        and LD among common variants is more stable than among pairs of rare variants.
+        (A) The CEU are inferred to have a stronger bottleneck than the YRI
+        10--100ka, reflecting the out-of-Africa event.
+        For fully linked loci (\(\rho=0\)) and \(\gamma=-2\) at both loci,
+        I compare the effects of
+        (B) standard interference, (C) site-wise dominance, and (D) gene-based
+        dominance on \(\sigma_d^1\).
+        As with epistasis, more severe bottlenecks have larger effects on signed LD.
+        (E, F) LD among common variants is more stable than among pairs of
+        uncommon variants. Colors are matched to conditions shown in B--D.
         Additional comparisons with epistasis and showing \(\sigma_d^2\) are
@@
 
-The moment system for \(\Psi_n\) readily incorporates variable population size as
-well as mutation and recombination rates and selection parameters that change
-over time. Here I focus on non-equilibrium population size history and consider
-scenarios relevant to human demographic history, including bottlenecks and
-expansions. I explore two simple models (Figure \ref{fig:toy}A), one with an
+The moment system for \(\Psi_n\) readily incorporates variable population size.
+I explore two simple size history models (Figure \ref{fig:toy}A), one with an
 instantaneous expansion and another with a bottleneck followed by recovery. I
@@
 genealogy reconstruction \citep{Speidel2019-nj} applied to the
-\citet{1000_Genomes_Project_Consortium2015-zq} dataset, and focus on size histories
-for the Yoruba from Ibidan, Nigeria (YRI) and Utahns of North and West European
-ancestry (CEU) (Figure \ref{fig:relate_dom}A).
+\citet{1000_Genomes_Project_Consortium2015-zq} dataset, and focus on size
+histories for the Yoruba from Ibidan, Nigeria (YRI) and Utahns of North and
+West European ancestry (CEU) (Figure \ref{fig:relate_dom}A).
 
@@
 epistasis, dominance within loci, and gene-based dominance. In general across
-each selection model, population size expansions are not expected to strongly
-affect \(\sigma_d^1\), whether that expansion occurs deeper in the past as in
-the simple expansion model or rapid expansion more recently, as in the YRI. On
-the other hand, population size reductions tend to push signed LD to more
-extreme values and subsequent recoveries or expansion again reduce the
-magnitude of LD. Under no selection condition tested here do population size
-changes cause expected LD to change sign, showing that while the magnitude of
-deviation of LD from zero is sensitive to population size history, interpreting
-general patterns of the observed sign of LD in data should not be strongly
-affected by population size history.
+each selection model, population size expansions do not strongly affect
+\(\sigma_d^1\), whether that expansion occurs deeper in the past as in the
+simple expansion model or rapid expansion more recently, as for YRI. On the
+other hand, population size reductions tend to push signed LD to more extreme
+values and subsequent recoveries or expansion again reduce the magnitude of LD.
+Under no selection condition tested here do population size changes cause
+expected LD to change sign, showing that while the magnitude of deviation of LD
+from zero is sensitive to population size history, interpretions of the
+observed sign of LD should be robust to population size history.
 
@@
         \textbf{LD in human protein-coding genes and annotated domains.}
-        (A) Gene-wide averages of signed LD are slightly positive between pairs
-        of both missense and synonymous mutations, considering pairs of mutations
+        (A) Gene-wide averages of signed LD are slightly positive
+        for both missense and synonymous mutations, considering pairs of mutations
         at matching distances. This positive, equal \(\sigma_d^1\) is also observed
@@
         (B) While there are relatively fewer pairs of loss-of-function mutations
-        within genes, causing larger measurement uncertainty, such pairs tend to have
-        negative average LD. Note that measurement noises for each class of
+        within genes, causing larger measurement uncertainty, they have
+        negative average LD. Measurement noises for each class of
         mutations overlap with zero and with each other, making it difficult to
@@
         of domains were matched to within-domain mutation pair distances.
-        (D) This pattern again holds for common variants.
-        (E, F) The signal of increased LD between missense variants within domains
-        and decreased LD outside of domains is driven largely by tightly linked
-        mutations (distances \(\lesssim 400\) bp), here showing African populations
-        in the \citet{1000_Genomes_Project_Consortium2015-zq}.
-        Additional comparisons are shown in the Supporing Information, and Thousand
-        Genomes population labels are described in Table~\ref{tab:1kgpops}.
+        (D--F) Rare and uncommon variants (with frequencies $<0.1$ have positive
+        LD, with synonymous LD exceeding missense LD in many populations. However,
+        common missense variants within domains have large positive LD and are
+        responsible for the pattern seen in (C).
     }
@@
 genes partitioned by functional annotations. Synonymous and missense mutations
-show similar levels of slighly positive signed LD when considering pairs of
+show similar levels of slightly positive signed LD when considering pairs of
 mutations within the same gene averaged over all autosomal chromosomes.
@@
 modes of selective interactions for loss-of-function and missense mutations
-(Figure \ref{fig:data}A--B). Within each population, measurement noise gives
+(Figure~\ref{fig:data}A~and~B). Within each population, measurement noise gives
 95\% confidence intervals that overlap with zero in each mutation class,
@@
 
-The similarity in signed LD between missense and synonymous mutations would
+The similarity in signed LD between missense and synonymous mutations might
 suggest that interference between missense mutations is minimal, or at least no
@@
 effects differ dramatically between pairs of mutations found in different
-intra-genic regions, with opposing effects canceling out when taking gene-wide
-averages. Due to the rarity of loss-of-function mutations, I only compare
-synonymous and missense mutations when looking at finer partitions of mutations
-within genes.
-
-Annotated conserved domains in protein-coding genes play a significant role in
-driving signals of positive LD within genes. Such protein-coding domains are
-conserved elements of genes, often associated with some known functional or
-structural feature of a protein \citep{Stanek2020-pa}. Purifying selection is
-expected to be stronger within conserved domains than within the same gene but
-outside of those domains. Indeed, the site-frequency spectrum (SFS) is skewed
-to lower frequencies for both missense and loss-of-function mutations within
-domains when compared to the same classes of mutations outside of domains, with
-much more negative values of Tajima's D within domains (Table
-\ref{tab:tajimasD}). On the other hand, no difference is observed for
-synonymous mutations whether within or outside domains, suggesting roughly
-equivalent effects of selection (either direct or linked) on synonymous
-variation.
-
-Pairs of missense mutations that both fall within the same functional domain
-have large positive LD that is elevated above that of pairs of synonymous
-mutations that both fall within the same domain
-(Figures~\ref{fig:data}C--F~and~\ref{fig:LDwithin}--\ref{fig:domainsEAS}). This
-difference between pairs of missense and synonymous variants within domains is
+intragenic regions. Due to the rarity of loss-of-function mutations, I only
+compare synonymous and missense mutations when looking at finer partitions of
+mutations within genes.
+
+Annotated conserved domains in protein-coding genes drive signals of positive
+LD between missense variants. Such protein-coding domains are conserved
+elements of genes, often associated with some known functional or structural
+feature of a protein \citep{Stanek2020-pa}. Purifying selection is expected to
+be stronger within conserved domains than within the same gene but outside of
+those domains.  Indeed, the site-frequency spectrum (SFS) is skewed to lower
+frequencies for both missense and loss-of-function mutations within domains
+when compared to the same classes of mutations outside of domains, with much
+more negative values of Tajima's D within domains
+(Tables~\ref{tab:tajimasD}~and~\ref{tab:msldfe}).  On the other hand, no
+difference is observed for synonymous mutations whether within or outside
+domains, suggesting roughly equivalent effects of selection (either direct or
+linked) on synonymous variation.
+
+Missense mutations within the same functional domain have large positive LD
+that is elevated above that of synonymous mutations within the same domain
+(Figures~\ref{fig:data}C~and~\ref{fig:LDwithin}--\ref{fig:domainsEAS}). This
+difference between missense and synonymous variants within domains is
 especially pronounced for linked pairs within a few hundred base pairs of each
-other (Figures~\ref{fig:data}E~and~\ref{fig:LDwithin}). Assuming most
-synonymous mutations are neutral, we would expect signed LD between missense
-mutations to be less than that between synonymous mutations under predictions
-from models of either Hill-Robertson interference, synergistic epistasis, and
-some models of dominance (Figures~\ref{fig:epistasis}~and~\ref{fig:dominance}).
-This observation is opposite to those expectations, suggesting a different
-prevailing interactive effect between nonsynonymous mutations within domains.
-
-The strength of selection on missense mutations within and outside of domains
-is observed to differ, leading to an excess of rare missense mutations within
-conserved domains (Table~\ref{tab:tajimasD}). LD is known to be sensitive to
-allele frequencies, with rare mutations showing large positive signed LD
+other (Figure~\ref{fig:LDwithin}).
+
+Selection is stronger against missense mutations within domains than outside
+domains, leading to an excess of rare missense mutations within conserved
+domains (Table~\ref{tab:tajimasD}). LD is known to be sensitive to allele
+frequencies, with rare mutations showing large positive signed LD
 \citep{Good2022-ot}. To test whether the signal of increased LD between
 missense mutations within domains is driven by rare variants, I considered
-subsets of pairs of mutations based on their derived allele frequences
-(Figures~\ref{fig:data}D~and~\ref{fig:domainsAFR}--\ref{fig:domainsEAS}). Rare
-and uncommon variants show large average LD for each class of mutations, but
-common variants recapitulate the opposing patterns of LD that is seen when
-averaging over pairs at all frequencies, as unconditioned statistics in the
-form of \(\sigma_d^1\) and \(\sigma_d^2\) are dominated by common variants.
+subsets of pairs of mutations based on their derived allele frequencies
+(Figures~\ref{fig:data}D--F~and~\ref{fig:domainsAFR}--\ref{fig:domainsEAS}).
+Rare and uncommon variants show large average LD for each class of mutations.
+However, common variants recapitulate the opposing patterns of LD that is seen
+when averaging over pairs at all frequencies, and the SFS for common missense
+and synonymous variants are similar (Figure~\ref{fig:sfs_proportions}), so that
+this signal is unlikely to be driven by subtle differences in allele
+frequencies between classes of mutations.
 
@@
 The large positive signal of LD for missense mutations within the same domain
-does not extend to pairs of missense mutations that are in different domains.
-Pairs of missense and synonymous mutations show nearly equal levels of LD close
+does not extend to pairs of missense mutations that span different domains.
+Missense and synonymous mutations show nearly equal levels of LD close
 to zero across domains, with missense mutations slightly more negative than
@@
 driving large LD in domains is therefore likely domain-specific. However, the
-average distance between pairs of mutations within domains is much smaller than
-between domains, and this observation may be primarily driven by the higher
-recombination distances between pairs of mutations across distinct domains.
-
-Pairs of mutations that both fall outside of annotated domains have the
-opposite pattern of signed LD to pairs of mutations falling within the same
+average distance between mutations within domains is much smaller than
+between domains, so this observation may be primarily driven by the higher
+recombination distances between mutations across distinct domains.
+
+Mutations that fall outside of annotated domains have the
+opposite pattern of signed LD to mutations within the same
 domain. For pairs of mutations outside of domains but with distances matched to
-those within domains, pairs of synonymous mutations have larger positive LD
+those within domains, synonymous mutations have larger positive LD
 than missense mutations. More distant pairs of mutations outside of domains,
@@
 (Figures~\ref{fig:LDgene}--\ref{fig:LDoutside}). Both synonymous and missense
-mutation pairs at distances greater than a few hundred base pairs have average
-LD that fluctuates around zero. However, for mutations outside domains,
+mutation pairs at distances greater than a few hundred bases have average
+LD fluctuating around zero. However, for mutations outside domains,
 synonymous variants separated by short distances have large positive LD, while
@@
 
-Previous theoretical and simulation studies have shown that interference and
-interactions between selected mutations reduce the efficacy of selection at
-linked loci, impacting substitution rates, the deleterious mutation load, and
-dynamics of segregating mutations
-\citep{Hill1968-vu,Birky1988-jm,Barton1995-mj,McVean2000-ox}. Interference and
-synergistic epistasis between moderately deleterious mutations are both
-expected to cause negative LD between selected mutations, which can be readily
-tested using population genetic data
-\citep{Sohail2017-zq,Sandler2021-of,Garcia2021-zn}. Here, I used a closed
-numerical approach to generate expectations for LD under a wide range of
-selective scenarios, and then compared patterns of LD in human populations
-between classes of coding mutations using unbiased estimators for LD from
-unphased genotypes.
-
-By taking broad genome- and gene-wide surveys of LD across functional classes
-of mutations, heterogeneous patterns of interactive effects that occur within
-genes can be missed. From gene-wide averages, LD between pairs of missense
-mutations does not appear to be different from pairs of synonymous variants,
-although LD between loss-of-function variants is more negative. Hill-Robertson
+%Previous theoretical and simulation studies have shown that interference and
+%interactions between selected mutations reduce the efficacy of selection at
+%linked loci, impacting substitution rates, the deleterious mutation load, and
+%dynamics of segregating mutations
+%\citep{Hill1968-vu,Birky1988-jm,Barton1995-mj,McVean2000-ox}. Interference and
+%synergistic epistasis between moderately deleterious mutations are both
+%expected to cause negative LD between selected mutations, which can be readily
+%tested using population genetic data
+%\citep{Sohail2017-zq,Sandler2021-of,Garcia2021-zn}. Here, I used a closed
+%numerical approach to generate expectations for LD under a wide range of
+%selective scenarios, and then compared patterns of LD in human populations
+%between classes of coding mutations using unbiased estimators for LD from
+%unphased genotypes.
+
+Broad genome- and gene-wide surveys of LD across functional classes of
+mutations miss heterogeneous patterns of interactive effects occurring within
+genes.
+From gene-wide averages, LD between missense
+mutations does not appear to differ from synonymous variants,
+while LD between loss-of-function variants is more negative. Hill-Robertson
 effects are expected to be strongest for slightly to moderately deleterious
@@
 expected to interfere with one another \citep{McVean2000-ox}. However,
-loss-of-function mutations do not typically fall within this regime, and
 inference of the distribution of fitness effects (DFE) for new loss-of-function
 variants shows that a large majority are strongly deleterious (Supporting
-Information).
+Information), so most will not strongly interfere with each other.
 
@@
 which are separated by larger distances on average than neighboring missense
-and synonymous mutations.
+and synonymous mutations (Figure~\ref{fig:mutDistances}).
 
 In addition to LD varying by distance, LD can also vary due to differences in
-allele frequencies among classes of mutations (as selection drives mutations to
-higher or lower frequencies). Matching distances between pairs as well as
-allele frequecies between classes of mutations helps to reduce these concerns.
-It is possible that recombination rates can vary between annotated regions,
-resulting in differing patterns of background selection, which can affect
-allele frequencies and LD (e.g., Figures~\ref{fig:bgs1}--\ref{fig:bgs4}). I did
-not condition on local recombination rates or inferred levels of background
-selection here, which is left to future work.
+allele frequencies among classes of mutations. Matching both distances between pairs
+and allele frequencies between classes of mutations reduces these
+concerns. It is further possible that recombination rates can vary between annotated
+regions, resulting in differing patterns of background selection, which can
+affect both allele frequencies and LD (e.g.,
+Figures~\ref{fig:bgs1}--\ref{fig:bgs4}). I did not condition on local
+recombination rates or inferred levels of background selection here.
 
@@
 \citep{Garcia2021-zn}. The similarity of their gene-wide LD observed in this
-study (Figure~\ref{fig:data}) might suggest that interference or interactions
-between missense mutations are minimal, or at least no stronger than those
-between synonymous mutations. However, averaging over all observed pairs of
+study might suggest that interference or interactions
+between missense mutations are minimal.
+However, averaging over all observed pairs of
 mutations within a gene masks element-specific interactive effects that drive
 LD in opposite directions. Nonsynonymous mutations found within conserved
-protein domains, identified as conserved subunits of a gene with some
-structural or functional role \citep{Stanek2020-pa}, are more strongly selected
+protein domains are more strongly selected
 against on average, but also have increased signed LD over synonymous mutations
@@
 
-There are a number of possible interaction scenarios that can result in
-positive signed LD between tightly linked loci. One explanation is a prevalence
+Different interaction scenarios that can result in
+positive signed LD between tightly linked loci. One possibility is a prevalence
 of pairs of compensatory mutations that are tolerated to co-segregate at high
@@
 
-We do not see an increase in LD between mutations that are found in different
-annotated domains within the same gene, which are on average at much larger
-recombination distances. Nor do we find increased LD outside of domains.
-Rather, for pairs of mutations outside of domains but matched to the same
-distance as those within the same domain, missense variants have considerably
-lower LD than synonymous variants. This difference between synonymous and
-missense pairs of variants largely disappears for SNPs separated by more than a
+There is no increase in LD between pairs of missense mutations found in
+different annotated domains or outside of domains.
+Rather, those missense variants have considerably
+lower LD than synonymous variants, and this difference between synonymous and
+missense variants disappears for SNPs separated by more than a
 few hundred base pairs (Figure~\ref{fig:data}E, F). This suggests that
@@
 
-Taken together, selection on segregating variants within protein-coding genes
-is nonuniform, with both the overall strength of selection and interactions
-between variants differing between annotated elements. Missense and
-loss-of-function mutations within conserved domains are subject to stronger
-selection, skewing the SFS to lower frequencies. Typical approaches for
-inferring the distribution of fitness effects from population genomic data
-average over these differences by considering all nonsynonymous mutations
-together \citep{Boyko2008-zk,Kim2017-xo}. If patterns of selection differ by
-both mutational class and location or if loss-of-function variants are in fact
-recessive, this can impact inferences of the DFE (Supporting Information). It
-would be straightforward to adapt DFE-inference methods to infer more detailed
+Taken together, selection and interactions in protein-coding genes are
+nonuniform, depending on mutation type and location within the gene. Typical
+approaches for inferring the distribution of fitness effects from population
+genomic data average over such differences by aggregating all observed
+nonsynonymous mutations \citep{Boyko2008-zk,Kim2017-xo}. It would be
+straightforward to adapt DFE-inference methods to infer more detailed
 representation of the heterogeneous effects of new mutations within genes by
 partitioning by missense and loss-of-function classes as well as by annotated
-domains. Similarly, the different modes of interaction between mutations across
-annotated domains mean that our standard models and simulation approaches may
-be too simple to capture the evolutionary trajectories and patterns of
-diversity that differ at finer scales.
-
-\subsection{Causes of positive LD between synonymous mutations}
-
-In a non-structured randomly mating population, neutral mutations are expected to have
-average signed LD of zero, but across all populations analyzed here, LD between
-synonymous mutations is positive. While selection on some subset of synonymous
-variants is possible, it should be much weaker on average than between missense
-mutations, and any interference between selected synonymous variants should
-lead to negative LD.
-
-Spatial population structure may be responsible for the increase of LD observed
-between synonymous variants. Neutral evolution in structured populations alone
-cannot create this effect. Under standard population genetic assumptions of
-constant mutation and recombination rates and independent mutation events,
-migration, admixture, and spatial structure do not result in nonzero \(\E[D]\).
-Rather, for a set of mutations with \emph{given} differences in allele
-frequencies between source populations, \(D\) can be nonzero in an admixed
-population between those particular mutations even when \(D\) is zero in the
-source populations \citep{Cavalli-Sforza1971-jv}. This result relies on
-conditioning on directional differences in allele frequencies between
-populations. Without conditioning, the expected difference in mean allele
-frequency between source populations is zero (\(\E[(p_1-p_2)(q_1-q_2)] = 0\)),
-as differences are equally likely to be positive or negative, and thus
-resulting \(\E[D]\) is still zero when taking genome-wide averages.
-
-Instead, \citet{Sohail2017-zq} and \citet{Sandler2021-of} used forward
-simulations under multi-population or explicit spatial models and a
-multiplicative fitness function to explore the joint effects of structure,
-assortative mating, and interference, finding that together these can lead to
-nonzero LD. For example, \citet{Sandler2021-of} found positive LD among neutral
-mutations in a simulation with admixture and linked selected mutations, showing
-that the combined effects of spatial structure and background selection can
-lead to positive average LD between neutral mutations. Positive LD between
-synonymous mutations is only observed at short distances in human data, which
-could be used to constrain such models that predict positive LD over varying
-distances.
-
-Nonrandom mutational processes provide an alternative explanation for positive
-LD between neutral variants separated by short distances, in which clusters of
-mutations occur simultaneously in the same mutational event. Such
-multinucleotide mutations have been shown to affect patterns of LD over short
-distances, on the order of 10s to 100s of base pairs \citep{Harris2014-zg}, and
-clustered mutational events may be common in humans \citep{Besenbacher2016-ac}.
-Indeed, a simple exponential model in which the fraction of mutations causing a
-multinucleotide mutation event decays with distance fits the observed patterns
-of \(\sigma_d^1\) between synonymous variants (Supporting Information,
-Figure~\ref{fig:mslmnms}), with the best-fit model requiring only a small
-fraction of mutations to cause multinucleotide mutation events. We may
-therefore treat positive LD as the baseline expectation for tightly linked
-variants primarily due to clustered mutations, so that subsequent selective
-processes and interactions cause LD to deviate from that expectation, as seen
-for missense mutations across different sub-elements of genes. It may also be
-more appropriate to compare the negative LD observed between linked
-loss-of-function variants to that positive baseline expectation instead of
-zero, which would imply that they are more recessive or have stronger
-synergistic epistasis than from inferences assuming a neutral expectation of
-zero. Additional analyses and extensive simulations will be required to tease
-apart the effects of population structure, selective interference at linked
-sites, and clustered mutational events on patterns of LD.
-
-\subsection{Challenges to distinguishing highly parameterized models of selective interactions}
+domains. Additionally, standard models and simulation approaches may be too
+simple to capture evolutionary trajectories and patterns of diversity that
+differ due to region-specific interactions.
+
+\subsection{Positive LD between synonymous mutations}
+
+In a non-structured randomly mating population, neutral mutations are expected
+to have average signed LD of zero, but across all populations analyzed here, LD
+between synonymous mutations is positive. While selection on some subset of
+synonymous variants is possible, it is likely weaker on average than
+between missense mutations, and any interference between selected synonymous
+variants should lead to negative LD.  Spatial population structure may be
+responsible for the increase of LD observed between synonymous variants. While
+neutral evolution alone in structured populations cannot cause positive LD,
+\citet{Sohail2017-zq} and \citet{Sandler2021-of} used forward simulations to
+show that population structure combined with selection at linked sites can
+induce positive LD between neutral mutations.
+
+Nonrandom mutational processes, in which clusters of mutations occur
+simultaneously in the same mutational event, can also lead to positive LD
+between neutral variants separated by short distances. Such multinucleotide
+mutations have been shown to affect patterns of LD on the order of 10s to 100s
+of base pairs \citep{Harris2014-zg}, and clustered mutational events may be
+common in humans \citep{Besenbacher2016-ac}. Indeed, a simple exponential model
+in which the fraction of mutations causing a multinucleotide mutation event
+decays with distance fits the observed patterns of \(\sigma_d^1\) between
+synonymous variants (Supporting Information, Figure~\ref{fig:mslmnms}), with
+the best-fit model requiring only a small fraction of mutations to involve
+multinucleotide mutation events.
+
+We may therefore treat positive LD as the baseline expectation for tightly
+linked variants due to either spatial structure or clustered mutations, so that
+subsequent selective processes and interactions cause LD to deviate from that
+expectation. It may then be more appropriate to compare the negative LD
+observed between linked loss-of-function variants to that positive baseline
+expectation instead of zero, which would imply that they are more recessive or
+have stronger synergistic epistasis than from inferences assuming a neutral
+expectation of zero. Additional analyses and simulations will be
+required to tease apart the effects of population structure, selective
+interference at linked sites, and clustered mutational events on patterns of
+LD.
+
+\subsection{Challenges to distinguishing modes of selective interactions from LD}
 
 When partitioning measurements of LD by mutation classes or regions within
-genes, the decreasing number of pairwise comparisons leads to large estimated
+genes, the decreasing number of pairwise comparisons leads to large
 measurement noise. Within each population, confidence intervals of observed
@@
 testing of significance difficult due to shared variation, as they cannot be
-treated as independent measurements. Extensive forward simulations will likely
+treated as independent measurements. Detailed simulations will likely
 be needed to more thoroughly assess significance.
 
-From a modeling perspective, in exploring or inferring multi-locus selective
-models, the number of plausible selection scenarios becomes quite large as we
-relax the strict assumptions of additivity and multiplicative interactions.
-The inclusion of dominance effects, epistasis, or other interactions leads to a
-rapid increase in the number of parameters to consider. This makes performing
-forward simulations that span the range of all such selective interaction
-scenarios burdensome.
-
-Instead, closed, numerical approaches to compute expectations for two-locus
-statistics under a wide range of selection models, as presented here and in
-\citet{Friedlander2022-bs}, allow us to explore this highly parameterized space
-of models far more efficiently, and it opens the possibility for performing
-likelihood-based inference using signed LD or other two-locus summaries of the
-data. For example, inferring the joint distribution of dominance and selection
-is underpowered using the SFS alone, but because signed LD is sensitive to the
-levels of dominance (Figure~\ref{fig:dominance}), inferring the DFE with
-dominance may be feasible using the joint distribution of allele frequencies
-and LD. The results presented in this paper likely do not cover the space of
-all possible two-locus models, and other unexplored models may result in
+From a modeling perspective, the space of plausible selection scenarios becomes
+large as we relax the strict assumptions of additivity and multiplicative
+interactions. This makes performing forward simulations that span the range of
+all such selective interaction scenarios burdensome. Instead, closed numerical
+approaches allow us to efficiently explore this highly parameterized space of
+models and to perform likelihood-based inference using signed LD or other
+two-locus summaries of the data. For example, inferring the joint distribution
+of dominance and selection is underpowered using the SFS alone, but because
+signed LD is sensitive to the levels of dominance (Figure~\ref{fig:dominance}),
+inferring the DFE with dominance may be feasible using the joint distribution
+of allele frequencies and LD. The results presented here do not cover the space
+of all possible two-locus models, and other unexplored models may result in
 similar patterns of signed LD.  Comparisons to empirical data should therefore
-be treated with caution, and additional careful modeling will allow a more firm
-determination of the interactive effects causing observed patterns of
-variation.
+be treated with caution.
 
@@
 \(\sigma_d^1\), is likely insufficient to confidently discriminate modes of
-selective interactions between linked mutations. Both dominance and epistasis
-can result in strongly negative LD, although the expected decay of LD under
-those scenarios can differ. Similarly, multiple interaction models can lead to
-positive signed LD. Among all interaction models, the extent of LD and rate of
-its decay also depends on the underlying distribution of selection coefficients
-among a class of mutations, which are unknown for a given pair of mutations, so
-that we must integrate over a distribution of fitness effects. This DFE,
-however, will have been inferred under a simple set of assumptions, such as
-additivity and interchangeability between sites within a gene, potentially
-biasing any inference using previously inferred DFEs to learn about patterns of
-interactions. Again, this underlines the need to jointly infer strengths and
-interactions of selected variants and to consider patterns of variation at
-finer genomic scales.
-
-Finally, exploring additional summaries of the two-locus sampling distribution
-should increase power to distinguish between interaction models. The
-numerical methods developed here provide the full two-locus sampling
-distribution, and expectations for a large family of informative two-locus
-statistics can be computed directly from this distribution. These expectations
-can then be compared directly to empirical observations, which can be taken
-from either phased or unphased data \citep{Ragsdale2020-nz}. Additional work
-performing such an analysis is therefore warranted, which should provide a path
-forward for distinguishing between modes of selective interactions.
+selective interactions that produce similar LD patterns. Among all interaction
+models, the extent of LD and rate of its decay also depends on the underlying
+distribution of selection coefficients among a class of mutations, which are
+unknown for a given pair of mutations, so that we must integrate over a
+distribution of fitness effects. This DFE, however, will have been inferred
+under a simple set of assumptions, such as additivity and interchangeability
+between sites within a gene, potentially biasing any inference using previously
+inferred DFEs to learn about patterns of interactions. Again, this underlines
+the need to jointly infer strengths and interactions of selected variants and
+to consider patterns of variation at finer genomic scales.
+
+Finally, expectations for a large family of informative two-locus statistics
+can be computed directly from the full two-locus sampling distribution, which
+can be compared to empirical observations from either phased or unphased data
+\citep{Ragsdale2020-nz}. Exploring additional patterns of correlations between
+mutations should uncover overlooked statistics that will improve our ability to
+distinguish between modes of selective interactions.
 
@@
 
-Many well-known properties of two-locus dynamics and equilibrium LD come from
+Many well-known properties of linkage disequilibrium come from
 early work on the multi-locus diffusion approximation \citep{Kimura1955-qe,
@@
 covariance measure of LD is defined as \(D = f_{AB} - f_{A}f_{B}\), where
-\(f_{AB}\) is the haplotype frequency of double-derived types carrying both
+\(f_{AB}\) is the haplotype frequency of types carrying both
 \(A\) and \(B\), and \(f_{A}\) and \(f_{B}\) are the marginal frequencies of
-the derived alleles at each locus. This covariance decays due to both drift and
+those alleles at each locus. This covariance decays due to both drift and
 recombination at a rate proportional to the inverse of the effective population
@@
 While \(\E[D] = 0\), the variance of \(D\) is non-zero, and \citet{Ohta1971-yd}
-presented their groundbreaking result that the variance of \(D\) under
+found that the variance of \(D\) under
 neutrality and steady-state demography, normalized by the joint heterozygosity
@@
 recursions were developed to compute the two-locus sampling distribution under
-neutrality, that is, the probability of observing given counts of two-locus
-haplotypes in a sample of size \(n\) \citep{Golding1984-pu}. This approach
-would continue to be developed and later form the foundation for the inference
-of local recombination rates from population genetic data
-\citep{Hudson2001-sg,McVean2004-gj}. More recently, \citet{Song2007-qk}
-computed \(\E[r^2]\) using a diffusion approximation approach, although their
-solution involves the summation of infinitely many terms and is restricted to
-neutrality and steady-state demography.
-
-To include selection, however, there have been relatively few advances beyond
+neutrality \citep{Golding1984-pu}, and this approach later formed the
+foundation for the inference of local recombination rates from population
+genetic data \citep{Hudson2001-sg,McVean2004-gj}. More recently,
+\citet{Song2007-qk} computed \(\E[r^2]\) using a diffusion approximation
+approach, although their solution involves the summation of infinitely many
+terms and is restricted to neutrality and steady-state demography.
+
+To include selection, there have been relatively few advances beyond
 the Monte Carlo simulation approach taken by \citet{Hill1966-gv}, albeit now
@@
 
-Numerical methods inhabit the space in between expensive discrete simulations
-and limited analytic solutions, providing a more efficient and practical method
-to compute expectations of two-locus diversity measures under a wider range of
+Numerical methods inhabit the space between expensive discrete simulations and
+limited analytic solutions, providing a more efficient and practical method to
+compute expectations of two-locus diversity measures under a wider range of
 parameters and demographic scenarios. \citet{Ragsdale2017-gg} used a finite
-differencing approach to numerically solve the two-locus diffusion equation
-with additive selection at either locus, although they focused on the
-applicability of two-locus statistics to demographic inference. More recently,
-\citet{Ragsdale2019-nt} extended the \citet{Hill1968-vu} system for \(\E[D^2]\)
-to compute arbitrary moments of the distribution of \(D\) for any number of
-populations connected by migration and admixture. They also showed that such a
-moments-construction can be used to solve for the two-locus sampling
-distribution within a single population, though it requires a moment-closure
-approximation for nonzero recombination and selection. Below, I extend this
-approach to model arbitrary diploid selection, which encompasses dominance,
-epistasis, and other forms of selective interactions between two loci. In a
-concurrent study to this paper, \citet{Friedlander2022-bs} developed numerical
-solutions to the same moments system, which they used to describe selected
-haplotype trajectories and the distortion of neutral diversity at loci variably
-linked to beneficial alleles that sweep to high frequencies under
-non-equilibrium demography.
+differences approach to numerically solve the two-locus diffusion equation with
+additive selection at either locus, and \citet{Ragsdale2019-nt} more recently
+extended the \citet{Hill1968-vu} system for \(\E[D^2]\) to compute arbitrary
+moments of the distribution of \(D\) for any number of populations connected by
+gene flow. They also showed that such a moment-construction can be used to
+solve for the two-locus sampling distribution within a single population,
+though it requires a moment-closure approximation for nonzero recombination and
+selection. Below, I extend this approach to model arbitrary diploid selection,
+which encompasses dominance, epistasis, and other forms of selective
+interactions between two loci. In a concurrent study to this paper,
+\citet{Friedlander2022-bs} developed numerical solutions to the same moment
+system, which they used to describe selected haplotype trajectories and the
+distortion of neutral diversity at loci variably linked to beneficial alleles
+that sweep to high frequencies under non-equilibrium demography.
 
@@
 site frequency spectrum (SFS) of a given sample size
-(Figure~\ref{fig:TwoLocusFS}). Instead of describing the density or number of
-mutations with a given allele count out of \(n\) samples, the two-locus
+(Figure~\ref{fig:TwoLocusFS}). The two-locus
 distribution \(\Psi_n\) stores the density or number of pairs of loci with
-observed haplotype counts, so that \(\Psi_n(i, j, k)\) is the number of pairs
+observed haplotype counts in a sample of size \(n\),
+so that \(\Psi_n(i, j, k)\) is the number of pairs
 for which we observe \(i\) copies of the \(AB\) haplotype, \(j\) of type
@@
 
-Under neutrality, a number of approaches exist to compute \(\Psi_n\), including
-the recursion due to \citet{Golding1984-pu} and \citet{Ethier1990-tb}, or more
-recent numerical approaches to the two-locus coalescent \citep{Kamm2016-ag} or
-diffusion approximation \citep{Ragsdale2017-gg}. Selection is most easily
-included using the forward-in-time diffusion equation
+Under neutrality, many approaches exist to compute \(\Psi_n\), including the
+recursion due to \citet{Golding1984-pu} and \citet{Ethier1990-tb} and recent
+numerical approaches \citep{Kamm2016-ag, Ragsdale2017-gg}. Selection is most
+easily included using the forward-in-time diffusion equation
 \citep{Kimura1955-qe,Hill1966-gv}, where a standard approach is to first solve
@@
 configurations in the full population, and then integrate \(\psi\) against the
-multinomial sampling function to obtain \(\Psi_n\).
-
-Alternatively, \citet{Ragsdale2019-nt} showed that there exists a system of
-ordinary differential equations directly on the entries of \(\Psi_n\). I
-briefly summarize this general approach below, but refer readers to that paper
-for detailed derivations of the drift, recombination, and mutation terms and
-the moment-closure approximation. Instead, here I focus on generalizing the
+multinomial sampling function to obtain \(\Psi_n\). Alternatively,
+\citet{Ragsdale2019-nt} showed that there exists a system of ordinary
+differential equations directly on the entries of \(\Psi_n\). I briefly
+summarize this general approach below, but refer readers to that paper for
+detailed derivations of the drift, recombination, and mutation terms and the
+moment-closure approximation. Instead, here I focus on generalizing the
 selection operator to include epistasis, dominance, and other forms of
@@
 + \mathcal{U}_{u}\Psi_n^t
-+ \mathcal{S}_{s_A, s_B, \ldots, h_A, h_B, \ldots}\Psi_{n+2}^t.
++ \mathcal{S}_{s,h}\Psi_{n+2}^t.
 \end{equation}
-Here, \(\mathcal{D}_{N(t)}\) is a sparse linear operator accounting for drift
+\(\mathcal{D}_{N(t)}\) is a sparse linear operator accounting for drift
 with population size \(N(t)\), \(\mathcal{R}\) accounts for recombination
-with per-generation recombination probability \(r\) between the two loci,
+with per-generation recombination probability \(r\) between loci,
 \(\mathcal{U}\) accounts for mutation, either under an infinite sites or
@@
 
-In the absence of selection and for fully linked loci (i.e., \(\rho=0\)), the
-system is closed and can be solved exactly. However, for nonzero recombination
-or selection, the entries of \(\Psi_n\) rely on the slightly larger sampling
+In the absence of selection and for fully linked loci (\(\rho=0\)), the system
+closes and can be solved exactly. However, for nonzero recombination or
+selection, the entries of \(\Psi_n\) rely on the slightly larger sampling
 distributions with sample sizes \(n+1\) (for recombination and additive
@@
 recombine with that chosen lineage, thus requiring \(\Psi_{n+1}^t\) to find
-\(\Psi_n^{t+1}\). Selection events similarly require extra lineages from the full
-population, which replace a chosen lineage that fails to reproduce with
+\(\Psi_n^{t+1}\). Selection events similarly require extra lineages from the
+full population, which replace a chosen lineage that fails to reproduce with
 probability proportional to its relative fitness.
 
-This requirement of extra lineages for nonzero recombination and selection
-means that the system in \eqref{eq:system} is not closed, so that we need a
-moment-closure approximation to solve for \(\Psi_n\). As in \citet{Ragsdale2019-nt}, a
-jackknife approximation is used to estimate \(\Psi_{n+l}\), for \(l=1\) or \(2\),
-from \(\Psi_n\) \citep[following the single-locus closure introduced in][]{Jouganous2017-pq}, so that \(\hat{\Psi}_{n+l}(i, j, k) = \mathcal{J}_{n, l}\Psi_n\), although other accurate closure approximations are possible
+This requirement of extra lineages for recombination and selection means that
+the system in \eqref{eq:system} is not closed, so that we need a moment-closure
+approximation to solve for \(\Psi_n\). As in \citet{Ragsdale2019-nt}, a
+jackknife approximation is used to estimate \(\Psi_{n+l}\), for \(l=1\) or
+\(2\), from \(\Psi_n\) \citep[following the single-locus closure introduced
+in][]{Jouganous2017-pq}, so that \(\hat{\Psi}_{n+l}(i, j, k) = \mathcal{J}_{n,
+l}\Psi_n\), although other accurate closure approximations are possible
 \citep{Friedlander2022-bs}. This emits a closed approximate system,
@@
 model for epistasis can be written as \[s_{AB} = (s_{A} + s_{B}) (1 +
-\epsilon),\] so that \(\epsilon > 0\) implies synergistic epistasis, while
-\(\epsilon < 0\) implies antagonistic epistasis.
+\epsilon),\] so that \(\epsilon > 0\) gives synergistic epistasis and
+\(\epsilon < 0\) gives antagonistic epistasis.
 
@@
 \(ab/ab\) genotype has relative fitness 1, this gives nine possible unique
-selection coefficients in the most general two-locus selection model. Note that
-\(AB/ab\) and \(Ab/aB\) genotypes need not have the same selection coefficient,
-which allows for simulation under a gene-based model of dominance (Table
-\ref{tab:selmodels}).
+selection coefficients in the most general two-locus selection model. Doubly
+heterozygous \(AB/ab\) and \(Ab/aB\) genotypes need not have the same selection
+coefficient (Table \ref{tab:selmodels}).
 
@@
 assume at most a single selection event occurs in any given generation. This
-means that to find \(\Psi_n^{t+1}\) under a general two-locus selection model,
-which encompasses dominance within either locus, gene-based dominance, or a
-combination of dominance and epistatic effects, we need \(\Psi_{n+2}^t\).
-Again, a full derivation and expressions for the general selection operator are
-given in the Supporting Information.
+means that to find \(\Psi_n^{t+1}\) under a general two-locus selection model
+we need \(\Psi_{n+2}^t\). Again, a full derivation and expressions for the
+general selection operator are given in the Supporting Information.
 
@@
 
-I considered four variable population size histories, two simple toy models and
+I consider four variable population size histories, two simple toy models and
 two inferred from human populations in African and Europe using \texttt{Relate}
-\citep{Speidel2019-nj}. For each size history scenario, I tracked the evolution
+\citep{Speidel2019-nj}. For each size history scenario, I track the evolution
 of \(\Psi_n(t)\) for varying selection models, plotting the trajectories of
@@
 \ref{fig:relate_dom}, and \ref{fig:toy_sd2}--\ref{fig:relate_dom_sd2}). The
-selection strength at both loci was fixed at either \(\gamma=-1\) or \(-10\)
+selection strength at both loci is fixed at either \(\gamma=-1\) or \(-10\)
 for the models with epistasis, or \(\gamma=-2\) for models with dominance, and
-recombination was set to zero.
-
-The simple size change models both had ancestral \(N_e=10,000\), with one a
+recombination is set to zero.
+
+The simple size change models both have ancestral \(N_e=10,000\), with one a
 \(3\)-fold population expansion that occurs 3,000 generations ago, and the
@@
 its initial size \(1,000\) generations ago. The size histories for YRI and CEU
-were inferred using \texttt{Relate} \citep{Speidel2019-nj} applied to the phase
+are inferred using \texttt{Relate} \citep{Speidel2019-nj} applied to the phase
 3 haplotype-phased autosomal data from
@@
 using \texttt{Relate} for population sizes in the very recent past (\(<3,000\)
-years, or \(\approx 100\) generations) diverged, so I truncated the history
-over this time period and assumed a constant size from the most recent
+years, or \(\approx 100\) generations) diverge, so I truncate the history
+over this time period and assume a constant size from the most recent
 non-diverged bin.
@@
 \citet{1000_Genomes_Project_Consortium2015-zq} (Thousand Genomes) data release,
-I subset the genotype VCFs to autosomal variants that were annotated as either
+I subset the genotype VCFs to autosomal variants that are annotated as either
 synonymous or nonsynonymous, including both missense mutations and more
@@
 Asian continental groups (five populations each, Table \ref{tab:1kgpops}).
-Signed LD is sensitive to ancestral state misidentification, so I only kept
+Signed LD is sensitive to ancestral state misidentification, so I only keep
 sites for which ancestral alleles were estimated with high confidence in both
@@
 In addition to ancestral state misidentification, measured LD is sensitive to
-phasing error, so I computed LD statistics using unphased genotypes following
+phasing error, so I compute LD statistics using unphased genotypes following
 \citet{Ragsdale2020-nz}. This approach provides unbiased estimates for pairwise
-LD, under the assumption that individuals are not inbred. I considered pairs of
+LD, under the assumption that individuals are not inbred. I consider pairs of
 mutations within the same mutation class (synonymous, missense, and
 loss-of-function) either within the same gene and inside or outside of
-annotated domains within the same protein-coding genes. I used a dataset of
+annotated domains within the same protein-coding genes. I use a dataset of
 annotated protein domains mapped to the hg19 human reference build compiled by
-\citet{Stanek2020-pa} to determine if a given mutation falls within a domain or
-not.
+\citet{Stanek2020-pa} to determine if a given mutation falls within our outside
+a conserved domain.
 
@@
 
-\section{Acknowledgements}\label{acknowledgements}
+\section{Acknowledgments}\label{acknowledgements}
 

### 2022-06-03 manuscript/supplement.tex
updates for minor revisions

@@
 size of \(N=1000\), a sequence length of 1Mb, with mutation and recombination
-rates both \(u=r=10^{-8}\). I considered four scenarios for selected mutaions:
+rates both \(u=r=10^{-8}\). I considered four scenarios for selected mutations:
 additive selection \(h=0.5\) with \(s=-0.0005\) and \(s=-0.01\) (\(\gamma=-1\)
@@
 selection pressures are observed to be greater within annotated conserved
-domains, with stronger purifying selection driving allele frqeuencies to
+domains, with stronger purifying selection driving allele frequencies to
 relatively lower levels among missense and loss-of-function variants. Since
@@
 \textbf{Tamija's $D$ for classes of coding mutations.} Here, we partitioned by
-synonymous, missense, and nonsense mutationsi, and considered all mutations gene-wide,
+synonymous, missense, and nonsense mutations, and considered all mutations gene-wide,
 mutations falling within annotated conserved elements,
@@
         pairs of synonymous and pairs of missense variants. This observation is
-        consistest when weighting missense variants to match distances between
+        consistent when weighting missense variants to match distances between
         pairs of synonymous variants and when weighting to match synonymous allele
@@
         has only a small effect on average signed LD, and gene-wide measurements
-        are roughy equal between these two mutation classes.
+        are roughly equal between these two mutation classes.
         Again, measurement noise for loss-of-function mutation pairs is high.
@@
         \label{tab:eassd1}
-        \textbf{Signed LD in EUR-labeled populations matched to distances
+        \textbf{Signed LD in EAS-labeled populations matched to distances
         and allele frequencies of synonymous mutations.}
@@
         E: \texttt{moments} expectations, D: simulated data.
-        Additional comparisons to simulated data are availabe at
+        Additional comparisons to simulated data are available at
         \texttt{https://github.com/apragsdale/two\_locus\_selection}.
@@
         \textbf{Background selection with slightly deleterious mutations.}
-        Using an individual based forward simulator \citep{Thornton2019-qc}, I
-        simulated many replicates of 1Mb regions with high mutation rates of
+        Using an individual-based forward simulator \citep{Thornton2019-qc}, I
+        simulated many replicates of 1Mb regions with a high mutation rate of
         selected mutations (see Section~\ref{sec:bgs} for details). Here,
         many linked additive weakly selected mutations slightly distort the
-        SFS for both selected and neutral mutations. \(\sigma_d^2\) is reduced
-        relative to two-locus expectations within background selection, and
-        Hill-Robertson interference effects are slightly diminished with
-        \(\sigma_d^1\) not as strongly negative as expected.
+        SFS for both (A) selected and (C) neutral mutations.
+        (B) In this scenario of background selection, \(\sigma_d^2\)
+        is reduced relative to two-locus expectations without linked selection.
+        (D) Hill-Robertson interference effects are slightly diminished, with
+        \(\sigma_d^1\) not as strongly negative as expected without additional
+        linked selected mutations.
     }
@@
         In the case of many linked strongly selected additive mutations, neither
-        selected nor neutral statistics deviate dramatically from two-locus expectations.
-        The SFS from the simulated data matches expectations, with a slight excess
-        of selected mutations. LD is also largely unaffected, with \(\sigma_d^2\)
+        selected nor neutral statistics deviate dramatically from expectations
+        for either the SFS (A, C) or two-locus statistics (B, D).
+        The SFS from the simulated data matches single-locus expectations.
+        LD is also largely unaffected, with \(\sigma_d^2\)
         matching expectations and \(\sigma_d^1\) fluctuating near zero as expected.
@@
         The presence of many linked weakly selected recessive mutations can cause
-        measues of diversity to have large increases over single- and two-locus
-        expectations. The SFS for both neutral and selected mutations both show
+        measures of diversity to have large increases over single- and two-locus
+        expectations. The SFS for both selected (A) and neutral (C) mutations both show
         a very large excess of common variants (note the log scale), and \(\sigma_d^2\)
+        decay curves (B)
         are also increased over two-locus expectations for both classes of mutations.
-        Still, \(\sigma_d^1\) remains 0 for neutral mutations and is not as strongly
-        negative for pairs of selected mutations.
+        (D) \(\sigma_d^1\) remains 0 for neutral mutations and is not as strongly
+        negative for pairs of selected mutations due to interference from linked selected
+        mutations.
     }
@@
         deviations from expectations as the case of weakly deleterious recessive
-        mutations. The neutral SFS is reduced, and we observe more selected mutations
-        at higher frequencies than expected. LD is also only weakly distorted from
-        two-locus expectations. Notably, we see positive LD (matching between
-        simulations and expectations) for strongly deleterious recessive mutations,
-        as predicted in \citet{Roze2021-cf}.
+        mutations (Fig.~\ref{fig:bgs3}.
+        (A) We observe more selected mutations at higher frequencies than expected,
+        and (C) the neutral SFS is reduced below single-locus expectations.
+        (B, D) LD is also only weakly distorted from
+        two-locus expectations. Notably, we see positive LD for strongly
+        deleterious recessive mutations (which matches between
+        simulations and expectations),
+        as previously predicted in \citet{Roze2021-cf}.
     }
@@
         while expansions tend to decrease squared LD.
-        Dashed lines show neutral expectations.
+        Black dashed lines show neutral expectations.
+        (A) The simulated size histories. (B-F) Blue solid lines
+        are expected squared LD under the expansion model, and gold
+        dashed lines are squared LD under the bottleneck/recovery model.
+        (B-D) Expectations are computed over all allele frequencies.
+        (E) Expected squared LD between uncommon variants, with $n_A$
+        and $n_B \leq 4$, in a sample size of 50 haploid copies.
+        (F) Expected LD squared LD between common variants
+        $p_A, p_B \gtrsim 10\%$.
     }
@@
         large positive signed LD, similar to compensatory mutations or
-        antagonistic epistasis. In general, common variants are more
+        antagonistic epistasis.
+        (A) The simulated size histories.
+        (B-D) Signed LD computed from pairs of mutations
+        across all allele frequencies.
+        (E, F) In general, common variants are more
         stable under population size changes than rare variants.
@@
         \(\sigma_d^2\) and expansions decrease squared LD.
+        (A) The simulated size histories.
+        (B-D) Squared LD computed from pairs of mutations at all allele frequencies.
+        (E, F) Squared LD computed from uncommon ($<10\%$) and common ($\gtrsim 10\%$)
+        mutations.
         Dashed lines show neutral expectations.
@@
         size history with with epistasis.}
+        (A) Size histories inferred using Relate \citep{Speidel2019-nj}.
+        (B-F)
         Similar to the instantaneous size changes in the toy models
         (Figure~\ref{fig:toy}), bottlenecks tend to make signed LD more extreme.
-        However, in the case of strong selection and antagonistic epistasis,
-        the CEU \(\sigma_d^1\) trajector decreases relative to the YRI
-        trajectory. Again, statistics for rare variants fluctuate more rapidly
+        However, in the case of strong selection and antagonistic epistasis (D),
+        the CEU \(\sigma_d^1\) trajectory decreases relative to the YRI
+        trajectory, showing that non-constant demographic history and epistasis can
+        have complicated interactions.
+        (E, F) Again, statistics for rare variants fluctuate more rapidly
         than common variants.
@@
         size history with with epistasis.}
+        (A) Inferred size histories used in simulations.
+        (B-D)
         Size reductions, such as bottlenecks in the recent past of Eurasian populations,
         increase squared LD.
+        (E, F) Uncommon and common variants can show strikingly different behaviors
+        under size change histories, here showing trajectories under the three
+        epistasis models from panels B-D.
         Dashed lines show neutral expectations.
@@
         size history with with dominance.}
+        (A) Inferred size histories used in simulations.
+        (B-D)
         Again, bottlenecks increase squared LD relative to populations that did
         not experience such strong size reductions in their history.
+        (E, F) Uncommon and common variants can show strikingly different behaviors
+        under size change histories, here showing trajectories under the three
+        dominance models from panels B-D.
         Dashed lines show neutral expectations.
@@
         At short distances between SNPs that both fall within sthe same conserved
-        annotated domain, missense mtuations show consistently
+        annotated domain, missense mutations show consistently
         larger signed LD than synonymous variants. This may be caused by
@@
         We observe increased LD between missense mutations within domains,
-        but decrased LD outside of domains at matched distances. This pattern
+        but decreased LD outside of domains at matched distances. This pattern
         holds when conditioning on non-rare variation (\(n_A, n_B \geq 9\),
@@
         We again observe increased LD between missense mutations within domains,
-        but decrased LD outside of domains at matched distances. And again,
+        but decreased LD outside of domains at matched distances. And again,
         this pattern holds when conditioning on non-rare variation
@@
         As with the African and European data,
-        wobserve increased LD between missense mutations within domains,
-        but decrased LD outside of domains at matched distances. And again,
+        we observe increased LD between missense mutations within domains,
+        but decreased LD outside of domains at matched distances. And again,
         this pattern holds when conditioning on non-rare variation
@@
     \label{fig:between_domains}
+\end{figure}
+
+\begin{figure}[ht!]
+    \centering
+    \includegraphics[width=\textwidth]{../figures/SFS_proportions}
+    \caption{
+        \textbf{SFS for common missense and synonymous variants.}
+        The full SFS (considering variants at all observed allele frequencies) is
+        skewed toward low frequencies for missense variants compared to synonymous
+        variants (shown in light colors). Tajima's $D$ is considerably more negative
+        for missense variants than synonymous variants for missense variants, whether
+        conditioning on mutations within annotated elements or outside
+        (Table~\ref{tab:tajimasD}). However, the SFS for common variants (with
+        observed MAF $\geq 10\%$) are similar between missense and synonymous variants
+        for each partition of the data. Thus, the excess of positive signed LD for
+        common variants within annotated domains is likely not to be driven by
+        subtle differences in allele frequencies between missense and synonymous
+        classes.
+    }
+    \label{fig:sfs_proportions}
 \end{figure}

### 2022-06-03 figures/data_coarse.py

@@ -93,7 +93,7 @@ def plot_domain_data(pops, categ, ax, legend=False):
     # within syn, within non, match syn, match non
     xx = np.array([1, 2, 3, 4])
     for i in range(len(pops) - 1):
-        xx = np.concatenate((xx, xx[-1] + 1 + np.array([1, 2, 3, 4])))
+        xx = np.concatenate((xx, xx[-1] + 3 + np.array([1, 2, 3, 4])))
 
     to_plot = [
         [
@@ -154,10 +154,10 @@ def plot_domain_data(pops, categ, ax, legend=False):
     ax.set_xticks(np.mean(xx.reshape(len(pops), 4), axis=1))
     ax.set_xticklabels(pops)
     if legend:
-        ax.legend(fontsize=5, frameon=False, loc="upper left")
+        ax.legend(fontsize=5, loc="upper left")
 
 
-fig = plt.figure(5, figsize=(6.5, 4.5))
+fig = plt.figure(5, figsize=(6.5, 4.0))
 fig.clf()
 
 # plot all stats within genes
@@ -244,7 +244,7 @@ ax1b.errorbar(
 
 ax1.set_ylabel(r"$\sigma_d^1$")
 ax1b.set_ylabel(r"$\sigma_d^1$")
-ax1b.set_xlabel("Populations")
+#ax1b.set_xlabel("Populations")
 
 ax1.set_xticks(
     [
@@ -309,120 +309,121 @@ pops = [pop_afr, pop_eur, pop_eas]
 ax2 = plt.subplot2grid(dims, (5, 0), rowspan=3)
 plot_domain_data(pops, "all", ax2, legend=True)
 ax2.set_ylabel("$\sigma_d^1$")
-ax2.set_xlabel("Populations")
+#ax2.set_xlabel("Populations")
 ax2.set_title("All pairs")
 ax2.set_ylim(-0.2, 1)
 
-#ax3 = plt.subplot2grid(dims, (5, 1), rowspan=3)
-#plot_domain_data(pops, "leq2", ax3)
-#ax3.set_xlabel("Populations")
-#ax3.set_title("$n_A, n_B \leq 2$")
-#
-#ax4 = plt.subplot2grid(dims, (5, 2), rowspan=3)
-#plot_domain_data(pops, "3to8", ax4)
-#ax4.set_xlabel("Populations")
-#ax4.set_title("$3 \leq n_A, n_B \leq 8$")
-
[…]

### 2022-06-03 figures/data_within_vs_between.py

@@ -1,0 +1,172 @@
+# Figure 5: LD within and between genes
+import sys
+import moments.TwoLocus
+import matplotlib.pylab as plt
+import numpy as np
+import pickle
+
+# set font sizes
+import matplotlib
+
+plt.rcParams["legend.title_fontsize"] = "xx-small"
+matplotlib.rc("xtick", labelsize=6)
+matplotlib.rc("ytick", labelsize=6)
+matplotlib.rc("axes", labelsize=7)
+matplotlib.rc("axes", titlesize=7)
+matplotlib.rc("legend", fontsize=6)
+
+from bokeh.palettes import Colorblind
+
+data = {}
+data_dom = {}
+data_btw = {}
+SE = {}
+SE_dom = {}
+SE_btw = {}
+
+populations = {
+    "Africa": ["ESN", "GWD", "LWK", "MSL", "YRI"],
+    "Europe": ["CEU", "FIN", "GBR", "IBS", "TSI"],
+    "East Asia": ["CDX", "CHB", "CHS", "JPT", "KHV"],
+}
+
+for continent, pops in populations.items():
+    for pop in pops:
+        ld_pop = pickle.load(
+            open(f"../analysis/parsed_data_v2/{pop}.unphased.all.bp", "rb")
+        )
+        data[pop] = ld_pop
+        se_pop = pickle.load(
+            open(
+                f"../analysis/parsed_data_v2/{pop}.unphased.all.SEs.bp",
+                "rb",
+            )
+        )
+        SE[pop] = se_pop
+
+        ld_pop = pickle.load(
+            open(f"../analysis/parsed_data_v2/{pop}.unphased.domains.bp", "rb")
+        )
+        data_dom[pop] = ld_pop
+        se_pop = pickle.load(
+            open(
+                f"../analysis/parsed_data_v2/{pop}.unphased.domains.SEs.bp",
+                "rb",
+            )
+        )
+        SE_dom[pop] = se_pop
+
+        ld_pop = pickle.load(
+            open(f"../analysis/parsed_data_v2/{pop}.unphased.between_domains.bp", "rb")
+        )
+        data_btw[pop] = ld_pop
+        se_pop = pickle.load(
+            open(
+                f"../analysis/parsed_data_v2/{pop}.unphased.between_domains.SEs.bp",
+                "rb",
+            )
+        )
+        SE_btw[pop] = se_pop
+
+ms = 2
+lw = 0.5
+markers = ["o", "s"]
[…]

### 2022-06-03 figures/demog_YRI_CEU.dominance.py

@@ -17,7 +17,10 @@ matplotlib.rc("axes", labelsize=7)
 matplotlib.rc("axes", titlesize=7)
 matplotlib.rc("legend", fontsize=6)
 
-from bokeh.palettes import Colorblind
+from bokeh.palettes import Colorblind, Category10
+from bokeh.palettes import YlGnBu, YlOrBr
+c1 = YlGnBu[8]
+c2 = YlOrBr[8]
 
 
 def piecewise_constant(
@@ -287,7 +290,7 @@ tt_CEU, sd1_neu_CEU = get_sd1(nu_CEU, Ts, n, Ne, name="CEU")
 ## plots
 
 lw = 1
-colors = Colorblind[8]
+colors = Category10[10]
 
 fig = plt.figure(4, figsize=(6.5, 3.5))
 fig.clf()
@@ -302,7 +305,7 @@ plot_relate_curve(
     N0,
     line_style="-",
     lw=lw,
-    color=colors[0],
+    color=c1[0],
     label="YRI",
     gen=1,
 )
@@ -312,7 +315,7 @@ plot_relate_curve(
     N1,
     line_style="--",
     lw=lw,
-    color=colors[1],
+    color=c2[0],
     label="CEU",
     gen=1,
 )
@@ -328,22 +331,22 @@ axes[0].set_title("Size histories")
 ## additivity
 print("plotting additivity panel")
 t, v = get_sd1(nu_YRI, Ts, 50, Ne, gamma=-2.0, name="YRI")
-axes[1].plot(1e6 - gen_time * t, v, "-", lw=lw, color=colors[0])
+axes[1].plot(1e6 - gen_time * t, v, "-", lw=lw, color=c1[1])
 t, v = get_sd1(nu_CEU, Ts, 50, Ne, gamma=-2.0, name="CEU")
-axes[1].plot(1e6 - gen_time * t, v, "--", lw=lw, color=colors[1])
+axes[1].plot(1e6 - gen_time * t, v, "--", lw=lw, color=c2[1])
 # axes[1].text(2e5, v[0] + 0.02, rf"$\gamma=-1$", va="center", ha="center", fontsize=7)
 
 # neutral baseline
 axes[1].plot(1e6 - gen_time * tt_YRI, sd1_neu_YRI, "k--", lw=0.5)
 
-axes[1].set_title("Interference")
[…]

### 2022-06-03 figures/demog_YRI_CEU.dominance.sd2.py

@@ -17,7 +17,10 @@ matplotlib.rc("axes", labelsize=7)
 matplotlib.rc("axes", titlesize=7)
 matplotlib.rc("legend", fontsize=6)
 
-from bokeh.palettes import Colorblind
+from bokeh.palettes import Colorblind, Category10
+from bokeh.palettes import YlGnBu, YlOrBr
+c1 = YlGnBu[8]
+c2 = YlOrBr[8]
 
 
 def piecewise_constant(
@@ -287,7 +290,7 @@ tt_CEU, sd2_neu_CEU = get_sd2(nu_CEU, Ts, n, Ne, name="CEU")
 ## plots
 
 lw = 1
-colors = Colorblind[8]
+colors = Category10[10]
 
 fig = plt.figure(4, figsize=(6.5, 3.5))
 fig.clf()
@@ -302,7 +305,7 @@ plot_relate_curve(
     N0,
     line_style="-",
     lw=lw,
-    color=colors[0],
+    color=c1[0],
     label="YRI",
     gen=1,
 )
@@ -312,7 +315,7 @@ plot_relate_curve(
     N1,
     line_style="--",
     lw=lw,
-    color=colors[1],
+    color=c2[0],
     label="CEU",
     gen=1,
 )
@@ -328,39 +331,42 @@ axes[0].set_title("Size histories")
 ## additivity
 print("plotting additivity panel")
 t, v = get_sd2(nu_YRI, Ts, 50, Ne, gamma=-2.0, name="YRI")
-axes[1].plot(1e6 - gen_time * t, v, "-", lw=lw, color=colors[0])
+axes[1].plot(1e6 - gen_time * t, v, "-", lw=lw, color=c1[1])
 t, v = get_sd2(nu_CEU, Ts, 50, Ne, gamma=-2.0, name="CEU")
-axes[1].plot(1e6 - gen_time * t, v, "--", lw=lw, color=colors[1])
+axes[1].plot(1e6 - gen_time * t, v, "--", lw=lw, color=c2[1])
 # axes[1].text(2e5, v[0] + 0.02, rf"$\gamma=-1$", va="center", ha="center", fontsize=7)
 
 # neutral baseline
 axes[1].plot(1e6 - gen_time * tt_YRI, sd2_neu_YRI, "k--", lw=0.5)
[…]

### 2022-06-03 figures/demog_YRI_CEU.py

@@ -17,8 +17,10 @@ matplotlib.rc("axes", labelsize=7)
 matplotlib.rc("axes", titlesize=7)
 matplotlib.rc("legend", fontsize=6)
 
-from bokeh.palettes import Colorblind
-
+from bokeh.palettes import Colorblind, Category10
+from bokeh.palettes import YlGnBu, YlOrBr
+c1 = YlGnBu[8]
+c2 = YlOrBr[8]
 
 def piecewise_constant(
     nus,
@@ -268,7 +270,7 @@ tt_CEU, sd1_neu_CEU = get_sd1(nu_CEU, Ts, n, Ne, name="CEU")
 ## plots
 
 lw = 1
-colors = Colorblind[8]
+colors = Category10[10]
 
 fig = plt.figure(4, figsize=(6.5, 3.5))
 fig.clf()
@@ -283,7 +285,7 @@ plot_relate_curve(
     N0,
     line_style="-",
     lw=lw,
-    color=colors[0],
+    color=c1[0],
     label="YRI",
     gen=1,
 )
@@ -293,7 +295,7 @@ plot_relate_curve(
     N1,
     line_style="--",
     lw=lw,
-    color=colors[1],
+    color=c2[0],
     label="CEU",
     gen=1,
 )
@@ -310,86 +312,86 @@ axes[0].set_title("Size histories")
 print("plotting additivity panel")
 # gamma = -1
 t, v = get_sd1(nu_YRI, Ts, 50, Ne, gamma=-1.0, name="YRI")
-axes[1].plot(1e6 - gen_time * t, v, "-", lw=lw, color=colors[0])
+axes[1].plot(1e6 - gen_time * t, v, "-", lw=lw, color=c1[1])
 t, v = get_sd1(nu_CEU, Ts, 50, Ne, gamma=-1.0, name="CEU")
-axes[1].plot(1e6 - gen_time * t, v, "--", lw=lw, color=colors[1])
+axes[1].plot(1e6 - gen_time * t, v, "--", lw=lw, color=c2[1])
 axes[1].text(2e5, v[0] + 0.02, rf"$\gamma=-1$", va="center", ha="center", fontsize=7)
 
 # gamma = -10
 t, v = get_sd1(nu_YRI, Ts, 50, Ne, n_large=70, gamma=-10.0, name="YRI")
[…]

### 2022-06-03 figures/demog_YRI_CEU.sd2.py

@@ -17,8 +17,10 @@ matplotlib.rc("axes", labelsize=7)
 matplotlib.rc("axes", titlesize=7)
 matplotlib.rc("legend", fontsize=6)
 
-from bokeh.palettes import Colorblind
-
+from bokeh.palettes import Colorblind, Category10
+from bokeh.palettes import YlGnBu, YlOrBr
+c1 = YlGnBu[8]
+c2 = YlOrBr[8]
 
 def piecewise_constant(
     nus,
@@ -268,7 +270,7 @@ tt_CEU, sd2_neu_CEU = get_sd2(nu_CEU, Ts, n, Ne, name="CEU")
 ## plots
 
 lw = 1
-colors = Colorblind[8]
+colors = Category10[10]
 
 fig = plt.figure(4, figsize=(6.5, 3.5))
 fig.clf()
@@ -283,7 +285,7 @@ plot_relate_curve(
     N0,
     line_style="-",
     lw=lw,
-    color=colors[0],
+    color=c1[0],
     label="YRI",
     gen=1,
 )
@@ -293,7 +295,7 @@ plot_relate_curve(
     N1,
     line_style="--",
     lw=lw,
-    color=colors[1],
+    color=c2[0],
     label="CEU",
     gen=1,
 )
@@ -310,64 +312,67 @@ axes[0].set_title("Size histories")
 print("plotting additivity panel")
 # gamma = -1
 t, v = get_sd2(nu_YRI, Ts, 50, Ne, gamma=-1.0, name="YRI")
-axes[1].plot(1e6 - gen_time * t, v, "-", lw=lw, color=colors[0])
+axes[1].plot(1e6 - gen_time * t, v, "-", lw=lw, color=c1[1])
 t, v = get_sd2(nu_CEU, Ts, 50, Ne, gamma=-1.0, name="CEU")
-axes[1].plot(1e6 - gen_time * t, v, "--", lw=lw, color=colors[1])
-#axes[1].text(2e5, v[0] + 0.02, rf"$\gamma=-1$", va="center", ha="center", fontsize=7)
+axes[1].plot(1e6 - gen_time * t, v, "--", lw=lw, color=c2[1])
+axes[1].text(2e5, v[0] -0.1, rf"$\gamma=-1$", va="center", ha="center", fontsize=7)
 
 # gamma = -10
[…]

### 2022-06-03 figures/demog_bottle_expansion.dominance.py

@@ -25,7 +25,10 @@ matplotlib.rc("axes", labelsize=7)
 matplotlib.rc("axes", titlesize=7)
 matplotlib.rc("legend", fontsize=6)
 
-from bokeh.palettes import Colorblind
+from bokeh.palettes import Colorblind, Category10
+from bokeh.palettes import YlGnBu, YlOrBr
+c1 = YlGnBu[8]
+c2 = YlOrBr[8]
 
 
 def piecewise_constant(
@@ -293,7 +296,7 @@ tt_bottle = ts_gens(data["t"])
 ## plots
 
 lw = 1
-colors = Colorblind[8]
+colors = Category10[10]
 
 fig = plt.figure(4, figsize=(6.5, 3.5))
 fig.clf()
@@ -306,7 +309,7 @@ plot_relate_curve(
     N_expand,
     line_style="-",
     lw=lw,
-    color=colors[0],
+    color=c1[0],
     label="Expansion",
     gen=1,
 )
@@ -316,7 +319,7 @@ plot_relate_curve(
     N_bottle,
     line_style="--",
     lw=lw,
-    color=colors[1],
+    color=c2[0],
     label="Bottleneck",
     gen=1,
 )
@@ -333,20 +336,20 @@ axes[0].set_title("Size histories")
 print("plotting additivity panel")
 axes[1].plot(tt_expand, sd1_neu_expand, "k--", lw=0.5)
 t, v = get_sd1(nus_expand, Ts_expand, 50, gamma=-2.0, name="expand")
-axes[1].plot(t, v, "-", lw=lw, color=colors[0])
+axes[1].plot(t, v, "-", lw=lw, color=c1[1])
 t, v = get_sd1(nus_bottle, Ts_bottle, 50, gamma=-2.0, name="bottle")
-axes[1].plot(t, v, "--", lw=lw, color=colors[1])
+axes[1].plot(t, v, "--", lw=lw, color=c2[1])
 axes[1].text(1000, v[0] + 0.02, rf"$\gamma=-2$", va="center", ha="center", fontsize=7)
 
-axes[1].set_title("Interference")
+axes[1].set_title("Interference ($h=0.5$)")
 
 ## simple dominance
 print("plotting simple dominance panel")
[…]

### 2022-06-03 figures/demog_bottle_expansion.dominance.sd2.py

@@ -25,7 +25,10 @@ matplotlib.rc("axes", labelsize=7)
 matplotlib.rc("axes", titlesize=7)
 matplotlib.rc("legend", fontsize=6)
 
-from bokeh.palettes import Colorblind
+from bokeh.palettes import Colorblind, Category10
+from bokeh.palettes import YlGnBu, YlOrBr
+c1 = YlGnBu[8]
+c2 = YlOrBr[8]
 
 
 def piecewise_constant(
@@ -293,7 +296,7 @@ tt_bottle = ts_gens(data["t"])
 ## plots
 
 lw = 1
-colors = Colorblind[8]
+colors = Category10[10]
 
 fig = plt.figure(4, figsize=(6.5, 3.5))
 fig.clf()
@@ -306,7 +309,7 @@ plot_relate_curve(
     N_expand,
     line_style="-",
     lw=lw,
-    color=colors[0],
+    color=c1[0],
     label="Expansion",
     gen=1,
 )
@@ -316,7 +319,7 @@ plot_relate_curve(
     N_bottle,
     line_style="--",
     lw=lw,
-    color=colors[1],
+    color=c2[0],
     label="Bottleneck",
     gen=1,
 )
@@ -334,21 +337,21 @@ print("plotting additivity panel")
 axes[1].plot(tt_expand, sd2_neu_expand, "k--", lw=0.5)
 axes[1].plot(tt_bottle, sd2_neu_bottle, "k--", lw=0.5)
 t, v = get_sd2(nus_expand, Ts_expand, 50, gamma=-2.0, name="expand")
-axes[1].plot(t, v, "-", lw=lw, color=colors[0])
+axes[1].plot(t, v, "-", lw=lw, color=c1[1])
 t, v = get_sd2(nus_bottle, Ts_bottle, 50, gamma=-2.0, name="bottle")
-axes[1].plot(t, v, "--", lw=lw, color=colors[1])
+axes[1].plot(t, v, "--", lw=lw, color=c2[1])
 axes[1].text(1000, v[0] + 0.025, rf"$\gamma=-2$", va="center", ha="center", fontsize=7)
 
-axes[1].set_title("Interference")
+axes[1].set_title("Interference ($h=0.5$)")
 
 ## simple dominance
 print("plotting simple dominance panel")
[…]

### 2022-06-03 figures/demog_bottle_expansion.py

@@ -25,7 +25,10 @@ matplotlib.rc("axes", labelsize=7)
 matplotlib.rc("axes", titlesize=7)
 matplotlib.rc("legend", fontsize=6)
 
-from bokeh.palettes import Colorblind
+import bokeh.palettes
+from bokeh.palettes import YlGnBu, YlOrBr
+c1 = YlGnBu[8]
+c2 = YlOrBr[8]
 
 
 def piecewise_constant(
@@ -273,7 +276,8 @@ tt_bottle = ts_gens(data["t"])
 ## plots
 
 lw = 1
-colors = Colorblind[8]
+#colors = bokeh.palettes.Colorblind[8]
+colors = bokeh.palettes.Category10[10]
 
 fig = plt.figure(4, figsize=(6.5, 3.5))
 fig.clf()
@@ -286,7 +290,7 @@ plot_relate_curve(
     N_expand,
     line_style="-",
     lw=lw,
-    color=colors[0],
+    color=c1[0],
     label="Expansion",
     gen=1,
 )
@@ -296,7 +300,7 @@ plot_relate_curve(
     N_bottle,
     line_style="--",
     lw=lw,
-    color=colors[1],
+    color=c2[0],
     label="Bottleneck",
     gen=1,
 )
@@ -313,72 +317,72 @@ axes[0].set_title("Size histories")
 print("plotting additivity panel")
 # gamma = -1
 t, v = get_sd1(nus_expand, Ts_expand, 50, gamma=-1.0, name="expand")
-axes[1].plot(t, v, "-", lw=lw, color=colors[0])
+axes[1].plot(t, v, "-", lw=lw, color=c1[1])
 t, v = get_sd1(nus_bottle, Ts_bottle, 50, gamma=-1.0, name="bottle")
-axes[1].plot(t, v, "--", lw=lw, color=colors[1])
+axes[1].plot(t, v, "--", lw=lw, color=c2[1])
 axes[1].text(1000, v[0] + 0.02, rf"$\gamma=-1$", va="center", ha="center", fontsize=7)
 
 # gamma = -10
 t, v = get_sd1(nus_expand, Ts_expand, 50, n_large=70, gamma=-10.0, name="expand")
-axes[1].plot(t, v, "-", lw=lw, color=colors[0])
+axes[1].plot(t, v, "-", lw=lw, color=c1[1])
[…]

### 2022-06-03 figures/demog_bottle_expansion.sd2.py

@@ -25,7 +25,10 @@ matplotlib.rc("axes", labelsize=7)
 matplotlib.rc("axes", titlesize=7)
 matplotlib.rc("legend", fontsize=6)
 
-from bokeh.palettes import Colorblind
+from bokeh.palettes import Colorblind, Category10
+from bokeh.palettes import YlGnBu, YlOrBr
+c1 = YlGnBu[8]
+c2 = YlOrBr[8]
 
 
 def piecewise_constant(
@@ -273,7 +276,7 @@ tt_bottle = ts_gens(data["t"])
 ## plots
 
 lw = 1
-colors = Colorblind[8]
+colors = Category10[10]
 
 fig = plt.figure(4, figsize=(6.5, 3.5))
 fig.clf()
@@ -286,7 +289,7 @@ plot_relate_curve(
     N_expand,
     line_style="-",
     lw=lw,
-    color=colors[0],
+    color=c1[0],
     label="Expansion",
     gen=1,
 )
@@ -296,7 +299,7 @@ plot_relate_curve(
     N_bottle,
     line_style="--",
     lw=lw,
-    color=colors[1],
+    color=c2[0],
     label="Bottleneck",
     gen=1,
 )
@@ -313,75 +316,75 @@ axes[0].set_title("Size histories")
 print("plotting additivity panel")
 # gamma = -1
 t, v = get_sd2(nus_expand, Ts_expand, 50, gamma=-1.0, name="expand")
-axes[1].plot(t, v, "-", lw=lw, color=colors[0])
+axes[1].plot(t, v, "-", lw=lw, color=c1[1])
 t, v = get_sd2(nus_bottle, Ts_bottle, 50, gamma=-1.0, name="bottle")
-axes[1].plot(t, v, "--", lw=lw, color=colors[1])
+axes[1].plot(t, v, "--", lw=lw, color=c2[1])
 axes[1].text(800, v[0] - 0.06, rf"$\gamma=-1$", va="center", ha="center", fontsize=7)
 
 # gamma = -10
 t, v = get_sd2(nus_expand, Ts_expand, 50, n_large=70, gamma=-10.0, name="expand")
-axes[1].plot(t, v, "-", lw=lw, color=colors[0])
+axes[1].plot(t, v, "-", lw=lw, color=c1[1])
[…]

### 2022-06-03 figures/dominance_vary_gamma.py

@@ -1,0 +1,57 @@
+import moments.TwoLocus
+import matplotlib.pylab as plt
+import numpy as np
+import pickle
+import gzip
+
+# set font sizes
+import matplotlib
+
+plt.rcParams["legend.title_fontsize"] = "xx-small"
+matplotlib.rc("xtick", labelsize=6)
+matplotlib.rc("ytick", labelsize=6)
+matplotlib.rc("axes", labelsize=7)
+matplotlib.rc("axes", titlesize=7)
+matplotlib.rc("legend", fontsize=6)
+
+from bokeh.palettes import Colorblind
+
+colors = Colorblind[8]
+
+gammas = -np.concatenate((np.logspace(-1, 1, 21), [15.0, 20.0]))
+
+rho = 0.0
+hs = [0.05, 0.25, 0.45]
+
+data = {}
+for h in hs:
+    print("h =", h)
+    data[h] = []
+    for gamma in gammas:
+        print("gamma =", gamma, end="")
+        if -gamma < 7.5:
+            n = 50
+        else:
+            n = 80
+        sel_params = moments.TwoLocus.Util.simple_dominance(gamma, h=h)
+        F = moments.TwoLocus.Demographics.equilibrium(
+            n, sel_params_general=sel_params, rho=rho
+        )
+        data[h].append(F)
+        print(", min=", F.min())
+    print()
+
+sd1s = {}
+for h in hs:
+    print(h)
+    sd1s[h] = []
+    for F in data[h]:
+        sd1s[h].append(F.D() / F.pi2())
+    print()
+
+import pickle, gzip
+
+pickle.dump(
+    {"spectra": data, "sd1": sd1s, "gammas": gammas},
+    gzip.open("data/dominance_vary_gamma.bp.gz", "wb+"),
+)

### 2022-06-03 figures/frequency_spectra.py

@@ -1,0 +1,102 @@
+import moments
+import pickle
+import numpy as np
+import matplotlib.pylab as plt
+
+populations = {
+    "Africa": ["YRI"],  # , "GWD", "LWK", "MSL", "YRI"],
+    "Europe": ["CEU"],  # , "FIN", "GBR", "IBS", "TSI"],
+    "East Asia": ["CHB"],  # , "CHB", "CHS", "JPT", "KHV"],
+}
+
+import bokeh.palettes
+
+colors = bokeh.palettes.Paired[8]
+
+
+def plot_mis_syn(ax, mis, syn, title=None, xlabel=False, ylabel=False, legend=False):
+    mis = mis.fold()
+    syn = syn.fold()
+
+    ax.plot(mis / mis.S(), ".-", color=colors[6], ms=2, lw=1, label="Missense (all)")
+    ax.plot(syn / syn.S(), ".--", color=colors[0], ms=2, lw=1, label="Synonymous (all)")
+
+    mis.mask[: int(mis.sample_sizes * 0.1)] = True
+    syn.mask[: int(mis.sample_sizes * 0.1)] = True
+    ax.plot(mis / mis.S(), ".-", color=colors[7], ms=2, lw=1, label="Missense (>10%)")
+    ax.plot(
+        syn / syn.S(), ".--", color=colors[1], ms=2, lw=1, label="Synonymous (>10%)"
+    )
+
+    if legend:
+        ax.legend(fontsize=6)
+    ax.set_yscale("log")
+    if ylabel:
+        ax.set_ylabel("Proportion")
+    if xlabel:
+        ax.set_xlabel("Minor allele frequency")
+    ax.set_title(title)
+
+
+fig = plt.figure(figsize=(8, 8))
+
+i = 0
+
+for c, ps in populations.items():
+    for p in ps:
+        fname = f"../analysis/parsed_data/{p}.frequency_spectra.bp"
+        data = pickle.load(open(fname, "rb"))
+
+        i += 1
+        ax = plt.subplot(3, 3, i)
+        if i > 6:
+            xlabel = True
+        else:
+            xlabel = False
+        if i == 1:
+            legend = True
+        else:
+            legend = False
+
+        # all
+        mis = moments.Spectrum(data["missense"]["all"])
+        syn = moments.Spectrum(data["synonymous"]["all"])
+        plot_mis_syn(
+            ax,
+            mis,
+            syn,
+            title=f"{p}, all mutations",
[…]
